### 2012-03-16 eQTL_2D/gpu_proposal.tex
Added two new projects - vitD stuff, and 2D scans on expression data

\documentclass{scrartcl}

\usepackage{hyperref}
\hypersetup{
linkcolor=blue}

\title{Using GPUs to explore the genetic architecture of complex traits}
\subtitle{Searching for functional interactions between mutations in gene expression}
\author{Gibran Hemani, Joseph Powell, Peter Visscher}
\date{}

\begin{document}
\maketitle

\paragraph{Introduction}

Phenotypic variation arises partly because of environmental influences, and partly due to genetic differences. Most traits of interest, for example common human diseases, are caused by a complex pattern of both. The human genome is a code consisting of 3 billion base pairs and from individual to individual most of these are identical, differing on average once every thousand bases. A very hot topic in genetics now is trying to map these polymorphisms to phenotypic variation.

A central question relating to this search is how exactly do mutations affect the phenotype? Most studies assume that each mutation acts independently from all others, and this leads to a computationally straightforward task of testing for association at each polymorphism in the genome with the phenotype. But theory provides an alternative hypothesis such that the effect of each causal mutation actually depends on which mutations are present elsewhere in the genome. Such genetic interactions (epistasis) are, in principle, much more difficult to identify than independent mutations because the problem passes into a higher dimensional parameter space. For this reason, although there is theoretical support for the importance of epistasis, it is seldom searched for empirically, even though the data to do so is abundantly available.

If we can understand how mutations are acting - whether that be independently or as part of complex interactions - then this will be useful for many aspects of biology and medicine including the mapping of causal mutations to phenotypes, developing more accurate disease risk predictors, and understanding the mechanisms that underlie evolution.

\paragraph{Methods}

We recently developed software that searches for epistasis exhaustively by parallelising the analysis on GPUs, using the OpenCL API {\tt epiGPU}. It has increased the search speed by more than 3 orders of magnitude, thus making this type of analysis tractable.

We have a rich dataset of RNA expression profiles for around 1000 individuals. We propose to search for interactions between genetic mutations in these individuals that might account for the differences in the expression of each of these RNA transcripts. In total, there are approximately 15000 RNA transcripts, and each transcript will take approximately 2 hours to analyse using {\tt epiGPU}, so access to a GPU cluster would make such a study possible.

\paragraph{Outcomes}

This type of analysis has never been done before, and it would provide an extremely thorough survey of the way in which mutations contribute towards complex traits. If this analysis fails to find much epistasis then this will be an important result in itself - providing empirical evidence in a continuing debate about the architecture of genetic variation. On the other hand if we discover that a large number of the polymorphisms associated with the trait do not act independently and are indeed involved in complex interactions then this will be the first empirical evidence for widespread epistasis.

\end{document}

### 2012-06-20 eQTL_2D/presentation/icqg4/icqg4_hemani.tex
website and icqg4 presentation

\documentclass{beamer}
%\usepackage{colortbl}
\usepackage[latin1]{inputenc}
\usepackage{comment}
%\usetheme{Warsaw}
\usetheme{Frankfurt}
%\usecolortheme{dove}
\usefonttheme{serif}

\usepackage[absolute,overlay]{textpos} 
\usepackage{booktabs}
\usepackage{color}
\usepackage{threeparttable}
\usepackage{multirow}
%\usepackage[table]{xcolor}
%\definecolor{tableShade}{HTML}{F1F5FA}
\usepackage[normalem]{ulem}

\title[Epistasis in GWAS]{2D or not 2D, that is the question}
\author{Gibran Hemani}
\institute{The Roslin Institute, University of Edinburgh \\
\vspace{.3cm}
Diamantina Institute \\and \\
Queensland Brain Institute,\\
University of Queensland}
\date{}
\begin{document}

\AtBeginSection[]
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[currentsection,currentsubsection]
  \end{frame}
}

\begin{frame}
\titlepage
\end{frame}

\begin{comment}

____________intro

talk about gwas design from an evolutionary perspective
- interesting paradox in quan genetic theory

what is a gwas?
mixed success. missing heritability

additive genetic variance paradigm
- additive effects
- heritability estimates (problems with mz and dz twin studies? occam's razor?)
- fitness vs morphological traits (response to selection, maintenance of additive variance)

epistasis
- define epistasis
- how to search (two stages, two dimensions, mention epiGPU)

can epistasis maintain additive variance?
- yes.
- it actually maintains non-additive variation much more
- how do we search for epistasis?

____________conclusions

is the observation of additive variance an illusion?
have we misused occam's razor?
here is the paradox:
if you can measure additive genetic variance, search for non-additive variance

____________acknowledgements

\end{comment}

\begin{frame}{}
\begin{itemize}
\item Is there theoretical reason to presume epistasis might play an important role in complex traits?
\item Is it computationally tractable in GWAS?
\item Is it statistically tractable in GWAS?
\item Anything detected empirically?
\end{itemize}
\end{frame}

\section{Polymorphisms under selection}
\subsection{}

\begin{frame}{Epistasis}
\begin{definition}
{\color{orange} The effect on the phenotype caused by locus A depends on the genotype at locus B }
%{\tiny \color{white} \newline - Carlborg and Haley 2004 }
\end{definition}
\end{frame}

\begin{frame}{Epistasis and additive variance}
\begin{center}
\includegraphics[height=7cm]{gp_va}
\end{center}
\end{frame}

\begin{frame}{}
\begin{itemize}
\item \textbf{Hypothesis:} Epistasis can maintain additive variance better than purely additive effects
\item How can maintained additive variance be detected most effectively?
\item What is the impact of low LD between causal variants and observed SNPs?
\end{itemize}
\end{frame}

\begin{frame}{Epistasis maintains common genetic variation}
\begin{center}
\includegraphics[width=10cm]{maps_freqs}
\end{center}
\end{frame}

\begin{frame}{Changes in variance under selection}
\begin{center}
\includegraphics[width=8cm]{sup_propadditive_det.png}
\end{center}
\end{frame}

\begin{frame}{Power to detect evolutionarily persistant variants}

\begin{columns}[c]

\column{.7\textwidth}

\includegraphics[width=8cm]{sup_heritbars_sim.png}

\column{.3\textwidth}

When LD between causal variants and observed SNPs is high, 2D scan is most powerful for 46/54 non-additive patterns

\end{columns}
\end{frame}

\begin{frame}{The paradox}
\begin{columns}[c]
\column{.5\textwidth}
\begin{itemize}
\item Epistasis can contribute towards the maintenance and observation of additive genetic variance
\item To find the missing additive variance, search for epistatic effects
\end{itemize}
\column{.5\textwidth} 
\includegraphics[height=6.5cm]{iceberg.jpg} \\
\end{columns}
\end{frame}

\section{Computational solutions}
\subsection{}

\begin{frame}{The problem}
Number of interactions: \\
\begin{equation}
n(n-1) / 2 \nonumber
\end{equation}
500000 SNPs $\rightarrow$ 125 billion interactions \\
\begin{itemize}
\item PLINK performs 5000 tests per second $\rightarrow$ 3 years
\item FastEpistasis performs 36000 tests per second $\rightarrow$ 5 months
\item Optimised C code 125000 tests per second $\rightarrow$ 6 weeks
\item Parallelised on 8-core CPU $\rightarrow$ 5 days
\end{itemize}
\end{frame}

\begin{frame}{CPU limits}
\begin{center}
\includegraphics[height=7cm]{clockspeed.pdf}
\end{center}
\end{frame}

\begin{frame}{Graphics Cards}
\begin{itemize}
\item Revenue of video game industry $>$ Hollywood film industry
\item Better games require better hardware
\begin{itemize}
\item Finer pixels rendered faster
\item Physics engines
\item Easily obtained by having many simple cores
\end{itemize}
\item Consumer level graphics cards have hundreds of very simple cores
\item Ideal for geometric parallelisation problems in science
\item Use OpenCL API to communicate with graphics cards
\end{itemize}
\end{frame}

\begin{frame}{GPU speeds}
\begin{center}
\includegraphics[height=7cm]{gpuoptimisation.pdf}
\end{center}
\end{frame}

\begin{frame}{epiGPU}
\begin{itemize}
\item http://sourceforge.net/projects/epigpu/
\item Free
\item Open source
\item Windows, Mac, Linux
\item NVIDIA and ATI graphics cards
\end{itemize}
\end{frame}

\begin{frame}{Other epistasis software}
\begin{itemize}
\item BiForce
\item GLIDE
\item GBOOST
\item TEAM
\item MDR-GPU
\item EPIBLASTER
\item SHEsisEPI
\item ...probably more
\end{itemize}
\end{frame}

\section{Thresholds}
\subsection{}

\begin{frame}{The curse of dimensionality}
As the dimensionality of the search increases the background noise drowns out all real biological signals
\end{frame}

\begin{frame}{Linkage disequilibrium}
\begin{center}
\includegraphics[width=10cm]{ld.png}
\end{center}
\begin{itemize}
\item SNPs are correlated $\rightarrow$ tests are not independent
\item Bonferroni correction is overly stringent
\end{itemize}
\end{frame}

\begin{frame}{}
\begin{center}
What is the correlation structure in a 2D search?
\end{center}
\end{frame}

\begin{frame}{Permutation analysis}
What is the \emph{effective} number of multiple tests that are being performed? (Churchill 1994)
\begin{center}
\includegraphics[width=10cm]{permutationanalysis.pdf}
\end{center}
\end{frame}

\begin{frame}{Epistatic permutations}
\begin{itemize}
\item How is the 2D threshold governed by:
\begin{itemize}
\item Number of SNPs being tested
\item Number of individuals in the study?
\end{itemize}
\item Difficult computational challenge
\begin{itemize}
\item Requires cluster of GPUs...
\item ...running for months!
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{A general formula for 2D thresholds?}
\begin{equation}
-\log_{10}(p_{T}) = a - \frac{a}{\beta_{1} N + \beta_{2} M} \nonumber
\label{eqn:general_threshold}
\end{equation}
\begin{eqnarray}
a & = & -\log_{10} \left (\frac{0.05}{t^2/2} \right) = 12.68 \nonumber
\end{eqnarray}
where $t = 693138$ is the number of independent regions in the genome (Dudbridge \& Gusnanto 2008)
\end{frame}

\begin{frame}{A general formula for 2D thresholds?}
\begin{center}
\includegraphics[height=7cm]{threshold_function.pdf}
\end{center}
\end{frame}

\section{2D eQTL}
\subsection{}

\begin{frame}{Impact of LD on observed variance}
\begin{center}
\includegraphics[width=5cm]{gpmaps_ld.png} \\
\end{center}
\end{frame}

\begin{frame}{Are epistatic effects controlling gene expression?}
\begin{itemize}
\item Study design:
\begin{itemize}
\item 846 healthy Australian adults
\item 546,192 SNPs
\item 5380 gene expression probes from whole blood
\end{itemize}
\item Plan:
\begin{itemize}
\item Do exhaustive search for 2 locus epistasis for each probe
\item Parallelise search across 100x GPU cluster
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Thresholds}
\begin{itemize}
\item 1641 complete scans were performed on permuted phenotype
\item Threshold predicted to be: \textbf{11.648}
\item Empirical estimate: \textbf{11.632}
\item Average correlation between 5380 probes: \textbf{0.265}
\item Overall threshold: $-\log_{10}p =$ \textbf{16.5}
\end{itemize}
\end{frame}

\begin{frame}{Very preliminary results}
\begin{itemize}
\item Is there much non-additive variance?
\item 71 probes have significant eQTLs after filtering for:
\begin{itemize}
\item Threshold of 16.5
\item Additive variance $\leq$ 60\% of total genetic variance
\item Relatively common SNPs (must have samples for all 9 genotype classes)
\item No LD between interacting SNPs
\end{itemize}
\item Total of 140 significant independent loci
\end{itemize}
\end{frame}

\begin{frame}{Very preliminary results}
\begin{itemize}
\item Proportion of the phenotypic variance explained:
\begin{itemize}
\item Total genetic: range 8.4\% - 17.1\%
\item Total non-additive: range 3.5\% - 9.2\%
\end{itemize}
\item 14\% cis-cis
\item 69\% cis-trans
\item 17\% trans-trans
\end{itemize}
\end{frame}

\begin{frame}{Some examples}
\begin{center}
\includegraphics[height=7.5cm]{epistasis_examples}
\end{center}
\end{frame}

\begin{frame}{Vandalising poetry}
\emph{\sout{To be} \textcolor{red}{2D} or not \sout{to be} \textcolor{red}{2D}, that is the question \\
Whether 'tis Nobler in the mind to suffer \\
The Slings and Arrows of outrageous Fortune, \\
Or to take Arms against a Sea of troubles} \\
\hspace{1cm}{\tiny - William Shakespeare, 1623} \\
\vspace{0.5in}
\pause
\emph{\sout{To be} \textcolor{red}{2D} or not \sout{to be} \textcolor{red}{2D} \\
is not my dilemma, \\
To break away from both worlds is not bravery} \\
\hspace{1cm}{\tiny - Rumi, 13th century} \\
\end{frame}

\section*{}

\begin{frame}{Acknowledgements}

\begin{columns}[c]

\column{.5\textwidth}

{\tiny
PhD Funding:
\begin{itemize}
\item The Roslin Institute
\item University of Edinburgh
\item Biosciences KTN
\item Monsanto / NCG
\end{itemize}
PhD Supervisors:
\begin{itemize}
\item Chris Haley
\item Sara Knott
\item John Woolliams
\end{itemize}
Important people:
\begin{itemize}
\item Ricardo Pong-Wong
\item Athanasios Theocharidis
\item Wenhua Wei
\end{itemize}
Compute resources
\begin{itemize}
\item ECDF - {\tt eddie}
\item Daresbury Labs - {\tt cseht}
\item IVEC - {\tt fornax}
\end{itemize}
}

\column{.5\textwidth}

{\tiny
Complex Trait Genomics Group

\begin{itemize}
\item Peter Visscher
\item Naomi Wray
\item Joseph Powell
\item Jian Yang
\item Allan Mcrae
\item Anita Goldinger
\item Hong Lee
\item Anna Vinkhuyzen
\item Guo-Bo Chen
\item Beben Benyamin
\item Zong Zhang
\item Enda Byrne
\item Marie-Jo Brion
\item Sven Stringer
\end{itemize}
}

\end{columns}
\end{frame}

\begin{frame}
\centering
Thanks!
\end{frame}

\begin{frame}{Sample size vs 2D threshold}

\begin{equation}
\lim_{n \to \infty} \bar{g} = 9 \nonumber
\end{equation}

\begin{eqnarray}
-\log \left ( f \left (
\frac{
 SS_{W}  (g - 1)^{-1}
}
{  SS_{B} (n - g)^{-1}
}; g - 1, n - g
\right ) \right )
& \propto &
 n \nonumber \\ 
& \propto &
SS_{W} / (SS_{W} + SS_{B}) \nonumber \\
& \propto &
-\log(g) \nonumber
\label{eq:ftest_n}
\end{eqnarray}

\end{frame}

\begin{frame}{GP maps from genetic algorithm}
\begin{center}
\includegraphics[width=8cm]{gpmaps2.pdf}
\end{center}
\end{frame}

\begin{frame}{Allele frequency trajectories}
\begin{center}
\includegraphics[width=8cm]{allelefreq_det.png}
\end{center}
\end{frame}

\begin{frame}{Changes in variance under selection}
\begin{center}
\includegraphics[width=10cm]{propadditive_det_grey.pdf} \\
\includegraphics[width=10cm]{Vg_det_grey.pdf}
\end{center}
\end{frame}

\begin{frame}{Power studies}
\begin{center}
\includegraphics[width=10cm]{powersimple.pdf} \\
\end{center}
\end{frame}

\begin{frame}{Epistatic patterns}
\begin{center}
\includegraphics[width=8cm]{sup_gpmaps.pdf}
\end{center}
\end{frame}

\begin{frame}{Allele frequency trajectories}
\begin{center}
\includegraphics[width=8cm]{sup_allelefreq_sim.png}
\end{center}
\end{frame}

\end{document}

### 2012-06-21 eQTL_2D/presentation/icqg4/icqg4_hemani.tex
icgq4 talk finished

@@
 
-\title[Epistasis in GWAS]{2D or not 2D, that is the question}
+\title[Epistasis in GWAS]{2D or not 2D, that is the question\\
+{\footnotesize Should we be looking for epistasis in GWAS?}}
 \author{Gibran Hemani}
@@
 \begin{itemize}
-\item Is there theoretical reason to presume epistasis might play an important role in complex traits?
-\item Is it computationally tractable in GWAS?
-\item Is it statistically tractable in GWAS?
-\item Anything detected empirically?
+\item \textbf{Simulations:} An evolutionary perspective on genetic variation
+\item \textbf{Computation:} Making epistatic searches possible
+\item \textbf{Thresholds:} What should we call `significant'?
+\item \textbf{Data:} What do we find empirically?
 \end{itemize}
@@
 
-\section{Polymorphisms under selection}
+\section{Evolutionary simulations}
 \subsection{}
@@
 \begin{frame}{}
-\begin{itemize}
-\item \textbf{Hypothesis:} Epistasis can maintain additive variance better than purely additive effects
-\item How can maintained additive variance be detected most effectively?
-\item What is the impact of low LD between causal variants and observed SNPs?
-\end{itemize}
+\textbf{Hypothesis:}\\ Epistasis can maintain additive variance better than purely additive effects
 \end{frame}
 
-\begin{frame}{Epistasis maintains common genetic variation}
+\begin{frame}{Epistasis maintains genetic variation}
 \begin{center}
-\includegraphics[width=10cm]{maps_freqs}
+\includegraphics[width=11cm]{maps_freqs}
 \end{center}
@@
 
-\begin{frame}{Changes in variance under selection}
+\begin{frame}{Most of the variance is non-additive}
 \begin{center}
@@
 
-\begin{frame}{Power to detect evolutionarily persistant variants}
+\begin{frame}{Impact of LD on observed variance}
+\begin{center}
+\includegraphics[width=5cm]{gpmaps_ld.png} \\
+\end{center}
+\end{frame}
 
+\begin{frame}{With high LD, 2D searches are most powerful}
 \begin{columns}[c]
-
 \column{.7\textwidth}
-
 \includegraphics[width=8cm]{sup_heritbars_sim.png}
-
 \column{.3\textwidth}
-
 When LD between causal variants and observed SNPs is high, 2D scan is most powerful for 46/54 non-additive patterns
-
 \end{columns}
@@
 \begin{itemize}
-\item PLINK performs 5000 tests per second $\rightarrow$ 3 years
+\item For 1000 individuals, PLINK performs 5000 tests per second $\rightarrow$ 3 years
 \item FastEpistasis performs 36000 tests per second $\rightarrow$ 5 months
@@
 \begin{frame}{Graphics Cards}
-\begin{itemize}
-\item Revenue of video game industry $>$ Hollywood film industry
-\item Better games require better hardware
-\begin{itemize}
-\item Finer pixels rendered faster
-\item Physics engines
-\item Easily obtained by having many simple cores
-\end{itemize}
-\item Consumer level graphics cards have hundreds of very simple cores
-\item Ideal for geometric parallelisation problems in science
-\item Use OpenCL API to communicate with graphics cards
-\end{itemize}
+\begin{center}
+\includegraphics[width=10cm]{gtx580}
+\end{center}
 \end{frame}
@@
 
-\begin{frame}{Other epistasis software}
+\begin{frame}{Other optimised epistasis software}
 \begin{itemize}
@@
 \item Bonferroni correction is overly stringent
+\item What is the correlation structure in a 2D search?
 \end{itemize}
-\end{frame}
-
-\begin{frame}{}
-\begin{center}
-What is the correlation structure in a 2D search?
-\end{center}
 \end{frame}
@@
 
-\begin{frame}{Epistatic permutations}
-\begin{itemize}
-\item How is the 2D threshold governed by:
-\begin{itemize}
-\item Number of SNPs being tested
-\item Number of individuals in the study?
-\end{itemize}
-\item Difficult computational challenge
-\begin{itemize}
-\item Requires cluster of GPUs...
-\item ...running for months!
-\end{itemize}
-\end{itemize}
-\end{frame}
-
 \begin{frame}{A general formula for 2D thresholds?}
 \begin{equation}
--\log_{10}(p_{T}) = a - \frac{a}{\beta_{1} N + \beta_{2} M} \nonumber
+-\log_{10}(p_{T}) = a - \frac{a}{f_{1}(N)+ f_{2}(M)} \nonumber
 \label{eqn:general_threshold}
@@
 \subsection{}
-
-\begin{frame}{Impact of LD on observed variance}
-\begin{center}
-\includegraphics[width=5cm]{gpmaps_ld.png} \\
-\end{center}
-\end{frame}
 
@@
 
-\begin{frame}{Vandalising poetry}
-\emph{\sout{To be} \textcolor{red}{2D} or not \sout{to be} \textcolor{red}{2D}, that is the question \\
-Whether 'tis Nobler in the mind to suffer \\
-The Slings and Arrows of outrageous Fortune, \\
-Or to take Arms against a Sea of troubles} \\
-\hspace{1cm}{\tiny - William Shakespeare, 1623} \\
-\vspace{0.5in}
-\pause
-\emph{\sout{To be} \textcolor{red}{2D} or not \sout{to be} \textcolor{red}{2D} \\
-is not my dilemma, \\
-To break away from both worlds is not bravery} \\
-\hspace{1cm}{\tiny - Rumi, 13th century} \\
-\end{frame}
-
 \section*{}
@@
 \item Biosciences KTN
-\item Monsanto / NCG
+\item Monsanto / Newsham Choice Genetics
 \end{itemize}

### 2012-06-21 eQTL_2D/presentation/icqg4/icqg4_hemani.tex
icgq4 talk finished

@@
 {\footnotesize Should we be looking for epistasis in GWAS?}}
-\author{Gibran Hemani}
+\author{Gibran Hemani, Sara Knott, Joseph Powell, Naomi Wray, Peter Visscher, Chris Haley}
 \institute{The Roslin Institute, University of Edinburgh \\
@@
 \end{itemize}
-\item Total of 140 significant independent loci
+\item Number of significant epistatic pairs: \textbf{140}
 \end{itemize}
@@
 \item IVEC - {\tt fornax}
+\item QBI cluster
 \end{itemize}
@@
 \item Sven Stringer
+\item Visit us at www.complextraitgenomics.com
 \end{itemize}
@@
 
+\begin{frame}{Narrow-sense heritability}
+\begin{figure}
+% \begin{minipage}[c]{0.38\textwidth}
+\includegraphics[width=6cm]{visscher2008_heritabiliy.jpg} \\
+{\tiny Visscher, 2008}
+\end{figure}
+\begin{itemize}
+\item Fitness related traits have lower heritabilites
+\item Additive variance appears to be maintained under selection
+\item Response to selection is lower than expected for fitness related traits
+\end{itemize}
+\end{frame}
+
 \begin{frame}{Allele frequency trajectories}

### 2012-10-13 eQTL_2D/presentation/20121012_fridayforum/20121012_fridayforum_hemani.tex
friday forum talk on 2D eQTL

--- 2012-06-21 eQTL_2D/presentation/icqg4/icqg4_hemani.tex
@@
 %\usecolortheme{dove}
-\usefonttheme{serif}
+\usefonttheme{professionalfonts}
 
@@
 
-\title[Epistasis in GWAS]{2D or not 2D, that is the question\\
-{\footnotesize Should we be looking for epistasis in GWAS?}}
-\author{Gibran Hemani, Sara Knott, Joseph Powell, Naomi Wray, Peter Visscher, Chris Haley}
-\institute{The Roslin Institute, University of Edinburgh \\
-\vspace{.3cm}
-Diamantina Institute \\and \\
-Queensland Brain Institute,\\
-University of Queensland}
-\date{}
+\title[Epistasis and eQTL]{Epistasis in eQTL studies}
+\author{Gibran Hemani}
+\date{\today}
 \begin{document}
@@
 
-talk about gwas design from an evolutionary perspective
-- interesting paradox in quan genetic theory
-
 what is a gwas?
 mixed success. missing heritability
-
-additive genetic variance paradigm
-- additive effects
-- heritability estimates (problems with mz and dz twin studies? occam's razor?)
-- fitness vs morphological traits (response to selection, maintenance of additive variance)
+endophenotypes
 
@@
 
-\begin{frame}{}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Intro to epistasis}
+\subsection{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Genome wide association studies}
+	\begin{center}
+		\includegraphics[height=7cm]{gwas.jpg} \\
+		{\tiny Wang 2009}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Additive variance paradigm}
+	\begin{equation}
+		V_{phenotype} = V_{{Additive} {Genetic}} + V_{{Everything} {Else}} \nonumber
+	\end{equation}
 	\begin{itemize}
-\item \textbf{Simulations:} An evolutionary perspective on genetic variation
-\item \textbf{Computation:} Making epistatic searches possible
-\item \textbf{Thresholds:} What should we call `significant'?
-\item \textbf{Data:} What do we find empirically?
+		\item Discover common mutations that affect phenotypes
+		\item Uncover genes and structural elements involved in disease
+		\item Understand the architecture of natural genetic variation
 	\end{itemize}
@@
 
-\section{Evolutionary simulations}
-\subsection{}
+\begin{frame}{Endophenotypes}
+	\begin{center}
+		\includegraphics[height=7cm]{broadgp.png}
+	\end{center}
+\end{frame}
 
 \begin{frame}{Epistasis}
+	\begin{center}
+		\includegraphics[width=8cm]{gpmaps2.pdf}
+	\end{center}
 	\begin{definition}
-{\color{orange} The effect on the phenotype caused by locus A depends on the genotype at locus B }
-%{\tiny \color{white} \newline - Carlborg and Haley 2004 }
+		The effect on the phenotype caused by locus A depends on the genotype at locus B
 	\end{definition}
@@
 
-\begin{frame}{Epistasis and additive variance}
+\begin{frame}{2D GWAS}
 	\begin{center}
-\includegraphics[height=7cm]{gp_va}
+		\includegraphics[height=7cm]{2dscan.png}
 	\end{center}
@@
 
-\begin{frame}{}
-\textbf{Hypothesis:}\\ Epistasis can maintain additive variance better than purely additive effects
-\end{frame}
-
-\begin{frame}{Epistasis maintains genetic variation}
-\begin{center}
-\includegraphics[width=11cm]{maps_freqs}
-\end{center}
-\end{frame}
-
-\begin{frame}{Most of the variance is non-additive}
-\begin{center}
-\includegraphics[width=8cm]{sup_propadditive_det.png}
-\end{center}
-\end{frame}
-
-\begin{frame}{Impact of LD on observed variance}
-\begin{center}
-\includegraphics[width=5cm]{gpmaps_ld.png} \\
-\end{center}
-\end{frame}
-
-\begin{frame}{With high LD, 2D searches are most powerful}
-\begin{columns}[c]
-\column{.7\textwidth}
-\includegraphics[width=8cm]{sup_heritbars_sim.png}
-\column{.3\textwidth}
-When LD between causal variants and observed SNPs is high, 2D scan is most powerful for 46/54 non-additive patterns
-\end{columns}
-\end{frame}
-
-\begin{frame}{The paradox}
-\begin{columns}[c]
-\column{.5\textwidth}
-\begin{itemize}
-\item Epistasis can contribute towards the maintenance and observation of additive genetic variance
-\item To find the missing additive variance, search for epistatic effects
-\end{itemize}
-\column{.5\textwidth} 
-\includegraphics[height=6.5cm]{iceberg.jpg} \\
-\end{columns}
-\end{frame}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{frame}{The problem}
-Number of interactions: \\
+	\begin{block}{Number of interactions}
 		\begin{equation}
@@
 		\end{equation}
-500000 SNPs $\rightarrow$ 125 billion interactions \\
-\begin{itemize}
-\item For 1000 individuals, PLINK performs 5000 tests per second $\rightarrow$ 3 years
-\item FastEpistasis performs 36000 tests per second $\rightarrow$ 5 months
-\item Optimised C code 125000 tests per second $\rightarrow$ 6 weeks
-\item Parallelised on 8-core CPU $\rightarrow$ 5 days
-\end{itemize}
+		500000 SNPs $\rightarrow$ 125 billion interactions
+	\end{block}
+
+	\begin{table}[ht]
+		\begin{center}
+			\begin{tabular}{lrl}
+				\hline
+				Software & Tests per second & Run time \\ 
+				\hline
+				PLINK & 5000 & 3 years \\ 
+				Optimised CPU code & 125000 & 6 weeks \\ 
+				Parallelised on 8-core CPU & 1000000 & 5 days \\ 
+			   \hline
+			\end{tabular}
+		\end{center}
+	\end{table}
 \end{frame}
@@
 		\includegraphics[height=7cm]{clockspeed.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Failed targets}
+	\begin{center}
+		\includegraphics[height=7cm]{clockrates.png}
 	\end{center}
@@
 \begin{frame}{epiGPU}	
+	\begin{block}{Download}
+		{\tt sourceforge.net/projects/epigpu/}
+	\end{block}
 	\begin{itemize}
-\item http://sourceforge.net/projects/epigpu/
 		\item Free
@@
 
-\begin{frame}{Other optimised epistasis software}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{2D eQTL}
+\subsection{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{Are epistatic effects controlling gene expression?}
+	\begin{block}{Study design}
 		\begin{itemize}
-\item BiForce
-\item GLIDE
-\item GBOOST
-\item TEAM
-\item MDR-GPU
-\item EPIBLASTER
-\item SHEsisEPI
-\item ...probably more
+			\item 846 healthy Australian adults
+			\item 546,192 SNPs
+			\item 5380 gene expression probes from whole blood
+		\end{itemize}
+	\end{block}
+	\begin{block}{Task}
+		Perform exhaustive search for 2 locus epistasis for each probe
+	\end{block}
+	\begin{itemize}
+		\item Parallelise search across 100x GPU cluster
+		\item 802,494,666,067,680 tests in total...
 	\end{itemize}
@@
 
-\section{Thresholds}
-\subsection{}
+\begin{frame}{Thresholds}
 
-\begin{frame}{The curse of dimensionality}
+	\begin{block}{Bonferroni correction}
+		\textbf{Adjust p-value by the number of tests that are being performed}\\
+		We want the probability of there being a false positive at the end of the experiment to be limited to 0.05
+	\end{block}
+	\begin{equation}
+		-\log_{10}\left ( \frac{0.05}{802\times10^{14}} \right) \approx 16.5 \nonumber
+	\end{equation}
+	\begin{block}{The curse of dimensionality}
 		As the dimensionality of the search increases the background noise drowns out all real biological signals
+	\end{block}
 \end{frame}
@@
 	\begin{center}
-\includegraphics[width=10cm]{ld.png}
+		\includegraphics[width=8cm]{ld.png}
 	\end{center}
@@
 
-\begin{frame}{Permutation analysis}
-What is the \emph{effective} number of multiple tests that are being performed? (Churchill 1994)
-\begin{center}
-\includegraphics[width=10cm]{permutationanalysis.pdf}
-\end{center}
-\end{frame}
-
-\begin{frame}{A general formula for 2D thresholds?}
-\begin{equation}
--\log_{10}(p_{T}) = a - \frac{a}{f_{1}(N)+ f_{2}(M)} \nonumber
-\label{eqn:general_threshold}
-\end{equation}
-\begin{eqnarray}
-a & = & -\log_{10} \left (\frac{0.05}{t^2/2} \right) = 12.68 \nonumber
-\end{eqnarray}
-where $t = 693138$ is the number of independent regions in the genome (Dudbridge \& Gusnanto 2008)
-\end{frame}
-
 \begin{frame}{A general formula for 2D thresholds?}
@@
 
-\section{2D eQTL}
-\subsection{}
-
-\begin{frame}{Are epistatic effects controlling gene expression?}
+\begin{frame}{More accurate thresholds?}
 	\begin{itemize}
-\item Study design:
+		\item 1641 complete scans were performed on permuted phenotype
 		\begin{itemize}
-\item 846 healthy Australian adults
-\item 546,192 SNPs
-\item 5380 gene expression probes from whole blood
+			\item This means we can see how extreme the p-values would go by chance alone, because any real biological connection between genotype and phenotype has been severed, but the correlation structure remains intact
 		\end{itemize}
-\item Plan:
-\begin{itemize}
-\item Do exhaustive search for 2 locus epistasis for each probe
-\item Parallelise search across 100x GPU cluster
-\end{itemize}
+		\item Threshold predicted to be: \textbf{11.648}
+		\item Empirical estimate: \textbf{11.632}
+		\item Average correlation between 5380 probes: \textbf{0.265}
+		\item Estimate of threshold for all probes: $-\log_{10}p =$ \textbf{15.3}
 	\end{itemize}
@@
 
-\begin{frame}{Thresholds}
+\begin{frame}{Initial glance at the results}
 	\begin{itemize}
-\item 1641 complete scans were performed on permuted phenotype
-\item Threshold predicted to be: \textbf{11.648}
-\item Empirical estimate: \textbf{11.632}
-\item Average correlation between 5380 probes: \textbf{0.265}
-\item Overall threshold: $-\log_{10}p =$ \textbf{16.5}
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Very preliminary results}
-\begin{itemize}
-\item Is there much non-additive variance?
 		\item 71 probes have significant eQTLs after filtering for:
 		\begin{itemize}
-\item Threshold of 16.5
+			\item Bonferroni threshold of 16.5
 			\item Additive variance $\leq$ 60\% of total genetic variance
@@
 		\end{itemize}
-\item Number of significant epistatic pairs: \textbf{140}
 	\end{itemize}
+	\begin{block}{Overall:}
+		Number of significant epistatic pairs after extremely conservative filtering: \textbf{140}!
+	\end{block}
 \end{frame}
 
-\begin{frame}{Very preliminary results}
+\begin{frame}{Initial glance at the results}
 	\begin{itemize}
@@
 
+\begin{frame}{Next steps}
+	\begin{itemize}
+		\item Do they make biological sense?
+		\item Replicate! We have a dataset with the same expression array
+		\begin{itemize}
+			\item Sample size is only 150, this will make things tricky!
+		\end{itemize}
+		\item Continue with the scan on remaining probes
+		\begin{itemize}
+			\item But unfortunately the cluster is currently being used by some astronomers...
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section*{}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 	\begin{columns}[c]
@@
 		\column{.5\textwidth}
-
 			{\tiny
-PhD Funding:
-\begin{itemize}
-\item The Roslin Institute
-\item University of Edinburgh
-\item Biosciences KTN
-\item Monsanto / Newsham Choice Genetics
-\end{itemize}
-PhD Supervisors:
-\begin{itemize}
-\item Chris Haley
-\item Sara Knott
-\item John Woolliams
-\end{itemize}
-Important people:
-\begin{itemize}
-\item Ricardo Pong-Wong
-\item Athanasios Theocharidis
-\item Wenhua Wei
-\end{itemize}
 				Compute resources
@@
 		\column{.5\textwidth}
-
 			{\tiny
 				Complex Trait Genomics Group
-
 				\begin{itemize}
@@
 					\item Naomi Wray
-\item Joseph Powell
+					\item {\color{red} Joseph Powell}
 					\item Jian Yang
@@
 			}
-
 		\end{columns}
@@
 \begin{frame}
+	\begin{columns}[ccc]
+		\begin{column}{0.3\textwidth} \end{column}
+		\begin{column}{0.3\textwidth}
 			\centering
-Thanks!
+			\textbf{Thank you!}
+		\end{column}
+		\begin{column}{0.3\textwidth}\end{column}
+	\end{columns}
 \end{frame}
@@
 			SS_{W}  (g - 1)^{-1}
-}
-{  SS_{B} (n - g)^{-1}
-}; g - 1, n - g
+		}{
+			SS_{B} (n - g)^{-1}
+		};
+		g - 1, n - g
 		\right ) \right )
@@
 			SS_{W} / (SS_{W} + SS_{B}) \nonumber \\
-& \propto &
--\log(g) \nonumber
-\label{eq:ftest_n}
+		& \propto & -\log(g)
+		\nonumber
 	\end{eqnarray}
@@
 
+\begin{frame}{Detection is underpowered}
+	\begin{center}
+		\includegraphics[height=7.5cm]{gpmaps_ld.png}
+	\end{center}
+\end{frame}
+
 \end{document}

### 2013-04-16 docs/manuscript/report_v1.tex
directory restructuring

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LaTeX: Epistatic control of gene expression
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Edit the title below to update the display in My Documents
%\title{Epistatic control of gene expression}
%
%%% Preamble
\documentclass[paper=a4, fontsize=11pt]{scrartcl}	% Article class of KOMA-script with 11pt font and a4 format
\usepackage[T1]{fontenc}
\usepackage{fourier}

\usepackage[english]{babel}											% English language/hyphenation
\usepackage[protrusion=true,expansion=true]{microtype}				% Better typography
\usepackage{amsmath,amsfonts,amsthm}								% Math packages
\usepackage[pdftex]{graphicx}										% Enable pdflatex
\usepackage{url}

%%% Custom sectioning (sectsty package)
\usepackage{sectsty}												% Custom sectioning (see below)
\allsectionsfont{\centering \normalfont\scshape}					% Change font of al section commands

%%% Custom headers/footers (fancyhdr package)
\usepackage{fancyhdr}		
\pagestyle{fancyplain}
\fancyhead{}														% No page header
\fancyfoot[L]{}														% Empty 
\fancyfoot[C]{}														% Empty
\fancyfoot[R]{\thepage}												% Pagenumbering
\renewcommand{\headrulewidth}{0pt}									% Remove header underlines
\renewcommand{\footrulewidth}{0pt}									% Remove footer underlines
\setlength{\headheight}{13.6pt}

%%% Equation and float numbering
\numberwithin{equation}{section}									% Equationnumbering: section.eq#
\numberwithin{figure}{section}										% Figurenumbering: section.fig#
\numberwithin{table}{section}										% Tablenumbering: section.tab#

%%% Maketitle metadata
\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} 					% Horizontal rule

\title{
		%\vspace{-1in} 	
		\usefont{OT1}{bch}{b}{n}
		\normalfont \normalsize \textsc{University of Queensland} \\ [25pt]
		\horrule{0.5pt} \\[0.4cm]
		\huge Epistatic control of gene expression in Humans\\[0.3cm]
        \huge *** Report V1 *** \\
		\horrule{2pt} \\[0.5cm]
}
\author{
		\normalfont 								\normalsize
        Joseph Powell\\[-3pt]		\normalsize
        Gibran Hemani\\[0.2cm]		\normalsize
		\today
}
\date{}

%%% Begin document
\begin{document}
\maketitle
\section{Introduction}
In humans expression quantitative trait loci (eQTL) have been extensively studied for their effects on transcript levels as well as their underlying effects on complex phenotypes such as disease susceptability. To date most studies have analysed SNPs independently and estimated the genotypic effect on mean expression levels assuming an additive mode of inheritance. However, epistasis (gene by gene interactions) is thought to play an important role in the control of gene expression principally through the interaction of transcription factors and the intra-cellular effects of proteins on gene regulation.

[add short paragraph outlining the types of analysis for epistasis - leading to the conclusions that 2D is the best...]

Recently we have shown that whilst most genetic variance for gene expression is likely to act in an additive manner, for many probes there exists significant non-additive genetic varaiance. [Gib: can you add something on the 'appearence of additive variance' in our system?]. The importance of epistasis for control of gene expression is largely unknown and has genrally not been studied in human populations. It is likely that this is due in part to the considerable compuation demands of running a 2D scan on 1000's of probes as well as a restrivive study-wide multiple testing burden.

[we can add a short paragraph outlining the cool work that Gib has done.]

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%

\section{Methods}

Data

\cite{pmid22563384}

\begin{align} 
	\begin{split}
	(x+y)^3 	&= (x+y)^2(x+y)\\
					&=(x^2+2xy+y^2)(x+y)\\
					&=(x^3+2x^2y+xy^2) + (x^2y+2xy^2+y^3)\\
					&=x^3+3x^2y+3xy^2+y^3
	\end{split}					
\end{align}

%%% Refs
\bibliographystyle{plain}
\bibliography{report_v1_refs}

%%% End document
\end{document}

### 2013-04-16 presentation/20120612_combio_seminar/combio_2012.tex
directory restructuring

--- 2012-06-20 eQTL_2D/presentation/icqg4/icqg4_hemani.tex
@@
 \usetheme{Frankfurt}
-%\usecolortheme{dove}
-\usefonttheme{serif}
-
+\usecolortheme{dove}
 \usepackage[absolute,overlay]{textpos} 
@@
 %\definecolor{tableShade}{HTML}{F1F5FA}
-\usepackage[normalem]{ulem}
 
-\title[Epistasis in GWAS]{2D or not 2D, that is the question}
+\title[Epistasis in GWAS]{Genetic interactions in complex traits}
 \author{Gibran Hemani}
-\institute{The Roslin Institute, University of Edinburgh \\
-\vspace{.3cm}
-Diamantina Institute \\and \\
-Queensland Brain Institute,\\
-University of Queensland}
+\institute{The Roslin Institute, University of Edinburgh \\Diamantina Institute, University of Queensland \\Queensland Brain Institute}
 \date{}
@@
 
-\begin{frame}{}
+\section[Introduction]{Genome wide association studies}
+\subsection{}
+
+\begin{frame}{Missing heritability in GWAS}
+\begin{center}
+\begin{figure}
+\includegraphics[width=7cm]{missing.jpg} \\
+{\tiny Maher, 2008}
+\end{figure}
+\end{center}
 \begin{itemize}
-\item Is there theoretical reason to presume epistasis might play an important role in complex traits?
-\item Is it computationally tractable in GWAS?
-\item Is it statistically tractable in GWAS?
-\item Anything detected empirically?
+\item Many variants found and replicated from huge studies
+\item Most have extremely small effect sizes
+\item Very small proportion of estimated additive variance
 \end{itemize}
@@
 
-\section{Polymorphisms under selection}
+\begin{frame}{Narrow-sense heritability}
+\begin{figure}
+% \begin{minipage}[c]{0.38\textwidth}
+\includegraphics[width=6cm]{visscher2008_heritabiliy.jpg} \\
+{\tiny Visscher, 2008}
+\end{figure}
+\begin{itemize}
+\item Fitness related traits have lower heritabilites
+\item Additive variance appears to be maintained under selection
+\item Response to selection is lower than expected for fitness related traits
+\end{itemize}
+\end{frame}
+
+\section{Epistasis}
 \subsection{}
@@
 
-\begin{frame}{Epistasis and additive variance}
+\begin{frame}{Examples}
+
+\begin{figure}[htb]
+ \centering
+ \begin{minipage}[c]{0.38\textwidth}
+  \centering
+{\tiny
+\begin{tabular}{cccc}
+& AA & Aa & aa \\\hline
+BB & $2\alpha + 2\beta$ & $\alpha + 2\beta$ & $2\beta$ \\
+Bb & $2\alpha + \beta$ & $\alpha + \beta$ & $\beta$ \\
+bb & $2\alpha$ & $\alpha$ & $0$ \\
+&&&\\
+\multicolumn{4}{l}{Additive + additive (not epistatic)}\\
+
+\end{tabular} }
+ \end{minipage}
+ \begin{minipage}[c]{0.38\textwidth}
+ \includegraphics[width=\textwidth]{additive.jpg}
+ \end{minipage}
+ \end{figure}
+
+\begin{figure}[htb]
+ \centering
+ \begin{minipage}[c]{0.38\textwidth}
+   \centering
+{\tiny
+\begin{tabular}{cccc}
+& AA & Aa & aa \\\hline
+BB & $\alpha$ & $\alpha$ & $\alpha$ \\
+Bb & $\alpha$ & $\alpha$ & $\alpha$ \\
+bb & $\alpha$ & $\alpha$ & $\beta$ \\
+&&&\\
+\multicolumn{4}{l}{Canalisation (epistatic)}\\
+\end{tabular} }
+ \end{minipage}
+ \begin{minipage}[c]{0.38\textwidth}
+ \includegraphics[width=\textwidth]{canalisation.jpg}
+ \end{minipage}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Two dimensional GWAS}
 \begin{center}
-\includegraphics[height=7cm]{gp_va}
+\includegraphics[width=7cm]{2dscan.png} \\
+{\tiny Carlborg 2004}
 \end{center}
@@
 
-\begin{frame}{}
+\begin{frame}{Impact of LD on observed variance}
+\begin{center}
+\includegraphics[width=5cm]{gpmaps_ld.png} \\
+\end{center}
+\end{frame}
+
+\begin{frame}{Can epistasis maintain additive variance?}
 \begin{itemize}
-\item \textbf{Hypothesis:} Epistasis can maintain additive variance better than purely additive effects
+\item Hypothesis: Epistasis can maintain additive variance better than purely additive effects
 \item How can maintained additive variance be detected most effectively?
@@
 
-\begin{frame}{Epistasis maintains common genetic variation}
+\begin{frame}{Epistatic patterns}
 \begin{center}
-\includegraphics[width=10cm]{maps_freqs}
+\includegraphics[width=8cm]{sup_gpmaps.pdf}
+\end{center}
+\end{frame}
+
+\begin{frame}{Allele frequency trajectories}
+\begin{center}
+\includegraphics[width=8cm]{sup_allelefreq_sim.png}
 \end{center}
@@
 \end{columns}
+\end{frame}
+
+\begin{frame}{Can epistasis maintain additive variance?}
+\begin{itemize}
+\item Hypothesis: Epistasis can maintain additive variance better than purely additive effects
+\begin{itemize}
+\item Yes, to a much greater extent
+\end{itemize}
+\item How can maintained additive variance be detected most effectively?
+\begin{itemize}
+\item 2 dimensional exhaustive searches
+\end{itemize}
+\item What is the impact of low LD between causal variants and observed SNPs?
+\begin{itemize}
+\item Upwards bias in reporting additive effects
+\end{itemize}
+\end{itemize}
 \end{frame}
@@
 
+\begin{frame}{CPUs and hitting the `power wall'}
+\begin{itemize}
+\item All electrical power consumed eventually radiates as heat
+\item Processor speed requires more power to improve
+\begin{itemize}
+\item Increased heat
+\item Increased circuit resistance
+\item Instability
+\end{itemize}
+\end{itemize}
+\end{frame}
+
 \begin{frame}{CPU limits}
@@
 \includegraphics[height=7cm]{clockspeed.pdf}
+\end{center}
+\end{frame}
+
+\begin{frame}{CPU limits}
+\begin{center}
+\includegraphics[height=6cm]{clockrates.png}
 \end{center}
@@
 \item Use OpenCL API to communicate with graphics cards
+
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Chip structures}
+\includegraphics[height=2.5cm]{cpugpu.png}
+\begin{itemize}
+\item Many cores that can process data simultaneously
+\item But all the cores are extremely simple
+\begin{itemize}
+\item Tiny L1 cache
+\item Little communication between cores
+\item Each core is relatively slow
+\item Cores are divided into local groups, all of which must do the same thing
+\end{itemize}
+
 \end{itemize}
@@
 
-\begin{frame}{Other epistasis software}
-\begin{itemize}
-\item BiForce
-\item GLIDE
-\item GBOOST
-\item TEAM
-\item MDR-GPU
-\item EPIBLASTER
-\item SHEsisEPI
-\item ...probably more
-\end{itemize}
-\end{frame}
+% mention epiGPU
 
@@
 \subsection{}
+
+\begin{frame}{Multiple testing}
+\begin{definition}
+{\color{orange} Frequentist $p$-value: The probability of obtaining a test statistic at least as extreme as the one that was actually observed, assuming that the null hypothesis is true.}
+\end{definition}
+\begin{itemize}
+\item \emph{e.g.} If 20 independent tests are performed then would expect one true null to have a $p$-value $\leq 0.05$.
+\item So if significance with a false positive rate of $\alpha = 0.05$ is desired, then the threshold for the family of tests should be adjusted to $\alpha_{F} = 0.05 / 20 = 0.0025$.
+\item 500000 SNPs being tested $\rightarrow \alpha_{F} = 0.05 / 500000 = 1 \times 10^{-7}$
+\item Epistasis $\rightarrow \alpha_{F} = \frac{0.05}{n(n-1)/2} \approx 4 \times 10^{-13}$
+\end{itemize}
+\end{frame}
 
@@
 
-\begin{frame}{}
-\begin{center}
-What is the correlation structure in a 2D search?
-\end{center}
-\end{frame}
-
 \begin{frame}{Permutation analysis}
@@
 \includegraphics[width=10cm]{permutationanalysis.pdf}
+\end{center}
+\end{frame}
+
+\begin{frame}{What is the correlation structure in a 2D search?}
+\begin{center}
+\includegraphics[width=7cm]{2dscan.png} \\
+{\tiny Carlborg 2004}
 \end{center}
@@
 \begin{equation}
--\log_{10}(p_{T}) = a - \frac{a}{\beta_{1} N + \beta_{2} M} \nonumber
+-\log_{10}(p_{T}) = a - a \left ( 
+ s_{1} \exp \left ( 
+  \frac {\ln(n) } { s_{2} } \right )
+ + s_{3} \exp \left (
+  \frac{ \ln(M)} {s_{4}} \right ) 
+\right )^{-1} \nonumber
 \label{eqn:general_threshold}
@@
 \begin{eqnarray}
-a & = & -\log_{10} \left (\frac{0.05}{t^2/2} \right) = 12.68 \nonumber
+a & = & 12.68 \nonumber \\
+s_{1} & = & 0.0258 \nonumber \\
+s_{2} & = & 1.72 \nonumber \\
+s_{3} & = & 0.0379 \nonumber \\
+s_{4} & = & 2.33. \nonumber
 \end{eqnarray}
-where $t = 693138$ is the number of independent regions in the genome (Dudbridge \& Gusnanto 2008)
 \end{frame}
@@
 \item Naomi Wray
-\item Joseph Powell
+\item \textbf{Joseph Powell}
 \item Jian Yang
@@
 
-\begin{frame}{Epistatic patterns}
-\begin{center}
-\includegraphics[width=8cm]{sup_gpmaps.pdf}
-\end{center}
-\end{frame}
-
-\begin{frame}{Allele frequency trajectories}
-\begin{center}
-\includegraphics[width=8cm]{sup_allelefreq_sim.png}
-\end{center}
-\end{frame}
-
 \end{document}

### 2013-04-21 docs/manuscript/report_v1.tex
added to the methods section of the report

@@
 
-Data
+\subsection{Discovery data}
 
-\cite{pmid22563384}
+The Brisbane Systems Genetics Study (BSGS) comprises 862 Individuals of European descent from 274 independent families \cite{pmid22563384}.
+DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland.
+Full details of genotyping procedures are given in Medland et al.
+\cite{Medland2009} Standard QC filters were applied and the remaining 528,509 SNPs were carried forward further analysis.
 
-\begin{align} 
-	\begin{split}
-	(x+y)^3 	&= (x+y)^2(x+y)\\
-					&=(x^2+2xy+y^2)(x+y)\\
-					&=(x^3+2x^2y+xy^2) + (x^2y+2xy^2+y^3)\\
-					&=x^3+3x^2y+3xy^2+y^3
-	\end{split}					
-\end{align}
+Gene expression profiles were generated from whole blood collected with PAXgene TM tubes (QIAGEN, Valencia, CA) using Illumina HT12-v4.0 bead arrays.
+The Illumina HT-12 v4.0 chip contains 47,323 probes, although some probes are not assigned to RefSeq genes.
+We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF > 0.05 within 1000 genomes data; did not map to a listed ref-seq gene; were not significantly expressed (based on a detection $p$-value < 0.05) in 100\% of samples.
+After this stringent QC ~7000 [need to get exact number] probes remained for 2d-eqtl mapping.
+
+\subsection{normalisation}
+
+Gene expression profiles were normalised and batch effects removed with the following procedures; Raw background expression were removed for each sample.
+Expression levels were then adjusted using Quantile and log2 transformation to standardise distribution between samples.To avoid inflation of the test statistic due to polygenic effects the following linear model was used to correct for batch and polygeneic effects;
+
+\begin{equation}
+y = \mu + c + p + s + a + g + e
+\label{eq:lm}
+\end{equation}
+where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectivly.
+$g$ is a random additive polygenic effect with a variance covariance matrix
+
+\begin{equation}
+G_{ijk} = \left \{ 
+\begin{array}{ll}
+\sigma _a ^2 + \sigma _e ^2&        j = k \\ 
+2\phi _{ijk} \sigma _a ^2& 			j \neq k \\
+\end{array} \right.
+\end{equation}
+
+The parameters $\sigma _a ^a$ and $\sigma _e ^2$ are variance components for additive background genetic and environmental effects respectively.
+Here, we are using family based pedigree information rather than SNP based IDB to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$.
+The residual, $e$, from equation \ref{eq:lm} are assumed to follow a multivariate normal distribution with a mean of zero.
+These residuals are corrected for batch, sex, generation and polygentic effects.
+Residuals were standardised to Z-scores uning a inverse-normal transformation and used as the adjusted phenotype for the pairise epistasis scan.
+
+\subsection{exhaustive 2d-eQTL analysis}
+We used epi-GPU software to perform an exhaustive scan for pairwise interactions, such that each SNP is tested against all other SNPs for statistical association with the expression values for each of the ~7000 probes.
+For each SNP-pair there are 9 possible genotype classes.
+We treat each genotype class as a fixed effect and fit an 8 d.f.
+$F$-test to test the following hypotheses:
+
+\begin{equation}
+H _0 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 = 0; 
+\end{equation}
+
+\begin{equation}
+H _1 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 > 0; 
+\label{eq:8df}
+\end{equation}
+
+where $\mu$ is the mean expression level and $x _{ij}$ is the pairwise genotype class mean for genotype $i$ at SNP 1 and genotype $j$ at locus B.
+This type of test does not parameterisze for specific types of epistasis, rather it tests for the joint genetic effects at two loci.
+This has been demonstrated to be statistically more efficient thwhen searching for a wide range of epistatic patterns (add refs), although will include any marginal effects of SNPs.
+\\[0.2cm]
+
+The complete exhaustive scan for 7,000 probes comprises of $\sim$9.8$e^{14}$ $F$-tests.
+Due to the very large number of tests [plus nature of epistasis ....] we used a highly conservative series of filtering steps to identify study-wide significant epistasis associations to follow forward for replication analysis.
+Initially a significance threshold of $-log10$ 15.5.
+This threshold is the studywide significance threshold determined by the number of independent tests for a 2D scan for a single probe / the number of probes analysed.
+This is likely to be conservative as this threshold assumes independance between probes.
+Following filtering on this threshold only SNP pairs with all 9 genotype classes and minium genotype class size > 5 individuals were kept.
+We then calculated the linkage disequilibrium between SNPs in a pair and removed any pairs with $r^2$ > 0.1.
+We also removed any pairs containing SNPs for which single marker additive or dominace effects had been identifyed in previous work (Powell et al.
+in Press).
+Sentinal SNP pairs we retained from epistasis eQTL 'peaks' comprising of multiple sets of pair-wise SNPs.
+At this stage $\sim$13,000 SNP pairs remained and were tested for the contribution of marginal and interaction terms by testing the 8d.f.
+\ref{eq:8df} verses a 4 d.f.
+model parameterizing for interaction terms only:
+
+\begin{equation}
+H _0: \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \bar x _i - \bar x _j + \mu) ^2 = 0;
+\end{equation}
+
+\begin{equation}
+H _1: \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \bar x _i - \bar x _j + \mu) ^2 > 0;
+\label{eq:4df}
+\end{equation}
+
+where $\bar x _i (\bar x _j)$ is the marginal class mean for genotype $i (j)$ at SNP A (B). Significance of interaction terms was determined using a bonferroni threshold of $0.05 / 13000 = 3.8e^{-6}$
+
+\subsection{replication}
+
+\subsection{simulation}
 

### 2013-04-23 impute_fine_map/epi_impute_analysis.R
added fun to collect the maximum association from the analysis of imputed blocks

@@ -14,6 +14,10 @@
 set3 <- load("/ibscratch/wrayvisscher/josephP/BSGS/Imputed/Epistasis/set3.RData")
 
 
+#=======================================================#
+#	pull in for information in the impute scan files	#
+#=======================================================#
+
 # Read in relavent imputed scan output (matched by probe and SNP 1 and SNP 2) 
 
 
@@ -22,7 +26,7 @@ impute_scan_filter.fun <- function(set3, dir) {
 	# dir :	 dir of the impute scan outputs	
 
 
-	out <- array(0, c(nrow(set3), 9))
+	out <- array(0, c(nrow(set3), 10))
 	for (i in 1:nrow(set3)) {
 		tmp <- read.table(paste(dir, "epi_impute_scan_", as.character(set3$probename[i]), "_", as.character(set3$snp1[i]), "_", as.character(set3$snp2[i]), ".txt", sep=""), header=T)
 
@@ -30,13 +34,14 @@ impute_scan_filter.fun <- function(set3, dir) {
 		index <- which(abs(tmp$rsq) < 0.1 & tmp$nclass==9 & tmp$minclass > 5)
 		tmp <- tmp[index,]
 
-		out[i,] <- as.matrix(tmp[which.max(tmp$intP),])
+		out[i,1:9] <- as.matrix(tmp[which.max(tmp$intP),])
+		out[i,10] <- nrow(tmp)
 		print(i)	
 
 	}
 
 	out <- as.data.frame(out)
-	names(out) <- names(tmp)	
+	names(out) <- c(names(tmp), "ntests")	
 	out <- cbind(set3, out)
 
 	return(out)
@@ -45,6 +50,9 @@ impute_scan_filter.fun <- function(set3, dir) {
 
 
 impute_results <- impute_scan_filter.fun(set3, dir)
+write.csv(umpute_results, "/ibscratch/wrayvisscher/josephP/BSGS/Imputed/Epistasis/impute_sig_results.csv", quote=F, row.names=F)
+
+
 
 
 

### 2013-04-23 docs/manuscript/report_v1.tex
added text on fine mapping and replication sections

@@
 \section{Introduction}
-In humans expression quantitative trait loci (eQTL) have been extensively studied for their effects on transcript levels as well as their underlying effects on complex phenotypes such as disease susceptability.
+In humans expression quantitative trait loci (eQTL) have been extensively studied for their effects on transcript levels as well as their underlying effects on complex phenotypes such as disease susceptibility.
 To date most studies have analysed SNPs independently and estimated the genotypic effect on mean expression levels assuming an additive mode of inheritance.
@@
 
-Recently we have shown that whilst most genetic variance for gene expression is likely to act in an additive manner, for many probes there exists significant non-additive genetic varaiance.
-[Gib: can you add something on the 'appearence of additive variance' in our system?].
-The importance of epistasis for control of gene expression is largely unknown and has genrally not been studied in human populations.
-It is likely that this is due in part to the considerable compuation demands of running a 2D scan on 1000's of probes as well as a restrivive study-wide multiple testing burden.
+Recently we have shown that whilst most genetic variance for gene expression is likely to act in an additive manner, for many probes there exists significant non-additive genetic variance.
+[Gib: can you add something on the 'appearance of additive variance' in our system?].
+The importance of epistasis for control of gene expression is largely unknown and has generally not been studied in human populations.
+It is likely that this is due in part to the considerable computation demands of running a 2D scan on 1000's of probes as well as a restrictive study-wide multiple testing burden.
 
@@
 Gene expression profiles were normalised and batch effects removed with the following procedures; Raw background expression were removed for each sample.
-Expression levels were then adjusted using Quantile and log2 transformation to standardise distribution between samples.To avoid inflation of the test statistic due to polygenic effects the following linear model was used to correct for batch and polygeneic effects;
+Expression levels were then adjusted using Quantile and log2 transformation to standardise distribution between samples.
+To avoid inflation of the test statistic due to polygenic effects the following linear model was used to correct for batch and polygenic effects;
 
@@
 \end{equation}
-where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectivly.
+where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively.
 $g$ is a random additive polygenic effect with a variance covariance matrix
@@
 The residual, $e$, from equation \ref{eq:lm} are assumed to follow a multivariate normal distribution with a mean of zero.
-These residuals are corrected for batch, sex, generation and polygentic effects.
-Residuals were standardised to Z-scores uning a inverse-normal transformation and used as the adjusted phenotype for the pairise epistasis scan.
+These residuals are corrected for batch, sex, generation and polygenic effects.
+Residuals were standardised to Z-scores using a inverse-normal transformation and used as the adjusted phenotype for the pairwise epistasis scan.
 
@@
 where $\mu$ is the mean expression level and $x _{ij}$ is the pairwise genotype class mean for genotype $i$ at SNP 1 and genotype $j$ at locus B.
-This type of test does not parameterisze for specific types of epistasis, rather it tests for the joint genetic effects at two loci.
-This has been demonstrated to be statistically more efficient thwhen searching for a wide range of epistatic patterns (add refs), although will include any marginal effects of SNPs.
+This type of test does not parameterize for specific types of epistasis, rather it tests for the joint genetic effects at two loci.
+This has been demonstrated to be statistically more efficient when searching for a wide range of epistatic patterns (add refs), although will include any marginal effects of SNPs.
 \\[0.2cm]
@@
 Initially a significance threshold of $-log10$ 15.5.
-This threshold is the studywide significance threshold determined by the number of independent tests for a 2D scan for a single probe / the number of probes analysed.
-This is likely to be conservative as this threshold assumes independance between probes.
-Following filtering on this threshold only SNP pairs with all 9 genotype classes and minium genotype class size > 5 individuals were kept.
+This threshold is the study-wide significance threshold determined by the number of independent tests for a 2D scan for a single probe / the number of probes analysed.
+This is likely to be conservative as this threshold assumes independence between probes.
+Following filtering on this threshold only SNP pairs with all 9 genotype classes and minimum genotype class size > 5 individuals were kept.
 We then calculated the linkage disequilibrium between SNPs in a pair and removed any pairs with $r^2$ > 0.1.
-We also removed any pairs containing SNPs for which single marker additive or dominace effects had been identifyed in previous work (Powell et al.
+We also removed any pairs containing SNPs for which single marker additive or dominance effects had been identified  in previous work (Powell et al.
 in Press).
-Sentinal SNP pairs we retained from epistasis eQTL 'peaks' comprising of multiple sets of pair-wise SNPs.
+Sentinel  SNP pairs we retained from epistasis eQTL 'peaks' comprising of multiple sets of pair-wise SNPs.
 At this stage $\sim$13,000 SNP pairs remained and were tested for the contribution of marginal and interaction terms by testing the 8d.f.
@@
 
+\subsubsection{Fine mapping}
+
+To assess the impact of pair-wise epistasis variants not present on the Illumina 610Q beadchip and better define the association regions, we imputed genotypes in the region 100kb upstream and downstream of the sentinel SNPs using the full reference panel from the 1000 Genomes Project.
+The quality of imputation was assessed by means of the $R^2$ statistic.
+Used a highly conservative threshold, we removed any SNPs with $R^2 < 0.9$.
+Thus, or each epistasis pair two imputed SNP blocks were carried forward for fine mapping analysis.
+Each pairwise combination of SNPs in blocks 1 and 2 were tested in a ful 8d.f.
+and reduced (interaction) 4d.f.
+model.
+As before any snp pairs not matching the criteria of $r^2$  < 0.1, number of genotype classes = 9, and minimum genotype class size > 5 were removed from the output.
+
+This approach allows us to fine map to the causal epistatic loci.
+Given the nature epistasis genetic variance, any snp pairs with large estimate of non-additive genetic variance are very unlikely to be rare.
+Therefore, fine mapped loci provide good functional candidates.
+However, as fine mapping includes a large number of additional tests increases in the interaction (4d.f.) $p$-values could be due to chance [need a better way of describing this].
+
 \subsection{replication}
 
-\subsection{simulation}
+We have attempted replication of the 2D scan significant results using three independent cohorts; Fehrmann, CHDWB and Rotterdam. Details of cohorts are as follows. \\[0.1cm]
+
+Fehrmann: $n$=1,469 \\[0.1cm]
+The Fehrmann dataset consists of whole peripheral blood samples of 1,469 unrelated individuals from the United Kingdom and the Netherlands.
+Some of these individuals are patients, while others are healthy controls.
+Individuals were genotyped using Illumina HumanHap300, HumanHap370 or the 610 Quad platform.
+RNA levels were quantified using both the Illumina H8v2 platform (N = 229) and the HT12v3 platform (N = 1,240).
+\\[0.1cm]
+
+CDHWB: $n$=189 \\ [0.1cm]
+The Center for Health Discovery and Well Being (CDHWB) Study is a population based cohort consisting of 189 individuals collected in Atlanta USA.
+Gene expression profiles were generated with Illumina HT-12 V3.0 arrays from whole blood collected from Tempus tubes that preserve RNA.
+Whole genome genotypes were measured using Illumina OmniQuad arrays.
+\\[0.1cm]
+
+Rotterdam: $n$=762 \\[0.1cm]
+The Rotterdam Study is a large population based cohort collected to investigate the prevalence, incidence, and risk factors of chronic diseases among  Caucasians aged 45 years and over.
+Participants were genotyped using the Illumina 610K quad arrays.
+Whole blood of 768 samples was collected (PAXgene)and RNA hybridized to the Illumina Human HT-12v4 chips.
+
+\subsection{Functional characterisation}
+
+[From Kostya]
+
+\section{Results}
+
+\subsection{2D Scan}
+
+\subsection{Additive vs non-additive}
+
+\subsection{Fine Mapping}
+
+\subsection{Functional characterisation}
 

### 2013-04-23 docs/manuscript/report_v1.tex
Manuscript updates - performed some analysis and made figures
- Added circos plots
- 3D bar plots of biggest interactions
- Characterised the type of interactions (cis-cis etc)
- Characterised genetic variance components
- Analysed replication in Greg Gibson's dataset
- Compared additive to non-additive genetic variance

@@
 
-[add short paragraph outlining the types of analysis for epistasis - leading to the conclusions that 2D is the best...]
+The detection of epistasis is significantly hampered by power issues.
+For example, there is a greater dependency on linkage disequilibrium (LD) between causal variants and observed SNPs in order to capture non-additive variance, and this problem is exacerbated further with epistasis because there must be high LD at all interacting loci.
+Another obstacle comes from the dependence on a much larger sample size in order to detect similar effect sizes compared to additive effects, because of the increased number of parameters in statistical tests for interaction terms, and the increased multiple testing correction for testing combinations of SNPs.
+Indeed, most examples of epistasis in humans come from candidate gene studies where multiple testing does not have a detrimental effect.
+
+Many approaches exist to detect epistasis, many of them are highly parametric, searching for a particular pattern of epistasis or only amongst SNPs with large marginal effects.
+However, it can be argued that the many patterns of epistasis will go undetected if the search is over-parameterised, and it has been shown that using a test that is agnostic of the epistatic pattern in an exhaustive manner can detect signals that would otherwise go undetected.
+One method to overcome the problem of excessive multiple testing is to search for interactions in traits, such as gene expression, that are likely to have very large effects, thus improving the power of the study.
 
 Recently we have shown that whilst most genetic variance for gene expression is likely to act in an additive manner, for many probes there exists significant non-additive genetic variance.
-[Gib: can you add something on the 'appearance of additive variance' in our system?].
 The importance of epistasis for control of gene expression is largely unknown and has generally not been studied in human populations.
 It is likely that this is due in part to the considerable computation demands of running a 2D scan on 1000's of probes as well as a restrictive study-wide multiple testing burden.
-
-[we can add a short paragraph outlining the cool work that Gib has done.]
+Due to modelling limitations, estimating the genetic variance due to non-additive genetic effects, and epistasis in particular, is often intractable.
+Thus, searching for epistatic effects directly may serve as a first step in quantifying the contribution of epistasis to complex traits.
 
@@
 
+We performed exhaustive two-dimensional searches on 7339 expression probes, identifying 432 significant pairwise interactions on 218 probes.
+We decomposed the genotype-phenotype map into parametric variance components and observed that there was no enrichment for any particular epistatic variance component amongst the significant SNPs (Table \ref{tab:vc}), demonstrating that a search that is agnostic of the pattern of epistasis is likely to detect things that would otherwise be missed.
+The majority of significant interactions were found to have a `cis-trans' effect (Table \ref{tab:type}; \emph{i.e.} one SNP was on the same chromosome as the probe while the other was on a different chromosome).
+Of the 218 probes, 78 had more than one significant interaction (Figure \ref{fig:circos})
+
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{rlr}
+		\hline
+		& Interaction type & Freq \\ 
+		\hline
+		1 & cis-cis &  32 \\ 
+		2 & cis-trans & 388 \\ 
+		3 & trans-trans &  12 \\ 
+		\hline
+	\end{tabular}
+	\label{tab:type}
+\end{table}
+
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{rlr}
+		\hline
+		& Main epistatic effect & Freq \\ 
+		\hline
+		1 & A x A &  99 \\ 
+		2 & A x D &  93 \\ 
+		3 & D x A & 117 \\ 
+		4 & D x D & 123 \\ 
+		\hline
+	\end{tabular}
+	\label{tab:vc}
+\end{table}
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/3dplots}
+	\caption{Genotype-phenotype maps for the nine most significant interactions.}
+	\label{fig:3dplots}
+\end{figure}
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/circos}
+	\caption{Genomic maps for 39 probes that have two or more significant interactions. The red dot denotes the chromosome where the probe resides and arcs denote interacting SNP pairs.}
+	\label{fig:circos}
+\end{figure}
+
+We attempted replication of the signals in an independent dataset [Greg Gibson - 139 individuals, 185 out of 432 interactions had both SNPs in common in the two datasets].
+We observed no significant $p$-values after adjusting for multiple testing (lowest $p$-value$=0.0099$ in replication dataset, threshold set at $0.05/185=0.0003$).
+We tested for enrichment of $p$-values but there was no evidence of departure from the null hypothesis of no effects (Figure \ref{fig:gg_replication}).
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/replication_qqplot}
+	\caption{Q-Q plot for interaction $p$-values (4 degrees of freedom) for 185 interactions that had SNPs in common in the replication dataset.}
+	\label{fig:gg_replication}
+\end{figure}
+
 \subsection{Additive vs non-additive}
+
+This study has focused on identifying statistical interactions, however all significant interactions exhibit relatively large additive genetic effects (Figure \ref{fig:proportion_additive}).
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/proportion_additive}
+	\caption{Proportion of genetic variance attributed to additive and non-additive effects, ranked in order of total genetic variance explained by the interacting SNP pair.}
+	\label{fig:proportion_additive}
+\end{figure}
 

### 2013-04-23 analysis/3dbars.R

@@ -1,9 +1,242 @@
 library(lattice)
 library(latticeExtra)
-?panel.3dbars
+library(noia)
+library(gridExtra)
+library(ggplot2)
+library(reshape2)
+library(plyr)
 
-tab <- as.table(matrix(-4:4, 3, 3))
-cloud(tab, panel.3d.cloud=panel.3dbars, col="black", col.facet=2:4, xbase=0.9, ybase=0.9)
+# Load in data
+
+load("~/repo/eQTL-2D/data/residuals_all.RData")
+load("~/repo/eQTL-2D/filtering/filtered_by_chr/sig_all_probes.RData")
+load("~/repo/eQTL-2D/data/clean_geno_final.RData")
 
 
+
+#' What is the biggest variance component for each interaction?
+#'
+#' For each interaction decompose the variance using NOIA and 
+#' find the epistatic VC with the largest variance
+#' @param sig A single row from sig
+#' @param geno Genotype matrix
+#' @param phen Residuals for each probe
+#' @param loud=FALSE Print full linear model?
+#' @export
+#' @return aa, ad, da or dd
+#' @alias
+#' @examples \dontrun{
+#' varianceComponents(sig[1,], xmat, resphen, TRUE)
+#'}
+varianceComponents <- function(sig, geno, phen, loud=FALSE)
+{
+	a <- linearRegression(
+		phen = phen[,sig$probeid],
+		gen  = cbind(geno[,sig$pos1], geno[, sig$pos2]) + 1
+	)
+	if(loud) print(a)
+	vars <- a$variances
+	vars <- vars[names(vars) %in% c("aa", "ad", "da", "dd")]
+
+	nom <- names(vars)[which.max(vars)]
+
+	nom <- unlist(strsplit(toupper(nom), split=""))
+	nom <- paste(nom, collapse=" x ")
+
+	return(nom)
+}
+
+#' What is the biggest variance component for each interaction?
+#'
+#' For each interaction decompose the variance using NOIA and 
+#' find the epistatic VC with the largest variance
+#' @param sig A single row from sig
+#' @param geno Genotype matrix
+#' @param phen Residuals for each probe
+#' @param loud=FALSE Print full linear model?
+#' @export
[…]

### 2013-04-23 analysis/circles.R

@@ -1,0 +1,157 @@
+library(ggbio)
+library(grid)
+library(gridExtra)
+library(plyr)
+library(GenomicRanges)
+
+
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL)
+{
+	require(grid)
+
+	# Make a list from the ... arguments and plotlist
+	plots <- c(list(...), plotlist)
+
+	numPlots = length(plots)
+
+	# If layout is NULL, then use 'cols' to determine layout
+	if (is.null(layout)) 
+	{
+		# Make the panel
+		# ncol: Number of columns of plots
+		# nrow: Number of rows needed, calculated from # of cols
+		layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+		ncol = cols, nrow = ceiling(numPlots/cols))
+	}
+
+	if (numPlots==1) 
+	{
+		print(plots[[1]])
+	} else {
+		# Set up the page
+		grid.newpage()
+		pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+
+		# Make each plot, in the correct location
+		for (i in 1:numPlots)
+		{
+			# Get the i,j matrix positions of the regions that contain this subplot
+			matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+
+			print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+			layout.pos.col = matchidx$col))
+		}
+	}
+}
+
+
+makeGr <- function()
+{
+	data("hg19Ideogram", package = "biovizBase")
+	chr.sub <- paste("chr", 1:22, sep = "")
+	new.names <- as.character(1:22)
+	names(new.names) <- paste("chr", new.names, sep = "")
+	hg19Ideo <- hg19Ideogram
+	hg19Ideo <- keepSeqlevels(hg19Ideogram, chr.sub)
+	hg19Ideo <- renameSeqlevels(hg19Ideo, new.names)
+	head(hg19Ideo)
+	gr <- GRanges(
+		seqnames = (1:22),
+		ranges = hg19Ideo@ranges
+	)
+	seqlengths(gr) <- seqlengths(hg19Ideo)
+	return(gr)
+}
+
+
+makeLinks <- function(x, gr)
+{
+	links1 <- GRanges(
+		seqnames = x$chr1,
+		IRanges(start = x$position1, width=1)
+	)
+	links2 <- GRanges(
[…]

### 2013-04-23 analysis/orig_analysis.R

@@ -33,7 +33,7 @@ sig$type[with(sig, chr1 != probechr & chr2 != probechr)] <- "trans-trans"
 # trans-trans: 24
 
 
-with(sig, tapply(pfull, type, mean))
+with(sig, tapply(propG, type, mean))
 library(ggplot2)
 
 ggplot(sig, aes(x=propG)) + geom_histogram() + facet_grid(.~type)

### 2013-04-23 data/merge_data.R

@@ -1,0 +1,51 @@
+mergeResiduals <- function()
+{
+	load("~/repo/eQTL-2D/data/residuals.RData")
+	hsq1 <- hsq
+	resphen1 <- resphen
+	probeinfo1 <- probeinfo
+	load("~/repo/eQTL-2D/data/residuals2.RData")
+	hsq <- c(hsq1, hsq)
+	resphen <- cbind(resphen1, resphen)
+	probeinfo <- rbind(probeinfo1, probeinfo)
+
+	save(hsq, probeinfo, resphen, file="~/repo/eQTL-2D/data/residuals_all.RData")
+}
+
+characteriseInteraction <- function(sig)
+{
+	sig$type <- NA
+	index <- with(sig, chr1 == probechr & chr2 == probechr)
+	sig$type[index] <- "cis-cis"
+	index <- with(sig, (chr1 == probechr & chr2 != probechr) | (chr1 != probechr & chr2 == probechr))
+	sig$type[index] <- "cis-trans"
+	index <- with(sig, chr1 != probechr & chr2 != probechr)
+	sig$type[index] <- "trans-trans"
+	return(sig)
+}
+
+loadSigData <- function()
+{
+	load("~/repo/eQTL-2D/filtering/filtered_by_chr/sig_all_probes.RData")
+	bim <- read.table("~/repo/eQTL-2D/data/clean_geno_final.bim", colClasses=c("character", "character", "numeric", "numeric", "character", "character"))
+	sig$position1 <- bim$V4[sig$pos1]
+	sig$position2 <- bim$V4[sig$pos2]
+	sig$snp1 <- as.character(sig$snp1)
+	sig$snp2 <- as.character(sig$snp2)
+	sig$probename <- as.character(sig$probename)
+	sig$probegene <- as.character(sig$probegene)
+
+	# map probe chromosomes
+	sig$probechr[sig$probegene == "MAP3K2"] <- 2
+	sig$probechr[sig$probegene == "NAPSB"] <- 19
+	sig$probechr[sig$probegene == "RPS26L"] <- 12
+	sig$probechr[sig$probegene == "SAPS2"] <- 22
+	sig$probechr[sig$probegene == "C15ORF28"] <- 22
+
+	sig <- characteriseInteraction(sig)
+
+	save(sig, file="~/repo/eQTL-2D/filtering/filtered_by_chr/sig_all_probes.RData")
+}
+
+temp <- mergeResiduals()
+sig <- loadSigData()

### 2013-04-24 docs/manuscript/report_v1.tex
minor chnages to the figure legends and text

@@
 		\hline
-		1 & cis-cis &  32 \\ 
-		2 & cis-trans & 388 \\ 
-		3 & trans-trans &  12 \\ 
+		1 & $cis-cis$ &  32 \\ 
+		2 & $cis-trans$ & 388 \\ 
+		3 & $trans-trans$ &  12 \\ 
 		\hline
 	\end{tabular}
+	\caption{Epistasis pair positions. Here $cis$ SNPs are on the same chromosome as the probe. $Trans$ SNPs are on a different chromosome}
 	\label{tab:type}
@@
 	\end{tabular}
+	\caption{The largest epistatic variance component was recorded for each pair. This table summerises how often each type of epistatic variance occurs}
 	\label{tab:vc}
@@
 	\includegraphics[width=15cm]{images/3dplots}
-	\caption{Genotype-phenotype maps for the nine most significant interactions.}
+	\caption{Genotype-phenotype maps for the nine most significant interactions. p1 and p2 are MAF for SNP 1 and SNP 2 respectively}
 	\label{fig:3dplots}

### 2013-04-24 docs/manuscript/report_v1.tex
added some text on the simulations / fine mapping. Need to wait until some more analyses finish before I can include full results here

@@
 model.
-As before any snp pairs not matching the criteria of $r^2$  < 0.1, number of genotype classes = 9, and minimum genotype class size > 5 were removed from the output.
+As before any SNP pairs not matching the criteria of $r^2$  < 0.1, number of genotype classes = 9, and minimum genotype class size > 5 were removed from the output.
 
@@
 Therefore, fine mapped loci provide good functional candidates.
-However, as fine mapping includes a large number of additional tests increases in the interaction (4d.f.) $p$-values could be due to chance [need a better way of describing this].
+Hereon, results from the original 2D scan using genotyped SNPs will be refered to as genotype pairs and results from the fine mapping 2D scan using imputed data will be termed imputed pairs.
+Thus, for each significant pair there will be a results from genotyped and imputed pairs.
+However, as fine mapping includes a large number of additional tests, any oberved increases in the interaction (4d.f.) $p$-values could be due to chance arising from multiple testing.
+To address this we ran a series of simulations to determine the empirical change from genotype interaction $p$-value to imputed interaction $p$-value expected in situations where no epistatic exists.
+For the following three scenarios; $cis-cis$, $cis-trans$ and $trans-trans$ we randomly selected a probe for which no significant epistatic genotype pairs had been identified and randomly selected two SNP blocks within the defined genomic regions.
+The SNP pair was tested using the 4 and 8d.f.
+models and then imputed SNPs were extracted from 100kb upstream and downstream regions.
+Pairwise interactions were tested for the SNPs in the two imputed blocks and the SNP pair with the minimum interaction $p$-value retained.
+This process was repeated 10,000 times for each of the three scenarios.
 

### 2013-05-14 docs/manuscript/report_v1.tex
minor changes to the report
included some tables and figure on the replication analysis

@@
 
-We attempted replication of the signals in an independent dataset [Greg Gibson - 139 individuals, 185 out of 432 interactions had both SNPs in common in the two datasets].
-We observed no significant $p$-values after adjusting for multiple testing (lowest $p$-value$=0.0099$ in replication dataset, threshold set at $0.05/185=0.0003$).
-We tested for enrichment of $p$-values but there was no evidence of departure from the null hypothesis of no effects (Figure \ref{fig:gg_replication}).
-
-\begin{figure}[p]
-	\centering
-	\includegraphics[width=15cm]{images/replication_qqplot}
-	\caption{Q-Q plot for interaction $p$-values (4 degrees of freedom) for 185 interactions that had SNPs in common in the replication dataset.}
-	\label{fig:gg_replication}
-\end{figure}
-
 \subsection{Additive vs non-additive}
@@
 
+\subsection{Replication}
+
+We attempted replication of the signals in an independent dataset [Greg Gibson - 139 individuals, 185 out of 432 interactions had both SNPs in common in the two datasets].
+We observed no significant $p$-values after adjusting for multiple testing (lowest $p$-value$=0.0099$ in replication dataset, threshold set at $0.05/185=0.0003$).
+We tested for enrichment of $p$-values but there was no evidence of departure from the null hypothesis of no effects (Figure \ref{fig:gg_replication}).
+
+Replication in Fehrmann and EGCUT
+
+Of the 432 probe by SNP pair associations supplied to the replication group(s) 387 and 391 were present at the desired criteria in Fehrmenn and EGCUT respectively.
+
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{rrrr}
+	\hline
+	& & Fehrmann & EGCUT \\
+	\hline
+	P value & Expected & Observed & Observed \\
+	\hline	  
+	0.1 & 38 & 51 & 67 \\
+	0.05 & 19 & 35 & 44 \\
+	0.01 & 4 & 22 & 26 \\
+	0.001 & <1 & 14 & 22 \\
+	0.0001 & ~0 & 7 & 17 \\
+	\hline
+	\end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrllrlrrrrr}
+  \hline
+chr1 & chr2 & pos1 & pos2 & snp1 & snp2 & probechr & probegene & snpcor & replication\_pnest & replication\_r & replication\_nclass & replication\_minclass \\ 
+  \hline
+19 &  19 & 483171 & 483155 & rs807491 & rs7254601 &  19 & TMEM149 & 0.02 & 81.55 & 0.05 &   9 &  16 \\ 
+19 &  19 & 481151 & 481145 & rs873870 & rs7252981 &  19 & ATP13A1 & -0.26 & 48.72 & -0.34 &   9 &  12 \\ 
+11 &  11 & 345268 & 345257 & rs7930237 & rs556895 &  11 & CTSC & -0.11 & 18.76 & 0.09 &   9 &   8 \\ 
+10 &  10 & 317649 & 317615 & rs2395095 & rs10824092 &  10 & ADK & 0.12 & 18.33 & 0.10 &   9 &  13 \\ 
+8 &   8 & 277063 & 277044 & rs2123758 & rs3889129 &   8 & NAPRT1 & 0.23 & 15.12 & 0.18 &   9 &  34 \\ 
+3 &   3 & 110710 & 110699 & rs16864367 & rs13069559 &   3 & MBNL1 & 0.29 & 10.30 & 0.23 &   9 &   9 \\ 
+14 &   3 & 416822 & 110699 & rs2614467 & rs13069559 &   3 & MBNL1 & -0.02 & 4.13 & 0.03 &   9 &  12 \\ 
+19 &   8 & 483165 & 252775 & rs8106959 & rs1843357 &  19 & TMEM149 & 0.02 & 3.72 & -0.00 &   9 &  10 \\ 
+19 &   3 & 483165 & 117001 & rs8106959 & rs10937361 &  19 & TMEM149 & 0.01 & 3.61 & -0.00 &   9 &  12 \\ 
+8 &   3 & 257325 & 117283 & rs8180944 & rs4553956 &   8 & HMBOX1 & 0.05 & 3.38 & -0.05 &   9 &   4 \\ 
+19 &   1 & 483165 & 38950 & rs8106959 & rs914940 &  19 & TMEM149 & 0.04 & 3.36 & -0.05 &   9 &   8 \\ 
+19 &   2 & 483165 & 80806 & rs8106959 & rs6718480 &  19 & TMEM149 & 0.02 & 3.31 & -0.02 &   9 &  10 \\ 
+19 &   6 & 483165 & 216524 & rs8106959 & rs6926382 &  19 & TMEM149 & -0.05 & 3.06 & -0.03 &   9 &  12 \\ 
+19 &  14 & 483165 & 414535 & rs8106959 & rs17719594 &  19 & TMEM149 & 0.04 & 3.06 & 0.02 &   9 &  12 \\ 
+5 &   3 & 155373 & 110699 & rs7710738 & rs13069559 &   3 & MBNL1 & 0.10 & 2.55 & -0.03 &   9 &   9 \\ 
+19 &  12 & 483165 & 379273 & rs8106959 & rs1401098 &  19 & TMEM149 & -0.01 & 2.41 & 0.04 &   9 &  11 \\ 
+17 &  15 & 450341 & 430903 & rs1533031 & rs12591171 &  17 & XAF1 & -0.04 & 2.38 & -0.03 &   9 &   9 \\ 
+14 &   8 & 406922 & 265704 & rs7152284 & rs2896452 &   8 & C8ORF59 & 0.02 & 2.18 & 0.07 &   9 &  17 \\ 
+19 &   5 & 483884 & 178750 & rs4803481 & rs2421050 &  19 & CEACAM21 & 0.07 & 2.16 & -0.02 &   9 &   9 \\ 
+7 &   1 & 242078 & 11854 & rs12532999 & rs12065581 &   1 & GPR177 & -0.08 & 2.13 & 0.04 &   9 &  10 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrllrlrrrrr}
+  \hline
+chr1 & chr2 & pos1 & pos2 & snp1 & snp2 & probechr & probegene & snpcor & replication\_pnest & replication\_r & replication\_nclass & replication\_minclass \\ 
+  \hline
+19 &  19 & 483171 & 483155 & rs807491 & rs7254601 &  19 & TMEM149 & 0.02 & 45.78 & -0.13 &   9 &   3 \\ 
+19 &  19 & 481151 & 481145 & rs873870 & rs7252981 &  19 & ATP13A1 & -0.26 & 42.37 & 0.37 &   9 &   3 \\ 
+10 &  10 & 317649 & 317615 & rs2395095 & rs10824092 &  10 & ADK & 0.12 & 21.21 & -0.04 &   9 &  11 \\ 
+3 &   3 & 110710 & 110699 & rs16864367 & rs13069559 &   3 & MBNL1 & 0.29 & 20.36 & 0.18 &   9 &   4 \\ 
+8 &   8 & 277063 & 277044 & rs2123758 & rs3889129 &   8 & NAPRT1 & 0.23 & 16.08 & 0.01 &   9 &  29 \\ 
+11 &  11 & 345268 & 345257 & rs7930237 & rs556895 &  11 & CTSC & -0.11 & 15.06 & 0.19 &   9 &   2 \\ 
+19 &   4 & 483165 & 137487 & rs8106959 & rs2351458 &  19 & TMEM149 & 0.03 & 9.61 & -0.02 &   9 &   4 \\ 
+19 &   6 & 483165 & 216524 & rs8106959 & rs6926382 &  19 & TMEM149 & -0.05 & 8.80 & -0.03 &   9 &   6 \\ 
+5 &   3 & 155373 & 110699 & rs7710738 & rs13069559 &   3 & MBNL1 & 0.10 & 7.89 & 0.05 &   9 &   4 \\ 
+19 &   1 & 483165 & 38950 & rs8106959 & rs914940 &  19 & TMEM149 & 0.04 & 6.96 & 0.01 &   9 &   3 \\ 
+17 &   3 & 450326 & 110699 & rs218671 & rs13069559 &   3 & MBNL1 & -0.03 & 5.82 & 0.00 &   9 &   4 \\ 
+6 &   3 & 207083 & 110699 & rs2030926 & rs13069559 &   3 & MBNL1 & 0.05 & 5.80 & -0.00 &   9 &   1 \\ 
+19 &  13 & 483165 & 380931 & rs8106959 & rs9509428 &  19 & TMEM149 & -0.01 & 5.75 & -0.01 &   9 &   2 \\ 
+7 &   3 & 236869 & 110699 & rs11981513 & rs13069559 &   3 & MBNL1 & 0.02 & 5.36 & 0.00 &   9 &   5 \\ 
+19 &   2 & 483165 & 80806 & rs8106959 & rs6718480 &  19 & TMEM149 & 0.02 & 5.15 & -0.07 &   9 &   1 \\ 
+4 &   3 & 126893 & 110699 & rs4392535 & rs13069559 &   3 & MBNL1 & 0.02 & 4.33 & -0.02 &   9 &   1 \\ 
+8 &   3 & 248082 & 110699 & rs4735830 & rs13069559 &   3 & MBNL1 & 0.03 & 4.21 & -0.05 &   9 &   4 \\ 
+19 &  11 & 483165 & 353621 & rs8106959 & rs471728 &  19 & TMEM149 & 0.04 & 3.62 & 0.04 &   9 &   4 \\ 
+19 &   8 & 483165 & 252775 & rs8106959 & rs1843357 &  19 & TMEM149 & 0.02 & 3.33 & -0.04 &   9 &   2 \\ 
+19 &  17 & 483165 & 462278 & rs8106959 & rs7213338 &  19 & TMEM149 & 0.04 & 3.14 & 0.04 &   9 &   7 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/replication_qqplot}
+	\caption{Q-Q plot for interaction $p$-values (4 degrees of freedom) for 185 interactions that had SNPs in common in the replication dataset.}
+	\label{fig:gg_replication}
+\end{figure}
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/EGCUT_vs_Fehrmann_all}
+	\caption{-log10 pvalues for all snp / probe pairs in Fehrnmann and EGCUT}
+	\label{fig:repl_match}
+\end{figure}
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/EGCUT_vs_Fehrmann_pval10}
+	\caption{-log10 pvalues for all snp / probe pairs with -log10 p values less than 10 in Fehrnmann and EGCUT datasets}
+	\label{fig:repl_match10}
+\end{figure}
+
 \subsection{Fine Mapping}
@@
 %%% Refs
+\clearpage
 \bibliographystyle{plain}

### 2013-05-19 docs/replication_analysis/analyse_replication.Rnw
analysing replication results

\documentclass[paper=a4, fontsize=11pt]{scrartcl}
\title{Analysis of initial replication results}

\begin{document}
\maketitle

\section{First look}

<<read_summary, eval=TRUE, echo=FALSE, warning=FALSE, cache=TRUE>>=
source("~/repo/eQTL-2D/replication/results/summary_script.R")
@

\end{document}

### 2013-05-19 replication/results/summary_script.R

@@ -1,151 +1,286 @@
 # Summary of the replication results
 # Results sent by Harm-Jan Westra	
 
-
-# Read in RData files
 library(reshape2)
 library(ggplot2)
-
-load("EGCUT_replication.RData")
-newsig$id <- with(newsig, paste(probename, snp1, snp2))
-EGCUT <- newsig
-
-load("FehrmannHT12v3_replication.RData")
-newsig$id <- with(newsig, paste(probename, snp1, snp2))
-Feh <- subset(newsig, select=-c(chr1, chr2, pos1, pos2, snp1, snp2, complete, probeid, probename, probechr, probegene, probehsq, minclasssize, snpcor, propG, propA, pnest, pfull, pint))
-rm(newsig)
-
-rep <- merge(EGCUT, Feh, by="id", suff=c("_EGCUT", "_Feh"), all=T)
-dim(rep)
-
-# Pairwise correlations between 
-cor(subset(rep, select=c(pnest, replication_pnest_EGCUT, replication_pnest_Feh)), use="pair")
+library(VennDiagram)
 
 
-panel.cor <- function(x, y, digits=2, prefix="", cex.cor)
+
+#'<brief desc>
+#'
+#'<full description>
+#' @param filename <what param does>
+#' @param  objname <what param does>
+#' @param  setname <what param does>
+#' @export
+#' @keywords
+#' @seealso
+#' @return
+#' @alias
+#' @examples \dontrun{
+#'
+#'}
+ReadOrig <- function(filename, objname, setname)
 {
-    usr <- par("usr"); on.exit(par(usr))
-    par(usr = c(0, 1, 0, 1))
-    r <- abs(cor(x, y, use="pair"))
-    txt <- format(c(r, 0.123456789), digits=digits)[1]
-    txt <- paste(prefix, txt, sep="")
-    if(missing(cex.cor)) cex <- 0.8/strwidth(txt)
+	load(filename)
+	sig <- get(objname)
 
-    test <- cor.test(x,y)
-    text(0.5, 0.5, txt, cex = cex)
+	sig <- subset(sig, select=c(chr1, chr2, snp1, snp2, probename, probegene, probechr, pfull, pnest))
+	sig$set <- setname
[…]

### 2013-05-20 filtering/filter_rdata_bychr.R
completed filtering, ready to send replication scripts

@@ -92,6 +92,12 @@ dim(res)
 if(nrow(res) == 1 & is.na(res$df1[1])) q()
 
 
+# Filter on threshold
+res <- subset(res, pfull > threshold)
+dim(res)
+if(nrow(res) == 0) q()
+
+
 # Remove non-autosomes
 res <- subset(res, chr1 %in% 1:22 & chr2 %in% 1:22)
 dim(res)

### 2013-05-20 filtering/filter_rdata_nonmarginal_sig.R

@@ -96,6 +96,12 @@ dim(res)
 if(nrow(res) == 1 & is.na(res$df1[1])) q()
 
 
+# Filter on threshold
+res <- subset(res, pfull > threshold)
+dim(res)
+if(nrow(res) == 0) q()
+
+
 # Remove non-autosomes
 res <- subset(res, chr1 %in% 1:22 & chr2 %in% 1:22)
 dim(res)
@@ -110,7 +116,7 @@ dim(marginal_list)
 
 
 # Filter so that any SNPs in snplist for a particular probe are removed
-marginal_list <- subset(marginal_list, probename == probeinfo$PROBE_ID[i])
+marginal_list <- subset(marginal_list, probename == res$probename[i])
 dim(marginal_list)
 if(nrow(marginal_list) != 0)
 {

### 2013-05-20 filtering/make_sig.R

@@ -25,5 +25,18 @@ SentinalSnp <- function(dat)
 sig <- SentinalSnp(sig)
 dim(sig)
 
+removeFactors <- function(x)
+{
+	nom <- names(x)
+	l <- nom[lapply(x, class)=="factor"]
+	for(i in 1:length(l))
+	{
+		x[[l[i]]] <- as.character(x[[l[i]]])
+	}
+	return(x)
+}
+
+sig <- removeFactors(sig)
+
 save(sig, file=outfile)
 

### 2013-05-20 filtering/make_sig_list.R

@@ -1,0 +1,78 @@
+
+removeFactors <- function(x)
+{
+    nom <- names(x)
+    l <- nom[lapply(x, class)=="factor"]
+    for(i in 1:length(l))
+    {
+	x[[l[i]]] <- as.character(x[[l[i]]])
+    }
+    return(x)
+}
+
+
+load("nonmarginal_interaction_list.RData")
+sig1 <- removeFactors(sig)
+load("old/sig.RData")
+sig <- removeFactors(sig)
+
+
+sig$code <- with(sig, paste(probename, snp1, snp2))
+sig1$code <- with(sig1, paste(probename, snp1, snp2))
+
+table(sig1$code %in% sig$code)
+table(sig$code %in% sig1$code)
+
+subset(sig, ! code %in% sig1$code)
+
+sig <- rbind(sig, sig1)
+sig <- subset(sig, !duplicated(code))
+dim(sig)
+
+table(table(sig$probename))
+
+signm <- sig
+signm$filter <- 2
+
+load("../../replication/run/interactions_list.RData")
+sig <- removeFactors(sig)
+sig$filter <- 1
+
+sig <- rbind(sig, subset(signm, select=-code))
+sig$code <- with(sig, paste(probename, snp1, snp2))
+table(duplicated(sig$code), sig$filter)
+
+sig <- subset(sig, !duplicated(code))
+
+
+# Get random list
+
+load("~/repo/eQTL-2D/data/residuals_all.RData")
+bim <- read.table("~/repo/eQTL-2D/data/clean_geno_final.bim", colClasses="character")
+
+sig2 <- sig
+sig2$filter <- 3
+bim <- subset(bim, V1 %in% 1:22)
+index1 <- sample(1:nrow(bim), nrow(sig2))
+index2 <- sample(1:nrow(bim), nrow(sig2))
+table(index1 %in% index2)
+
+sig2$chr1 <- as.numeric(bim$V1[index1])
+sig2$snp1 <- bim$V2[index1]
+sig2$pos1 <- index1
+sig2$chr2 <- as.numeric(bim$V1[index2])
+sig2$snp2 <- bim$V2[index2]
+sig2$pos2 <- index2
+sig2$pfull <- sig2$pint <- sig2$df1 <- sig2$df2 <- sig2$minclasssize <- sig2$propG <- sig2$propA <- NA
+
+sig2$code <- with(sig2, paste(probename, snp1, snp2))
+head(sig2)
[…]

### 2013-05-20 filtering/submit_filter_nonmarginal_sig.sh

@@ -6,7 +6,7 @@
 #$ -t 1-7339
 #$ -o job_reports/
 #$ -e job_reports/
-#$ -l h_vmem=5G
+#$ -l h_vmem=10G
 
 
 set -e

### 2013-05-20 replication/R/functions.R

@@ -138,7 +138,7 @@ DataChecks <- function(probes, geno)
 
 #' Replication statistical tests
 #'
-#' Calculates allele frequencies, correlation between SNPs, class sizes
+#' Calculates allele frequencies, correlation between SNPs, class sizes, variance components
 #'
 #' @param geno \code{matrix} of genotype data
 #' @param probes \code{data.frame} of probes

### 2013-05-20 docs/replication_analysis/analyse_replication.Rnw

@@
-\documentclass[paper=a4, fontsize=11pt]{scrartcl}
-\title{Analysis of initial replication results}
+% grDevices::pdf.options(useDingbats = FALSE)
+% require(knitr)
+% opts_knit$set(concordance = TRUE)
+% knit('report.Rnw', encoding='UTF-8')
+
+\documentclass{article}
+\usepackage{subfig}
+\usepackage{rotating}
 
 \begin{document}
+
+\title{The effect of inbreeding on BMI and height}
+
 \maketitle
 
-\section{First look}
+<<initilise, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE>>=
 
-<<read_summary, eval=TRUE, echo=FALSE, warning=FALSE, cache=TRUE>>=
-source("~/repo/eQTL-2D/replication/results/summary_script.R")
+# Summary of the replication results
+# Results sent by Harm-Jan Westra
+
+library(reshape2)
+library(ggplot2)
+library(VennDiagram)
+library(xtable)
+
+#'<brief desc>
+#'
+#'<full description>
+#' @param filename <what param does>
+#' @param  objname <what param does>
+#' @param  setname <what param does>
+#' @export
+#' @keywords
+#' @seealso
+#' @return
+#' @alias
+#' @examples \dontrun{
+#'
+#'}
+ReadOrig <- function(filename, objname, setname)
+{
+	load(filename)
+	sig <- get(objname)
+
+	sig <- subset(sig, select=c(chr1, chr2, snp1, snp2, probename, probegene, probechr, pfull, pnest))
+	sig$set <- setname
+	sig$code <- with(sig, paste(probename, snp1, snp2))
+	return(sig)
+}
+
+#'<brief desc>
+#'
+#'<full description>
+#' @param filename <what param does>
+#' @param  objname <what param does>
+#' @param  setname <what param does>
+#' @export
+#' @keywords
+#' @seealso
+#' @return
+#' @alias
+#' @examples \dontrun{
+#'
+#'}
+ReadRep <- function(filename, objname, setname)
+{
+	load(filename)
+	sig <- get(objname)
+
+	a <- sum(sig$replication_r^2 > 0.1)
+	# cat(a, "pairs with high LD\n")
+
+	b <- sum(sig$replication_nclass != 9)
+	# cat(b, "pairs with < 9 classes\n")
+
+	sig <- subset(sig, replication_r^2 < 0.1 & replication_nclass == 9, select=c(chr1, chr2, snp1, snp2, probename, probegene, probechr, replication_pfull, replication_pnest))
+
+	names(sig) <- c("chr1", "chr2", "snp1", "snp2", "probename", "probegene", "probechr", "pfull", "pnest")
+	sig$set <- setname
+	sig$code <- with(sig, paste(probename, snp1, snp2))
+	return(sig)
+}
+
+confInt <- function(n, alpha)
+{
+	k <- c(1:n)
+	upper <- -log10(qbeta(alpha/2,k,n+1-k))
+	lower <- -log10(qbeta((1-alpha/2),k,n+1-k))
+	expect <- -log10((k-0.5)/n)
+	return(data.frame(expect, lower, upper))
+}
+
+qqDat <- function(sig, alpha)
+{
+	sig <- sig[order(sig$pnest, decreasing=T), ]
+	ci <- confInt(nrow(sig), alpha)
+	sig <- data.frame(sig, ci)
+	return(sig)
+}
+
+panelCor <- function(x, y, digits=2, prefix="", cex.cor)
+{
+	usr <- par("usr"); on.exit(par(usr))
+	par(usr = c(0, 1, 0, 1))
+	r <- cor(x, y, use="pair")
+	txt <- format(c(r, 0.123456789), digits=digits)[1]
+	txt <- paste(prefix, txt, sep="")
+	if(missing(cex.cor))
+	{
+		cex <- 0.8/strwidth(txt)
+	}
+	test <- cor.test(x,y)
+	text(0.5, 0.5, txt, cex = cex)
+}
+
+uniteData <- function(..., pval)
+{
+	l <- list(...)
+
+	# Get all pairs in common
+	plist <- lapply(l, function(x) { return(x$code)})
+	plist <- Reduce(intersect, plist)
+
+	l <- lapply(l, function(x)
+	{
+		x <- subset(x, code %in% plist)
+		x <- x[order(x$code), ]
+	})
+
+	a <- data.frame(code=l[[1]]$code)
+	for(i in 1:length(l))
+	{
+		a <- data.frame(a, l[[i]][[pval]])
+		names(a)[i+1] <- l[[i]]$set[1]
+	}
+	return(a)
+}
+
+plotCor <- function(ud, m=NULL)
+{
+	ud <- subset(ud, select=-c(code))
+	ud <- as.matrix(ud)
+	ud[is.infinite(ud)] <- NA
+	a <- pairs(ud, lower.panel=panel.smooth, upper.panel=panelCor, main=m)
+	return(a)
+}
+
+makeBonfTable <- function(fehr, egcut, marginal_list, top)
+{
+	# Choose the top 20 SNPs based on pnest from each set
+	f <- fehr[order(fehr$pnest, decreasing=TRUE), ][1:top, ]
+	e <- egcut[order(egcut$pnest, decreasing=TRUE), ][1:top, ]
+
+	# Show the Bonferroni significant replications
+	index <- f$pnest > -log10(0.05/nrow(fehr))
+	f$set2 <- f$set
+	f$set2[index] <- paste(f$set[index], "*", sep="")
+	index <- e$pnest > -log10(0.05/nrow(fehr))
+	e$set2 <- e$set
+	e$set2[index] <- paste(e$set[index], "*", sep="")
+
+	# Find the interactions common to both SNP sets
+	index <- f$code %in% e$code
+	nom <- f$code[index]
+	f1 <- subset(f, code %in% nom)
+	e1 <- subset(e, code %in% nom)
+	f1 <- f1[order(f1$code), ]
+	e1 <- e1[order(e1$code), ]
+	f1$set2 <- paste(f1$set2, "+", e1$set2)
+	fe <- rbind(f1, subset(e, ! code %in% nom), subset(f, ! code %in% nom))
+
+	# Characterise the cis/trans types
+	fe$type <- "cis-cis"
+	fe$type[fe$chr1 == fe$probechr & fe$chr2 != fe$probechr] <- "cis-trans"
+	fe$type[fe$chr2 == fe$probechr & fe$chr1 != fe$probechr] <- "trans-cis"
+	fe$type[fe$chr2 != fe$probechr & fe$chr1 != fe$probechr] <- "trans-trans"
+
+	# Characterise new SNPs vs known SNPs
+	fe$code1 <- paste(fe$snp1, fe$probename)
+	fe$code2 <- paste(fe$snp2, fe$probename)
+	marginal_list$code <- paste(marginal_list$snp, marginal_list$probename)
+	fe$margins <- "known-known"
+	fe$margins[fe$code1 %in% marginal_list$code & ! fe$code2 %in% marginal_list$code] <- "known-new"
+	fe$margins[! fe$code1 %in% marginal_list$code & fe$code2 %in% marginal_list$code] <- "new-known"
+	fe$margins[! fe$code1 %in% marginal_list$code & ! fe$code2 %in% marginal_list$code] <- "new-new"
+
+	# Tidy up the table
+	fe$pg <- with(fe, paste(probegene, " (chr", probechr, ")", sep=""))
+	fe$s1 <- with(fe, paste(snp1, " (chr", chr1, ")", sep=""))
+	fe$s2 <- with(fe, paste(snp2, " (chr", chr2, ")", sep=""))
+
+	fe <- subset(fe, select=c(pg, s1, s2, set2, type, margins))
+	fe <- fe[order(fe$pg, fe$set2), ]
+	names(fe) <- c("Probe gene", "SNP1", "SNP2", "Replication", "Type", "Previous associations")
+
+	return(fe)
+}
+
+# Load Data
+
+load("~/repo/eQTL-2D/filtering/marginal_lists/marginal_list.RData")
+bsgs <- ReadOrig(
+	"~/repo/eQTL-2D/replication/run/interactions_list.RData",
+	"sig",
+	"BSGS"
+)
+egcut <- ReadRep(
+	"~/repo/eQTL-2D/replication/results/EGCUT_replication.RData",
+	"newsig",
+	"EGCUT"
+)
+fehr <- ReadRep(
+	"~/repo/eQTL-2D/replication/results/FehrmannHT12v3_replication.RData",
+	"newsig",
+	"Ferhmann"
+)
+
+@
+
+<<qqplots, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, fig.cap="Q-Q plots of interaction p-values from replication dataset", fig.subcap=c("All signals", "Signals excluding 20 most significant replications"), out.width='.4\\linewidth'>>=
+
+# Q-Q confidence intervals
+egcut <- qqDat(egcut, 0.05)
+fehr <- qqDat(fehr, 0.05)
+
+# Bonferroni corrected significant replication
+egcut_b <- subset(egcut, pnest > -log10(0.05 / nrow(egcut)))
+fehr_b <- subset(fehr, pnest > -log10(0.05 / nrow(fehr)))
+
+# FDR corrected significant replication
+egcut_f <- subset(egcut, pnest > upper)
+fehr_f <- subset(fehr, pnest > upper)
+
+# Q-Q plot
+
+# Show bonferroni correction
+dat1 <- rbind(fehr, egcut)
+dat1$ex <- FALSE
+dat1$ex[dat1$pnest > -log10(0.05/(nrow(dat1)/2))] <- TRUE
+
+qqplot1 <- ggplot(dat1) +
+	geom_ribbon(aes(x=expect, ymin=lower, ymax=upper), colour="white", alpha=0.5) +
+	geom_abline(intercept=0, slope=1) +
+	geom_point(aes(x=expect, y=pnest, colour=ex)) +
+	geom_hline(yintercept=-log10(0.05/(nrow(dat1)/2)), linetype="dotted") +
+	facet_grid(set ~ .) +
+	labs(colour="Above 5% Bonferroni?", y="Observed", x="Expected")
+
+# Show FDR correction
+dat2 <- rbind(fehr[-c(1:20), ], egcut[-c(1:20), ])
+dat2$ex <- FALSE
+dat2$ex[dat2$pnest > dat2$upper] <- TRUE
+
+qqplot2 <- ggplot(dat2) +
+	geom_ribbon(aes(x=expect, ymin=lower, ymax=upper), colour="white", alpha=0.5) +
+	geom_abline(intercept=0, slope=1) +
+	geom_point(aes(x=expect, y=pnest, colour=ex), size=1.5) +
+	facet_grid(set ~ .) +
+	labs(colour="Above FDR 5% CI?", y="Observed", x="Expected")
+
+qqplot1
+qqplot2
+
+@
+
+<<overlapplots, eval=TRUE, echo=FALSE, warning=FALSE, cache=TRUE, message=FALSE, fig.cap='Overlap in signals between replication datasets', fig.subcap=c('Bonferroni significance', '5\\% FDR significance'), out.width='.4\\linewidth'>>=
+
+# Characterise overlap
+# Significant in only Fehr, Significant in only EGCUT, Significant in both
+
+# Make venn diagram
+
+bonf.venn <- draw.triple.venn(
+	area1 = nrow(bsgs),
+	area2 = nrow(fehr_b),
+	area3 = nrow(egcut_b),
+	n12   = sum(fehr_b$code %in% bsgs$code),
+	n23  = sum(egcut_b$code %in% fehr_b$code),
+	n13  = sum(egcut_b$code %in% bsgs$code),
+	n123 = sum(egcut_b$code %in% fehr_b$code),
+	category=c("BSGS", "Fehrmann", "EGCUT"),
+	fill = c("blue", "red", "green"),
+	alpha=c(0.1, 0.1, 0.1),
+	lty = "blank",
+	cex = 2,
+	cat.cex = 2,
+	cat.col = c("blue", "red", "green"),
+	ind=FALSE
+)
+
+fdr.venn <- draw.triple.venn(
+	area1 = nrow(bsgs),
+	area2 = nrow(fehr_f),
+	area3 = nrow(egcut_f),
+	n12   = sum(fehr_f$code %in% bsgs$code),
+	n23  = sum(egcut_f$code %in% fehr_f$code),
+	n13  = sum(egcut_f$code %in% bsgs$code),
+	n123 = sum(egcut_f$code %in% fehr_f$code),
+	category=c("BSGS", "Fehrmann", "EGCUT"),
+	fill = c("blue", "red", "green"),
+	alpha=c(0.1, 0.1, 0.1),
+	lty = "blank",
+	cex = 2,
+	cat.cex = 2,
+	cat.col = c("blue", "red", "green"),
+	ind=FALSE
+)
+
+plot.new() 
+grid.draw(bonf.venn)
+
+plot.new() 
+grid.draw(fdr.venn)
+
+@
+
+<<corpfull, eval=TRUE, echo=FALSE, warning=FALSE, cache=TRUE, fig.cap='Full genetic effects correlate between datasets (8d.f.
+p-values)', fig.subcap=c('All signals', 'Signals that surpass 5\\% FDR in both datasets'), out.width='.4\\linewidth'>>=
+
+# Plot correlations of pfull and pnest
+
+pfull_all <- uniteData(bsgs, egcut, fehr, pval="pfull")
+pfull_f <- uniteData(bsgs, egcut_f, fehr_f, pval="pfull")
+pnest_all <- uniteData(bsgs, egcut, fehr, pval="pnest")
+pnest_f <- uniteData(bsgs, egcut_f, fehr_f, pval="pnest")
+
+plotCor(pfull_all)
+plotCor(pfull_f)
+
+@
+
+<<corpnest, eval=TRUE, echo=FALSE, warning=FALSE, cache=TRUE, fig.cap='Full genetic effects correlate between datasets (8d.f.
+p-values)', fig.subcap=c('All signals', 'Signals that surpass 5\\% FDR in both datasets'), out.width='.4\\linewidth'>>=
+
+plotCor(pnest_all)
+plotCor(pnest_f)
+
+@
+
+<<interactiontable, cache=TRUE, eval=TRUE, echo=FALSE, results='asis', size='tiny'>>=
+# List of top interactions
+# show bonferroni correction, previous associations, type
+
+bonf.tab <- makeBonfTable(fehr, egcut, marginal_list, 20)
+overlap <- table(gsub(bonf.tab$Replication, pattern="\\*", replacement=""))
+dum <- bonf.tab$Type
+dum[dum == "trans-cis"] <- "cis-trans"
+type <- table(dum)
+prev <- table(bonf.tab$Prev)
+
+print(xtable(bonf.tab, caption="The 20 most significant interaction terms from both replication sets.
+* denotes Bonferroni level significance in the named data set."), include.rownames=FALSE, latex.environments=c("center", "footnotesize"), floating.environment="sidewaystable")
 @

### 2013-05-22 docs/report1/report_v1.tex
Reorganised documents folder

--- 2013-05-14 docs/manuscript/report_v1.tex
@@
 \usepackage{url}
+\usepackage{rotating}
+\usepackage{natbib}
 
@@
 
-The Brisbane Systems Genetics Study (BSGS) comprises 862 Individuals of European descent from 274 independent families \cite{pmid22563384}.
+The Brisbane Systems Genetics Study (BSGS) comprises 862 Individuals of European descent from 274 independent families \citep{pmid22563384}.
 DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland.
 Full details of genotyping procedures are given in Medland et al.
-\cite{Medland2009} Standard QC filters were applied and the remaining 528,509 SNPs were carried forward further analysis.
+\citep{Medland2009} Standard QC filters were applied and the remaining 528,509 SNPs were carried forward further analysis.
 
@@
 
-\begin{table}[ht]
+\begin{sidewaystable}[ht]
 \centering
+{\footnotesize
 \begin{tabular}{rrrrrllrlrrrrr}
   \hline
-chr1 & chr2 & pos1 & pos2 & snp1 & snp2 & probechr & probegene & snpcor & replication\_pnest & replication\_r & replication\_nclass & replication\_minclass \\ 
+chr1 & chr2 & pos1 & pos2 & snp1 & snp2 & probechr & probegene & snpcor & rep\_pnest & rep\_r & rep\_nclass & rep\_minclass \\ 
   \hline
@@
 \end{tabular}
-\end{table}
+}
+\end{sidewaystable}
 
-\begin{table}[ht]
+\begin{sidewaystable}[ht]
 \centering
+{\footnotesize
 \begin{tabular}{rrrrrllrlrrrrr}
   \hline
-chr1 & chr2 & pos1 & pos2 & snp1 & snp2 & probechr & probegene & snpcor & replication\_pnest & replication\_r & replication\_nclass & replication\_minclass \\ 
+chr1 & chr2 & pos1 & pos2 & snp1 & snp2 & probechr & probegene & snpcor & rep\_pnest & rep\_r & replication\_nclass & rep\_minclass \\ 
   \hline
@@
 \end{tabular}
-\end{table}
+}
+\end{sidewaystable}
 
@@
 
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/pfull_rep_pvals}
+	\caption{8df pvals for 391 interactions}
+	\label{fig:cor_8df}
+\end{figure}
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/pnest_rep_pvals}
+	\caption{4df pvals for 391 interactions}
+	\label{fig:cor_4df}
+\end{figure}
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/qqplot_replication}
+	\caption{QQ plot excluding 50 largest pvals which really distort the plot}
+	\label{fig:cor_8df}
+\end{figure}
+
+% color_epi_snps.pdf
+% EpiSNPs_cell_type_spec_for_top_tissues.pdf
+% EpiSNPs_H3K4me3_Overlap.pdf
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/color_epi_snps}
+	\caption{Global view of all interactions}
+	\label{fig:cor_8df}
+\end{figure}
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/EpiSNPs_H3K4me3_Overlap}
+	\caption{}
+	\label{fig:cor_8df}
+\end{figure}
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=15cm]{images/EpiSNPs_cell_type_spec_for_top_tissues}
+	\caption{}
+	\label{fig:cor_8df}
+\end{figure}
+
 \subsection{Fine Mapping}

### 2013-05-22 analysis/3dbars.R

@@ -246,33 +246,6 @@ dev.off()
 
 sig$probeid <- match(sig$probename, colnames(resphen))
 
-a <- subset(bsgs, probegene == "TMEM149")
-
-l <- list()
-for(i in 1:nrow(a))
-{
-	l[[i]] <- plot3D(a[i,], xmat, resphen)[[3]]
-}
-
-
-b <- subset(bsgs, probegene == "MBNL1")
-
-l2 <- list()
-for(i in 1:nrow(b))
-{
-	l2[[i]] <- plot3D(a[i,], xmat, resphen)[[3]]
-}
-
-pdf("~/repo/eQTL-2D/analysis/TMEM149.pdf", width=15, height=15)
-do.call(grid.arrange, l)
-dev.off()
-
-pdf("~/repo/eQTL-2D/analysis/MBNL1.pdf", width=15, height=15)
-do.call(grid.arrange, l2)
-dev.off()
-
-
-
 sig2 <- subset(sig, rep == "Bonf")
 sig2$probeid <- match(sig2$probename, colnames(resphen))
 grid.arrange(

### 2013-05-22 analysis/circles.R

@@ -102,79 +102,6 @@ qqDat <- function(sig, alpha)
 	return(sig)
 }
 
-replicationOverlap <- function(bsgs, fehr, egcut)
-{
-	# Make a variable that indicates level of replication
-	# - not present
-	# - no replication
-	# - 5% FDR in one
-	# - 5% FDR in two
-	# - Bonferroni in one
-	# - Bonferroni in two
-
-	egcut$bonf <- FALSE
-	egcut$bonf[egcut$pnest > -log10(0.05/nrow(egcut))] <- TRUE
-	fehr$bonf <- FALSE
-	fehr$bonf[fehr$pnest > -log10(0.05/nrow(fehr))] <- TRUE
-
-	egcut$fdr <- FALSE
-	egcut$fdr[egcut$pnest > egcut$upper] <- TRUE
-	fehr$fdr <- FALSE
-	fehr$fdr[fehr$pnest > fehr$upper] <- TRUE
-
-
-	bsgs$rep <- "None"
-	bsgs$sets <- 0
-
-	# FDR in one
-
-	bsgs$rep[bsgs$code %in% egcut$code[egcut$fdr] | bsgs$code %in% fehr$code[fehr$fdr]] <- "FDR"
-	bsgs$sets[bsgs$code %in% egcut$code[egcut$fdr] | bsgs$code %in% fehr$code[fehr$fdr]] <- 1
-	bsgs$sets[bsgs$code %in% egcut$code[egcut$fdr] & bsgs$code %in% fehr$code[fehr$fdr]] <- 2
-
-	bsgs$rep[bsgs$code %in% egcut$code[egcut$bonf] | bsgs$code %in% fehr$code[fehr$bonf]] <- "Bonf"
-	bsgs$sets[bsgs$code %in% egcut$code[egcut$bonf] | bsgs$code %in% fehr$code[fehr$bonf]] <- 1
-	bsgs$sets[bsgs$code %in% egcut$code[egcut$bonf] & bsgs$code %in% fehr$code[fehr$bonf]] <- 2
-
-	table(bsgs$rep, bsgs$sets)
-	return(bsgs)
-}
-
-newAssoc <- function(sig, marginal_list)
-{
-	sig$code1 <- paste(sig$snp1, sig$probename)
-	sig$code2 <- paste(sig$snp2, sig$probename)
-	marginal_list$code <- paste(marginal_list$snp, marginal_list$probename)
-	sig$known1 <- "known"
[…]

### 2013-05-22 analysis/functional_analysis.R

@@ -1,0 +1,61 @@
+library(lattice)
+library(latticeExtra)
+library(gridExtra)
+library(ggplot2)
+library(reshape2)
+library(plyr)
+
+
+plot3D <- function(res, geno, phen, z=45)
+{
+	a <- tapply(phen[,res$probeid], list(geno[,res$pos1], geno[, res$pos2]), mean)
+	b <- table(geno[,res$pos1], geno[, res$pos2])
+	p <- cloud(
+		a, 
+		panel.3d.cloud=panel.3dbars, 
+		col="black", 
+		col.facet=c("#e5f5e0", "#A1D99B", "#31A354"), 
+		xbase=0.6, ybase=0.6,
+		xlab="SNP1", ylab="SNP2", zlab="y",
+		default.scales=list(arrows=F),
+		screen = list(z = z, x = -60, y = 3),
+		main = paste(res$chr1, res$chr2)
+	)
+	return(list(a,b,p))
+}
+
+
+
+
+
+
+load("~/repo/eQTL-2D/analysis/replication_summary.RData")
+
+a <- subset(sig, probegene == "TMEM149")
+
+l <- list()
+for(i in 1:nrow(a))
+{
+	l[[i]] <- plot3D(a[i,], xmat, resphen)[[3]]
+}
+
+
+b <- subset(sig, probegene == "MBNL1")
+
+l2 <- list()
+for(i in 1:nrow(b))
+{
+	l2[[i]] <- plot3D(a[i,], xmat, resphen)[[3]]
+}
+
+pdf("~/repo/eQTL-2D/analysis/TMEM149.pdf", width=15, height=15)
+do.call(grid.arrange, l)
+dev.off()
+
+pdf("~/repo/eQTL-2D/analysis/MBNL1.pdf", width=15, height=15)
+do.call(grid.arrange, l2)
+dev.off()
+
+
+
+

### 2013-05-27 docs/replication_analysis/analyse_replication.Rnw
Analysing new replication results

--- 2013-05-19 docs/replication_analysis/analyse_replication.Rnw
@@
-\documentclass[paper=a4, fontsize=11pt]{scrartcl}
-\title{Analysis of initial replication results}
+% grDevices::pdf.options(useDingbats = FALSE)
+% require(knitr)
+% opts_knit$set(concordance = TRUE)
+% knit('report.Rnw', encoding='UTF-8')
+
+\documentclass{article}
+\usepackage{subfig}
+\usepackage{rotating}
 
 \begin{document}
+
+\title{The effect of inbreeding on BMI and height}
+
 \maketitle
 
-\section{First look}
+<<initialise, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE>>=
 
-<<read_summary, eval=TRUE, echo=FALSE, warning=FALSE, cache=TRUE>>=
-source("~/repo/eQTL-2D/replication/results/summary_script.R")
+read_chunk("~/repo/eQTL-2D/docs/replication_analysis/analyse_replication.R")
+
+@
+
+<<qqplots, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, fig.cap="Q-Q plots of interaction p-values from replication dataset", fig.subcap=c("All signals", "Signals excluding 20 most significant replications"), out.width='.4\\linewidth'>>=
+
+qqplot1
+qqplot2
+
+@
+
+<<overlapplots, eval=TRUE, echo=FALSE, warning=FALSE, cache=TRUE, message=FALSE, fig.cap='Overlap in signals between replication datasets', fig.subcap=c('Bonferroni significance', '5\\% FDR significance'), out.width='.4\\linewidth'>>=
+
+# Characterise overlap
+# Significant in only Fehr, Significant in only EGCUT, Significant in both
+
+# Make venn diagram
+
+plot.new() 
+grid.draw(bonf.venn)
+
+plot.new() 
+grid.draw(fdr.venn)
+
+@
+
+<<corpfull, eval=TRUE, echo=FALSE, warning=FALSE, cache=TRUE, fig.cap='Full genetic effects correlate between datasets (8d.f.
+p-values)', fig.subcap=c('All signals', 'Signals that surpass 5\\% FDR in both datasets'), out.width='.4\\linewidth'>>=
+
+plotCor(pfull_all)
+plotCor(pfull_f)
+
+@
+
+<<corpnest, eval=TRUE, echo=FALSE, warning=FALSE, cache=TRUE, fig.cap='Full genetic effects correlate between datasets (8d.f.
+p-values)', fig.subcap=c('All signals', 'Signals that surpass 5\\% FDR in both datasets'), out.width='.4\\linewidth'>>=
+
+plotCor(pnest_all)
+plotCor(pnest_f)
+
+@
+
+<<interactiontable, cache=TRUE, eval=TRUE, echo=FALSE, results='asis', size='tiny'>>=
+
+print(xtable(bonf.tab, caption="The 20 most significant interaction terms from both replication sets.
+* denotes Bonferroni level significance in the named data set."), include.rownames=FALSE, latex.environments=c("center", "footnotesize"), floating.environment="sidewaystable")
 @

### 2013-05-27 analysis/circles.R

@@ -141,8 +141,9 @@ multipleSnps <- function(sig)
 
 # Load data 
 
-load("~/repo/eQTL-2D/analysis/replication_summary.RData")
+load("~/repo/eQTL-2D/analysis/interaction_list_replication_summary.RData")
 sig <- linkColour(sig)
+sig <- subset(sig, filter != 3)
 
 # Circles 
 
@@ -150,8 +151,8 @@ gr <- makeGr()
 index <- table(sig$probename)
 sig_mult <- subset(sig, probename %in% names(index)[index > 2])
 
-sig_mult <- subset(sig, probegene == "MBNL1")
-links <- makeLinks(sig_mult[1:13,], gr)
+# sig_mult <- subset(sig, probegene == "MBNL1")
+# links <- makeLinks(sig_mult[1:13,], gr)
 
 a <- dlply(sig_mult, .(probename), .progress="text", function(x)
 {
@@ -162,10 +163,8 @@ a <- dlply(sig_mult, .(probename), .progress="text", function(x)
 	return(a)
 })
 
-plot(a[[1]])
-
-pdf(file="~/repo/eQTL-2D/analysis/images/circles_replication.pdf", width=25, height=15)
-multiplot(plotlist=a, cols=6)
+pdf(file="~/repo/eQTL-2D/analysis/images/circles_replication2.pdf", width=25, height=20)
+multiplot(plotlist=a, cols=8)
 dev.off()
 
 

### 2013-05-27 analysis/combine_data.R

@@ -1,46 +1,55 @@
 library(noia)
 library(plyr)
+library(biomaRt)
 
 ReadOrig <- function(filename, objname, setname)
 {
 	load(filename)
 	sig <- get(objname)
 
-	sig <- subset(sig, select=c(chr1, chr2, snp1, snp2, pos1, pos2, probename, probegene, probechr, pfull, pnest))
+	sig <- subset(sig, select=c(chr1, chr2, snp1, snp2, pos1, pos2, probename, probegene, probechr, pfull, pnest, filter))
 	sig$set <- setname
 	sig$code <- with(sig, paste(probename, snp1, snp2))
 	return(sig)
 }
 
-
-ReadRep <- function(filename, objname, setname)
+FormatRep <- function(sig)
 {
-	load(filename)
-	sig <- get(objname)
-
 	a <- sum(sig$replication_r^2 > 0.1)
 	# cat(a, "pairs with high LD\n")
 
 	b <- sum(sig$replication_nclass != 9)
 	# cat(b, "pairs with < 9 classes\n")
 
-	sig <- subset(sig, replication_r^2 < 0.1 & replication_nclass == 9, select=c(chr1, chr2, snp1, snp2, pos1, pos2, probename, probegene, probechr, replication_pfull, replication_pnest))
+	vc <- apply(subset(sig, replication_r^2 < 0.1 & replication_nclass == 9, select=c(.a, a., .d, d., aa, ad, da, dd)), 1, list)
+	sig <- subset(sig, replication_r^2 < 0.1 & replication_nclass == 9, select=c(chr1, chr2, snp1, snp2, probename, probegene, probechr, replication_pfull, replication_pnest, filter, gcm, gcs))
 
-	names(sig) <- c("chr1", "chr2", "snp1", "snp2", "pos1", "pos2", "probename", "probegene", "probechr", "pfull", "pnest")
-	sig$set <- setname
+
+	names(sig) <- c("chr1", "chr2", "snp1", "snp2", "probename", "probegene", "probechr", "pfull", "pnest", "filter", "gcm", "gcs")
+	sig$vc <- vc
 	sig$code <- with(sig, paste(probename, snp1, snp2))
+	sig <- sig[order(sig$pnest, decreasing=T), ]
[…]

### 2013-05-27 analysis/functional_analysis.R

@@ -327,11 +327,3 @@ ggplot(prop, aes(y=value, x=index)) +
 	coord_flip() 
 ggsave("~/repo/eQTL-2D/analysis/images/proportion_additive.pdf", width=10, height=10)
 
-
-
-
-
-
-
-
-

### 2013-05-27 replication/results/summary_script2.R

@@ -1,8 +1,3 @@
-library(lattice)
-library(latticeExtra)
-library(gridExtra)
-
-
 summariseRep <- function(filename)
 {
 	load(filename)

### 2013-05-27 analysis/gpmaps.R
Kostya: use analysis/interaction_list_replication_summary.RData

@@ -1,0 +1,46 @@
+library(lattice)
+library(latticeExtra)
+library(gridExtra)
+library(ggplot2)
+library(reshape2)
+library(plyr)
+
+plot3dHairballsRep <- function(sig, pg, cissnp, z=45)
+{
+	# make table of mean phenotypes
+	# AA cis vs 1:nalleles trans
+	# Aa cis vs 1:nalleles trans
+	# aa cis vs 1:nalleles trans
+
+	s <- subset(sig, probegene==pg)
+	pn <- s$probename[1]
+	snps <- unique(c(s$snp1, s$snp2))
+	stopifnot(cissnp %in% snps)
+	tsnps <- snps[! snps %in% cissnp]
+	x1 <- xmat[, match(cissnp, bim$V2)]
+	x <- xmat[, match(tsnps, bim$V2)]
+	y <- resphen[, match(pn, colnames(resphen))]
+
+	temp1 <- subset(s, select=c(snp1, chr1))
+	temp2 <- subset(s, select=c(snp2, chr2))
+	names(temp1) <- names(temp2) <- c("snp", "chr")
+	chrkey <- rbind(temp1, temp2)
+	chrkey <- subset(chrkey, !duplicated(snp))
+
+	l <- list()
+	for(i in 1:ncol(x))
+	{
+		print(table(x1, x[,i]))
+		gp <- tapply(y, list(x1, x[,i]), function(x) mean(x, na.rm=T))
+		gp <- gp - min(gp, na.rm=T)
+		print(gp)
+		title <- paste(s$probegene[1], "chr", subset(chrkey, snp == cissnp)$chr, "x", subset(chrkey, snp == tsnps[i])$chr)
+		l[[i]] <- plot3dGp(gp, title, cissnp, tsnps[i], z)
+	}
+	do.call(grid.arrange, l)
+}
+
+
+load("~/repo/eQTL-2D/analysis/replication2_summary.RData")
+
+

### 2013-05-28 docs/manuscript/nature_letter.tex
writing manuscript

\documentclass{article}

\title{Epistasis identified to be widespread in the control of human gene expression}

\begin{document}

\maketitle

\begin{abstract}
A long standing question in human genetics is the extent to which epistasis, where one mutation's effect on a trait depends on other mutations present in the genome, contributes to complex traits. Though epistasis is known to arise from artificial gene manipulation studies in model organisms, and examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits. Its absence from empirical findings could be due to its unimportance in the genetic control of complex traits, or it may be too technically difficult to detect due to statistical power and computational issues. Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, abundant evidence for epistasis is found. In a modest sample size of 842 individuals with data on 7339 gene expression levels in whole blood, we find that after correcting for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions. We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset. Several genes are controlled by epistatic interactions whereby one cis-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs, for example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by trans-SNPs on 18 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction. This study presents the first strong evidence for the widespread existence of epistasis among natural polymorphisms in humans.
\end{abstract}

\section{Main}

The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits. Typically, SNPs are treated as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying hundreds of causal variants. Yet outside the prism of association studies there is widespread evidence for epistasis, for example at the molecular level in artificial double gene knockouts in experimental organisms and at the evolutionary level in fitness adaptation and speciation. Though its importance is frequently the subject of debate there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits.

There are two major obstacles in detecting epistasis. First, the detection of epistasis is significantly hampered by power issues. For example, there is a greater dependency on linkage disequilibrium (LD) between causal variants and observed SNPs in order to capture non-additive genetic variance, and this problem is exacerbated further with epistasis because there must be high LD at all interacting loci. Relatedly, the dependence on a much larger sample size in order to detect similar effect sizes compared to additive effects, because of the increased number of parameters in statistical tests for interaction terms, and the increased multiple testing correction for testing combinations of SNPs.  Second, one must extend genome-wide association scans into two or more dimensions in order to detect interacting SNPs, resulting in an overwhelming computational burden. We overcome this problem by distributing the search grid across graphics processing unit compute clusters.

Results

- Hundreds of significant pairwise interactions controlling gene expression replicate in two independent studies

- The genotype-phenotype maps for the most significant replications are consistent in all three datasets

- The proportion of the variance that is due to interactions is generally smaller than additive. There is an ascertainment bias in that there are no genetic interactions strong enough to be detected on their own, and large additive effects are required to push them above the threshold

- There are several examples of gene expression being controlled by a cis-SNP which in turn is modulated by multiple trans-SNPs. The effects are consistent

\section{Methods summary}

\end{document}

### 2013-05-28 docs/manuscript/nature_letter.tex
writing continues

@@
 
-\title{Epistasis identified to be widespread in the control of human gene expression}
+\title{Epistasis is widespread in the control of human gene expression}
 
@@
 Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, abundant evidence for epistasis is found.
-In a modest sample size of 842 individuals with data on 7339 gene expression levels in whole blood, we find that after correcting for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
+In a modest sample size of 842 individuals with data on 7339 gene expression levels in whole blood, we found that after correcting for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
 We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset.
-Several genes are controlled by epistatic interactions whereby one cis-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs, for example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by trans-SNPs on 18 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction.
+Several genes are controlled by multi-locus epistatic interactions whereby one cis-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs, for example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by trans-SNPs on 18 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction.
 This study presents the first strong evidence for the widespread existence of epistasis among natural polymorphisms in humans.
@@
 Typically, SNPs are treated as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying hundreds of causal variants.
-Yet outside the prism of association studies there is widespread evidence for epistasis, for example at the molecular level in artificial double gene knockouts in experimental organisms and at the evolutionary level in fitness adaptation and speciation.
-Though its importance is frequently the subject of debate there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits.
+Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale in artificial double gene knockouts in experimental organisms but also at the evolutionary scale in fitness adaptation and speciation.
+Though its importance is frequently the subject of debate so far there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits.
 
-There are two major obstacles in detecting epistasis.
-First, the detection of epistasis is significantly hampered by power issues.
-For example, there is a greater dependency on linkage disequilibrium (LD) between causal variants and observed SNPs in order to capture non-additive genetic variance, and this problem is exacerbated further with epistasis because there must be high LD at all interacting loci.
-Relatedly, the dependence on a much larger sample size in order to detect similar effect sizes compared to additive effects, because of the increased number of parameters in statistical tests for interaction terms, and the increased multiple testing correction for testing combinations of SNPs.
-Second, one must extend genome-wide association scans into two or more dimensions in order to detect interacting SNPs, resulting in an overwhelming computational burden.
-We overcome this problem by distributing the search grid across graphics processing unit compute clusters.
+The detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium between causal SNPs and observed SNPs, increased model complexity in fitting interaction terms, and more extreme significance thresholds to account for increased multiple testing.
+When genetic effect sizes are small, as is expected in most complex traits of interest, the power to detect epistasis diminishes rapidly.
+To overcome this problem we searched for epistasis controlling gene expression levels.
+Though they number in the thousands, these traits are typically heritable but much less polygenic, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis if it should exist.
+
+We searched for pairwise epistasis exhaustively in the BSGS dataset which comprises 842 individuals who are genotyped at 528,000 SNPs and who have gene expression levels measured in whole blood samples for 7339 genes.
+Recent hardware and software advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
+We used permutation analysis to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ for $p$-values at the 5\% family-wise error rate (FWER).
 

### 2013-05-29 docs/manuscript/nature_letter.tex
writing continues

@@
 
-\title{Epistasis is widespread in the control of human gene expression}
+\title{Evidence for widespread epistasis in the control of human gene expression}
 
@@
 Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, abundant evidence for epistasis is found.
-In a modest sample size of 842 individuals with data on 7339 gene expression levels in whole blood, we found that after correcting for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
+In a cohort of 842 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
 We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset.
 Several genes are controlled by multi-locus epistatic interactions whereby one cis-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs, for example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by trans-SNPs on 18 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction.
-This study presents the first strong evidence for the widespread existence of epistasis among natural polymorphisms in humans.
+This study presents the first strong evidence for the widespread existence of epistasis emerging from natural genetic variation in humans.
 \end{abstract}
@@
 
-The detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium between causal SNPs and observed SNPs, increased model complexity in fitting interaction terms, and more extreme significance thresholds to account for increased multiple testing.
+The detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs, increased model complexity in fitting interaction terms, and more extreme significance thresholds to account for increased multiple testing.
 When genetic effect sizes are small, as is expected in most complex traits of interest, the power to detect epistasis diminishes rapidly.
-To overcome this problem we searched for epistasis controlling gene expression levels.
+There are two ways to overcome this problem, either use extremely large sample sizes, or use traits that are likely to have large effect sizes.
+Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we opted for the latter approach and searched for epistasis controlling gene expression levels.
 Though they number in the thousands, these traits are typically heritable but much less polygenic, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis if it should exist.
 
-We searched for pairwise epistasis exhaustively in the BSGS dataset which comprises 842 individuals who are genotyped at 528,000 SNPs and who have gene expression levels measured in whole blood samples for 7339 genes.
+We searched for pairwise epistasis exhaustively in the BSGS dataset which comprises 842 individuals who are genotyped at 528,509 SNPs and who have gene expression levels measured in whole blood samples for 7339 genes.
 Recent hardware and software advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
-We used permutation analysis to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ for $p$-values at the 5\% family-wise error rate (FWER).
+We used permutation analysis to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
+SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
+Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test.
+Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
+All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least 5 data points in all nine genotype classes.
+If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained.
+Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair, resulting in 549 significant interactions after Bonferroni correction for multiple testing.
+
+The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for whole blood, Fehrmann ($n=1240$) and EGCUT ($n=891$).
+There was substantial evidence for the same effects being present in all three datasets.
+Of the 480 original pairs that passed filtering in Fehrmann and EGCUT, 39 were significant at the Bonferroni level in at least one, and of these 18 were significant in both.
+The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}).
+
+We also attempted replication in a third dataset, CHDWB, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power there was no evidence for replication.
+
+\section{Notes}
 

### 2013-05-29 analysis/functional_analysis.R

@@ -310,9 +310,10 @@ plot3dProbe(sig, "CTSC", xmat, resphen, z=135)
 
 # Distribution of additive vs non-additive variance
 
-sig$varA <- sig$.a + sig$a.
-sig$varD <- sig$.d + sig$d.
-sig$varI <- sig$aa + sig$ad + sig$da + sig$dd
+a <- as.data.frame(do.call(rbind, sig$vc))
+sig$varA <- a$.a + a$a.
+sig$varD <- a$.d + a$d.
+sig$varI <- a$aa + a$ad + a$da + a$dd
 prop <- subset(sig, select=c(varA, varI, varD))
 prop <- prop[order(prop$varA+prop$varI+prop$varD, decreasing=T),]
 prop$index <- 1:nrow(prop)
@@ -321,9 +322,34 @@ prop <- prop[nrow(prop):1, ]
 prop$variable <- factor(prop$variable, levels=c("varI", "varD", "varA"))
 levels(prop$variable) <- c("Interaction", "Dominance", "Additive")
 ggplot(prop, aes(y=value, x=index)) + 
-	geom_bar(stat="identity", width=1, size=0.0, colour="black", aes(fill=variable)) +
-	scale_fill_brewer("Proportion of phenotypic variance") +
-	ylab("Phenotypic variance") +
+	geom_bar(stat="identity", width=1, size=0.0, aes(fill=variable)) +
+	scale_fill_brewer("Variance component") +
+	ylab("Phenotypic variance") + xlab("") +
 	coord_flip() 
 ggsave("~/repo/eQTL-2D/analysis/images/proportion_additive.pdf", width=10, height=10)
+
+
+a <- as.data.frame(do.call(rbind, sig$vc))
+sig$varA <- a$.a + a$a.
+sig$varD <- a$.d + a$d.
+sig$varI <- a$aa + a$ad + a$da + a$dd
+sig$varG <- with(sig, varA + varD + varI)
+sig$varA <- sig$varA / sig$varG
+sig$varD <- sig$varD / sig$varG
+sig$varI <- sig$varI / sig$varG
+prop <- subset(sig, select=c(varA, varI, varD))
+prop <- prop[order(prop$varA, decreasing=T),]
+prop$index <- 1:nrow(prop)
+prop <- melt(prop, id=c("index"))
+prop <- prop[nrow(prop):1, ]
+prop$variable <- factor(prop$variable, levels=c("varI", "varD", "varA"))
[…]

### 2013-05-29 docs/manuscript/nature_letter.tex
progress continues on writing manuscript
need to add kostya's figure for functional annotation to supplementary

@@
 \documentclass{article}
+\usepackage[superscript,biblabel]{cite}
+\usepackage{graphicx}
 
-\title{Evidence for widespread epistasis in the control of human gene expression}
+\graphicspath{{../../analysis/images/}}
+\bibliographystyle{plain}
+
+\title{Epistasis is widespread in the genetic control of transcription in humans}
 
@@
 \begin{abstract}
-A long standing question in human genetics is the extent to which epistasis, where one mutation's effect on a trait depends on other mutations present in the genome, contributes to complex traits.
-Though epistasis is known to arise from artificial gene manipulation studies in model organisms, and examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits.
-Its absence from empirical findings could be due to its unimportance in the genetic control of complex traits, or it may be too technically difficult to detect due to statistical power and computational issues.
-Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, abundant evidence for epistasis is found.
+A long standing question in human genetics is the extent to which epistasis, the phenomenon whereby one mutation's effect on a trait depends on other mutations present in the genome, contributes to complex traits.
+Though epistasis is known to arise from artificial gene manipulation studies in model organisms, and occasional examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits.
+Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits, but we hypothesised that it has previously been too technically difficult to detect due to statistical power and computational issues.
+Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, evidence for abundant epistasis is found.
 In a cohort of 842 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
 We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset.
-Several genes are controlled by multi-locus epistatic interactions whereby one cis-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs, for example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by trans-SNPs on 18 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction.
+Several genes are controlled by multi-locus epistatic interactions whereby one cis-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
+For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by trans-SNPs on 18 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction.
 This study presents the first strong evidence for the widespread existence of epistasis emerging from natural genetic variation in humans.
@@
 
+\subsection{Introduction}
 The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits.
@@
 
-The detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs, increased model complexity in fitting interaction terms, and more extreme significance thresholds to account for increased multiple testing.
+The detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs, increased model complexity in fitting interaction terms, and more extreme significance thresholds to account for increased multiple testing \cite{Hemani2013}.
 When genetic effect sizes are small, as is expected in most complex traits of interest, the power to detect epistasis diminishes rapidly.
@@
 Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we opted for the latter approach and searched for epistasis controlling gene expression levels.
-Though they number in the thousands, these traits are typically heritable but much less polygenic, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis if it should exist.
+Though they number in the thousands, these traits are typically heritable but much less polygenic, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis should it exist.
 
+\subsection{Results}
+
+\subsubsection{Initial search}
 We searched for pairwise epistasis exhaustively in the BSGS dataset which comprises 842 individuals who are genotyped at 528,509 SNPs and who have gene expression levels measured in whole blood samples for 7339 genes.
 Recent hardware and software advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
-We used permutation analysis to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
+We used permutation analysis to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (Methods).
 SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
@@
 If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained.
-Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair, resulting in 549 significant interactions after Bonferroni correction for multiple testing.
+Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 549 significant interactions after Bonferroni correction for multiple testing.
 
-The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for whole blood, Fehrmann ($n=1240$) and EGCUT ($n=891$).
-There was substantial evidence for the same effects being present in all three datasets.
-Of the 480 original pairs that passed filtering in Fehrmann and EGCUT, 39 were significant at the Bonferroni level in at least one, and of these 18 were significant in both.
-The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}).
+\subsubsection{Replication}
+It is intrinsically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than for independent additive effects.
+The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, Fehrmann ($n=1240$) and EGCUT ($n=891$).
+There was substantial evidence for overlap of interaction effects in all three datasets.
+Of the 480 original pairs that passed filtering in Fehrmann and EGCUT, 421 of the SNP pairs showed replication for the interaction test at the 5\% false discovery rate (FDR) level in at least one dataset, and 148 in both datasets (Supplementary Figure \ref{fig:qqplotfdr}).
+The congruence of the epistatic networks in both datasets is shown in Figure \ref{fig:fireworks}.
+We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and of these 18 were significant in both.
+The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}), lending strong evidence for the veracity of these signals.
+We also attempted replication in the CHDWB dataset, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power there was no evidence for replication.
 
-We also attempted replication in a third dataset, CHDWB, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power there was no evidence for replication.
+\subsubsection{Patterns of epistasis}
+
+\subsubsection{Characterisation of genetic variation}
+
+\subsubsection{Functional annotation}
 
@@
 
+- Functional annotation shows that proteins are not being changed at the codon level
+- There is strong enrichment for enhancers
+- There is tissue specific transcription regulation for cis SNPs - these are regulatory
+- Trans SNPs are not found in tissue specific transcription regions.
+- General enrichment for regulatory motif modifications
+- Names of specific factors that are enriched.
+
+- No explicit mechanism arises from looking at GP maps
+
+- Chromosome interactions?
+
+\clearpage
 \section{Methods summary}
 
+\clearpage
+\section{Figure legends}
+
+\begin{figure}
+	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A shows the overall network of genetic interactions (edges) for each probe (red nodes) and their controlling SNPs (black nodes) from the discovery set.
+Panel B shows those interactions that were present at the 5\% FDR level in at least one replication dataset, and panel C shows which interactions were significant in both replication datasets.}
+	\label{fig:fireworks}
+\end{figure}
+
+\begin{figure}
+	\caption{\textbf{Replication of genotype-phenotype maps in 2 cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets.
+Each tile represents the expression level for that SNP-SNP genotype class.
+Phenotypes for each interaction within each cohort are scaled to be between 0 and 1.}
+	\label{fig:gpmaps}
+\end{figure}
+
+\begin{figure}
+	\caption{\textbf{Comparison of genetic variances} Panel A shows the proportion of phenotypic variation that is captured by additive, dominant and interaction components for all 529 significant interactions, ordered by total phenotypic variance explained by the SNP pair.
+Panel B shows the proportion of the genetic variance that is attributable to the three variance components, ordered by the proportion that is additive.}
+	\label{fig:variancecomponents}
+\end{figure}
+
+\clearpage
+\section*{}
+\setcounter{figure}{0}
+
+\begin{figure}
+\centering
+\includegraphics[width=5in]{hairballs_all_reps.pdf}
+\caption{}
+\end{figure}
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=5in]{gpbonfrep.pdf}
+\caption{}
+\end{figure}
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=5in]{variance_components.pdf}
+\caption{}
+\end{figure}
+\clearpage
+
+\section{Full Methods}
+
+\subsection{Discovery data}
+
+The Brisbane Systems Genetics Study (BSGS) comprises 862 Individuals of European descent from 274 independent families \cite{pmid22563384}.
+DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland.
+Full details of genotyping procedures are given in Medland et al.
+\cite{Medland2009} Standard QC filters were applied and the remaining 528,509 SNPs were carried forward further analysis.
+
+Gene expression profiles were generated from whole blood collected with PAXgene TM tubes (QIAGEN, Valencia, CA) using Illumina HT12-v4.0 bead arrays.
+The Illumina HT-12 v4.0 chip contains 47,323 probes, although some probes are not assigned to RefSeq genes.
+We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF > 0.05 within 1000 genomes data; did not map to a listed ref-seq gene; were not significantly expressed (based on a detection $p$-value < 0.05) in 100\% of samples.
+After this stringent QC ~7000 [need to get exact number] probes remained for 2d-eqtl mapping.
+
+\subsection{normalisation}
+
+Gene expression profiles were normalised and batch effects removed with the following procedures; Raw background expression were removed for each sample.
+Expression levels were then adjusted using Quantile and log2 transformation to standardise distribution between samples.
+To avoid inflation of the test statistic due to polygenic effects the following linear model was used to correct for batch and polygenic effects;
+
+\begin{equation}
+y = \mu + c + p + s + a + g + e
+\label{eq:lm}
+\end{equation}
+where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively.
+$g$ is a random additive polygenic effect with a variance covariance matrix
+
+\begin{equation}
+G_{ijk} = \left \{ 
+\begin{array}{ll}
+\sigma _a ^2 + \sigma _e ^2&        j = k \\ 
+2\phi _{ijk} \sigma _a ^2& 			j \neq k \\
+\end{array} \right.
+\end{equation}
+
+The parameters $\sigma _a ^a$ and $\sigma _e ^2$ are variance components for additive background genetic and environmental effects respectively.
+Here, we are using family based pedigree information rather than SNP based IDB to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$.
+The residual, $e$, from equation \ref{eq:lm} are assumed to follow a multivariate normal distribution with a mean of zero.
+These residuals are corrected for batch, sex, generation and polygenic effects.
+Residuals were standardised to Z-scores using a inverse-normal transformation and used as the adjusted phenotype for the pairwise epistasis scan.
+
+\subsection{exhaustive 2d-eQTL analysis}
+We used epi-GPU software to perform an exhaustive scan for pairwise interactions, such that each SNP is tested against all other SNPs for statistical association with the expression values for each of the ~7000 probes.
+For each SNP-pair there are 9 possible genotype classes.
+We treat each genotype class as a fixed effect and fit an 8 d.f.
+$F$-test to test the following hypotheses:
+
+\begin{equation}
+H _0 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 = 0; 
+\end{equation}
+
+\begin{equation}
+H _1 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 > 0; 
+\label{eq:8df}
+\end{equation}
+
+where $\mu$ is the mean expression level and $x _{ij}$ is the pairwise genotype class mean for genotype $i$ at SNP 1 and genotype $j$ at locus B.
+This type of test does not parameterize for specific types of epistasis, rather it tests for the joint genetic effects at two loci.
+This has been demonstrated to be statistically more efficient when searching for a wide range of epistatic patterns (add refs), although will include any marginal effects of SNPs.
+\\[0.2cm]
+
+The complete exhaustive scan for 7,000 probes comprises of $\sim$9.8$e^{14}$ $F$-tests.
+Due to the very large number of tests [plus nature of epistasis ....] we used a highly conservative series of filtering steps to identify study-wide significant epistasis associations to follow forward for replication analysis.
+Initially a significance threshold of $-log10$ 15.5.
+This threshold is the study-wide significance threshold determined by the number of independent tests for a 2D scan for a single probe / the number of probes analysed.
+This is likely to be conservative as this threshold assumes independence between probes.
+Following filtering on this threshold only SNP pairs with all 9 genotype classes and minimum genotype class size > 5 individuals were kept.
+We then calculated the linkage disequilibrium between SNPs in a pair and removed any pairs with $r^2$ > 0.1.
+We also removed any pairs containing SNPs for which single marker additive or dominance effects had been identified  in previous work (Powell et al.
+in Press).
+Sentinel  SNP pairs we retained from epistasis eQTL `peaks' comprising of multiple sets of pair-wise SNPs.
+At this stage $\sim$13,000 SNP pairs remained and were tested for the contribution of marginal and interaction terms by testing the 8d.f.
+\ref{eq:8df} verses a 4 d.f.
+model parameterizing for interaction terms only:
+
+\begin{equation}
+H _0: \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \bar x _i - \bar x _j + \mu) ^2 = 0;
+\end{equation}
+
+\begin{equation}
+H _1: \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \bar x _i - \bar x _j + \mu) ^2 > 0;
+\label{eq:4df}
+\end{equation}
+
+where $\bar x _i (\bar x _j)$ is the marginal class mean for genotype $i (j)$ at SNP A (B). Significance of interaction terms was determined using a bonferroni threshold of $0.05 / 13000 = 3.8e^{-6}$
+
+We have attempted replication of the 2D scan significant results using three independent cohorts; Fehrmann, CHDWB and Rotterdam. Details of cohorts are as follows. \\[0.1cm]
+
+Fehrmann: $n$=1,469 \\[0.1cm]
+The Fehrmann dataset consists of whole peripheral blood samples of 1,469 unrelated individuals from the United Kingdom and the Netherlands.
+Some of these individuals are patients, while others are healthy controls.
+Individuals were genotyped using Illumina HumanHap300, HumanHap370 or the 610 Quad platform.
+RNA levels were quantified using both the Illumina H8v2 platform (N = 229) and the HT12v3 platform (N = 1,240).
+\\[0.1cm]
+
+CDHWB: $n$=189 \\ [0.1cm]
+The Center for Health Discovery and Well Being (CDHWB) Study is a population based cohort consisting of 189 individuals collected in Atlanta USA.
+Gene expression profiles were generated with Illumina HT-12 V3.0 arrays from whole blood collected from Tempus tubes that preserve RNA.
+Whole genome genotypes were measured using Illumina OmniQuad arrays.
+\\[0.1cm]
+
+Rotterdam: $n$=762 \\[0.1cm]
+The Rotterdam Study is a large population based cohort collected to investigate the prevalence, incidence, and risk factors of chronic diseases among  Caucasians aged 45 years and over.
+Participants were genotyped using the Illumina 610K quad arrays.
+Whole blood of 768 samples was collected (PAXgene)and RNA hybridized to the Illumina Human HT-12v4 chips.
+
+\clearpage
+\section{References}
+\bibliography{refs}
+
+\clearpage
+\section{Supplementary Figures}
+\makeatletter 
+\renewcommand{\thefigure}{S\@arabic\c@figure}
+
+\begin{figure}
+\caption{\textbf{Q-Q plots of interaction $p$-values in two independent datasets}}
+\label{fig:qqplotbonf}
+\end{figure}
+
+\begin{figure}
+\caption{\textbf{Q-Q plots of replicated interaction $p$-values excluding the 20 most extreme $p$-values}}
+\label{fig:qqplotfdr}
+\end{figure}
+
 \end{document}

### 2013-05-29 analysis/circles.R

@@ -167,23 +167,51 @@ pdf(file="~/repo/eQTL-2D/analysis/images/circles_replication2.pdf", width=25, he
 multiplot(plotlist=a, cols=8)
 dev.off()
 
+key <- subset(sig, !duplicated(probename), select=c(probename, probegene))
+nom <- subset(key, duplicated(probegene))$probename
 
+sig2 <- subset(sig, ! probename %in% nom)
 
 # More circles
-a <- subset(sig, select=c(snp1, snp2))
+a <- subset(sig2, select=c(snp1, snp2))
 names(a) <- c("from", "to")
 a$thickness <- 1
-a$col <- as.character(sig$rep)
+a$col <- as.character(sig2$rep)
 a$col[a$col == 0] <- "white"
 a$col[a$col == 1] <- "black"
 a$col[a$col == 2] <- "black"
+a$col2 <- a$col
+a$col2[sig2$rep == 1] <- "white"
+a$col3 <- as.character(sig2$rep)
+a$col3[a$col3 == 0] <- ""
+a$col3[a$col3 == 1] <- "black"
+a$col3[a$col3 == 2] <- "red"
+a$thickness2 <- a$thickness
+a$thickness2[a$col3 == "red"] <- 1.5
+a$cistrans <- "blue"
+a$cistrans[sig2$c]
+
 
 pdf(file="~/repo/eQTL-2D/analysis/images/hairballs_all.pdf")
-qgraph(a, esize=1, edge.color=a$thickness, label.cex=0, vsize=0.2, arrows=FALSE)
+qgraph(a, curve=0.1, borders=FALSE, esize=1, color="grey", edge.color=a$thickness, labels=FALSE, vsize=0.2, arrows=FALSE)
 dev.off()
 pdf(file="~/repo/eQTL-2D/analysis/images/hairballs_rep.pdf")
-qgraph(a, esize=1, color=a$col, edge.color=a$col, label.cex=0, vsize=0.2, arrows=FALSE)
+qgraph(a, curve=0.1, borders=FALSE, esize=1, color="grey", edge.color=a$col, labels=FALSE, vsize=0.2, arrows=FALSE)
 dev.off()
+pdf(file="~/repo/eQTL-2D/analysis/images/hairballs_rep2.pdf")
+qgraph(a, curve=0.1, borders=FALSE, esize=1, color="grey", edge.color=a$col2, labels=FALSE, vsize=0.2, arrows=FALSE)
+dev.off()
[…]

### 2013-05-29 analysis/functional_analysis.R

@@ -5,6 +5,45 @@ library(ggplot2)
 library(reshape2)
 library(plyr)
 
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL)
+{
+	require(grid)
+
+	# Make a list from the ... arguments and plotlist
+	plots <- c(list(...), plotlist)
+
+	numPlots = length(plots)
+
+	# If layout is NULL, then use 'cols' to determine layout
+	if (is.null(layout)) 
+	{
+		# Make the panel
+		# ncol: Number of columns of plots
+		# nrow: Number of rows needed, calculated from # of cols
+		layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+		ncol = cols, nrow = ceiling(numPlots/cols))
+	}
+
+	if (numPlots==1) 
+	{
+		print(plots[[1]])
+	} else {
+		# Set up the page
+		grid.newpage()
+		pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+
+		# Make each plot, in the correct location
+		for (i in 1:numPlots)
+		{
+			# Get the i,j matrix positions of the regions that contain this subplot
+			matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+
+			print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+			layout.pos.col = matchidx$col))
+		}
+	}
+}
+
 
 plot3d <- function(res, geno, phen, z=-45)
 {
@@ -321,11 +360,11 @@ prop <- melt(prop, id=c("index"))
 prop <- prop[nrow(prop):1, ]
 prop$variable <- factor(prop$variable, levels=c("varI", "varD", "varA"))
 levels(prop$variable) <- c("Interaction", "Dominance", "Additive")
-ggplot(prop, aes(y=value, x=index)) + 
+p1 <- ggplot(prop, aes(y=value, x=index)) + 
 	geom_bar(stat="identity", width=1, size=0.0, aes(fill=variable)) +
 	scale_fill_brewer("Variance component") +
 	ylab("Phenotypic variance") + xlab("") +
-	coord_flip() 
+	coord_flip() + theme(legend.position = "none")
 ggsave("~/repo/eQTL-2D/analysis/images/proportion_additive.pdf", width=10, height=10)
 
 
@@ -344,12 +383,15 @@ prop <- melt(prop, id=c("index"))
 prop <- prop[nrow(prop):1, ]
[…]

### 2013-05-29 docs/manuscript/nature_letter.tex
modified variance_components image to have no strange lines

@@
 \usepackage{graphicx}
+\usepackage{float}
 
@@
 
-\begin{figure}
+\begin{figure}[H]
 	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A shows the overall network of genetic interactions (edges) for each probe (red nodes) and their controlling SNPs (black nodes) from the discovery set.
@@
 
-\begin{figure}
+\begin{figure}[H]
 	\caption{\textbf{Replication of genotype-phenotype maps in 2 cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets.
@@
 
-\begin{figure}
+\begin{figure}[H]
 	\caption{\textbf{Comparison of genetic variances} Panel A shows the proportion of phenotypic variation that is captured by additive, dominant and interaction components for all 529 significant interactions, ordered by total phenotypic variance explained by the SNP pair.
@@
 \clearpage
-\section*{}
+\section{Figures}
 \setcounter{figure}{0}
 
-\begin{figure}
+\begin{figure}[H]
 \centering
@@
 \section{Supplementary Figures}
+\setcounter{figure}{0}
 \makeatletter 
 \renewcommand{\thefigure}{S\@arabic\c@figure} 
+\makeatletter
 
-\begin{figure}
-\caption{\textbf{Q-Q plots of interaction $p$-values in two independent datasets}}
+\begin{figure}[H]
+\includegraphics[width=5in]{qqbonf}
+\caption{\textbf{Q-Q plots of interaction $p$-values in two independent datasets} The interaction $p$-values for the 480 SNP pairs that passed filtering in the two replication datasets.
+Green points represent $p$-values that surpass Bonferroni correction}
 \label{fig:qqplotbonf}
@@
 \begin{figure}
-\caption{\textbf{Q-Q plots of replicated interaction $p$-values excluding the 20 most extreme $p$-values}}
+\includegraphics[width=5in]{qqfdr}
+\caption{\textbf{Q-Q plots of replicated interaction $p$-values excluding the 20 most extreme $p$-values} The left panel (Null) shows the replication $p$-values from 529 randomly drawn SNP pairs in the Fehrmann (top row) and EGCUT (bottom row) datasets.
+The right panel (Observed) shows the interaction $p$-values from the 460 least significant pairs that pass filtering in the two replication datasets (\emph{i.e.} excluding the 20 most extreme $p$-values for clarity).
+Dark blue points represent $p$-values that surpass the 5\% FDR level.}
 \label{fig:qqplotfdr}
@@
 
+\begin{figure}
+\caption{\textbf{}}
+\end{figure}
+
 \end{document}

### 2013-05-29 analysis/functional_analysis.R

@@ -138,7 +138,7 @@ plot3dHairballs <- function(sig, pg, cissnp, resphen, xmat, bim, z=45)
 #=================================================================================================#
 
 
-load("~/repo/eQTL-2D/analysis/replication_summary.RData")
+load("~/repo/eQTL-2D/analysis/interaction_list_replication_summary.RData")
 load("~/repo/eQTL-2D/data/residuals_all.RData")
 load("~/repo/eQTL-2D/data/clean_geno_final.RData")
 
@@ -361,10 +361,12 @@ prop <- prop[nrow(prop):1, ]
 prop$variable <- factor(prop$variable, levels=c("varI", "varD", "varA"))
 levels(prop$variable) <- c("Interaction", "Dominance", "Additive")
 p1 <- ggplot(prop, aes(y=value, x=index)) + 
-	geom_bar(stat="identity", width=1, size=0.0, aes(fill=variable)) +
+	geom_bar(stat="identity", aes(fill=variable, colour=variable), position=position_stack(width=0)) +
 	scale_fill_brewer("Variance component") +
+	scale_colour_brewer("Variance component") +
 	ylab("Phenotypic variance") + xlab("") +
-	coord_flip() + theme(legend.position = "none")
+	coord_flip() + theme(legend.position = "none") +
+	theme(axis.text.y=element_text(size=0), axis.ticks.y=element_line(size=0))
 ggsave("~/repo/eQTL-2D/analysis/images/proportion_additive.pdf", width=10, height=10)
 
 
@@ -384,14 +386,16 @@ prop <- prop[nrow(prop):1, ]
 prop$variable <- factor(prop$variable, levels=c("varI", "varD", "varA"))
 levels(prop$variable) <- c("Interaction", "Dominance", "Additive")
 p2 <- ggplot(prop, aes(y=value, x=index)) + 
-	geom_bar(stat="identity", width=1, size=0.0, aes(fill=variable)) +
+	geom_bar(stat="identity", aes(fill=variable, colour=variable), position=position_stack(width=0)) +
 	scale_fill_brewer("Variance component") +
+	scale_colour_brewer("Variance component") +
 	ylab("Proportion of genetic variance") + xlab("") +
-	coord_flip() + theme(legend.justification=c(1,0), legend.position=c(1,0))
+	coord_flip() + 
+	theme(legend.justification=c(1,0), legend.position=c(1,0)) + 
[…]

### 2013-05-29 docs/manuscript/nature_letter.tex
writing continues. around 400 words left...

@@
 \subsubsection{Replication}
-It is intrinsically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than for independent additive effects.
+Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than it is for independent additive effects.
 The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, Fehrmann ($n=1240$) and EGCUT ($n=891$).
@@
 The congruence of the epistatic networks in both datasets is shown in Figure \ref{fig:fireworks}.
-We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and of these 18 were significant in both.
-The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}), lending strong evidence for the veracity of these signals.
+We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and 18 were significant in both.
+The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}).
 We also attempted replication in the CHDWB dataset, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power there was no evidence for replication.
 
-\subsubsection{Patterns of epistasis}
-
-\subsubsection{Characterisation of genetic variation}
-
 \subsubsection{Functional annotation}
-
-\section{Notes}
-
-Results
-
-- Hundreds of significant pairwise interactions controlling gene expression replicate in two independent studies
-
-- The genotype-phenotype maps for the most significant replications are consistent in all three datasets
-
-- The proportion of the variance that is due to interactions is generally smaller than additive.
-There is an ascertainment bias in that there are no genetic interactions strong enough to be detected on their own, and large additive effects are required to push them above the threshold
-
-- There are several examples of gene expression being controlled by a cis-SNP which in turn is modulated by multiple trans-SNPs. The effects are consistent
-
 - Functional annotation shows that proteins are not being changed at the codon level
@@
 - Names of specific factors that are enriched.
+- Chromosome interactions?
 
+\subsubsection{Patterns of epistasis}
+- There are several examples of gene expression being controlled by a cis-SNP which in turn is modulated by multiple trans-SNPs. The effects are consistent 
 - No explicit mechanism arises from looking at GP maps
+- Describe MBNL1
 
-- Chromosome interactions?
+\subsubsection{Characterisation of genetic variation}
+- The proportion of the variance that is due to interactions is generally smaller than additive.
+There is an ascertainment bias in that there are no genetic interactions strong enough to be detected on their own, and large additive effects are required to push them above the threshold
 
@@
 \begin{figure}[H]
-	\caption{\textbf{Replication of genotype-phenotype maps in 2 cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets.
-Each tile represents the expression level for that SNP-SNP genotype class.
-Phenotypes for each interaction within each cohort are scaled to be between 0 and 1.}
+	\caption{\textbf{Replication of genotype-phenotype maps in two cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
+Each tile represents the expression level for that two-locus genotype class.
+Phenotypes for each interaction within each cohort are scaled to be between 0 and 1.
+Phenotypes are for gene transcript levels, where the gene's name is labelled in the panel columns.
+Panel rows are labelled for the dataset.
+There is a general trend of the GP maps replicating across all three datasets.}
 	\label{fig:gpmaps}
@@
 
-The Brisbane Systems Genetics Study (BSGS) comprises 862 Individuals of European descent from 274 independent families \cite{pmid22563384}.
+The Brisbane Systems Genetics Study (BSGS) comprises 862 individuals of European descent from 274 independent families \cite{pmid22563384}.
 DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland.
@@
 
-\subsection{normalisation}
+\subsection{Normalisation}
 

### 2013-05-29 analysis/functional_analysis.R

@@ -127,7 +127,7 @@ plot3dHairballs <- function(sig, pg, cissnp, resphen, xmat, bim, z=45)
 		gp <- tapply(y, list(x1, x[,i]), function(x) mean(x, na.rm=T))
 		gp <- gp - min(gp, na.rm=T)
 		print(gp)
-		title <- paste(s$probegene[1], "chr", subset(chrkey, snp == cissnp)$chr, "x", subset(chrkey, snp == tsnps[i])$chr)
+		title <- paste("Chromosome", subset(chrkey, snp == cissnp)$chr, "x", subset(chrkey, snp == tsnps[i])$chr)
 		l[[i]] <- plot3dGp(gp, title, cissnp, tsnps[i], z)
 	}
 	do.call(grid.arrange, l)

### 2013-05-29 analysis/gpmaps.R

@@ -137,7 +137,7 @@ rotateGp <- function(mat)
 
 plotTileGp <- function(l)
 {
-	p <- ggplot(l, aes(x=snp2, y=snp1, fill=y)) + geom_tile() + facet_grid(cohort ~ code2) + theme(strip.text = element_text(size = 6))
+	p <- ggplot(l, aes(x=snp2, y=snp1, fill=y)) + geom_tile() + facet_grid(cohort ~ code3) + theme(strip.text = element_text(size = 6))
 	return(p)
 }
 
@@ -165,23 +165,55 @@ l <- GpMod(l, 6, "BSGS", t)
 l <- GpMod(l, 7, "BSGS", rotateGp, rotateGp)
 l <- GpMod(l, 9, "BSGS", rotateGp, rotateGp)
 
+index <- match(mbnl1$code, sig$code)
+mbnl1$code4 <- paste(sig$chr1[index], "x", sig$chr2[index])
+mbnl1 <- subset(mbnl1, !is.na(y))
+mbnl1$code3 <- mbnl1$code
 
 pdf(file="~/repo/eQTL-2D/analysis/images/gpBonfRep.pdf", width=20, height=4)
 plotTileGp(l)
 dev.off()
 
+temp <- subset(sig, probegene == "MBNL1")
+temp <- subset(temp, !duplicated(paste(chr1, chr2)) & !is.na(pnest_egcut) & !is.na(pnest_fehr))
+mbnl1 <- datHeatmapGp(temp)
+mbnl1 <- GpMod(mbnl1, 1, "BSGS", flipGpCol)
+mbnl1 <- GpMod(mbnl1, 2, "BSGS", rotateGp, rotateGp)
+mbnl1 <- GpMod(mbnl1, 3, "BSGS", flipGpCol)
+mbnl1 <- GpMod(mbnl1, 4, "BSGS", rotateGp, rotateGp)
+mbnl1 <- GpMod(mbnl1, 5, "BSGS", flipGpCol)
+mbnl1 <- GpMod(mbnl1, 6, "BSGS", flipGpRow)
+mbnl1 <- GpMod(mbnl1, 6, "Fehrmann", flipGpCol)
+mbnl1 <- GpMod(mbnl1, 7, "BSGS", rotateGp, rotateGp)
+mbnl1 <- GpMod(mbnl1, 8, "BSGS", rotateGp, rotateGp)
+mbnl1 <- GpMod(mbnl1, 9, "BSGS", flipGpCol)
[…]

### 2013-05-29 docs/manuscript/nature_letter.tex
statistical full methods mostly complete. Still need functional annotation methods

@@
 Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, evidence for abundant epistasis is found.
-In a cohort of 842 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
+In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
 We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset.
@@
 \subsubsection{Initial search}
-We searched for pairwise epistasis exhaustively in the BSGS dataset which comprises 842 individuals who are genotyped at 528,509 SNPs and who have gene expression levels measured in whole blood samples for 7339 genes.
+We searched for pairwise epistasis exhaustively in the BSGS dataset which comprises 846 individuals who are genotyped at 528,509 SNPs and who have gene expression levels measured in whole blood samples for 7339 genes.
 Recent hardware and software advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
@@
 
-The Brisbane Systems Genetics Study (BSGS) comprises 862 individuals of European descent from 274 independent families \cite{pmid22563384}.
+The Brisbane Systems Genetics Study (BSGS) comprises 846 individuals of European descent from 274 independent families \cite{pmid22563384}.
 DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland.
 Full details of genotyping procedures are given in Medland et al.
-\cite{Medland2009} Standard QC filters were applied and the remaining 528,509 SNPs were carried forward further analysis.
+\cite{Medland2009} Standard QC filters were applied and the remaining 528,509 SNPs were carried forward for further analysis.
 
@@
 The Illumina HT-12 v4.0 chip contains 47,323 probes, although some probes are not assigned to RefSeq genes.
-We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF > 0.05 within 1000 genomes data; did not map to a listed ref-seq gene; were not significantly expressed (based on a detection $p$-value < 0.05) in 100\% of samples.
-After this stringent QC ~7000 [need to get exact number] probes remained for 2d-eqtl mapping.
+We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF > 0.05 within 1000 genomes data; did not map to a listed ref-seq gene; were not significantly expressed (based on a detection $p$-value < 0.05) in at least 90\% of samples.
+After this stringent QC 7339 probes remained for 2D-eQTL mapping.
 
@@
 
-Gene expression profiles were normalised and batch effects removed with the following procedures; Raw background expression were removed for each sample.
-Expression levels were then adjusted using Quantile and log2 transformation to standardise distribution between samples.
-To avoid inflation of the test statistic due to polygenic effects the following linear model was used to correct for batch and polygenic effects;
+Gene expression profiles were normalised and adjusted for batch and polygenic effects.
+Profiles were first adjusted for raw background expression in each sample.
+Expression levels were then adjusted using quantile and log2 transformation to standardise distributions between samples.
+Batch and polygenic effects were adjusted using the linear model
 
 \begin{equation}
-y = \mu + c + p + s + a + g + e
+y = \mu + \beta_{1}c + \beta_{2}p + \beta_{3}s + \beta_{4}a + g + e
 \label{eq:lm}
 \end{equation}
-where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively.
-$g$ is a random additive polygenic effect with a variance covariance matrix
-
+where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively, fitted as fixed effects; and $g$ is a random additive polygenic effect with a variance covariance matrix
 \begin{equation}
@@
 \end{equation}
+The parameters $\sigma_a^2$ and $\sigma_e^2$ are variance components for additive background genetic and environmental effects respectively.
+Here, we are using family based pedigree information rather than SNP based IDB to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$.
+The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero.
+Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan.
 
-The parameters $\sigma _a ^a$ and $\sigma _e ^2$ are variance components for additive background genetic and environmental effects respectively.
-Here, we are using family based pedigree information rather than SNP based IDB to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$.
-The residual, $e$, from equation \ref{eq:lm} are assumed to follow a multivariate normal distribution with a mean of zero.
-These residuals are corrected for batch, sex, generation and polygenic effects.
-Residuals were standardised to Z-scores using a inverse-normal transformation and used as the adjusted phenotype for the pairwise epistasis scan.
+\subsection{Exhaustive 2D-eQTL analysis}
 
-\subsection{exhaustive 2d-eQTL analysis}
-We used epi-GPU software to perform an exhaustive scan for pairwise interactions, such that each SNP is tested against all other SNPs for statistical association with the expression values for each of the ~7000 probes.
-For each SNP-pair there are 9 possible genotype classes.
+We used epiGPU software to perform an exhaustive scan for pairwise interactions, such that each SNP is tested against all other SNPs for statistical association with the expression values for each of the 7339 probes.
+For each SNP pair there are 9 possible genotype classes.
 We treat each genotype class as a fixed effect and fit an 8 d.f.
@@
 This has been demonstrated to be statistically more efficient when searching for a wide range of epistatic patterns (add refs), although will include any marginal effects of SNPs.
-\\[0.2cm]
 
-The complete exhaustive scan for 7,000 probes comprises of $\sim$9.8$e^{14}$ $F$-tests.
-Due to the very large number of tests [plus nature of epistasis ....] we used a highly conservative series of filtering steps to identify study-wide significant epistasis associations to follow forward for replication analysis.
-Initially a significance threshold of $-log10$ 15.5.
-This threshold is the study-wide significance threshold determined by the number of independent tests for a 2D scan for a single probe / the number of probes analysed.
-This is likely to be conservative as this threshold assumes independence between probes.
-Following filtering on this threshold only SNP pairs with all 9 genotype classes and minimum genotype class size > 5 individuals were kept.
-We then calculated the linkage disequilibrium between SNPs in a pair and removed any pairs with $r^2$ > 0.1.
-We also removed any pairs containing SNPs for which single marker additive or dominance effects had been identified  in previous work (Powell et al.
-in Press).
-Sentinel  SNP pairs we retained from epistasis eQTL `peaks' comprising of multiple sets of pair-wise SNPs.
-At this stage $\sim$13,000 SNP pairs remained and were tested for the contribution of marginal and interaction terms by testing the 8d.f.
-\ref{eq:8df} verses a 4 d.f.
-model parameterizing for interaction terms only:
+The complete exhaustive scan for 7339 probes comprises $1.03 \times 10^{15}$ $F$-tests.
+Due to the very large number of tests it was necessary to devise an appropriate threshold.
+We performed a further 1600 exhaustive 2D scans on permuted phenotypes to generate a null distribution of the extreme $p$-values expected to be obtained from this number of multiple tests given the correlation structure between the SNPs.
+We took the most extreme $p$-value from each of the 1600 scans and set the 5\% FWER to be the 95\% most extreme of these $p$-values, $T_{*} = 2.13 \times 10^{-12}$.
+The effective number of tests in one 2D scan being performed is therefore $N_{*} = 0.05 / T_{*} \approx 2.33 \times 10^{10}$.
+To correct for the testing of multiple traits we established an experiment wide threshold of $T = 0.05 / (N_{*} \times 7339) = 2.91 \times 10^{-16}$.
+This is likely to be conservative as it assumes independence between probes.
 
-\begin{equation}
-H _0: \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \bar x _i - \bar x _j + \mu) ^2 = 0;
-\end{equation}
+Following filtering on this threshold only SNP pairs with at least 5 data points in all 9 genotype classes were kept.
+We then calculated the LD between interacting SNPs and removed any pairs with $r^2 > 0.1$.
+If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained.
+At this stage 6404 filtered SNP pairs remained.
+We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.0 \times 10^{-10}$) was removed, creating a second set of 4751 filtered SNP pairs.
+To ensure that interacting SNPs were driven by epistasis and not marginal effects we performed a nested ANOVA on each filtered pair to test if the interaction terms were significant.
+We did this by contrasting the full genetic model (8 d.f.) against the reduced marginal effects model which included the additive and dominance terms at both SNPs (4 d.f.).
+Thus, a 4 d.f.
+$F$-test was performed on the residual genetic variation, representing the contribution of epistatic variance.
+Significance of epistasis was determined using a Bonferroni threshold of $0.05 / (6404+4751) = 4.48 \times 10^{-6}$.
+This resulted in 432 and 117 SNP pairs with significant interaction terms from filters 1 and 2, respectively.
 
-\begin{equation}
-H _1: \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \bar x _i - \bar x _j + \mu) ^2 > 0;
-\label{eq:4df}
-\end{equation}
+\subsection{Replication}
+We have attempted replication of the 2D scan significant results using three independent cohorts; Fehrmann, EGCUT, and CHDWB. Details of cohorts are as follows. \\[0.1cm]
 
-where $\bar x _i (\bar x _j)$ is the marginal class mean for genotype $i (j)$ at SNP A (B). Significance of interaction terms was determined using a bonferroni threshold of $0.05 / 13000 = 3.8e^{-6}$
-
-We have attempted replication of the 2D scan significant results using three independent cohorts; Fehrmann, CHDWB and Rotterdam. Details of cohorts are as follows. \\[0.1cm]
-
-Fehrmann: $n$=1,469 \\[0.1cm]
-The Fehrmann dataset consists of whole peripheral blood samples of 1,469 unrelated individuals from the United Kingdom and the Netherlands.
+Fehrmann: $n=1240$ \\[0.1cm]
+The Fehrmann dataset consists of whole peripheral blood samples of 1240 unrelated individuals from the United Kingdom and the Netherlands.
 Some of these individuals are patients, while others are healthy controls.
-Individuals were genotyped using Illumina HumanHap300, HumanHap370 or the 610 Quad platform.
-RNA levels were quantified using both the Illumina H8v2 platform (N = 229) and the HT12v3 platform (N = 1,240).
+Individuals were genotyped using the Illumina 610 Quad platform.
+RNA levels were quantified using the HT12v3 platform.
 \\[0.1cm]
 
-CDHWB: $n$=189 \\ [0.1cm]
-The Center for Health Discovery and Well Being (CDHWB) Study is a population based cohort consisting of 189 individuals collected in Atlanta USA.
+EGCUT: $n=891$ \\[0.1cm]
+The Estonian Genome Center of the University of Tartu (EGCUT) study consists of whole blood samples of 891 unrelated individuals from Estonia.
+They were genotyped using the Illumina 610 Quad platform.
+RNA levels were quantified using the HT12v3 platform.
+
+CDHWB: $n=139$ \\ [0.1cm]
+The Center for Health Discovery and Well Being (CDHWB) Study is a population based cohort consisting of 139 individuals of European descent collected in Atlanta USA.
 Gene expression profiles were generated with Illumina HT-12 V3.0 arrays from whole blood collected from Tempus tubes that preserve RNA.
@@
 \\[0.1cm]
-
-Rotterdam: $n$=762 \\[0.1cm]
-The Rotterdam Study is a large population based cohort collected to investigate the prevalence, incidence, and risk factors of chronic diseases among  Caucasians aged 45 years and over.
-Participants were genotyped using the Illumina 610K quad arrays.
-Whole blood of 768 samples was collected (PAXgene)and RNA hybridized to the Illumina Human HT-12v4 chips.
 

### 2013-05-29 docs/manuscript/nature_letter.tex
added variance analysis scripts

@@
 Recent hardware and software advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
-We used permutation analysis to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (Methods).
+We used permutation analysis to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
 SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
@@
 If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained.
-Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 549 significant interactions after Bonferroni correction for multiple testing.
+Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 549 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
 
@@
 \subsubsection{Functional annotation}
+Of the 549 interactions, 66 were cis-cis acting (both SNPs were on the same chromosome as the expression gene), 470 were cis-trans acting, and 3 were trans-trans acting.
+In total the 549 interactions comprised 835 unique SNPs, which we analysed for functional enrichment.
+We observed that there was significant enrichment of cis-acting SNPs in transcriptionally active regions for haematopoetic cell lines only, as measured by localisation within H3K4ME3 regions ($p = ?$).
+
 - Functional annotation shows that proteins are not being changed at the codon level
@@
 There is an ascertainment bias in that there are no genetic interactions strong enough to be detected on their own, and large additive effects are required to push them above the threshold
-
-\clearpage
-\section{Methods summary}
 
@@
 
-where $\mu$ is the mean expression level and $x _{ij}$ is the pairwise genotype class mean for genotype $i$ at SNP 1 and genotype $j$ at locus B.
+where $\mu$ is the mean expression level and $x _{ij}$ is the pairwise genotype class mean for genotype $i$ at SNP 1 and genotype $j$ at SNP 2.
 This type of test does not parameterize for specific types of epistasis, rather it tests for the joint genetic effects at two loci.
-This has been demonstrated to be statistically more efficient when searching for a wide range of epistatic patterns (add refs), although will include any marginal effects of SNPs.
+This has been demonstrated to be statistically more efficient when searching for a wide range of epistatic patterns, although will include any marginal effects of SNPs.
 
 The complete exhaustive scan for 7339 probes comprises $1.03 \times 10^{15}$ $F$-tests.
-Due to the very large number of tests it was necessary to devise an appropriate threshold.
-We performed a further 1600 exhaustive 2D scans on permuted phenotypes to generate a null distribution of the extreme $p$-values expected to be obtained from this number of multiple tests given the correlation structure between the SNPs.
+We used permutation analysis to estimate an appropriate significance threshold for the study.
+To do this we performed a further 1600 exhaustive 2D scans on permuted phenotypes to generate a null distribution of the extreme $p$-values expected to be obtained from this number of multiple tests given the correlation structure between the SNPs.
 We took the most extreme $p$-value from each of the 1600 scans and set the 5\% FWER to be the 95\% most extreme of these $p$-values, $T_{*} = 2.13 \times 10^{-12}$.
@@
 \subsection{Replication}
-We have attempted replication of the 2D scan significant results using three independent cohorts; Fehrmann, EGCUT, and CHDWB. Details of cohorts are as follows. \\[0.1cm]
+We have attempted replication of the 529 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB. Details of the cohorts are as follows. \\[0.1cm]
 

### 2013-05-29 analysis/variance_analysis.R

@@ -1,0 +1,35 @@
+sig$type <- "cis-cis"
+sig$type[with(sig, chr1 == probechr & chr2 != probechr)] <- "cis-trans"
+sig$type[with(sig, chr1 != probechr & chr2 == probechr)] <- "cis-trans"
+sig$type[with(sig, chr1 != probechr & chr2 != probechr)] <- "trans-trans"
+
+for(i in 1:nrow(sig))
+{
+	sig$vc_fehr[[i]] <- sig$vc_fehr[[i]][[1]]
+	sig$vc_egcut[[i]] <- sig$vc_egcut[[i]][[1]]
+}
+
+bsgs <- as.data.frame(do.call(rbind, sig$vc))
+fehr <- as.data.frame(do.call(rbind, sig$vc_fehr))
+egcut <- as.data.frame(do.call(rbind, sig$vc_egcut))
+
+bsgs$varA <- bsgs$a. + bsgs$.a
+fehr$varA <- fehr$a. + fehr$.a
+egcut$varA <- egcut$a. + egcut$.a
+
+bsgs$varD <- bsgs$d. + bsgs$.d
+fehr$varD <- fehr$d. + fehr$.d
+egcut$varD <- egcut$d. + egcut$.d
+
+bsgs$varI <- bsgs$aa + bsgs$ad + bsgs$da + bsgs$dd
+fehr$varI <- fehr$aa + fehr$ad + fehr$da + fehr$dd
+egcut$varI <- egcut$aa + egcut$ad + egcut$da + egcut$dd
+
+
+varA <- data.frame(bsgs = bsgs$varA, fehr = fehr$varA, egcut = egcut$varA)
+varD <- data.frame(bsgs = bsgs$varD, fehr = fehr$varD, egcut = egcut$varD)
+varI <- data.frame(bsgs = bsgs$varI, fehr = fehr$varI, egcut = egcut$varI)
+
+pairs(varA)
+pairs(varD)
+pairs(varI)

### 2013-05-30 docs/manuscript/nature_letter.tex
majority of writing is complete. need to finalise discussion

@@
 \begin{abstract}
-A long standing question in human genetics is the extent to which epistasis, the phenomenon whereby one mutation's effect on a trait depends on other mutations present in the genome, contributes to complex traits.
-Though epistasis is known to arise from artificial gene manipulation studies in model organisms, and occasional examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits.
+A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits.
+Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms, and some examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits.
 Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits, but we hypothesised that it has previously been too technically difficult to detect due to statistical power and computational issues.
-Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, evidence for abundant epistasis is found.
+Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, evidence for multiple instances of epistasis is found.
 In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
 We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset.
-Several genes are controlled by multi-locus epistatic interactions whereby one cis-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
-For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by trans-SNPs on 18 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction.
+Several genes are controlled by multi-locus epistatic interactions whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
+For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
 This study presents the first strong evidence for the widespread existence of epistasis emerging from natural genetic variation in humans.
@@
 Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale in artificial double gene knockouts in experimental organisms but also at the evolutionary scale in fitness adaptation and speciation.
-Though its importance is frequently the subject of debate so far there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits.
+Though its importance is frequently the subject of debate, to date there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits.
 
@@
 
-\subsection{Results}
-
-\subsubsection{Initial search}
-We searched for pairwise epistasis exhaustively in the BSGS dataset which comprises 846 individuals who are genotyped at 528,509 SNPs and who have gene expression levels measured in whole blood samples for 7339 genes.
+\subsection{Initial search in discovery set}
+We searched for pairwise epistasis exhaustively in the BSGS dataset, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 genes.
 Recent hardware and software advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
@@
 Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
-All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least 5 data points in all nine genotype classes.
+All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least five data points in all nine genotype classes.
 If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained.
@@
 
-\subsubsection{Replication}
+\subsection{Replication}
 Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than it is for independent additive effects.
@@
 Of the 480 original pairs that passed filtering in Fehrmann and EGCUT, 421 of the SNP pairs showed replication for the interaction test at the 5\% false discovery rate (FDR) level in at least one dataset, and 148 in both datasets (Supplementary Figure \ref{fig:qqplotfdr}).
-The congruence of the epistatic networks in both datasets is shown in Figure \ref{fig:fireworks}.
+The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}.
 We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and 18 were significant in both.
@@
 
-\subsubsection{Functional annotation}
-Of the 549 interactions, 66 were cis-cis acting (both SNPs were on the same chromosome as the expression gene), 470 were cis-trans acting, and 3 were trans-trans acting.
+\subsection{Functional annotation}
+Of the 549 interactions, 66 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 470 were \emph{cis}-\emph{trans} acting, and 13 were \emph{trans}-\emph{trans} acting.
 In total the 549 interactions comprised 835 unique SNPs, which we analysed for functional enrichment.
-We observed that there was significant enrichment of cis-acting SNPs in transcriptionally active regions for haematopoetic cell lines only, as measured by localisation within H3K4ME3 regions ($p = ?$).
+We observed that there was significant enrichment of \emph{cis}-acting SNPs in transcriptionally active regions for haematopoetic cell lines only, as measured by localisation within H3K4me3 active regions (the three tissues with the strongest enrichment are significant at $p < 1 \times 10^{-4}$, after adjusting for multiple testing), but \emph{trans}-acting SNPs did not show tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:h3k4me3}).
 
 - Functional annotation shows that proteins are not being changed at the codon level
+
 - There is strong enrichment for enhancers
+
 - There is tissue specific transcription regulation for cis SNPs - these are regulatory
+
 - Trans SNPs are not found in tissue specific transcription regions.
+
 - General enrichment for regulatory motif modifications
+
 - Names of specific factors that are enriched.
+
 - Chromosome interactions?
 
-\subsubsection{Patterns of epistasis}
-- There are several examples of gene expression being controlled by a cis-SNP which in turn is modulated by multiple trans-SNPs. The effects are consistent 
-- No explicit mechanism arises from looking at GP maps
-- Describe MBNL1
+\subsection{Patterns of epistasis}
+Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions, but our analysis shows this is unlikely to be sufficient.
+The majority of interactions comprised of one SNP that had a previously known association and one that had no previous association (476 out of 549).
+Only 9 interactions were between SNPs that both had marginal effects and 64 were between SNPs that had no known marginal effects.
+Additionally, we observed that the largest epistatic variance component for the 529 interactions were proportionately divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms ($p = 0.22$ for divergence from expectation), thus choosing a particular parameterisation \emph{a priori} may not be warranted.
+This is expected as these traditional epistatic variance components are statistical formulations rather than biologically meaningful patterns.
 
-\subsubsection{Characterisation of genetic variation}
-- The proportion of the variance that is due to interactions is generally smaller than additive.
-There is an ascertainment bias in that there are no genetic interactions strong enough to be detected on their own, and large additive effects are required to push them above the threshold
+We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common form of epistasis detected involved a homozygous \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
+For example, MBNL1 has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit this masking pattern.
+Each of these interactions have evidence for replication in at least one dataset and eight have evidence for replication in both datasets.
+Though the exact mechanism underlying this interaction requires further research, we observed that nine of the 14 \emph{trans} SNPs are located in intronic regions (significantly enriched, $p = 3.11 \times 10^{-9}$) and MBNL1 is involved in RNA modification and regulation of splicing, suggesting a putative mechanism by which epistatic interactions may regulate transcription.
+
+\subsection{Contribution of epistatic variance}
+Though we have discovered many instances of epistasis, quantifying the relative importance of epistasis remains an open question.
+In this study we are able to identify 249 probes with at least one significant interaction given our experiment-wide threshold.
+Using the same threshold ($p < 2.91 \times 10^{-16}$) for the same data searching for only additive or dominance effects returns 517 probes out of the 7339 analysed with at least one significant effect, therefore the number of instances of epistasis are substantial.
+However, although many epistatic effects are observed, in terms of their contribution to complex traits a more important metric is the proportion of the variance that they explain.
+Ideally one would assess this question from a whole genome level but this is intractable for non-additive variance outside of in humans, but quantifying the ascertained effects detected in this study demonstrates that additive variance is overall a much larger component than epistatic variance.
+We observe that of the significant interactions the average additive contribution to the total phenotypic variance is 11.3\%, over three times higher than the epistatic contribution of 3.7\% (Figure \ref{fig:variancecomponents}).
+From an experiment-wide perspective, the proportion of the total phenotypic variance of all 7339 probes explained by additive effects significant at the $p < 2.91 \times 10^{-16}$ threshold is 2.7\% (?????), whereas the epistasic signals detected in this study only explains 0.27\%.
+It must be stressed that this comparison strongly favours additive variance because the power of a 1 d.f.
+test exceeds that of an 8 d.f.
+test, and the non-additive variance at causal variants is expected to be grossly underestimated by observed SNPs in comparison to estimates for additive variance.
+
+\subsection{Discussion}
+Most genetic studies focus on the additive effects of polymorphisms, and there is a
 
@@
 Full details of genotyping procedures are given in Medland et al.
-\cite{Medland2009} Standard QC filters were applied and the remaining 528,509 SNPs were carried forward for further analysis.
+\cite{Medland2009} Standard QC filters were applied and the remaining 528,509 autosomal SNPs were carried forward for further analysis.
 
@@
 The Illumina HT-12 v4.0 chip contains 47,323 probes, although some probes are not assigned to RefSeq genes.
-We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF > 0.05 within 1000 genomes data; did not map to a listed ref-seq gene; were not significantly expressed (based on a detection $p$-value < 0.05) in at least 90\% of samples.
+We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF > 0.05 within 1000 genomes data; did not map to a listed ref-seq gene; were not significantly expressed (based on a detection $p$-value $< 0.05$) in at least 90\% of samples.
 After this stringent QC 7339 probes remained for 2D-eQTL mapping.
@@
 At this stage 6404 filtered SNP pairs remained.
-We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.0 \times 10^{-10}$) was removed, creating a second set of 4751 filtered SNP pairs.
+We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs.
 To ensure that interacting SNPs were driven by epistasis and not marginal effects we performed a nested ANOVA on each filtered pair to test if the interaction terms were significant.
@@
 \subsection{Replication}
-We have attempted replication of the 529 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB. Details of the cohorts are as follows. \\[0.1cm]
+We have attempted replication of the 529 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB. Details of the cohorts are as follows.
 
-Fehrmann: $n=1240$ \\[0.1cm]
+\paragraph{Fehrmann: $n=1240$}
 The Fehrmann dataset consists of whole peripheral blood samples of 1240 unrelated individuals from the United Kingdom and the Netherlands.
@@
 RNA levels were quantified using the HT12v3 platform.
-\\[0.1cm]
 
-EGCUT: $n=891$ \\[0.1cm]
+\paragraph{EGCUT: $n=891$}
 The Estonian Genome Center of the University of Tartu (EGCUT) study consists of whole blood samples of 891 unrelated individuals from Estonia.
@@
 
-CDHWB: $n=139$ \\ [0.1cm]
+\paragraph{CDHWB: $n=139$}
 The Center for Health Discovery and Well Being (CDHWB) Study is a population based cohort consisting of 139 individuals of European descent collected in Atlanta USA.
@@
 Whole genome genotypes were measured using Illumina OmniQuad arrays.
-\\[0.1cm]
 
@@
 \end{figure}
+\clearpage
 
@@
 \end{figure}
+\clearpage
 
 \begin{figure}
-\caption{\textbf{}}
+	\includegraphics[width=5in]{circles_replication2}
+	\caption{\textbf{Gene expression traits with four or more genetic interactions} Circle plots represent the genomic positions for SNPs (linking lines) and expression probes (red points).
+Chromosomes are represented by black blocks and ordered from 1 to 22 clockwise, starting from the top.
+Grey lines represent no evidence for replication, blue lines denote replication in at least one dataset, and red lines denote replication in two datasets.
+Most interactions are characterised as being \emph{cis}-\emph{trans} to the expression probe.}
+	\label{fig:circleplots}
 \end{figure}
+\clearpage
+
+\begin{figure}
+	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3.
+Here we show that there is }
+	\label{fig:h3k4me3}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\includegraphics[width=5in]{MBNL1.pdf}
+	\caption{\textbf{Genotype-phenotype maps for 14 interactions controlling MBNL1} Each bar represents the mean phenotypic value for individuals in that genotype class.
+The rs13069559 SNP typically has a \emph{cis}-additive decreasing effect on the expression of MBNL1, but in many of these interactions the \emph{cis} effect is masked when the \emph{trans} SNP is homozygous.}
+	\label{fig:mbnl13d}
+\end{figure}
+\clearpage
 

### 2013-05-30 analysis/circles.R

@@ -149,7 +149,7 @@ sig <- subset(sig, filter != 3)
 
 gr <- makeGr()
 index <- table(sig$probename)
-sig_mult <- subset(sig, probename %in% names(index)[index > 2])
+sig_mult <- subset(sig, probename %in% names(index)[index > 3])
 
 # sig_mult <- subset(sig, probegene == "MBNL1")
 # links <- makeLinks(sig_mult[1:13,], gr)
@@ -164,7 +164,7 @@ a <- dlply(sig_mult, .(probename), .progress="text", function(x)
 })
 
 pdf(file="~/repo/eQTL-2D/analysis/images/circles_replication2.pdf", width=25, height=20)
-multiplot(plotlist=a, cols=8)
+multiplot(plotlist=a, cols=6)
 dev.off()
 
 key <- subset(sig, !duplicated(probename), select=c(probename, probegene))

### 2013-05-30 analysis/combine_data.R

@@ -203,12 +203,21 @@ bsgs <- getVcBreakdown(bsgs, xmat, resphen)
 bsgs <- gpMaps(bsgs, xmat, resphen)
 
 # Get associations for marginal SNPs
-bsgs <- marginalSnpAssociations(bsgs, marginal_list, 1e-2, probeinfo_all)
+bsgs <- marginalSnpAssociations(bsgs, marginal_list, 1.29e-11, probeinfo_all)
 
 
 # Get replication pvals
 sig <- mergeBsgsRep(bsgs, fehr, egcut)
 sig <- posData(sig)
+
+
+# The VC lists are weird
+for(i in 1:nrow(sig))
+{
+	sig$vc_fehr[[i]] <- sig$vc_fehr[[i]][[1]]
+	sig$vc_egcut[[i]] <- sig$vc_egcut[[i]][[1]]
+}
+
 
 # Remove filter==3
 sig_all <- sig

### 2013-05-30 analysis/gpmaps.R

@@ -14,8 +14,9 @@ plot3dGp <- function(gp, title="", snp1="SNP1", snp2="SNP2", z=-45)
 		col="black", 
 		col.facet=c("#e5f5e0", "#A1D99B", "#31A354"), 
 		xbase=0.6, ybase=0.6,
-		xlab=snp1, ylab=snp2, zlab="y",
-		default.scales=list(arrows=F),
+		xlab=list(label=snp1, cex=0.8), ylab=list(label=snp2, cex=0.8), zlab="",
+		default.scales=list(arrows=F, z = list(cex = 0), y = list(cex = 0.5), x=list(cex=0.5)),
+		cex.title = 0.5,
 		screen = list(z = z, x = -60, y = 3),
 		main = title
 	)
@@ -58,7 +59,7 @@ plot3dHairballsRep <- function(sig, pg, cissnp, cohort, z=45)
 		}
 		gp <- gp - min(gp, na.rm=T)
 
-		title <- paste(s$probegene[1], "chr", subset(chrkey, snp == cissnp)$chr, "x", subset(chrkey, snp == snp2)$chr)
+		title <- paste("chr", subset(chrkey, snp == cissnp)$chr, "x", subset(chrkey, snp == snp2)$chr)
 		l[[i]] <- plot3dGp(gp, title, cissnp, snp2, z)
 	}
 	do.call(grid.arrange, l)
@@ -165,11 +166,6 @@ l <- GpMod(l, 6, "BSGS", t)
 l <- GpMod(l, 7, "BSGS", rotateGp, rotateGp)
 l <- GpMod(l, 9, "BSGS", rotateGp, rotateGp)
 
-index <- match(mbnl1$code, sig$code)
-mbnl1$code4 <- paste(sig$chr1[index], "x", sig$chr2[index])
-mbnl1 <- subset(mbnl1, !is.na(y))
-mbnl1$code3 <- mbnl1$code
-
 pdf(file="~/repo/eQTL-2D/analysis/images/gpBonfRep.pdf", width=20, height=4)
 plotTileGp(l)
 dev.off()
@@ -207,11 +203,14 @@ plotTileGp(mbnl1)
 
 
 
-plot3dHairballsRep(sig, "TMEM149", "rs8106959", "", z=45)
+plot3dHairballsRep(sig, "TMEM149", "rs8106959", "", z=-45)
 plot3dHairballsRep(sig, "TMEM149", "rs8106959", "_egcut", z=45)
[…]

### 2013-05-30 analysis/variance_analysis.R

@@ -1,3 +1,6 @@
+load("~/repo/eQTL-2D/analysis/interaction_list_replication_summary.RData")
+
+
 sig$type <- "cis-cis"
 sig$type[with(sig, chr1 == probechr & chr2 != probechr)] <- "cis-trans"
 sig$type[with(sig, chr1 != probechr & chr2 == probechr)] <- "cis-trans"
@@ -33,3 +36,31 @@ varI <- data.frame(bsgs = bsgs$varI, fehr = fehr$varI, egcut = egcut$varI)
 pairs(varA)
 pairs(varD)
 pairs(varI)
+
+
+# Proportion of SNPs with known marginal effects
+with(sig, table(is.na(marginal_gene1), is.na(marginal_gene2)))
+
+
+biggestVC <- function(vc)
+{
+	a <- apply(subset(vc, select=c(aa, ad, da, dd)), 1, which.max)
+	a[a == 1] <- "aa"
+	a[a == 2] <- "ad"
+	a[a == 3] <- "ad"
+	a[a == 4] <- "dd"
+	vc$mainI <- a
+	return(vc)
+}
+
+bsgs <- biggestVC(bsgs)
+
+tab <- table(bsgs$mainI)
+chisq.test(x=tab / sum(tab), y=c(0.25, 0.5, 0.25))
+
+
+
+
+
+
+

### 2013-05-30 docs/manuscript/nature_letter.tex
there may be evidence for chromosome x chromosome interactions

@@
 Only 9 interactions were between SNPs that both had marginal effects and 64 were between SNPs that had no known marginal effects.
-Additionally, we observed that the largest epistatic variance component for the 529 interactions were proportionately divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms ($p = 0.22$ for divergence from expectation), thus choosing a particular parameterisation \emph{a priori} may not be warranted.
+Additionally, we observed that the largest epistatic variance component for the 529 interactions were proportionately divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms ($p = 0.22$ for divergence from expectation).
 This is expected as these traditional epistatic variance components are statistical formulations rather than biologically meaningful patterns.
 
-We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common form of epistasis detected involved a homozygous \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
-For example, MBNL1 has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit this masking pattern.
+We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common form of epistasis detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
+For example, MBNL1 has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit a masking pattern such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
 Each of these interactions have evidence for replication in at least one dataset and eight have evidence for replication in both datasets.
@@
 We observe that of the significant interactions the average additive contribution to the total phenotypic variance is 11.3\%, over three times higher than the epistatic contribution of 3.7\% (Figure \ref{fig:variancecomponents}).
-From an experiment-wide perspective, the proportion of the total phenotypic variance of all 7339 probes explained by additive effects significant at the $p < 2.91 \times 10^{-16}$ threshold is 2.7\% (?????), whereas the epistasic signals detected in this study only explains 0.27\%.
+From an experiment-wide perspective, the proportion of the total phenotypic variance of all 7339 probes explained by additive effects significant at the $p < 2.91 \times 10^{-16}$ threshold is 1.9\%, whereas the epistatic signals detected in this study only explains 0.27\%.
 It must be stressed that this comparison strongly favours additive variance because the power of a 1 d.f.

### 2013-05-30 analysis/chromosome_interactions/simulation.R

@@ -1,0 +1,89 @@
+
+# For each of the significant interactions see if the two positions are within 1kb of any of the chromosome interactions
+
+ciOverlap <- function(ci, sig, win)
+{
+	ci$index <- 1:nrow(ci)
+	sig$int1 <- NA
+	sig$int2 <- NA
+	for(i in 1:nrow(sig))
+	{
+		cat(i, "of", nrow(sig), "\n")
+		chr1 <- sig$chr1[i]
+		chr2 <- sig$chr2[i]
+		pos1 <- sig$pos1[i]
+		pos2 <- sig$pos2[i]
+
+		sa <- subset(ci, 
+			loci1_chromosome == chr1 & 
+			loci2_chromosome == chr2 &
+			abs(loci1_position - pos1) <= win &
+			abs(loci2_position - pos2) <= win
+		)
+
+		sb <- subset(ci, 
+			loci1_chromosome == chr2 & 
+			loci2_chromosome == chr1 &
+			abs(loci1_position - pos2) <= win &
+			abs(loci2_position - pos1) <= win
+		)
+
+		if(nrow(sa) > 0)
+		{
+			sig$int1[i] <- list(sa$index)
+			print("Found!")
+		}
+		if(nrow(sb) > 0)
+		{
+			sig$int2[i] <- list(sb$index)
+			print("Found!")
+		}
+	}
+	return(sig)
+}
+
+createFake <- function(bim, n)
+{
+	a <- sample(1:nrow(bim), n, replace=FALSE)
+	b <- sample(1:nrow(bim), n, replace=FALSE)
+
+	dat <- data.frame(chr1=bim$V1[a], pos1=bim$V4[a], chr2=bim$V1[b], pos2=bim$V4[b])
+	return(dat)
+}
+
+
+#=============================================================#
+#=============================================================#
+
+
+arguments <- commandArgs(T)
+
+jid <- as.numeric(arguments[1])
+n <- as.numeric(arguments[2])
+win <- as.numeric(arguments[3])
+cifile <- arguments[4]
+bimfile <- arguments[5]
+outroot <- arguments[6]
+
+output <- paste(outroot, n, win, jid, sep="_")
+
+#=============================================================#
+#=============================================================#
+
+
+ci <- read.csv(cifile, header=T)
+bim <- read.table(bimfile, colClasses=c("numeric", "character", "numeric", "numeric", "character", "character"))
+
+head(ci)
[…]

### 2013-05-30 analysis/chromosome_interactions/simulation.sh

@@ -1,0 +1,26 @@
+#!/bin/bash
+
+#$ -N ci
+#$ -cwd
+#$ -l h_vmem=1G
+#$ -t 1-1000
+#$ -S /bin/bash
+#$ -o job_reports/
+#$ -e job_reports/
+
+set -e
+
+if [ -n "${1}" ]; then
+  echo "${1}"
+  SGE_TASK_ID=${1}
+fi
+
+i=${SGE_TASK_ID}
+
+cifile="~/repo/eQTL-2D/data/supFile3_K562_interactingLoci_clusters.csv"
+bimfile="~/repo/eQTL-2D/data/clean_geno_final.bim"
+n=549
+win=25000
+outroot="~/repo/eQTL-2D/analysis/chromosome_interactions/out"
+
+R --no-save --args ${i} ${n} ${win} ${cifile} ${bimfile} ${outroot} < simulation.R

### 2013-05-30 analysis/functional_analysis.R

@@ -399,3 +399,83 @@ pdf(file="~/repo/eQTL-2D/analysis/images/variance_components.pdf", width=10, hei
 multiplot(p1, p2, cols=2)
 dev.off()
 
+
+
+
+# Chromosome interactions
+
+ci <- read.csv("supFile3_K562_interactingLoci_clusters.csv", header=T)
+dim(ci)
+head(ci)
+table(ci$cluster)
+
+# For each of the significant interactions see if the two positions are within 1kb of any of the chromosome interactions
+
+ciOverlap <- function(ci, sig, win)
+{
+	ci$index <- 1:nrow(ci)
+	sig$int1 <- NA
+	sig$int2 <- NA
+	for(i in 1:nrow(sig))
+	{
+		cat(i, "of", nrow(sig), "\n")
+		chr1 <- sig$chr1[i]
+		chr2 <- sig$chr2[i]
+		pos1 <- sig$pos1[i]
+		pos2 <- sig$pos2[i]
+
+		sa <- subset(ci, 
+			loci1_chromosome == chr1 & 
+			loci2_chromosome == chr2
+		)
+		index <- with(sa, abs(loci1_position - pos1) <= win & abs(loci2_position - pos2) <= win)
+
+		sa <- subset(sa,
+		)
+
+		sb <- subset(ci, 
+			loci1_chromosome == chr2 & 
+			loci2_chromosome == chr1 &
+			abs(loci1_position - pos2) <= win &
+			abs(loci2_position - pos1) <= win
+		)
+
+		if(nrow(sa) > 0)
+		{
+			sig$int1[i] <- list(sa$index)
+			print("Found!")
+		}
+		if(nrow(sb) > 0)
+		{
+			sig$int2[i] <- list(sb$index)
+			print("Found!")
+		}
+	}
+	return(sig)
+}
+
+
+a <- ciOverlap(ci, sig, 10000)
+table(is.na(a$int1))
+
+
+# What are the chances of finding any?
+
+bim <- read.table("~/repo/eQTL-2D/data/clean_geno_final.bim", colClasses=c("numeric", "character", "numeric", "numeric", "character", "character"))
+
+createFake <- function(bim, n)
+{
+	a <- sample(1:nrow(bim), n, replace=FALSE)
+	b <- sample(1:nrow(bim), n, replace=FALSE)
+
[…]

### 2013-05-30 analysis/chromosome_interactions/simulation.R
run simulations to get null dist

@@ -82,7 +82,7 @@ head(fake)
 
 b <- ciOverlap(ci, fake, win)
 
-write.table(sum(!is.na(b$int1) | !is.na(b$int2)), file=output)
+write.table(sum(!is.na(b$int1) | !is.na(b$int2)), file=output, row=F, col=F, qu=F)
 
 
 

### 2013-05-30 analysis/chromosome_interactions/simulation.R
run simulations to get null dist

@@ -83,7 +83,3 @@ head(fake)
 b <- ciOverlap(ci, fake, win)
 
 write.table(sum(!is.na(b$int1) | !is.na(b$int2)), file=output, row=F, col=F, qu=F)
-
-
-
-

### 2013-05-30 docs/manuscript/nature_letter.tex
updated manuscript

@@
 For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
+Additionally, 69 of the genetic interactions overlap with known chromosome interactions ($p < 1.0 \times 10^{-4}$).
 This study presents the first strong evidence for the widespread existence of epistasis emerging from natural genetic variation in humans.
@@
 
+We cross referenced our list of interactions with a map of chromosome interacting regions in K562 blood cell lines.
+Forty-nine epistatic interactions mapped to within 50kb of a chromosome interaction, and 69 mapped to within 2Mb ($p < 1.0 \times 10^{-4}$ in both cases).
+
 - Functional annotation shows that proteins are not being changed at the codon level
@@
 - Names of specific factors that are enriched.
-
-- Chromosome interactions?
 

### 2013-05-30 analysis/functional_analysis.R

@@ -455,9 +455,11 @@ ciOverlap <- function(ci, sig, win)
 }
 
 
-a <- ciOverlap(ci, sig, 10000)
+a <- ciOverlap(ci, sig, 250000)
 table(is.na(a$int1))
 
+b <- ciOverlap(ci, sig, 1000000)
+table(is.na(b$int1))
 
 # What are the chances of finding any?
 

### 2013-05-30 docs/manuscript/nature_letter.tex
added kostya's enhancer image

@@
 We cross referenced our list of interactions with a map of chromosome interacting regions in K562 blood cell lines.
-Forty-nine epistatic interactions mapped to within 50kb of a chromosome interaction, and 69 mapped to within 2Mb ($p < 1.0 \times 10^{-4}$ in both cases).
+Forty-nine epistatic interactions mapped to within 500kb of a chromosome interaction, and 69 mapped to within 2Mb ($p < 1.0 \times 10^{-4}$ in both cases, Supplementary Figure \ref{fig:chromosomeinteractions}).
 
-- Functional annotation shows that proteins are not being changed at the codon level
-
-- There is strong enrichment for enhancers
-
-- There is tissue specific transcription regulation for cis SNPs - these are regulatory
-
-- Trans SNPs are not found in tissue specific transcription regions.
+There is strong enrichment for SNPs to be localised in enhancer binding regions (Supplementary Figure \ref{fig:enhancers}). This enrichment is consistent for both \emph{cis} and \emph{trans} SNPs.
 
@@
 - Names of specific factors that are enriched.
+
+- Functional annotation shows that proteins are not being changed at the codon level
 
@@
 \begin{figure}
-	\includegraphics[width=5in]{MBNL1.pdf}
+	\includegraphics[width=5in]{MBNL1}
 	\caption{\textbf{Genotype-phenotype maps for 14 interactions controlling MBNL1} Each bar represents the mean phenotypic value for individuals in that genotype class.
@@
 
+\begin{figure}
+	\includegraphics[width=5in]{chromosome_interactions}
+	\caption{\textbf{Number of overlaps between chromosome interactions and epistatic interactions} Interacting chromosome regions may be a possible mechanism underlying epistatic interactions.
+The number of epistatic interactions within 20kb, 500kb, 2Mb and 10Mb of known chromosome interacting regions are shown by red vertical lines.
+The histograms represent the null distribution based on random sampling of 10000 datasets for each window size.}
+	\label{fig:chromosomeinteractions}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\includegraphics[width=5in]{enhancers}
+	\caption{\textbf{There is enrichment for enhancer sequences for \emph{cis} and \emph{trans} SNPs}}
+	\label{fig:enhancers}
+\end{figure}
+\clearpage
+
 \end{document}

### 2013-05-30 analysis/functional_analysis.R

@@ -455,11 +455,20 @@ ciOverlap <- function(ci, sig, win)
 }
 
 
-a <- ciOverlap(ci, sig, 250000)
-table(is.na(a$int1))
 
-b <- ciOverlap(ci, sig, 1000000)
-table(is.na(b$int1))
+counts <- rep(0, 4)
+
+a1 <- ciOverlap(ci, sig, 10000)
+counts[1] <- sum(!is.na(a1$int1))
+
+a2 <- ciOverlap(ci, sig, 250000)
+counts[2] <- sum(!is.na(a2$int1))
+
+a3 <- ciOverlap(ci, sig, 1000000)
+counts[3] <- sum(!is.na(a3$int1))
+
+a4 <- ciOverlap(ci, sig, 5000000)
+counts[4] <- sum(!is.na(a4$int1))
 
 # What are the chances of finding any?
 
@@ -477,6 +486,39 @@ createFake <- function(bim, n)
 fake <- createFake(bim, 549)
 
 b <- ciOverlap(ci, fake, 2000000)
+
+# Permutations performed on cluster. Plot results
+
+load("~/repo/eQTL-2D/analysis/chromosome_interactions/collate_549_5e+06.RData")
+perm10mb <- a
+load("~/repo/eQTL-2D/analysis/chromosome_interactions/collate_549_1e+06.RData")
+perm2mb <- a
+load("~/repo/eQTL-2D/analysis/chromosome_interactions/collate_549_250000.RData")
+perm500kb <- a
+load("~/repo/eQTL-2D/analysis/chromosome_interactions/collate_549_10000.RData")
+perm20kb <- a
+
+perms <- data.frame(
+	n=c(sample(c(0,1), 10000, replace=T), perm2mb, sample(c(0,1), 10000, replace=T), sample(c(0,1), 10000, replace=T)), 
+	window=rep(c("10Mb", "2Mb", "500kb", "20kb"), each=10000))
+
+perms <- data.frame(
+	n=c(perm10mb, perm2mb, perm50kb, perm20kb), 
+	window=rep(c("10Mb", "2Mb", "500kb", "20kb"), each=10000))
+
+observed <- data.frame(
+	n=c(88, 69, 49, 8), 
+	window=rep(c("10Mb", "2Mb", "500kb", "20kb")))
+
+perms$window <- factor(perms$window, levels=c("10Mb", "2Mb", "500kb", "20kb"))
[…]

### 2013-05-30 docs/manuscript/nature_letter.tex
added new graphs

@@
 In total the 549 interactions comprised 835 unique SNPs, which we analysed for functional enrichment.
-We observed that there was significant enrichment of \emph{cis}-acting SNPs in transcriptionally active regions for haematopoetic cell lines only, as measured by localisation within H3K4me3 active regions (the three tissues with the strongest enrichment are significant at $p < 1 \times 10^{-4}$, after adjusting for multiple testing), but \emph{trans}-acting SNPs did not show tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:h3k4me3}).
+We observed that there was significant enrichment of \emph{cis}-acting SNPs in transcriptionally active regions for haematopoetic cell lines only, as measured by localisation within H3K4me3 active regions (the three tissues with the strongest enrichment are significant at $p < 1 \times 10^{-4}$, after adjusting for multiple testing, Supplementary Figure \ref{fig:transh3k4me3}), but \emph{trans}-acting SNPs did not show tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:transh3k4me3}).
 
@@
 \begin{figure}
+	\includegraphics[width=5in]{cis_h3k4me3}
 	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3.
 Here we show that there is }
-	\label{fig:h3k4me3}
+	\label{fig:cish3k4me3}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=5in]{trans_h3k4me3}
+	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3.
+Here we show that there is }
+	\label{fig:transh3k4me3}
 \end{figure}

### 2013-05-30 analysis/functional_analysis.R
awaiting permutations to finish

@@ -498,9 +498,11 @@ perm500kb <- a
 load("~/repo/eQTL-2D/analysis/chromosome_interactions/collate_549_10000.RData")
 perm20kb <- a
 
-perms <- data.frame(
-	n=c(sample(c(0,1), 10000, replace=T), perm2mb, sample(c(0,1), 10000, replace=T), sample(c(0,1), 10000, replace=T)), 
-	window=rep(c("10Mb", "2Mb", "500kb", "20kb"), each=10000))
+binom.test(x=69, n=549, p=mean(perm2mb)/549)$p.value
+
+# perms <- data.frame(
+# 	n=c(sample(c(0,1), 10000, replace=T), perm2mb, sample(c(0,1), 10000, replace=T), sample(c(0,1), 10000, replace=T)), 
+# 	window=rep(c("10Mb", "2Mb", "500kb", "20kb"), each=10000))
 
 perms <- data.frame(
 	n=c(perm10mb, perm2mb, perm50kb, perm20kb), 

### 2013-05-30 docs/manuscript/nature_letter.tex
awaiting permutations to finish
chromosome interactions completed. need to see which ones are hits

@@
 For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-Additionally, 69 of the genetic interactions overlap with known chromosome interactions ($p < 1.0 \times 10^{-4}$).
+Additionally, 69 of the genetic interactions overlap with known chromosome interactions ($p < 1.1 \times 10^{-70}$).
 This study presents the first strong evidence for the widespread existence of epistasis emerging from natural genetic variation in humans.
@@
 \subsection{Discussion}
-Most genetic studies focus on the additive effects of polymorphisms, and there is a
+
+There are a likely to be a large number of mechanisms by which epistasis may occur, as evidenced by the functional analaysis of our discovery set.
+Therefore it is not surprising that epistasis should be widespread.
+Further research is required to elucidate specific mechanisms by which statistical epistasis manifests from molecular action, and to quantify the contribution of epistasis to the genetic variance of complex traits.
 

### 2013-05-30 analysis/functional_analysis.R

@@ -495,7 +495,7 @@ load("~/repo/eQTL-2D/analysis/chromosome_interactions/collate_549_1e+06.RData")
 perm2mb <- a
 load("~/repo/eQTL-2D/analysis/chromosome_interactions/collate_549_250000.RData")
 perm500kb <- a
-load("~/repo/eQTL-2D/analysis/chromosome_interactions/collate_549_10000.RData")
+load("~/repo/eQTL-2D/analysis/chromosome_interactions/collate_549_5000.RData")
 perm20kb <- a
 
 binom.test(x=69, n=549, p=mean(perm2mb)/549)$p.value
@@ -505,7 +505,7 @@ binom.test(x=69, n=549, p=mean(perm2mb)/549)$p.value
 # 	window=rep(c("10Mb", "2Mb", "500kb", "20kb"), each=10000))
 
 perms <- data.frame(
-	n=c(perm10mb, perm2mb, perm50kb, perm20kb), 
+	n=c(perm10mb, perm2mb, perm500kb, perm20kb), 
 	window=rep(c("10Mb", "2Mb", "500kb", "20kb"), each=10000))
 
 observed <- data.frame(

### 2013-05-31 docs/manuscript/nature_letter.tex
manuscript has been vetted by kostya and joseph

--- 2013-05-30 docs/manuscript/nature_letter.tex
@@
 \begin{abstract}
-A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits.
-Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms, and some examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits.
-Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits, but we hypothesised that it has previously been too technically difficult to detect due to statistical power and computational issues.
-Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, evidence for multiple instances of epistasis is found.
+A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits \cite{Phillips1998, Carlborg2004, Hill2008a, Crow2010}.
+Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
+Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Greene2009, Hemani2013}.
+Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design where any effects are expected to be large \cite{Powell2012}, evidence for multiple instances of epistasis is found.
 In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
-We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset.
-Several genes are controlled by multi-locus epistatic interactions whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
+We attempted replication in two independent datasets \cite{Metspalu2004, Fehrmann2011} and 421 show evidence of significance in at least one dataset.
+We provide evidence of functional enrichment for the interacting SNPs, for instance 49 of the genetic interactions are located within 500kb of known chromosome interactions \cite{Lan2012} ($p < 1.1 \times 10^{-70}$).
+One hundred and twenty-nine genes are controlled by multi-locus epistatic interactions whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
 For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-Additionally, 69 of the genetic interactions overlap with known chromosome interactions ($p < 1.0 \times 10^{-4}$).
-This study presents the first strong evidence for the widespread existence of epistasis emerging from natural genetic variation in humans.
+This study presents the first strong evidence for the widespread existence of epistatic genetic variation emerging from natural genetic variation in humans.
 \end{abstract}
@@
 The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits.
-Typically, SNPs are treated as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying hundreds of causal variants.
-Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale in artificial double gene knockouts in experimental organisms but also at the evolutionary scale in fitness adaptation and speciation.
-Though its importance is frequently the subject of debate, to date there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits.
+Typically, SNPs are treated as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying thousands of causal variants \cite{Hindorff2010, Visscher2012}.
+Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale in artificial double gene knockouts in experimental organisms \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
+Though its importance is frequently the subject of debate \cite{Phillips1998, Carlborg2004, Hill2008a, Crow2010}, to date there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits in humans \cite{Strange2010, Evans2011}.
 
-The detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs, increased model complexity in fitting interaction terms, and more extreme significance thresholds to account for increased multiple testing \cite{Hemani2013}.
-When genetic effect sizes are small, as is expected in most complex traits of interest, the power to detect epistasis diminishes rapidly.
-There are two ways to overcome this problem, either use extremely large sample sizes, or use traits that are likely to have large effect sizes.
+Detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2002}.
+When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Yang2011}, the power to detect epistasis diminishes rapidly.
+There are two ways to overcome this problem, either by using extremely large sample sizes \cite{LangoAllen2010}, or use traits that are likely to have large effect sizes.
 Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we opted for the latter approach and searched for epistasis controlling gene expression levels.
-Though they number in the thousands, these traits are typically heritable but much less polygenic, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis should it exist.
+Transcription levels can be measured for thousands of genes and these traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis should it exist.
 
 \subsection{Initial search in discovery set}
-We searched for pairwise epistasis exhaustively in the BSGS dataset, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 genes.
-Recent hardware and software advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
-We used permutation analysis to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
+We searched for pairwise epistasis exhaustively in the Brisbane Systems Genetics Study (BSGS) dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes.
+Recent hardware \cite{Fan:2004:GCH:1048933.1049991} and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
+We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
 SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
-Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test.
+Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Evans2006, Hemani2013}.
 Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
@@
 \subsection{Replication}
-Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than it is for independent additive effects.
-The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, Fehrmann ($n=1240$) and EGCUT ($n=891$).
-There was substantial evidence for overlap of interaction effects in all three datasets.
-Of the 480 original pairs that passed filtering in Fehrmann and EGCUT, 421 of the SNP pairs showed replication for the interaction test at the 5\% false discovery rate (FDR) level in at least one dataset, and 148 in both datasets (Supplementary Figure \ref{fig:qqplotfdr}).
-The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}.
+Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
+The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
+We observed substantial replication of interaction signals in both datasets.
+Of the 480 original pairs that passed filtering (Methods) in Fehrmann and EGCUT, 421 of the SNP pairs showed replication for the interaction test at the 5\% false discovery rate (FDR) level in at least one dataset, and 148 in both datasets (Supplementary Figure \ref{fig:qqplotfdr}).
+The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations.
 We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and 18 were significant in both.
 The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}).
-We also attempted replication in the CHDWB dataset, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power there was no evidence for replication.
+A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication.
+
+\subsection{Patterns of epistasis}
+Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2002, Cordell2009}, but our analysis shows this is unlikely to be the most effective framework for its detection.
+The majority of interactions comprised of one SNP that had a previously known association and one that had no previous association in the dataset \cite{Powell2013} (476 out of 549).
+Only 9 interactions were between SNPs that both had marginal effects while 64 were between SNPs that had no known marginal effects.
+Additionally, we observed that the largest epistatic variance component for the 529 interactions were divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by change ($p = 0.22$ for divergence from expectation).
+This is perhaps to be expected because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.
+
+We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
+For example, MBNL1 (involved in RNA modification and regulation of splicing) has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
+Each of these interactions have evidence for replication in at least one dataset and eight have evidence for replication in both datasets (Supplementary Figure \ref{fig:circleplots}).
+We observed that nine of the 14 \emph{trans} SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$), suggesting a putative mechanism by which epistatic interactions may regulate transcription.
 
 \subsection{Functional annotation}
-Of the 549 interactions, 66 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 470 were \emph{cis}-\emph{trans} acting, and 13 were \emph{trans}-\emph{trans} acting.
+Of the 549 interactions, 66 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 470 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
 In total the 549 interactions comprised 835 unique SNPs, which we analysed for functional enrichment.
-We observed that there was significant enrichment of \emph{cis}-acting SNPs in transcriptionally active regions for haematopoetic cell lines only, as measured by localisation within H3K4me3 active regions (the three tissues with the strongest enrichment are significant at $p < 1 \times 10^{-4}$, after adjusting for multiple testing, Supplementary Figure \ref{fig:transh3k4me3}), but \emph{trans}-acting SNPs did not show tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:transh3k4me3}).
-
-We cross referenced our list of interactions with a map of chromosome interacting regions in K562 blood cell lines.
-Forty-nine epistatic interactions mapped to within 500kb of a chromosome interaction, and 69 mapped to within 2Mb ($p < 1.0 \times 10^{-4}$ in both cases, Supplementary Figure \ref{fig:chromosomeinteractions}).
-
-There is strong enrichment for SNPs to be localised in enhancer binding regions (Supplementary Figure \ref{fig:enhancers}). This enrichment is consistent for both \emph{cis} and \emph{trans} SNPs.
-
-- General enrichment for regulatory motif modifications
-
-- Names of specific factors that are enriched.
+We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013}.
+There was significant enrichment for \emph{cis}-acting SNPs in haematopoeitic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment are significant at after adjusting for multiple testing, Supplementary Figure \ref{fig:transh3k4me3}).
+However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:transh3k4me3}).
+This difference between \emph{cis} and \emph{trans} SNPs suggests that there is a range of molecular mechanisms by which epistasis might arise.
+There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a} (Supplementary Figure \ref{fig:enhancers}).
+This enrichment is consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$).
+In particular, there was substantial enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs ($p = 1 \times 10^{-40}$).
 
@@
 
-\subsection{Patterns of epistasis}
-Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions, but our analysis shows this is unlikely to be sufficient.
-The majority of interactions comprised of one SNP that had a previously known association and one that had no previous association (476 out of 549).
-Only 9 interactions were between SNPs that both had marginal effects and 64 were between SNPs that had no known marginal effects.
-Additionally, we observed that the largest epistatic variance component for the 529 interactions were proportionately divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms ($p = 0.22$ for divergence from expectation).
-This is expected as these traditional epistatic variance components are statistical formulations rather than biologically meaningful patterns.
-
-We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common form of epistasis detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
-For example, MBNL1 has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit a masking pattern such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
-Each of these interactions have evidence for replication in at least one dataset and eight have evidence for replication in both datasets.
-Though the exact mechanism underlying this interaction requires further research, we observed that nine of the 14 \emph{trans} SNPs are located in intronic regions (significantly enriched, $p = 3.11 \times 10^{-9}$) and MBNL1 is involved in RNA modification and regulation of splicing, suggesting a putative mechanism by which epistatic interactions may regulate transcription.
+We also demonstrate another putative novel mechanism by which biological function can lead to epistatic genetic variance.
+We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012}.
+Forty-nine epistatic SNP pairs mapped to within 500kb of a chromosome interaction, and 69 mapped to within 2Mb ($p < 1.1 \times 10^{-70}$), (Supplementary Figure \ref{fig:chromosomeinteractions}), but decyphering the exact cellular processes underlying this interaction requires further research.
 
 \subsection{Contribution of epistatic variance}
-Though we have discovered many instances of epistasis, quantifying the relative importance of epistasis remains an open question.
+Though we have discovered many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question.
 In this study we are able to identify 249 probes with at least one significant interaction given our experiment-wide threshold.
-Using the same threshold ($p < 2.91 \times 10^{-16}$) for the same data searching for only additive or dominance effects returns 517 probes out of the 7339 analysed with at least one significant effect, therefore the number of instances of epistasis are substantial.
-However, although many epistatic effects are observed, in terms of their contribution to complex traits a more important metric is the proportion of the variance that they explain.
-Ideally one would assess this question from a whole genome level but this is intractable for non-additive variance outside of in humans, but quantifying the ascertained effects detected in this study demonstrates that additive variance is overall a much larger component than epistatic variance.
-We observe that of the significant interactions the average additive contribution to the total phenotypic variance is 11.3\%, over three times higher than the epistatic contribution of 3.7\% (Figure \ref{fig:variancecomponents}).
-From an experiment-wide perspective, the proportion of the total phenotypic variance of all 7339 probes explained by additive effects significant at the $p < 2.91 \times 10^{-16}$ threshold is 1.9\%, whereas the epistatic signals detected in this study only explains 0.27\%.
-It must be stressed that this comparison strongly favours additive variance because the power of a 1 d.f.
-test exceeds that of an 8 d.f.
-test, and the non-additive variance at causal variants is expected to be grossly underestimated by observed SNPs in comparison to estimates for additive variance.
+Using the same threshold ($p < 2.91 \times 10^{-16}$) for the same data but searching for only additive effects returns 517 probes out of the 7339 analysed with at least one significant expression quantitative trait locus (eQTL) \cite{Powell2012}, therefore we argue that the number of instances of epistasis are substantial.
+However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}.
+Ideally one would approach this question from a whole genome level \cite{Visscher2008} but this is intractable for non-additive variance components.
+Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance as has been argued previously \cite{Hill2008a, Crow2010}.
+We observe that of the significant interactions detected the average additive contribution to the total detected genetic variance is 72.7\%, over three times higher than the epistatic contribution of 24.1\% (Figure \ref{fig:variancecomponents}).
+
+From an experiment-wide perspective, the proportion of the total phenotypic variance for all 7339 probes explained by additive effects significant at the $p < 2.91 \times 10^{-16}$ threshold, as reported in an independent study on the same dataset that focused on additive effects only \cite{Powell2012}, is 1.9\%.
+In contrast, the epistatic signals detected in this study explains 0.27\%, approximately seven times lower than the additive variance.
+There are two caveats to this comparison, firstly the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
+Secondly the non-additive variance at causal variants is expected to be grossly underestimated by observed SNPs in comparison to estimates for additive variance \cite{Weir2008, Hemani2013}, due to differences in the rate of decay estiamted variance between observed and causal SNPs as LD decreases.
+Therefore this is likely to be a lower bound on the estimate of the relative contribution of epistasis.
 
 \subsection{Discussion}
-Most genetic studies focus on the additive effects of polymorphisms, and there is a
+Overall, we have demonstrated that it is possible to identify and replicate epistasis in complex traits amongst common human variants.
+The functional analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation, thus it is not surprising that instances of epistasis should be widespread.
+Further research into such genetic effects may provide a useful portal to understanding molecular mechanisms with greater clarity.
+With data and computational techniques now widely available the search for epistasis in larger datasets for traits of broader interest is warranted.
 
@@
 \begin{figure}[H]
-	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A shows the overall network of genetic interactions (edges) for each probe (red nodes) and their controlling SNPs (black nodes) from the discovery set.
-Panel B shows those interactions that were present at the 5\% FDR level in at least one replication dataset, and panel C shows which interactions were significant in both replication datasets.}
+	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A represents all 549 genetic interactions (edges) discovered in the discovery set.
+Panel B shows those interactions that were present at the 5\% FDR level in at least one replication dataset, and panel C shows which interactions were significant in both replication datasets.
+Red nodes represent probes and black nodes represent SNPs.
+The congruence between datasets is indicitive of the consistent epistatic instances in independent populations.}
 	\label{fig:fireworks}
@@
 \begin{figure}[H]
-	\caption{\textbf{Replication of genotype-phenotype maps in two cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
+	\caption{\textbf{Replication of genotype-phenotype (GP) maps in two cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
 Each tile represents the expression level for that two-locus genotype class.
+Phenotypes are for gene transcript levels, where the gene symbol is labelled in the panel columns.
+Panel rows are labelled for the dataset.
 Phenotypes for each interaction within each cohort are scaled to be between 0 and 1.
-Phenotypes are for gene transcript levels, where the gene's name is labelled in the panel columns.
-Panel rows are labelled for the dataset.
 There is a general trend of the GP maps replicating across all three datasets.}
@@
 \begin{figure}[H]
-	\caption{\textbf{Comparison of genetic variances} Panel A shows the proportion of phenotypic variation that is captured by additive, dominant and interaction components for all 529 significant interactions, ordered by total phenotypic variance explained by the SNP pair.
-Panel B shows the proportion of the genetic variance that is attributable to the three variance components, ordered by the proportion that is additive.}
+	\caption{\textbf{Comparison of genetic variances} Panel A shows the proportion of phenotypic variation ($x$-axis) that is captured by additive, dominant and interaction components for all 529 significant interactions, ordered by total phenotypic variance explained by the SNP pair along the $y$-axis.
+Panel B shows the proportion of the genetic variance ($x$-axis) that is attributable to the three variance components, ordered by the proportion that is additive.}
 	\label{fig:variancecomponents}
@@
 Full details of genotyping procedures are given in Medland et al.
-\cite{Medland2009} Standard QC filters were applied and the remaining 528,509 autosomal SNPs were carried forward for further analysis.
+\cite{Medland2009} Standard quality control (QC) filters were applied and the remaining 528,509 autosomal SNPs were carried forward for further analysis.
 
@@
 For each SNP pair there are 9 possible genotype classes.
-We treat each genotype class as a fixed effect and fit an 8 d.f.
-$F$-test to test the following hypotheses:
+We treat each genotype class as a fixed effect and fit an 8 \emph{d.f.} $F$-test to test the following hypotheses:
 
@@
 To ensure that interacting SNPs were driven by epistasis and not marginal effects we performed a nested ANOVA on each filtered pair to test if the interaction terms were significant.
-We did this by contrasting the full genetic model (8 d.f.) against the reduced marginal effects model which included the additive and dominance terms at both SNPs (4 d.f.).
-Thus, a 4 d.f.
-$F$-test was performed on the residual genetic variation, representing the contribution of epistatic variance.
+We did this by contrasting the full genetic model (8 \emph{d.f.}) against the reduced marginal effects model which included the additive and dominance terms at both SNPs (4 \emph{d.f.}).
+Thus, a 4 \emph{d.f.} $F$-test was performed on the residual genetic variation, representing the contribution of epistatic variance.
 Significance of epistasis was determined using a Bonferroni threshold of $0.05 / (6404+4751) = 4.48 \times 10^{-6}$.
@@
 	\includegraphics[width=5in]{qqbonf}
-	\caption{\textbf{Q-Q plots of interaction $p$-values in two independent datasets} The interaction $p$-values for the 480 SNP pairs that passed filtering in the two replication datasets.
-Green points represent $p$-values that surpass Bonferroni correction}
+	\caption{\textbf{Q-Q plots of interaction $p$-values in two independent datasets} Of the 549 discovery interactions, interaction $p$-values (4 \emph{d.f.} test) were obtained for 480 SNP pairs that passed filtering in the two replication datasets.
+The $p$-values plotted along the $y$-axis, and the expected $p$-values are plotted along the $x$-axis.
+Green points represent $p$-values that surpass Bonferroni correction.}
 	\label{fig:qqplotbonf}
@@
 	\includegraphics[width=5in]{qqfdr}
-	\caption{\textbf{Q-Q plots of replicated interaction $p$-values excluding the 20 most extreme $p$-values} The left panel (Null) shows the replication $p$-values from 529 randomly drawn SNP pairs in the Fehrmann (top row) and EGCUT (bottom row) datasets.
+	\caption{\textbf{Q-Q plots of replicated interaction $p$-values excluding the 20 most extreme $p$-values} The left panel (Null) shows the replication $p$-values from 549 randomly drawn SNP pairs in the Fehrmann (top row) and EGCUT (bottom row) datasets.
 The right panel (Observed) shows the interaction $p$-values from the 460 least significant pairs that pass filtering in the two replication datasets (\emph{i.e.} excluding the 20 most extreme $p$-values for clarity).

### 2013-05-31 docs/manuscript/nature_letter.tex
The manuscript is in a form that can be sent. Just waiting for Kostya's final analyses and graphs

@@
 \graphicspath{{../../analysis/images/}}
-\bibliographystyle{plain}
+\bibliographystyle{naturemag}
 
 \title{Epistasis is widespread in the genetic control of transcription in humans}
+\date{}
+\author{Gibran Hemani, Konstantin Shakhbazov, Joseph E Powell}
 
@@
 \begin{abstract}
-A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits \cite{Phillips1998, Carlborg2004, Hill2008a, Crow2010}.
+A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits \cite{Carlborg2004, Hill2008a, Crow2010}.
 Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
-Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Greene2009, Hemani2013}.
+Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}.
 Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design where any effects are expected to be large \cite{Powell2012}, evidence for multiple instances of epistasis is found.
 In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
-We attempted replication in two independent datasets \cite{Metspalu2004, Fehrmann2011} and 421 show evidence of significance in at least one dataset.
+We attempted replication in two independent datasets \cite{Metspalu2004, Fehrmann2011} and 421 showed evidence of significance in at least one dataset.
 We provide evidence of functional enrichment for the interacting SNPs, for instance 49 of the genetic interactions are located within 500kb of known chromosome interactions \cite{Lan2012} ($p < 1.1 \times 10^{-70}$).
@@
 For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-This study presents the first strong evidence for the widespread existence of epistatic genetic variation emerging from natural genetic variation in humans.
+This study presents the first strong evidence for the widespread existence of epistatic genetic effects emerging from natural genetic variation in humans.
 \end{abstract}
 
-\section{Main}
+\section{Main text}
 
-\subsection{Introduction}
+% \subsection{Introduction}
 The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits.
-Typically, SNPs are treated as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying thousands of causal variants \cite{Hindorff2010, Visscher2012}.
+Typically, SNPs are treated as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying thousands of causal variants \cite{Visscher2012}.
 Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale in artificial double gene knockouts in experimental organisms \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
-Though its importance is frequently the subject of debate \cite{Phillips1998, Carlborg2004, Hill2008a, Crow2010}, to date there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits in humans \cite{Strange2010, Evans2011}.
+Though its importance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}, to date there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits in humans \cite{Strange2010, Evans2011}.
 
-Detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2002}.
-When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Yang2011}, the power to detect epistasis diminishes rapidly.
+Detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009}.
+When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
 There are two ways to overcome this problem, either by using extremely large sample sizes \cite{LangoAllen2010}, or use traits that are likely to have large effect sizes.
@@
 
-\subsection{Initial search in discovery set}
+% \subsection{Initial search in discovery set}
 We searched for pairwise epistasis exhaustively in the Brisbane Systems Genetics Study (BSGS) dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes.
-Recent hardware \cite{Fan:2004:GCH:1048933.1049991} and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
+Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
 We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
 SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
-Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Evans2006, Hemani2013}.
+Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}.
 Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
@@
 
-\subsection{Replication}
+% \subsection{Replication}
 Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
@@
 The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}).
-A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication.
+A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication.
 
-\subsection{Patterns of epistasis}
-Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2002, Cordell2009}, but our analysis shows this is unlikely to be the most effective framework for its detection.
+% \subsection{Patterns of epistasis}
+Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows this is unlikely to be the most effective framework for its detection.
 The majority of interactions comprised of one SNP that had a previously known association and one that had no previous association in the dataset \cite{Powell2013} (476 out of 549).
@@
 We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
-For example, MBNL1 (involved in RNA modification and regulation of splicing) has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
+For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
 Each of these interactions have evidence for replication in at least one dataset and eight have evidence for replication in both datasets (Supplementary Figure \ref{fig:circleplots}).
@@
 
-\subsection{Functional annotation}
+% \subsection{Functional annotation}
 Of the 549 interactions, 66 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 470 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
@@
 However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:transh3k4me3}).
-This difference between \emph{cis} and \emph{trans} SNPs suggests that there is a range of molecular mechanisms by which epistasis might arise.
+This difference between \emph{cis} and \emph{trans} SNPs suggests that there is a range of molecular mechanisms by which epistasis might arise beyond tissue specific transcription.
 There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a} (Supplementary Figure \ref{fig:enhancers}).
 This enrichment is consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$).
-In particular, there was substantial enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs ($p = 1 \times 10^{-40}$).
-
-- Functional annotation shows that proteins are not being changed at the codon level
+In particular, there was substantial enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs, a known regulator of transcription in haematopoietic cells \cite{Tsai1994} ($p = 1 \times 10^{-40}$).
 
@@
 
-\subsection{Contribution of epistatic variance}
-Though we have discovered many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question.
+% \subsection{Contribution of epistatic variance}
+Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question.
 In this study we are able to identify 249 probes with at least one significant interaction given our experiment-wide threshold.
@@
 
-\subsection{Discussion}
+% \subsection{Discussion}
 Overall, we have demonstrated that it is possible to identify and replicate epistasis in complex traits amongst common human variants.
@@
 \clearpage
-\section{Figure legends}
+\section{Figures}
 
 \begin{figure}[H]
+	\includegraphics[width=5in]{hairballs_all_reps.pdf}
 	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A represents all 549 genetic interactions (edges) discovered in the discovery set.
@@
 \end{figure}
+\clearpage
 
 \begin{figure}[H]
+	\includegraphics[width=5in]{gpbonfrep.pdf}
 	\caption{\textbf{Replication of genotype-phenotype (GP) maps in two cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
@@
 \end{figure}
+\clearpage
 
 \begin{figure}[H]
+	\includegraphics[width=5in]{variance_components.pdf}
 	\caption{\textbf{Comparison of genetic variances} Panel A shows the proportion of phenotypic variation ($x$-axis) that is captured by additive, dominant and interaction components for all 529 significant interactions, ordered by total phenotypic variance explained by the SNP pair along the $y$-axis.
@@
 \end{figure}
-
-\clearpage
-\section{Figures}
-\setcounter{figure}{0}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=5in]{hairballs_all_reps.pdf}
-	\caption{}
-\end{figure}
 \clearpage
 
-\begin{figure}
-	\centering
-	\includegraphics[width=5in]{gpbonfrep.pdf}
-	\caption{}
-\end{figure}
-\clearpage
+% \clearpage
+% \section{Figures}
+% \setcounter{figure}{0}
 
-\begin{figure}
-	\centering
-	\includegraphics[width=5in]{variance_components.pdf}
-	\caption{}
-\end{figure}
-\clearpage
+% \begin{figure}[H]
+% 	\centering
+% 	\includegraphics[width=5in]{hairballs_all_reps.pdf}
+% 	\caption{}
+% \end{figure}
+% \clearpage
 
-\section{Full Methods}
+% \begin{figure}
+% 	\centering
+% 	\includegraphics[width=5in]{gpbonfrep.pdf}
+% 	\caption{}
+% \end{figure}
+% \clearpage
 
-\subsection{Discovery data}
-
-The Brisbane Systems Genetics Study (BSGS) comprises 846 individuals of European descent from 274 independent families \cite{pmid22563384}.
-DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland.
-Full details of genotyping procedures are given in Medland et al.
-\cite{Medland2009} Standard quality control (QC) filters were applied and the remaining 528,509 autosomal SNPs were carried forward for further analysis.
-
-Gene expression profiles were generated from whole blood collected with PAXgene TM tubes (QIAGEN, Valencia, CA) using Illumina HT12-v4.0 bead arrays.
-The Illumina HT-12 v4.0 chip contains 47,323 probes, although some probes are not assigned to RefSeq genes.
-We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF > 0.05 within 1000 genomes data; did not map to a listed ref-seq gene; were not significantly expressed (based on a detection $p$-value $< 0.05$) in at least 90\% of samples.
-After this stringent QC 7339 probes remained for 2D-eQTL mapping.
-
-\subsection{Normalisation}
-
-Gene expression profiles were normalised and adjusted for batch and polygenic effects.
-Profiles were first adjusted for raw background expression in each sample.
-Expression levels were then adjusted using quantile and log2 transformation to standardise distributions between samples.
-Batch and polygenic effects were adjusted using the linear model
-
-\begin{equation}
-y = \mu + \beta_{1}c + \beta_{2}p + \beta_{3}s + \beta_{4}a + g + e
-\label{eq:lm}
-\end{equation}
-where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively, fitted as fixed effects; and $g$ is a random additive polygenic effect with a variance covariance matrix
-\begin{equation}
-G_{ijk} = \left \{ 
-\begin{array}{ll}
-\sigma _a ^2 + \sigma _e ^2&        j = k \\ 
-2\phi _{ijk} \sigma _a ^2& 			j \neq k \\
-\end{array} \right.
-\end{equation}
-The parameters $\sigma_a^2$ and $\sigma_e^2$ are variance components for additive background genetic and environmental effects respectively.
-Here, we are using family based pedigree information rather than SNP based IDB to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$.
-The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero.
-Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan.
-
-\subsection{Exhaustive 2D-eQTL analysis}
-
-We used epiGPU software to perform an exhaustive scan for pairwise interactions, such that each SNP is tested against all other SNPs for statistical association with the expression values for each of the 7339 probes.
-For each SNP pair there are 9 possible genotype classes.
-We treat each genotype class as a fixed effect and fit an 8 \emph{d.f.} $F$-test to test the following hypotheses:
-
-\begin{equation}
-H _0 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 = 0; 
-\end{equation}
-
-\begin{equation}
-H _1 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 > 0; 
-\label{eq:8df}
-\end{equation}
-
-where $\mu$ is the mean expression level and $x _{ij}$ is the pairwise genotype class mean for genotype $i$ at SNP 1 and genotype $j$ at SNP 2.
-This type of test does not parameterize for specific types of epistasis, rather it tests for the joint genetic effects at two loci.
-This has been demonstrated to be statistically more efficient when searching for a wide range of epistatic patterns, although will include any marginal effects of SNPs.
-
-The complete exhaustive scan for 7339 probes comprises $1.03 \times 10^{15}$ $F$-tests.
-We used permutation analysis to estimate an appropriate significance threshold for the study.
-To do this we performed a further 1600 exhaustive 2D scans on permuted phenotypes to generate a null distribution of the extreme $p$-values expected to be obtained from this number of multiple tests given the correlation structure between the SNPs.
-We took the most extreme $p$-value from each of the 1600 scans and set the 5\% FWER to be the 95\% most extreme of these $p$-values, $T_{*} = 2.13 \times 10^{-12}$.
-The effective number of tests in one 2D scan being performed is therefore $N_{*} = 0.05 / T_{*} \approx 2.33 \times 10^{10}$.
-To correct for the testing of multiple traits we established an experiment wide threshold of $T = 0.05 / (N_{*} \times 7339) = 2.91 \times 10^{-16}$.
-This is likely to be conservative as it assumes independence between probes.
-
-Following filtering on this threshold only SNP pairs with at least 5 data points in all 9 genotype classes were kept.
-We then calculated the LD between interacting SNPs and removed any pairs with $r^2 > 0.1$.
-If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained.
-At this stage 6404 filtered SNP pairs remained.
-We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs.
-To ensure that interacting SNPs were driven by epistasis and not marginal effects we performed a nested ANOVA on each filtered pair to test if the interaction terms were significant.
-We did this by contrasting the full genetic model (8 \emph{d.f.}) against the reduced marginal effects model which included the additive and dominance terms at both SNPs (4 \emph{d.f.}).
-Thus, a 4 \emph{d.f.} $F$-test was performed on the residual genetic variation, representing the contribution of epistatic variance.
-Significance of epistasis was determined using a Bonferroni threshold of $0.05 / (6404+4751) = 4.48 \times 10^{-6}$.
-This resulted in 432 and 117 SNP pairs with significant interaction terms from filters 1 and 2, respectively.
-
-\subsection{Replication}
-We have attempted replication of the 529 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB. Details of the cohorts are as follows.
-
-\paragraph{Fehrmann: $n=1240$}
-The Fehrmann dataset consists of whole peripheral blood samples of 1240 unrelated individuals from the United Kingdom and the Netherlands.
-Some of these individuals are patients, while others are healthy controls.
-Individuals were genotyped using the Illumina 610 Quad platform.
-RNA levels were quantified using the HT12v3 platform.
-
-\paragraph{EGCUT: $n=891$}
-The Estonian Genome Center of the University of Tartu (EGCUT) study consists of whole blood samples of 891 unrelated individuals from Estonia.
-They were genotyped using the Illumina 610 Quad platform.
-RNA levels were quantified using the HT12v3 platform.
-
-\paragraph{CDHWB: $n=139$}
-The Center for Health Discovery and Well Being (CDHWB) Study is a population based cohort consisting of 139 individuals of European descent collected in Atlanta USA.
-Gene expression profiles were generated with Illumina HT-12 V3.0 arrays from whole blood collected from Tempus tubes that preserve RNA.
-Whole genome genotypes were measured using Illumina OmniQuad arrays.
-
-\clearpage
-\section{References}
-\bibliography{refs}
+% \begin{figure}
+% 	\centering
+% 	\includegraphics[width=5in]{variance_components.pdf}
+% 	\caption{}
+% \end{figure}
 
@@
 \end{figure}
+
 \clearpage
+\section{References}
+\bibliography{refs}
 

### 2013-06-06 docs/manuscript/nature_letter.tex
starting second draft of manuscript

@@
 Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design where any effects are expected to be large \cite{Powell2012}, evidence for multiple instances of epistasis is found.
-In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions.
-We attempted replication in two independent datasets \cite{Metspalu2004, Fehrmann2011} and 421 showed evidence of significance in at least one dataset.
-We provide evidence of functional enrichment for the interacting SNPs, for instance 49 of the genetic interactions are located within 500kb of known chromosome interactions \cite{Lan2012} ($p < 1.1 \times 10^{-70}$).
-One hundred and twenty-nine genes are controlled by multi-locus epistatic interactions whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
+In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 434 significant pairwise epistatic interactions acting on the expression levels of 165 genes.
+Twenty-nine interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing.
+Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects ($\lambda_{GC} = 1.35$, $p = 0.04$).
+We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of known regions of \emph{in vitro} chromosome interactions \cite{LiebermanAiden2009} ($p < 1.8 \times 10^{-10}$).
+Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
 For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-This study presents the first strong evidence for the widespread existence of epistatic genetic effects emerging from natural genetic variation in humans.
+This study presents the first strong evidence for the widespread existence of epistatic genetic effects emerging from natural genetic variation in humans, and makes inference on its contribution.
 \end{abstract}
@@
 The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits.
-Typically, SNPs are treated as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying thousands of causal variants \cite{Visscher2012}.
-Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale in artificial double gene knockouts in experimental organisms \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
-Though its importance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}, to date there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits in humans \cite{Strange2010, Evans2011}.
+Typically, SNPs are modelled as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying thousands of causal variants \cite{Visscher2012}.
+To date there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its importance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}.
+Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
+Here we demonstrate that instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.
 
@@
 When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
-There are two ways to overcome this problem, either by using extremely large sample sizes \cite{LangoAllen2010}, or use traits that are likely to have large effect sizes.
-Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we opted for the latter approach and searched for epistasis controlling gene expression levels.
-Transcription levels can be measured for thousands of genes and these traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis should it exist.
+There are two ways to overcome this problem, either by using extremely large sample sizes \cite{LangoAllen2010}, or by analysing traits that are likely to have large effect sizes.
+Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels.
+Transcription levels can be measured for thousands of genes.
+These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.
 
 % \subsection{Initial search in discovery set}
-We searched for pairwise epistasis exhaustively in the Brisbane Systems Genetics Study (BSGS) dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes.
-Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
-We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
-SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
-Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}.
-Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
-All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least five data points in all nine genotype classes.
-If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained.
-Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 549 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
+In a dataset of 846 individuals genotyped at 528509 SNPs, we tested every pair of SNPs for genetic interactions on each of 7339 expression traits in whole blood.
+After stringent filtering and multiple testing correction we had identified 501 putative genetic interactions influencing 238 gene expression levels.
+Of these, 434 passed filtering in two independent datasets of which 29 had interaction terms surpassing a Bonferroni corrected level of significance (Figure \ref{fig:gpmaps}, Table \ref{tab:bonferroni}).
 
@@
 Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows this is unlikely to be the most effective framework for its detection.
-The majority of interactions comprised of one SNP that had a previously known association and one that had no previous association in the dataset \cite{Powell2013} (476 out of 549).
+The majority of interactions comprised of one SNP that had a previously known association and one that had no previous association in the dataset \cite{Powell2013} (476 out of 434).
 Only 9 interactions were between SNPs that both had marginal effects while 64 were between SNPs that had no known marginal effects.
@@
 % \subsection{Functional annotation}
-Of the 549 interactions, 66 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 470 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
-In total the 549 interactions comprised 835 unique SNPs, which we analysed for functional enrichment.
+Of the 434 interactions, 66 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 470 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
+In total the 434 interactions comprised 835 unique SNPs, which we analysed for functional enrichment.
 We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013}.
@@
 
+\subsection{Methods Summary}
+We searched for pairwise epistasis exhaustively in the Brisbane Systems Genetics Study (BSGS) dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes.
+Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
+We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
+SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
+Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}.
+Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
+All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least five data points in all nine genotype classes.
+If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained.
+Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
+Of these, 434 passed filtering in both replication datasets.
+
 \clearpage
@@
 	\includegraphics[width=5in]{hairballs_all_reps.pdf}
-	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A represents all 549 genetic interactions (edges) discovered in the discovery set.
+	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A represents all 434 genetic interactions (edges) discovered in the discovery set.
 Panel B shows those interactions that were present at the 5\% FDR level in at least one replication dataset, and panel C shows which interactions were significant in both replication datasets.
@@
 	\includegraphics[width=5in]{qqbonf}
-	\caption{\textbf{Q-Q plots of interaction $p$-values in two independent datasets} Of the 549 discovery interactions, interaction $p$-values (4 \emph{d.f.} test) were obtained for 480 SNP pairs that passed filtering in the two replication datasets.
+	\caption{\textbf{Q-Q plots of interaction $p$-values in two independent datasets} Of the 434 discovery interactions, interaction $p$-values (4 \emph{d.f.} test) were obtained for 480 SNP pairs that passed filtering in the two replication datasets.
 The $p$-values plotted along the $y$-axis, and the expected $p$-values are plotted along the $x$-axis.
@@
 	\includegraphics[width=5in]{qqfdr}
-	\caption{\textbf{Q-Q plots of replicated interaction $p$-values excluding the 20 most extreme $p$-values} The left panel (Null) shows the replication $p$-values from 549 randomly drawn SNP pairs in the Fehrmann (top row) and EGCUT (bottom row) datasets.
+	\caption{\textbf{Q-Q plots of replicated interaction $p$-values excluding the 20 most extreme $p$-values} The left panel (Null) shows the replication $p$-values from 434 randomly drawn SNP pairs in the Fehrmann (top row) and EGCUT (bottom row) datasets.
 The right panel (Observed) shows the interaction $p$-values from the 460 least significant pairs that pass filtering in the two replication datasets (\emph{i.e.} excluding the 20 most extreme $p$-values for clarity).

### 2013-06-06 analysis/combine_data.R

@@ -219,6 +219,21 @@ for(i in 1:nrow(sig))
 }
 
 
+# Get sentinal SNPs and remove probes that tag the same genes
+sig$code2 <- with(sig, paste(chr1, chr2, probegene))
+sig <- sig[order(sig$code2, sig$pnest_egcut, decreasing=T), ]
+sig <- subset(sig, !duplicated(code2) | filter == 3)
+
+
+# Make sure that probechr, chr1 and chr2 are in 1:22
+sig <- subset(sig, ((probechr %in% 1:22 & chr1 %in% 1:22 & chr2 %in% 1:22) | filter == 3))
+
+
+# Make sure all SNPs are rsIDs
+sig <- subset(sig, snp1 %in% grep("rs", snp1, value=TRUE) & snp2 %in% grep("rs", snp2, value=TRUE))
+table(sig$filter != 3)
+
+
 # Remove filter==3
 sig_all <- sig
 sig <- subset(sig, filter != 3)
@@ -226,24 +241,4 @@ sig <- subset(sig, filter != 3)
 sig_rep1 <- subset(sig, pnest_fehr > upper_fehr | pnest_egcut > upper_egcut)
 sig_rep2 <- subset(sig, pnest_fehr > upper_fehr & pnest_egcut > upper_egcut)
 
-
 save(sig, sig_all, sig_rep1, sig_rep2, file="~/repo/eQTL-2D/analysis/interaction_list_replication_summary.RData")
-
-
-sig_all$code2 <- with(sig_all, paste(chr1, chr2, probegene))
-sig$code2 <- with(sig, paste(chr1, chr2, probegene))
-sig_rep1$code2 <- with(sig_rep1, paste(chr1, chr2, probegene))
-sig_rep2$code2 <- with(sig_rep2, paste(chr1, chr2, probegene))
-
-sig_all <- subset(sig_all, !duplicated(code2) | filter == 3)
-sig <- subset(sig, !duplicated(code2))
-sig_rep1 <- subset(sig_rep1, !duplicated(code2))
-sig_rep2 <- subset(sig_rep2, !duplicated(code2))
-
-dim(sig_all)
-dim(sig)
-dim(sig_rep1)
-dim(sig_rep2)
-
-save(sig, sig_all, sig_rep1, sig_rep2, file="~/repo/eQTL-2D/analysis/interaction_list_replication_summary_pruned.RData")
-

### 2013-06-06 analysis/functional_analysis.R

@@ -344,8 +344,6 @@ plot3dProbe(sig, "CTSC", xmat, resphen, z=135)
 #=================================================================================================#
 #=================================================================================================#
 
-
-
 # Distribution of additive vs non-additive variance
 
 a <- as.data.frame(do.call(rbind, sig$vc))
@@ -399,14 +397,10 @@ multiplot(p1, p2, cols=2)
 dev.off()
 
 
-
+#=================================================================================================#
+#=================================================================================================#
 
 # Chromosome interactions
-
-ci <- read.csv("~/repo/eQTL-2D/data/supFile3_K562_interactingLoci_clusters.csv", header=T)
-dim(ci)
-head(ci)
-table(ci$cluster)
 
 # For each of the significant interactions see if the two positions are within 1kb of any of the chromosome interactions
 ciOverlap <- function(ci, sig, win)
@@ -457,28 +451,24 @@ ciOverlap <- function(ci, sig, win)
 }
 
 
+load("~/repo/eQTL-2D/analysis/interaction_list_meta_analysis.RData")
+ci <- read.csv("~/repo/eQTL-2D/data/supFile3_K562_interactingLoci_clusters.csv", header=T)
+dim(ci)
+head(ci)
+table(ci$cluster)
+
+sig2 <- subset(meta, filter != 3)
 
 counts <- rep(0, 4)
-
-sig$code2 <- with(sig, paste(probegene, chr1, chr2))
-sig2 <- sig[order(sig$code2, sig$pnest_fehr, decreasing=T), ]
-sig2 <- subset(sig2, !duplicated(sig2$code2))
-dim(sig2)
-head(sig2)
-
-sig3 <- subset(sig2, chr1 != chr2)
-
 a1 <- ciOverlap(ci, sig2, 10000)
 counts[1] <- sum((!is.na(a1$int1) | !is.na(a1$int2)))
-
 a2 <- ciOverlap(ci, sig2, 250000)
 counts[2] <- sum((!is.na(a2$int1) | !is.na(a2$int2)))
-
 a3 <- ciOverlap(ci, sig2, 1000000)
 counts[3] <- sum((!is.na(a3$int1) | !is.na(a3$int2)))
-
-a4 <- ciOverlap(ci, sig3, 5000000)
+a4 <- ciOverlap(ci, sig2, 5000000)
[…]

### 2013-06-06 analysis/meta_analysis.R

@@ -224,7 +224,7 @@ createAverageGp <- function(gcm, gcs)
 
 # Read in data
 
-load("~/repo/eQTL-2D/analysis/interaction_list_replication_summary_pruned.RData")
+load("~/repo/eQTL-2D/analysis/interaction_list_replication_summary.RData")
 
 load("~/repo/eQTL-2D/replication/results/replication_GrngHT12v3.RData")
 newsig$code <- with(newsig, paste(probename, snp1, snp2))
@@ -243,18 +243,12 @@ names(egcut) <- newsig$code
 # Perform meta analysis on replication data
 
 sig_all <- performMeta(sig_all, egcut, fehr)
-sig <- performMeta(sig, egcut, fehr)
-sig_rep1 <- performMeta(sig_rep1, egcut, fehr)
-sig_rep2 <- performMeta(sig_rep2, egcut, fehr)
 
 thresh <- -log10(0.05 / 442)
 with(sig_all, table(filter, pnest_meta > thresh))
 with(sig_all, table(filter, pnest_meta2 > thresh))
 with(sig_all, table(filter, pnest_meta_vc > thresh))
 with(sig_all, table(filter, is.na(pnest_meta)))
-
-save(sig_all, sig, sig_rep1, sig_rep2, file="~/repo/eQTL-2D/analysis/interaction_list_replication_summary_pruned.RData")
-
 
 
 #=============================================================#
@@ -270,9 +264,9 @@ ggsave(file="~/repo/eQTL-2D/analysis/images/qqMetaAll.pdf", width=15, height=7.5
 qqPlot2(meta, thresh)
 ggsave(file="~/repo/eQTL-2D/analysis/images/qqMeta.pdf", width=7.5, height=15)
 
-with(meta, table(filter, pnest_meta > upper))
-with(meta, table(filter, pnest_fehr > upper_fehr))
-with(meta, table(filter, pnest_egcut > upper_egcut))
+with(meta, table(filter == 3, pnest_meta > upper))
+with(meta, table(filter == 3, pnest_fehr > upper_fehr))
+with(meta, table(filter == 3, pnest_egcut > upper_egcut))
 
 save(meta, file="~/repo/eQTL-2D/analysis/interaction_list_meta_analysis.RData")
 

### 2013-06-07 docs/manuscript/nature_letter.tex
LD check - as expected in BSGS therefore no evidence for population stratification

@@
 Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}.
-Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design where any effects are expected to be large \cite{Powell2012}, evidence for multiple instances of epistasis is found.
-In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 434 significant pairwise epistatic interactions acting on the expression levels of 165 genes.
+Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, evidence for multiple instances of epistasis is found.
+In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 501 significant pairwise epistatic interactions acting on the expression levels of 223 genes.
 Twenty-nine interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing.
-Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects ($\lambda_{GC} = 1.35$, $p = 0.04$).
-We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of known regions of \emph{in vitro} chromosome interactions \cite{LiebermanAiden2009} ($p < 1.8 \times 10^{-10}$).
+Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects (316 of 434 above the 2.5\% confidence interval for uniformly distributed $p$-values, $p < 1.0 \times 10^{-16}$).
+We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known intra-cellular chromosome interactions \cite{LiebermanAiden2009} ($p < 1.8 \times 10^{-10}$).
 Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
 For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-This study presents the first strong evidence for the widespread existence of epistatic genetic effects emerging from natural genetic variation in humans, and makes inference on its contribution.
+This study presents the first strong evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans, and demonstrates its significant contribution to the variance of gene expression.
 \end{abstract}
@@
 The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits.
-Typically, SNPs are modelled as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying thousands of causal variants \cite{Visscher2012}.
-To date there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its importance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}.
-Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
-Here we demonstrate that instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.
+Typically, SNPs are treated using an additive model, assumed to contribute linearly, independently, and cumulatively to the mean of a trait.
+This has been successful in identifying thousands of causal variants \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its importance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}.
+Outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
+Here we demonstrate that multiple instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.
 
@@
 Transcription levels can be measured for thousands of genes.
-These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.
+These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that many genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.
 
-% \subsection{Initial search in discovery set}
+% \subsection{Initial search in discovery set and replication}
 In a dataset of 846 individuals genotyped at 528509 SNPs, we tested every pair of SNPs for genetic interactions on each of 7339 expression traits in whole blood.
 After stringent filtering and multiple testing correction we had identified 501 putative genetic interactions influencing 238 gene expression levels.
-Of these, 434 passed filtering in two independent datasets of which 29 had interaction terms surpassing a Bonferroni corrected level of significance (Figure \ref{fig:gpmaps}, Table \ref{tab:bonferroni}).
+Of the 501 discovery interactions, data was available for 434 in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
+We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 29 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}).
+These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}), demonstrating that these complex genetic patterns are common even across independent populations.
+Additionally, there was extreme enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication.
+We observed that 316 of the remaining 405 discovery SNPs had replication interaction $p$-values exceeding the 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$).
+A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication.
+Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
+Therefore
 
-% \subsection{Replication}
-Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
-The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
-We observed substantial replication of interaction signals in both datasets.
-Of the 480 original pairs that passed filtering (Methods) in Fehrmann and EGCUT, 421 of the SNP pairs showed replication for the interaction test at the 5\% false discovery rate (FDR) level in at least one dataset, and 148 in both datasets (Supplementary Figure \ref{fig:qqplotfdr}).
-The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations.
 We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and 18 were significant in both.
 The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}).
-A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication.
 
@@
 Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
+The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
 Of these, 434 passed filtering in both replication datasets.

### 2013-06-07 analysis/check_ld.R

@@ -1,0 +1,33 @@
+load("clean_geno_final.RData")
+load("../analysis/interaction_list_meta_analysis.RData")
+
+fam$index <- 1:nrow(fam)
+fam_ind <- subset(fam, !duplicated(V1))
+dim(fam)
+dim(fam_ind)
+
+
+snps <- with(subset(meta, filter!=3 & !duplicated(probegene)), unique(c(snp1)))
+
+i <- fam_ind$index
+j <- match(snps, bim$V2)
+
+xmat[1:10,1:10]
+xmat_ind <- xmat[i, j]
+dim(xmat_ind)
+
+
+r <- cor(xmat_ind, use="pair")^2
+r[1:10,1:10]
+
+l <- r[lower.tri(r, diag=F)]
+length(l)
+mean(l)
+range(l)
+
+1/length(i)
+
+range(l)
+hist(l)
+
+

### 2013-06-07 docs/manuscript/nature_letter.tex
updated manuscript
need to provide data files for kostya
- new interaction SNPs (meta)
- list of chromosome interactions

@@
 The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits.
-Typically, SNPs are treated using an additive model, assumed to contribute linearly, independently, and cumulatively to the mean of a trait.
-This has been successful in identifying thousands of causal variants \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its importance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}.
-Outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
+Typically, SNPs are treated using an additive model where they are assumed to contribute linearly, independently, and cumulatively to the mean of a trait.
+This has been successful in identifying thousands of associations \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}.
+Outside the prism of human association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
 Here we demonstrate that multiple instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.
@@
 When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
-There are two ways to overcome this problem, either by using extremely large sample sizes \cite{LangoAllen2010}, or by analysing traits that are likely to have large effect sizes.
+There are two simple ways to overcome this problem.
+One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes.
 Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels.
@@
 % \subsection{Initial search in discovery set and replication}
-In a dataset of 846 individuals genotyped at 528509 SNPs, we tested every pair of SNPs for genetic interactions on each of 7339 expression traits in whole blood.
+In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood.
 After stringent filtering and multiple testing correction we had identified 501 putative genetic interactions influencing 238 gene expression levels.
-Of the 501 discovery interactions, data was available for 434 in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
+Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
 We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 29 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}).
-These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}), demonstrating that these complex genetic patterns are common even across independent populations.
-Additionally, there was extreme enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication.
-We observed that 316 of the remaining 405 discovery SNPs had replication interaction $p$-values exceeding the 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$).
+These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).
+
+Additionally, there was extreme enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication (Figure \ref{fig:qqMeta}).
+We observed that 316 of the remaining 405 discovery SNPs had replication interaction $p$-values exceeding the 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Supplementary Figure \ref{fig:qqMetaNonsig}).
+The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations.
 A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication.
 Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
-Therefore
-
-We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and 18 were significant in both.
-The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}).
+Therefore these results are very encouraging with regards to the detection and replication of epistasis.
 
 % \subsection{Patterns of epistasis}
-Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows this is unlikely to be the most effective framework for its detection.
-The majority of interactions comprised of one SNP that had a previously known association and one that had no previous association in the dataset \cite{Powell2013} (476 out of 434).
+Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows that this is unlikely to be the most effective strategy for its detection.
+The majority of our discovery interactions comprised of one SNP that had a previously known association and one SNP that had no previous association in the dataset \cite{Powell2013} (439 out of 501).
 Only 9 interactions were between SNPs that both had marginal effects while 64 were between SNPs that had no known marginal effects.
-Additionally, we observed that the largest epistatic variance component for the 529 interactions were divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by change ($p = 0.22$ for divergence from expectation).
+Additionally, we observed that the largest epistatic variance component for the 501 interactions were divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by change ($p = 0.22$ for divergence from expectation).
 This is perhaps to be expected because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.
@@
 % \subsection{Functional annotation}
-Of the 434 interactions, 66 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 470 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
-In total the 434 interactions comprised 835 unique SNPs, which we analysed for functional enrichment.
+Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
+In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment.
 We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013}.
@@
 However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:transh3k4me3}).
-This difference between \emph{cis} and \emph{trans} SNPs suggests that there is a range of molecular mechanisms by which epistasis might arise beyond tissue specific transcription.
+This difference between \emph{cis} and \emph{trans} SNPs suggests that there is a range of molecular mechanisms by which epistasis might arise and the cis-SNPs may provide tissue specificity in these interactions.
 There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a} (Supplementary Figure \ref{fig:enhancers}).
@@
 We also demonstrate another putative novel mechanism by which biological function can lead to epistatic genetic variance.
-We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012}.
-Forty-nine epistatic SNP pairs mapped to within 500kb of a chromosome interaction, and 69 mapped to within 2Mb ($p < 1.1 \times 10^{-70}$), (Supplementary Figure \ref{fig:chromosomeinteractions}), but decyphering the exact cellular processes underlying this interaction requires further research.
+It has been shown chromosomal regions  We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{LiebermanAiden2009, Lan2012}.
+Fifty epistatic interactions mapped to within 2Mb ($p < 1.1 \times 10^{-70}$), (Supplementary Figure \ref{fig:chromosomeinteractions}), but decyphering the exact cellular processes underlying this interaction requires further research.
 
@@
 \subsection{Methods Summary}
-We searched for pairwise epistasis exhaustively in the Brisbane Systems Genetics Study (BSGS) dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes.
+We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes.
 Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.

### 2013-06-10 docs/manuscript/methods.tex

\documentclass{article}
\usepackage[superscript,biblabel]{cite}
\bibliographystyle{naturemag}

\title{Online methods}
\date{}
\author{}

\begin{document}
\maketitle

\section{Discovery data}

The Brisbane Systems Genetics Study (BSGS) comprises 846 individuals of European descent from 274 independent families \cite{Powell2012}. DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland. Full details of genotyping procedures are given in Medland et al. \cite{Medland2009} Standard quality control (QC) filters were applied and the remaining 528,509 autosomal SNPs were carried forward for further analysis.

Gene expression profiles were generated from whole blood collected with PAXgene TM tubes (QIAGEN, Valencia, CA) using Illumina HT12-v4.0 bead arrays. The Illumina HT-12 v4.0 chip contains 47,323 probes, although some probes are not assigned to RefSeq genes. We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF $ > 0.05$ within 1000 genomes data; did not map to a listed RefSeq gene; were not significantly expressed (based on a detection $p$-value $< 0.05$) in at least 90\% of samples. After this stringent QC 7339 probes remained for 2D-eQTL mapping.

\section{Normalisation}

Gene expression profiles were normalised and adjusted for batch and polygenic effects. Profiles were first adjusted for raw background expression in each sample. Expression levels were then adjusted using quantile and log2 transformation to standardise distributions between samples. Batch and polygenic effects were adjusted using the linear model

\begin{equation}
y = \mu + \beta_{1}c + \beta_{2}p + \beta_{3}s + \beta_{4}a + g + e
\label{eq:lm}
\end{equation}
where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively, fitted as fixed effects; and $g$ is a random additive polygenic effect with a variance covariance matrix 
\begin{equation}
G_{ijk} = \left \{ 
\begin{array}{ll}
\sigma _a ^2 + \sigma _e ^2&        j = k \\ 
2\phi _{ijk} \sigma _a ^2& 			j \neq k \\
\end{array} \right.
\end{equation}
The parameters $\sigma_a^2$ and $\sigma_e^2$ are variance components for additive background genetic and environmental effects respectively. Here, we are using family based pedigree information rather than SNP based IDB to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$. The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero. Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan. The GenABEL package for R was used to perform the normalisation \cite{Aulchenko2007}.

\section{Exhaustive 2D-eQTL analysis}

We used epiGPU \cite{Hemani2011} software to perform an exhaustive scan for pairwise interactions, such that each SNP is tested against all other SNPs for statistical association with the expression values for each of the 7339 probes. For each SNP pair there are 9 possible genotype classes. We treat each genotype class as a fixed effect and fit an 8 \emph{d.f.} $F$-test to test the following hypotheses:

\begin{equation}
H _0 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 = 0; 
\end{equation}

\begin{equation}
H _1 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 > 0; 
\label{eq:8df}
\end{equation}

where $\mu$ is the mean expression level and $x _{ij}$ is the pairwise genotype class mean for genotype $i$ at SNP 1 and genotype $j$ at SNP 2. This type of test does not parameterize for specific types of epistasis, rather it tests for the joint genetic effects at two loci. This has been demonstrated to be statistically more efficient when searching for a wide range of epistatic patterns, although will also include any marginal effects of SNPs which must be dealt with post-hoc \cite{Hemani2013}.

The complete exhaustive scan for 7339 probes comprises $1.03 \times 10^{15}$ $F$-tests. We used permutation analysis to estimate an appropriate significance threshold for the study. To do this we performed a further 1600 exhaustive 2D scans on permuted phenotypes to generate a null distribution of the extreme $p$-values expected to be obtained from this number of multiple tests given the correlation structure between the SNPs. We took the most extreme $p$-value from each of the 1600 scans and set the 5\% FWER to be the 95\% most extreme of these $p$-values, $T_{*} = 2.13 \times 10^{-12}$. The effective number of tests in one 2D scan being performed is therefore $N_{*} = 0.05 / T_{*} \approx 2.33 \times 10^{10}$. To correct for the testing of multiple traits we established an experiment wide threshold of $T = 0.05 / (N_{*} \times 7339) = 2.91 \times 10^{-16}$. This is likely to be conservative as it assumes independence between probes.

Following filtering on this threshold only SNP pairs with at least 5 data points in all 9 genotype classes were kept. We then calculated the LD between interacting SNPs and removed any pairs with $r^2 > 0.1$. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained. At this stage 6404 filtered SNP pairs remained. We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs. To ensure that interacting SNPs were driven by epistasis and not marginal effects we performed a nested ANOVA on each filtered pair to test if the interaction terms were significant. We did this by contrasting the full genetic model (8 \emph{d.f.}) against the reduced marginal effects model which included the additive and dominance terms at both SNPs (4 \emph{d.f.}). Thus, a 4 \emph{d.f.} $F$-test was performed on the residual genetic variation, representing the contribution of epistatic variance. Significance of epistasis was determined using a Bonferroni threshold of $0.05 / (6404+4751) = 4.48 \times 10^{-6}$. This resulted in 432 and 117 SNP pairs with significant interaction terms from filters 1 and 2, respectively.

\section{Replication}
We have attempted replication of the 529 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB. It was required that LD $r^2 < 0.1$ between interacting SNPs, and all nine genotype classes had at least 5 individuals present in order to proceed with statistical testing for replication. Details of the cohorts are as follows.

\paragraph{Fehrmann: $n=1240$}
The Fehrmann dataset \cite{Fehrmann2011} consists of whole peripheral blood samples of 1240 unrelated individuals from the United Kingdom and the Netherlands. Some of these individuals are patients, while others are healthy controls. Individuals were genotyped using the Illumina 610 Quad platform. RNA levels were quantified using the HT12v3 platform.

\paragraph{EGCUT: $n=891$}
The Estonian Genome Center of the University of Tartu (EGCUT) study \cite{Metspalu2004} consists of whole blood samples of 891 unrelated individuals from Estonia. They were genotyped using the Illumina 610 Quad platform. RNA levels were quantified using the HT12v3 platform.

\paragraph{CDHWB: $n=139$}
The Center for Health Discovery and Well Being (CDHWB) Study \cite{Preininger2013} is a population based cohort consisting of 139 individuals of European descent collected in Atlanta USA. Gene expression profiles were generated with Illumina HT-12 V3.0 arrays from whole blood collected from Tempus tubes that preserve RNA. Whole genome genotypes were measured using Illumina OmniQuad arrays.

\section{References}
\bibliography{refs}

\end{document}

### 2013-06-10 docs/manuscript/nature_letter.tex

@@
 \usepackage{float}
+\usepackage{multirow}
 
@@
 Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, evidence for multiple instances of epistasis is found.
-In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 501 significant pairwise epistatic interactions acting on the expression levels of 223 genes.
-Twenty-nine interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing.
+In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 501 significant pairwise epistatic interactions acting on the expression levels of 238 genes.
+Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing.
 Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects (316 of 434 above the 2.5\% confidence interval for uniformly distributed $p$-values, $p < 1.0 \times 10^{-16}$).
-We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known intra-cellular chromosome interactions \cite{LiebermanAiden2009} ($p < 1.8 \times 10^{-10}$).
+We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p < 1.8 \times 10^{-10}$).
 Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
@@
 Outside the prism of human association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
-Here we demonstrate that multiple instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.
+Here we demonstrate that amongst common polymorphisms multiple instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.
 
@@
 Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
-We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 29 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}).
+We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}).
 These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).
@@
 Additionally, there was extreme enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication (Figure \ref{fig:qqMeta}).
-We observed that 316 of the remaining 405 discovery SNPs had replication interaction $p$-values exceeding the 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Supplementary Figure \ref{fig:qqMetaNonsig}).
+We observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Supplementary Figure \ref{fig:qqMetaNonsig}).
 The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations.
 A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication.
-Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
+It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
 Therefore these results are very encouraging with regards to the detection and replication of epistasis.
@@
 Only 9 interactions were between SNPs that both had marginal effects while 64 were between SNPs that had no known marginal effects.
-Additionally, we observed that the largest epistatic variance component for the 501 interactions were divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by change ($p = 0.22$ for divergence from expectation).
-This is perhaps to be expected because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.
+Additionally, we observed that the largest epistatic variance component for the 501 interactions were divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by chance ($p = 0.22$ for divergence from expectation).
+This is not surprising because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.
 
 We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
-For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 14 \emph{trans}-SNPs that exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
-Each of these interactions have evidence for replication in at least one dataset and eight have evidence for replication in both datasets (Supplementary Figure \ref{fig:circleplots}).
-We observed that nine of the 14 \emph{trans} SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$), suggesting a putative mechanism by which epistatic interactions may regulate transcription.
+For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
+We observed that nine of the 14 masking SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$), and each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
 
@@
 
-We also demonstrate another putative novel mechanism by which biological function can lead to epistatic genetic variance.
-It has been shown chromosomal regions  We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{LiebermanAiden2009, Lan2012}.
-Fifty epistatic interactions mapped to within 2Mb ($p < 1.1 \times 10^{-70}$), (Supplementary Figure \ref{fig:chromosomeinteractions}), but decyphering the exact cellular processes underlying this interaction requires further research.
+We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance.
+It has been shown that chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
+We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that fifty epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}), but decyphering the exact cellular processes underlying this interaction requires further research.
 
@@
 Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question.
-In this study we are able to identify 249 probes with at least one significant interaction given our experiment-wide threshold.
-Using the same threshold ($p < 2.91 \times 10^{-16}$) for the same data but searching for only additive effects returns 517 probes out of the 7339 analysed with at least one significant expression quantitative trait locus (eQTL) \cite{Powell2012}, therefore we argue that the number of instances of epistasis are substantial.
+In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold.
+How does this compare to the number of traits controlled by additive effects?
+The BSGS dataset has been previously analysed for additive effects at all expression traits \cite{Powell2012}, and if we take all the additive eQTLs that were significant at the epistatic threshold of $p < 2.91 \times 10^{-16}$ we find that 453 gene expression levels out of the 7339 analysed had at least one significant expression quantitative trait locus (eQTL).
+Therefore it can be argued that the number of instances of detectable epistasis are substantial.
+
 However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}.
 Ideally one would approach this question from a whole genome level \cite{Visscher2008} but this is intractable for non-additive variance components.
-Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance as has been argued previously \cite{Hill2008a, Crow2010}.
-We observe that of the significant interactions detected the average additive contribution to the total detected genetic variance is 72.7\%, over three times higher than the epistatic contribution of 24.1\% (Figure \ref{fig:variancecomponents}).
-
-From an experiment-wide perspective, the proportion of the total phenotypic variance for all 7339 probes explained by additive effects significant at the $p < 2.91 \times 10^{-16}$ threshold, as reported in an independent study on the same dataset that focused on additive effects only \cite{Powell2012}, is 1.9\%.
-In contrast, the epistatic signals detected in this study explains 0.27\%, approximately seven times lower than the additive variance.
-There are two caveats to this comparison, firstly the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
-Secondly the non-additive variance at causal variants is expected to be grossly underestimated by observed SNPs in comparison to estimates for additive variance \cite{Weir2008, Hemani2013}, due to differences in the rate of decay estiamted variance between observed and causal SNPs as LD decreases.
-Therefore this is likely to be a lower bound on the estimate of the relative contribution of epistasis.
+Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}.
+Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes.
+By contrast, the epistatic signals detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance.
+This is broadly in line with theory \cite{Hill2008a}, but there are two caveats to this comparison.
+Firstly the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
+Secondly the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.
+Therefore this is likely to be a lower bound on the estimate of the relative contribution of epistasis to the genetic variance of gene expression.
 
@@
 Overall, we have demonstrated that it is possible to identify and replicate epistasis in complex traits amongst common human variants.
-The functional analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation, thus it is not surprising that instances of epistasis should be widespread.
+The functional analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation, thus it is not surprising that multiple instances of epistasis should be discovered in spite of statistical power limitations.
 Further research into such genetic effects may provide a useful portal to understanding molecular mechanisms with greater clarity.
@@
 \clearpage
+\section{Tables}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rlllrr}
+  \hline
+ & & & & \multicolumn{2}{c}{$-\log_{10} p$} \\
+ & Gene (chr.) & SNP 1 (chr.) & SNP 2 (chr.) & Discovery & Replication \\
+  \hline
+1 & ADK (10)  & rs2395095 (10)  & rs10824092 (10)  & 6.69 & 39.82 \\
+  2 & ATP13A1 (19)  & rs4284750 (19)  & rs873870 (19)  & 5.30 & 14.23 \\
+  3 & C21ORF57 (21)  & rs9978658 (21)  & rs11701361 (21)  & 9.42 & 21.67 \\
+  4 & CSTB (21)  & rs9979356 (21)  & rs3761385 (21)  & 11.99 & 42.27 \\
+  5 & CTSC (11)  & rs7930237 (11)  & rs556895 (11)  & 7.16 & 33.53 \\
+  6 & FN3KRP (17)  & rs898095 (17)  & rs9892064 (17)  & 16.16 & 59.95 \\
+  7 & GAA (17)  & rs11150847 (17)  & rs12602462 (17)  & 13.91 & 32.60 \\
+  8 & HNRPH1 (5)  & rs6894268 (5)  & rs4700810 (5)  & 15.38 & 10.37 \\
+  9 & LAX1 (1)  & rs1891432 (1)  & rs10900520 (1)  & 19.16 & 29.24 \\
+  10 & MBNL1 (3)  & rs16864367 (3)  & rs13079208 (3)  & 13.49 & 41.56 \\
+  11 & MBNL1 (3)  & rs7710738 (5)  & rs13069559 (3)  & 7.92 & 9.28 \\
+  12 & MBNL1 (3)  & rs2030926 (6)  & rs13069559 (3)  & 7.10 & 5.53 \\
+  13 & MBNL1 (3)  & rs2614467 (14)  & rs13069559 (3)  & 5.74 & 5.30 \\
+  14 & MBNL1 (3)  & rs218671 (17)  & rs13069559 (3)  & 7.63 & 5.23 \\
+  15 & MBNL1 (3)  & rs11981513 (7)  & rs13069559 (3)  & 7.71 & 4.58 \\
+  16 & MBP (18)  & rs8092433 (18)  & rs4890876 (18)  & 5.40 & 28.73 \\
+  17 & NAPRT1 (8)  & rs2123758 (8)  & rs3889129 (8)  & 8.45 & 30.77 \\
+  18 & NCL (2)  & rs7563453 (2)  & rs4973397 (2)  & 7.31 & 12.70 \\
+  19 & PRMT2 (21)  & rs2839372 (21)  & rs11701058 (21)  & 4.81 & 4.06 \\
+  20 & RPL13 (16)  & rs352935 (16)  & rs2965817 (16)  & 4.98 & 17.24 \\
+  21 & SNORD14A (11)  & rs2634462 (11)  & rs6486334 (11)  & 7.31 & 23.22 \\
+  22 & TMEM149 (19)  & rs807491 (19)  & rs7254601 (19)  & 12.16 & 145.78 \\
+  23 & TMEM149 (19)  & rs8106959 (19)  & rs6926382 (6)  & 5.80 & 10.72 \\
+  24 & TMEM149 (19)  & rs8106959 (19)  & rs914940 (1)  & 6.22 & 9.20 \\
+  25 & TMEM149 (19)  & rs8106959 (19)  & rs2351458 (4)  & 7.30 & 8.00 \\
+  26 & TMEM149 (19)  & rs8106959 (19)  & rs6718480 (2)  & 8.55 & 7.36 \\
+  27 & TMEM149 (19)  & rs8106959 (19)  & rs1843357 (8)  & 6.21 & 6.00 \\
+  28 & TMEM149 (19)  & rs8106959 (19)  & rs9509428 (13)  & 9.44 & 4.47 \\
+  29 & TRA2A (7)  & rs7776572 (7)  & rs11770192 (7)  & 8.23 & 4.09 \\
+  30 & VASP (19)  & rs1264226 (19)  & rs2276470 (19)  & 5.09 & 4.95 \\
+   \hline
+\end{tabular}
+\caption{Epistatic interactions significant at the Bonferroni level in two replication sets.
+All $p$-values are for 4 \emph{d.f.} interaction tests (exact in Discovery and meta analysis for Replication).}
+\label{tab:bonferroni}
+\end{table}
+
+\clearpage
 \section{Figures}
@@
 \begin{figure}[H]
-	\includegraphics[width=5in]{hairballs_all_reps.pdf}
+	\centering
+	\includegraphics[width=5in]{graph_of_interactions.pdf}
 	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A represents all 434 genetic interactions (edges) discovered in the discovery set.
@@
 \begin{figure}[H]
-	\includegraphics[width=5in]{gpbonfrep.pdf}
-	\caption{\textbf{Replication of genotype-phenotype (GP) maps in two cohorts} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
-Each tile represents the expression level for that two-locus genotype class.
-Phenotypes are for gene transcript levels, where the gene symbol is labelled in the panel columns.
-Panel rows are labelled for the dataset.
-Phenotypes for each interaction within each cohort are scaled to be between 0 and 1.
+	\centering
+	\includegraphics[width=2.5in]{gpbonfrep.pdf}
+	\caption{\textbf{Replication of genotype-phenotype (GP) maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
+Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class.
+Phenotypes are for gene transcript levels (dark coloured tiles = low expression, light coloured tiles = high expression).
+Columns of GP maps are for each independent population.
+Rows of GP maps are for each of 30 significantly replicated interactions, corresponding to the rows in Table \ref{tab:bonferroni}.
 There is a general trend of the GP maps replicating across all three datasets.}
@@
 \begin{figure}[H]
-	\includegraphics[width=5in]{variance_components.pdf}
-	\caption{\textbf{Comparison of genetic variances} Panel A shows the proportion of phenotypic variation ($x$-axis) that is captured by additive, dominant and interaction components for all 529 significant interactions, ordered by total phenotypic variance explained by the SNP pair along the $y$-axis.
-Panel B shows the proportion of the genetic variance ($x$-axis) that is attributable to the three variance components, ordered by the proportion that is additive.}
-	\label{fig:variancecomponents}
+	\centering
+	\includegraphics[width=2.5in]{qqMeta.pdf}
+	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} Top panel shows all 434 discovery SNPs that were tested for interactions.
+Observed $p$-values ($y$-axis, $-\log_{10}$ scale) are plotted against the expected $p$-values ($x$-axis, $-\log_{10}$ scale).
+The multiple testing correction threshold for significance is denoted by a dotted line.
+Bottom panel shows the same data as the top panel but excluding the 30 interactions that were significant at the Bonferroni level in the replication datasets.
+The shaded grey area represents the 5\% confidence interval for the expected distribution of $p$-values.
+Dark blue points represent $p$-values that exceed the confidence interval, light blue are within the confidence interval.}
+	\label{fig:qqMeta}
 \end{figure}
 \clearpage
-
-% \clearpage
-% \section{Figures}
-% \setcounter{figure}{0}
-
-% \begin{figure}[H]
-% 	\centering
-% 	\includegraphics[width=5in]{hairballs_all_reps.pdf}
-% 	\caption{}
-% \end{figure}
-% \clearpage
-
-% \begin{figure}
-% 	\centering
-% 	\includegraphics[width=5in]{gpbonfrep.pdf}
-% 	\caption{}
-% \end{figure}
-% \clearpage
-
-% \begin{figure}
-% 	\centering
-% 	\includegraphics[width=5in]{variance_components.pdf}
-% 	\caption{}
-% \end{figure}
 
@@
 \begin{figure}[H]
-	\includegraphics[width=5in]{qqbonf}
-	\caption{\textbf{Q-Q plots of interaction $p$-values in two independent datasets} Of the 434 discovery interactions, interaction $p$-values (4 \emph{d.f.} test) were obtained for 480 SNP pairs that passed filtering in the two replication datasets.
-The $p$-values plotted along the $y$-axis, and the expected $p$-values are plotted along the $x$-axis.
-Green points represent $p$-values that surpass Bonferroni correction.}
-	\label{fig:qqplotbonf}
+	\centering
+	\includegraphics[width=5in]{qqMetaNonsig}
+	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets, excluding the 30 points significant at the Bonferroni level} The right panel (Null) shows the interaction $p$-values from a meta analysis across two independent datasets on 434 randomly drawn SNP pairs.
+The left panel (Empirical) shows the interaction $p$-values from the 404 putative interactions that were not significant at the Bonferroni correction threshold.
+Dark blue points represent $p$-values that surpass the 5\% FDR level, as in Figure \ref{fig:qqMeta}.}
+\label{fig:qqMetaNonsig}
 \end{figure}
@@
 \begin{figure}
-	\includegraphics[width=5in]{qqfdr}
-	\caption{\textbf{Q-Q plots of replicated interaction $p$-values excluding the 20 most extreme $p$-values} The left panel (Null) shows the replication $p$-values from 434 randomly drawn SNP pairs in the Fehrmann (top row) and EGCUT (bottom row) datasets.
-The right panel (Observed) shows the interaction $p$-values from the 460 least significant pairs that pass filtering in the two replication datasets (\emph{i.e.} excluding the 20 most extreme $p$-values for clarity).
-Dark blue points represent $p$-values that surpass the 5\% FDR level.}
-\label{fig:qqplotfdr}
-\end{figure}
-\clearpage
-
-\begin{figure}
+	\centering
 	\includegraphics[width=5in]{circles_replication2}
@@
 \begin{figure}
+	\centering
 	\includegraphics[width=5in]{cis_h3k4me3}
@@
 \begin{figure}
+	\centering
 	\includegraphics[width=5in]{trans_h3k4me3}
@@
 \begin{figure}
+	\centering
 	\includegraphics[width=5in]{MBNL1}
@@
 \begin{figure}
+	\centering
 	\includegraphics[width=5in]{chromosome_interactions}
@@
 \begin{figure}
+	\centering
 	\includegraphics[width=5in]{enhancers}

### 2013-06-10 docs/manuscript/methods.tex
completed draft of manuscript

@@
 We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs.
-To ensure that interacting SNPs were driven by epistasis and not marginal effects we performed a nested ANOVA on each filtered pair to test if the interaction terms were significant.
+
+To ensure that interacting SNPs were driven by epistasis and not marginal effects we performed a nested ANOVA on each pair in the filtered set to test if the interaction terms were significant.
 We did this by contrasting the full genetic model (8 \emph{d.f.}) against the reduced marginal effects model which included the additive and dominance terms at both SNPs (4 \emph{d.f.}).
@@
 
+All variance decompositions were calculated using the NOIA method \cite{Alvarez-Castro2008}.
+
 \section{Replication}
 We have attempted replication of the 529 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB.
-It was required that LD $r^2 < 0.1$ between interacting SNPs, and all nine genotype classes had at least 5 individuals present in order to proceed with statistical testing for replication.
+It was required that LD $r^2 < 0.1$ between interacting SNPs, and all nine genotype classes had at least 5 individuals present in order to proceed with statistical testing for replication in both datasets.
+We also excluded any putative SNPs that had discordant allele frequencies in any of the datasets.
 Details of the cohorts are as follows.
@@
 Whole genome genotypes were measured using Illumina OmniQuad arrays.
+Due to the small sample size, most SNP pairs did not pass filtering in this dataset and so we have excluded it from the rest of the analysis.
+
+\subsection{Meta Analysis}
+The 4 \emph{d.f.} interaction $p$-values for each independent replication dataset were calculated using the same statistical test as was performed in the discovery dataset.
+We then took the interaction $p$-values from EGCUT and Fehrmann and calculated a joint $p$-value as $-2 \log p_{1} - 2 \log p_{2} \sim \chi^{2}_{4 d.f.}$.
 

### 2013-06-10 analysis/meta_analysis.R

@@ -324,8 +324,12 @@ af_bsgs <- 1 - do.call(rbind, lapply(sig_all$gcs, alleleFreq))
 af_egcut <- do.call(rbind, lapply(sig_all$gcs_egcut, alleleFreq))
 af_fehr <- do.call(rbind, lapply(sig_all$gcs_fehr, alleleFreq))
 
+pdf("~/repo/eQTL-2D/analysis/images/alleleFreq.pdf")
 pairs(data.frame(BSGS = af_bsgs[,1], EGCUT = af_egcut[,1], Fehrmann = af_fehr[,1]))
-pairs(data.frame(BSGS = af_bsgs[,2], EGCUT = af_egcut[,2], Fehrmann = af_fehr[,2]))
+dev.off()
+
+
+pairs(data.frame(BSGS = c(af_bsgs), EGCUT = c(af_egcut), Fehrmann = c(af_fehr)))
 
 
 x_fehr <- cbind(abs(af_bsgs[,1] - af_fehr[,1]), abs(af_bsgs[,2] - af_fehr[,2]))

### 2013-06-10 docs/manuscript/nature_letter.tex
completed draft of manuscript
Added new data lists for kostya to perform functional annotations on
Updated number of chromosome interactions in manuscript to 44

@@
 It has been shown that chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
-We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that fifty epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}), but decyphering the exact cellular processes underlying this interaction requires further research.
+We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}), but decyphering the exact cellular processes underlying this interaction requires further research.
 
@@
 
+\begin{figure}
+	\centering
+	\includegraphics[width=5in]{alleleFreq}
+	\caption{\textbf{Comparison of allele frequencies for 781 SNPs involved in genetic interactions across independent populations} Outliers were removed from the analysis as part of the filtering stage during replication.}
+	\label{fig:allelefreq}
+\end{figure}
+
 \clearpage

### 2013-06-10 analysis/functional_analysis.R

@@ -505,6 +505,18 @@ ggplot() +
 ggsave(file="~/repo/eQTL-2D/analysis/images/chromosome_interactions.pdf")
 
 
+
+chr_interaction_list <- subset(a4, !is.na(int1) | !is.na(int2), select=c(probegene, probechr, snp1, snp2, chr1, chr2, position1, position2, pnest, pnest_meta))
+
+meta <- subset(meta, filter != 3)
+meta_replicated <- subset(meta, pnest_meta > upper)
+
+save(sig, chr_interaction_list, meta, meta_replicated, file="~/repo/eQTL-2D/analysis/data_lists.RData")
+
+
+
+
+
 getInteractionList <- function(a, ci)
 {
 	require(plyr)

### 2013-06-11 docs/manuscript/nature_letter.tex
changed title
awaiting final graphs from kostya

@@
 
-\title{Epistasis is widespread in the genetic control of transcription in humans}
+\title{Detection and replication of epistasis influencing transcription in humans}
 \date{}
@@
 Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects (316 of 434 above the 2.5\% confidence interval for uniformly distributed $p$-values, $p < 1.0 \times 10^{-16}$).
-We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p < 1.8 \times 10^{-10}$).
+We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
 Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
@@
 Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes.
-By contrast, the epistatic signals detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance.
-This is broadly in line with theory \cite{Hill2008a}, but there are two caveats to this comparison.
-Firstly the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
-Secondly the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.
-Therefore this is likely to be a lower bound on the estimate of the relative contribution of epistasis to the genetic variance of gene expression.
+By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance.
+This is broadly in line with theory \cite{Hill2008a}, but there are three caveats to this comparison.
+Firstly, the ratio of additive to epistatic variance may differ at different effect sizes.
+Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
+And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.
 
@@
 	\centering
-	\includegraphics[width=5in]{graph_of_interactions.pdf}
-	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A represents all 434 genetic interactions (edges) discovered in the discovery set.
-Panel B shows those interactions that were present at the 5\% FDR level in at least one replication dataset, and panel C shows which interactions were significant in both replication datasets.
-Red nodes represent probes and black nodes represent SNPs.
-The congruence between datasets is indicitive of the consistent epistatic instances in independent populations.}
+	\includegraphics[width=5in]{Grey_graph_of_interactions_2_lists.pdf}
+	\caption{\textbf{Replication of epistatic networks in two independent cohorts} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown.
+316 Interactions that had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data are shown with black nodes for SNPs and red nodes for gene expression traits.
+The interactions that fell within the confidence interval and showed no evidence of replication are greyed out.
+There is strong congruence of interactions between independent populations.}
 	\label{fig:fireworks}

### 2013-06-12 docs/manuscript/nature_letter.tex
submitted second draft of manuscript

@@
 Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}.
-Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, evidence for multiple instances of epistasis is found.
+Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, multiple instances of epistasis are found.
 In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 501 significant pairwise epistatic interactions acting on the expression levels of 238 genes.
@@
 Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects (316 of 434 above the 2.5\% confidence interval for uniformly distributed $p$-values, $p < 1.0 \times 10^{-16}$).
-We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
+We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
 Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
 For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-This study presents the first strong evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans, and demonstrates its significant contribution to the variance of gene expression.
+This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans that replicate in independent populations.
 \end{abstract}
@@
 In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood.
-After stringent filtering and multiple testing correction we had identified 501 putative genetic interactions influencing 238 gene expression levels.
+After stringent filtering and multiple testing correction (Methods) we had identified 501 putative genetic interactions influencing 238 gene expression levels.
 Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
@@
 Only 9 interactions were between SNPs that both had marginal effects while 64 were between SNPs that had no known marginal effects.
-Additionally, we observed that the largest epistatic variance component for the 501 interactions were divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by chance ($p = 0.22$ for divergence from expectation).
+Additionally, we observed that the largest epistatic variance component for the 501 interactions was divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by chance ($p = 0.22$ for divergence from expectation).
 This is not surprising because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.
@@
 For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
-We observed that nine of the 14 masking SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$), and each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
+We observed that nine of the 14 masking SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$).
+Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
 
@@
 We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013}.
-There was significant enrichment for \emph{cis}-acting SNPs in haematopoeitic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment are significant at after adjusting for multiple testing, Supplementary Figure \ref{fig:transh3k4me3}).
+There was significant enrichment for \emph{cis}-acting SNPs in haematopoeitic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing, Supplementary Figure \ref{fig:transh3k4me3}).
 However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:transh3k4me3}).
-This difference between \emph{cis} and \emph{trans} SNPs suggests that there is a range of molecular mechanisms by which epistasis might arise and the cis-SNPs may provide tissue specificity in these interactions.
+This difference between \emph{cis} and \emph{trans} SNPs suggests that there is a range of molecular mechanisms by which epistasis might arise and the \emph{cis}-SNPs may provide tissue specificity in these interactions.
 There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a} (Supplementary Figure \ref{fig:enhancers}).
@@
 We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance.
-It has been shown that chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
-We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}), but decyphering the exact cellular processes underlying this interaction requires further research.
+It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
+We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
 
@@
 Overall, we have demonstrated that it is possible to identify and replicate epistasis in complex traits amongst common human variants.
-The functional analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation, thus it is not surprising that multiple instances of epistasis should be discovered in spite of statistical power limitations.
+The functional analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation.
 Further research into such genetic effects may provide a useful portal to understanding molecular mechanisms with greater clarity.
@@
 Of these, 434 passed filtering in both replication datasets.
+A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.
 
@@
 The interactions that fell within the confidence interval and showed no evidence of replication are greyed out.
-There is strong congruence of interactions between independent populations.}
+There is good congruence of interactions between independent populations.}
 	\label{fig:fireworks}

### 2013-06-12 docs/manuscript/nature_letter.tex
added coverletter, sent copy of manuscript to big people

@@
 \date{}
-\author{Gibran Hemani, Konstantin Shakhbazov, Joseph E Powell}
+\author{Gibran Hemani, Joseph E Powell, Konstantin Shakhbazov, \\
+Greg Gibson, Grant W Montgomery, Peter M Visscher}
 
@@
 \begin{abstract}
-A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits \cite{Carlborg2004, Hill2008a, Crow2010}.
+A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010}.
 Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
-Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}.
-Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, multiple instances of epistasis are found.
-In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 501 significant pairwise epistatic interactions acting on the expression levels of 238 genes.
+Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}.
+Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found.
+In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between single nucleotide polymorphisms (SNPs) acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
 Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing.
-Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects (316 of 434 above the 2.5\% confidence interval for uniformly distributed $p$-values, $p < 1.0 \times 10^{-16}$).
-We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
-Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs.
-For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans that replicate in independent populations.
+Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects ($p < 1.0 \times 10^{-16}$).
+There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
+Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs.
+For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
+This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans.
+
 \end{abstract}
@@
 % \subsection{Introduction}
-The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits.
-Typically, SNPs are treated using an additive model where they are assumed to contribute linearly, independently, and cumulatively to the mean of a trait.
-This has been successful in identifying thousands of associations \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}.
-Outside the prism of human association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
-Here we demonstrate that amongst common polymorphisms multiple instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.
+In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently, and cumulatively to the mean of a trait.
+This framework has been successful in identifying thousands of associations \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}.
+Outside the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
 
-Detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009}.
+Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power.
+For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects.
 When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
@@
 We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
+Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.
 
@@
 The functional analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation.
-Further research into such genetic effects may provide a useful portal to understanding molecular mechanisms with greater clarity.
-With data and computational techniques now widely available the search for epistasis in larger datasets for traits of broader interest is warranted.
+Further research into such epistatic effects may provide a useful portal to understanding molecular mechanisms and complex trait variation with greater clarity.
+With computational techniques and data now widely available the search for epistasis in larger datasets for traits of broader interest is warranted.
 

### 2013-06-14 docs/manuscript/nature_letter.tex
work on manuscript

@@
 \usepackage{float}
-\usepackage{multirow}
+\usepackage{threeparttable}
 
@@
 \begin{abstract}
-A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010}.
+Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome.
+The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics.
 Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
@@
 In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between single nucleotide polymorphisms (SNPs) acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
-Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing.
+Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following stringent Bonferroni correction for multiple testing.
 Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects ($p < 1.0 \times 10^{-16}$).
@@
 	\centering
-\begin{tabular}{rlllrr}
+	\begin{threeparttable}
+		\caption{Epistatic interactions significant at the Bonferroni level in two replication sets}
+		\label{tab:bonferroni}
+		{\footnotesize
+		\begin{tabular}{rlllrrrr}
   \hline
- & & & & \multicolumn{2}{c}{$-\log_{10} p$} \\
- & Gene (chr.) & SNP 1 (chr.) & SNP 2 (chr.) & Discovery & Replication \\
+ & Gene (chr.) & SNP 1 (chr.) & SNP 2 (chr.) & BSGS\tnote{2} & Fehrmann\tnote{3} & EGCUT\tnote{3} & Meta\tnote{4} \\
   \hline
-1 & ADK (10)  & rs2395095 (10)  & rs10824092 (10)  & 6.69 & 39.82 \\
-  2 & ATP13A1 (19)  & rs4284750 (19)  & rs873870 (19)  & 5.30 & 14.23 \\
-  3 & C21ORF57 (21)  & rs9978658 (21)  & rs11701361 (21)  & 9.42 & 21.67 \\
-  4 & CSTB (21)  & rs9979356 (21)  & rs3761385 (21)  & 11.99 & 42.27 \\
-  5 & CTSC (11)  & rs7930237 (11)  & rs556895 (11)  & 7.16 & 33.53 \\
-  6 & FN3KRP (17)  & rs898095 (17)  & rs9892064 (17)  & 16.16 & 59.95 \\
-  7 & GAA (17)  & rs11150847 (17)  & rs12602462 (17)  & 13.91 & 32.60 \\
-  8 & HNRPH1 (5)  & rs6894268 (5)  & rs4700810 (5)  & 15.38 & 10.37 \\
-  9 & LAX1 (1)  & rs1891432 (1)  & rs10900520 (1)  & 19.16 & 29.24 \\
-  10 & MBNL1 (3)  & rs16864367 (3)  & rs13079208 (3)  & 13.49 & 41.56 \\
-  11 & MBNL1 (3)  & rs7710738 (5)  & rs13069559 (3)  & 7.92 & 9.28 \\
-  12 & MBNL1 (3)  & rs2030926 (6)  & rs13069559 (3)  & 7.10 & 5.53 \\
-  13 & MBNL1 (3)  & rs2614467 (14)  & rs13069559 (3)  & 5.74 & 5.30 \\
-  14 & MBNL1 (3)  & rs218671 (17)  & rs13069559 (3)  & 7.63 & 5.23 \\
-  15 & MBNL1 (3)  & rs11981513 (7)  & rs13069559 (3)  & 7.71 & 4.58 \\
-  16 & MBP (18)  & rs8092433 (18)  & rs4890876 (18)  & 5.40 & 28.73 \\
-  17 & NAPRT1 (8)  & rs2123758 (8)  & rs3889129 (8)  & 8.45 & 30.77 \\
-  18 & NCL (2)  & rs7563453 (2)  & rs4973397 (2)  & 7.31 & 12.70 \\
-  19 & PRMT2 (21)  & rs2839372 (21)  & rs11701058 (21)  & 4.81 & 4.06 \\
-  20 & RPL13 (16)  & rs352935 (16)  & rs2965817 (16)  & 4.98 & 17.24 \\
-  21 & SNORD14A (11)  & rs2634462 (11)  & rs6486334 (11)  & 7.31 & 23.22 \\
-  22 & TMEM149 (19)  & rs807491 (19)  & rs7254601 (19)  & 12.16 & 145.78 \\
-  23 & TMEM149 (19)  & rs8106959 (19)  & rs6926382 (6)  & 5.80 & 10.72 \\
-  24 & TMEM149 (19)  & rs8106959 (19)  & rs914940 (1)  & 6.22 & 9.20 \\
-  25 & TMEM149 (19)  & rs8106959 (19)  & rs2351458 (4)  & 7.30 & 8.00 \\
-  26 & TMEM149 (19)  & rs8106959 (19)  & rs6718480 (2)  & 8.55 & 7.36 \\
-  27 & TMEM149 (19)  & rs8106959 (19)  & rs1843357 (8)  & 6.21 & 6.00 \\
-  28 & TMEM149 (19)  & rs8106959 (19)  & rs9509428 (13)  & 9.44 & 4.47 \\
-  29 & TRA2A (7)  & rs7776572 (7)  & rs11770192 (7)  & 8.23 & 4.09 \\
-  30 & VASP (19)  & rs1264226 (19)  & rs2276470 (19)  & 5.09 & 4.95 \\
+1 & ADK (10)  & rs2395095 (10)  & rs10824092 (10)  & 6.69\tnote{1} & 18.33\tnote{1} & 21.21\tnote{1} & 39.82\tnote{1} \\
+  2 & ATP13A1 (19)  & rs4284750 (19)  & rs873870 (19)  & 5.30 & 12.18 & 3.25 & 14.23 \\
+  3 & C21ORF57 (21)  & rs9978658 (21)  & rs11701361 (21)  & 9.42 & 6.08 & 16.36 & 21.67 \\
+  4 & CSTB (21)  & rs9979356 (21)  & rs3761385 (21)  & 11.99 & 25.20 & 16.72 & 42.27 \\
+  5 & CTSC (11)  & rs7930237 (11)  & rs556895 (11)  & 7.16 & 18.76 & 15.06 & 33.53 \\
+  6 & FN3KRP (17)  & rs898095 (17)  & rs9892064 (17)  & 16.16 & 28.24 & 29.39 & 59.95 \\
+  7 & GAA (17)  & rs11150847 (17)  & rs12602462 (17)  & 13.91 & 19.98 & 12.99 & 32.60 \\
+  8 & HNRPH1 (5)  & rs6894268 (5)  & rs4700810 (5)  & 15.38 & 8.55 & 3.01 & 10.37 \\
+  9 & LAX1 (1)  & rs1891432 (1)  & rs10900520 (1)  & 19.16 & 18.60 & 11.22 & 29.24 \\
+  10 & MBNL1 (3)  & rs16864367 (3)  & rs13079208 (3)  & 13.49 & 16.25 & 24.74 & 41.56 \\
+  11 & MBNL1 (3)  & rs7710738 (5)  & rs13069559 (3)  & 7.92 & 2.55 & 7.89 & 9.28 \\
+  12 & MBNL1 (3)  & rs2030926 (6)  & rs13069559 (3)  & 7.10 & 0.91 & 5.80 & 5.53 \\
+  13 & MBNL1 (3)  & rs2614467 (14)  & rs13069559 (3)  & 5.74 & 4.13 & 2.22 & 5.30 \\
+  14 & MBNL1 (3)  & rs218671 (17)  & rs13069559 (3)  & 7.63 & 0.62 & 5.82 & 5.23 \\
+  15 & MBNL1 (3)  & rs11981513 (7)  & rs13069559 (3)  & 7.71 & 0.43 & 5.36 & 4.58 \\
+  16 & MBP (18)  & rs8092433 (18)  & rs4890876 (18)  & 5.40 & 7.06 & 21.91 & 28.73 \\
+  17 & NAPRT1 (8)  & rs2123758 (8)  & rs3889129 (8)  & 8.45 & 15.12 & 16.08 & 30.77 \\
+  18 & NCL (2)  & rs7563453 (2)  & rs4973397 (2)  & 7.31 & 7.51 & 6.33 & 12.70 \\
+  19 & PRMT2 (21)  & rs2839372 (21)  & rs11701058 (21)  & 4.81 & 0.69 & 4.47 & 4.06 \\
+  20 & RPL13 (16)  & rs352935 (16)  & rs2965817 (16)  & 4.98 & 3.79 & 14.41 & 17.24 \\
+  21 & SNORD14A (11)  & rs2634462 (11)  & rs6486334 (11)  & 7.31 & 13.11 & 10.96 & 23.22 \\
+  22 & TMEM149 (19)  & rs807491 (19)  & rs7254601 (19)  & 12.16 & 81.55 & 45.78 & 145.78 \\
+  23 & TMEM149 (19)  & rs8106959 (19)  & rs6926382 (6)  & 5.80 & 3.06 & 8.80 & 10.72 \\
+  24 & TMEM149 (19)  & rs8106959 (19)  & rs914940 (1)  & 6.22 & 3.36 & 6.96 & 9.20 \\
+  25 & TMEM149 (19)  & rs8106959 (19)  & rs2351458 (4)  & 7.30 & 0.04 & 9.61 & 8.00 \\
+  26 & TMEM149 (19)  & rs8106959 (19)  & rs6718480 (2)  & 8.55 & 3.31 & 5.15 & 7.36 \\
+  27 & TMEM149 (19)  & rs8106959 (19)  & rs1843357 (8)  & 6.21 & 3.72 & 3.33 & 6.00 \\
+  28 & TMEM149 (19)  & rs8106959 (19)  & rs9509428 (13)  & 9.44 & 0.10 & 5.75 & 4.47 \\
+  29 & TRA2A (7)  & rs7776572 (7)  & rs11770192 (7)  & 8.23 & 3.19 & 1.89 & 4.09 \\
+  30 & VASP (19)  & rs1264226 (19)  & rs2276470 (19)  & 5.09 & 0.94 & 5.14 & 4.95 \\
    \hline
 		\end{tabular}
-\caption{Epistatic interactions significant at the Bonferroni level in two replication sets.
-All $p$-values are for 4 \emph{d.f.} interaction tests (exact in Discovery and meta analysis for Replication).}
-\label{tab:bonferroni}
+		}
+		\begin{tablenotes}
+			\item [1] $-\log_{10} p$-values for 4 \emph{d.f.} interaction tests
+			\item [2] Discovery dataset
+			\item [3] Independent replication dataset
+			\item [4] Meta analysis of interaction terms between replication datasets only
+		\end{tablenotes}
+	\end{threeparttable}
 \end{table}

### 2013-06-14 analysis/gpmaps.R

@@ -182,7 +182,7 @@ bsig <- subset(sig, pnest_meta > -log10(0.05/434))
 bsig <- bsig[order(bsig$probegene), ]
 dim(bsig)
 
-tab <- subset(bsig, select=c(probegene, probechr, snp1, chr1, snp2, chr2, pnest, pnest_meta))
+tab <- subset(bsig, select=c(probegene, probechr, snp1, chr1, snp2, chr2, pnest, pnest_fehr, pnest_egcut, pnest_meta))
 rownames(tab) <- 1:nrow(tab)
 
 tab$snp1 <- paste(tab$snp1, " (", tab$chr1, ") ", sep="")
@@ -191,9 +191,8 @@ tab$probegene <- paste(tab$probegene, " (", tab$probechr, ") ", sep="")
 tab <- subset(tab, select=-c(chr1, chr2, probechr))
 tab
 
-names(tab) <- c("Gene (chromosome)", "SNP 1 (chromosome)", "SNP 2 (chromosome)", "Discovery", "Replication")
-
-xtable(tab, digits = c(0, 0, 0, 0, 2, 2))
+names(tab) <- c("Gene (chr.)", "SNP 1 (chr.)", "SNP 2 (chr.)", "Discovery", "Fehrmann", "EGCUT", "Combined replication")
+xtable(tab, digits = c(0, 0, 0, 0, 2, 2, 2, 2))
 
 
 

### 2013-06-18 docs/manuscript/nature_letter.tex
modifying text

@@
 Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}.
-Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found.
-In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between single nucleotide polymorphisms (SNPs) acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
+Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
+In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
 Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following stringent Bonferroni correction for multiple testing.

### 2013-06-18 docs/manuscript/nature_letter.tex
manuscript updated

@@
 One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes.
-Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels.
+Because our focus was to ascertain the extent to which instances of epistasis occur amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels.
 Transcription levels can be measured for thousands of genes.
@@
 In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood.
-After stringent filtering and multiple testing correction (Methods) we had identified 501 putative genetic interactions influencing 238 gene expression levels.
+After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing 238 gene expression levels.
 Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
@@
 
-Additionally, there was extreme enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication (Figure \ref{fig:qqMeta}).
+Additionally, there was enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication (Figure \ref{fig:qqMeta}).
 We observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Supplementary Figure \ref{fig:qqMetaNonsig}).
@@
 
+Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
 We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
 For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
-We observed that nine of the 14 masking SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$).
+We observed that nine of the 14 masking SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$, Supplementary Figure \ref{fig:snplocations}).
 Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
@@
 % \subsection{Functional annotation}
-Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
-In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment.
-We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013}.
-There was significant enrichment for \emph{cis}-acting SNPs in haematopoeitic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing, Supplementary Figure \ref{fig:transh3k4me3}).
-However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:transh3k4me3}).
-This difference between \emph{cis} and \emph{trans} SNPs suggests that there is a range of molecular mechanisms by which epistasis might arise and the \emph{cis}-SNPs may provide tissue specificity in these interactions.
+In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment (Methods).
+We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}).
+There was significant enrichment for \emph{cis}-acting SNPs in haematopoeitic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing).
+However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues).
+This difference between \emph{cis} and \emph{trans} SNPs suggests a range of molecular mechanisms by which epistasis might arise and the \emph{cis}-SNPs may provide tissue specificity in these interactions.
 There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a} (Supplementary Figure \ref{fig:enhancers}).
@@
 It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
-We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
+We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
 Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.
@@
 In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold.
-How does this compare to the number of traits controlled by additive effects?
+How does this compare to the number of traits influenced by additive effects?
 The BSGS dataset has been previously analysed for additive effects at all expression traits \cite{Powell2012}, and if we take all the additive eQTLs that were significant at the epistatic threshold of $p < 2.91 \times 10^{-16}$ we find that 453 gene expression levels out of the 7339 analysed had at least one significant expression quantitative trait locus (eQTL).
@@
 However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}.
-Ideally one would approach this question from a whole genome level \cite{Visscher2008} but this is intractable for non-additive variance components.
+Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components.
 Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}.
@@
 By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance.
-This is broadly in line with theory \cite{Hill2008a}, but there are three caveats to this comparison.
+It must be noted that there are at least three caveats to this comparison.
 Firstly, the ratio of additive to epistatic variance may differ at different effect sizes.
@@
 	\includegraphics[width=5in]{Grey_graph_of_interactions_2_lists.pdf}
-	\caption{\textbf{Replication of epistatic networks in two independent cohorts} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown.
-316 Interactions that had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data are shown with black nodes for SNPs and red nodes for gene expression traits.
-The interactions that fell within the confidence interval and showed no evidence of replication are greyed out.
-There is good congruence of interactions between independent populations.}
+	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).
+316 Interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data, but the remaining 118 interactions that did not replicate are depicted in grey.
+It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
 	\label{fig:fireworks}
@@
 	\includegraphics[width=2.5in]{gpbonfrep.pdf}
-	\caption{\textbf{Replication of genotype-phenotype (GP) maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
+	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
 Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class.
@@
 	\centering
-	\includegraphics[width=5in]{cis_h3k4me3}
-	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3.
-Here we show that there is }
-	\label{fig:cish3k4me3}
+	\includegraphics[width=5in]{OverlapAnnotation}
+	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 5kb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic regions were covered by the 501 significant interactions.
+There is enrichment for \emph{cis}-acting SNPs in promotor regions, and \emph{cis}- and \emph{trans}-acting SNPs in intronic regions.}
+	\label{fig:snplocations}
 \end{figure}
+\clearpage
 
@@
 	\centering
-	\includegraphics[width=5in]{trans_h3k4me3}
-	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3.
-Here we show that there is }
-	\label{fig:transh3k4me3}
+	\includegraphics[width=5in]{501_pairs_snp_list_Grid}
+	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3 \cite{Trynka2013}.
+Enrichment $p$-values are calculated using permutation analysis for 34 different cell types ($y$-axis) in four tissues (Rows of boxes).
+There is enrichment for \emph{cis}-acting SNPs in Haematopoeitic tissue types only.
+\emph{Trans}-acting SNPs have no tissue specificity.}
+	\label{fig:h3k4me3}
 \end{figure}

### 2013-06-18 docs/manuscript/nature_letter.tex
gib and joseph final changes to ms

@@
 Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
-Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}.
+Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}.
 Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
 In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
-Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following stringent Bonferroni correction for multiple testing.
-Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects ($p < 1.0 \times 10^{-16}$).
+Three hundred and forty-five interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} at a false dicovery rate (FDR) of 2.5\%, with 30 significant at a conservative $p < 0.05$ Bonferroni level.
 There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
@@
 % \subsection{Introduction}
-In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently, and cumulatively to the mean of a trait.
+In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently and cumulatively to the mean of a trait.
 This framework has been successful in identifying thousands of associations \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}.
@@
 % \subsection{Initial search in discovery set and replication}
-In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood.
-After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing 238 gene expression levels.
+In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood.
+After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes.
 Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
@@
 
-Additionally, there was enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication (Figure \ref{fig:qqMeta}).
-We observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Supplementary Figure \ref{fig:qqMetaNonsig}).
+In addition, we observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}).
 The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations.
-A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication.
+A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication.
 It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
-Therefore these results are very encouraging with regards to the detection and replication of epistasis.
+Therefore these results are encouraging with regards to the detection and replication of epistasis.
 
@@
 The majority of our discovery interactions comprised of one SNP that had a previously known association and one SNP that had no previous association in the dataset \cite{Powell2013} (439 out of 501).
-Only 9 interactions were between SNPs that both had marginal effects while 64 were between SNPs that had no known marginal effects.
+Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects.
 Additionally, we observed that the largest epistatic variance component for the 501 interactions was divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by chance ($p = 0.22$ for divergence from expectation).
@@
 We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}).
-There was significant enrichment for \emph{cis}-acting SNPs in haematopoeitic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing).
+There was significant enrichment for \emph{cis}-acting SNPs in haematopoietic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing).
 However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues).
 This difference between \emph{cis} and \emph{trans} SNPs suggests a range of molecular mechanisms by which epistasis might arise and the \emph{cis}-SNPs may provide tissue specificity in these interactions.
-There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a} (Supplementary Figure \ref{fig:enhancers}).
-This enrichment is consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$).
-In particular, there was substantial enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs, a known regulator of transcription in haematopoietic cells \cite{Tsai1994} ($p = 1 \times 10^{-40}$).
+There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a}, consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$).
+In particular, there was enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs, a known regulator of transcription in haematopoietic cells \cite{Tsai1994} ($p = 1 \times 10^{-40}$).
 
@@
 It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
-We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
+We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
 Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.
@@
 Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes.
-By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance.
+By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance (Methods).
 It must be noted that there are at least three caveats to this comparison.
@@
 \subsection{Methods Summary}
-We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes.
+We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7,339 probes representing 6,158 RefSeq genes.
 Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
@@
 	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).
-316 Interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data, but the remaining 118 interactions that did not replicate are depicted in grey.
+Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data, but the remaining 89 interactions that did not replicate are depicted in grey.
 It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
@@
 Columns of GP maps are for each independent population.
-Rows of GP maps are for each of 30 significantly replicated interactions, corresponding to the rows in Table \ref{tab:bonferroni}.
-There is a general trend of the GP maps replicating across all three datasets.}
+Rows of GP maps are for each of 30 significantly replicated interactions at the Bonferroni level, corresponding to the rows in Table \ref{tab:bonferroni}.
+There is a clear trend of the GP maps replicating across all three datasets.}
 	\label{fig:gpmaps}
@@
 	\includegraphics[width=2.5in]{qqMeta.pdf}
-	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} Top panel shows all 434 discovery SNPs that were tested for interactions.
+	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} The top panel shows all 434 discovery SNPs that were tested for interactions.
 Observed $p$-values ($y$-axis, $-\log_{10}$ scale) are plotted against the expected $p$-values ($x$-axis, $-\log_{10}$ scale).
-The multiple testing correction threshold for significance is denoted by a dotted line.
-Bottom panel shows the same data as the top panel but excluding the 30 interactions that were significant at the Bonferroni level in the replication datasets.
+The multiple testing correction threshold for significance following Bonferroni correction is denoted by a dotted line.
+The bottom panel shows the same data as the top panel but excluding the 30 interactions that were significant at the Bonferroni level in the replication datasets.
 The shaded grey area represents the 5\% confidence interval for the expected distribution of $p$-values.
@@
 The left panel (Empirical) shows the interaction $p$-values from the 404 putative interactions that were not significant at the Bonferroni correction threshold.
-Dark blue points represent $p$-values that surpass the 5\% FDR level, as in Figure \ref{fig:qqMeta}.}
+Dark blue points represent $p$-values that surpass the 2.5\% FDR level, as in Figure \ref{fig:qqMeta}.}
 \label{fig:qqMetaNonsig}
@@
 	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3 \cite{Trynka2013}.
-Enrichment $p$-values are calculated using permutation analysis for 34 different cell types ($y$-axis) in four tissues (Rows of boxes).
-There is enrichment for \emph{cis}-acting SNPs in Haematopoeitic tissue types only.
+Enrichment $p$-values are calculated using permutation analysis for 34 different cell types ($y$-axis) in four tissue types (Rows of boxes).
+There is enrichment for \emph{cis}-acting SNPs in Haematopoietic tissue types only.
 \emph{Trans}-acting SNPs have no tissue specificity.}
@@
 The number of epistatic interactions within 20kb, 500kb, 2Mb and 10Mb of known chromosome interacting regions are shown by red vertical lines.
-The histograms represent the null distribution based on random sampling of 10000 datasets for each window size.}
+The histograms represent the null distribution based on random sampling of 10,000 datasets for each window size.}
 	\label{fig:chromosomeinteractions}
@@
 \clearpage
-
-\begin{figure}
-	\centering
-	\includegraphics[width=5in]{enhancers}
-	\caption{\textbf{There is enrichment for enhancer sequences for \emph{cis} and \emph{trans} SNPs}}
-	\label{fig:enhancers}
-\end{figure}
 

### 2013-06-18 docs/manuscript/methods.tex
work on manuscript
manuscript updated
working on methods

@@
 
+\section{Additive vs epistatic variance comparison}
+
+To compare the relative contribution to the phenotypic variance of gene expression levels between additive and epistatic effects we are constrained by the problem that non-additive variance components for a phenotype cannot be calculated directly.
+Here, we only have SNP pairs that exceed a threshold of $p < 2.91 \times 10^{-16} = T_{e}$.
+A strong conclusion cannot be made about the genome-wide variance contribution, but we can compare the variance explained by SNP effects at this threshold for additive scans and epistatic scans.
+
+In Powell \emph{et al} 2012 \cite{Powell2012} an expression quantitative trait locus (eQTL) study was performed searching for additive effects in the same BSGS dataset as was used for the discovery here.
+Using the threshold $T_{e}$ for the additive eQTL study, 453 of the 7339 probes analysed here had at least one significant additive effect.
+Assuming that the phenotypic variance for each of the probes is normalised to 1, the total phenotypic variance of all 7339 explained by the significant additive effects was 1.73\%.
+
+Following the same procedure, at the threshold $T_{e}$ there were 238 gene expression probes with at least one significant pairwise epistatic interaction out of the 7339 tested.
+In total the proportion of the phenotypic variance explained by the epistatic effects at these SNP pairs was 0.25\%.
+
+\section{Functional enrichment analysis}
+
+\subsection{Tissue specific transcriptionally active regions}
+
+We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions.
+We subsequently refer to these 731 SNPs as the IndexSNPs A recent paper shows that H3K4me3 chromatin marks are the most cell-type-specific marks in terms of co-localization with previously published GWAS loci \cite{Trynka2013}.
+The idea behind the method is that variants related to a particular phenotype may affect cell-type-specific gene expression by changing regulatory elements in cell types relevant to that phenotype.
+Hence, overlap between associated variants and chromatin marks should occur preferentially in the relevant cell type(s).
+Our analysis tested for enrichment of these chromatin marks in 34 different cell types from 4 different broad tissue types.
+
+For each IndexSNP, a locus region was defined, bounded on either side of the IndexSNP by the most distant SNP within 250 kb that was in LD (r2 > 0.8).
+For each SNP in a locus region regulatory activity scores were calculated as the height of the nearest H3K4me3 mark divided by distance from the SNP to the H3K4me3 mark.
+The SNP with the highest score within each locus region in a given tissue was designated the BestSNP, which served as the score representing that locus.
+
+Cell-type-specificity scores per locus were estimated by normalizing BestSNP scores so that the sum of scores for a given locus across all cell types equaled 1.
+Cell-type-specificity scores per tissue were defined by summing normalized BestSNP scores across all loci within a given tissue.
+10,000 sets of SNPs (matched to the IndexSNP regions having the same total number of H3K4me3 peaks) were sampled (from among background SNPs provided with the software) to estimate null distributions of cell-type-specificity scores per locus and per tissue.
+P-values for cell-type-specificity scores summed across all BestSNPs for each tissue (the observed per-tissue score) were estimated as the proportion of random SNP sets with a per-tissue score exceeding the observed per- tissue score.
+We identified loci with BestSNP cell-type-specificity scores falling at or above the 95th percentile of the corresponding null distribution as demonstrating greater than expected specificity within a given cell type (50).
+Figure S20 shows p-values for the cell-type-specific overlap of H3K4me3 marks and IndexSNP regions for each cell type.
+Four cell/tissue types showed significant overlap at nominal p ≤ 0.05: for EduYears, anterior caudate (p = 0.0089), CD4+ naive primary cells (p = 0.032), hippocampus middle (p = 0.05) and muscle satellite cultured cells (p = 0.0236); and for College, anterior caudate (p = 0.0007).
+Additionally, for College the mid-frontal lobe showed marginal enrichment at p = 0.0502.
+Only the anterior caudate results for College survive correction for multiple hypothesis testing.
+The results from the analysis of overlap between H3K4me4 chromatin marks and education-associated SNP regions suggest gene expression regulatory function for those loci in specific cell types.
+In particular, we note that anterior caudate tissue appears enriched for both EduYears and College phenotypes, although only the latter survives multiple testing correction.
+Figure S21 shows cell-type-specificity scores per locus in the four nominally significant tissues and 95th-percentile threshold (dashed red line).
+Loci above the threshold were identified as showing greater than chance specificity within that particular cell type.
+Table S19 identifies these enriched loci, along with distance to the nearest chromatin mark.
+
+\subsection{Chromosome interactions}
+
+I like when chromosomes interact because I like epistasis. Also there is a cool figure with lots of circles.
+
+\subsection{Transcription factor enrichment}
+
+\subsection{SNP colocalisation with genic features}
+
 \section{References}

### 2013-06-18 analysis/gpmaps.R

@@ -200,7 +200,7 @@ xtable(tab, digits = c(0, 0, 0, 0, 2, 2, 2, 2))
 
 l <- datHeatmapGp(bsig)
 l <- GpMod(l, 2, "EGCUT", rotateGp, rotateGp, rotateGp)
-l <- GpMod(l, 3, "EGCUT", flipGpRow, flipGpCol)
+l <- GpMod(l, 3, "EGCUT", t)
 l <- GpMod(l, 5, "BSGS", flipGpCol)
 l <- GpMod(l, 6, "EGCUT", flipGpRow)
 l <- GpMod(l, 10, "Fehrmann", flipGpCol)
@@ -218,7 +218,7 @@ plot1 <- plotTileGp(subset(l, code3 %in% 1:15))
 plot2 <- plotTileGp(subset(l, code3 %in% 16:30))
 grid.arrange(plot1, plot2, ncol=2)
 dev.off()
-
+plot1
 
 
 temp <- subset(sig, probegene == "MBNL1")

### 2013-06-19 docs/manuscript/methods.tex
updating methods

@@
 
-We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions.
+We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013, Rietvald2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions.
 We subsequently refer to these 731 SNPs as the IndexSNPs A recent paper shows that H3K4me3 chromatin marks are the most cell-type-specific marks in terms of co-localization with previously published GWAS loci \cite{Trynka2013}.
@@
 10,000 sets of SNPs (matched to the IndexSNP regions having the same total number of H3K4me3 peaks) were sampled (from among background SNPs provided with the software) to estimate null distributions of cell-type-specificity scores per locus and per tissue.
-P-values for cell-type-specificity scores summed across all BestSNPs for each tissue (the observed per-tissue score) were estimated as the proportion of random SNP sets with a per-tissue score exceeding the observed per- tissue score.
-We identified loci with BestSNP cell-type-specificity scores falling at or above the 95th percentile of the corresponding null distribution as demonstrating greater than expected specificity within a given cell type (50).
-Figure S20 shows p-values for the cell-type-specific overlap of H3K4me3 marks and IndexSNP regions for each cell type.
-Four cell/tissue types showed significant overlap at nominal p ≤ 0.05: for EduYears, anterior caudate (p = 0.0089), CD4+ naive primary cells (p = 0.032), hippocampus middle (p = 0.05) and muscle satellite cultured cells (p = 0.0236); and for College, anterior caudate (p = 0.0007).
-Additionally, for College the mid-frontal lobe showed marginal enrichment at p = 0.0502.
-Only the anterior caudate results for College survive correction for multiple hypothesis testing.
-The results from the analysis of overlap between H3K4me4 chromatin marks and education-associated SNP regions suggest gene expression regulatory function for those loci in specific cell types.
-In particular, we note that anterior caudate tissue appears enriched for both EduYears and College phenotypes, although only the latter survives multiple testing correction.
-Figure S21 shows cell-type-specificity scores per locus in the four nominally significant tissues and 95th-percentile threshold (dashed red line).
-Loci above the threshold were identified as showing greater than chance specificity within that particular cell type.
-Table S19 identifies these enriched loci, along with distance to the nearest chromatin mark.
+P-values for cell-type-specificity scores summed across all BestSNPs for each tissue (the observed per-tissue score) were estimated as the proportion of random SNP sets with a per-tissue score exceeding the observed per-tissue score.
+We identified loci with BestSNP cell-type-specificity scores falling at or above the 95th percentile of the corresponding null distribution as demonstrating greater than expected specificity within a given cell type \cite{Trynka2013}.
 
@@
 
-I like when chromosomes interact because I like epistasis. Also there is a cool figure with lots of circles.
+It has been shown \cite{Lieberman-Aiden2009} that different regions on different chromosomes or within chromosomes spatially colocalise within the cell.
+We shall refer to the colocalisation of two chromosome regions as a chromosome interaction.
+A map of all known pairwise chromosome interactions for K562 blood cell lines was recently produced \cite{Lan2012}, and we hypothesised that part of the underlying biological mechanism behind some of the 501 epistatic interactions may arise from chromosome interactions.
+We found that 44 of the putative epistatic interactions were amongst SNPs that were within 5Mb of known chromosome interactions.
+This means that SNP A was no more than 2.5Mb from the focal point of the chromosome interaction on chromosome A, and SNP B was no more than 2.5Mb from the focal point on chromosome B.
+
+We performed simulations to test how extreme the observation of 44 epistatic interactions overlapping with chromosome interactions is compared to chance.
+Chromosome interactions fall within functional genomic regions \cite{Lieberman-Aiden2009, Lan2012}, and the SNPs in our epistatic interactions are enriched for functional genomic regions.
+Therefore, we designed the simulations to ensure that the null distribution was of chromosome interactions between SNPs enriched for functional genomic regions but with no known epistatic interactions.
+To do this we used the 731 SNPs that form the 501 putative epistatic interactions and randomly shuffled them to create new sets of 501 pairs, disallowing any SNP combinations that were in the original set.
+Therefore, each new random set was enriched for functional regions but had no genetic interactions.
+We scanned the map of chromosome interactions for overlaps with the new sets and then repeated the random shuffling process.
+We performed 10,000 such permutations to generate a null distribution of chromosome interaction overlaps.
+
+We repeated this process, searching for overlaps within 1Mb, 250kb, and 10kb.
 
@@
 
+We tested for enrichment of genic features for the 731 SNPs that comprise the 501 epistatic interactions.
+For each of the 731 SNPs we calculated LD with all neighbouring SNPs within a radius of 1kb and kept all neighbouring SNPs with LD $r^2 > 0.8$.
+We then cross-referenced the remaining regional SNPs with the annotated reference genome (version HG19) querying whether the regional SNPs fell in intergenic, promotor, intronic, exonic, splice sites, 5' UTR or 3' UTR positions.
+Therefore a particular query SNP might have regional SNPs in multiple genic features.
+
+We then performed the whole querying process for each of the 528,509 SNPs present in the SNP chip used in the scan, and used the results from this second analysis to establish the expected proportion of SNPs for each genic feature.
+
 \section{References}

### 2013-06-20 docs/manuscript/methods.tex
updated manuscript and methods

@@
 
-We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013, Rietvald2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions.
+We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions.
 We subsequently refer to these 731 SNPs as the IndexSNPs A recent paper shows that H3K4me3 chromatin marks are the most cell-type-specific marks in terms of co-localization with previously published GWAS loci \cite{Trynka2013}.
@@
 
-For each IndexSNP, a locus region was defined, bounded on either side of the IndexSNP by the most distant SNP within 250 kb that was in LD (r2 > 0.8).
+For each IndexSNP, a locus region was defined, bounded on either side of the IndexSNP by the most distant SNP within 250 kb that was in LD ($r^{2} > 0.8$).
 For each SNP in a locus region regulatory activity scores were calculated as the height of the nearest H3K4me3 mark divided by distance from the SNP to the H3K4me3 mark.
@@
 
+\subsection{SNP colocalisation with genomic features}
+\label{sub:snp_colocalisation_with_genomic_features}
+
+We tested for enrichment of genomic features for the 687 IndexSNPs that comprise the 434 epistatic interactions with data present in discovery and replication datasets.
+For each of the 687 IndexSNPs we calculated LD with all regional SNPs within a radius of 0.5Mb and kept all regional SNPs with LD $r^{2} > 0.8$.
+We then cross-referenced the remaining regional SNPs with the annotated reference genome (version HG19, GENCODE annotation version 16 \cite{Harrow2012}) querying whether the regional SNPs fell in intergenic, promotor, intronic, exonic, splice sites, 5' UTR or 3' UTR positions.
+Therefore a particular IndexSNP might cover multiple genomic features.
+
+We then performed the whole querying process for each of the 528,509 SNPs present in the SNP chip used in the scan, and used the results from this second analysis to establish a null distribution for the expected proportion of SNPs for each genomic feature.
+We calculated $p$-values for enrichment of each of the seven genomic features independently, and for \emph{cis}- and \emph{trans}-SNPs separately, using a binomial test.
+For each genomic feature we used the expected proportion of SNPs as the expected probability of ``success'' ($p$).
+Here, a success is defined as an IndexSNP residing in a region that includes the genomic feature.
+The observed number of successes for each IndexSNP ($k$) out of the total count of IndexSNPs ($n$) was then modelled as $\Pr (X = k) = {n \choose k} p^{k}(1-p)^{n-k}$.
+
 \subsection{Transcription factor enrichment}
 
-\subsection{SNP colocalisation with genic features}
+To test for enrichment of transcription factor binding sites (TFBS) we followed a procedure similar to that described in Section \ref{sub:snp_colocalisation_with_genomic_features}.
+For each of the 687 IndexSNPs we extracted regional SNPs as previously described.
+We then used the PWMEnrich package in Bioconductor (http://www.bioconductor.org/packages/2.12/bioc/html/PWMEnrich.html) to identify which TFBSs each of the regional SNPs for one IndexSNP falls in (within a radius of 250bp).
+Thus, the number of occurrences of a particular TFBS was counted for each IndexSNP.
+We used the ``Threshold-free affinity'' method for identifying TFBSs \cite{Stormo2000}.
 
-We tested for enrichment of genic features for the 731 SNPs that comprise the 501 epistatic interactions.
-For each of the 731 SNPs we calculated LD with all neighbouring SNPs within a radius of 1kb and kept all neighbouring SNPs with LD $r^2 > 0.8$.
-We then cross-referenced the remaining regional SNPs with the annotated reference genome (version HG19) querying whether the regional SNPs fell in intergenic, promotor, intronic, exonic, splice sites, 5' UTR or 3' UTR positions.
-Therefore a particular query SNP might have regional SNPs in multiple genic features.
-
-We then performed the whole querying process for each of the 528,509 SNPs present in the SNP chip used in the scan, and used the results from this second analysis to establish the expected proportion of SNPs for each genic feature.
+We constructed a null distribution of expected TFBS occurrences based on the same null hypothesis as described in Section \ref{sub:snp_colocalisation_with_genomic_features} - the probability of an IndexSNP covering a particular TFBS is identical to any of the 528,509 SNPs in the discovery SNP chip.
+To do this, we performed the same procedure for each SNP in the discovery SNP chip as was performed for each IndexSNP to obtain an expected probability of covering a particular TFBS.
+We then tested the IndexSNPs for enrichment of each TFBS independently, and for \emph{cis}- and \emph{trans}-SNPs separately.
+$p$-values were obtained using Z-scores, calculated by using a normal approximation to the sum of binomial random variables representing motif hits along the sequence \cite{HoSui2005}.
 

### 2013-06-20 docs/manuscript/nature_letter.tex
updated manuscript and methods

@@
 \author{Gibran Hemani, Joseph E Powell, Konstantin Shakhbazov, \\
-Greg Gibson, Grant W Montgomery, Peter M Visscher}
+Harm-Jan Westra, Anjali Henders, Allan F. McRae, \\
+Andres Metspalu, Greg Gibson, Tonu Esko, Lude Franke,  \\
+Nick Martin, Grant W Montgomery, Peter M Visscher}
 
@@
 For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
-We observed that nine of the 14 masking SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$, Supplementary Figure \ref{fig:snplocations}).
 Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
+We see similar epistatic networks involving multiple \emph{trans}-acting SNPs for
 
@@
 We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance.
-It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
+It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}, and that `transcriptional factories' can organise across distant chromosomal regions \cite{Osborne2004}.
 We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
@@
 Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
-All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least five data points in all nine genotype classes.
+All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects.
+All SNP pairs were required to have at least five data points in all nine genotype classes.
 If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained.
@@
 A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.
+
+\subsection{Acknowledgements}
+
+We are grateful to the volunteers for their generous participation in these studies, and thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.
+
+The University of Queensland group is supported by the Australian Research Council (DP130102666, DE130100691), the Australian National Health and Medical Research Council (APP1011506, APP1047956, APP1048853) and the National Institutes of Health (GM099568, GM075091).
+
+The QIMR researchers acknowledge funding from the Australian National Health and Medical Research Council (grants 241944, 389875, 389891, 389892, 389938, 442915, 442981, 496739, 496688 and 552485), the and the National Institutes of Health (grants AA07535, AA10248, AA014041, AA13320, AA13321, AA13326 and DA12854).
+
+This work could not have been completed without access to high performance GPGPU compute clusters.
+We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au).
+We also acknowledge Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.
+
+The CHDWB study funding support from the Georgia Institute of Technology Research Foundation.
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript
+
+The Fehrmann study was supported by grants from the Celiac Disease Consortium (an innovative cluster approved by the Netherlands Genomics Initiative and partly funded by the Dutch Government (grant BSIK03009), the Netherlands Organization for Scientific Research (NWO-VICI grant 918.66.620, NWO-VENI grant 916.10.135 to L.F.), the Dutch Digestive Disease Foundation (MLDS WO11-30), and a Horizon Breakthrough grant from the Netherlands Genomics Initiative (grant 92519031 to L.F.).
+This project was supported by the Prinses Beatrix Fonds, VSB fonds, H.
+Kersten and M.
+Kersten (Kersten Foundation), The Netherlands ALS Foundation, and J.R.
+van Dijk and the Adessium Foundation.
+The research leading to these results has received funding from the European Community’s Health Seventh Framework Programme (FP7/2007-2013) under grant agreement 259867.
+
+The EGCUT study received targeted financing from Estonian Government SF0180142s08, Center of Excellence in Genomics (EXCEGEN) and University of Tartu (SP1GVARENG).
+We acknowledge EGCUT technical personnel, especially Mr V.
+Soo and S.
+Smit.
+Data analyzes were carried out in part in the High Performance Computing Center of University of Tartu.
 

### 2013-06-20 analysis/check_dprime.R

@@ -37,6 +37,8 @@ ciscis$gap <- abs(ciscis$position1 - ciscis$position2) / 1000000
 
 ciscis
 
+table(ciscis$gap < 1)
+
 
 #============================================================#
 #============================================================#
@@ -71,3 +73,5 @@ for(i in 1:nrow(ciscis))
 	cat(i, " - ", l[i], "\n")
 }
 
+sort(l^2)
+mean(l^2)

### 2013-06-21 docs/manuscript/nature_letter.tex
worky work

@@
 \date{}
-\author{Gibran Hemani, Joseph E Powell, Konstantin Shakhbazov, \\
+\author{Gibran Hemani, Konstantin Shakhbazov, \\
 Harm-Jan Westra, Anjali Henders, Allan F. McRae, \\
-Andres Metspalu, Greg Gibson, Tonu Esko, Lude Franke,  \\
-Nick Martin, Grant W Montgomery, Peter M Visscher}
+Andres Metspalu, Greg Gibson, Nick Martin, Tonu Esko, \\
+Lude Franke, Grant W Montgomery, Peter M Visscher, Joseph E Powell}
 
@@
 	\centering
-	\includegraphics[width=5in]{Grey_graph_of_interactions_2_lists.pdf}
+	\includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
 	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).

### 2013-06-21 analysis/gpmaps.R

@@ -205,7 +205,7 @@ l <- GpMod(l, 5, "BSGS", flipGpCol)
 l <- GpMod(l, 6, "EGCUT", flipGpRow)
 l <- GpMod(l, 10, "Fehrmann", flipGpCol)
 l <- GpMod(l, 11, "EGCUT", flipGpRow)
-l <- GpMod(l, 12, "EGCUT", flipGpRow)
+l <- GpMod(l, 12, "EGCUT", flipGpRow) 
 l <- GpMod(l, 13, "EGCUT", flipGpRow)
 l <- GpMod(l, 15, "EGCUT", flipGpRow)
 l <- GpMod(l, 22, "BSGS", flipGpRow)
@@ -213,7 +213,7 @@ l <- GpMod(l, 22, "EGCUT", flipGpRow)
 l <- GpMod(l, 22, "Fehrmann", flipGpRow)
 
 
-pdf(file="~/repo/eQTL-2D/analysis/images/gpBonfRep.pdf", width=7, height=14)
+pdf(file="~/repo/eQTL-2D/analysis/images/gpBonfRep.pdf", width=4, height=7)
 plot1 <- plotTileGp(subset(l, code3 %in% 1:15))
 plot2 <- plotTileGp(subset(l, code3 %in% 16:30))
 grid.arrange(plot1, plot2, ncol=2)

### 2013-06-21 analysis/qq_meta.R

@@ -218,31 +218,7 @@ createAverageGp <- function(gcm, gcs)
 
 
 
-
-#=============================================================#
-#=============================================================#
-
-# Read in data
-
-load("~/repo/eQTL-2D/analysis/interaction_list_replication_summary.RData")
-
-load("~/repo/eQTL-2D/replication/results/replication_GrngHT12v3.RData")
-newsig$code <- with(newsig, paste(probename, snp1, snp2))
-fehr <- mod
-names(fehr) <- newsig$code
-
-load("~/repo/eQTL-2D/replication/results/replication_EGCUT.RData")
-newsig$code <- with(newsig, paste(probename, snp1, snp2))
-egcut <- mod
-names(egcut) <- newsig$code
-
-
-#=============================================================#
-#=============================================================#
-
-# Perform meta analysis on replication data
-
-sig_all <- performMeta(sig_all, egcut, fehr)
+load(file="~/repo/eQTL-2D/analysis/interaction_list_meta_analysis.RData")
 
 thresh <- -log10(0.05 / 442)
 with(sig_all, table(filter, pnest_meta > thresh))
@@ -256,19 +232,18 @@ with(sig_all, table(filter, is.na(pnest_meta)))
 
 # Make Q-Q plots for meta analysis
 
-meta <- makeQqDat(sig_all, 0.05, "pnest_meta")
+
 qqPlot(meta, thresh)
-ggsave(file="~/repo/eQTL-2D/analysis/images/qqMetaNonsig.pdf", width=15, height=7.5)
+ggsave(file="~/repo/eQTL-2D/analysis/images/qqMetaNonsig.pdf", width=5, height=10)
 qqPlot(meta, max(meta$pnest_meta))
-ggsave(file="~/repo/eQTL-2D/analysis/images/qqMetaAll.pdf", width=15, height=7.5)
+ggsave(file="~/repo/eQTL-2D/analysis/images/qqMetaAll.pdf", width=5, height=10)
 qqPlot2(meta, thresh)
-ggsave(file="~/repo/eQTL-2D/analysis/images/qqMeta.pdf", width=7.5, height=15)
+ggsave(file="~/repo/eQTL-2D/analysis/images/qqMeta.pdf", width=5, height=10)
 
 with(meta, table(filter == 3, pnest_meta > upper))
 with(meta, table(filter == 3, pnest_fehr > upper_fehr))
 with(meta, table(filter == 3, pnest_egcut > upper_egcut))
 
[…]

### 2013-06-21 analysis/variance_analysis.R

@@ -1,3 +1,6 @@
+library(reshape2)
+library(ggplot2)
+
 load("~/repo/eQTL-2D/analysis/interaction_list_replication_summary.RData")
 
 
@@ -12,22 +15,24 @@ for(i in 1:nrow(sig))
 	sig$vc_egcut[[i]] <- sig$vc_egcut[[i]][[1]]
 }
 
-bsgs <- as.data.frame(do.call(rbind, sig$vc))
-fehr <- as.data.frame(do.call(rbind, sig$vc_fehr))
-egcut <- as.data.frame(do.call(rbind, sig$vc_egcut))
+bsgs <- do.call(rbind, sig$vc)
+fehr <- do.call(rbind, sig$vc_fehr)
+egcut <- do.call(rbind, sig$vc_egcut)
 
-bsgs$varA <- bsgs$a. + bsgs$.a
-fehr$varA <- fehr$a. + fehr$.a
-egcut$varA <- egcut$a. + egcut$.a
+standardiseVg <- function(dat)
+{
+	vG <- apply(dat, 1, sum)
+	dat <- dat / vG
+	dat <- as.data.frame(dat)
+	dat$varA <- dat$a. + dat$.a
+	dat$varD <- dat$d. + dat$.d
+	dat$varI <- dat$aa + dat$ad + dat$da + dat$dd
+	return(dat)
+}
 
-bsgs$varD <- bsgs$d. + bsgs$.d
-fehr$varD <- fehr$d. + fehr$.d
-egcut$varD <- egcut$d. + egcut$.d
-
-bsgs$varI <- bsgs$aa + bsgs$ad + bsgs$da + bsgs$dd
-fehr$varI <- fehr$aa + fehr$ad + fehr$da + fehr$dd
-egcut$varI <- egcut$aa + egcut$ad + egcut$da + egcut$dd
-
+bsgs <- standardiseVg(bsgs)
+fehr <- standardiseVg(fehr)
+egcut <- standardiseVg(egcut)
 
 varA <- data.frame(bsgs = bsgs$varA, fehr = fehr$varA, egcut = egcut$varA)
 varD <- data.frame(bsgs = bsgs$varD, fehr = fehr$varD, egcut = egcut$varD)
@@ -37,9 +42,94 @@ pairs(varA)
 pairs(varD)
 pairs(varI)
 
+vA <- melt(varA)
+vA$vc <- "A"
+vD <- melt(varD)
[…]

### 2013-06-24 docs/manuscript/methods.tex
working on GP maps

@@
 \documentclass{article}
+\usepackage{url}
 \usepackage[superscript,biblabel]{cite}
@@
 
+\tableofcontents
+
 \section{Discovery data}
+
+\subsection{Data description}
 
@@
 
-\section{Normalisation}
+\subsection{Normalisation}
 
@@
 The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero.
-Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan.
+Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan to remove any skewness and avoid results being driven by outliers.
 The GenABEL package for R was used to perform the normalisation \cite{Aulchenko2007}.
@@
 The effective number of tests in one 2D scan being performed is therefore $N_{*} = 0.05 / T_{*} \approx 2.33 \times 10^{10}$.
-To correct for the testing of multiple traits we established an experiment wide threshold of $T = 0.05 / (N_{*} \times 7339) = 2.91 \times 10^{-16}$.
+To correct for the testing of multiple traits we established an experiment wide threshold of $T_{e} = 0.05 / (N_{*} \times 7339) = 2.91 \times 10^{-16}$.
 This is likely to be conservative as it assumes independence between probes.
@@
 Following filtering on this threshold only SNP pairs with at least 5 data points in all 9 genotype classes were kept.
-We then calculated the LD between interacting SNPs and removed any pairs with $r^2 > 0.1$.
+We then calculated the LD between interacting SNPs and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition.
 If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained.
@@
 \section{Replication}
-We have attempted replication of the 529 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB.
-It was required that LD $r^2 < 0.1$ between interacting SNPs, and all nine genotype classes had at least 5 individuals present in order to proceed with statistical testing for replication in both datasets.
+
+\subsection{Data description}
+
+We have attempted replication of the 501 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB.
+It was required that LD $r^2 < 0.1$ and $D'^{2} < 0.1$ between interacting SNPs, and all nine genotype classes had at least 5 individuals present in order to proceed with statistical testing for replication in both datasets.
 We also excluded any putative SNPs that had discordant allele frequencies in any of the datasets.
@@
 Whole genome genotypes were measured using Illumina OmniQuad arrays.
-Due to the small sample size, most SNP pairs did not pass filtering in this dataset and so we have excluded it from the rest of the analysis.
+Due to the small sample size, most SNP pairs did not pass filtering in this dataset (185 remained) and so we have excluded it from the rest of the analysis.
 
 \subsection{Meta Analysis}
+
 The 4 \emph{d.f.} interaction $p$-values for each independent replication dataset were calculated using the same statistical test as was performed in the discovery dataset.
-We then took the interaction $p$-values from EGCUT and Fehrmann and calculated a joint $p$-value as $-2 \log p_{1} - 2 \log p_{2} \sim \chi^{2}_{4 d.f.}$.
+We then took the interaction $p$-values from EGCUT and Fehrmann and calculated a joint $p$-value using Fisher's method of combining $p$-values for a meta analysis as $-2 \ln p_{1} - 2 \ln p_{2} \sim \chi^{2}_{4 d.f.}$.
+As in the discovery analysis, all gene expression levels were normalised using rank transformation to avoid skew or outliers in the distribution.
 
@@
 
-We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions.
+We employed a recently published method (\url{http://www.broadinstitute.org/mpg/epigwas/}) \cite{Trynka2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions.
 We subsequently refer to these 731 SNPs as the IndexSNPs A recent paper shows that H3K4me3 chromatin marks are the most cell-type-specific marks in terms of co-localization with previously published GWAS loci \cite{Trynka2013}.
@@
 For each of the 687 IndexSNPs we extracted regional SNPs as previously described.
-We then used the PWMEnrich package in Bioconductor (http://www.bioconductor.org/packages/2.12/bioc/html/PWMEnrich.html) to identify which TFBSs each of the regional SNPs for one IndexSNP falls in (within a radius of 250bp).
+We then used the PWMEnrich package in Bioconductor (\url{http://www.bioconductor.org/packages/2.12/bioc/html/PWMEnrich.html}) to identify which TFBSs each of the regional SNPs for one IndexSNP falls in (within a radius of 250bp).
 Thus, the number of occurrences of a particular TFBS was counted for each IndexSNP.
@@
 
+\subsection{Defining previously identified SNP associations}
+
+The discovery dataset (BSGS) had previously been analysed for additive and dominant marginal effects for all gene expression levels \cite{Powell2012, Powell2013}.
+To define SNPs that had been previously detected to have effects for a particular gene expression level we used a significance threshold accounting for multiple testing across SNPs and expression probes, $T_{m} = 0.05 / (528509 \times 7339) = 1.29 \times 10^{-11}$.
+From this, we found that only nine of the 501 discovery interactions had known main effects, 64 were between SNPs that had no known marginal effects, and 439 were between a SNP with a known marginal effect and a SNP with no known marginal effect.
+
 \section{References}

### 2013-06-24 docs/manuscript/nature_letter.tex
working on GP maps

@@
 \usepackage{threeparttable}
+\usepackage{authblk}
+
+\makeatletter
+\renewcommand\AB@affilsepx{. \protect\Affilfont}
+\makeatother
 
@@
 \date{}
-\author{Gibran Hemani, Konstantin Shakhbazov, \\
-Harm-Jan Westra, Anjali Henders, Allan F. McRae, \\
-Andres Metspalu, Greg Gibson, Nick Martin, Tonu Esko, \\
-Lude Franke, Grant W Montgomery, Peter M Visscher, Joseph E Powell}
+\author[1,2,*]{Gibran Hemani}
+\author[1,2]{Konstantin Shakhbazov}
+\author[3]{Harm-Jan Westra}
+\author[4]{Anjali K Henders}
+\author[1,2]{Allan F. McRae}
+\author[5]{Andres Metspalu}
+\author[6]{Greg Gibson}
+\author[4]{Nick G Martin}
+\author[5,7,8]{Tonu Esko}
+\author[3]{Lude Franke}
+\author[4]{Grant W Montgomery}
+\author[1,2]{Peter M Visscher}
+\author[1,2]{Joseph E Powell}
+
+{\tiny
+
+\affil[1]{University of Queensland Diamantina Institute, University of Queensland, Princess Alexandra Hospital, Brisbane, Queensland, Australia}
+
+\affil[2]{The University of Queensland, Queensland Brain Institute, Brisbane, QLD, Australia}
+
+\affil[3]{Department of Genetics, University Medical Center Groningen, University of Groningen, Hanzeplein 1, Groningen, the Netherlands}
+
+\affil[4]{Queensland Institute of Medical Research, Brisbane, Queensland, Australia}
+
+\affil[5]{Estonian Genome Center, University of Tartu, Tartu, 51010, Estonia}
+
+\affil[6]{School of Biology and Centre for Integrative Genomics, Georgia Institute of Technology, Atlanta, Georgia United States of America}
+
+\affil[7]{Divisions of Endocrinology, Children's Hospital, Boston, MA, 02115, US}
+
+\affil[8]{Medical and Population Genetics, Broad Institute, Cambridge, MA, 02142, US}
+
+\affil[*]{Corresponding author: g.hemani@uq.edu.au}
+
+}
 
@@
 \begin{abstract}
+
 Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome.
 The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics.
-Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
-Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}.
+Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been reported in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
+Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}.
 Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
 In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
-Three hundred and forty-five interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} at a false dicovery rate (FDR) of 2.5\%, with 30 significant at a conservative $p < 0.05$ Bonferroni level.
+We tested the discovery interactions for replication in two independent data sets \cite{Metspalu2004, Fehrmann2011}.
+Three hundred and forty-five interactions had replication interaction $p$-values that were more extreme than the 2.5\% confidence interval of the distribution under the null hypothesis of no epistasis, with 30 significant at a conservative $p < 0.05$ Bonferroni level.
 There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
@@
 In addition, we observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}).
-The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations.
+The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets.
 A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication.
-It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
+It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is up to four orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
 Therefore these results are encouraging with regards to the detection and replication of epistasis.
@@
 Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows that this is unlikely to be the most effective strategy for its detection.
-The majority of our discovery interactions comprised of one SNP that had a previously known association and one SNP that had no previous association in the dataset \cite{Powell2013} (439 out of 501).
+The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \cite{Powell2013} (439 out of 501, Methods).
 Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects.
-Additionally, we observed that the largest epistatic variance component for the 501 interactions was divided amongst additive $\times$ additive, additive $\times$ dominance and dominance $\times$ dominance terms proportional to what is expected by chance ($p = 0.22$ for divergence from expectation).
+Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst additive $\times$ additive, additive $\times$ dominance, dominance $\times$ additive and dominance $\times$ dominance ($p = 0.22$ for departure from expectation).
 This is not surprising because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.
@@
 We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
-For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect.
+For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an effect (Supplementary Figure \ref{fig:MBNL1}.
 Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
-We see similar epistatic networks involving multiple \emph{trans}-acting SNPs for
+We see similar epistatic networks involving multiple \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5}), and CAST (Supplementary Figure \ref{fig:CAST}).
 
@@
 However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues).
-This difference between \emph{cis} and \emph{trans} SNPs suggests a range of molecular mechanisms by which epistasis might arise and the \emph{cis}-SNPs may provide tissue specificity in these interactions.
+This difference between \emph{cis} and \emph{trans} SNPs suggests different roles in which epistasis might arise where the \emph{cis}-SNPs provide tissue specificity in these interactions.
 There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a}, consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$).
-In particular, there was enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs, a known regulator of transcription in haematopoietic cells \cite{Tsai1994} ($p = 1 \times 10^{-40}$).
 
-We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance.
-It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}, and that `transcriptional factories' can organise across distant chromosomal regions \cite{Osborne2004}.
-We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
+We also demonstrate spacial organisation of interacting loci suggesting a mechanism by which biological function can lead to epistatic genetic variance.
+It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
+We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
 Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.
+Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Osborne2004, Rieder2013}.
 
@@
 Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components.
-Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}.
-Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes.
-By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance (Methods).
-It must be noted that there are at least three caveats to this comparison.
+Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}.
+Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that on average they explain 1.73\% of the phenotypic variance of each of the 7339 probes.
+By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain 0.25\% of phenotypic variance, approximately seven times lower than the additive variance (Methods).
+There are at least three caveats to this comparison.
 Firstly, the ratio of additive to epistatic variance may differ at different effect sizes.
@@
 Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
-We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
+We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
 SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
@@
 
-We are grateful to the volunteers for their generous participation in these studies, and thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.
+We are grateful to the volunteers for their generous participation in these studies. We thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.
+
+This work could not have been completed without access to high performance GPGPU compute clusters.
+We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au).
+We also acknowledge Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.
 
@@
 The QIMR researchers acknowledge funding from the Australian National Health and Medical Research Council (grants 241944, 389875, 389891, 389892, 389938, 442915, 442981, 496739, 496688 and 552485), the and the National Institutes of Health (grants AA07535, AA10248, AA014041, AA13320, AA13321, AA13326 and DA12854).
-
-This work could not have been completed without access to high performance GPGPU compute clusters.
-We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au).
-We also acknowledge Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.
+We thank Anthony Caracella and Lisa Bowdler for technical assistance with the micro-array hybridisations.
 

### 2013-06-24 analysis/gpmaps.R

@@ -7,22 +7,6 @@ library(plyr)
 library(xtable)
 
 
-plot3dGp <- function(gp, title="", snp1="SNP1", snp2="SNP2", z=-45)
-{
-	p <- cloud(
-		gp, 
-		panel.3d.cloud=panel.3dbars, 
-		col="black", 
-		col.facet=c("#e5f5e0", "#A1D99B", "#31A354"), 
-		xbase=0.6, ybase=0.6,
-		xlab=list(label=snp1, cex=0.8), ylab=list(label=snp2, cex=0.8), zlab="",
-		default.scales=list(arrows=F, z = list(cex = 0), y = list(cex = 0.5), x=list(cex=0.5)),
-		cex.title = 0.5,
-		screen = list(z = z, x = -60, y = 3),
-		main = title
-	)
-}
-
 
 plot3dHairballsRep <- function(sig, pg, cissnp, cohort, z=45)
 {
@@ -99,6 +83,11 @@ datHeatmapGp <- function(sig)
 		g$code <- a$code[i]
 		g$code2 <- a$probegene[i]
 		g$code3 <- i
+		g$s1 <- a$snp1[i]
+		g$s2 <- a$snp2[i]
+		g$chr1 <- a$chr1[i]
+		g$chr2 <- a$chr2[i]
+		g$index <- i
 		g <- GpMod(g, i, "BSGS", rotateGp, rotateGp)
 		l[[i]] <- g
 	}
@@ -167,10 +156,8 @@ plotTileGp2 <- function(l)
 }
 
 
-convertToScientific <- function(x)
-{
-
-}
+#=================================================================================================#
+#=================================================================================================#
 
 
 load("~/repo/eQTL-2D/analysis/interaction_list_meta_analysis.RData")
@@ -194,6 +181,9 @@ tab
 names(tab) <- c("Gene (chr.)", "SNP 1 (chr.)", "SNP 2 (chr.)", "Discovery", "Fehrmann", "EGCUT", "Combined replication")
 xtable(tab, digits = c(0, 0, 0, 0, 2, 2, 2, 2))
 
+
+#=================================================================================================#
+#=================================================================================================#
 
 
 # Plots
[…]

### 2013-06-25 docs/manuscript/nature_letter.tex
updates: manuscript close to completion
images updated for
- 3d plots
- circle plots
- genomic features

organised scripts more sensibly

answered most comments from peter and grant on manuscript and methods

just need to calculate variances for replication datasets now

@@
 \usepackage{threeparttable}
+\usepackage{threeparttablex}
 \usepackage{authblk}
+\usepackage{multirow}
+\usepackage{longtable}
+\usepackage{lscape}
 
@@
 \date{}
+
 \author[1,2,*]{Gibran Hemani}
@@
 
-{\tiny
-
 \affil[1]{University of Queensland Diamantina Institute, University of Queensland, Princess Alexandra Hospital, Brisbane, Queensland, Australia}
-
 \affil[2]{The University of Queensland, Queensland Brain Institute, Brisbane, QLD, Australia}
-
 \affil[3]{Department of Genetics, University Medical Center Groningen, University of Groningen, Hanzeplein 1, Groningen, the Netherlands}
-
 \affil[4]{Queensland Institute of Medical Research, Brisbane, Queensland, Australia}
-
 \affil[5]{Estonian Genome Center, University of Tartu, Tartu, 51010, Estonia}
-
 \affil[6]{School of Biology and Centre for Integrative Genomics, Georgia Institute of Technology, Atlanta, Georgia United States of America}
-
 \affil[7]{Divisions of Endocrinology, Children's Hospital, Boston, MA, 02115, US}
-
 \affil[8]{Medical and Population Genetics, Broad Institute, Cambridge, MA, 02142, US}
-
 \affil[*]{Corresponding author: g.hemani@uq.edu.au}
-
-}
 
@@
 \maketitle
+
+\clearpage
 
@@
 We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
-For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an effect (Supplementary Figure \ref{fig:MBNL1}.
+For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an effect (Supplementary Figure \ref{fig:MBNL1_3D}.
 Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
-We see similar epistatic networks involving multiple \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5}), and CAST (Supplementary Figure \ref{fig:CAST}).
+We see similar epistatic networks involving multiple \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149_3D}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1_3D}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5_3D}), and CAST (Supplementary Figure \ref{fig:CAST_3D}).
 
@@
 We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
-Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.
-Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Osborne2004, Rieder2013}.
+Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Osborne2004, Rieder2012}.
 
@@
 Chromosomes are represented by black blocks and ordered from 1 to 22 clockwise, starting from the top.
-Grey lines represent no evidence for replication, blue lines denote replication in at least one dataset, and red lines denote replication in two datasets.
+Grey lines represent no evidence for replication, blue lines denote interactions that are outside the 97.5\% confidence interval or the Q-Q plot (Figure \ref{fig:qqMeta}), and red lines denote replication at the Bonferroni correction level.
 Most interactions are characterised as being \emph{cis}-\emph{trans} to the expression probe.}
@@
 	\centering
-	\includegraphics[width=5in]{OverlapAnnotation}
-	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 5kb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic regions were covered by the 501 significant interactions.
-There is enrichment for \emph{cis}-acting SNPs in promotor regions, and \emph{cis}- and \emph{trans}-acting SNPs in intronic regions.}
+	\includegraphics[width=5in]{OLapWithGenomicFeaturesFinal}
+	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 1kb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic features (\emph{x}-axis) were covered by the SNPs that compose the 501 significant interactions.
+Green bars represent the proportion (\emph{y}-axis) of the 528,509 SNPs used in the analysis that fall within the range of the different genomic features.
+There is enrichment for \emph{cis}-acting SNPs (red bars) in promotor regions, but \emph{trans}-acting SNPs (blue bars) are not enriched for genomic features.}
 	\label{fig:snplocations}
@@
 	\centering
-	\includegraphics[width=5in]{MBNL1}
-	\caption{\textbf{Genotype-phenotype maps for 14 interactions controlling MBNL1} Each bar represents the mean phenotypic value for individuals in that genotype class.
+	\includegraphics[width=5in]{MBNL1_3D}
+	\caption{\textbf{Genotype-phenotype maps for 14 interactions influencing the expression of MBNL1} Each bar represents the mean phenotypic value for individuals in that genotype class.
 The rs13069559 SNP typically has a \emph{cis}-additive decreasing effect on the expression of MBNL1, but in many of these interactions the \emph{cis} effect is masked when the \emph{trans} SNP is homozygous.}
-	\label{fig:mbnl13d}
+	\label{fig:MBNL1_3D}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\centering
+	\includegraphics[width=5in]{TMEM149_3D}
+	\caption{\textbf{Genotype-phenotype maps for 19 interactions influencing the expression of TMEM149} Each bar represents the mean phenotypic value for individuals in that genotype class.
+The rs13069559 SNP typically has a \emph{cis}-additive decreasing effect on the expression of TMEM149, but in many of these interactions the \emph{cis} effect is masked when the \emph{trans} SNP is homozygous.}
+	\label{fig:TMEM149_3D}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\centering
+	\includegraphics[width=5in]{NAPRT1_3D}
+	\caption{\textbf{Genotype-phenotype maps for 8 interactions influencing the expression of NAPRT1} Each bar represents the mean phenotypic value for individuals in that genotype class.}
+	\label{fig:NAPRT1_3D}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\centering
+	\includegraphics[width=5in]{TRAPPC5_3D}
+	\caption{\textbf{Genotype-phenotype maps for 16 interactions influencing the expression of TRAPPC5} Each bar represents the mean phenotypic value for individuals in that genotype class.}
+	\label{fig:TRAPPC5_3D}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\centering
+	\includegraphics[width=5in]{CAST_3D}
+	\caption{\textbf{Genotype-phenotype maps for 15 interactions influencing the expression of CAST} Each bar represents the mean phenotypic value for individuals in that genotype class.}
+	\label{fig:CAST_3D}
 \end{figure}
@@
 \clearpage
+\section{Supplementary Tables}
+\setcounter{table}{0}
+\makeatletter 
+\renewcommand{\thetable}{S\@arabic\c@table} 
+\makeatletter
+
+\begin{landscape}
+
+{\tiny
+
+\begin{ThreePartTable}
+\begin{longtable}{|llr|lrrl|lrrl|rrrr|r|}
+
+\caption{\textbf{Details on 501 interactions discovered in BSGS dataset}}\\
+
+\hline
+\multicolumn{3}{|c|}{Expression trait} & 
+\multicolumn{4}{|c|}{SNP 1} & 
+\multicolumn{4}{|c|}{SNP 2} & 
+\multicolumn{4}{|c|}{Interaction statistic / $-\log_{10}p$-values} & \\
+\hline
+Gene ID\tnote{a} & 
+Probe ID\tnote{b} & 
+Chr. & 
+rs ID & 
+Chr. & 
+Pos/Mb\tnote{c} & 
+Association\tnote{d} & 
+rs ID & 
+Chr. & 
+Pos/Mb\tnote{c} & 
+Association\tnote{d} & 
+BSGS\tnote{e} & 
+Fehrmann\tnote{f} & 
+EGCUT\tnote{f} & 
+Meta\tnote{g} & 
+Distance / Mb\tnote{h} \\
+\hline
+\endfirsthead
+
+\multicolumn{16}{c}{{\bfseries \tablename\ \thetable{} -- continued from previous page}} \\
+\hline
+\multicolumn{3}{|c|}{Expression trait} & 
+\multicolumn{4}{|c|}{SNP 1} & 
+\multicolumn{4}{|c|}{SNP 2} & 
+\multicolumn{4}{|c|}{Interaction statistic / $-\log_{10}p$-values} & \\
+\hline
+Gene ID\tnote{a} & 
+Probe ID\tnote{b} & 
+Chr. & 
+rs ID & 
+Chr. & 
+Pos/Mb\tnote{c} & 
+Association\tnote{d} & 
+rs ID & 
+Chr. & 
+Pos/Mb\tnote{c} & 
+Association\tnote{d} & 
+BSGS\tnote{e} & 
+Fehrmann\tnote{f} & 
+EGCUT\tnote{f} & 
+Meta\tnote{g} & 
+Distance / Mb\tnote{h} \\
+\hline
+\endhead
+
+\hline
+\multicolumn{16}{r}{{Continued on next page}} \\
+\endfoot
+
+\hline \hline
+\endlastfoot
+
+ABCA7 & ILMN\_1743205 & 19 & rs3752237 & 19 & 1047161 & ABCA7 & rs596183 & 6 & 158100199 &  & 5.82 & 0.38 & 0.02 & 0.09 &  \\
+  ABCA7 & ILMN\_1743205 & 19 & rs3752237 & 19 & 1047161 & ABCA7 & rs914737 & 9 & 139522101 &  & 5.50 &  &  &  &  \\
+  ABCC3 & ILMN\_1677814 & 17 & rs9455 & 17 & 48771135 & ABCC3 & rs4732202 & 7 & 136057883 &  & 6.10 & 0.02 & 1.81 & 0.95 &  \\
+  ACAT1 & ILMN\_1800008 & 11 & rs227064 & 11 & 108207393 & ACAT1 & rs4744894 & 9 & 72001517 &  & 6.59 & 1.04 & 1.78 & 2.02 &  \\
+  ADCK1 & ILMN\_1698777 & 14 & rs12431896 & 14 & 78088813 & ADCK1 & rs4833241 & 4 & 122933691 &  & 5.59 & 0.36 & 1.14 & 0.87 &  \\
+  ADCK1 & ILMN\_1698777 & 14 & rs8058066 & 16 & 88462550 &  & rs12431896 & 14 & 78088813 & ADCK1 & 6.58 & 2.04 & 0.83 & 2.05 &  \\
+  ADK & ILMN\_2358626 & 10 & rs2395095 & 10 & 76446305 & ADK & rs10824092 & 10 & 75929517 &  & 6.69 & 18.33 & 21.21 & 39.82 & 0.517 \\
+  AGAP6 & ILMN\_3239130 & 10 & rs2611512 & 10 & 51515534 & AGAP8 & rs2547996 & 5 & 95174319 &  & 6.22 &  &  &  &  \\
+  AHSA2 & ILMN\_1798308 & 2 & rs10881585 & 9 & 137112421 &  & rs842647 & 2 & 61119471 &  & 7.15 & 1.83 & 1.93 & 2.88 &  \\
+  AHSA2 & ILMN\_1798308 & 2 & rs2523971 & 6 & 29938258 & HLA-G & rs1177303 & 2 & 61388355 & AHSA2 & 5.45 & 0.92 & 0.64 & 0.94 &  \\
+  AKTIP & ILMN\_1665982 & 16 & rs2896940 & 16 & 57721127 &  & rs13332406 & 16 & 53489705 & AKTIP & 6.91 & 0.16 & 0.99 & 0.57 & 4.231 \\
+  AKTIP & ILMN\_1665982 & 16 & rs7189819 & 16 & 53536345 & AKTIP & rs1362032 & 7 & 125543391 &  & 5.93 & 0.71 & 0.20 & 0.42 &  \\
+  AKTIP & ILMN\_1665982 & 16 & rs7189819 & 16 & 53536345 & AKTIP & rs1473017 & 4 & 179323762 &  & 6.18 & 0.27 & 0.30 & 0.23 &  \\
+  ALDH3A2 & ILMN\_2401641 & 17 & rs3760489 & 17 & 19581009 & ALDH3A2 & rs11720112 & 3 & 161996349 &  & 6.26 & 0.33 & 1.37 & 1.01 &  \\
+  ANG & ILMN\_1760727 & 14 & rs9322855 & 14 & 21153299 & ANG & rs4866516 & 5 & 3032625 &  & 5.75 & 0.02 & 0.20 & 0.04 &  \\
+  ANPEP & ILMN\_1763837 & 15 & rs11073891 & 15 & 90363995 & ANPEP & rs3823523 & 7 & 154511163 &  & 5.85 & 0.44 & 1.09 & 0.90 &  \\
+  ANPEP & ILMN\_1763837 & 15 & rs11073891 & 15 & 90363995 & ANPEP & rs6846031 & 4 & 178019148 &  & 6.31 & 0.47 & 0.17 & 0.26 &  \\
+  AP3B1 & ILMN\_1768867 & 5 & rs6453374 & 5 & 77508159 & AP3B1 & rs4684443 & 3 & 4818792 &  & 5.94 & 0.05 &  &  &  \\
+  APPL2 & ILMN\_1765076 & 12 & rs935251 & 12 & 105580918 & APPL2 & rs2769594 & 9 & 87918528 &  & 5.60 & 0.80 & 1.02 & 1.16 &  \\
+  ARL17B & ILMN\_3231952 & 17 & rs12947580 & 17 & 75768225 &  & rs8079215 & 17 & 44064851 & ARL17B & 5.96 &  &  &  & 31.703 \\
+  ARL17B & ILMN\_3231952 & 17 & rs2834541 & 21 & 35932619 &  & rs8079215 & 17 & 44064851 & ARL17B & 6.65 &  &  &  &  \\
+  ARL17B & ILMN\_3231952 & 17 & rs8079215 & 17 & 44064851 & ARL17B & rs1950646 & 14 & 94722497 &  & 7.64 &  &  &  &  \\
+  ARL17B & ILMN\_3231952 & 17 & rs8079215 & 17 & 44064851 & ARL17B & rs2197777 & 12 & 125831219 &  & 6.26 &  &  &  &  \\
+  ARL17B & ILMN\_3231952 & 17 & rs8079215 & 17 & 44064851 & ARL17B & rs2684789 & 15 & 99492045 &  & 5.98 &  &  &  &  \\
+  ARL17B & ILMN\_3231952 & 17 & rs8079215 & 17 & 44064851 & ARL17B & rs9834627 & 3 & 191203546 &  & 5.72 &  &  &  &  \\
+  ATP13A1 & ILMN\_2134224 & 19 & rs4284750 & 19 & 19810050 &  & rs873870 & 19 & 19738554 &  & 5.30 & 12.18 & 3.25 & 14.23 & 0.071 \\
+  BID & ILMN\_1763386 & 22 & rs8919 & 22 & 18213057 & BID & rs9804943 & 12 & 129906275 &  & 5.84 & 0.06 & 0.40 & 0.14 &  \\
+  BID & ILMN\_2372413 & 22 & rs181405 & 22 & 18233000 &  & rs10888267 & 1 & 248059423 &  & 6.60 & 0.87 & 0.16 & 0.50 &  \\
+  C11ORF17 & ILMN\_1752988 & 11 & rs2568061 & 11 & 8886260 & C11ORF17 & rs6553184 & 4 & 189150656 &  & 5.66 & 1.15 & 0.04 & 0.54 &  \\
+  C13ORF18 & ILMN\_2196550 & 13 & rs2110603 & 16 & 6259852 &  & rs674754 & 13 & 46913416 & C13ORF18 & 6.66 & 0.28 & 0.28 & 0.22 &  \\
+  C13ORF18 & ILMN\_2196550 & 13 & rs674754 & 13 & 46913416 & C13ORF18 & rs6857876 & 4 & 153610164 &  & 3.87 & 0.38 & 0.50 & 0.43 &  \\
+  C14ORF173 & ILMN\_2393450 & 14 & rs11089825 & 22 & 37575398 &  & rs4983382 & 14 & 105189504 & C14ORF173 & 6.02 & 0.60 & 0.84 & 0.85 &  \\
+  C14ORF173 & ILMN\_2393450 & 14 & rs3935344 & 15 & 92276674 &  & rs4983382 & 14 & 105189504 & C14ORF173 & 5.98 & 0.31 & 0.28 & 0.24 &  \\
+  C14ORF173 & ILMN\_2393450 & 14 & rs4983382 & 14 & 105189504 & C14ORF173 & rs10754644 & 1 & 238724741 &  & 7.15 & 0.42 & 0.34 & 0.35 &  \\
+  C14ORF4 & ILMN\_1804396 & 14 & rs1293455 & 18 & 13819673 &  & rs2655991 & 14 & 77574438 &  & 4.87 &  &  &  &  \\
+  C14ORF4 & ILMN\_1804396 & 14 & rs2655991 & 14 & 77574438 &  & rs10972462 & 9 & 35427324 &  & 4.32 &  &  &  &  \\
+  C14ORF4 & ILMN\_1804396 & 14 & rs2655991 & 14 & 77574438 &  & rs6445340 & 3 & 63371601 &  & 4.40 &  &  &  &  \\
+  C14ORF4 & ILMN\_1804396 & 14 & rs2655991 & 14 & 77574438 &  & rs9787151 & 1 & 63179138 &  & 4.05 &  &  &  &  \\
+  C14ORF4 & ILMN\_1804396 & 14 & rs4793445 & 17 & 70416307 &  & rs2655991 & 14 & 77574438 &  & 3.85 &  &  &  &  \\
+  C14ORF4 & ILMN\_1804396 & 14 & rs6010061 & 22 & 51151724 &  & rs2655991 & 14 & 77574438 &  & 4.61 &  &  &  &  \\
+  C14ORF4 & ILMN\_1804396 & 14 & rs7245800 & 19 & 52083552 &  & rs2655991 & 14 & 77574438 &  & 4.69 &  &  &  &  \\
+  C17ORF60 & ILMN\_1747347 & 17 & rs9907897 & 17 & 63502633 &  & rs7405659 & 17 &  & C17ORF60 & 6.79 & 0.53 & 0.05 & 0.19 &  \\
+  C1ORF86 & ILMN\_1726989 & 1 & rs2334323 & 6 & 110577257 &  & rs2257182 & 1 & 2082566 & C1ORF86 & 5.90 & 0.01 & 0.50 & 0.13 &  \\
+  C1ORF86 & ILMN\_2097790 & 1 & rs2279474 & 18 & 46384412 &  & rs2460002 & 1 & 2119833 & C1ORF86 & 5.65 &  & 0.03 &  &  \\
+  C1ORF86 & ILMN\_2097790 & 1 & rs7188668 & 16 & 25711358 &  & rs2460002 & 1 & 2119833 & C1ORF86 & 5.59 & 0.29 & 0.50 & 0.37 &  \\
+  C21ORF57 & ILMN\_1795836 & 21 & rs4819271 & 21 & 48052838 &  & rs901964 & 12 & 48676038 & ZNF641 & 4.91 & 0.65 & 0.08 & 0.28 &  \\
+  C21ORF57 & ILMN\_1795836 & 21 & rs9978658 & 21 & 48027084 &  & rs11701361 & 21 & 47764477 &  & 9.42 & 6.08 & 16.36 & 21.67 & 0.263 \\
+  C5ORF4 & ILMN\_1728742 & 5 & rs1122762 & 18 & 45866512 &  & rs286595 & 5 & 154348552 & C5ORF4 & 5.55 & 0.72 & 0.04 & 0.27 &  \\
+  C8ORF59 & ILMN\_1653205 & 8 & rs12429804 & 13 & 36577930 &  & rs2896452 & 8 & 86102223 & C8ORF59 & 5.49 & 0.29 & 0.02 & 0.07 &  \\
+  C8ORF59 & ILMN\_1653205 & 8 & rs12454561 & 18 & 31272238 &  & rs2896452 & 8 & 86102223 & C8ORF59 & 5.45 & 0.31 &  &  &  \\
+  C8ORF59 & ILMN\_1653205 & 8 & rs2896452 & 8 & 86102223 & C8ORF59 & rs1004564 & 4 & 55242625 &  & 7.62 & 0.38 & 0.18 & 0.21 &  \\
+  C8ORF59 & ILMN\_1653205 & 8 & rs7152284 & 14 & 52273663 &  & rs2896452 & 8 & 86102223 & C8ORF59 & 5.67 & 2.18 & 0.07 & 1.33 &  \\
+  C8ORF59 & ILMN\_1653205 & 8 & rs8051751 & 16 & 7188323 &  & rs2896452 & 8 & 86102223 & C8ORF59 & 5.79 & 1.39 & 0.18 & 0.87 &  \\
+  C9ORF72 & ILMN\_1741881 & 9 & rs10122902 & 9 & 27556780 & C9ORF72 & rs2526698 & 1 & 242029101 &  & 6.36 & 0.96 & 0.01 & 0.37 &  \\
+  CABC1 & ILMN\_1731064 & 1 & rs12765847 & 10 & 4353908 &  & rs3738725 & 1 & 227174210 & CABC1 & 6.36 & 0.94 & 0.00 & 0.34 &  \\
+  CARD9 & ILMN\_1712532 & 9 & rs4266763 & 9 & 139289825 & INPP5E & rs684040 & 1 & 82128660 &  & 5.81 &  &  &  &  \\
+  CARD9 & ILMN\_1712532 & 9 & rs4573661 & 11 & 6026661 &  & rs4077515 & 9 & 139266496 & INPP5E & 6.61 & 0.09 & 0.86 & 0.42 &  \\
+  CAST & ILMN\_1717234 & 5 & rs1157079 & 20 & 6778978 &  & rs7733671 & 5 & 96000269 & CAST & 7.07 & 0.23 & 0.96 & 0.62 &  \\
+  CAST & ILMN\_1717234 & 5 & rs12463169 & 19 & 17321669 &  & rs7733671 & 5 & 96000269 & CAST & 5.73 & 0.02 & 2.85 & 1.75 &  \\
+  CAST & ILMN\_1717234 & 5 & rs12599264 & 16 & 81840122 &  & rs7733671 & 5 & 96000269 & CAST & 7.00 &  &  &  &  \\
+  CAST & ILMN\_1717234 & 5 & rs12719343 & 5 & 125369113 &  & rs7733671 & 5 & 96000269 & CAST & 7.68 & 0.36 & 1.57 & 1.20 & 29.369 \\
+  CAST & ILMN\_1717234 & 5 & rs1410575 & 9 & 78255630 &  & rs7733671 & 5 & 96000269 & CAST & 6.55 & 0.13 & 1.34 & 0.78 &  \\
+  CAST & ILMN\_1717234 & 5 & rs166444 & 8 & 78392770 &  & rs7733671 & 5 & 96000269 & CAST & 7.01 & 0.27 & 0.52 & 0.37 &  \\
+  CAST & ILMN\_1717234 & 5 & rs17648036 & 15 & 27311111 &  & rs7733671 & 5 & 96000269 & CAST & 7.81 & 0.97 & 0.03 & 0.41 &  \\
+  CAST & ILMN\_1717234 & 5 & rs17818702 & 11 & 86107920 &  & rs7733671 & 5 & 96000269 & CAST & 6.62 & 1.15 & 0.59 & 1.09 &  \\
+  CAST & ILMN\_1717234 & 5 & rs227402 & 14 & 70496867 &  & rs7733671 & 5 & 96000269 & CAST & 6.12 & 0.11 & 0.01 & 0.01 &  \\
+  CAST & ILMN\_1717234 & 5 & rs2822124 & 21 & 15166804 &  & rs7733671 & 5 & 96000269 & CAST & 6.87 &  &  &  &  \\
+  CAST & ILMN\_1717234 & 5 & rs3757155 & 6 & 136458593 &  & rs7733671 & 5 & 96000269 & CAST & 7.24 & 0.07 & 0.33 & 0.12 &  \\
+  CAST & ILMN\_1717234 & 5 & rs4503014 & 7 & 31149140 &  & rs7733671 & 5 & 96000269 & CAST & 5.88 & 0.92 & 1.56 & 1.72 &  \\
+  CAST & ILMN\_1717234 & 5 & rs7474890 & 10 & 59590078 &  & rs7733671 & 5 & 96000269 & CAST & 6.74 & 0.49 & 0.12 & 0.23 &  \\
+  CAST & ILMN\_1717234 & 5 & rs7733671 & 5 & 96000269 & CAST & rs10802643 & 1 & 238120177 &  & 7.42 & 0.75 & 0.78 & 0.93 &  \\
+  CAST & ILMN\_1717234 & 5 & rs7733671 & 5 & 96000269 & CAST & rs12650909 & 4 & 170192890 &  & 7.42 & 0.23 & 0.78 & 0.50 &  \\
+  CAST & ILMN\_1717234 & 5 & rs7733671 & 5 & 96000269 & CAST & rs2203733 & 2 & 224093101 &  & 6.07 & 0.22 & 0.87 & 0.54 &  \\
+  CAST & ILMN\_1717234 & 5 & rs7733671 & 5 & 96000269 & CAST & rs2641772 & 3 & 195531841 &  & 6.93 & 0.19 & 0.26 & 0.15 &  \\
+  CAT & ILMN\_1651705 & 11 & rs872311 & 18 & 66175386 &  & rs11032695 & 11 & 34447586 & CAT & 6.41 & 0.26 & 0.30 & 0.22 &  \\
+  CCDC88B & ILMN\_1772208 & 11 & rs2353203 & 19 & 17099980 &  & rs541207 & 11 & 64125142 & CCDC88B & 5.68 & 0.33 & 0.37 & 0.31 &  \\
+  CCDC88B & ILMN\_1772208 & 11 & rs694739 & 11 & 64097233 & CCDC88B & rs12771349 & 10 & 96998193 &  & 5.62 & 0.23 & 0.18 & 0.14 &  \\
+  CD36 & ILMN\_1784863 & 7 & rs3211834 & 7 & 80280117 &  & rs1254900 & 2 & 85816334 & VAMP8 & 6.93 & 0.15 & 0.01 & 0.02 &  \\
+  CD55 & ILMN\_1800540 & 1 & rs750801 & 11 & 76033374 &  & rs6700168 & 1 & 207502534 & CD55 & 5.09 & 0.08 & 0.03 & 0.02 &  \\
+  CD93 & ILMN\_1704730 & 20 & rs1884655 & 20 & 23074375 & CD93 & rs10255470 & 7 & 157182040 &  & 6.06 & 1.74 & 0.24 & 1.20 &  \\
+  CD93 & ILMN\_1704730 & 20 & rs1884655 & 20 & 23074375 & CD93 & rs4696726 & 4 & 7992632 &  & 5.71 & 0.13 & 0.80 & 0.42 &  \\
+  CD93 & ILMN\_1704730 & 20 & rs1884655 & 20 & 23074375 & CD93 & rs7622580 & 3 & 196721395 &  & 5.56 & 0.04 & 0.27 & 0.08 &  \\
+  CD93 & ILMN\_1704730 & 20 & rs1884655 & 20 & 23074375 & CD93 & rs838875 & 12 & 125145394 &  & 6.31 & 0.24 & 1.67 & 1.16 &  \\
+  CD93 & ILMN\_1704730 & 20 & rs1884655 & 20 & 23074375 & CD93 & rs9576388 & 13 & 38434472 &  & 7.88 & 0.71 & 0.22 & 0.45 &  \\
+  CD93 & ILMN\_1704730 & 20 & rs2868504 & 20 & 37771578 &  & rs1884655 & 20 & 23074375 & CD93 & 5.71 & 0.64 & 0.75 & 0.81 & 14.697 \\
+  CD93 & ILMN\_1704730 & 20 & rs4813479 & 20 & 23076914 & CD93 & rs10925747 & 1 & 238899903 &  & 7.43 &  &  &  &  \\
+  CD93 & ILMN\_1704730 & 20 & rs4813479 & 20 & 23076914 & CD93 & rs2873420 & 8 & 136500554 &  & 7.02 &  &  &  &  \\
+  CD93 & ILMN\_1704730 & 20 & rs4813479 & 20 & 23076914 & CD93 & rs4328531 & 18 & 74439542 &  & 6.13 &  &  &  &  \\
+  CD93 & ILMN\_1704730 & 20 & rs4813479 & 20 & 23076914 & CD93 & rs4789981 & 17 & 77264482 &  & 6.08 &  &  &  &  \\
+  CDC16 & ILMN\_2339796 & 13 & rs861544 & 14 & 104162263 &  & rs7324744 & 13 & 115008038 & CDC16 & 5.46 & 0.21 & 0.14 & 0.11 &  \\
+  CDK5R1 & ILMN\_1730928 & 17 & rs9905940 & 17 & 46614102 & HOXB2 & rs11655031 & 17 & 30833162 & CDK5R1 & 5.47 & 0.95 & 0.07 & 0.45 & 15.781 \\
+  CEACAM21 & ILMN\_1745949 & 19 & rs200609 & 20 & 51956250 &  & rs4803481 & 19 & 42066556 & CEACAM21 & 6.15 & 0.90 & 0.12 & 0.48 &  \\
+  CEACAM21 & ILMN\_1745949 & 19 & rs4803481 & 19 & 42066556 & CEACAM21 & rs2421050 & 5 & 158943044 &  & 6.67 & 2.16 & 0.16 & 1.44 &  \\
+  CEP192 & ILMN\_1703754 & 18 & rs6505780 & 18 & 13069782 & CEP192 & rs13132719 & 4 & 180265266 &  & 5.75 & 0.15 & 0.24 & 0.12 &  \\
+  CEP63 & ILMN\_1787808 & 3 & rs3825569 & 14 & 101350298 &  & rs13079012 & 3 & 134247706 & ANAPC13 & 6.36 & 0.23 & 0.10 & 0.09 &  \\
+  CES1 & ILMN\_2359945 & 16 & rs8192935 & 16 & 55861794 & CES1 & rs772788 & 2 & 235248562 &  & 5.65 &  &  &  &  \\
+  CHPT1 & ILMN\_2202940 & 12 & rs591967 & 13 & 38838122 &  & rs2695290 & 12 & 102087844 & CHPT1 & 5.74 & 0.72 & 0.20 & 0.44 &  \\
+  CHPT1 & ILMN\_2202940 & 12 & rs6539014 & 12 & 102277782 &  & rs867578 & 11 & 81937002 &  & 4.75 & 0.92 & 0.02 & 0.36 &  \\
+  CLEC12A & ILMN\_1663142 & 12 & rs429790 & 16 & 84471642 &  & rs7313235 & 12 & 10132283 & CLEC12A & 5.55 & 0.07 & 1.28 & 0.67 &  \\
+  CLEC12A & ILMN\_2403228 & 12 & rs7305054 & 12 & 10156646 &  & rs3903088 & 10 & 134236688 &  & 7.54 & 0.95 & 0.36 & 0.73 &  \\
+  CLTB & ILMN\_1674609 & 5 & rs17129799 & 11 & 96929337 &  & rs6863172 & 5 & 175595960 & CLTB & 5.55 &  & 0.27 &  &  \\
+  CNN2 & ILMN\_1770290 & 19 & rs3752237 & 19 & 1047161 & ABCA7 & rs169130 & 16 & 63121080 &  & 7.56 & 0.07 & 0.07 & 0.02 &  \\
+  CNN2 & ILMN\_1770290 & 19 & rs3752237 & 19 & 1047161 & ABCA7 & rs7336017 & 13 & 67713633 &  & 6.33 & 1.92 & 0.28 & 1.39 &  \\
+  CPSF1 & ILMN\_1654545 & 8 & rs4333645 & 8 & 145569535 &  & rs1455268 & 4 & 61738094 &  & 6.34 & 0.10 & 0.01 & 0.01 &  \\
+  CPVL & ILMN\_1682928 & 7 & rs12596791 & 16 & 26115562 &  & rs245884 & 7 & 29188475 & CPVL & 5.74 & 0.06 & 0.57 & 0.23 &  \\
+  CPVL & ILMN\_1682928 & 7 & rs2835998 & 21 & 39202070 &  & rs245884 & 7 & 29188475 & CPVL & 5.55 & 0.19 & 0.03 & 0.04 &  \\
+  CRIPT & ILMN\_1813256 & 2 & rs2131290 & 4 & 188859908 &  & rs1531133 & 2 & 46843631 & CRIPT & 5.47 & 0.28 & 0.10 & 0.12 &  \\
+  CRLS1 & ILMN\_1737685 & 20 & rs6139887 & 20 & 5986234 & CRLS1 & rs1473927 & 5 & 62406408 &  & 6.18 & 0.10 & 0.36 & 0.15 &  \\
+  CSTB & ILMN\_1761797 & 21 & rs9979356 & 21 & 45230974 &  & rs3761385 & 21 & 45198355 &  & 11.99 & 25.20 & 16.72 & 42.27 & 0.033 \\
+  CTNNA1 & ILMN\_1804854 & 5 & rs924943 & 18 & 69500505 &  & rs176382 & 5 & 138226767 & CTNNA1 & 5.74 & 0.02 & 0.41 & 0.11 &  \\
+  CTSC & ILMN\_1696347 & 11 & rs2457684 & 11 & 88139983 & CTSC & rs7079264 & 10 & 108679892 &  & 5.67 & 0.92 & 0.74 & 1.03 &  \\
+  CTSC & ILMN\_1696347 & 11 & rs5752236 & 22 & 26250645 &  & rs7128352 & 11 & 88087357 & CTSC & 5.84 & 0.49 & 0.80 & 0.73 &  \\
+  CTSC & ILMN\_2242463 & 11 & rs7930237 & 11 & 88117962 &  & rs556895 & 11 & 88077479 &  & 7.16 & 18.76 & 15.06 & 33.53 & 0.040 \\
+  CWF19L1 & ILMN\_1651886 & 10 & rs7108734 & 11 & 11456027 &  & rs12784396 & 10 & 102027407 & CWF19L1 & 5.42 & 0.21 & 0.01 & 0.03 &  \\
+  CYBRD1 & ILMN\_1712305 & 2 & rs2592948 & 4 & 129994690 &  & rs888427 & 2 & 172368120 & CYBRD1 & 5.89 & 0.23 & 0.53 & 0.34 &  \\
+  CYBRD1 & ILMN\_1712305 & 2 & rs7852475 & 9 & 140698856 &  & rs888427 & 2 & 172368120 & CYBRD1 & 5.68 & 0.20 & 0.02 & 0.04 &  \\
+  CYBRD1 & ILMN\_2087692 & 2 & rs11257679 & 10 & 12318284 &  & rs888427 & 2 & 172368120 & CYBRD1 & 5.81 & 0.39 & 1.87 & 1.47 &  \\
+  CYBRD1 & ILMN\_2087692 & 2 & rs6137908 & 20 & 23344590 &  & rs888427 & 2 & 172368120 & CYBRD1 & 5.53 & 0.05 & 0.83 & 0.36 &  \\
+  CYBRD1 & ILMN\_2087692 & 2 & rs888427 & 2 & 172368120 & CYBRD1 & rs7591849 & 2 & 160112881 &  & 5.85 & 0.87 & 0.10 & 0.44 & 12.255 \\
+  CYP27A1 & ILMN\_1704985 & 2 & rs6021982 & 20 & 36571928 &  & rs933994 & 2 & 219650616 & CYP27A1 & 5.42 & 0.29 & 0.86 & 0.60 &  \\
+  DAB2 & ILMN\_2128428 & 5 & rs7778910 & 7 & 110451383 &  & rs835223 & 5 & 39381357 & DAB2 & 5.44 & 0.48 & 0.41 & 0.44 &  \\
+  DCAKD & ILMN\_1811648 & 17 & rs9900173 & 17 & 43111688 &  & rs1343244 & 6 & 82076988 &  & 9.12 & 0.00 & 0.58 & 0.14 &  \\
+  DDT & ILMN\_1690982 & 22 & rs5760102 & 22 & 24248761 & DDT & rs2378341 & 3 & 187475208 &  & 5.62 & 0.64 & 0.25 & 0.42 &  \\
+  DDX58 & ILMN\_1797001 & 9 & rs4937097 & 11 & 125962645 &  & rs7042042 & 9 & 32451144 &  & 5.31 & 0.61 & 0.29 & 0.44 &  \\
+  DEM1 & ILMN\_1783996 & 1 & rs10120023 & 9 & 137810259 & COQ10A & rs2519515 & 7 & 88204888 &  & 5.47 & 0.08 & 0.41 & 0.16 &  \\
+  DEM1 & ILMN\_1783996 & 1 & rs12363827 & 11 & 106703727 &  & rs10120023 & 9 & 137810259 & COQ10A & 6.39 & 0.77 & 0.02 & 0.29 &  \\
+  DHRS9 & ILMN\_1733998 & 2 & rs1519956 & 12 & 89468283 &  & rs7566044 & 2 & 169960422 & DHRS9 & 6.00 & 0.06 & 1.17 & 0.58 &  \\
+  DHRS9 & ILMN\_1733998 & 2 & rs1528529 & 7 & 147132505 &  & rs7566044 & 2 & 169960422 & DHRS9 & 6.48 & 0.37 & 0.34 & 0.32 &  \\
+  DHRS9 & ILMN\_2384181 & 2 & rs2831914 & 21 & 29959453 &  & rs2161037 & 2 & 169893419 & DHRS9 & 5.51 & 0.88 & 0.04 & 0.37 &  \\
+  DHRS9 & ILMN\_2384181 & 2 & rs7661304 & 4 & 187776431 &  & rs2161037 & 2 & 169893419 & DHRS9 & 7.64 & 0.05 & 0.11 & 0.03 &  \\
+  DIP2B & ILMN\_1755589 & 12 & rs11080134 & 17 & 29161503 &  & rs11169322 & 12 & 50610976 & LASS5 & 4.65 & 0.32 & 0.05 & 0.10 &  \\
+  DIP2B & ILMN\_1755589 & 12 & rs11169335 & 12 & 50636364 & LASS5 & rs2872008 & 7 & 153134888 &  & 4.87 &  & 0.58 &  &  \\
+  DIP2B & ILMN\_1755589 & 12 & rs338585 & 19 & 41711815 &  & rs7134595 & 12 & 50730458 & LASS5 & 5.31 & 0.30 & 0.22 & 0.19 &  \\
+  DIP2B & ILMN\_1755589 & 12 & rs7134595 & 12 & 50730458 & LASS5 & rs1808634 & 8 & 61971140 &  & 4.40 & 0.37 &  &  &  \\
+  DIP2B & ILMN\_1755589 & 12 & rs7312252 & 12 & 50744171 & LASS5 & rs4532958 & 10 & 115214154 &  & 5.03 & 0.09 & 0.02 & 0.01 &  \\
+  DIP2B & ILMN\_1755589 & 12 & rs871257 & 12 & 117994348 &  & rs12427378 & 12 & 51074199 & LASS5 & 5.92 & 0.48 & 0.00 & 0.11 & 66.920 \\
+  DNAJB6 & ILMN\_1793770 & 7 & rs2286842 & 7 & 157216093 &  & rs3779589 & 7 & 157163614 & DNAJB6 & 5.79 & 0.23 & 1.45 & 0.97 & 0.052 \\
+  DPH3 & ILMN\_2349610 & 3 & rs12232308 & 15 & 93409054 &  & rs1566972 & 3 & 16320360 & DPH3 & 6.17 & 1.58 & 0.27 & 1.12 &  \\
+  ECGF1 & ILMN\_2109708 & 22 & rs140522 & 22 & 50971266 & ECGF1 & rs4891884 & 18 & 64004670 &  & 4.81 & 0.15 & 1.18 & 0.70 &  \\
+  ECHDC2 & ILMN\_1671568 & 1 & rs4234091 & 2 & 241911027 &  & rs11206043 & 1 & 53402552 & ECHDC2 & 6.19 & 0.22 & 0.35 & 0.22 &  \\
+  ECHDC2 & ILMN\_1671568 & 1 & rs5992637 & 22 & 17675900 &  & rs11206043 & 1 & 53402552 & ECHDC2 & 5.58 & 0.64 & 0.16 & 0.35 &  \\
+  EHD4 & ILMN\_1720083 & 15 & rs10403312 & 19 & 53244938 &  & rs1048166 & 15 & 42192040 & EHD4 & 6.98 & 0.90 & 0.47 & 0.79 &  \\
+  EIF2B2 & ILMN\_1713380 & 14 & rs6567288 & 18 & 60218334 &  & rs175450 & 14 & 75590340 & EIF2B2 & 5.56 & 0.23 & 0.11 & 0.10 &  \\
+  EIF5A & ILMN\_1794522 & 17 & rs7216490 & 17 & 7221707 & EIF5A & rs1269096 & 14 & 99603119 &  & 5.44 & 0.56 & 0.08 & 0.24 &  \\
+  EIF5A & ILMN\_1794522 & 17 & rs7216490 & 17 & 7221707 & EIF5A & rs1553474 & 2 & 49359676 &  & 5.55 & 0.28 & 0.59 & 0.41 &  \\
+  EIF5A & ILMN\_1794522 & 17 & rs7216490 & 17 & 7221707 & EIF5A & rs2197210 & 8 & 129624067 &  & 6.36 & 0.08 & 0.05 & 0.02 &  \\
+  EIF5A & ILMN\_1794522 & 17 & rs7216490 & 17 & 7221707 & EIF5A & rs4471434 & 11 & 126387391 &  & 5.52 & 0.05 & 1.12 & 0.53 &  \\
+  EMR2 & ILMN\_2353633 & 19 & rs2827076 & 21 & 23196249 &  & rs9305048 & 19 & 14879034 & EMR2 & 6.51 & 0.36 & 0.04 & 0.11 &  \\
+  EMR2 & ILMN\_2353633 & 19 & rs6132112 & 20 & 18761714 &  & rs9305048 & 19 & 14879034 & EMR2 & 5.56 & 0.45 & 0.40 & 0.41 &  \\
+  EMR2 & ILMN\_2353633 & 19 & rs9305048 & 19 & 14879034 & EMR2 & rs3007765 & 13 & 102480759 &  & 6.03 &  &  &  &  \\
+  EPHX2 & ILMN\_1709237 & 8 & rs1107764 & 11 & 127909396 &  & rs13269963 & 8 & 27400604 & EPHX2 & 5.70 & 0.20 & 0.58 & 0.35 &  \\
+  ERICH1 & ILMN\_1731001 & 8 & rs10894861 & 11 & 134611176 &  & rs12115088 & 8 & 578742 & ERICH1 & 5.43 & 0.25 & 1.20 & 0.81 &  \\
+  ERICH1 & ILMN\_1731001 & 8 & rs5766218 & 22 & 45337329 &  & rs4735900 & 8 & 607161 & ERICH1 & 6.11 & 0.20 & 0.11 & 0.09 &  \\
+  ERICH1 & ILMN\_1731001 & 8 & rs726145 & 18 & 31187910 &  & rs12115088 & 8 & 578742 & ERICH1 & 5.65 & 0.29 & 0.04 & 0.08 &  \\
+  ERICH1 & ILMN\_2104696 & 8 & rs4735895 & 8 & 600729 & ERICH1 & rs1517297 & 4 & 182786760 &  & 5.63 & 0.67 & 1.03 & 1.06 &  \\
+  EXOC3 & ILMN\_1789419 & 5 & rs187076 & 10 & 55228462 &  & rs12188164 & 5 & 428236 & EXOC3 & 6.83 & 0.74 & 0.19 & 0.44 &  \\
+  FAHD1 & ILMN\_2246661 & 16 & rs1560104 & 16 & 12708208 &  & rs344363 & 16 & 1972548 & FAHD1 & 5.61 &  & 1.38 &  & 10.736 \\
+  FCN1 & ILMN\_1668063 & 9 & rs12580388 & 12 & 129591144 &  & rs10120023 & 9 & 137810259 & COQ10A & 6.33 & 0.27 & 0.30 & 0.23 &  \\
+  FEZ2 & ILMN\_1739586 & 2 & rs2356400 & 19 & 44321776 &  & rs13406184 & 2 & 36791226 & FEZ2 & 5.78 & 0.14 & 0.33 & 0.16 &  \\
+  FEZ2 & ILMN\_1739586 & 2 & rs969010 & 4 & 159963132 &  & rs11691600 & 2 & 36810133 & FEZ2 & 6.59 & 0.14 & 0.28 & 0.14 &  \\
+  FGD2 & ILMN\_2115005 & 6 & rs4803848 & 19 & 46205050 &  & rs831486 & 6 & 37001267 & FGD2 & 5.69 & 0.12 & 0.25 & 0.11 &  \\
+  FGD2 & ILMN\_2115005 & 6 & rs902634 & 10 & 133943951 &  & rs831489 & 6 & 36999682 & FGD2 & 5.49 & 1.20 & 0.11 & 0.66 &  \\
+  FLJ20489 & ILMN\_1778144 & 12 & rs17615703 & 12 & 117036766 &  & rs3782908 & 12 & 48169526 & FLJ20489 & 5.81 & 0.06 & 0.70 & 0.29 & 68.867 \\
+  FLJ20489 & ILMN\_1778144 & 12 & rs3782908 & 12 & 48169526 & FLJ20489 & rs897511 & 4 & 167695661 &  & 5.53 & 0.03 & 0.11 & 0.02 &  \\
+  FLJ20489 & ILMN\_1778144 & 12 & rs4792199 & 17 & 7992118 &  & rs3782908 & 12 & 48169526 & FLJ20489 & 5.74 & 0.19 & 0.02 & 0.04 &  \\
+  FLJ20489 & ILMN\_1778144 & 12 & rs4984440 & 15 & 97033129 &  & rs3782908 & 12 & 48169526 & FLJ20489 & 6.49 & 0.31 & 0.47 & 0.36 &  \\
+  FLJ20489 & ILMN\_1778144 & 12 & rs7204135 & 16 & 50626195 &  & rs3782908 & 12 & 48169526 & FLJ20489 & 6.90 & 0.38 & 0.17 & 0.21 &  \\
+  FLJ20718 & ILMN\_1763663 & 16 & rs9325634 & 21 & 43818790 &  & rs2287197 & 16 & 50106594 & FLJ20718 & 6.04 & 0.14 & 0.95 & 0.53 &  \\
+  FLJ43093 & ILMN\_2123450 & 6 & rs17112712 & 14 & 107276627 &  & rs6906101 & 6 & 36667610 & FLJ43093 & 5.48 & 0.39 & 0.06 & 0.13 &  \\
+  FLJ43093 & ILMN\_2123450 & 6 & rs6906101 & 6 & 36667610 & FLJ43093 & rs13214069 & 6 & 32705248 &  & 5.44 & 0.00 & 0.64 & 0.18 & 3.962 \\
+  FN3KRP & ILMN\_1652333 & 17 & rs898095 & 17 & 80890638 &  & rs9892064 & 17 & 80827903 &  & 16.16 & 28.24 & 29.39 & 59.95 & 0.063 \\
+  FUCA1 & ILMN\_1752728 & 1 & rs4971478 & 2 & 1346063 &  & rs12744386 & 1 & 24168019 & FUCA1 & 6.41 & 0.01 & 0.30 & 0.06 &  \\
+  FXYD5 & ILMN\_2309848 & 19 & rs1633921 & 19 & 35695200 &  & rs788178 & 13 & 98328559 &  & 3.70 & 0.09 & 0.41 & 0.17 &  \\
+  FXYD5 & ILMN\_2309848 & 19 & rs17398183 & 20 & 55609148 &  & rs2285515 & 19 & 35660450 & FXYD5 & 6.58 & 0.03 & 0.48 & 0.15 &  \\
+  FXYD5 & ILMN\_2309848 & 19 & rs2285515 & 19 & 35660450 & FXYD5 & rs11739594 & 5 & 141709563 &  & 5.70 & 0.07 & 0.17 & 0.05 &  \\
+  FXYD5 & ILMN\_2309848 & 19 & rs2285515 & 19 & 35660450 & FXYD5 & rs13067700 & 3 & 95331048 &  & 6.00 & 0.09 & 0.51 & 0.22 &  \\
+  FXYD5 & ILMN\_2309848 & 19 & rs2285515 & 19 & 35660450 & FXYD5 & rs17036504 & 2 & 47567329 &  & 6.10 & 0.28 &  &  &  \\
+  G3BP2 & ILMN\_2381758 & 4 & rs10230232 & 7 & 29390239 &  & rs1553985 & 4 & 76554604 &  & 5.19 & 0.08 & 0.37 & 0.14 &  \\
+  GAA & ILMN\_2410783 & 17 & rs11150847 & 17 & 78153130 &  & rs12602462 & 17 & 78146016 &  & 13.91 & 19.98 & 12.99 & 32.60 & 0.007 \\
+  GAA & ILMN\_2410783 & 17 & rs8068856 & 17 & 78100731 & GAA & rs10902506 & 12 & 132678089 &  & 5.65 & 0.11 & 0.39 & 0.17 &  \\
+  GAPT & ILMN\_1675191 & 5 & rs10070522 & 5 & 57786110 & GAPT & rs7605821 & 2 & 235695228 &  & 5.85 & 0.01 & 0.78 & 0.28 &  \\
+  GAPT & ILMN\_1675191 & 5 & rs7082031 & 10 & 128038717 &  & rs10070522 & 5 & 57786110 & GAPT & 5.72 & 0.26 & 0.11 & 0.11 &  \\
+  GATS & ILMN\_1699631 & 7 & rs1147447 & 14 & 66460742 &  & rs2950520 & 7 & 99827148 & GATS & 5.47 & 0.83 & 0.63 & 0.87 &  \\
+  GATS & ILMN\_1699631 & 7 & rs2425256 & 20 & 35056572 &  & rs2950520 & 7 & 99827148 & GATS & 6.22 &  & 0.42 &  &  \\
+  GDPD3 & ILMN\_1774901 & 16 & rs3809624 & 16 & 30102802 & GDPD3 & rs2197465 & 14 & 48572632 &  & 6.57 & 0.38 & 0.35 & 0.33 &  \\
+  GDPD3 & ILMN\_1774901 & 16 & rs7204270 & 16 & 30156963 & GDPD3 & rs1015111 & 4 & 128972357 &  & 5.86 & 0.55 & 0.09 & 0.24 &  \\
+  GNLY & ILMN\_1790692 & 2 & rs4145072 & 13 & 110899955 &  & rs7577293 & 2 & 85935282 & GNLY & 5.78 & 0.02 & 0.45 & 0.13 &  \\
+  GPN3 & ILMN\_3239426 & 12 & rs7198646 & 16 & 26084476 &  & rs7960552 & 12 & 111164237 & GPN3 & 5.72 &  &  &  &  \\
+  GPR162 & ILMN\_1730816 & 12 & rs1860563 & 16 & 6478898 &  & rs2707210 & 12 & 6902002 & GPR162 & 5.49 & 0.36 & 0.46 & 0.39 &  \\
+  GPR162 & ILMN\_1730816 & 12 & rs2272500 & 12 & 79685913 &  & rs2707210 & 12 & 6902002 & GPR162 & 5.07 & 0.25 & 0.03 & 0.06 & 72.784 \\
+  GPR162 & ILMN\_1730816 & 12 & rs2707210 & 12 & 6902002 & GPR162 & rs4740848 & 9 & 6554558 &  & 5.47 & 0.25 & 0.06 & 0.07 &  \\
+  GPR162 & ILMN\_1730816 & 12 & rs2707210 & 12 & 6902002 & GPR162 & rs9827054 & 3 & 188880113 &  & 6.21 & 0.96 & 0.06 & 0.44 &  \\
+  GPR177 & ILMN\_1660549 & 1 & rs11057383 & 12 & 124369421 &  & rs12065581 & 1 & 68732819 & GPR177 & 5.45 & 0.72 & 0.67 & 0.81 &  \\
+  GPR177 & ILMN\_1660549 & 1 & rs12527241 & 6 & 120468039 &  & rs12065581 & 1 & 68732819 & GPR177 & 5.76 & 0.17 & 0.40 & 0.22 &  \\
+  GPR177 & ILMN\_1660549 & 1 & rs12532999 & 7 & 127939793 &  & rs12065581 & 1 & 68732819 & GPR177 & 6.50 & 0.79 & 1.43 & 1.50 &  \\
+  GPR177 & ILMN\_1660549 & 1 & rs725613 & 16 & 11169683 &  & rs12065581 & 1 & 68732819 & GPR177 & 5.43 & 0.31 & 0.11 & 0.13 &  \\
+  GPR177 & ILMN\_1660549 & 1 & rs9575097 & 13 & 82986268 &  & rs12065581 & 1 & 68732819 & GPR177 & 6.04 & 0.95 & 0.21 & 0.60 &  \\
+  GPR177 & ILMN\_2283325 & 1 & rs6566669 & 18 & 70506011 &  & rs12065581 & 1 & 68732819 & GPR177 & 5.86 & 0.24 & 0.34 & 0.23 &  \\
+  GPR177 & ILMN\_2283325 & 1 & rs9290426 & 3 & 171399321 &  & rs12065581 & 1 & 68732819 & GPR177 & 6.50 & 0.01 & 0.24 & 0.04 &  \\
+  GSDMB & ILMN\_2347193 & 17 & rs11557467 & 17 & 38028634 & GSDMB & rs4965745 & 15 & 101508261 &  & 5.88 & 0.68 & 0.20 & 0.41 &  \\
+  GSTM1 & ILMN\_2391861 & 1 & rs12248673 & 10 & 53192833 &  & rs11101992 & 1 & 110266754 & GSTM1 & 6.11 & 0.27 & 0.19 & 0.16 &  \\
+  GSTM1 & ILMN\_2391861 & 1 & rs1547574 & 13 & 85344527 &  & rs11101992 & 1 & 110266754 & GSTM1 & 5.91 & 0.27 & 1.14 & 0.79 &  \\
+  GSTM2 & ILMN\_2201580 & 1 & rs6492807 & 13 & 96159560 &  & rs3754446 & 1 & 110253241 & GSTM1 & 6.77 &  &  &  &  \\
+  H1F0 & ILMN\_1757467 & 22 & rs139898 & 22 & 38399979 &  & rs4853333 & 2 & 77919015 &  & 6.36 & 0.52 & 0.66 & 0.65 &  \\
+  H1F0 & ILMN\_1757467 & 22 & rs139898 & 22 & 38399979 &  & rs6497007 & 15 & 85877017 &  & 6.52 & 0.27 & 0.31 & 0.23 &  \\
+  H1F0 & ILMN\_1757467 & 22 & rs139898 & 22 & 38399979 &  & rs9983949 & 21 & 19532546 &  & 5.70 & 0.25 & 0.48 & 0.32 &  \\
+  HBG1 & ILMN\_1796678 & 11 & rs11078523 & 17 & 4523167 &  & rs2855039 & 11 & 5271671 & HBG2 & 5.47 & 0.00 & 0.66 & 0.19 &  \\
+  HBG1 & ILMN\_1796678 & 11 & rs12975066 & 19 & 35723501 &  & rs2855039 & 11 & 5271671 & HBG2 & 5.98 & 0.15 & 0.24 & 0.12 &  \\
+  HBG1 & ILMN\_1796678 & 11 & rs2855039 & 11 & 5271671 & HBG2 & rs12042181 & 1 & 213088494 & LQK1 & 6.78 & 0.08 & 0.52 & 0.21 &  \\
+  HBG1 & ILMN\_1796678 & 11 & rs2855039 & 11 & 5271671 & HBG2 & rs12503379 & 4 & 141533832 &  & 6.42 & 0.01 & 0.46 & 0.11 &  \\
+  HBG2 & ILMN\_2084825 & 11 & rs11078523 & 17 & 4523167 &  & rs16912979 & 11 & 5309695 & HBG2 & 6.06 & 0.01 & 0.41 & 0.10 &  \\
+  HBG2 & ILMN\_2084825 & 11 & rs12975066 & 19 & 35723501 &  & rs2855039 & 11 & 5271671 & HBG2 & 5.77 & 0.08 & 0.13 & 0.05 &  \\
+  HBG2 & ILMN\_2084825 & 11 & rs2855039 & 11 & 5271671 & HBG2 & rs12042181 & 1 & 213088494 & LQK1 & 6.84 & 0.06 & 0.54 & 0.21 &  \\
+  HBG2 & ILMN\_2084825 & 11 & rs2855039 & 11 & 5271671 & HBG2 & rs12503379 & 4 & 141533832 &  & 5.98 & 0.00 & 0.46 & 0.10 &  \\
+  HDAC7 & ILMN\_3266186 & 12 & rs2109029 & 16 & 6036851 &  & rs4760636 & 12 & 48173352 & HDAC7 & 5.75 &  &  &  &  \\
+  HEBP1 & ILMN\_1802557 & 12 & rs3782567 & 12 & 13145613 & HEBP1 & rs17686635 & 8 & 135220622 &  & 5.98 & 0.15 & 0.59 & 0.32 &  \\
+  HEXDC & ILMN\_1741180 & 17 & rs1942719 & 18 & 71237270 &  & rs7213057 & 17 & 80378939 & HEXDC & 5.81 & 1.61 & 0.34 & 1.22 &  \\
+  HLA-DRA & ILMN\_2157441 & 6 & rs4899635 & 14 & 77532672 &  & rs7192 & 6 & 32411646 & HLA-DRB6 & 5.94 & 0.90 & 0.16 & 0.52 &  \\
+  HLA-F & ILMN\_1762861 & 6 & rs11660982 & 18 & 75467313 &  & rs2523404 & 6 & 29695713 & HLA-H & 5.69 & 1.00 & 0.47 & 0.86 &  \\
+  HMBOX1 & ILMN\_1720059 & 8 & rs12435486 & 14 & 98670849 &  & rs7837237 & 8 & 28876221 & HMBOX1 & 6.54 & 0.92 & 1.11 & 1.34 &  \\
+  HMBOX1 & ILMN\_1720059 & 8 & rs2837803 & 21 & 42112794 &  & rs4732890 & 8 & 28751381 & HMBOX1 & 6.62 & 0.05 & 1.01 & 0.46 &  \\
+  HMBOX1 & ILMN\_1720059 & 8 & rs4765451 & 12 & 127237464 &  & rs8180944 & 8 & 28904086 & HMBOX1 & 5.80 & 0.39 & 3.13 & 2.52 &  \\
+  HMBOX1 & ILMN\_1720059 & 8 & rs587639 & 8 & 132725731 &  & rs7837237 & 8 & 28876221 & HMBOX1 & 6.58 & 0.55 & 0.34 & 0.44 & 103.850 \\
+  HMBOX1 & ILMN\_1720059 & 8 & rs8180944 & 8 & 28904086 & HMBOX1 & rs4553956 & 3 & 189533772 &  & 6.88 & 3.38 & 0.03 & 2.20 &  \\
+  HMBOX1 & ILMN\_1720059 & 8 & rs8180944 & 8 & 28904086 & HMBOX1 & rs7810884 & 7 & 158276926 &  & 6.12 & 0.34 & 0.66 & 0.52 &  \\
+  HMBOX1 & ILMN\_1720059 & 8 & rs9521666 & 13 & 110897444 &  & rs8180944 & 8 & 28904086 & HMBOX1 & 5.45 & 0.67 & 0.26 & 0.45 &  \\
+  HNRPH1 & ILMN\_2101920 & 5 & rs6894268 & 5 & 179032488 &  & rs4700810 & 5 & 178991794 &  & 15.38 & 8.55 & 3.01 & 10.37 & 0.041 \\
+  HSPC157 & ILMN\_3194087 & 1 & rs555812 & 16 & 88882257 &  & rs4654783 & 1 & 22439520 & HSPC157 & 5.51 &  &  &  &  \\
+  HSPC157 & ILMN\_3194087 & 1 & rs6063164 & 20 & 46486900 &  & rs4654783 & 1 & 22439520 & HSPC157 & 6.51 &  &  &  &  \\
+  HSPC157 & ILMN\_3194087 & 1 & rs662739 & 12 & 121229893 &  & rs4654783 & 1 & 22439520 & HSPC157 & 6.61 &  &  &  &  \\
+  HSPC157 & ILMN\_3194087 & 1 & rs7088558 & 10 & 101884937 & CWF19L1 & rs4654783 & 1 & 22439520 & HSPC157 & 6.48 &  &  &  &  \\
+  IL32 & ILMN\_1778010 & 16 & rs1554999 & 16 & 3115628 & IL32 & rs4759890 & 12 & 131757163 &  & 6.90 & 0.19 & 0.50 & 0.29 &  \\
+  IL32 & ILMN\_2368530 & 16 & rs765044 & 19 & 2560423 &  & rs1554999 & 16 & 3115628 & IL32 & 5.53 & 0.69 & 0.23 & 0.44 &  \\
+  INPP5E & ILMN\_1811301 & 9 & rs8044524 & 16 & 81603771 &  & rs1127152 & 9 & 139335599 & INPP5E & 5.58 & 1.46 & 0.84 & 1.55 &  \\
+  JAZF1 & ILMN\_1682727 & 7 & rs757355 & 12 & 47970693 &  & rs849341 & 7 & 28288174 &  & 8.16 & 0.02 & 0.26 & 0.05 &  \\
+  KCNJ15 & ILMN\_1675756 & 21 & rs2186344 & 21 & 39606769 & KCNJ15 & rs424299 & 11 & 5570771 &  & 5.64 & 0.65 & 0.13 & 0.33 &  \\
+  KIR2DS5 & ILMN\_1691803 & 19 & rs649216 & 19 & 55324635 & KIR2DL1 & rs6419960 & 4 & 189055298 &  & 4.74 & 0.46 & 0.89 & 0.77 &  \\
+  KTELC1 & ILMN\_1811104 & 3 & rs4349034 & 13 & 84597119 &  & rs727905 & 3 & 119119433 & KTELC1 & 5.53 & 0.08 & 0.80 & 0.37 &  \\
+  KTELC1 & ILMN\_1811104 & 3 & rs6815953 & 4 & 183109012 &  & rs6414283 & 3 & 119195913 & KTELC1 & 5.45 & 0.64 & 0.08 & 0.28 &  \\
+  L3MBTL2 & ILMN\_2336109 & 22 & rs4822006 & 22 & 41519362 & L3MBTL2 & rs1294338 & 1 & 233438952 &  & 5.88 & 0.33 & 0.04 & 0.09 &  \\
+  LAP3 & ILMN\_1683792 & 4 & rs7042087 & 9 & 132602868 &  & rs7658240 & 4 & 17588950 & LAP3 & 5.72 & 0.24 & 0.47 & 0.31 &  \\
+  LAX1 & ILMN\_1769782 & 1 & rs1891432 & 1 & 203877662 &  & rs10900520 & 1 & 203780591 &  & 19.16 & 18.60 & 11.22 & 29.24 & 0.097 \\
+  LDLRAP1 & ILMN\_1809040 & 1 & rs1552032 & 15 & 59971635 &  & rs6687605 & 1 & 25889632 & LDLRAP1 & 6.00 &  &  &  &  \\
+  LGALS9 & ILMN\_2412214 & 17 & rs12450521 & 17 & 26083392 &  & rs11749727 & 5 & 179608360 &  & 5.16 & 0.35 & 0.40 & 0.34 &  \\
+  LILRA5 & ILMN\_2357419 & 19 & rs3859532 & 19 & 54827248 & LILRA5 & rs714789 & 18 & 71561497 &  & 6.13 & 0.23 & 0.03 & 0.05 &  \\
+  LINS1 & ILMN\_2338197 & 15 & rs11247226 & 15 & 101120963 & LINS1 & rs1278387 & 10 & 127804531 &  & 5.89 &  & 0.13 &  &  \\
+  LRRC25 & ILMN\_2150196 & 19 & rs6009951 & 22 & 51151350 &  & rs8101804 & 19 & 18496107 & LRRC25 & 5.68 & 0.11 & 0.35 & 0.15 &  \\
+  LY86 & ILMN\_1807825 & 6 & rs977785 & 6 & 6588881 & LY86 & rs1543675 & 1 & 78946879 &  & 5.61 & 0.13 & 0.15 & 0.07 &  \\
+  LYZ & ILMN\_1815205 & 12 & rs2168029 & 12 & 69734641 & LYZ & rs11981725 & 7 & 154137150 &  & 5.95 & 0.15 & 0.03 & 0.03 &  \\
+  LYZ & ILMN\_2162972 & 12 & rs177820 & 18 & 77276964 &  & rs2168029 & 12 & 69734641 & LYZ & 5.71 & 0.49 & 0.03 & 0.16 &  \\
+  LYZ & ILMN\_2162972 & 12 & rs2168029 & 12 & 69734641 & LYZ & rs2253135 & 9 & 130319560 &  & 6.31 & 0.61 & 0.36 & 0.49 &  \\
+  MAD1L1 & ILMN\_2358069 & 7 & rs7783715 & 7 & 1923385 & MAD1L1 & rs6414306 & 3 & 127011798 &  & 5.62 & 0.25 & 0.88 & 0.59 &  \\
+  MAD2L1BP & ILMN\_1694711 & 6 & rs7983718 & 13 & 103203146 &  & rs1096699 & 6 & 43528441 & MAD2L1BP & 5.93 & 0.63 & 1.11 & 1.09 &  \\
+  MAP1LC3A & ILMN\_1776188 & 20 & rs974607 & 21 & 29435869 &  & rs6060034 & 20 & 33351864 & MAP1LC3A & 5.78 & 1.18 &  &  &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs10869600 & 9 & 78225815 &  & rs13069559 & 3 & 152187431 & MBNL1 & 7.96 & 0.79 & 0.27 & 0.54 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs11053043 & 12 & 9932070 &  & rs13069559 & 3 & 152187431 & MBNL1 & 6.70 & 0.08 & 2.21 & 1.37 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs1164596 & 13 & 97100681 &  & rs13069559 & 3 & 152187431 & MBNL1 & 7.38 & 1.43 & 0.63 & 1.34 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs11981513 & 7 & 94648239 &  & rs13069559 & 3 & 152187431 & MBNL1 & 7.71 & 0.43 & 5.36 & 4.58 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs16864367 & 3 & 152234166 &  & rs13079208 & 3 & 152116652 &  & 13.49 & 16.25 & 24.74 & 41.56 & 0.118 \\
+  MBNL1 & ILMN\_2313158 & 3 & rs2030926 & 6 & 114067127 &  & rs13069559 & 3 & 152187431 & MBNL1 & 7.10 & 0.91 & 5.80 & 5.53 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs218671 & 17 & 6604708 &  & rs13069559 & 3 & 152187431 & MBNL1 & 7.63 & 0.62 & 5.82 & 5.23 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs2213360 & 22 & 34291750 &  & rs13069559 & 3 & 152187431 & MBNL1 & 6.05 & 0.52 & 0.72 & 0.70 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs2305802 & 19 & 16038535 &  & rs13069559 & 3 & 152187431 & MBNL1 & 6.94 & 1.67 &  &  &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs2614467 & 14 & 99770138 &  & rs13069559 & 3 & 152187431 & MBNL1 & 5.74 & 4.13 & 2.22 & 5.30 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs4392535 & 4 & 41513423 &  & rs13069559 & 3 & 152187431 & MBNL1 & 8.39 & 0.02 & 4.33 & 3.02 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs4735830 & 8 & 895841 &  & rs13069559 & 3 & 152187431 & MBNL1 & 6.74 & 0.32 & 4.21 & 3.38 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs4939558 & 18 & 46278591 &  & rs1522374 & 3 & 152235530 &  & 7.72 & 0.03 & 0.27 & 0.07 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs6128397 & 20 & 57253132 &  & rs16864367 & 3 & 152234166 &  & 7.22 & 1.34 & 1.15 & 1.73 &  \\
+  MBNL1 & ILMN\_2313158 & 3 & rs7710738 & 5 & 22101322 &  & rs13069559 & 3 & 152187431 & MBNL1 & 7.92 & 2.55 & 7.89 & 9.28 &  \\
+  MBP & ILMN\_2331544 & 18 & rs6079849 & 20 & 15462611 &  & rs2051344 & 18 & 74715653 & MBP & 6.26 & 0.10 & 0.03 & 0.02 &  \\
+  MBP & ILMN\_2398939 & 18 & rs139568 & 22 & 42210985 &  & rs2051344 & 18 & 74715653 & MBP & 5.56 & 0.03 & 0.23 & 0.05 &  \\
+  MBP & ILMN\_2398939 & 18 & rs2051344 & 18 & 74715653 & MBP & rs1125539 & 3 & 155204939 &  & 5.79 & 0.02 & 0.76 & 0.27 &  \\
+  MBP & ILMN\_2398939 & 18 & rs2051344 & 18 & 74715653 & MBP & rs2619046 & 5 & 55097534 &  & 6.03 & 0.15 & 0.50 & 0.26 &  \\
+  MBP & ILMN\_2398939 & 18 & rs4805021 & 19 & 33436367 &  & rs2051344 & 18 & 74715653 & MBP & 5.82 & 0.03 & 0.47 & 0.14 &  \\
+  MBP & ILMN\_2398939 & 18 & rs8092433 & 18 & 74747424 &  & rs4890876 & 18 & 74732087 &  & 5.40 & 7.06 & 21.91 & 28.73 & 0.015 \\
+  MEGF9 & ILMN\_2290118 & 9 & rs13039689 & 20 & 51922071 &  & rs966396 & 9 & 123453281 & MEGF9 & 4.63 & 1.13 & 1.33 & 1.71 &  \\
+  MFN2 & ILMN\_1651385 & 1 & rs7989895 & 13 & 109401737 &  & rs4846085 & 1 & 12050634 & MFN2 & 5.76 & 0.61 & 0.25 & 0.41 &  \\
+  MGC13057 & ILMN\_1787526 & 2 & rs12718598 & 7 & 50428445 & MGC13057 & rs11725347 & 4 & 171860973 &  & 5.81 & 0.13 & 0.30 & 0.14 &  \\
+  MGC13057 & ILMN\_1787526 & 2 & rs674608 & 18 & 69070772 &  & rs12718598 & 7 & 50428445 & MGC13057 & 5.57 & 0.07 & 1.03 & 0.50 &  \\
+  MGC13057 & ILMN\_1787526 & 2 & rs8058318 & 16 & 82628245 &  & rs12718598 & 7 & 50428445 & MGC13057 & 7.05 & 0.11 & 0.12 & 0.05 &  \\
+  MGC72104 & ILMN\_1688318 & 20 & rs845787 & 20 & 26197931 & MGC72104 & rs2660665 & 8 & 137526799 &  & 4.17 & 0.05 & 0.08 & 0.02 &  \\
+  MGST3 & ILMN\_1751956 & 1 & rs740441 & 17 & 55779644 &  & rs4147592 & 1 & 165600146 & MGST3 & 5.45 & 0.57 & 0.27 & 0.40 &  \\
+  MPZL2 & ILMN\_1752932 & 11 & rs1805 & 11 & 118076069 & MPZL2 & rs11771552 & 7 & 154708716 &  & 5.90 & 0.01 & 0.23 & 0.04 &  \\
+  MPZL2 & ILMN\_1752932 & 11 & rs7316716 & 12 & 19953193 &  & rs1805 & 11 & 118076069 & MPZL2 & 5.64 & 0.97 & 1.08 & 1.35 &  \\
+  MRPL36 & ILMN\_1800197 & 5 & rs17469061 & 10 & 8436432 &  & rs750495 & 5 & 1782046 & MRPL36 & 6.89 & 0.34 & 0.18 & 0.19 &  \\
+  MRPL43 & ILMN\_2258774 & 10 & rs6564769 & 16 & 80641040 &  & rs2863095 & 10 & 102746503 & MRPL43 & 5.71 & 0.26 &  &  &  \\
+  MRPL52 & ILMN\_1713966 & 14 & rs1950857 & 14 & 26710271 &  & rs3811188 & 14 &  & MRPL52 & 6.56 & 0.14 & 0.44 & 0.22 &  \\
+  MRPS10 & ILMN\_1663664 & 6 & rs10955512 & 8 & 110202230 &  & rs722269 & 6 & 42194916 & MRPS10 & 7.48 & 0.46 & 0.70 & 0.64 &  \\
+  MRPS10 & ILMN\_1663664 & 6 & rs11698155 & 20 & 15063214 &  & rs2395803 & 6 & 42158596 & MRPS10 & 6.85 & 0.31 & 0.63 & 0.46 &  \\
+  MRPS10 & ILMN\_1663664 & 6 & rs1420537 & 16 & 52453567 &  & rs13217993 & 6 & 42164401 & MRPS10 & 6.21 & 0.41 & 0.25 & 0.28 &  \\
+  MTMR15 & ILMN\_2152178 & 15 & rs7178375 & 15 & 31215935 & MTMR10 & rs12431444 & 14 & 42068689 &  & 5.18 & 1.87 & 1.87 & 2.86 &  \\
+  MX1 & ILMN\_1662358 & 21 & rs459498 & 21 & 42795027 &  & rs11160227 & 14 & 95514596 &  & 6.31 & 0.46 & 0.52 & 0.50 &  \\
+  MX1 & ILMN\_1662358 & 21 & rs459498 & 21 & 42795027 &  & rs4973801 & 3 & 26706382 &  & 5.83 & 0.11 & 0.50 & 0.23 &  \\
+  MX1 & ILMN\_1662358 & 21 & rs459498 & 21 & 42795027 &  & rs8130120 & 21 & 29363604 &  & 6.78 & 0.29 & 0.92 & 0.65 & 13.431 \\
+  MYBPC3 & ILMN\_1781184 & 11 & rs10134030 & 14 & 61593110 &  & rs1317149 & 11 & 47486885 & MYBPC3 & 5.56 & 0.13 & 0.46 & 0.23 &  \\
+  MYBPC3 & ILMN\_1781184 & 11 & rs7322768 & 13 & 109550561 &  & rs7124681 & 11 & 47529947 & MYBPC3 & 5.70 & 0.04 & 0.08 & 0.02 &  \\
+  MYOM1 & ILMN\_1680344 & 18 & rs4798075 & 18 & 3247256 & MYOM1 & rs2737422 & 8 & 134485237 &  & 6.02 & 0.74 & 0.15 & 0.40 &  \\
+  N4BP1 & ILMN\_2201966 & 16 & rs12444224 & 16 & 87580855 &  & rs11649236 & 16 & 48632478 & N4BP1 & 5.54 & 2.00 & 0.59 & 1.77 & 38.948 \\
+  NAAA & ILMN\_1668605 & 4 & rs2707575 & 7 & 147638723 &  & rs6826085 & 4 & 76870229 & NAAA & 5.65 & 0.20 & 0.03 & 0.04 &  \\
+  NAAA & ILMN\_2391512 & 4 & rs2071856 & 22 & 37770630 &  & rs6826085 & 4 & 76870229 & NAAA & 5.46 & 0.27 & 0.43 & 0.30 &  \\
+  NAPRT1 & ILMN\_1710752 & 8 & rs2123758 & 8 & 144663661 & NAPRT1 & rs2786014 & 1 & 234897243 &  & 6.08 & 0.07 & 0.48 & 0.18 &  \\
+  NAPRT1 & ILMN\_1710752 & 8 & rs2123758 & 8 & 144663661 & NAPRT1 & rs3889129 & 8 & 144613680 &  & 8.45 & 15.12 & 16.08 & 30.77 & 0.050 \\
+  NAPRT1 & ILMN\_1710752 & 8 & rs2123758 & 8 & 144663661 & NAPRT1 & rs4862705 & 4 & 187445552 &  & 5.62 & 1.27 & 0.19 & 0.81 &  \\
+  NAPRT1 & ILMN\_1710752 & 8 & rs2123758 & 8 & 144663661 & NAPRT1 & rs6455553 & 6 & 167811764 &  & 6.12 & 0.87 & 0.76 & 1.01 &  \\
+  NAPRT1 & ILMN\_1710752 & 8 & rs2123758 & 8 & 144663661 & NAPRT1 & rs700276 & 7 & 146189057 &  & 6.86 & 1.10 & 2.58 & 2.77 &  \\
+  NAPRT1 & ILMN\_1710752 & 8 & rs2123758 & 8 & 144663661 & NAPRT1 & rs7571561 & 2 & 213386267 &  & 6.03 & 0.13 & 0.47 & 0.23 &  \\
+  NAPRT1 & ILMN\_1710752 & 8 & rs2208123 & 22 & 48214812 &  & rs2123758 & 8 & 144663661 & NAPRT1 & 6.60 & 0.29 & 0.88 & 0.63 &  \\
+  NAPRT1 & ILMN\_1710752 & 8 & rs4743420 & 9 & 103488089 &  & rs2123758 & 8 & 144663661 & NAPRT1 & 5.50 & 0.12 & 0.17 & 0.08 &  \\
+  NAPSA & ILMN\_1784040 & 19 & rs1405655 & 19 & 50882619 & NAPSB & rs930280 & 9 & 98391111 &  & 5.58 & 0.82 & 0.10 & 0.40 &  \\
+  NAPSB & ILMN\_2109416 & 19 & rs1405655 & 19 & 50882619 & NAPSB & rs10882406 & 10 & 95976932 &  & 5.58 & 0.67 & 1.10 & 1.12 &  \\
+  NAPSB & ILMN\_2109416 & 19 & rs1405655 & 19 & 50882619 & NAPSB & rs7577137 & 2 & 234721287 &  & 5.58 & 2.11 & 0.44 & 1.71 &  \\
+  NCL & ILMN\_2121437 & 2 & rs7563453 & 2 & 232301670 &  & rs4973397 & 2 & 232291471 &  & 7.31 & 7.51 & 6.33 & 12.70 & 0.010 \\
+  NDUFA12 & ILMN\_1737738 & 12 & rs2746971 & 22 & 37101890 &  & rs11107847 & 12 & 95386791 & NDUFA12 & 3.88 & 0.39 & 0.18 & 0.22 &  \\
+  NMT2 & ILMN\_1656378 & 10 & rs10906857 & 10 & 15239498 & NMT2 & rs12490878 & 3 & 183114008 &  & 6.84 & 0.42 & 0.34 & 0.35 &  \\
+  NOD2 & ILMN\_1762594 & 16 & rs2967636 & 19 & 7067773 &  & rs9302752 & 16 & 50719103 & NOD2 & 5.90 & 0.24 & 0.04 & 0.06 &  \\
+  NRBF2 & ILMN\_3237385 & 10 & rs11063498 & 12 & 5209048 &  & rs7923609 & 10 & 65133822 & NRBF2 & 5.45 &  &  &  &  \\
+  NRBF2 & ILMN\_3237385 & 10 & rs2375269 & 11 & 69876894 &  & rs7923609 & 10 & 65133822 & NRBF2 & 5.53 &  &  &  &  \\
+  NRBF2 & ILMN\_3237385 & 10 & rs6025645 & 20 & 56157341 &  & rs7923609 & 10 & 65133822 & NRBF2 & 5.45 &  &  &  &  \\
+  NRBF2 & ILMN\_3237385 & 10 & rs6517815 & 21 & 19819016 &  & rs7923609 & 10 & 65133822 & NRBF2 & 6.11 &  &  &  &  \\
+  NRD1 & ILMN\_1800897 & 1 & rs4852124 & 2 & 240680022 &  & rs6588415 & 1 & 52334047 &  & 6.13 & 0.47 & 0.05 & 0.17 &  \\
+  NUDT18 & ILMN\_1787885 & 8 & rs5017351 & 11 & 25453482 &  & rs1005901 & 8 & 21964378 & NUDT18 & 5.44 & 0.03 & 0.46 & 0.15 &  \\
+  OAS1 & ILMN\_1658247 & 12 & rs11613438 & 12 & 113480510 &  & rs1047944 & 6 & 163997467 &  & 8.59 & 1.27 & 1.55 & 2.03 &  \\
+  OAS1 & ILMN\_1658247 & 12 & rs13311 & 12 & 113448652 &  & rs2072133 & 12 & 113409260 &  & 4.13 & 4.12 & 0.81 & 3.86 & 0.039 \\
+  OAS1 & ILMN\_1675640 & 12 & rs2892233 & 19 & 49160255 &  & rs3741981 & 12 &  & OAS1 & 4.38 & 0.87 & 0.46 & 0.76 &  \\
+  OPTN & ILMN\_2381899 & 10 & rs7192613 & 16 & 74286646 &  & rs17512962 & 10 & 13169066 & OPTN & 5.64 & 0.42 & 0.06 & 0.14 &  \\
+  OSBPL5 & ILMN\_2307032 & 11 & rs2829679 & 21 & 26662543 &  & rs998639 & 11 & 3149249 & OSBPL5 & 5.00 & 0.36 & 0.00 & 0.07 &  \\
+  OSTF1 & ILMN\_1742456 & 9 & rs17780195 & 17 & 70624189 &  & rs2273770 & 9 & 77755469 & OSTF1 & 5.42 & 0.16 & 0.87 & 0.49 &  \\
+  OSTF1 & ILMN\_1742456 & 9 & rs2273770 & 9 & 77755469 & OSTF1 & rs7718088 & 5 & 179590952 &  & 5.42 & 1.20 & 0.08 & 0.62 &  \\
+  OVGP1 & ILMN\_1734542 & 1 & rs10802822 & 1 & 240132968 &  & rs1264898 & 1 & 111992823 & OVGP1 & 5.43 & 0.13 & 1.48 & 0.88 & 128.140 \\
+  OVGP1 & ILMN\_1734542 & 1 & rs347331 & 3 & 140148107 &  & rs1264894 & 1 & 111969719 & OVGP1 & 6.04 & 0.25 & 1.21 & 0.82 &  \\
+  PAM & ILMN\_2313901 & 5 & rs28092 & 5 & 102149795 & PAM & rs784600 & 1 & 40139553 & HPCAL4 & 5.59 & 0.66 & 0.44 & 0.59 &  \\
+  PCYOX1L & ILMN\_1815951 & 5 & rs2438490 & 5 & 148726162 & PCYOX1L & rs2731939 & 3 & 21395989 &  & 6.20 & 0.19 & 0.26 & 0.16 &  \\
+  PEX5 & ILMN\_1660232 & 12 & rs10444467 & 12 & 128052636 &  & rs4329748 & 12 & 7364442 & PEX5 & 5.85 & 0.09 & 0.71 & 0.32 & 120.688 \\
+  PEX5 & ILMN\_1660232 & 12 & rs7495797 & 15 & 27246462 &  & rs4329748 & 12 & 7364442 & PEX5 & 5.74 & 0.34 & 0.09 & 0.13 &  \\
+  PFAAP5 & ILMN\_1797893 & 13 & rs131969 & 22 & 49151303 &  & rs7328733 & 13 & 33126737 & PFAAP5 & 5.64 & 0.87 & 0.36 & 0.67 &  \\
+  PGLYRP1 & ILMN\_1704870 & 19 & rs12982353 & 19 & 46529456 & PGLYRP1 & rs1263806 & 14 & 21982957 &  & 6.51 & 0.03 & 0.65 & 0.24 &  \\
+  PHCA & ILMN\_1812552 & 11 & rs493642 & 11 & 123097386 &  & rs10736812 & 11 & 76708086 & PHCA & 5.51 & 0.36 & 0.90 & 0.70 & 46.389 \\
+  PIK3IP1 & ILMN\_1719986 & 22 & rs4141404 & 22 & 31675185 & PIK3IP1 & rs2065841 & 1 & 61728597 &  & 5.60 & 0.20 & 0.01 & 0.03 &  \\
+  PISD & ILMN\_1793934 & 22 & rs470072 & 22 & 32263131 & PISD & rs10498313 & 14 & 30398876 &  & 5.23 & 0.02 & 0.87 & 0.33 &  \\
+  PISD & ILMN\_1793934 & 22 & rs6518752 & 22 & 31999127 & PISD & rs954627 & 1 & 18236681 &  & 7.11 & 0.00 & 1.19 & 0.48 &  \\
+  PISD & ILMN\_1793934 & 22 & rs715572 & 22 & 33234931 &  & rs6518754 & 22 & 32097775 & PISD & 4.12 & 0.05 & 0.42 & 0.15 & 1.137 \\
+  PNKD & ILMN\_1774604 & 2 & rs6869411 & 5 & 158781604 &  & rs4672884 & 2 & 219182481 & PNKD & 6.35 & 0.16 & 0.04 & 0.04 &  \\
+  PNPLA7 & ILMN\_1662587 & 9 & rs11639998 & 16 & 4527109 &  & rs928046 & 9 & 140487108 & PNPLA7 & 5.15 & 0.31 & 0.78 & 0.56 &  \\
+  PPFIBP2 & ILMN\_1675656 & 11 & rs911019 & 20 & 49668255 &  & rs4758001 & 11 & 7559930 & PPFIBP2 & 4.44 & 0.29 & 0.33 & 0.26 &  \\
+  PPP2R3C & ILMN\_1662617 & 14 & rs12914603 & 15 & 58350896 &  & rs11156875 & 14 & 35619816 & PPP2R3C & 5.81 & 0.12 & 0.42 & 0.19 &  \\
+  PPP2R5A & ILMN\_1738784 & 1 & rs10930170 & 2 & 166399467 &  & rs12120009 & 1 & 212447167 & PPP2R5A & 5.63 & 0.72 & 0.48 & 0.66 &  \\
+  PPP2R5A & ILMN\_1738784 & 1 & rs12423255 & 12 & 123595064 &  & rs12120009 & 1 & 212447167 & PPP2R5A & 5.72 & 0.08 & 0.95 & 0.46 &  \\
+  PPP2R5A & ILMN\_1738784 & 1 & rs1889083 & 13 & 66222691 &  & rs12120009 & 1 & 212447167 & PPP2R5A & 5.61 & 0.36 & 0.13 & 0.17 &  \\
+  PPP2R5A & ILMN\_1738784 & 1 & rs682334 & 11 & 107417238 &  & rs12120009 & 1 & 212447167 & PPP2R5A & 5.65 & 1.69 & 0.28 & 1.21 &  \\
+  PPP2R5A & ILMN\_1738784 & 1 & rs7757871 & 6 & 135030045 &  & rs12120009 & 1 & 212447167 & PPP2R5A & 5.95 & 0.37 & 0.06 & 0.12 &  \\
+  PPP2R5A & ILMN\_1738784 & 1 & rs7871178 & 9 & 27148475 &  & rs12120009 & 1 & 212447167 & PPP2R5A & 5.72 & 0.16 & 0.30 & 0.16 &  \\
+  PRDX5 & ILMN\_1711606 & 11 & rs8019823 & 14 & 95040482 &  & rs11600990 & 11 & 64082807 & PRDX5 & 6.43 & 0.81 & 0.14 & 0.44 &  \\
+  PRKCB1 & ILMN\_1713603 & 16 & rs2188355 & 16 & 23867776 &  & rs10492793 & 16 & 12639800 &  & 7.34 & 0.53 & 0.11 & 0.25 & 11.228 \\
+  PRMT2 & ILMN\_1675038 & 21 & rs1029231 & 21 & 47931653 & C21ORF57 & rs958127 & 18 & 31497346 &  & 5.60 & 0.19 & 0.03 & 0.04 &  \\
+  PRMT2 & ILMN\_1675038 & 21 & rs2839372 & 21 & 48063862 &  & rs11701058 & 21 & 47776382 & C21ORF57 & 4.81 & 0.69 & 4.47 & 4.06 & 0.287 \\
+  PSMB1 & ILMN\_1789176 & 6 & rs3862607 & 11 & 121774705 &  & rs13207114 & 6 & 170877444 & PSMB1 & 5.79 &  & 0.44 &  &  \\
+  PSMB1 & ILMN\_1789176 & 6 & rs4890648 & 18 & 43983954 &  & rs6928843 & 6 & 170890384 & PSMB1 & 5.14 & 0.00 & 0.26 & 0.04 &  \\
+  PSMB1 & ILMN\_1789176 & 6 & rs6060930 & 20 & 30347832 &  & rs9295415 & 6 & 170823379 & PSMB1 & 5.44 & 0.44 & 0.21 & 0.27 &  \\
+  PSMB1 & ILMN\_1789176 & 6 & rs6928843 & 6 & 170890384 & PSMB1 & rs2769689 & 1 & 225797957 &  & 4.58 & 1.95 & 0.64 & 1.78 &  \\
+  PSMB1 & ILMN\_1789176 & 6 & rs7299749 & 12 & 131727816 &  & rs13207114 & 6 & 170877444 & PSMB1 & 5.42 & 1.18 & 0.32 & 0.86 &  \\
+  PWP1 & ILMN\_1743049 & 12 & rs2353567 & 14 & 95478823 &  & rs11036212 & 11 & 5221825 & PTDSS1 & 5.00 & 0.03 & 0.48 & 0.15 &  \\
+  PWP1 & ILMN\_1743049 & 12 & rs4969205 & 17 & 76598123 &  & rs11036212 & 11 & 5221825 & PTDSS1 & 5.90 & 0.80 & 0.08 & 0.38 &  \\
+  PWP1 & ILMN\_1743049 & 12 & rs631562 & 11 & 126852438 &  & rs11036212 & 11 & 5221825 & PTDSS1 & 5.70 & 0.02 & 0.40 & 0.11 &  \\
+  QDPR & ILMN\_1672443 & 4 & rs4946705 & 6 & 106348246 &  & rs10020773 & 4 & 17526682 & QDPR & 5.75 & 1.03 & 1.25 & 1.55 &  \\
+  RAB3IP & ILMN\_1803197 & 12 & rs241730 & 22 & 33375704 &  & rs7305307 & 12 & 70235726 &  & 6.55 & 0.25 & 0.08 & 0.09 &  \\
+  RABAC1 & ILMN\_2207363 & 19 & rs1075728 & 19 & 42462788 & RABAC1 & rs7951628 & 11 & 120161117 &  & 6.42 & 0.28 & 0.84 & 0.59 &  \\
+  RBL2 & ILMN\_1756999 & 16 & rs9931702 & 16 & 53526551 & AKTIP & rs1863464 & 15 & 26938488 &  & 6.38 & 0.03 & 0.31 & 0.08 &  \\
+  RCN1 & ILMN\_1800276 & 11 & rs10879131 & 12 & 41147155 &  & rs4922579 & 11 & 32136436 & RCN1 & 5.23 & 0.58 & 0.37 & 0.47 &  \\
+  RCN1 & ILMN\_1800276 & 11 & rs4922579 & 11 & 32136436 & RCN1 & rs11166957 & 8 & 141177468 &  & 4.32 & 0.41 & 0.09 & 0.17 &  \\
+  RCN1 & ILMN\_1800276 & 11 & rs4922579 & 11 & 32136436 & RCN1 & rs1341899 & 1 & 102740645 &  & 5.40 & 0.04 & 0.26 & 0.07 &  \\
+  RERE & ILMN\_1802380 & 1 & rs4982958 & 14 & 24987865 &  & rs301819 & 1 & 8501786 & RERE & 5.66 & 0.61 & 1.23 & 1.17 &  \\
+  RERE & ILMN\_1802380 & 1 & rs7697290 & 4 & 135248366 &  & rs301819 & 1 & 8501786 & RERE & 5.74 & 0.14 & 0.10 & 0.06 &  \\
+  RERE & ILMN\_2327795 & 1 & rs11085829 & 19 & 13174312 &  & rs301819 & 1 & 8501786 & RERE & 5.12 & 0.21 & 0.33 & 0.21 &  \\
+  RERE & ILMN\_2327795 & 1 & rs3852011 & 3 & 112844086 &  & rs301819 & 1 & 8501786 & RERE & 5.71 & 0.08 & 0.60 & 0.26 &  \\
+  RNASE6 & ILMN\_1780533 & 14 & rs11628398 & 14 & 21182800 & RNASE6 & rs7324365 & 13 & 100601327 &  & 5.48 & 0.42 & 0.21 & 0.26 &  \\
+  RNASE6 & ILMN\_1780533 & 14 & rs6603134 & 19 & 8106521 &  & rs11628398 & 14 & 21182800 & RNASE6 & 5.11 & 0.09 & 0.22 & 0.08 &  \\
+  RNF167 & ILMN\_1794726 & 17 & rs238230 & 17 & 4875566 &  & rs4884857 & 13 & 54668512 &  & 4.37 &  &  &  &  \\
+  RNF167 & ILMN\_1794726 & 17 & rs400688 & 17 & 4839930 & RNF167 & rs11706900 & 3 & 36348968 &  & 5.59 & 0.71 & 0.46 & 0.64 &  \\
+  RNPEP & ILMN\_1738347 & 1 & rs1107121 & 21 & 46127549 &  & rs2819365 & 1 & 201983242 &  & 6.27 & 0.11 & 0.30 & 0.13 &  \\
+  RNPEP & ILMN\_1738347 & 1 & rs8071611 & 17 & 67153386 &  & rs2819365 & 1 & 201983242 &  & 4.32 & 1.48 & 0.52 & 1.28 &  \\
+  RPL13 & ILMN\_2413278 & 16 & rs352935 & 16 & 89648580 &  & rs2965817 & 16 & 89513234 &  & 4.98 & 3.79 & 14.41 & 17.24 & 0.135 \\
+  RPL23AP7 & ILMN\_2222750 & 2 & rs1401202 & 16 & 80320056 &  & rs4849261 & 2 & 114450028 & RPL23AP7 & 5.55 & 0.13 & 0.73 & 0.38 &  \\
+  RPL36AL & ILMN\_2189933 & 14 & rs3007033 & 14 & 50103816 & RPL36AL & rs17495030 & 9 & 138038093 &  & 5.46 & 0.09 & 0.06 & 0.02 &  \\
+  RPL36AL & ILMN\_2189936 & 14 & rs4900928 & 14 & 50020817 & RPL36AL & rs1502991 & 6 & 66137260 &  & 5.86 & 0.32 & 0.20 & 0.19 &  \\
+  RPL8 & ILMN\_1764721 & 8 & rs2958482 & 8 & 145984615 & RPL8 & rs1619856 & 1 & 234585790 &  & 4.59 & 0.10 & 0.37 & 0.15 &  \\
+  RPL8 & ILMN\_1764721 & 8 & rs4143674 & 20 & 4741304 &  & rs2958482 & 8 & 145984615 & RPL8 & 4.33 & 0.13 & 0.45 & 0.22 &  \\
+  SEC13 & ILMN\_3297880 & 3 & rs4889214 & 16 & 80913946 &  & rs696221 & 3 & 10342876 & SEC13 & 6.48 &  &  &  &  \\
+  SEMA4A & ILMN\_1702787 & 1 & rs17085428 & 5 & 95388015 &  & rs7695 & 1 & 156147326 & SEMA4A & 5.70 & 0.22 & 1.73 & 1.17 &  \\
+  SESN3 & ILMN\_1694027 & 11 & rs12147460 & 14 & 104412137 &  & rs684856 & 11 & 94906111 & SESN3 & 5.50 & 0.02 & 0.51 & 0.15 &  \\
+  SESN3 & ILMN\_1694027 & 11 & rs355391 & 15 & 46591793 &  & rs684856 & 11 & 94906111 & SESN3 & 5.67 & 0.31 & 0.06 & 0.10 &  \\
+  SESN3 & ILMN\_1694027 & 11 & rs684856 & 11 & 94906111 & SESN3 & rs7004947 & 8 & 134606425 &  & 5.60 & 0.21 & 0.51 & 0.31 &  \\
+  SH3BGRL2 & ILMN\_1762764 & 6 & rs10838191 & 11 & 43893658 &  & rs1354034 & 3 & 56849749 & PPBP & 5.52 & 0.70 & 0.12 & 0.35 &  \\
+  SH3BGRL2 & ILMN\_1762764 & 6 & rs2545385 & 5 & 66383979 &  & rs1354034 & 3 & 56849749 & PPBP & 5.97 & 0.20 & 0.51 & 0.30 &  \\
+  SH3BGRL2 & ILMN\_1762764 & 6 & rs6845304 & 4 & 88280502 &  & rs1354034 & 3 & 56849749 & PPBP & 5.23 & 0.32 & 0.71 & 0.53 &  \\
+  SH3GLB2 & ILMN\_2158336 & 9 & rs1034120 & 21 & 18196922 &  & rs17455517 & 9 & 131785369 & SH3GLB2 & 7.40 & 0.22 & 0.18 & 0.13 &  \\
+  SIRPG & ILMN\_1771801 & 20 & rs1535883 & 20 & 1612819 & SIRPG & rs6842739 & 4 & 60489510 &  & 5.74 & 0.29 & 0.18 & 0.17 &  \\
+  SLC22A18 & ILMN\_2382505 & 11 & rs11673260 & 19 & 52181798 &  & rs367035 & 11 & 2923826 & SLC22A18 & 5.47 & 0.09 & 0.24 & 0.09 &  \\
+  SLC22A18 & ILMN\_2382505 & 11 & rs367035 & 11 & 2923826 & SLC22A18 & rs3110874 & 7 & 153224179 &  & 5.70 & 0.15 & 0.10 & 0.06 &  \\
+  SLC22A18 & ILMN\_2382505 & 11 & rs367035 & 11 & 2923826 & SLC22A18 & rs3772054 & 2 & 241678528 &  & 6.15 & 0.39 & 0.13 & 0.19 &  \\
+  SLC41A3 & ILMN\_2356111 & 3 & rs1912136 & 11 & 24616743 &  & rs6771703 & 3 & 125801067 & SLC41A3 & 5.88 & 1.10 & 0.82 & 1.24 &  \\
+  SLC45A4 & ILMN\_1745778 & 8 & rs6985508 & 8 & 142337734 & SLC45A4 & rs7701916 & 5 & 174598073 &  & 5.95 & 0.86 & 0.07 & 0.40 &  \\
+  SLC46A3 & ILMN\_1658639 & 13 & rs949805 & 17 & 55602091 &  & rs7981190 & 13 & 29259349 & SLC46A3 & 5.52 & 0.09 & 0.58 & 0.26 &  \\
+  SMG7 & ILMN\_1706553 & 1 & rs8035259 & 15 & 97403923 &  & rs10911353 & 1 & 183489203 & SMG7 & 6.52 & 0.17 & 0.09 & 0.06 &  \\
+  SMOX & ILMN\_1775380 & 20 & rs8118315 & 20 & 4161500 & SMOX & rs11677815 & 2 & 65800982 &  & 5.68 & 0.39 & 0.62 & 0.52 &  \\
+  SNHG8 & ILMN\_3309349 & 4 & rs1105621 & 9 & 133050233 &  & rs705837 & 4 & 119225940 & SNHG8 & 6.11 &  &  &  &  \\
+  SNORD14A & ILMN\_1799381 & 11 & rs1520429 & 15 & 46259108 &  & rs214097 & 11 & 17291499 & SNORD14A & 6.60 & 0.29 & 1.03 & 0.72 &  \\
+  SNORD14A & ILMN\_1799381 & 11 & rs2634462 & 11 & 17339127 &  & rs6486334 & 11 & 17015557 &  & 7.31 & 13.11 & 10.96 & 23.22 & 0.324 \\
+  SNORD89 & ILMN\_3238662 & 2 & rs10445863 & 2 & 115929241 &  & rs750783 & 2 & 101889306 & SNORD89 & 6.08 &  &  &  & 14.040 \\
+  SNORD89 & ILMN\_3238662 & 2 & rs11605822 & 11 & 122986326 &  & rs750783 & 2 & 101889306 & SNORD89 & 5.96 &  &  &  &  \\
+  SNORD89 & ILMN\_3238662 & 2 & rs2135064 & 5 & 26778066 &  & rs750783 & 2 & 101889306 & SNORD89 & 6.33 &  &  &  &  \\
+  SNUPN & ILMN\_1733932 & 15 & rs8134646 & 21 & 46376528 & SNUPN & rs7185362 & 16 & 81888905 &  & 6.45 & 0.13 & 1.41 & 0.83 &  \\
+  SNUPN & ILMN\_2364535 & 15 & rs8134646 & 21 & 46376528 & SNUPN & rs1472075 & 3 & 193706323 &  & 5.59 & 0.34 & 0.00 & 0.06 &  \\
+  SPATA5L1 & ILMN\_1729179 & 15 & rs1131620 & 19 & 41117869 &  & rs4774580 & 15 & 45652086 & SPATA5L1 & 5.44 &  &  &  &  \\
+  STARD10 & ILMN\_1717052 & 11 & rs2221406 & 13 & 90174526 &  & rs1000620 & 11 & 72509713 &  & 5.65 & 0.67 & 0.12 & 0.33 &  \\
+  STYXL1 & ILMN\_2210729 & 7 & rs4073164 & 14 & 104947517 &  & rs17685 & 7 & 75616105 & STYXL1 & 5.88 & 0.57 & 0.17 & 0.31 &  \\
+  SULF2 & ILMN\_2345142 & 20 & rs11700063 & 20 & 46153148 & SULF2 & rs939294 & 4 & 180439236 &  & 5.51 & 0.46 & 0.24 & 0.30 &  \\
+  SULT1A4 & ILMN\_2336133 & 16 & rs1463965 & 18 & 74332954 &  & rs3785354 & 16 & 28550667 & TUFM & 7.05 & 0.01 & 0.05 & 0.00 &  \\
+  SULT1A4 & ILMN\_2336133 & 16 & rs2836657 & 21 & 40119768 &  & rs3785354 & 16 & 28550667 & TUFM & 5.83 &  &  &  &  \\
+  SURF6 & ILMN\_1778032 & 9 & rs6099626 & 20 & 56013994 &  & rs3118663 & 9 & 136281753 & SURF6 & 6.14 & 0.26 & 0.16 & 0.14 &  \\
+  SYTL2 & ILMN\_2336609 & 11 & rs1375719 & 13 & 103410782 &  & rs485485 & 11 & 85495269 & SYTL2 & 5.47 & 0.28 & 0.31 & 0.24 &  \\
+  THBS3 & ILMN\_1804663 & 1 & rs1939875 & 11 & 95422867 &  & rs4072037 & 1 & 155162067 & THBS3 & 5.55 & 0.03 & 0.15 & 0.03 &  \\
+  THBS3 & ILMN\_1804663 & 1 & rs8014956 & 14 & 20687978 &  & rs2049805 & 1 & 155194980 & THBS3 & 5.65 & 0.31 & 0.76 & 0.55 &  \\
+  TIPRL & ILMN\_1781457 & 1 & rs2823245 & 21 & 16745523 &  & rs1320993 & 1 & 168154599 & TIPRL & 5.22 & 0.07 & 0.40 & 0.15 &  \\
+  TMED4 & ILMN\_1804148 & 7 & rs1940400 & 11 & 132389627 &  & rs17725246 & 7 & 44581986 & TMED4 & 5.70 & 0.06 & 1.34 & 0.70 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs2839013 & 21 & 47248981 &  & rs8106959 & 19 & 36219525 & TMEM149 & 8.11 & 0.16 & 0.48 & 0.26 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs5762235 & 22 & 27925288 &  & rs8106959 & 19 & 36219525 & TMEM149 & 6.79 &  &  &  &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs6090518 & 20 & 45207005 &  & rs8106959 & 19 & 36219525 & TMEM149 & 11.09 & 0.76 &  &  &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs807491 & 19 & 36268923 & SNX26 & rs7254601 & 19 & 36147315 & TMEM149 & 12.16 & 81.55 & 45.78 & 145.78 & 0.122 \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs10508289 & 10 & 4799159 &  & 8.12 & 1.55 & 3.09 & 3.67 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs10819626 & 9 & 133025756 &  & 8.02 & 0.40 & 0.99 & 0.80 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs10937361 & 3 & 188359436 &  & 8.39 & 3.61 & 1.18 & 3.78 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs1401098 & 12 & 128884559 &  & 7.37 & 2.41 & 1.00 & 2.52 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs1557335 & 18 & 64268976 &  & 6.95 & 0.08 & 0.07 & 0.03 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs17719594 & 14 & 90932598 &  & 6.93 & 3.06 & 0.77 & 2.87 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs1843357 & 8 & 13822381 &  & 6.21 & 3.72 & 3.33 & 6.00 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs2351458 & 4 & 113317583 &  & 7.30 & 0.04 & 9.61 & 8.00 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs2539000 & 7 & 147619772 &  & 6.70 & 1.57 & 1.52 & 2.27 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs2731711 & 5 & 171792273 &  & 5.92 & 0.19 & 0.33 & 0.19 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs471728 & 11 & 129595460 &  & 8.89 & 0.90 & 3.62 & 3.51 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs6718480 & 2 & 233879066 &  & 8.55 & 3.31 & 5.15 & 7.36 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs6926382 & 6 & 161683974 &  & 5.80 & 3.06 & 8.80 & 10.72 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs7213338 & 17 & 80357420 &  & 5.49 & 0.07 & 3.14 & 2.10 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs914940 & 1 & 242889492 &  & 6.22 & 3.36 & 6.96 & 9.20 &  \\
+  TMEM149 & ILMN\_1786426 & 19 & rs8106959 & 19 & 36219525 & TMEM149 & rs9509428 & 13 & 21473952 &  & 9.44 & 0.10 & 5.75 & 4.47 &  \\
+  TMEM63A & ILMN\_1719649 & 1 & rs1254086 & 13 & 72890603 &  & rs4149226 & 1 & 226027323 & TMEM63A & 5.60 &  &  &  &  \\
+  TMEM80 & ILMN\_1708482 & 11 & rs1548475 & 19 & 58058246 &  & rs4963126 & 11 & 656845 & TMEM80 & 5.79 & 0.64 & 0.12 & 0.32 &  \\
+  TNPO3 & ILMN\_1683811 & 7 & rs1537146 & 9 & 4859303 &  & rs10488630 & 7 & 128593948 & IRF5 & 5.61 & 0.11 & 0.15 & 0.07 &  \\
+  TNPO3 & ILMN\_1683811 & 7 & rs199793 & 20 & 22287303 &  & rs10488630 & 7 & 128593948 & IRF5 & 5.52 & 1.03 & 0.17 & 0.62 &  \\
+  TRA2A & ILMN\_1731043 & 7 & rs7776572 & 7 & 23528927 &  & rs11770192 & 7 & 23498358 &  & 8.23 & 3.19 & 1.89 & 4.09 & 0.031 \\
+  TRAPPC4 & ILMN\_1814650 & 11 & rs1278760 & 13 & 113531675 &  & rs3916581 & 11 & 118887887 & TRAPPC4 & 5.61 & 0.28 & 0.40 & 0.29 &  \\
+  TRAPPC4 & ILMN\_1814650 & 11 & rs1793823 & 11 & 131018917 &  & rs3916581 & 11 & 118887887 & TRAPPC4 & 5.52 & 0.93 & 0.01 & 0.36 & 12.131 \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs10059004 & 5 & 166970604 &  & 5.97 & 0.21 & 1.60 & 1.07 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs1023095 & 8 & 132022957 &  & 6.92 & 0.37 & 0.87 & 0.68 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs1375714 & 6 & 156404902 &  & 7.79 & 0.12 & 0.18 & 0.08 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs1393299 & 1 & 242329791 &  & 6.43 & 0.63 & 0.47 & 0.59 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs17763599 & 19 & 2369415 &  & 6.38 & 0.21 & 0.24 & 0.16 & 5.389 \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs4968328 & 17 & 57495457 &  & 6.51 & 0.50 & 0.38 & 0.44 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs7313362 & 12 & 129644342 &  & 7.08 & 0.04 & 0.65 & 0.25 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs7694997 & 4 & 9947811 &  & 5.86 & 0.20 & 0.36 & 0.22 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs7800935 & 7 & 146690926 &  & 6.27 & 0.15 & 0.33 & 0.16 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs17159840 & 19 & 7758194 & TRAPPC5 & rs856638 & 14 & 85439550 &  & 6.73 & 0.24 & 0.07 & 0.08 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs380708 & 22 & 22740855 &  & rs17159840 & 19 & 7758194 & TRAPPC5 & 7.58 &  &  &  &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs3916995 & 21 & 45128454 &  & rs17159840 & 19 & 7758194 & TRAPPC5 & 7.73 & 0.85 & 0.78 & 1.01 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs6040514 & 20 & 11272861 &  & rs17159840 & 19 & 7758194 & TRAPPC5 & 8.10 & 0.51 & 0.55 & 0.56 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs7246264 & 19 & 7762978 &  & rs10179572 & 2 & 228504503 &  & 6.71 & 0.14 & 0.02 & 0.02 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs7246264 & 19 & 7762978 &  & rs12921440 & 16 & 30408765 &  & 7.34 & 0.14 & 0.26 & 0.13 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs7246264 & 19 & 7762978 &  & rs1887778 & 9 & 134635088 & RAPGEF1 & 7.05 & 0.08 & 0.86 & 0.40 &  \\
+  TRAPPC5 & ILMN\_2372639 & 19 & rs7246264 & 19 & 7762978 &  & rs963354 & 3 & 157393770 &  & 7.41 & 0.36 & 0.90 & 0.69 &  \\
+  TREM1 & ILMN\_1688231 & 6 & rs10862975 & 12 & 85749398 &  & rs2395771 & 6 & 41264577 & TREM1 & 5.42 & 0.11 & 0.25 & 0.11 &  \\
+  TREM1 & ILMN\_1688231 & 6 & rs12412964 & 10 & 108256422 &  & rs2395771 & 6 & 41264577 & TREM1 & 5.92 & 1.20 & 1.23 & 1.69 &  \\
+  TRIM38 & ILMN\_1697971 & 6 & rs2527180 & 7 & 158808416 &  & rs2032447 & 6 & 26044369 & TRIM38 & 6.46 & 0.04 & 0.91 & 0.39 &  \\
+  TSPAN14 & ILMN\_1785060 & 10 & rs968726 & 17 & 27194634 &  & rs10748526 & 10 & 82273079 & TSPAN14 & 6.00 & 0.07 & 0.18 & 0.06 &  \\
+  TSPAN32 & ILMN\_1718621 & 11 & rs10838738 & 11 & 47663049 & MYBPC3 & rs12800998 & 11 & 2317951 & TSPAN32 & 5.01 &  &  &  & 45.345 \\
+  TSPAN32 & ILMN\_2389970 & 11 & rs12800998 & 11 & 2317951 & TSPAN32 & rs620607 & 6 & 137947208 &  & 5.51 &  &  &  &  \\
+  TYMP & ILMN\_3223126 & 22 & rs140522 & 22 & 50971266 & ECGF1 & rs1198819 & 2 & 238746880 &  & 6.34 &  &  &  &  \\
+  TYMP & ILMN\_3223126 & 22 & rs470119 & 22 & 50966914 & ECGF1 & rs4783126 & 16 & 85147633 &  & 6.13 &  &  &  &  \\
+  UBASH3A & ILMN\_2338348 & 21 & rs1893592 & 21 & 43855067 & UBASH3A & rs7201194 & 16 & 83600397 &  & 5.91 & 0.59 & 0.42 & 0.52 &  \\
+  UBASH3A & ILMN\_2338348 & 21 & rs1893592 & 21 & 43855067 & UBASH3A & rs7512594 & 1 & 214514361 &  & 6.01 & 0.48 & 1.29 & 1.10 &  \\
+  USP36 & ILMN\_1697227 & 17 & rs2279308 & 17 & 76794981 & USP36 & rs7225546 & 17 & 75151717 &  & 5.71 & 0.03 & 0.14 & 0.03 & 1.643 \\
+  VASP & ILMN\_1743646 & 19 & rs1264226 & 19 & 46063167 &  & rs2276470 & 19 & 45974668 &  & 5.09 & 0.94 & 5.14 & 4.95 & 0.088 \\
+  VNN2 & ILMN\_1678939 & 6 & rs10435352 & 7 & 105252718 &  & rs1883613 & 6 & 133077063 & VNN2 & 5.64 & 0.84 & 0.15 & 0.46 &  \\
+  VNN2 & ILMN\_1678939 & 6 & rs13044386 & 20 & 9116155 &  & rs1883617 & 6 & 133072650 & VNN2 & 5.44 & 0.39 & 0.69 & 0.57 &  \\
+  VNN2 & ILMN\_1678939 & 6 & rs134447 & 22 & 49927332 &  & rs1883617 & 6 & 133072650 & VNN2 & 5.72 &  &  &  &  \\
+  VNN2 & ILMN\_1678939 & 6 & rs216495 & 11 & 16834510 &  & rs1883617 & 6 & 133072650 & VNN2 & 5.77 & 0.33 & 0.19 & 0.19 &  \\
+  VNN3 & ILMN\_1804935 & 6 & rs10278073 & 7 & 151662184 &  & rs2267952 & 6 & 133067782 & VNN3 & 6.44 & 0.16 & 0.74 & 0.41 &  \\
+  VNN3 & ILMN\_1804935 & 6 & rs1443946 & 8 & 73006453 &  & rs2267952 & 6 & 133067782 & VNN3 & 5.74 & 0.23 & 0.48 & 0.31 &  \\
+  VNN3 & ILMN\_1804935 & 6 & rs348462 & 9 & 75547169 &  & rs2267952 & 6 & 133067782 & VNN3 & 6.44 & 0.31 & 0.17 & 0.17 &  \\
+  VNN3 & ILMN\_1804935 & 6 & rs7157055 & 14 & 83262064 &  & rs2267952 & 6 & 133067782 & VNN3 & 5.82 & 0.03 & 0.19 & 0.04 &  \\
+  VNN3 & ILMN\_2387680 & 6 & rs2823165 & 21 & 16594253 &  & rs2267952 & 6 & 133067782 & VNN3 & 6.12 & 0.73 & 1.15 & 1.21 &  \\
+  VNN3 & ILMN\_2387680 & 6 & rs9596457 & 13 & 51692548 &  & rs2267952 & 6 & 133067782 & VNN3 & 4.83 & 0.46 & 0.05 & 0.16 &  \\
+  VSTM1 & ILMN\_1763455 & 19 & rs10500316 & 19 & 54553697 & VSTM1 & rs4552100 & 18 & 71024750 &  & 5.60 & 0.53 & 0.54 & 0.57 &  \\
+  VSTM1 & ILMN\_1763455 & 19 & rs10500316 & 19 & 54553697 & VSTM1 & rs7895870 & 10 & 123098249 &  & 5.71 & 0.48 & 0.17 & 0.26 &  \\
+  VSTM1 & ILMN\_1763455 & 19 & rs9625870 & 22 & 30261219 &  & rs10500316 & 19 & 54553697 & VSTM1 & 5.88 & 0.81 & 1.38 & 1.47 &  \\
+  WDR48 & ILMN\_1762103 & 3 & rs1388935 & 4 & 188927822 &  & rs6778963 & 3 & 39091812 & WDR48 & 5.88 & 0.19 & 0.13 & 0.09 &  \\
+  WDR48 & ILMN\_1762103 & 3 & rs1887778 & 9 & 134635088 & RAPGEF1 & rs883349 & 3 & 39067925 & WDR48 & 6.34 & 0.57 & 1.35 & 1.22 &  \\
+  WDR48 & ILMN\_1762103 & 3 & rs9554833 & 13 & 102624790 &  & rs7619193 & 3 & 39044116 & WDR48 & 5.85 & 0.18 & 0.61 & 0.35 &  \\
+  WDR6 & ILMN\_1669484 & 3 & rs12362253 & 11 & 123371708 &  & rs11715581 & 3 & 49194331 & WDR6 & 4.86 & 1.64 & 1.43 & 2.25 &  \\
+  XAF1 & ILMN\_2370573 & 17 & rs1533031 & 17 & 6673170 & XAF1 & rs12591171 & 15 & 93119799 &  & 5.48 & 2.38 & 0.17 & 1.63 &  \\
+  ZFP90 & ILMN\_1684628 & 16 & rs909446 & 21 & 37040648 &  & rs1182968 & 16 & 68573945 & ZFP90 & 5.79 & 0.09 & 0.36 & 0.15 &  \\
+  ZNF500 & ILMN\_1700238 & 16 & rs4823723 & 22 & 48283177 &  & rs2290560 & 16 & 4799041 & ZNF500 & 5.29 & 0.67 & 0.27 & 0.46 &  \\
+  ZYX & ILMN\_1701875 & 7 & rs6056281 & 20 & 8935312 &  & rs2242601 & 7 & 143093824 & ZYX & 6.04 & 0.26 & 0.01 & 0.05 &  \\
+   \hline
+\end{longtable}
+
+% Gene ID\tnote{a} & 
+% Probe ID\tnote{b} & 
+% Chr. & 
+% rs ID & 
+% Chr. & 
+% Position\tnote{c} & 
+% Association\tnote{d} & 
+% rs ID & 
+% Chr. & 
+% Position\tnote{c} & 
+% Association\tnote{d} & 
+% BSGS\tnote{e} & 
+% Fehrmann\tnote{f} & 
+% EGCUT\tnote{f} & 
+% Meta\tnote{g} & 
+% Distance / Mb\tnote{h} \\
+
+\begin{tablenotes}
+\item[a] Phenotypes are expression levels of RefSeq Genes
+\item[b] Illumina probe ID used to measure gene expression
+\item[c] Physical SNP position in base pairs (HG19)
+\item[d] RefSeq Gene ID of gene expression level that is influenced by the SNP (BSGS discovery dataset, significance threshold $= 1.29 \times 10^{-11}$)
+\item[e] Interaction $-\log_{10} p$-value from discovery dataset
+\item[f] Interaction $-\log_{10} p$-value from replication dataset
+\item[g] Interaction $-\log_{10} p$-value from meta analysis of replication datasets only
+\item[h] Distance in Mb between interacting SNPs for \emph{cis}-\emph{cis} acting SNP pairs
+
+\end{tablenotes}
+
+\end{ThreePartTable}
+}
+\end{landscape}
+
+\clearpage
 \section{References}

### 2013-06-25 analysis/gpmaps.R

@@ -7,48 +7,9 @@ library(plyr)
 library(xtable)
 
 
+#=================================================================================================#
+#=================================================================================================#
 
-plot3dHairballsRep <- function(sig, pg, cissnp, cohort, z=45)
-{
-	# make table of mean phenotypes
-	# AA cis vs 1:nalleles trans
-	# Aa cis vs 1:nalleles trans
-	# aa cis vs 1:nalleles trans
-
-	s <- subset(sig, probegene==pg)
-	snps <- unique(c(s$snp1, s$snp2))
-	stopifnot(cissnp %in% snps)
-	tsnps <- snps[! snps %in% cissnp]
-
-	# transpose the GP map for all where the cis snp is snp2
-
-	temp1 <- subset(s, select=c(snp1, chr1))
-	temp2 <- subset(s, select=c(snp2, chr2))
-	names(temp1) <- names(temp2) <- c("snp", "chr")
-	chrkey <- rbind(temp1, temp2)
-	chrkey <- subset(chrkey, !duplicated(snp))
-
-	nom <- paste("gcm", cohort, sep="")
-
-	l <- list()
-	for(i in 1:nrow(s))
-	{
-		print(gp <- s[[nom]][[i]])
-		snp1 <- cissnp
-		snp2 <- s$snp2[i]
-
-		if(s$snp1[i] != cissnp)
-		{
-			gp <- t(gp)
-			snp2 <- s$snp1[i]
-		}
-		gp <- gp - min(gp, na.rm=T)
-
-		title <- paste("chr", subset(chrkey, snp == cissnp)$chr, "x", subset(chrkey, snp == snp2)$chr)
-		l[[i]] <- plot3dGp(gp, title, cissnp, snp2, z)
-	}
-	do.call(grid.arrange, l)
-}
 
 normaliseGp <- function(g)
 {
@@ -141,18 +102,62 @@ plotTileGp <- function(l)
 	return(p)
 }
 
-plotTileGp2 <- function(l)
+
+plot3dGp <- function(gp, title="", snp1="SNP1", snp2="SNP2", z=-45)
 {
-	p <- ggplot(l, aes(x=snp2, y=snp1, fill=y)) + 
[…]

### 2013-06-25 analysis/hairballs_old.R

@@ -1,0 +1,62 @@
+library(ggbio)
+library(grid)
+library(gridExtra)
+library(plyr)
+library(GenomicRanges)
+library(qgraph)
+
+
+# This is code to make hairball plots
+# Redundant - Kostya did it better
+
+
+key <- subset(sig, !duplicated(probename), select=c(probename, probegene))
+nom <- subset(key, duplicated(probegene))$probename
+
+sig2 <- subset(sig, ! probename %in% nom)
+
+# More circles
+a <- subset(sig2, select=c(snp1, snp2))
+names(a) <- c("from", "to")
+a$thickness <- 1
+a$col <- as.character(sig2$rep)
+a$col[a$col == 0] <- "white"
+a$col[a$col == 1] <- "black"
+a$col[a$col == 2] <- "black"
+a$col2 <- a$col
+a$col2[sig2$rep == 1] <- "white"
+a$col3 <- as.character(sig2$rep)
+a$col3[a$col3 == 0] <- ""
+a$col3[a$col3 == 1] <- "black"
+a$col3[a$col3 == 2] <- "red"
+a$thickness2 <- a$thickness
+a$thickness2[a$col3 == "red"] <- 1.5
+a$cistrans <- "blue"
+a$cistrans[sig2$c]
+
+
+pdf(file="~/repo/eQTL-2D/analysis/images/hairballs_all.pdf")
+qgraph(a, curve=0.1, borders=FALSE, esize=1, color="grey", edge.color=a$thickness, labels=FALSE, vsize=0.2, arrows=FALSE)
+dev.off()
+pdf(file="~/repo/eQTL-2D/analysis/images/hairballs_rep.pdf")
+qgraph(a, curve=0.1, borders=FALSE, esize=1, color="grey", edge.color=a$col, labels=FALSE, vsize=0.2, arrows=FALSE)
+dev.off()
+pdf(file="~/repo/eQTL-2D/analysis/images/hairballs_rep2.pdf")
+qgraph(a, curve=0.1, borders=FALSE, esize=1, color="grey", edge.color=a$col2, labels=FALSE, vsize=0.2, arrows=FALSE)
+dev.off()
+pdf(file="~/repo/eQTL-2D/analysis/images/hairballs_allrep.pdf")
+qgraph(a, curve=0.1, borders=FALSE, esize=a$thickness2, color="grey", edge.color=a$col3, labels=FALSE, vsize=0.2, arrows=FALSE)
+dev.off()
+
[…]

### 2013-06-25 docs/manuscript/nature_letter.tex
completed variance analysis

@@
 By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain 0.25\% of phenotypic variance, approximately seven times lower than the additive variance (Methods).
-There are at least three caveats to this comparison.
-Firstly, the ratio of additive to epistatic variance may differ at different effect sizes.
+There are several caveats to this comparison.
+Firstly, the ratio of additive to epistatic variance may differ at different effect sizes, and our estimate is determined by the threshold used.
 Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
@@
 	\label{fig:allelefreq}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=5in]{compare_vc}
+	\caption{\textbf{Comparison of variance distribution from different cohorts} How does the estimated variance decomposition change in different cohorts?
+The cumulative proportion of the genetic variance that is additive, dominant, or epistatic for each putative interaction is shown on the $y$-axis.
+BSGS has 501 interactions whereas Fehrmann and EGCUT have 434 ($x$-axis).
+The variance estimates are ordered from lowest additive to highest additive proportion.
+This is done independently for each cohort to depict the distribution of estimated effects.}
+	\label{fig:compare_vc}
 \end{figure}
@@
 \clearpage
-\section{References}
+% \section{References}
 \bibliography{refs}

### 2013-06-25 analysis/variance_analysis.R

@@ -1,23 +1,10 @@
 library(reshape2)
 library(ggplot2)
 
-load("~/repo/eQTL-2D/analysis/interaction_list_replication_summary.RData")
 
+#================================================================================#
+#================================================================================#
 
-sig$type <- "cis-cis"
-sig$type[with(sig, chr1 == probechr & chr2 != probechr)] <- "cis-trans"
-sig$type[with(sig, chr1 != probechr & chr2 == probechr)] <- "cis-trans"
-sig$type[with(sig, chr1 != probechr & chr2 != probechr)] <- "trans-trans"
-
-for(i in 1:nrow(sig))
-{
-	sig$vc_fehr[[i]] <- sig$vc_fehr[[i]][[1]]
-	sig$vc_egcut[[i]] <- sig$vc_egcut[[i]][[1]]
-}
-
-bsgs <- do.call(rbind, sig$vc)
-fehr <- do.call(rbind, sig$vc_fehr)
-egcut <- do.call(rbind, sig$vc_egcut)
 
 standardiseVg <- function(dat)
 {
@@ -30,52 +17,122 @@ standardiseVg <- function(dat)
 	return(dat)
 }
 
-bsgs <- standardiseVg(bsgs)
-fehr <- standardiseVg(fehr)
-egcut <- standardiseVg(egcut)
 
-varA <- data.frame(bsgs = bsgs$varA, fehr = fehr$varA, egcut = egcut$varA)
-varD <- data.frame(bsgs = bsgs$varD, fehr = fehr$varD, egcut = egcut$varD)
-varI <- data.frame(bsgs = bsgs$varI, fehr = fehr$varI, egcut = egcut$varI)
+sortVar <- function(bsgs, fehr, egcut, cohort)
+{
+	varA <- data.frame(bsgs = bsgs$varA, fehr = fehr$varA, egcut = egcut$varA)
+	varD <- data.frame(bsgs = bsgs$varD, fehr = fehr$varD, egcut = egcut$varD)
+	varI <- data.frame(bsgs = bsgs$varI, fehr = fehr$varI, egcut = egcut$varI)
 
-pairs(varA)
-pairs(varD)
-pairs(varI)
+	varA$index <- 1:nrow(varA)
+	varA <- varA[order(varA[[cohort]], decreasing=T), ]
+	varD <- varD[varA$index, ]
+	varI <- varI[varA$index, ]
[…]

### 2013-06-26 docs/manuscript/methods.tex
more work

@@
 For each of the 687 IndexSNPs we calculated LD with all regional SNPs within a radius of 0.5Mb and kept all regional SNPs with LD $r^{2} > 0.8$.
-We then cross-referenced the remaining regional SNPs with the annotated reference genome (version HG19, GENCODE annotation version 16 \cite{Harrow2012}) querying whether the regional SNPs fell in intergenic, promotor, intronic, exonic, splice sites, 5' UTR or 3' UTR positions.
+We then cross-referenced the remaining regional SNPs with the annotated chromatin structure reference \cite{Hoffman2012}) querying whether the regional SNPs fell in Predicted promoter region including TSS (TSS), Predicted promoter flanking region (PF), Predicted enhancer (E), Predicted weak enhancer or open chromatin cis regulatory element (WE), CTCF enriched element (CTCF), Predicted transcribed region (T), or Predicted Repressed or Low Activity region (R) positions.
 Therefore a particular IndexSNP might cover multiple genomic features.

### 2013-06-26 docs/manuscript/nature_letter.tex
more work

@@
 In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood.
-After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes.
+After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}).
 Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
@@
 We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au).
-We also acknowledge Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.
+We also thank Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.
 
@@
 	\centering
-	\includegraphics[width=5in]{OLapWithGenomicFeaturesFinal}
-	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 1kb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic features (\emph{x}-axis) were covered by the SNPs that compose the 501 significant interactions.
+	\includegraphics[width=5in]{ChromatinStatesOverlap}
+	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 1Mb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic features (\emph{x}-axis) were covered by the SNPs that compose the 501 significant interactions.
 Green bars represent the proportion (\emph{y}-axis) of the 528,509 SNPs used in the analysis that fall within the range of the different genomic features.
-There is enrichment for \emph{cis}-acting SNPs (red bars) in promotor regions, but \emph{trans}-acting SNPs (blue bars) are not enriched for genomic features.}
+There is enrichment for \emph{cis}-acting SNPs (red bars) in promotor regions, but \emph{trans}-acting SNPs (blue bars) are not enriched for genomic features.
+The labels on the $x$-axis are as follows: TSS = Predicted promoter region including TSS, PF = Predicted promoter flanking region, E = Predicted enhancer, WE = Predicted weak enhancer or open chromatin cis regulatory element, CTCF = CTCF enriched element, T = Predicted transcribed region, R = Predicted Repressed or Low Activity region}
 	\label{fig:snplocations}
@@
 \begin{ThreePartTable}
+\label{tab:results}
 \begin{longtable}{|llr|lrrl|lrrl|rrrr|r|}

### 2013-06-26 analysis/gpmaps.R

@@ -246,6 +246,15 @@ plot3dGpGrid(TRAPPC5, "BSGS", 19, c(-45, 225, -45, -45, -45, 225))
 dev.off()
 
 
+
+
+SH3BGRL2 <- subset(datHeatmapGp(subset(sig, probegene == "SH3BGRL2")), cohort == "BSGS")
+plot3dGpGrid(SH3BGRL2, "BSGS", 8, c(NA))
+
+THBS3 <- subset(datHeatmapGp(subset(sig, probegene == "THBS3")), cohort == "BSGS")
+plot3dGpGrid(THBS3, "BSGS", 1, c(NA))
+
+
 #=================================================================================================#
 #=================================================================================================#
 

### 2013-06-27 docs/manuscript/nature_letter.tex

@@
 Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
-In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
+In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
 We tested the discovery interactions for replication in two independent data sets \cite{Metspalu2004, Fehrmann2011}.
@@
 For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects.
-When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
+Thus, when combined with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
 There are two simple ways to overcome this problem.
-One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes.
+One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants.
 Because our focus was to ascertain the extent to which instances of epistasis occur amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels.
@@
 % \subsection{Initial search in discovery set and replication}
-In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood.
+In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in peripheral blood.
 After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}).
@@
 
-In addition, we observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}).
+In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the distribution under the null distribution of no epistatic effects ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}).
 The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets.
-A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication.
-It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is up to four orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
+A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 27 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication.
+It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
 Therefore these results are encouraging with regards to the detection and replication of epistasis.
@@
 Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects.
-Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst additive $\times$ additive, additive $\times$ dominance, dominance $\times$ additive and dominance $\times$ dominance ($p = 0.22$ for departure from expectation).
+Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst additive $\times$ additive, additive $\times$ dominance, dominance $\times$ additive and dominance $\times$ dominance at the discovery stage ($p = 0.22$ for departure from expectation).
 This is not surprising because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.
@@
 Overall, we have demonstrated that it is possible to identify and replicate epistasis in complex traits amongst common human variants.
-The functional analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation.
-Further research into such epistatic effects may provide a useful portal to understanding molecular mechanisms and complex trait variation with greater clarity.
+The bioinformatic analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation.
+Further research into such epistatic effects may provide a useful framework to understanding molecular mechanisms and complex trait variation in greater detail.
 With computational techniques and data now widely available the search for epistasis in larger datasets for traits of broader interest is warranted.
@@
 \subsection{Methods Summary}
-We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7,339 probes representing 6,158 RefSeq genes.
+We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in peripheral blood samples for 7,339 probes representing 6,158 RefSeq genes.
 Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
@@
 Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
-The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
+The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
 Of these, 434 passed filtering in both replication datasets.

### 2013-06-27 replication/results/summary_script2.R
modified summary script

@@ -24,10 +24,3 @@ fehr <- summariseRep("~/repo/eQTL-2D/replication/results/replication_GrngHT12v3.
 egcut <- summariseRep("~/repo/eQTL-2D/replication/results/replication_EGCUT.RData")
 
 save(fehr, egcut, file="~/repo/eQTL-2D/replication/results/replication2_summarised.RData")
-
-
-
-load("~/repo/eQTL-2D/replication/results/replication_GrngHT12v3.RData")
-
-
-

### 2013-06-27 docs/manuscript/methods.tex

@@
 
-Gene expression profiles were generated from whole blood collected with PAXgene TM tubes (QIAGEN, Valencia, CA) using Illumina HT12-v4.0 bead arrays.
+Gene expression profiles were generated from peripheral blood collected with PAXgene TM tubes (QIAGEN, Valencia, CA) using Illumina HT12-v4.0 bead arrays.
 The Illumina HT-12 v4.0 chip contains 47,323 probes, although some probes are not assigned to RefSeq genes.
@@
 \paragraph{Fehrmann: $n=1240$}
-The Fehrmann dataset \cite{Fehrmann2011} consists of whole peripheral blood samples of 1240 unrelated individuals from the United Kingdom and the Netherlands.
+The Fehrmann dataset \cite{Fehrmann2011} consists of peripheral blood samples of 1240 unrelated individuals from the United Kingdom and the Netherlands.
 Some of these individuals are patients, while others are healthy controls.
@@
 \paragraph{EGCUT: $n=891$}
-The Estonian Genome Center of the University of Tartu (EGCUT) study \cite{Metspalu2004} consists of whole blood samples of 891 unrelated individuals from Estonia.
+The Estonian Genome Center of the University of Tartu (EGCUT) study \cite{Metspalu2004} consists of peripheral blood samples of 891 unrelated individuals from Estonia.
 They were genotyped using the Illumina 610 Quad platform.
@@
 The Center for Health Discovery and Well Being (CDHWB) Study \cite{Preininger2013} is a population based cohort consisting of 139 individuals of European descent collected in Atlanta USA.
-Gene expression profiles were generated with Illumina HT-12 V3.0 arrays from whole blood collected from Tempus tubes that preserve RNA.
+Gene expression profiles were generated with Illumina HT-12 V3.0 arrays from peripheral blood collected from Tempus tubes that preserve RNA.
 Whole genome genotypes were measured using Illumina OmniQuad arrays.
-Due to the small sample size, most SNP pairs did not pass filtering in this dataset (185 remained) and so we have excluded it from the rest of the analysis.
+Due to the small sample size, most SNP pairs did not pass filtering in this dataset (27 SNP pairs remained) and so we have excluded it from the rest of the analysis.
 

### 2013-06-27 docs/manuscript/methods.tex
submitted draft to peter and grant

@@
 
-\section{Additive vs epistatic variance comparison}
+\section{Additive and non-additive variance estimation}
+
+\subsection{Fixed effects}
 
@@
 In total the proportion of the phenotypic variance explained by the epistatic effects at these SNP pairs was 0.25\%.
+
+\subsection{Pedigree estimates}
+
+The gene expression levels for MBNL1, TMEM149, NAPRT1, TRAPPC5 and CAST are influenced by large \emph{cis}-\emph{trans} epistatic networks (eight interactions or more).
+Though it is not possible to orthogonally estimate the non-additive genetic variance for non-clonal populations, an approximation of a component of non-additive variance can be estimated using pedigree information.
+The BSGS data is comprised of some related individuals and standard quantitative genetic analysis was used to calculate the additive and dominance variance components for each gene expression phenotype in Powell \emph{et al} 2013 \cite{Powell2013}.
+The dominance effect is likely to capture additive $\times$ additive genetic variance plus some fraction of other epistatic variance components.
+We found that the aforementioned genes had dominance variance component estimates within the top 5\% of all 17,994 gene expression probes that were analysed in Powell \emph{et al} 2013.
 

### 2013-06-27 docs/manuscript/nature_letter.tex
submitted draft to peter and grant

@@
 Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
-We see similar epistatic networks involving multiple \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149_3D}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1_3D}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5_3D}), and CAST (Supplementary Figure \ref{fig:CAST_3D}).
+We see similar epistatic networks involving multiple (eight or more) \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149_3D}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1_3D}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5_3D}), and CAST (Supplementary Figure \ref{fig:CAST_3D}).
+We observed that these five gene expression phenotypes had whole-genome non-additive variance component estimates within the 95th percentile of the 17,994 gene expression phenotypes that were analysed in Powell \emph{et al} 2013 (Supplementary Table \ref{tab:vc}, Methods) \cite{Powell2013}.
 
@@
 Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that on average they explain 1.73\% of the phenotypic variance of each of the 7339 probes.
-By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain 0.25\% of phenotypic variance, approximately seven times lower than the additive variance (Methods).
+By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain 0.25\% of phenotypic variance, approximately seven times lower than the additive variance.
+If we use the effect estimates taken from the Fehrmann or EGCUT datasets to perform the same comparison we obtain ratios of additive to epistatic variance of 36:1 and 34:1, respectively (Supplementary Figure \ref{fig:compare_vc}, Methods).
 There are several caveats to this comparison.
@@
 % \subsection{Discussion}
-Overall, we have demonstrated that it is possible to identify and replicate epistasis in complex traits amongst common human variants.
+Overall, we have demonstrated that it is possible to identify and replicate epistasis in complex traits amongst common human variants, despite the relative contribution of pairwise epistasis to phenotypic variation being small.
 The bioinformatic analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation.
@@
 	\centering
-	\includegraphics[width=5in]{compare_vc}
-	\caption{\textbf{Comparison of variance distribution from different cohorts} How does the estimated variance decomposition change in different cohorts?
-The cumulative proportion of the genetic variance that is additive, dominant, or epistatic for each putative interaction is shown on the $y$-axis.
+	\includegraphics[width=5in]{compare_vc2}
+	\caption{\textbf{Comparison of variance explained by additive, dominant and epistatic effects from different cohorts} How does the estimated variance decomposition change in different cohorts?
+The proportion of the phenotypic variance that is additive (A), dominant (D), or epistatic (I) for each putative interaction is shown on the $y$-axis (Note: different scales for each row).
 BSGS has 501 interactions whereas Fehrmann and EGCUT have 434 ($x$-axis).
-The variance estimates are ordered from lowest additive to highest additive proportion.
+The variance estimates in each plot are ordered from lowest additive to highest.
 This is done independently for each cohort to depict the distribution of estimated effects.}
@@
 
-% Gene ID\tnote{a} & 
-% Probe ID\tnote{b} & 
-% Chr. & 
-% rs ID & 
-% Chr. & 
-% Position\tnote{c} & 
-% Association\tnote{d} & 
-% rs ID & 
-% Chr. & 
-% Position\tnote{c} & 
-% Association\tnote{d} & 
-% BSGS\tnote{e} & 
-% Fehrmann\tnote{f} & 
-% EGCUT\tnote{f} & 
-% Meta\tnote{g} & 
-% Distance / Mb\tnote{h} \\
-
 \begin{tablenotes}
@@
 
+\begin{table}[ht]
+\begin{threeparttable}
+\caption{\textbf{Estimation of additive and non-additive variance components from pedigree information}}
+\label{tab:vc}
+\begin{tabular}{llrrrr}
+
+\hline
+& & \multicolumn{2}{c}{Additive} & \multicolumn{2}{c}{Non-additive} \\
+Gene & Probe & Variance & s.e. & Variance & s.e. \\
+\hline
+NAPRT1 & ILMN\_1710752 & 0.37 & 0.03 & 0.14 & 0.05 \\
+  TMEM149 & ILMN\_1786426 & 0.41 & 0.04 & 0.09 & 0.04 \\
+  MBNL1 & ILMN\_2313158 & 0.18 & 0.03 & 0.11 & 0.04 \\
+  TRAPPC5 & ILMN\_2372639 & 0.32 & 0.04 & 0.13 & 0.05 \\
+  CAST & ILMN\_1717234 & 0.31 & 0.03 & 0.10 & 0.04 \\
+\hline
+
+\end{tabular}
+\end{threeparttable}
+\end{table}
+
 \clearpage

### 2013-06-27 analysis/variance_analysis.R

@@ -6,14 +6,20 @@ library(ggplot2)
 #================================================================================#
 
 
-standardiseVg <- function(dat)
+summariseVg <- function(dat)
 {
-	vG <- apply(dat, 1, sum)
-	dat <- dat / vG
 	dat <- as.data.frame(dat)
 	dat$varA <- dat$a. + dat$.a
 	dat$varD <- dat$d. + dat$.d
 	dat$varI <- dat$aa + dat$ad + dat$da + dat$dd
+	return(dat)	
+}
+
+standardiseVg <- function(dat)
+{
+	vG <- apply(dat, 1, sum)
+	dat <- dat / vG
+	dat <- summariseVg(dat)
 	return(dat)
 }
 
@@ -112,12 +118,15 @@ bsgs <- do.call(rbind, sig$vc)
 fehr <- do.call(rbind, sig$vc_fehr)
 egcut <- do.call(rbind, sig$vc_egcut)
 
-bsgs <- standardiseVg(bsgs)
-fehr <- standardiseVg(fehr)
-egcut <- standardiseVg(egcut)
+bsgs <- summariseVg(bsgs)
+fehr <- summariseVg(fehr)
+egcut <- summariseVg(egcut)
 
 dat <- sortVar(bsgs, fehr, egcut, "fehr")
 dat2 <- sortVar2(bsgs, fehr, egcut)
+
+dat3 <- dat2[with(dat2, order(dataset, Component, Variance)), ]
+dat3$index <- rep(1:501, 9)
 
 
 #================================================================================#
@@ -127,11 +136,19 @@ dat2 <- sortVar2(bsgs, fehr, egcut)
 plotvarG(dat2)
 ggsave(file="~/repo/eQTL-2D/analysis/images/compare_vc.pdf", width=7, height=7)
 
+ggplot(dat3, aes(y=Variance, x=index)) +
+	geom_bar(stat="identity", position=position_stack(width=0)) +
+	ylab("Proportion of phenotypic variance explained") + 
+	xlab("") +
+	facet_grid(Component ~ dataset, scales="free_y")
+ggsave(file="~/repo/eQTL-2D/analysis/images/compare_vc2.pdf", width=7, height=7)
+
+
 #================================================================================#
 #================================================================================#
 
 
-
+ggplot(dat2, aes(x=Variance)) + geom_histogram(binwidth=0.01) + facet_grid(dataset ~ Component)
 
 
 

### 2013-07-03 docs/manuscript/nature_letter.tex

@@
 \author[3]{Harm-Jan Westra}
-\author[4]{Anjali K Henders}
+\author[4,5,6]{Tonu Esko}
+\author[7]{Anjali K Henders}
 \author[1,2]{Allan F. McRae}
-\author[5]{Andres Metspalu}
-\author[6]{Greg Gibson}
-\author[4]{Nick G Martin}
-\author[5,7,8]{Tonu Esko}
+\author[2]{Jian Yang}
+\author[8]{Greg Gibson}
+\author[7]{Nick G Martin}
+\author[4]{Andres Metspalu}
 \author[3]{Lude Franke}
-\author[4]{Grant W Montgomery}
-\author[1,2]{Peter M Visscher}
-\author[1,2]{Joseph E Powell}
+\author[7,+]{Grant W Montgomery}
+\author[1,2,+]{Peter M Visscher}
+\author[1,2,+]{Joseph E Powell}
 
@@
 \affil[3]{Department of Genetics, University Medical Center Groningen, University of Groningen, Hanzeplein 1, Groningen, the Netherlands}
-\affil[4]{Queensland Institute of Medical Research, Brisbane, Queensland, Australia}
-\affil[5]{Estonian Genome Center, University of Tartu, Tartu, 51010, Estonia}
-\affil[6]{School of Biology and Centre for Integrative Genomics, Georgia Institute of Technology, Atlanta, Georgia United States of America}
-\affil[7]{Divisions of Endocrinology, Children's Hospital, Boston, MA, 02115, US}
+\affil[4]{Estonian Genome Center, University of Tartu, Tartu, 51010, Estonia}
+\affil[5]{School of Biology and Centre for Integrative Genomics, Georgia Institute of Technology, Atlanta, Georgia United States of America}
+\affil[6]{Divisions of Endocrinology, Children's Hospital, Boston, MA, 02115, US}
+\affil[7]{Queensland Institute of Medical Research, Brisbane, Queensland, Australia}
 \affil[8]{Medical and Population Genetics, Broad Institute, Cambridge, MA, 02142, US}
+\affil[+]{These authors contributed equally}
 \affil[*]{Corresponding author: g.hemani@uq.edu.au}
@@
 
-\section{Main text}
+\section*{Main text}
 
@@
 Transcription levels can be measured for thousands of genes.
-These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that many genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.
+These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus many genetic effects are relatively large, maximising the chance at detecting epistasis, should it exist.
 
@@
 A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 27 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication.
-It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
+It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
 Therefore these results are encouraging with regards to the detection and replication of epistasis.
@@
 We see similar epistatic networks involving multiple (eight or more) \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149_3D}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1_3D}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5_3D}), and CAST (Supplementary Figure \ref{fig:CAST_3D}).
-We observed that these five gene expression phenotypes had whole-genome non-additive variance component estimates within the 95th percentile of the 17,994 gene expression phenotypes that were analysed in Powell \emph{et al} 2013 (Supplementary Table \ref{tab:vc}, Methods) \cite{Powell2013}.
+We observed that from pedigree analysis these five gene expression phenotypes had non-additive variance component estimates within the 95th percentile of the 17,994 gene expression phenotypes that were analysed previously \cite{Powell2013} (Supplementary Table \ref{tab:vc}, Methods).
 
@@
 However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues).
-This difference between \emph{cis} and \emph{trans} SNPs suggests different roles in which epistasis might arise where the \emph{cis}-SNPs provide tissue specificity in these interactions.
+This difference between \emph{cis} and \emph{trans} SNPs suggests different roles in epistatic interactions where the \emph{cis}-SNPs provide tissue specificity in these interactions.
 There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a}, consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$).
@@
 
-\subsection{Methods Summary}
+\subsection*{Methods Summary}
 We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in peripheral blood samples for 7,339 probes representing 6,158 RefSeq genes.
-Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
+Recent hardware and software \cite{Hemani2011} advances that use graphics processing units (GPUs) made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
 We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
@@
 
-\subsection{Acknowledgements}
+\subsection*{Acknowledgements}
 
@@
 
-The University of Queensland group is supported by the Australian Research Council (DP130102666, DE130100691), the Australian National Health and Medical Research Council (APP1011506, APP1047956, APP1048853) and the National Institutes of Health (GM099568, GM075091).
+The University of Queensland group is supported by the Australian National Health and Medical Research Council (NHMRC) grants 389892, 496667, 613601, 1010374 and 1046880, the Australian Research Council (ARC) grant (DE130100691), and by National Institutes of Health (NIH) grants GM057091 and GM099568.
 
@@
 Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class.
-Phenotypes are for gene transcript levels (dark coloured tiles = low expression, light coloured tiles = high expression).
+Phenotypes are for gene transcript levels (dark coloured tiles = high expression, light coloured tiles = low expression).
 Columns of GP maps are for each independent population.
@@
 	\centering
-	\includegraphics[width=5in]{ChromatinStatesOverlap}
+	\includegraphics[width=5in]{ChromatinStatesOverlapFixedOrder}
 	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 1Mb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic features (\emph{x}-axis) were covered by the SNPs that compose the 501 significant interactions.
@@
 There is enrichment for \emph{cis}-acting SNPs (red bars) in promotor regions, but \emph{trans}-acting SNPs (blue bars) are not enriched for genomic features.
-The labels on the $x$-axis are as follows: TSS = Predicted promoter region including TSS, PF = Predicted promoter flanking region, E = Predicted enhancer, WE = Predicted weak enhancer or open chromatin cis regulatory element, CTCF = CTCF enriched element, T = Predicted transcribed region, R = Predicted Repressed or Low Activity region}
+The labels on the $x$-axis are as follows: E = Predicted enhancer, CTCF = CTCF enriched element, WE = Predicted weak enhancer or open chromatin cis regulatory element, PF = Predicted promoter flanking region, TSS = Predicted promoter region including transcriptional start site, T = Predicted transcribed region, R = Predicted Repressed or Low Activity region}
 	\label{fig:snplocations}
@@
 	\centering
-	\includegraphics[width=5in]{501_pairs_snp_list_Grid}
+	\includegraphics[width=5in]{H3K4me3Enrichment501_list}
 	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3 \cite{Trynka2013}.
@@
 
-ABCA7 & ILMN\_1743205 & 19 & rs3752237 & 19 & 1047161 & ABCA7 & rs596183 & 6 & 158100199 &  & 5.82 & 0.38 & 0.02 & 0.09 &  \\
+ABCA7 & ILMN\_1743205 & 19 & rs3752237 & 19 & 1047161 & ABCA7 & rs596183 & 6 & 158100199 &  & 5.82 & 0.38\tnote{i} & 0.02\tnote{i} & 0.09\tnote{j} &  \\
   ABCA7 & ILMN\_1743205 & 19 & rs3752237 & 19 & 1047161 & ABCA7 & rs914737 & 9 & 139522101 &  & 5.50 &  &  &  &  \\
@@
 \item[h] Distance in Mb between interacting SNPs for \emph{cis}-\emph{cis} acting SNP pairs
+\item[i] $p$-values are absent if the interaction did not pass the QC filtering in the replication dataset
+\item[j] Meta analysis $p$-values are absent if the interaction did not pass the QC filtering in either replication dataset
 
@@
 \begin{threeparttable}
-\caption{\textbf{Estimation of additive and non-additive variance components from pedigree information}}
+\caption{\textbf{Estimation of additive and non-additive variance components from pedigree information} Taken from previous analysis in Powell et al 2013 \cite{Powell2013}}
 \label{tab:vc}

### 2013-07-04 docs/manuscript/methods.tex
responding to co-author comments

@@
 We used epiGPU \cite{Hemani2011} software to perform an exhaustive scan for pairwise interactions, such that each SNP is tested against all other SNPs for statistical association with the expression values for each of the 7339 probes.
+This uses the massively parallel computational architecture of graphical processing units (GPUs) to speed up the exhaustive search.
 For each SNP pair there are 9 possible genotype classes.
@@
 This resulted in 432 and 117 SNP pairs with significant interaction terms from filters 1 and 2, respectively.
+
+We ruled out population stratification as a possible cause of inflated test statistics.
+To test for cryptic relatedness driving the interaction terms we tested for increased LD among the SNPs \cite{Yang2011}.
+We calculated the mean of the off-diagonal elements of the correlation matrix of all unique SNPs from the 501 interactions (731 SNPs) using only unrelated individuals, $\bar{r^2} = 0.0039$.
+This is not significantly different from the null hypothesis of zero (sampling error $= 1/n_{\textrm{unrelated}} = 0.0039$).
 
@@
 We shall refer to the colocalisation of two chromosome regions as a chromosome interaction.
-A map of all known pairwise chromosome interactions for K562 blood cell lines was recently produced \cite{Lan2012}, and we hypothesised that part of the underlying biological mechanism behind some of the 501 epistatic interactions may arise from chromosome interactions.
+A map of pairwise chromosome interactions for K562 blood cell lines was recently produced \cite{Lan2012}, and we hypothesised that part of the underlying biological mechanism behind some of the 501 epistatic interactions may arise from chromosome interactions.
 We found that 44 of the putative epistatic interactions were amongst SNPs that were within 5Mb of known chromosome interactions.
@@
 We then cross-referenced the remaining regional SNPs with the annotated chromatin structure reference \cite{Hoffman2012}) querying whether the regional SNPs fell in Predicted promoter region including TSS (TSS), Predicted promoter flanking region (PF), Predicted enhancer (E), Predicted weak enhancer or open chromatin cis regulatory element (WE), CTCF enriched element (CTCF), Predicted transcribed region (T), or Predicted Repressed or Low Activity region (R) positions.
-Therefore a particular IndexSNP might cover multiple genomic features.
+Therefore a particular IndexSNP might cover multiple genomic features through LD.
 

### 2013-07-04 docs/manuscript/nature_letter.tex
responding to co-author comments

@@
 \affil[1]{University of Queensland Diamantina Institute, University of Queensland, Princess Alexandra Hospital, Brisbane, Queensland, Australia}
-\affil[2]{The University of Queensland, Queensland Brain Institute, Brisbane, QLD, Australia}
+\affil[2]{Queensland Brain Institute, The University of Queensland, Brisbane, QLD, Australia}
 \affil[3]{Department of Genetics, University Medical Center Groningen, University of Groningen, Hanzeplein 1, Groningen, the Netherlands}
 \affil[4]{Estonian Genome Center, University of Tartu, Tartu, 51010, Estonia}
-\affil[5]{School of Biology and Centre for Integrative Genomics, Georgia Institute of Technology, Atlanta, Georgia United States of America}
+\affil[5]{Medical and Population Genetics, Broad Institute, Cambridge, MA, 02142, US}
 \affil[6]{Divisions of Endocrinology, Children's Hospital, Boston, MA, 02115, US}
 \affil[7]{Queensland Institute of Medical Research, Brisbane, Queensland, Australia}
-\affil[8]{Medical and Population Genetics, Broad Institute, Cambridge, MA, 02142, US}
+\affil[8]{School of Biology and Centre for Integrative Genomics, Georgia Institute of Technology, Atlanta, Georgia United States of America}
 \affil[+]{These authors contributed equally}
@@
 Three hundred and forty-five interactions had replication interaction $p$-values that were more extreme than the 2.5\% confidence interval of the distribution under the null hypothesis of no epistasis, with 30 significant at a conservative $p < 0.05$ Bonferroni level.
-There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
+There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
 Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs.
@@
 In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment (Methods).
-We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}).
+We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,tri-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}).
 There was significant enrichment for \emph{cis}-acting SNPs in haematopoietic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing).
@@
 This difference between \emph{cis} and \emph{trans} SNPs suggests different roles in epistatic interactions where the \emph{cis}-SNPs provide tissue specificity in these interactions.
-There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a}, consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$).
+There is also enrichment for \emph{cis}-SNPs to be localised in regions with active chromatin states \cite{Hoffman2012} (Supplementary Figure
+\ref{fig:snplocations}).
 
@@
 By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain 0.25\% of phenotypic variance, approximately seven times lower than the additive variance.
-If we use the effect estimates taken from the Fehrmann or EGCUT datasets to perform the same comparison we obtain ratios of additive to epistatic variance of 36:1 and 34:1, respectively (Supplementary Figure \ref{fig:compare_vc}, Methods).
 There are several caveats to this comparison.
@@
 All SNP pairs were required to have at least five data points in all nine genotype classes.
-If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained.
+If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained.
 Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
@@
 	\centering
-	\includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
-	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).
-Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data, but the remaining 89 interactions that did not replicate are depicted in grey.
-It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
-	\label{fig:fireworks}
-\end{figure}
-\clearpage
-
-\begin{figure}[H]
-	\centering
 	\includegraphics[width=2.5in]{gpbonfrep.pdf}
@@
 	\centering
+	\includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
+	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).
+Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data, but the remaining 89 interactions that did not replicate are depicted in grey.
+It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
+	\label{fig:fireworks}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+	\centering
 	\includegraphics[width=2.5in]{qqMeta.pdf}
@@
 	\includegraphics[width=5in]{ChromatinStatesOverlapFixedOrder}
-	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 1Mb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic features (\emph{x}-axis) were covered by the SNPs that compose the 501 significant interactions.
+	\caption{\textbf{Location of SNPs relative to genomic features} We used chromatin segmentation \cite{Hoffman2012} as a method for labelling genomic features.
+All SNPs within 1Mb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic features (\emph{x}-axis) were covered by the SNPs that compose the 501 significant interactions.
 Green bars represent the proportion (\emph{y}-axis) of the 528,509 SNPs used in the analysis that fall within the range of the different genomic features.
@@
 Enrichment $p$-values are calculated using permutation analysis for 34 different cell types ($y$-axis) in four tissue types (Rows of boxes).
+The dotted red line denotes significance (Bonferroni correction for 34 cell types, $x$-axis).
 There is enrichment for \emph{cis}-acting SNPs in Haematopoietic tissue types only.

### 2013-07-06 docs/manuscript/methods.tex
submitting to nature arghghggh

@@
 Profiles were first adjusted for raw background expression in each sample.
-Expression levels were then adjusted using quantile and log2 transformation to standardise distributions between samples.
+Expression levels were then adjusted using quantile and $\log_2$ transformation to standardise distributions between samples.
 Batch and polygenic effects were adjusted using the linear model
@@
 The parameters $\sigma_a^2$ and $\sigma_e^2$ are variance components for additive background genetic and environmental effects respectively.
-Here, we are using family based pedigree information rather than SNP based IDB to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$.
+Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$.
 The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero.
@@
 \end{equation}
-
 where $\mu$ is the mean expression level and $x _{ij}$ is the pairwise genotype class mean for genotype $i$ at SNP 1 and genotype $j$ at SNP 2.
@@
 Following filtering on this threshold only SNP pairs with at least 5 data points in all 9 genotype classes were kept.
-We then calculated the LD between interacting SNPs and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition.
-If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained.
+We then calculated the LD between interacting SNPs (amongst unrelated individuals within the discovery sample and also from 1000 genomes data) and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition.
+If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained.
 At this stage 6404 filtered SNP pairs remained.
@@
 Significance of epistasis was determined using a Bonferroni threshold of $0.05 / (6404+4751) = 4.48 \times 10^{-6}$.
-This resulted in 432 and 117 SNP pairs with significant interaction terms from filters 1 and 2, respectively.
+This resulted in 406 and 95 SNP pairs with significant interaction terms from filters 1 and 2, respectively.
 
@@
 
-We have attempted replication of the 501 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB.
-It was required that LD $r^2 < 0.1$ and $D'^{2} < 0.1$ between interacting SNPs, and all nine genotype classes had at least 5 individuals present in order to proceed with statistical testing for replication in both datasets.
+We attempted replication of the 501 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB.
+It was required that LD $r^2 < 0.1$ and $D'^{2} < 0.1$ between interacting SNPs (as measured in the replication sample directly), and all nine genotype classes had at least 5 individuals present in order to proceed with statistical testing for replication in both datasets.
 We also excluded any putative SNPs that had discordant allele frequencies in any of the datasets.
@@
 Some of these individuals are patients, while others are healthy controls.
-Individuals were genotyped using the Illumina 610 Quad platform.
-RNA levels were quantified using the HT12v3 platform.
+Individuals were genotyped using the Illumina HumanHap300, Illumina HumanHap370CNV, and Illumina 610 Quad platforms.
+RNA levels were quantified using the Illumina HT-12 V3.0 platform.
 
@@
 The Estonian Genome Center of the University of Tartu (EGCUT) study \cite{Metspalu2004} consists of peripheral blood samples of 891 unrelated individuals from Estonia.
-They were genotyped using the Illumina 610 Quad platform.
-RNA levels were quantified using the HT12v3 platform.
+They were genotyped using the Illumina HumanHap370CNV platform.
+RNA levels were quantified using the Illumina HT-12 V3.0 platform.
 
@@
 Whole genome genotypes were measured using Illumina OmniQuad arrays.
-Due to the small sample size, most SNP pairs did not pass filtering in this dataset (27 SNP pairs remained) and so we have excluded it from the rest of the analysis.
+Due to the small sample size, most SNP pairs did not pass filtering in this dataset (20 SNP pairs remained) and so we have excluded it from the rest of the analysis.
 
@@
 We then took the interaction $p$-values from EGCUT and Fehrmann and calculated a joint $p$-value using Fisher's method of combining $p$-values for a meta analysis as $-2 \ln p_{1} - 2 \ln p_{2} \sim \chi^{2}_{4 d.f.}$.
-As in the discovery analysis, all gene expression levels were normalised using rank transformation to avoid skew or outliers in the distribution.
+As in the discovery analysis, all gene expression levels were normalised using rank transformation to avoid skew or outliers in the distribution \cite{Westra2011}.
 
@@
 We employed a recently published method (\url{http://www.broadinstitute.org/mpg/epigwas/}) \cite{Trynka2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions.
-We subsequently refer to these 731 SNPs as the IndexSNPs A recent paper shows that H3K4me3 chromatin marks are the most cell-type-specific marks in terms of co-localization with previously published GWAS loci \cite{Trynka2013}.
-The idea behind the method is that variants related to a particular phenotype may affect cell-type-specific gene expression by changing regulatory elements in cell types relevant to that phenotype.
-Hence, overlap between associated variants and chromatin marks should occur preferentially in the relevant cell type(s).
-Our analysis tested for enrichment of these chromatin marks in 34 different cell types from 4 different broad tissue types.
-
-For each IndexSNP, a locus region was defined, bounded on either side of the IndexSNP by the most distant SNP within 250 kb that was in LD ($r^{2} > 0.8$).
-For each SNP in a locus region regulatory activity scores were calculated as the height of the nearest H3K4me3 mark divided by distance from the SNP to the H3K4me3 mark.
-The SNP with the highest score within each locus region in a given tissue was designated the BestSNP, which served as the score representing that locus.
-
-Cell-type-specificity scores per locus were estimated by normalizing BestSNP scores so that the sum of scores for a given locus across all cell types equaled 1.
-Cell-type-specificity scores per tissue were defined by summing normalized BestSNP scores across all loci within a given tissue.
-10,000 sets of SNPs (matched to the IndexSNP regions having the same total number of H3K4me3 peaks) were sampled (from among background SNPs provided with the software) to estimate null distributions of cell-type-specificity scores per locus and per tissue.
-P-values for cell-type-specificity scores summed across all BestSNPs for each tissue (the observed per-tissue score) were estimated as the proportion of random SNP sets with a per-tissue score exceeding the observed per-tissue score.
-We identified loci with BestSNP cell-type-specificity scores falling at or above the 95th percentile of the corresponding null distribution as demonstrating greater than expected specificity within a given cell type \cite{Trynka2013}.
+The exact method used to perform this analysis has been described previously \cite{Rietveld2013}.
+Briefly, we tested the hypothesis that the 731 SNPs were more likely to be in transcriptionally active regions (as measured by chromatin marks) than a random set of SNPs selected from the same SNP chip.
+This hypothesis was tested for 34 cell types across four broad tissue types (haematopoietic, gastrointenstinal, musculoskeletal and endocrine, and brain).
 
@@
 We scanned the map of chromosome interactions for overlaps with the new sets and then repeated the random shuffling process.
-We performed 10,000 such permutations to generate a null distribution of chromosome interaction overlaps.
+We performed 1,000 such permutations to generate a null distribution of chromosome interaction overlaps.
 

### 2013-07-06 docs/manuscript/nature_letter.tex
submitting to nature arghghggh

@@
 \affil[1]{University of Queensland Diamantina Institute, University of Queensland, Princess Alexandra Hospital, Brisbane, Queensland, Australia}
-\affil[2]{Queensland Brain Institute, The University of Queensland, Brisbane, QLD, Australia}
+\affil[2]{Queensland Brain Institute, University of Queensland, Brisbane, QLD, Australia}
 \affil[3]{Department of Genetics, University Medical Center Groningen, University of Groningen, Hanzeplein 1, Groningen, the Netherlands}
@@
 In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently and cumulatively to the mean of a trait.
-This framework has been successful in identifying thousands of associations \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}.
+This framework has been successful in identifying thousands of associations \cite{Visscher2012}.
+But to date, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}, there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}.
 Outside the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
@@
 In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in peripheral blood.
-After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}).
+After stringent filtering and multiple testing correction (5\% significance threshold $ p < 2.91 \times 10^{-16}$, Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}).
 Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
-We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}).
+We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.0001$, Table \ref{tab:bonferroni}).
 These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).
@@
 The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets.
-A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 27 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication.
+A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}).
 It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
@@
 We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
-For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an effect (Supplementary Figure \ref{fig:MBNL1_3D}.
-Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
+For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an effect (Supplementary Figure \ref{fig:MBNL1_3D}).
+Each of these interactions has evidence for replication in at least one dataset and six are significantly replicated at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
 We see similar epistatic networks involving multiple (eight or more) \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149_3D}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1_3D}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5_3D}), and CAST (Supplementary Figure \ref{fig:CAST_3D}).
@@
 However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues).
-This difference between \emph{cis} and \emph{trans} SNPs suggests different roles in epistatic interactions where the \emph{cis}-SNPs provide tissue specificity in these interactions.
-There is also enrichment for \emph{cis}-SNPs to be localised in regions with active chromatin states \cite{Hoffman2012} (Supplementary Figure
-\ref{fig:snplocations}).
-
-We also demonstrate spacial organisation of interacting loci suggesting a mechanism by which biological function can lead to epistatic genetic variance.
+This difference between \emph{cis} and \emph{trans} SNPs suggests different roles in epistatic interactions where tissue specificity is provided by the \emph{cis} SNPs.
+There is also enrichment for \emph{cis}-SNPs to be localised in regions with regulatory genomic features as measured by chromatin states \cite{Hoffman2012} (Supplementary Figure \ref{fig:snplocations}).
+
+We also demonstrate physical organisation of interacting loci within the cell, suggesting a mechanism by which biological function can lead to epistatic genetic variance.
 It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
@@
 \subsection*{Methods Summary}
-We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in peripheral blood samples for 7,339 probes representing 6,158 RefSeq genes.
+We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs.
+Each individual had gene expression levels measured in peripheral blood at 47,323 probes.
+Only the probes that passed quality control and had significant expression in $\geq 90\%$ of individuals were used in the analysis (7,339 probes representing 6,158 RefSeq genes).
 Recent hardware and software \cite{Hemani2011} advances that use graphics processing units (GPUs) made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
@@
 Phenotypes are for gene transcript levels (dark coloured tiles = high expression, light coloured tiles = low expression).
-Columns of GP maps are for each independent population.
+Columns of GP maps are for each independent dataset.
 Rows of GP maps are for each of 30 significantly replicated interactions at the Bonferroni level, corresponding to the rows in Table \ref{tab:bonferroni}.
@@
 	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).
-Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data, but the remaining 89 interactions that did not replicate are depicted in grey.
+Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data The remaining 89 interactions that did not replicate are depicted in grey.
 It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
@@
 	\centering
+	\includegraphics[width=5in]{qqgg}
+	\caption{\textbf{Q-Q plot of interaction $p$-values in the CDHWB dataset} Twenty of the 501 discovery SNP pairs passed filtering in the CDHWB dataset (mainly due to small sample size).
+There is no evidence for enrichment of interaction terms, most likely due to insufficient power given the limited sample size.}
+	\label{fig:qqgg}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\centering
 	\includegraphics[width=5in]{MBNL1_3D}
 	\caption{\textbf{Genotype-phenotype maps for 14 interactions influencing the expression of MBNL1} Each bar represents the mean phenotypic value for individuals in that genotype class.
-The rs13069559 SNP typically has a \emph{cis}-additive decreasing effect on the expression of MBNL1, but in many of these interactions the \emph{cis} effect is masked when the \emph{trans} SNP is homozygous.}
+The rs13069559 SNP typically has a \emph{cis}-additive decreasing effect on the expression of MBNL1, but in many of these interactions the \emph{cis} effect is masked when the \emph{trans} SNP is homozygous for the masking allele.}
 	\label{fig:MBNL1_3D}
@@
 	\caption{\textbf{Genotype-phenotype maps for 19 interactions influencing the expression of TMEM149} Each bar represents the mean phenotypic value for individuals in that genotype class.
-The rs13069559 SNP typically has a \emph{cis}-additive decreasing effect on the expression of TMEM149, but in many of these interactions the \emph{cis} effect is masked when the \emph{trans} SNP is homozygous.}
+The rs13069559 SNP typically has a \emph{cis}-additive decreasing effect on the expression of TMEM149, but in many of these interactions the \emph{cis} effect is masked when the \emph{trans} SNP is homozygous for the masking allele.}
 	\label{fig:TMEM149_3D}
@@
 The number of epistatic interactions within 20kb, 500kb, 2Mb and 10Mb of known chromosome interacting regions are shown by red vertical lines.
-The histograms represent the null distribution based on random sampling of 10,000 datasets for each window size.}
+The histograms represent the null distribution based on random sampling of 1,000 datasets for each window size.}
 	\label{fig:chromosomeinteractions}

### 2013-07-24 docs/presentations/20130724_seattle/seattle_grant_meeting.tex
written presentation for seattle meeting

\documentclass{beamer}
%\usepackage{colortbl}
\usepackage[latin1]{inputenc}
\usepackage{comment}
%\usetheme{Warsaw}
\usetheme{Frankfurt}
% \usecolortheme{dove}
\usepackage[absolute,overlay]{textpos} 
\usepackage{booktabs}
\usepackage{color}
\usepackage{threeparttable}
\usepackage{multirow}
\usepackage{rotating}

%\usepackage[table]{xcolor}
%\definecolor{tableShade}{HTML}{F1F5FA}

\title[Epistasis in expression]{Detection and replication of epistasis influencing human transcription}

\author{Gibran Hemani \and
Konstantin Shakhbazov \and \\
Grant W Montgomery \and
Peter M Visscher \and
Joseph E Powell}

\institute{Queensland Brain Institute, University of Queensland \and 
University of Queensland Diamantina Institute}

\date{}

\begin{document}

\AtBeginSection[]
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[currentsection,currentsubsection]
  \end{frame}
}

\begin{frame}
\titlepage
\end{frame}

\section{Epistasis}
\subsection{}

\begin{frame}{Epistasis}
	\begin{definition}
		The effect on the phenotype caused by locus A depends on the genotype at locus B
	\end{definition}
\end{frame}

\begin{frame}{Two dimensional GWAS}
	\begin{center}
		\includegraphics[width=7cm]{2dscan.png} \\
		{\tiny Carlborg 2004}
	\end{center}
\end{frame}

\begin{frame}{Impact of LD on detecting epistasis}
	\begin{center}
		\includegraphics[width=5cm]{gpmaps_ld.png} \\
	\end{center}
\end{frame}

\begin{frame}{Multiple testing problem}
	\begin{block}{Curse of dimensionality}
		As the dimensionality of the search increases the background noise drowns out all real biological signals
	\end{block}

	\begin{equation}
		N_{\textrm{tests}} = \frac{m \times (m - 1)}{2} \nonumber
	\end{equation}
	\emph{e.g.} $500000k$ SNPs $\rightarrow 1.25 \times 10^{11}$ tests
\end{frame}

\section{Study design}
\subsection{}

\begin{frame}{Expression traits likely have larger effect sizes}
	\begin{center}
		\includegraphics[width=7cm]{broadgp}
	\end{center}
\end{frame}

\begin{frame}{Discovery data}
	\begin{itemize}
		\item BSGS data - 842 individuals
		\item Gene expression on whole blood
		\item 7339 traits with $n \geq 90\%$
		\item 528,509 SNPs
	\end{itemize}
\end{frame}

\begin{frame}{Replication data}
	\begin{itemize}
		\item Fehrmann
		\begin{itemize}
			\item $n = 1240$
			\item Identical SNP chip and expression chip
		\end{itemize}
		\item EGCUT
		\begin{itemize}
			\item $n = 891$
			\item Identical SNP chip and expression chip
		\end{itemize}
		\item CHDWB
		\begin{itemize}
			\item $n = 139$
			\item Different SNP chip, same expression chip
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Computation}
	\begin{block}{Total number of tests}
		528,509 pairwise SNPs $\times$ 7,339 traits $=$ 1.02 quadrillion tests
	\end{block}
	\begin{block}{epiGPU software}
		Performs $\sim$12 million association tests per second
	\end{block}
	\begin{block}{GPU clusters}
		Supercomputers with 10s or 100s of GPUs can do this in a few weeks
	\end{block}
\end{frame}

\begin{frame}{Analysis outline}
	\begin{enumerate}
		\item Discovery scan
		\item Filtering of results based on threshold etc
		\item Filtering based on interaction vs genetic effects
		\item Replication in independent samples
	\end{enumerate}
\end{frame}

\begin{frame}{Discovery and filtering}
	\begin{block}{}
		Perform 8 d.f. test for full genetic effect (additive + dominance + epistasis) at each SNP pair
	\end{block}
	\begin{enumerate}
		\item Significance threshold $T = 2.91 \times 10^{-16}$
		\item Remove SNP pairs with any class size $< 5$
		\item Remove SNP pairs with LD $r^2 > 0.1$ or $D' > 0.1$
		\item Keep the sentinel SNP pair for each chromosome $\times$ chromosome $\times$ trait
		\item 11155 SNP pairs remain
		\item Perform nested test of full genetic model (8 d.f.) vs marginal model (a + d, 4 d.f.)
		\item Keep 4 d.f. interaction effects with $p < 0.05 / 11155$
	\end{enumerate}
	\begin{block}{}
		501 significant interaction SNP pairs
	\end{block}
\end{frame}

\section{Results}
\subsection{}

\begin{frame}{501 significant interactions in 238 expression traits}
	\begin{block}{Genomic positions}
		\begin{itemize}
			\item 47 \emph{cis}-\emph{cis}
			\item 441 \emph{cis}-\emph{trans}
			\item 13 \emph{trans}-\emph{trans}
		\end{itemize}
	\end{block}
	\pause
	\begin{block}{Marginal effects ($p < 1.0 \times 10^{-10}$)}
		\begin{itemize}
			\item 9 between two main effects
			\item 428 with only one main effect
			\item 64 with no main effects
		\end{itemize}
	\end{block}
	\pause
	\begin{block}{Largest epistatic variance component}
		\begin{itemize}
			\item 120 A x A
			\item 255 A x D
			\item 126 D x D
		\end{itemize}
	\end{block}
\end{frame}

\begin{frame}{Replication}
	\begin{itemize}
		\item Only 20 SNP pairs passed filtering in CHDWB
		\item 434 SNPs pairs passed QC in both EGCUT and Fehrmann
		\item 30 were significant for interaction $p$-values ($p < 0.05/434$) in EGCUT and Fehrmann
	\end{itemize}
\end{frame}

% q-q plots
\begin{frame}{Q-Q plots of replication interaction $p$-values}
	\begin{center}
		\includegraphics[height=7cm]{qqMeta}
	\end{center}
\end{frame}

\begin{frame}{Bonferroni level replicated GP maps}
	\begin{center}
		\includegraphics[width=11cm]{gpBonfRep}
	\end{center}
\end{frame}

\begin{frame}{Map of interactions}
	\begin{center}
		\includegraphics[height=7.5cm]{pale_gray_graph_of_interactions_2_lists}
	\end{center}
\end{frame}

\begin{frame}{TMEM149}
	\begin{center}
		\includegraphics[height=7.5cm]{TMEM149_3d}
	\end{center}
\end{frame}

\begin{frame}{Chromosome interactions}
	\begin{center}
		\includegraphics[height=7cm]{chromosome_interactions}
	\end{center}
\end{frame}

\begin{frame}{Contribution relative to additive effects}
	\begin{block}{At the same threshold ($2.91 \times 10^{-16}$)}
		\begin{itemize}
			\item 453 expression traits have a significant additive effect
			\item 238 have a significant interaction effect
		\end{itemize}
	\end{block}
	\pause
	\begin{block}{At the same threshold ($2.91 \times 10^{-16}$)}
		\begin{itemize}
			\item Significant additive effects explain 1.73\% of phenotypic variance of 7339 traits
			\item Significant epistatic effects explain 0.25\% (seven times less)
		\end{itemize}
	\end{block}
\end{frame}

\section{Acknowledgements}
\subsection{}

\begin{frame}{Acknowledgements}
	\begin{columns}[c]
		\column{0.5\textwidth}
			\begin{itemize}
				\item University of Queensland
				\begin{itemize}
					\item Joseph Powell
					\item Konstantin Shakhbazov
					\item Allan McRae
					\item Jian Yang
					\item Peter Visscher
				\end{itemize}
				\item QIMR
				\begin{itemize}
					\item Grant Montgomery
					\item Nicholas Martin
					\item Anjali Henders
				\end{itemize}
				\item Estonian Genomics Centre
				\begin{itemize}
					\item Tonu Esko
					\item Andres Metspalu
				\end{itemize}
			\end{itemize}
		\column{0.5\textwidth}
			\begin{itemize}
				\item Georgia Tech University
				\begin{itemize}
					\item Greg Gibson
				\end{itemize}
				\item University of Groningen
				\begin{itemize}
					\item Harm-Jan Westra
					\item Lude Franke
				\end{itemize}
				\item Computer resources
				\begin{itemize}
					\item iVEC at University of Western Australia
					\item MASSIVE project
					\item QBI IT Team
				\end{itemize}
			\end{itemize}
	\end{columns}
\end{frame}

\end{document}

### 2013-08-21 docs/protocols_ms/epistasis_protocols.tex
added first draft of a protocols manuscript

\documentclass{article}
\usepackage[superscript,biblabel]{cite}
\usepackage{graphicx}
\usepackage{float}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage{authblk}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{lscape}

\makeatletter
\renewcommand\AB@affilsepx{. \protect\Affilfont}
\makeatother

%\graphicspath{{../../analysis/images/}}
\bibliographystyle{naturemag}

\title{How to find epistasis, Bitches!}
\date{}

\author[1,2]{Gibran Hemani}
\author[1,2]{Joseph E Powell}

\affil[1]{University of Queensland Diamantina Institute, University of Queensland, Princess Alexandra Hospital, Brisbane, Queensland, Australia}
\affil[2]{Queensland Brain Institute, University of Queensland, Brisbane, QLD, Australia}

\begin{document}

\maketitle

\clearpage

\section{Abstract}

\clearpage

\section{Introduction}

Epistasis is the process whereby the effect of one genetic loci on a phenotype is modified by the genotypes carried at other unlinked loci. As a topic in human genetics it has been widely discussed but rarely investigated in depth due to a series of computational and statistical challenges. Recently, a number of these challenges have been overcome and we have recently shown that through careful application of these methods (wide scale) epistasis can be detected for gene expression phenotypes in humans (ref). [I think we should try and say something about the processes needed to interoperate the output of a epistasis analysis]

\subsection{Challenges}

[Initially outline the challenges, then address each in turn] \\
\newline
1. Computational \\ 
\newline
[I think it would be worth re-discussing some of the computational problems of 2d scans. and highlighting the software + hardware solutions] \\
\newline
One of the first obstacles that make epistatic searches difﬁcult is the computational demands of the statistical analyses. When searching for independent additive effects, as is done for the majority of GWA studies, each SNP is tested for association with the phenotype. However, in order to most powerfully identify epistatic effects, the search must be increased to multiple dimensions (ref). Here the scale of the computational demand increases by $~(n^x)/2$ where $n$ is the numer of snps and $x$ is the number of epistatsis dimensions fitted. For example, testing for 2 loci interactions using SNP data from a 1 million SNP chip would require $1000000 * 999999/2 \approx 5e11$ individual tests. \\     
\newline 
2. Model choice \\
\newline
From a 2d 8df biallelic SNP model there are 4 epistatic variance components; additive x additive, additive x
dominance, dominance x additive and dominance x dominance. Within the literature there has been considerable discussion as to the likely degree of epistatsis variance components and whether statistical power could be increased by parameterizing models for only a subset. \\  
\newline
3. Statistical \\
A complete exhaustive scan of $m$ phenotypes and $n$ SNPs comprises of $((n*(n-1))^2/2)*m$ $8df$ F-tests. Given the high correlation structure inherent in genotype data as well as between multiple phenotypes, choices regarding multiple testing need to be carefully considered. [what can we say about tails of the distributions?]  \\
\newline
4. Interpretation / filtering

The total output from an

\end{document}

### 2013-08-21 docs/protocols_ms/epistasis_protocols.tex
edits to first draft of a protocols manuscript

@@
 
-[Initially outline the challenges, then address each in turn] \\
+[Initially outline the challenges in the introduction. These can be address through the methods section] \\
 \newline
@@
 Within the literature there has been considerable discussion as to the likely degree of epistatsis variance components and whether statistical power could be increased by parameterizing models for only a subset.
+We have shown then estimates of parameters repressenting the 4 variance components are proportionally represented amongst the significant epistatis pairs.
+This implys a full model is preferential, particularly if we are agonostic about the mechanism by which epistasis might arise.
 \\
@@
 3. Statistical \\
+\newline
 A complete exhaustive scan of $m$ phenotypes and $n$ SNPs comprises of $((n*(n-1))^2/2)*m$ $8df$ F-tests.
 Given the high correlation structure inherent in genotype data as well as between multiple phenotypes, choices regarding multiple testing need to be carefully considered.
-[what can we say about tails of the distributions?]  \\
+[what can we say about tails of the distributions?]
+[Perhaps this is a good point to outline out a 2-step procedure where main effects are initially corrected for] \\
+The 8df F-test includes parameters for main (additive) effects of the two SNPs.
+Therefore it is important to test that test statistic isn't driven my marginal effects - something that is quite likely for a trait such as gene expression.
+Thus one should really consider a nested test.
+\\
 \newline
-4. Interpretation / filtering
+4. Interpretation / filtering \\
+\newline
+Things to consider and address during the analysis and output, particularly given    
+\newline
+\section{Methods}
 
-The total output from an
+I think one approach for this paper (which is in keeping with other Nature Protocol manuscripts) is to go through the whole study design and show how we have done things - discussing choices in terms of the challenges mentioned above.
+For many steps we can show example (pseudo) code or give it in the supporting material
+
+\subsection{Stages of an epistasis analysis}
+
+\subsubsection{Data}
+
+QC \\
+Normalisation (distributon with respect to models used) \\
+Statistical model choice \\
+
+\subsubsection{Computational / Statistics}
+
+\subsubsection{Multiple testing}
 

### 2013-08-21 docs/protocols_ms/epistasis_protocols.tex
edits to first draft of a protocols manuscript

@@
 However, in order to most powerfully identify epistatic effects, the search must be increased to multiple dimensions (ref).
-Here the scale of the computational demand increases by $~(n^x)/2$ where $n$ is the numer of snps and $x$ is the number of epistatsis dimensions fitted.
+Here the scale of the computational demand increases by $~(n^x)/2$ where $n$ is the number of SNPs and $x$ is the number of epistasis dimensions fitted.
 For example, testing for 2 loci interactions using SNP data from a 1 million SNP chip would require $1000000 * 999999/2 \approx 5e11$ individual tests.
@@
 dominance, dominance x additive and dominance x dominance.
-Within the literature there has been considerable discussion as to the likely degree of epistatsis variance components and whether statistical power could be increased by parameterizing models for only a subset.
-We have shown then estimates of parameters repressenting the 4 variance components are proportionally represented amongst the significant epistatis pairs.
-This implys a full model is preferential, particularly if we are agonostic about the mechanism by which epistasis might arise.
+Within the literature there has been considerable discussion as to the likely degree of epistasis variance components and whether statistical power could be increased by parameterizing models for only a subset.
+We have shown then estimates of parameters representing the 4 variance components are proportionally represented amongst the significant epistatic pairs.
+This implies a full model is preferential, particularly if we are agonistic about the mechanism by which epistasis might arise.
 \\
@@
 \newline
-Things to consider and address during the analysis and output, particularly given    
+Things to consider and address during the analysis and output, particularly given the massive multiple testing and inability to easily evaluate the tails of the test-statistic distribution.
+Why we require at least 5 (or more) individuals in the smallest classt size for the genotypes.
+The choice of number is partly based total sample size.
+Assuming Hardy-Weinberg equilibrium, the frequency of the smallest class size is $p_1^2 * p_2^2$ where $p_1$ and $p_2$ are the frequencies of the minor alleles for SNPs 1 and 2 respectivly.
 \newline
@@
 
-QC \\
-Normalisation (distributon with respect to models used) \\
-Statistical model choice \\
+QC - probably want to use a hight MAF cut-off than a standard GWAS in order to as anything with low MAF is going to increase the probability of $p_1^2 * p_2^2 * n$ being less than a required level. \\
+Normalisation (distribution with respect to models used).
+Normal data.
+Is it worth discussing the potential problems caused by a skew in $y$ and its effect on the tail of the test statistic distribution?
+\\
 
@@
 
+Reiterate the distinction between CPU and GPU.
+I guess we can re-calculate some of the speed differences based on current GPU hardware and show speed difference with relation to cpu based software.
+(God place for a figure)
+
+Outline the 2 stage models; 8df followed by (nested) 4df; or initial correction for main effects followed by a interection term.
+This would also be the place to point out the importance of a) running a full 2d scan rather than initially search for marginal effects and b) including all epistasis variance component parameters in the interection model.
+
 \subsubsection{Multiple testing}
 
+\subsubsection{Potential artifacts}
+
 \end{document}

### 2013-08-21 docs/protocols_ms/epistasis_protocols.tex
minor edits to first draft of a protocols manuscript

@@
 Things to consider and address during the analysis and output, particularly given the massive multiple testing and inability to easily evaluate the tails of the test-statistic distribution.
-Why we require at least 5 (or more) individuals in the smallest classt size for the genotypes.
+Why we require at least 5 (or more) individuals in the smallest class size for the genotypes.
 The choice of number is partly based total sample size.
-Assuming Hardy-Weinberg equilibrium, the frequency of the smallest class size is $p_1^2 * p_2^2$ where $p_1$ and $p_2$ are the frequencies of the minor alleles for SNPs 1 and 2 respectivly.
+Assuming Hardy-Weinberg equilibrium, the frequency of the smallest class size is $p_1^2 * p_2^2$ where $p_1$ and $p_2$ are the frequencies of the minor alleles for SNPs 1 and 2 respectively.
 \newline
@@
 
-Outline the 2 stage models; 8df followed by (nested) 4df; or initial correction for main effects followed by a interection term.
-This would also be the place to point out the importance of a) running a full 2d scan rather than initially search for marginal effects and b) including all epistasis variance component parameters in the interection model.
+Outline the 2 stage models; 8df followed by (nested) 4df; or initial correction for main effects followed by a interaction term.
+This would also be the place to point out the importance of a) running a full 2d scan rather than initially search for marginal effects and b) including all epistasis variance component parameters in the interaction model.
 

### 2013-08-21 docs/protocols_ms/epistasis_protocols.tex
added methods sections to first draft of a protocols manuscript

@@
 
-QC - probably want to use a hight MAF cut-off than a standard GWAS in order to as anything with low MAF is going to increase the probability of $p_1^2 * p_2^2 * n$ being less than a required level. \\
+QC - probably want to use a high MAF cut-off than a standard GWAS in order to as anything with low MAF is going to increase the probability of $p_1^2 * p_2^2 * n$ being less than a required level. \\
 Normalisation (distribution with respect to models used).
@@
 I guess we can re-calculate some of the speed differences based on current GPU hardware and show speed difference with relation to cpu based software.
-(God place for a figure)
+(Good place for a figure)
 
@@
 
+How to deal with multiple testing [what is the current state of your manuscript?].
+Using GPUs it's not unrealistic that a researcher with access to a GPU cluster could run a permutation analysis for a single phenotype 2D scan.
+However, that should compare to your theoretically derived threshold.
+Therefore, the problem lies when we have multiple, correlated phenotypes such as expression.
+
 \subsubsection{Potential artifacts}
 
+Remove SNP pairs that are in LD greater than (threshold). Could lead in appearance of effects. Threshold determined based on....?
+
+Population stratification and cryptic relatedness - look for inflation of LD between snp pairs.
+
+\section{Discussion}
+
 \end{document}

### 2013-08-21 docs/protocols_ms/epistasis_protocols.tex
updated doc added

@@
 As a topic in human genetics it has been widely discussed but rarely investigated in depth due to a series of computational and statistical challenges.
-Recently, a number of these challenges have been overcome and we have recently shown that through careful application of these methods (wide scale) epistasis can be detected for gene expression phenotypes in humans (ref).
+In the last few years a number of these challenges have been overcome and we have recently shown that through careful application of these methods (wide scale) epistasis can be detected for gene expression phenotypes in humans (ref).
 [I think we should try and say something about the processes needed to interoperate the output of a epistasis analysis]
@@
 
-[Initially outline the challenges in the introduction. These can be address through the methods section] \\
+[Initially outline the challenges in the introduction. These can be addressed through the methods section] \\
 \newline
@@
 dominance, dominance x additive and dominance x dominance.
-Within the literature there has been considerable discussion as to the likely degree of epistasis variance components and whether statistical power could be increased by parameterizing models for only a subset.
-We have shown then estimates of parameters representing the 4 variance components are proportionally represented amongst the significant epistatic pairs.
+Within the literature there has been considerable discussion as to the likely degree of epistasis variance components and whether statistical power could be increased by parameterizing models for only a subset of these parameters.
+We have shown that estimates of parameters representing the 4 variance components are proportionally represented amongst the significant epistatic pairs.
 This implies a full model is preferential, particularly if we are agonistic about the mechanism by which epistasis might arise.

### 2013-08-31 docs/manuscript/nature_letter.tex
manuscript has been submitted to nature
more stuff on signs

@@
 \clearpage
-\section{Tables}
+\section*{Tables}
 
@@
 \clearpage
-\section{Figures}
+\section*{Figures}
 
@@
 	\centering
-	\includegraphics[width=2.5in]{gpbonfrep.pdf}
+	% \includegraphics[width=2.5in]{gpbonfrep.pdf}
 	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
@@
 \end{figure}
-\clearpage
 
@@
 	\centering
-	\includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
-	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).
-Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data The remaining 89 interactions that did not replicate are depicted in grey.
-It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
-	\label{fig:fireworks}
-\end{figure}
-\clearpage
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=2.5in]{qqMeta.pdf}
+	% \includegraphics[width=2.5in]{qqMeta.pdf}
 	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} The top panel shows all 434 discovery SNPs that were tested for interactions.
@@
 \end{figure}
-\clearpage
-
-\clearpage
-\section{Supplementary Figures}
+
+\begin{figure}[H]
+	\centering
+	% \includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
+	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).
+Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data The remaining 89 interactions that did not replicate are depicted in grey.
+It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
+	\label{fig:fireworks}
+\end{figure}
+
+\clearpage
+
+\bibliography{refs}
+\clearpage
+
+\section*{Supplementary Figures}
 \setcounter{figure}{0}
@@
 \clearpage
-\section{Supplementary Tables}
+\section*{Supplementary Tables}
 \setcounter{table}{0}
@@
 
-\clearpage
-% \section{References}
-\bibliography{refs}
+\begin{table}[ht]
+\begin{threeparttable}
+\caption{\textbf{Concordance of sign of epistatic variance components between discovery and replication datasets}}
+\label{tab:sign}
+\begin{tabular}{lllrrrr}
+
+  \hline
+Test & Interactions & Dataset & $n$ & Expected & Observed & $p$-value \\
+  \hline
+1\tnote{a} & All & EGCUT & 434 & 217.00 & 306 & $6.69 \times 10^{18}$ \\
+&  & Fehrmann & 434 & 217.00 & 278 & $5.04 \times 10^{09}$ \\
+&  & Both & 434 & 108.50 & 221 & $5.56 \times 10^{31}$ \\
+& Significant & EGCUT & 30 & 15.00 & 25 & $3.25 \times 10^{04}$ \\
+&  & Fehrmann & 30 & 15.00 & 24 & $1.43 \times 10^{03}$ \\
+&  & Both & 30 & 7.50 & 22 & $3.76 \times 10^{08}$ \\
+  \hline
+2\tnote{b} & All & EGCUT & 434 & 54.25 & 92 & $4.22 \times 10^{07}$ \\
+&  & Fehrmann & 434 & 54.25 & 79 & $6.18 \times 10^{04}$ \\
+&  & Both & 434 & 6.78 & 30 & $2.55 \times 10^{11}$ \\
+& Significant & EGCUT & 30 & 3.75 & 19 & $9.46 \times 10^{11}$ \\
+&  & Fehrmann & 30 & 3.75 & 19 & $9.46 \times 10^{11}$ \\
+&  & Both & 30 & 0.47 & 18 & $2.23 \times 10^{25}$ \\
+  \hline
+3\tnote{c} & All & EGCUT & 434 & 27.12 & 34 & $1.65 \times 10^{01}$ \\
+&  & Fehrmann & 434 & 27.12 & 35 & $1.35 \times 10^{01}$ \\
+&  & Both & 434 & 1.70 &  2 & $6.89 \times 10^{01}$ \\
+& Significant & EGCUT & 30 & 1.88 &  8 & $3.92 \times 10^{04}$ \\
+&  & Fehrmann & 30 & 1.88 &  9 & $6.22 \times 10^{05}$ \\
+&  & Both & 30 & 0.12 &  1 & $1.11 \times 10^{01}$ \\
+  \hline
+4\tnote{d} & All & EGCUT & 1133 & 566.50 & 775 & $7.10 \times 10^{36}$ \\
+&  & Fehrmann & 1133 & 566.50 & 726 & $1.90 \times 10^{21}$ \\
+&  & Both & 1133 & 283.25 & 562 & $1.39 \times 10^{70}$ \\
+& Significant & EGCUT & 73 & 36.50 & 55 & $1.69 \times 10^{05}$ \\
+&  & Fehrmann & 73 & 36.50 & 55 & $1.69 \times 10^{05}$ \\
+&  & Both & 73 & 18.25 & 46 & $7.86 \times 10^{12}$ \\
+   \hline
+\end{tabular}
+
+\begin{tablenotes}
+\item[a] The sign of the most significant epistatic variance component in discovery is the same as the corresponding variance component in the replication data.
+\item[b] The largest epistatic variance component in the discovery is the same as in the replication with the same sign in both.
+\item[c] The sign of all four epistatic variance components are identical in the discovery and the replication.
+\item[d] The sign of all epistatic variance components in the discovery with $p < 0.05$ are the same as the corresponding variance components in the replication data.
+\end{tablenotes}
+
+\end{threeparttable}
+\end{table}
 

### 2013-08-31 analysis/null_tests.R

@@ -1,0 +1,115 @@
+library(plyr)
+
+####
+
+# Take a SNP / probe combination in replication dataset
+# Choose another SNP randomly from remaining SNPs
+# Check that it doesn't already exist as a real signal
+# Perform test
+# Do this 1000 times
+# Keep the first 434
+
+
+modifySig <- function(sig)
+{
+	sig$comb1 <- with(sig, paste(probename, snp1))
+	sig$comb2 <- with(sig, paste(probename, snp2))
+	sig$comb3 <- with(sig, paste(snp1, snp2))
+	sig$comb4 <- with(sig, paste(snp2, snp1))
+	return(sig)
+}
+
+
+chooseRandomSnpPair <- function(sigm)
+{
+	sig <- sigm
+	flag <- 1
+	while(flag == 1)
+	{
+		a <- sample(1:nrow(sig), 2, replace=FALSE)
+		ran1 <- a[1]
+		ran2 <- a[2]
+		probename <- sig$probename[ran1]
+		snp1 <- sig$snp1[ran1]
+		snp2 <- sig$snp2[ran2]
+		t1 <- paste(probename, snp1)
+		t2 <- paste(probename, snp2)
+		t3 <- paste(snp1, snp2)
+		t4 <- paste(snp2, snp1)
+		if(
+			t2 %in% sig$comb2 |
+			t3 %in% sig$comb3 |
+			t4 %in% sig$comb4 |
+			snp1 == snp2
+		) {
+			flag <- 1
+		} else {
+			flag <- 0
+		}
+	}
+	return(list(probename, snp1, snp2))
+}
+
+
+replicationTests <- function(genolist, phenlist, ids)
+{
+	a <- list()
+
+	for(i in 1:3)
+	{
+		geno <- genolist[[i]]
+		probes <- phenlist[[i]]
+		snp1 <- geno[, colnames(geno) == ids[[2]]]
+		snp2 <- geno[, colnames(geno) == ids[[3]]]
+		probe <- probes[, colnames(probes) == ids[[1]]]
+
+		# Statistical tests
+		fullmod <- lm(probe ~ as.factor(snp1) * as.factor(snp2))
+		margmod <- lm(probe ~ as.factor(snp1) + as.factor(snp2))
+		inttest <- anova(margmod, fullmod)
+		a[[i]] <- inttest$P[2]
+	}
+	p <- data.frame(c(ids, a))
[…]

### 2013-08-31 analysis/organise_rep.R

@@ -489,7 +489,7 @@ pairs(f3[,2:4])
 ##
 
 genolist <- gl3
-save(genolist, phenlist, file="bsgs_egcut_fehr_data.RData")
+save(genolist, phenlist, file="~/repo/eQTL-2D/data/bsgs_egcut_fehr_data.RData")
 
 noia <- doAllTests(genolist, phenlist, sig, 1)
 bnoia <- doAllTests(genolist, phenlist, bsig, 1)

### 2013-08-31 analysis/qqplot.R

@@ -136,3 +136,17 @@ qqplot3 <- ggplot(subset(dat3, filter != 3)) +
 	scale_colour_brewer(type="qual", palette=3) +
 	theme(legend.position="none")
 ggsave(qqplot3, file="~/repo/eQTL-2D/analysis/images/qqgg.pdf")
+
+
+
+####
+
+# Take a SNP / probe combination in replication dataset
+# Choose another SNP randomly from remaining SNPs
+# Check that it doesn't already exist as a real signal
+# Perform test
+# Do this 1000 times
+# Keep the first 434
+
+
+

### 2013-08-31 analysis/sign_enrichment.R

@@ -1,5 +1,6 @@
 library(noia)
 library(plyr)
+library(xtable)
 
 
 # Largest discovery has the same sign in replication
@@ -30,8 +31,11 @@ test1 <- function(noiad)
 	l <- list()
 	l$els <- els
 	l$p1 <- p1
-	l$p1 <- p2
+	l$p2 <- p2
 	l$p12 <- p12
+	l$e1 <- 0.5 * nrow(els)
+	l$e2 <- 0.5 * nrow(els)
+	l$e12 <- 0.25 * nrow(els)
 	return(l)
 }
 
@@ -62,8 +66,11 @@ test2 <- function(noiad)
 	l <- list()
 	l$els <- els
 	l$p1 <- p1
-	l$p1 <- p2
+	l$p2 <- p2
 	l$p12 <- p12
+	l$e1 <- 1/8 * nrow(els)
+	l$e2 <- 1/8 * nrow(els)
+	l$e12 <- 1/64 * nrow(els)
 	return(l)
 }
 
@@ -79,21 +86,25 @@ test3 <- function(noiad)
 		for(i in 2:3)
 		{
 			y <- subset(x, dataset == i)
-			r[[i-1]] <- sum(s == sign(y$E[5:8]))
+			r[[i-1]] <- sum(s == sign(y$E[5:8])) == 4
 		}
 
 		out <- data.frame(r)
 		names(out) <- c("r1", "r2")
 		return(out)
 	})
-	p1 <- binom.test(n = nrow(els), x = sum(els$r1 == 4), p = 1/16)$p.value
-	p2 <- binom.test(n = nrow(els), x = sum(els$r2 == 4), p = 1/16)$p.value
-	p12 <- binom.test(n = nrow(els), x = sum(els$r1 == 4 & els$r2 == 4), p = 1/256)$p.value
+	p1 <- binom.test(n = nrow(els), x = sum(els$r1), p = 1/16)$p.value
+	p2 <- binom.test(n = nrow(els), x = sum(els$r2), p = 1/16)$p.value
+	p12 <- binom.test(n = nrow(els), x = sum(els$r1 & els$r2), p = 1/256)$p.value
+
 	l <- list()
 	l$els <- els
 	l$p1 <- p1
-	l$p1 <- p2
+	l$p2 <- p2
 	l$p12 <- p12
+	l$e1 <- 1/16 * nrow(els)
+	l$e2 <- 1/16 * nrow(els)
+	l$e12 <- 1/256 * nrow(els)
[…]

### 2013-09-01 docs/manuscript/nature_letter.tex
added the diff files

--- 2013-07-06 docs/manuscript/nature_letter.tex
@@
 This work could not have been completed without access to high performance GPGPU compute clusters.
-We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au).
+We acknowledge iVEC for the use of advanced computing resources located at iVEC\@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au).
 We also thank Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.
@@
 van Dijk and the Adessium Foundation.
-The research leading to these results has received funding from the European Community’s Health Seventh Framework Programme (FP7/2007-2013) under grant agreement 259867.
+The research leading to these results has received funding from the European Community's Health Seventh Framework Programme (FP7/2007-2013) under grant agreement 259867.
 
@@
 
-\clearpage
-% \section{References}
-\bibliography{refs}
-
 \end{document}

### 2013-09-01 docs/manuscript/methods.tex
most recent version

@@
 \begin{equation}
-G_{ijk} = \left \{ 
+G_{jk} = \left \{ 
 \begin{array}{ll}
-\sigma _a ^2 + \sigma _e ^2&        j = k \\ 
+\sigma _a ^2             &          j = k \\ 
 2\phi _{ijk} \sigma _a ^2& 			j \neq k \\
@@
 \end{equation}
-The parameters $\sigma_a^2$ and $\sigma_e^2$ are variance components for additive background genetic and environmental effects respectively.
+The parameter $\sigma_a^2$ is the variance component for additive background genetic.
 Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$.
@@
 \section{Exhaustive 2D-eQTL analysis}
+
+\subsection{Two stage search}
 
@@
 If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained.
-At this stage 6404 filtered SNP pairs remained.
-We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs.
+At this stage 6404 filtered SNP pairs remained (filter 1).
+We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs (filter 2).
 
@@
 All variance decompositions were calculated using the NOIA method \cite{Alvarez-Castro2008}.
+
+\subsection{Type 1 error rate}
+
+Using a Bonferroni correction of $0.05$ in the second stage of the two stage discovery scan implies a type 1 error rate of $\alpha = 0.05$.
+However, this could be underestimated because the number tests performed in the second stage depends on the number of tests in the first stage, and this depends on statistical power and model choice.
+We performed simulations to estimate the type 1 error rate of this study design.
+
+We assumed a null model where there was one true additive effect and 7 other terms with no effect.
+To simulate a test statistic we simulated 8 z-scores, $z_1 \sim N(\sqrt{NCP}, 1)$ and $z_{2..8} \sim N(0,1)$.
+Thus $z_{full} = \sum^8_{i=1}{z_i} \sim \chi^{2}_{8}$ (representing the 8 d.f.
+test) and $z_{int} = \sum^8_{i=5}{z_i} \sim \chi^{2}_{4}$ (representing the 4 d.f.
+test where the null hypothesis of no epistasis is true).
+For a particular value of $NCP$ we simulated 100,000 $z$ values, and calculated the $p_{full}$-value for the $z_{full}$ test statistic.
+The $n_{int}$ test statistics with $p_{full} < 2.31 \times 10^{-16}$ were kept for the second stage, where the type 1 error rate of stage 2 was calculated as the proportion of $p_{int} < 0.05 / n_{int}$.
+The power at stage 1 was calculated as $n_{int} / 100,000$.
+This procedure was performed for a range of $NCP$ parameters that represented power ranging from $\sim 0$ to $\sim 1$.
 
@@
 As in the discovery analysis, all gene expression levels were normalised using rank transformation to avoid skew or outliers in the distribution \cite{Westra2011}.
+
+\subsection{Concordance of direction of effects}
+
+We used four methods to calculate the concordance of the direction of effects between the discovery and replication datasets.
+
+\begin{enumerate}
+
+\item [Test 1] Is the most significant epistatic effect in the discovery set in the same direction as the same epistatic effect in the replication sets?
+We decomposed the genetic variance into 8 orthogonal effects, four of which are epistatic ($A \times A$, $A \times D$, $D \times A$, $D \times D$).
+The sign of the epistatic effect that had the largest variance in the discovery was recorded, and then was compared to the same epistatic effect in the two replication datasets (regardless of whether or not the same epistatic effect was the largest in the replication datasets).
+The probability of the sign being the same in one dataset is $1/2$.
+The probability of the sign being the same in two is $1/4$.
+
+\item [Test 2] Is the most significant epistatic effect in the discovery the same as the largest epistatic effect in the replication set with the sign being concordant.
+As in Test 1, but this time we required that the largest effect was the same in the discovery and the replication, and that they had the same sign (\emph{e.g.} if the largest effect in the discovery is $A \times A$, with a positive effect, then concordance is achieved if the same is true in the replication).
+The probability of one replication dataset being concordant by chance is $1/8$, and concordance in both is $1/64$.
+
+\item [Test 3] Are all 4 epistatic effects in the same direction in the discovery and the replication datasets?
+This is a very conservative test as it is not realistic to assume that all four epistatic effects are significant in every interaction.
+The probability of concordance in one replication dataset is $1/16$, and the probability of concordance in both replication datasets is $1/256$.
+
+\item [Test 4] Do the epistatic effects that are significant at nominal $p < 0.05$ in the discovery have the same direction of effect as in the replication?
+Here we count all the epistatic variance components in the discovery that have $p < 0.05$ (1133 amongst the 434 discovery SNP pairs, $i.e.$ each SNP pair has at least 1 and at most 4 significant epistatic variance components).
+Then we compare the direction of the effect in the replication dataset.
+The probability of the sign being the same in one dataset for any one significant effect is $1/2$.
+The probability of the sign being the same in two is $1/4$.
+
+\end{enumerate}
+
+The probability of observing the number of concordant signs is calculated using a binomial test.
 

### 2013-09-01 docs/manuscript/nature_letter.tex
most recent version

--- 2013-08-31 docs/manuscript/nature_letter.tex
@@
 \author[8]{Greg Gibson}
-\author[7]{Nick G Martin}
+\author[7]{Nicholas G Martin}
 \author[4]{Andres Metspalu}
@@
 The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics.
-Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been reported in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
+Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and examples have been reported in other species \cite{Carlborg2006}, few convincing examples with independent replication exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
 Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}.
@@
 In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
-We tested the discovery interactions for replication in two independent data sets \cite{Metspalu2004, Fehrmann2011}.
-Three hundred and forty-five interactions had replication interaction $p$-values that were more extreme than the 2.5\% confidence interval of the distribution under the null hypothesis of no epistasis, with 30 significant at a conservative $p < 0.05$ Bonferroni level.
-There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
+Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31})$ and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 434$.
+There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
 Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs.
 For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans.
+This study presents the first evidence for multiple instances of natural genetic polymorphisms interacting to influence human traits.
 
@@
 But to date, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}, there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}.
-Outside the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
+Beyond the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
 
@@
 Transcription levels can be measured for thousands of genes.
-These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus many genetic effects are relatively large, maximising the chance at detecting epistasis, should it exist.
+These traits are largely heritable but on average less polygenic than high level phenotypes, thus many genetic effects are relatively large \cite{Powell2013}, maximising the chance at detecting epistasis, should it exist.
 
 % \subsection{Initial search in discovery set and replication}
-In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in peripheral blood.
-After stringent filtering and multiple testing correction (5\% significance threshold $ p < 2.91 \times 10^{-16}$, Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}).
+In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions.
+First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (5\% significance threshold $ p < 2.91 \times 10^{-16}$, Methods).
+Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $\alpha \approx 0.07$, Methods, Supplementary Figure \ref{fig:type1}).
+Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}).
 Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
-We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.0001$, Table \ref{tab:bonferroni}).
+We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/434$, Table \ref{tab:bonferroni}).
 These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).
-
-In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the distribution under the null distribution of no epistatic effects ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}).
+We decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect in the independent datasets (Methods).
+The most significant epistatic effect in the discovery sample was in the same direction in both replication datasets in 22 out of the 30 significantly replicated interactions ($p = 3.76 \times 10^{-8}$, Supplementary Table \ref{tab:sign}).
+
+In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}).
+Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}).
 The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets.
@@
 Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst additive $\times$ additive, additive $\times$ dominance, dominance $\times$ additive and dominance $\times$ dominance at the discovery stage ($p = 0.22$ for departure from expectation).
-This is not surprising because the patterns of epistasis used for statistical decomposition are not designed to resemble biological function \cite{Cockerham1954}.
+This is not surprising because the patterns of epistasis used for statistical decomposition (\emph{i.e.} $A \times A$, $A \times D$, $D \times A$, $D \times D$) are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.
 
@@
 It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
-We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
+We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 5Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
 Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Osborne2004, Rieder2012}.
@@
 The BSGS dataset has been previously analysed for additive effects at all expression traits \cite{Powell2012}, and if we take all the additive eQTLs that were significant at the epistatic threshold of $p < 2.91 \times 10^{-16}$ we find that 453 gene expression levels out of the 7339 analysed had at least one significant expression quantitative trait locus (eQTL).
-Therefore it can be argued that the number of instances of detectable epistasis are substantial.
+Therefore it can be argued that the number of instances of detectable epistasis is substantial.
 
@@
 	\centering
+	\includegraphics[width=5in]{type1}
+	\caption{\textbf{Type 1 error rate of two stage design} In stage 1 SNPs are tested for full genetic effects (8 d.f.) and those that surpass a threshold for multiple testing are then tested for significant interaction terms in stage 2.
+These interaction $p$-values are then adjusted (Bonferroni) for the total number of tests that passed stage 1.
+The type 1 error rate of this two stage design is dependent on the power, which is not known empirically.}
+\label{fig:type1}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+	\centering
 	\includegraphics[width=5in]{qqMetaNonsig}
@@
 \begin{ThreePartTable}
-\label{tab:results}
 \begin{longtable}{|llr|lrrl|lrrl|rrrr|r|}
 
-\caption{\textbf{Details on 501 interactions discovered in BSGS dataset}}\\
+\caption{\textbf{Details on 501 interactions discovered in BSGS dataset}}  \label{tab:results} \\
 
@@
 
+\begin{table}[ht]
+{\footnotesize
+\begin{threeparttable}
+\caption{\textbf{Details on linkage disequilibrium and relative positions of all discovery \emph{cis-cis} interactions}}
+\label{tab:ld}
+\begin{tabular}{llllrrrrr}
+
+  \hline
+Chr & Gene & SNP 1 & SNP 2 & Position 1 & Position 2 & Distance / Mb & $R^2$ & $D'$ \\
+  \hline
+19 & TMEM149 & rs807491 & rs7254601 & 36268923 & 36147315.0 & 0.122 & 0.000 & 0.001 \\
+  17 & FN3KRP & rs898095 & rs9892064 & 80890638 & 80827903.0 & 0.063 & 0.063 & 0.088 \\
+  21 & CSTB & rs9979356 & rs3761385 & 45230974 & 45198355.0 & 0.033 & 0.041 & 0.066 \\
+  3 & MBNL1 & rs16864367 & rs13079208 & 152234166 & 152116652.0 & 0.118 & 0.041 & 0.117 \\
+  10 & ADK & rs2395095 & rs10824092 & 76446305 & 75929517.0 & 0.517 & 0.013 & 0.020 \\
+  11 & CTSC & rs7930237 & rs556895 & 88117962 & 88077479.0 & 0.040 & 0.012 & 0.045 \\
+  17 & GAA & rs11150847 & rs12602462 & 78153130 & 78146016.0 & 0.007 & 0.000 & 0.001 \\
+  8 & NAPRT1 & rs2123758 & rs3889129 & 144663661 & 144613680.0 & 0.050 & 0.053 & 0.060 \\
+  1 & LAX1 & rs1891432 & rs10900520 & 203877662 & 203780591.0 & 0.097 & 0.065 & 0.106 \\
+  18 & MBP & rs8092433 & rs4890876 & 74747424 & 74732087.0 & 0.015 & 0.035 & 0.053 \\
+  11 & SNORD14A & rs2634462 & rs6486334 & 17339127 & 17015557.0 & 0.324 & 0.008 & 0.012 \\
+  21 & C21ORF57 & rs9978658 & rs11701361 & 48027084 & 47764477.0 & 0.263 & 0.032 & 0.065 \\
+  16 & RPL13 & rs352935 & rs2965817 & 89648580 & 89513234.0 & 0.135 & 0.054 & 0.060 \\
+  19 & ATP13A1 & rs4284750 & rs873870 & 19810050 & 19738554.0 & 0.071 & 0.008 & 0.015 \\
+  2 & NCL & rs7563453 & rs4973397 & 232301670 & 232291471.0 & 0.010 & 0.027 & 0.029 \\
+  5 & HNRPH1 & rs6894268 & rs4700810 & 179032488 & 178991794.0 & 0.041 & 0.000 & 0.001 \\
+  19 & VASP & rs1264226 & rs2276470 & 46063167 & 45974668.0 & 0.088 & 0.018 & 0.022 \\
+  7 & TRA2A & rs7776572 & rs11770192 & 23528927 & 23498358.0 & 0.031 & 0.064 & 0.064 \\
+  21 & PRMT2 & rs2839372 & rs11701058 & 48063862 & 47776382.0 & 0.287 & 0.100 & 0.122 \\
+  12 & OAS1 & rs13311 & rs2072133 & 113448652 & 113409260.0 & 0.039 & 0.002 & 0.016 \\
+  16 & N4BP1 & rs12444224 & rs11649236 & 87580855 & 48632478.0 & 38.948 & 0.007 & 0.021 \\
+  5 & CAST & rs12719343 & rs7733671 & 125369113 & 96000269.0 & 29.369 & 0.001 & 0.001 \\
+  7 & DNAJB6 & rs2286842 & rs3779589 & 157216093 & 157163614.0 & 0.052 & 0.005 & 0.006 \\
+  1 & OVGP1 & rs10802822 & rs1264898 & 240132968 & 111992823.0 & 128.140 & 0.008 & 0.030 \\
+  20 & CD93 & rs2868504 & rs1884655 & 37771578 & 23074375.0 & 14.697 & 0.000 & 0.002 \\
+  11 & PHCA & rs493642 & rs10736812 & 123097386 & 76708086.0 & 46.389 & 0.002 & 0.008 \\
+  21 & MX1 & rs459498 & rs8130120 & 42795027 & 29363604.0 & 13.431 & 0.000 & 0.000 \\
+  16 & AKTIP & rs2896940 & rs13332406 & 57721127 & 53489705.0 & 4.231 & 0.000 & 0.001 \\
+  17 & CDK5R1 & rs9905940 & rs11655031 & 46614102 & 30833162.0 & 15.781 & 0.000 & 0.000 \\
+  2 & CYBRD1 & rs888427 & rs7591849 & 172368120 & 160112881.0 & 12.255 & 0.000 & 0.000 \\
+  8 & HMBOX1 & rs587639 & rs7837237 & 132725731 & 28876221.0 & 103.850 & 0.001 & 0.001 \\
+  11 & TRAPPC4 & rs1793823 & rs3916581 & 131018917 & 118887887.0 & 12.131 & 0.001 & 0.002 \\
+  12 & PEX5 & rs10444467 & rs4329748 & 128052636 & 7364442.0 & 120.688 & 0.000 & 0.000 \\
+  12 & FLJ20489 & rs17615703 & rs3782908 & 117036766 & 48169526.0 & 68.867 & 0.001 & 0.002 \\
+  16 & PRKCB1 & rs2188355 & rs10492793 & 23867776 & 12639800.0 & 11.228 & 0.000 & 0.000 \\
+  14 & MRPL52 & rs1950857 & rs3811188 & 26710271 & 23299135.0 & 3.411 & 0.002 & 0.004 \\
+  17 & C17ORF60 & rs9907897 & rs7405659 & 63502633 & 59874129.0 & 3.629 & 0.004 & 0.011 \\
+  6 & FLJ43093 & rs6906101 & rs13214069 & 36667610 & 32705248.0 & 3.962 & 0.000 & 0.000 \\
+  19 & TRAPPC5 & rs17159840 & rs17763599 & 7758194 & 2369415.0 & 5.389 & 0.000 & 0.000 \\
+  22 & PISD & rs715572 & rs6518754 & 33234931 & 32097775.0 & 1.137 & 0.001 & 0.003 \\
+  12 & DIP2B & rs871257 & rs12427378 & 117994348 & 51074199.0 & 66.920 & 0.001 & 0.001 \\
+  12 & GPR162 & rs2272500 & rs2707210 & 79685913 & 6902002.0 & 72.784 & 0.003 & 0.005 \\
+  17 & USP36 & rs2279308 & rs7225546 & 76794981 & 75151717.0 & 1.643 & 0.000 & 0.000 \\
+   \hline
+\end{tabular}
+\end{threeparttable}
+}
+\end{table}
+
 \end{document}

### 2013-09-02 docs/manuscript/methods.tex
minor edits to the re-written manuscript and methods

@@
 \sigma _a ^2             &          j = k \\ 
-2\phi _{ijk} \sigma _a ^2& 			j \neq k \\
+2\phi _{jk} \sigma _a ^2& 			j \neq k \\
 \end{array} \right.
@@
 The parameter $\sigma_a^2$ is the variance component for additive background genetic.
-Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{ijk}$ is the kinship coefficient between individuals $j$ and $k$.
+Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{jk}$ is the kinship coefficient between individuals $j$ and $k$.
 The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero.
@@
 
+\subsubsection{Stage 1}
+
 The complete exhaustive scan for 7339 probes comprises $1.03 \times 10^{15}$ $F$-tests.
@@
 
-Following filtering on this threshold only SNP pairs with at least 5 data points in all 9 genotype classes were kept.
+\subsubsection{Filtering}
+
+We used two approaches to filter SNPs from stage 1 to be tested for significant interaction effects in stage 2.
+
+\paragraph{Filter 1} After keeping SNP pairs that surpassed the $2.91 \times 10^{-16}$ threshold in stage 1 only SNP pairs with at least 5 data points in all 9 genotype classes were kept.
 We then calculated the LD between interacting SNPs (amongst unrelated individuals within the discovery sample and also from 1000 genomes data) and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition.
 If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained.
-At this stage 6404 filtered SNP pairs remained (filter 1).
-We then performed a second filtering screen that was identical to the first but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed, creating a second set of 4751 filtered SNP pairs (filter 2).
+At this stage 6404 filtered SNP pairs remained.
+
+\paragraph{Filter 2} We also performed a second filtering screen applied to the list of SNP pairs from stage 1 that was identical to filter 1 but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed \cite{Powell2013}, creating a second set of 4751 unique filtered SNP pairs.
+
+\subsubsection{Stage 2}
 
@@
 This resulted in 406 and 95 SNP pairs with significant interaction terms from filters 1 and 2, respectively.
-
-We ruled out population stratification as a possible cause of inflated test statistics.
-To test for cryptic relatedness driving the interaction terms we tested for increased LD among the SNPs \cite{Yang2011}.
-We calculated the mean of the off-diagonal elements of the correlation matrix of all unique SNPs from the 501 interactions (731 SNPs) using only unrelated individuals, $\bar{r^2} = 0.0039$.
-This is not significantly different from the null hypothesis of zero (sampling error $= 1/n_{\textrm{unrelated}} = 0.0039$).
-
-All variance decompositions were calculated using the NOIA method \cite{Alvarez-Castro2008}.
 
@@
 This procedure was performed for a range of $NCP$ parameters that represented power ranging from $\sim 0$ to $\sim 1$.
+
+\subsection{Population stratification}
+
+We ruled out population stratification as a possible cause of inflated test statistics.
+To test for cryptic relatedness driving the interaction terms we tested for increased LD among the SNPs \cite{Yang2011}.
+We calculated the mean of the off-diagonal elements of the correlation matrix of all unique SNPs from the 501 interactions (731 SNPs) using only unrelated individuals, $\bar{r^2} = 0.0039$.
+This is not significantly different from the null hypothesis of zero (sampling error $= 1/n_{\textrm{unrelated}} = 0.0039$).
+
+\subsection{Probe mapping}
+
+To avoid possibility that epistatic signals might arise due to expression probes hybridising in multiple locations we verified that probe sequences for genes with significant interactions mapped to only a single location.
+As an initial verification we performed a BLAST search of the full probe sequence against 1000 genomes phase 1 version 3 human genome reference and ensured that only one genomic location aligned significantly ($p < 0.05$).
+As a second step, to mitigate the possibility of weak hybridisation elsewhere in the genome we divided the probe sequence into three sections (1-25bp, 13-37bp, 26-50bp) and performed a BLAST search of these probe sequence fragments.
+No probe sequemces or probe sequence fragments mapped to positions other than the single expected genomic target ($p < 0.05$).
 
@@
 
-The probability of observing the number of concordant signs is calculated using a binomial test.
+The probability of observing the number of concordant signs is calculated using a binomial test. All variance decompositions were calculated using the NOIA method \cite{Alvarez-Castro2008}.
 

### 2013-09-02 docs/manuscript/nature_letter.tex
minor edits to the re-written manuscript and methods

@@
 In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
-Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31})$ and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 434$.
+Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 434$.
 There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
@@
 We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/434$, Table \ref{tab:bonferroni}).
-These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).
-We decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect in the independent datasets (Methods).
-The most significant epistatic effect in the discovery sample was in the same direction in both replication datasets in 22 out of the 30 significantly replicated interactions ($p = 3.76 \times 10^{-8}$, Supplementary Table \ref{tab:sign}).
+To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods).
+Sign concordance between the discovery and both replication datasets was observed in 22 out of the 30 significantly replicated interactions (expected value $= 7.5$ under the null hypothesis of no interactions, $p = 3.76 \times 10^{-8}$).
 

### 2013-09-02 docs/manuscript/Response_to_review_comments.tex
added tex file for response to reviewers

\documentclass{article}
\usepackage{url}
\usepackage[superscript,biblabel]{cite}

\bibliographystyle{naturemag}

\title{Response to review comments}
\date{}
\author{}

\begin{document}
\maketitle

\section{Reviewer 1}

Hemani et al. have performed a complete genome-scan for pair-wise epistatic eQTL effects using expression data for ∼7400 transcripts assayed with microarrays and RNA extracted from PBMCs of ∼850 individuals genotyped for ∼500,000 SNPs. After filtering on LD (within and between SNP pairs) the authors report 501 SNPs pairs that are significantly (after Bonferroni correction) influencing the expression level of a target gene. The authors show that - for 30 of 434/501 “replicatable” SNP pairs - the epistatic component of the effect remains significant (after Bonferroni correction) in a confirmation cohort comprising expression data obtained with the same array on PBMCs of $>$ 2,000 individuals. Moreover, the ranked p-values of the top 316/404 exceed the 97.5\% confidence surface occupied by QQ-plots under the null hypothesis, supporting the veracity of the majority of the corresponding epistatic effects. The majority of interacting SNP pairs include a cis- and a trans-acting SNPs. Several instances are described where multiple trans-SNPs modify the effect of a cis-SNP. The authors show that SNPs in high LD with the interacting cis-SNPs (but not trans-SNPs), tend to be located in transcriptionally active chromatin regions, particularly in haematopoeitic cell lines. Moreover, the authors show that interacting SNPs tend to be located in the vicinity of chromosome segments shown to physically interact in the nucleus of K562 blood cells. The authors make an attempt to quantify the relative importance of additive vs epistatic effects in determining gene expression level. In conclusion, this work provides strong evidence that pair-wise epistasis contributes significantly to variation in transcript levels, encouraging similar efforts applied to other complex traits including diseases.

The topic of epistasis is of great interest to the field of complex trait genetics. The paper is in general clearly and succinctly written. The methodology is sound. The results are convincing.

I only have minor comments: \\
1. Abstract: “2.5\% confidence interval of the distribution under the null hypothesis” ... I don’t think that this is what the authors mean. Should be rephrased here and throughput the text.

2. Abstract: “... within 2MB of regions ... “ Some contradictions here and throughout the text. Is it 2 or 2.5? According to Suppl. Fig. 11, its should be 2.5Mb

3. Main text, page 3: “... 316 of the remaining 404 discovered SNPs ...” come confusiuon here and in other places. These are 404 SNP pairs not SNPs.

4. I am not sure that trimming SNPs based on LD within pairs completely avoided haplotype effects. Yet, as the majority of interactions are cistrans this is not a measure issue. It would be nice to provide an idea of the average distance (+ range) for cis-cis interactions.

5. Suppl. Methods: Equ. 2: Is it correct that the diagonal includes both $\sigma^2$ $A + \sigma^2 E$?

6. Suppl. Methods: Page 4, line 1: How do the authors explain that the epistatic component was significant for a much larger proportion of SNP pairs for which one or both SNP had a highly significant marginal effect?

7. Suppl. Fig. 1: I would have liked to see the right panel (null) generated not with random SNPs but with the reshuffled 434 or 404 SNP pairs. The legend refers to 2.5\% FDR ... Is this truly FDR?

Finally, despite the inclusion and use of Hi-C data, the paper really does not elaborate about possible molecular mechanisms that might underlie the observed epistatic effects. Were the trans-SNPs more often causing cis-eQTL
effects than expected by chance alone? Were they (and SNP in LD) enriched in coding SNP? Were they located in the vicinity of specific types of genes (transcription factors)? Etc.

\section{Reviewer 2}

Authors carry out epistasis analyses on human eQTL GWAS data. Across 7339 gene expression levels in blood in a cohort of 846 individuals, they detect 501 pairwise SNP interactions, some of which replicate in at least one of two replication data sets. Authors perform some bioinformatic enrichment analyses of interaction SNPs.

Abstract and introduction background: epistasis has been reported in many mapping studies of natural trait variation in multiple species, including for gene expression levels.

What is the evidence that transcription levels are less polygenic than higher level phenotypes?

Page 3, \"remarkable similarity in GP maps\" needs to be quantified.

Page 4, \"cis-cis\" interactions are defined as \"both SNPs on same chromosome as expression gene\". These can be very far away and unlikely to be cis, especially if filter of any SNPs in LD is applied here. Interaction results between SNPs on same chromosome are frequently artefactual due to small sample size of \"recombinant\" haplotype classes because of LD.

Page 4, genes and SNPs involved in very many interactions are not expected given the sparseness of interactions detected, and are likely to reflect technical artefacts.

Bottom of page 5 and top of page 6, enrichement analyses are weakly informative at best. From weak enrichment of cis-acting SNPs vs trans-acting SNPs for transcriptionally active regions in haematopoietic cells it seems unreasonable to draw conclusions about their biological relevance.

Page 5, there is no justification for applying interaction threshold to additive effects. Should match false discovery rates or effect sizes but not thresholds for classes with very different statistical properties in regards to multiple testing and power. Leads to huge underestimate of additive effects.

Methods to identify epistatic QTL are confusing. Test of full vs null model should capture significant additive, epistatic, and/or dominance effects and post hoc methods could be used to disentangle which terms are contributing, but significance after post hoc filters hard to evaluate. Significance threshold for the full vs null model is cited in main text on page 3 before stating that 501 interactions were discovered. This is not exactly appropriate, as this was threshold for full vs null, not the criteria used to determine if there was significant epistasis. It is not clear how many of the 501 interactions are actually significant, nor what the false discovery rate is for this set. It is unclear what \"filters 1 and 2\" are on (methods page 4).

Should at least use additive-by-additive epistasis model alongside the full model, to increase statistical power and generate better context via comparison to previous work, where this is what is standardly done.

It is not clear how filtering out SNPs with significant additive or dominant effects (methods page 3) is consistent with results in the first full paragraph of page 4 (main text), which notes many interaction SNP pairs with significant main effects.

An appropriate FDR threshold for the tests of the (full) model vs the (additive and dominance) model would be more informative than the Bonferroni threshold used for the post hoc determination of epistatic pairs.

How is the \"null distribution of no epistatic effects\" (bottom of page 3) determined?

(top of page 4) Is the dependence on LD between observed SNPs and causal variants the most noteworthy explanation for the lack of replication between discovery and replication samples.

What is the relevance of the statement that \"patterns of epistasis used for statistical decomposition are not designed to resemble biological function\" in the context of that paragraph (end of first full paragraph page 4).

\end{document}

### 2013-09-02 docs/manuscript/methods.tex
updated manuscript and added extra table on sign enrichment

@@
 
-\begin{enumerate}
-
-\item [Test 1] Is the most significant epistatic effect in the discovery set in the same direction as the same epistatic effect in the replication sets?
+\paragraph{Test 1} Is the most significant epistatic effect in the discovery set in the same direction as the same epistatic effect in the replication sets?
 We decomposed the genetic variance into 8 orthogonal effects, four of which are epistatic ($A \times A$, $A \times D$, $D \times A$, $D \times D$).
@@
 
-\item [Test 2] Is the most significant epistatic effect in the discovery the same as the largest epistatic effect in the replication set with the sign being concordant.
+\paragraph{Test 2} Is the most significant epistatic effect in the discovery the same as the largest epistatic effect in the replication set with the sign being concordant.
 As in Test 1, but this time we required that the largest effect was the same in the discovery and the replication, and that they had the same sign (\emph{e.g.} if the largest effect in the discovery is $A \times A$, with a positive effect, then concordance is achieved if the same is true in the replication).
@@
 
-\item [Test 3] Are all 4 epistatic effects in the same direction in the discovery and the replication datasets?
-This is a very conservative test as it is not realistic to assume that all four epistatic effects are significant in every interaction.
-The probability of concordance in one replication dataset is $1/16$, and the probability of concordance in both replication datasets is $1/256$.
-
-\item [Test 4] Do the epistatic effects that are significant at nominal $p < 0.05$ in the discovery have the same direction of effect as in the replication?
+\paragraph{Test 3} Do the epistatic effects that are significant at nominal $p < 0.05$ in the discovery have the same direction of effect as in the replication?
 Here we count all the epistatic variance components in the discovery that have $p < 0.05$ (1133 amongst the 434 discovery SNP pairs, $i.e.$ each SNP pair has at least 1 and at most 4 significant epistatic variance components).
@@
 
-\end{enumerate}
+\paragraph{Test 4} If we count how many of the 4 epistatic effects are concordant between the discovery and replication data for each interaction then is this significant from what we expect by chance?
+There can be either 0, 1, 2, 3 or 4 concordant signs at each interaction, each with expectation of $p=1/16, 4/16, 6/16, 4/16, 1/16$ under the null, respectively.
+Observed counts are multinomially distributed, and we tested if the observed proportions were statistically different from the expected proportions using an approximation of the multinomial test \cite{Williams1976}.
 
-The probability of observing the number of concordant signs is calculated using a binomial test. All variance decompositions were calculated using the NOIA method \cite{Alvarez-Castro2008}.
+The probability of observing the number of concordant signs in tests 1-3 is calculated using a binomial test.
+All variance decompositions were calculated using the NOIA method \cite{Alvarez-Castro2008}.
 

### 2013-09-02 docs/manuscript/nature_letter.tex
updated manuscript and added extra table on sign enrichment

@@
 
-Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
+Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene, median distance between interacting SNPs is 1.83Mb), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
 We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
@@
   \hline
-Test & Interactions & Dataset & $n$ & Expected & Observed & $p$-value \\
-  \hline
-1\tnote{a} & All & EGCUT & 434 & 217.00 & 306 & $6.69 \times 10^{18}$ \\
-&  & Fehrmann & 434 & 217.00 & 278 & $5.04 \times 10^{09}$ \\
-&  & Both & 434 & 108.50 & 221 & $5.56 \times 10^{31}$ \\
-& Significant & EGCUT & 30 & 15.00 & 25 & $3.25 \times 10^{04}$ \\
-&  & Fehrmann & 30 & 15.00 & 24 & $1.43 \times 10^{03}$ \\
-&  & Both & 30 & 7.50 & 22 & $3.76 \times 10^{08}$ \\
-  \hline
-2\tnote{b} & All & EGCUT & 434 & 54.25 & 92 & $4.22 \times 10^{07}$ \\
-&  & Fehrmann & 434 & 54.25 & 79 & $6.18 \times 10^{04}$ \\
-&  & Both & 434 & 6.78 & 30 & $2.55 \times 10^{11}$ \\
-& Significant & EGCUT & 30 & 3.75 & 19 & $9.46 \times 10^{11}$ \\
-&  & Fehrmann & 30 & 3.75 & 19 & $9.46 \times 10^{11}$ \\
-&  & Both & 30 & 0.47 & 18 & $2.23 \times 10^{25}$ \\
-  \hline
-3\tnote{c} & All & EGCUT & 434 & 27.12 & 34 & $1.65 \times 10^{01}$ \\
-&  & Fehrmann & 434 & 27.12 & 35 & $1.35 \times 10^{01}$ \\
-&  & Both & 434 & 1.70 &  2 & $6.89 \times 10^{01}$ \\
-& Significant & EGCUT & 30 & 1.88 &  8 & $3.92 \times 10^{04}$ \\
-&  & Fehrmann & 30 & 1.88 &  9 & $6.22 \times 10^{05}$ \\
-&  & Both & 30 & 0.12 &  1 & $1.11 \times 10^{01}$ \\
-  \hline
-4\tnote{d} & All & EGCUT & 1133 & 566.50 & 775 & $7.10 \times 10^{36}$ \\
-&  & Fehrmann & 1133 & 566.50 & 726 & $1.90 \times 10^{21}$ \\
-&  & Both & 1133 & 283.25 & 562 & $1.39 \times 10^{70}$ \\
-& Significant & EGCUT & 73 & 36.50 & 55 & $1.69 \times 10^{05}$ \\
-&  & Fehrmann & 73 & 36.50 & 55 & $1.69 \times 10^{05}$ \\
-&  & Both & 73 & 18.25 & 46 & $7.86 \times 10^{12}$ \\
+Test & Interactions\tnote{a} & Dataset & $n$\tnote{b} & Expected\tnote{c} & Observed\tnote{d} & $p$-value \\
+  \hline
+1\tnote{e} & All & EGCUT & 434 & 217.00 & 306 & $6.69 \times 10^{-18}$ \\
+&  & Fehrmann & 434 & 217.00 & 278 & $5.04 \times 10^{-9}$ \\
+&  & Both & 434 & 108.50 & 221 & $5.56 \times 10^{-31}$ \\
+& Significant & EGCUT & 30 & 15.00 & 25 & $3.25 \times 10^{-4}$ \\
+&  & Fehrmann & 30 & 15.00 & 24 & $1.43 \times 10^{-3}$ \\
+&  & Both & 30 & 7.50 & 22 & $3.76 \times 10^{-8}$ \\
+  \hline
+2\tnote{f} & All & EGCUT & 434 & 54.25 & 92 & $4.22 \times 10^{-7}$ \\
+&  & Fehrmann & 434 & 54.25 & 79 & $6.18 \times 10^{-4}$ \\
+&  & Both & 434 & 6.78 & 30 & $2.55 \times 10^{-11}$ \\
+& Significant & EGCUT & 30 & 3.75 & 19 & $9.46 \times 10^{-11}$ \\
+&  & Fehrmann & 30 & 3.75 & 19 & $9.46 \times 10^{-11}$ \\
+&  & Both & 30 & 0.47 & 18 & $2.23 \times 10^{-25}$ \\
+  \hline
+3\tnote{g} & All & EGCUT & 1133 & 566.50 & 775 & $7.10 \times 10^{-36}$ \\
+&  & Fehrmann & 1133 & 566.50 & 726 & $1.90 \times 10^{-21}$ \\
+&  & Both & 1133 & 283.25 & 562 & $1.39 \times 10^{-70}$ \\
+& Significant & EGCUT & 73 & 36.50 & 55 & $1.69 \times 10^{-5}$ \\
+&  & Fehrmann & 73 & 36.50 & 55 & $1.69 \times 10^{-5}$ \\
+&  & Both & 73 & 18.25 & 46 & $7.86 \times 10^{-12}$ \\
    \hline
@@
 \begin{tablenotes}
-\item[a] The sign of the most significant epistatic variance component in discovery is the same as the corresponding variance component in the replication data.
-\item[b] The largest epistatic variance component in the discovery is the same as in the replication with the same sign in both.
-\item[c] The sign of all four epistatic variance components are identical in the discovery and the replication.
-\item[d] The sign of all epistatic variance components in the discovery with $p < 0.05$ are the same as the corresponding variance components in the replication data.
+\item[a] ``All'' denotes 434 discovery interactions and ``Significant'' denotes 30 interactions with significant replication $p$-values
+\item[b] Number of tests for concordance
+\item[c] Expected number of concordant cases under the null hypothesis of no interactions
+\item[d] Observed number of concordant cases
+\item[e] The sign of the most significant epistatic variance component in discovery is the same as the corresponding variance component in the replication data.
+\item[f] The largest epistatic variance component in the discovery is the same as in the replication with the same sign in both.
+\item[g] The sign of all epistatic variance components in the discovery with $p < 0.05$ are the same as the corresponding variance components in the replication data.
+\end{tablenotes}
+
+\end{threeparttable}
+\end{table}
+
+\begin{table}[ht]
+\begin{threeparttable}
+\caption{\textbf{Concordance of sign of epistatic variance components between discovery and replication datasets using test 4}}
+\label{tab:sign2}
+\begin{tabular}{llrrrrrrr}
+\hline
+Interactions\tnote{a} & Dataset & $n$\tnote{b} & 0\tnote{c} & 1\tnote{c} & 2\tnote{c} & 3\tnote{c} & 4\tnote{c} & $p$ \\
+\hline
+Expected\tnote{d} & - & - & 0.06 & 0.25 & 0.38 & 0.25 & 0.06 & - \\
+All & EGCUT & 434 & 0.06 & 0.22 & 0.41 & 0.23 & 0.08 & $0.194$ \\
+All & Fehrmann & 434 & 0.07 & 0.22 & 0.39 & 0.24 & 0.08 & $0.385$ \\
+All & Combined & 868 & 0.07 & 0.22 & 0.40 & 0.23 & 0.08 & $0.0448$ \\
+Significant & EGCUT & 30 & 0.07 & 0.03 & 0.30 & 0.33 & 0.27 & $4.72 \times 10^{-4}$ \\
+Significant & Fehrmann & 30 & 0.03 & 0.07 & 0.33 & 0.27 & 0.30 & $6.69 \times 10^{-4}$ \\
+Significant & Combined & 60 & 0.05 & 0.05 & 0.32 & 0.30 & 0.28 & $5.49 \times 10^{-8}$ \\
+\hline\end{tabular}
+
+\begin{tablenotes}
+\item[a] ``All'' denotes 434 discovery interactions and ``Significant'' denotes 30 interactions with significant replication $p$-values.
+\item[b] Number of tests for concordance.
+\item[c] Proportion of tests that have 0, 1, 2, 3 or 4 concordant signs between discovery and replication.
+\item[d] Expected proportion of concordant signs under the null hypothesis of no epistasis.
 \end{tablenotes}
@@
 \begin{tabular}{llllrrrrr}
-
 \hline
@@
 \hline
-19 & TMEM149 & rs807491 & rs7254601 & 36268923 & 36147315.0 & 0.122 & 0.000 & 0.001 \\
-  17 & FN3KRP & rs898095 & rs9892064 & 80890638 & 80827903.0 & 0.063 & 0.063 & 0.088 \\
-  21 & CSTB & rs9979356 & rs3761385 & 45230974 & 45198355.0 & 0.033 & 0.041 & 0.066 \\
-  3 & MBNL1 & rs16864367 & rs13079208 & 152234166 & 152116652.0 & 0.118 & 0.041 & 0.117 \\
-  10 & ADK & rs2395095 & rs10824092 & 76446305 & 75929517.0 & 0.517 & 0.013 & 0.020 \\
-  11 & CTSC & rs7930237 & rs556895 & 88117962 & 88077479.0 & 0.040 & 0.012 & 0.045 \\
-  17 & GAA & rs11150847 & rs12602462 & 78153130 & 78146016.0 & 0.007 & 0.000 & 0.001 \\
-  8 & NAPRT1 & rs2123758 & rs3889129 & 144663661 & 144613680.0 & 0.050 & 0.053 & 0.060 \\
-  1 & LAX1 & rs1891432 & rs10900520 & 203877662 & 203780591.0 & 0.097 & 0.065 & 0.106 \\
-  18 & MBP & rs8092433 & rs4890876 & 74747424 & 74732087.0 & 0.015 & 0.035 & 0.053 \\
-  11 & SNORD14A & rs2634462 & rs6486334 & 17339127 & 17015557.0 & 0.324 & 0.008 & 0.012 \\
-  21 & C21ORF57 & rs9978658 & rs11701361 & 48027084 & 47764477.0 & 0.263 & 0.032 & 0.065 \\
-  16 & RPL13 & rs352935 & rs2965817 & 89648580 & 89513234.0 & 0.135 & 0.054 & 0.060 \\
-  19 & ATP13A1 & rs4284750 & rs873870 & 19810050 & 19738554.0 & 0.071 & 0.008 & 0.015 \\
-  2 & NCL & rs7563453 & rs4973397 & 232301670 & 232291471.0 & 0.010 & 0.027 & 0.029 \\
-  5 & HNRPH1 & rs6894268 & rs4700810 & 179032488 & 178991794.0 & 0.041 & 0.000 & 0.001 \\
-  19 & VASP & rs1264226 & rs2276470 & 46063167 & 45974668.0 & 0.088 & 0.018 & 0.022 \\
-  7 & TRA2A & rs7776572 & rs11770192 & 23528927 & 23498358.0 & 0.031 & 0.064 & 0.064 \\
-  21 & PRMT2 & rs2839372 & rs11701058 & 48063862 & 47776382.0 & 0.287 & 0.100 & 0.122 \\
-  12 & OAS1 & rs13311 & rs2072133 & 113448652 & 113409260.0 & 0.039 & 0.002 & 0.016 \\
-  16 & N4BP1 & rs12444224 & rs11649236 & 87580855 & 48632478.0 & 38.948 & 0.007 & 0.021 \\
-  5 & CAST & rs12719343 & rs7733671 & 125369113 & 96000269.0 & 29.369 & 0.001 & 0.001 \\
-  7 & DNAJB6 & rs2286842 & rs3779589 & 157216093 & 157163614.0 & 0.052 & 0.005 & 0.006 \\
-  1 & OVGP1 & rs10802822 & rs1264898 & 240132968 & 111992823.0 & 128.140 & 0.008 & 0.030 \\
-  20 & CD93 & rs2868504 & rs1884655 & 37771578 & 23074375.0 & 14.697 & 0.000 & 0.002 \\
-  11 & PHCA & rs493642 & rs10736812 & 123097386 & 76708086.0 & 46.389 & 0.002 & 0.008 \\
-  21 & MX1 & rs459498 & rs8130120 & 42795027 & 29363604.0 & 13.431 & 0.000 & 0.000 \\
-  16 & AKTIP & rs2896940 & rs13332406 & 57721127 & 53489705.0 & 4.231 & 0.000 & 0.001 \\
-  17 & CDK5R1 & rs9905940 & rs11655031 & 46614102 & 30833162.0 & 15.781 & 0.000 & 0.000 \\
-  2 & CYBRD1 & rs888427 & rs7591849 & 172368120 & 160112881.0 & 12.255 & 0.000 & 0.000 \\
-  8 & HMBOX1 & rs587639 & rs7837237 & 132725731 & 28876221.0 & 103.850 & 0.001 & 0.001 \\
-  11 & TRAPPC4 & rs1793823 & rs3916581 & 131018917 & 118887887.0 & 12.131 & 0.001 & 0.002 \\
-  12 & PEX5 & rs10444467 & rs4329748 & 128052636 & 7364442.0 & 120.688 & 0.000 & 0.000 \\
-  12 & FLJ20489 & rs17615703 & rs3782908 & 117036766 & 48169526.0 & 68.867 & 0.001 & 0.002 \\
-  16 & PRKCB1 & rs2188355 & rs10492793 & 23867776 & 12639800.0 & 11.228 & 0.000 & 0.000 \\
-  14 & MRPL52 & rs1950857 & rs3811188 & 26710271 & 23299135.0 & 3.411 & 0.002 & 0.004 \\
-  17 & C17ORF60 & rs9907897 & rs7405659 & 63502633 & 59874129.0 & 3.629 & 0.004 & 0.011 \\
-  6 & FLJ43093 & rs6906101 & rs13214069 & 36667610 & 32705248.0 & 3.962 & 0.000 & 0.000 \\
-  19 & TRAPPC5 & rs17159840 & rs17763599 & 7758194 & 2369415.0 & 5.389 & 0.000 & 0.000 \\
-  22 & PISD & rs715572 & rs6518754 & 33234931 & 32097775.0 & 1.137 & 0.001 & 0.003 \\
-  12 & DIP2B & rs871257 & rs12427378 & 117994348 & 51074199.0 & 66.920 & 0.001 & 0.001 \\
-  12 & GPR162 & rs2272500 & rs2707210 & 79685913 & 6902002.0 & 72.784 & 0.003 & 0.005 \\
-  17 & USP36 & rs2279308 & rs7225546 & 76794981 & 75151717.0 & 1.643 & 0.000 & 0.000 \\
+19 & TMEM149 & rs807491 & rs7254601 & 36268923 & 36147315 & 0.122 & 0.000 & 0.001 \\
+17 & FN3KRP & rs898095 & rs9892064 & 80890638 & 80827903 & 0.063 & 0.063 & 0.088 \\
+21 & CSTB & rs9979356 & rs3761385 & 45230974 & 45198355 & 0.033 & 0.041 & 0.066 \\
+3 & MBNL1 & rs16864367 & rs13079208 & 152234166 & 152116652 & 0.118 & 0.041 & 0.117 \\
+10 & ADK & rs2395095 & rs10824092 & 76446305 & 75929517 & 0.517 & 0.013 & 0.020 \\
+11 & CTSC & rs7930237 & rs556895 & 88117962 & 88077479 & 0.040 & 0.012 & 0.045 \\
+17 & GAA & rs11150847 & rs12602462 & 78153130 & 78146016 & 0.007 & 0.000 & 0.001 \\
+8 & NAPRT1 & rs2123758 & rs3889129 & 144663661 & 144613680 & 0.050 & 0.053 & 0.060 \\
+1 & LAX1 & rs1891432 & rs10900520 & 203877662 & 203780591 & 0.097 & 0.065 & 0.106 \\
+18 & MBP & rs8092433 & rs4890876 & 74747424 & 74732087 & 0.015 & 0.035 & 0.053 \\
+11 & SNORD14A & rs2634462 & rs6486334 & 17339127 & 17015557 & 0.324 & 0.008 & 0.012 \\
+21 & C21ORF57 & rs9978658 & rs11701361 & 48027084 & 47764477 & 0.263 & 0.032 & 0.065 \\
+16 & RPL13 & rs352935 & rs2965817 & 89648580 & 89513234 & 0.135 & 0.054 & 0.060 \\
+19 & ATP13A1 & rs4284750 & rs873870 & 19810050 & 19738554 & 0.071 & 0.008 & 0.015 \\
+2 & NCL & rs7563453 & rs4973397 & 232301670 & 232291471 & 0.010 & 0.027 & 0.029 \\
+5 & HNRPH1 & rs6894268 & rs4700810 & 179032488 & 178991794 & 0.041 & 0.000 & 0.001 \\
+19 & VASP & rs1264226 & rs2276470 & 46063167 & 45974668 & 0.088 & 0.018 & 0.022 \\
+7 & TRA2A & rs7776572 & rs11770192 & 23528927 & 23498358 & 0.031 & 0.064 & 0.064 \\
+21 & PRMT2 & rs2839372 & rs11701058 & 48063862 & 47776382 & 0.287 & 0.100 & 0.122 \\
+12 & OAS1 & rs13311 & rs2072133 & 113448652 & 113409260 & 0.039 & 0.002 & 0.016 \\
+16 & N4BP1 & rs12444224 & rs11649236 & 87580855 & 48632478 & 38.948 & 0.007 & 0.021 \\
+5 & CAST & rs12719343 & rs7733671 & 125369113 & 96000269 & 29.369 & 0.001 & 0.001 \\
+7 & DNAJB6 & rs2286842 & rs3779589 & 157216093 & 157163614 & 0.052 & 0.005 & 0.006 \\
+1 & OVGP1 & rs10802822 & rs1264898 & 240132968 & 111992823 & 128.140 & 0.008 & 0.030 \\
+20 & CD93 & rs2868504 & rs1884655 & 37771578 & 23074375 & 14.697 & 0.000 & 0.002 \\
+11 & PHCA & rs493642 & rs10736812 & 123097386 & 76708086 & 46.389 & 0.002 & 0.008 \\
+21 & MX1 & rs459498 & rs8130120 & 42795027 & 29363604 & 13.431 & 0.000 & 0.000 \\
+16 & AKTIP & rs2896940 & rs13332406 & 57721127 & 53489705 & 4.231 & 0.000 & 0.001 \\
+17 & CDK5R1 & rs9905940 & rs11655031 & 46614102 & 30833162 & 15.781 & 0.000 & 0.000 \\
+2 & CYBRD1 & rs888427 & rs7591849 & 172368120 & 160112881 & 12.255 & 0.000 & 0.000 \\
+8 & HMBOX1 & rs587639 & rs7837237 & 132725731 & 28876221 & 103.850 & 0.001 & 0.001 \\
+11 & TRAPPC4 & rs1793823 & rs3916581 & 131018917 & 118887887 & 12.131 & 0.001 & 0.002 \\
+12 & PEX5 & rs10444467 & rs4329748 & 128052636 & 7364442 & 120.688 & 0.000 & 0.000 \\
+12 & FLJ20489 & rs17615703 & rs3782908 & 117036766 & 48169526 & 68.867 & 0.001 & 0.002 \\
+16 & PRKCB1 & rs2188355 & rs10492793 & 23867776 & 12639800 & 11.228 & 0.000 & 0.000 \\
+14 & MRPL52 & rs1950857 & rs3811188 & 26710271 & 23299135 & 3.411 & 0.002 & 0.004 \\
+17 & C17ORF60 & rs9907897 & rs7405659 & 63502633 & 59874129 & 3.629 & 0.004 & 0.011 \\
+6 & FLJ43093 & rs6906101 & rs13214069 & 36667610 & 32705248 & 3.962 & 0.000 & 0.000 \\
+19 & TRAPPC5 & rs17159840 & rs17763599 & 7758194 & 2369415 & 5.389 & 0.000 & 0.000 \\
+22 & PISD & rs715572 & rs6518754 & 33234931 & 32097775 & 1.137 & 0.001 & 0.003 \\
+12 & DIP2B & rs871257 & rs12427378 & 117994348 & 51074199 & 66.920 & 0.001 & 0.001 \\
+12 & GPR162 & rs2272500 & rs2707210 & 79685913 & 6902002 & 72.784 & 0.003 & 0.005 \\
+17 & USP36 & rs2279308 & rs7225546 & 76794981 & 75151717 & 1.643 & 0.000 & 0.000 \\
 \hline

### 2013-09-02 analysis/sign_enrichment.R

@@ -80,41 +80,6 @@ test3 <- function(noiad)
 	{
 		x <- mutate(x)
 		a <- subset(x, dataset == 1)
-		s <- sign(x$E[5:8])
-
-		r <- list()
-		for(i in 2:3)
-		{
-			y <- subset(x, dataset == i)
-			r[[i-1]] <- sum(s == sign(y$E[5:8])) == 4
-		}
-
-		out <- data.frame(r)
-		names(out) <- c("r1", "r2")
-		return(out)
-	})
-	p1 <- binom.test(n = nrow(els), x = sum(els$r1), p = 1/16)$p.value
-	p2 <- binom.test(n = nrow(els), x = sum(els$r2), p = 1/16)$p.value
-	p12 <- binom.test(n = nrow(els), x = sum(els$r1 & els$r2), p = 1/256)$p.value
-
-	l <- list()
-	l$els <- els
-	l$p1 <- p1
-	l$p2 <- p2
-	l$p12 <- p12
-	l$e1 <- 1/16 * nrow(els)
-	l$e2 <- 1/16 * nrow(els)
-	l$e12 <- 1/256 * nrow(els)
-	return(l)
-
-}
-
-test4 <- function(noiad)
-{
-	els <- ddply(noiad, .(interaction), function(x)
-	{
-		x <- mutate(x)
-		a <- subset(x, dataset == 1)
 		sig <- which(x$p[5:8] < 0.1)
 		s <- sign(x$E[sig])
 
@@ -144,6 +109,39 @@ test4 <- function(noiad)
 	return(l)
 }
 
+testMult <- function(noiad)
+{
+	els <- ddply(noiad, .(interaction), function(x)
+	{
+		x <- mutate(x)
+		a <- subset(x, dataset == 1)
+		s <- sign(x$E[5:8])
+
+		r <- list()
+		for(i in 2:3)
+		{
+			y <- subset(x, dataset == i)
+			r[[i-1]] <- sum(s == sign(y$E[5:8]))
+		}
+
+		out <- data.frame(r)
+		names(out) <- c("r1", "r2")
+		return(out)
+	})
+
+
+	l <- list()
+	l$els <- els
+	l$p1 <- NA
+	l$p2 <- NA
+	l$p12 <- NA
+	l$e1 <- 1/16 * nrow(els)
+	l$e2 <- 1/16 * nrow(els)
+	l$e12 <- 1/256 * nrow(els)
+	return(l)
+
+}
+
[…]

### 2013-09-06 docs/manuscript/manuscript.tex
rebuttal documents sent to peter
work continues on rebuttal

--- 2013-09-02 docs/manuscript/nature_letter.tex
@@
 A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}).
-It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}.
+It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects (Supplementary Figure \ref{fig:ld_mean}) \cite{Weir2008, Hemani2013}, and for higher power terms of LD $r$ the sampling variance increases (Methods, Supplementary Figure \ref{fig:ld_sd}).
 Therefore these results are encouraging with regards to the detection and replication of epistasis.
@@
 	\centering
-	% \includegraphics[width=2.5in]{gpbonfrep.pdf}
+	\includegraphics[width=2.5in]{gpbonfrep.pdf}
 	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown.
@@
 	\centering
-	% \includegraphics[width=2.5in]{qqMeta.pdf}
+	\includegraphics[width=2.5in]{qqMeta.pdf}
 	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} The top panel shows all 434 discovery SNPs that were tested for interactions.
@@
 	\centering
-	% \includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
+	\includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
 	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).
@@
 	\label{fig:qqgg}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\centering
+	\includegraphics[width=5in]{ld_sampling_mean}
+	\caption{\textbf{Sampling mean for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$.
+For a given a population value of LD $r$ (columns of plots), plotted is the sample mean ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines).
+As true $r$ reduces the statistical power to detect epistatic variants drops dramatically under the assumption that statistical power is proportional to higher moments of $r$.}
+	\label{fig:ld_mean}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\centering
+	\includegraphics[width=5in]{ld_sampling_sd}
+	\caption{\textbf{Sampling standard deviation for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$.
+For a given a population value of LD $r$ (columns of plots), plotted is the sampling standard deviation ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines).
+As the power term of $r$ increases the sampling variance also increases.
+Supposing that there is sufficiently high $r^x$ in the discovery sample for detection of epistasis, the replication sample is less likely to have similarly high $r^x$ as $x$ increases, leading to an expectation of reduced replication rates.}
+	\label{fig:ld_sd}
 \end{figure}

### 2013-09-06 analysis/ld_sampling_var.R

@@ -29,6 +29,7 @@ sampleSnp <- function(r, n, p1, q1)
 	x22 <- r * sqrt(p1*q1*p2*q2) + p2*q2
 
 	xs <- c(x11, x12, x21, x22)
+
 	if(any(xs < 0))
 	{
 		return(NA)
@@ -62,10 +63,10 @@ doSim <- function(param)
 # Create parameters
 param <- expand.grid(
 	r = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.99), 
-	n = c(900, 1200, 20000), 
-	p1 = c(0.25, 0.5, 0.75), 
-	q1 = c(0.25, 0.5, 0.75), 
-	sim = 1:100, r_obs = NA
+	n = c(50, 900, 1200, 20000), 
+	p1 = c(0.1, 0.2, 0.3, 0.4, 0.5), 
+	q1 = c(0.1, 0.2, 0.3, 0.4, 0.5), 
+	sim = 1:50, r_obs = NA
 )
 param <- subset(param, p1 >= q1)
 
@@ -77,12 +78,12 @@ param <- doSim(param)
 
 
 # Summarise results
-p <- subset(param, !is.na(r_obs), select=c(r, r_obs, n))
+p <- subset(param, !is.na(r_obs), select=c(r, r_obs, n, p1, q1))
 ps <- rbind(p, p, p, p, p)
 ps$pow <- rep(c(1,2,4,6,8), each=nrow(p))
 ps$r_obs <- ps$r_obs^ps$pow
 
-ps_summary <- ddply(ps, .(pow, r, n), summarise, 
+ps_summary <- ddply(subset(ps), .(pow, r, n), summarise, 
 	N    = length(r_obs),
 	sd   = sd(r_obs),
 	se   = sd(r_obs) / sqrt(length(r_obs)),
@@ -91,21 +92,39 @@ ps_summary <- ddply(ps, .(pow, r, n), summarise,
 	qu3  = quantile(r_obs, prob=0.75, na.rm=T)	
 )
 
+subset(ps_summary, r==0.5 & pow==2)
 
 #=================================================================#
 
 # Plot
-ggplot(ps_summary, aes(x = pow, y = sd)) +
+ggplot(subset(ps_summary, n > 100), aes(x = pow, y = sd)) +
 geom_point() +
 geom_line(aes(colour=factor(n))) +
 facet_grid(. ~ r) + 
-labs(x = "Power term", y = "Standard deviation", colour = "Sample size") 
[…]

### 2013-09-09 epistasis_pages/index.Rmd
Supporting web page to be hosted on github pages

---
title: Epistasis Detection and Replication
subtitle: 
author: Gibran Hemani, Konstantin Shakhbazov
github: {user: , repo: , branch: }
framework: minimal
mode: selfcontained
widgets: [polycharts]
highlighter: highlight.js
hitheme: solarized_light
assets:
  css: 
    - "http://fonts.googleapis.com/css?family=Open+Sans"
    - "http://fonts.googleapis.com/css?family=Open+Sans+Condensed:700"
---

# Epistasis Detection adn Replication

This is web page of a paper [Detection and replication of epistasis influencing transcription in humans](http://www.nature.com)

# Abstract

Epistasis is the phenomenon whereby one polymorphism’s effect on a trait depends on other polymorphisms present in the genome. The ex- tent to which epistasis influences complex traits1 and contributes to their variation2,3 is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms,4,5 and some examples have been reported in other species,6 few convincing examples exist for epistasis amongst natural polymorphisms in human traits.7,8 Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits,2,3 but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues.9 Here we show that, using advanced computation techniques10 and a gene expression study design, many instances of epis- tasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interac- tions between common SNPs acting on the expression levels of 238 genes (p < 2.91 × 10−16). We tested the discovery interactions for replication in two independent data sets.11,12 Three hundred and forty-five interac- tions had replication interaction p-values that were more extreme than the 2.5% confidence interval of the distribution under the null hypothesis of no epistasis, with 30 significant at a conservative p < 0.05 Bonferroni level. There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 2Mb of re- gions of known intra-cellular chromosome interactions13 (p = 1.8×10−10). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one cis-acting SNP is modulated by several trans- acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by trans-SNPs on 14 different chro- mosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction. This study presents the first evidence for multiple instances of epistatic genetic effects emerging from natural genetic varia- tion in humans.

### 2013-09-10 ld_simulations/extractSnps.R
performing ld simulations

@@ -1,0 +1,117 @@
+# Usage:
+
+# R --no-save --args snplist.txt /ibscratch/wrayvisscher/imputation/arichu/data/imputed/chr\*/arichu_1kg_p1v3_\* outputfile < extractSnps.R
+
+# - The first argument is the SNP list (just a list of rs SNP ids and nothing else)
+# - The second argument specifies the root name for the genotype data (in binary plink format). Replace the chromosome number with "\*"
+# - The third argument is the output file root name that you want to save everything to.
+
+
+library(snpStats)
+
+
+findChromosomes <- function(snp, plinkrt)
+{
+        cmd <- paste("grep -P -m 1 '", snp, "\t' ", plinkrt, ".bim | head -n 1 | cut -d \":\" -f 2 | cut -f 1", sep="")
+    cat(snp)
+    chr <- system(cmd, intern=TRUE)
+    cat(" ", chr, "\n")
+    return(chr)
+}
+
+
+findChromosomesAll <- function(snplist, plinkrt)
+{
+    chr <- snplist
+    for(i in 1:length(snplist))
+    {
+	a <- findChromosomes(snplist[i], plinkrt)
+	chr[i] <- ifelse(is.null(a), NA, a)
+    }
+    dat <- data.frame(snp = snplist, chr = as.numeric(chr))
+    dat <- subset(dat, !is.na(chr))
+    dat$snp <- as.character(dat$snp)
+    dat <- dat[order(dat$chr), ]
+    return(dat)
+}
+
+
+extractSnps <- function(snpnames, plinkrt)
+{
+    require(snpStats)
+    rawdata <- read.plink(bed=plinkrt, select.snps=snpnames)
+    return(rawdata)
+}
+
+
+extractSnpsAll <- function(snpdat, plinkrt)
+{
+    chr <- unique(snpdat$chr)
+    l <- length(chr)
+
+    cat(chr[1], ":", snpdat$snp[snpdat$chr == chr[1]], "\n")
+    dat <- extractSnps(snpdat$snp[snpdat$chr == chr[1]], gsub("\\*", chr[1], plinkrt))
+    if(l == 1) return(dat)
+    for(i in 2:l)
+    {
+	cat(chr[i], ":", snpdat$snp[snpdat$chr == chr[i]], "\n")
[…]

### 2013-09-10 docs/manuscript/methods.tex

@@
 \usepackage[superscript,biblabel]{cite}
+\usepackage{amsmath}
 \bibliographystyle{naturemag}
@@
 
-\subsubsection{Filtering}
-
-We used two approaches to filter SNPs from stage 1 to be tested for significant interaction effects in stage 2.
+\paragraph{Filtering} We used two approaches to filter SNPs from stage 1 to be tested for significant interaction effects in stage 2.
 
@@
 
+\section{Effects of LD on detection and replication}
+
+The power to detect genetic effects, when the observed markers are in LD with the causal variants, is proportional to $r^x$.
+For additive effects $x = 2$, but for non-additive effects $x$ is larger, \emph{i.e.} $x = 4$ for dominance or $A \times A$, $x = 6$ for $A \times D$ or $D \times A$, and $x = 8$ for $D \times D$.
+Many biologically realistic GP maps may be comprised of all 8 variance components \cite{Hemani2013}.
+
+This is important for both detection and for replication of epistasis.
+For detection, if the epistatic effect includes the $D \times D$ term then if the two causal variants are tagged by observed markers that are each in LD $r = 0.9$, then if the true variance is $V_{t}$ then the observed variance $V_{o}$ at the markers will be $0.9^{8}V_{t} = 0.43V_{t}$.
+Therefore, it is important to consider the sampling variation of $\hat{r}^x$ in a sample given some true population value of $r$.
+
+\subsection{Simulations}
+
+For some values of fixed population parameters, $p_{1}$ (minor allele frequency at observed marker), $q_{1}$ (minor allele frequency at causal variant), and $r$ (LD between marker and causal variant), the expected haplotype frequencies are
+
+\begin{align}
+h_{11} &= r \sqrt{p_{1}q_{1}p_{2}q_{2}} + p_{1}q_{1} \\
+h_{12} &= p_{1}q_{2} - r \sqrt{p_{1}q_{1}p_{2}q_{2}} \\
+h_{21} &= p_{2}q_{1} - r \sqrt{p_{1}q_{1}p_{2}q_{2}} \\
+h_{22} &= r \sqrt{p_{1}q_{1}p_{2}q_{2}} + p_{2}q_{2}
+\end{align}
+where $p_{2} = 1 - p_{1}$ and $q_{2} = 1 - q_{1}$.
+For a range of population parameters we randomly sampled $2n$ haplotypes where the expected haplotype frequencies were $h_{11}, h_{12}, h_{21}, h_{22}$.
+From the sample haplotype frequencies we then calculated sample estimates of $\hat{r}$ where
+
+\begin{equation}
+\hat{r} = \frac{\hat{h}_{11} - \hat{p}_{1}\hat{q}_{1}}{\sqrt{\hat{p}_{1}\hat{q}_{1}\hat{p}_{2}\hat{q}_{2}}}
+\end{equation}
+
+For each value of combination of the parameters $p_{1}, q_{1}, r, n$ 1000 simulations were performed and the sampling mean and sampling standard deviation of $\hat{r}, \hat{r}^2, \hat{r}^4, \hat{r}^6, \hat{r}^8$ were recorded.
+
+\subsection{Interpretation}
+
+These simulations show that sampling variance of $r^x$ increases as $x$ increases.
+Detection of epistatis is highly dependent upon high $r^2$.
+Amongst the discovery SNPs there will be a mixture of interactions where observed SNPs are either in true high LD with causal variants, or will have highly inflated sample $\hat{r}^{x}$ compared to the population $r^x$.
+Supposing that SNPs
+
 \section{Additive and non-additive variance estimation}

### 2013-09-12 docs/manuscript/methods.tex
work continues

@@
 
-\subsection{Simulations}
+\subsection{Simulation 1}
 
@@
 For each value of combination of the parameters $p_{1}, q_{1}, r, n$ 1000 simulations were performed and the sampling mean and sampling standard deviation of $\hat{r}, \hat{r}^2, \hat{r}^4, \hat{r}^6, \hat{r}^8$ were recorded.
+It was observed that sampling variance increases for increasing $x$ in $\hat{r}^x$.
+
+\subsection{Simulation 2}
+
+We assume that the discovery SNP pairs are ascertained (from a very large number of tests) have high $\hat{r}$ between observed SNPs and causal variants because otherwise power of detection would be low.
+We can hypothesis that the distribution of $\hat{r}$ in this ascertained sample will be a mixture of $r$ that is high and $r$ that is lower but with ascertained higher values from sampling.
+Therefore, we would expect those with truly high $r$ to have a higher replication rate in independent datasets, and those with ascertained high $\hat{r}$ to have lower replication because resampling is unlikely to result in the same extreme ascertainment.
+To obtain empirical estimates of $\hat{r}$ in discovery and replication datasets we conducted the following simulation.
+
+\begin{enumerate}
+\item Using 1000 genomes data (phase 1, version 3, 379 European samples) we selected the 528,509 ``markers'' used in the original discovery analysis, plus 100,000 randomly chosen ``causal variants'' (CVs) with minor allele frequence $> 0.05$.
+\item The 379 individuals were split into discovery (190) and replication (189) sets.
+\item For each CV the marker with the maximum $\hat{r}^2_{D}$ from the marker panel was recorded in the discovery set. This marker was known as the ``discovery marker'' (DM).
+\item The $\hat{r}^2_{R}$ for each CV/DM pair was then calculated in the replication set where the discovery LD was ascertained to be high, such that $\hat{r}^2_{D} > 0.9$.
+\end{enumerate}
+
+We observed that there was an average decrease in $\hat{r}^x_{R}$ relative to $\hat{r}^x_{D}$, and that this decrease was larger with increasing $x$.
+We observed that $(\hat{r}^2_{R} - \hat{r}^2_{D}) / \hat{r}^2_{D} = 0.029$ whereas $(\hat{r}^8_{R} - \hat{r}^8_{D}) / \hat{r}^8_{D} = 0.092$.
+The average drop in in replication $\hat{r}^8$ was 3 times higher than the drop in $\hat{r}^2$.
 
@@
 
-These simulations show that sampling variance of $r^x$ increases as $x$ increases.
-Detection of epistatis is highly dependent upon high $r^2$.
+Simulation 1 shows that sampling variance of $r^x$ increases as $x$ increases.
+Detection of epistatis is highly dependent upon high $\hat{r}$.
 Amongst the discovery SNPs there will be a mixture of interactions where observed SNPs are either in true high LD with causal variants, or will have highly inflated sample $\hat{r}^{x}$ compared to the population $r^x$.
-Supposing that SNPs
+Simulation 2 shows that as $x$ gets larger, the average decrease in $\hat{r}^x$ between discovery and replication becomes larger, likely to be a result of ascertained high $\hat{r}$ in the discovery and increased sampling variance with increasing $x$ in the replication.
+These results demonstrate that assuming all else is equal, sampling variance of $r$ alone will reduce the replication rate of epistatic effects compared to additive effects.
 

### 2013-09-12 ld_simulations/getMaxLd.R

@@ -171,6 +171,7 @@ b_all <- rbind(b2, b4, b6, b8)
 
 library(ggplot2)
 
+
 ggplot(b_all, aes(x=max_rsq, y=replication_rsq)) +
 geom_point() +
 geom_abline(intercept=0, slope=1, colour="blue") +
@@ -180,15 +181,31 @@ labs(y = "Replication", x = "Discovery")
 ggsave("~/repo/eQTL-2D/analysis/images/ld_ascertained.png")
 
 
-di <- sort(b$max_rsq)
-re <- sort(b$replication_rsq)
-di <- sort(di[sample(1:length(di), length(re), replace=FALSE)])
-plot(di, re)
-plot(di^8, re^8)
+library(plyr)
 
-length(di)
-length(re)
+a <- ddply(b_all, .(pow), function(x)
+{
+	x <- mutate(x)
+	m <- mean(x$max_rsq - x$replication_rsq, na.rm=T)
+	return(m)
+})
 
-quantile(b$max_rsq)^8
-quantile(b$replication_rsq, na.rm=T)^8
+
+a <- ddply(b_all, .(pow), function(x)
+{
+	x <- mutate(x)
+	m <- mean((x$replication_rsq - x$max_rsq) / x$max_rsq, na.rm=T)
+
+	return(m)
+})
+a
+
+b_all$rdiff <- b_all$replication_rsq - b_all$max_rsq
+b_all$prdiff <- (b_all$replication_rsq - b_all$max_rsq) / b_all$max_rsq
+ggplot(b_all, aes(x=factor(pow), y=prdiff)) + 
+geom_boxplot() +
+labs(y = expression( (r[R]^x - r[D]^x)/r[D]^x) , x = "Power term")
+ggsave("~/repo/eQTL-2D/analysis/images/ld_reduction.pdf")
+
+
 

### 2013-09-13 docs/manuscript/manuscript.tex
The main text and methods have been updated.

@@
 A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}).
-It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice because the dependence on LD between observed SNPs and causal variants is up to three orders of magnitude higher than it is for independent additive effects (Supplementary Figure \ref{fig:ld_mean}) \cite{Weir2008, Hemani2013}, and for higher power terms of LD $r$ the sampling variance increases (Methods, Supplementary Figure \ref{fig:ld_sd}).
+
+It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice for several reasons.
+For example, LD between causal variants and observed markers plays an important role.
+Not only is the dependence on LD much greater for epistatic effects than for additive effects (Supplementary Figure \ref{fig:ld_sampling_mean}), but when estimating epistatic variance it is more sensitive to changes in LD between observed SNPs and causal variants between independent samples when compared to additive effects (Supplementary Figure \ref{fig:ld_sampling_sd}), and this has a direct effect on statistical power for replication.
+Indeed, in addition to heterogeneity between populations, sampling variance of LD $r$ alone leads to over three fold decrease in observed variance for the most extreme case of $D \times D$ effects when compared to the average decrease for additive variance on average (Methods, Supplementary Figure \ref{fig:ld_reduction}).
 Therefore these results are encouraging with regards to the detection and replication of epistasis.
@@
 Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question.
-In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold.
-How does this compare to the number of traits influenced by additive effects?
-The BSGS dataset has been previously analysed for additive effects at all expression traits \cite{Powell2012}, and if we take all the additive eQTLs that were significant at the epistatic threshold of $p < 2.91 \times 10^{-16}$ we find that 453 gene expression levels out of the 7339 analysed had at least one significant expression quantitative trait locus (eQTL).
-Therefore it can be argued that the number of instances of detectable epistasis is substantial.
-
-However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}.
+In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance.
+Taking published results from a standard eQTL study \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the additive variance of a locus was $2.1\%$ or greater.
+Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.
+
+In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}.
 Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components.
 Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}.
-Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that on average they explain 1.73\% of the phenotypic variance of each of the 7339 probes.
-By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain 0.25\% of phenotypic variance, approximately seven times lower than the additive variance.
+Taking the additive effects detected in Powell \emph{et al} (2012) where the minimum additive variance was $2.1\%$, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$.
+By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain a total of 0.22\% of phenotypic variance, approximately ten times lower than the additive variance.
 There are several caveats to this comparison.
-Firstly, the ratio of additive to epistatic variance may differ at different effect sizes, and our estimate is determined by the threshold used.
+Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used.
 Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
@@
 As true $r$ reduces the statistical power to detect epistatic variants drops dramatically under the assumption that statistical power is proportional to higher moments of $r$.}
-	\label{fig:ld_mean}
+	\label{fig:ld_sampling_mean}
 \end{figure}
@@
 Supposing that there is sufficiently high $r^x$ in the discovery sample for detection of epistasis, the replication sample is less likely to have similarly high $r^x$ as $x$ increases, leading to an expectation of reduced replication rates.}
-	\label{fig:ld_sd}
+	\label{fig:ld_sampling_sd}
+\end{figure}
+\clearpage
+
+\begin{figure}
+	\centering
+	\includegraphics[width=5in]{ld_reduction}
+	\caption{\textbf{Reduction in LD as estimated in replication data after ascertaining for high LD in discovery data} 100,000 ``unobserved'' causal variants (CVs) were tested for LD against a panel of 528,509 ``observed'' discovery markers (DMs).
+DM/CV pairs with LD $r^2 > 0.9$ were then tested in an independent sample.
+Simulation results of the proportional decrease between discovery and replication datasets in LD ($y$-axis) of $\hat{r}^2, \hat{r}^4, \hat{r}^6, \hat{r}^8$ ($x$-axis) are shown, where $\hat{r}^{x}_{D}$ and $\hat{r}^{x}_{R}$ are the sample LD measurements in the discovery and replication datasets, respectively.
+The average proportional decrease in the replication $\hat{r}^{x}_{R}$ was $2.8\%, 5.3\%, 7.4\%$ and $9.2\%$ for $x=2,4,6$ and $8$, respectively.}
+	\label{fig:ld_reduction}
 \end{figure}

### 2013-09-13 docs/manuscript/methods.tex
The main text and methods have been updated.

@@
 Simulation 2 shows that as $x$ gets larger, the average decrease in $\hat{r}^x$ between discovery and replication becomes larger, likely to be a result of ascertained high $\hat{r}$ in the discovery and increased sampling variance with increasing $x$ in the replication.
-These results demonstrate that assuming all else is equal, sampling variance of $r$ alone will reduce the replication rate of epistatic effects compared to additive effects.
+These results demonstrate that if all else is equal, the impact of sampling variance of $r$ alone will reduce the replication rate of epistatic effects compared to additive effects.
 

### 2013-09-13 analysis/variance_analysis.R

@@ -15,10 +15,10 @@ hist(v$V1)
 min(v$V1)
 
 # Read in additive effects list
-b <- read.table("filtering/marginal_lists/FDR_0_05_genotyped_eQTL.txt", head=T)
+b <- read.table("~/repo/eQTL-2D/filtering/marginal_lists/FDR_0_05_genotyped_eQTL.txt", head=T)
 
 # Read in probenames used in analysis
-probenames <- scan("data/probenames_used_in_analysis.txt", what="character")
+probenames <- scan("~/repo/eQTL-2D/data/probenames_used_in_analysis.txt", what="character")
 
 # Calculate additive variance for each additive effect
 b$va <- with(b, 2 * FREQ1 * (1-FREQ1) * EFFECT^2)
@@ -26,8 +26,13 @@ b$va <- with(b, 2 * FREQ1 * (1-FREQ1) * EFFECT^2)
 # Get all additive effects for 7339 probes that are larger than the minimum epistatic variance
 b1 <- subset(b, TRAIT %in% probenames & va >= min(v$V1))
 
+# Remove duplicated genes
+load("~/repo/eQTL-2D/data/residuals_all.RData")
+genes <- subset(probeinfo, select=c(PROBE_ID, ILMN_GENE))
+b1 <- merge(b1, genes, by.x="TRAIT", by.y="PROBE_ID", all.x=TRUE)
+
 # Choose the highest additive for each chr x trait combination
-b1$code <- with(b1, paste(CHR, TRAIT))
+b1$code <- with(b1, paste(CHR, ILMN_GENE))
 b1 <- b1[order(b1$va, decreasing=T), ]
 b1 <- b1[!duplicated(b1$code), ]
 dim(b1)

### 2013-09-13 ld_simulations/getMaxLd.R

@@ -204,7 +204,7 @@ b_all$rdiff <- b_all$replication_rsq - b_all$max_rsq
 b_all$prdiff <- (b_all$replication_rsq - b_all$max_rsq) / b_all$max_rsq
 ggplot(b_all, aes(x=factor(pow), y=prdiff)) + 
 geom_boxplot() +
-labs(y = expression( (r[R]^x - r[D]^x)/r[D]^x) , x = "Power term")
+labs(y = expression( (hat(r)[R]^x - hat(r)[D]^x)/hat(r)[D]^x) , x = "Power term")
 ggsave("~/repo/eQTL-2D/analysis/images/ld_reduction.pdf")
 
 

### 2013-09-13 docs/manuscript/manuscript.tex
updated

@@
 Transcription levels can be measured for thousands of genes.
-These traits are largely heritable but on average less polygenic than high level phenotypes, thus many genetic effects are relatively large \cite{Powell2013}, maximising the chance at detecting epistasis, should it exist.
+These traits are largely heritable but on average have been shown to be less polygenic than high level phenotypes \cite{Powell2013}, thus many genetic effects are relatively large \cite{Powell2012}, maximising the chance at detecting epistasis, should it exist.
 

### 2013-09-13 docs/manuscript/manuscript.tex
minor edits to methods and manuscript

@@
 \usepackage{lscape}
+\usepackage{lineno}
 
@@
 \makeatother
+\linenumber
 
@@
 In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
-Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 434$.
+Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$.
 There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
@@
 In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions.
-First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (5\% significance threshold $ p < 2.91 \times 10^{-16}$, Methods).
+First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods).
 Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $\alpha \approx 0.07$, Methods, Supplementary Figure \ref{fig:type1}).
@@
 Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
-We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/434$, Table \ref{tab:bonferroni}).
+We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/501$, Table \ref{tab:bonferroni}).
 To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods).
@@
 Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
-And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.
+And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance.
+This is due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.
 

### 2013-09-13 docs/manuscript/methods.tex
minor edits to methods and manuscript

@@
 \usepackage{amsmath}
+\usepackage{lineno}
+
 \bibliographystyle{naturemag}
@@
 \author{}
+\linenumber
 
@@
 
-\paragraph{Filter 2} We also performed a second filtering screen applied to the list of SNP pairs from stage 1 that was identical to filter 1 but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) was removed \cite{Powell2013}, creating a second set of 4751 unique filtered SNP pairs.
+\paragraph{Filter 2} We also performed a second filtering screen applied to the list of SNP pairs from stage 1 that was identical to filter 1 but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) were removed \cite{Powell2013}, creating a second set of 4751 unique filtered SNP pairs.
 

### 2013-09-13 docs/manuscript/manuscript.tex
updated response to reviewers

@@
 \usepackage{lineno}
+\linenumbers
 
@@
 \makeatother
-\linenumber
 
@@
 
-In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}).
+In addition, using the meta analysis from the replication samples only, we observed that 316 of the remaining 404 discovery SNP pairs had replication interaction $p$-values more extreme than the 2.5\% confidence interval of the quantile-quantile plot against the null hypothesis of no interactions where $p$-values are assumed to be uniformly distributed ($p << 1.0 \times 10^{-16}$, Figure \ref{fig:qqMeta} and Supplementary Figure \ref{fig:qqMetaNonsig}).
 Concordance of the direction of the effect of the largest variance component was also highly significant ($p = 5.71 \times 10^{-31}$, Supplementary Table \ref{tab:sign}).

### 2013-09-16 docs/manuscript/manuscript.tex
changed some things in response to peter's comments

@@
 \author[7]{Anjali K Henders}
-\author[1,2]{Allan F. McRae}
+\author[1,2]{Allan F McRae}
 \author[2]{Jian Yang}
@@
 In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
-Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$.
+Replication of these interactions in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$.
 There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
@@
 For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-This study presents the first evidence for multiple instances of natural genetic polymorphisms interacting to influence human traits.
+This study presents the first evidence for multiple instances of segregating common polymorphisms interacting to influence human traits.
 
@@
 One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants.
-Because our focus was to ascertain the extent to which instances of epistasis occur amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels.
-Transcription levels can be measured for thousands of genes.
-These traits are largely heritable but on average have been shown to be less polygenic than high level phenotypes \cite{Powell2013}, thus many genetic effects are relatively large \cite{Powell2012}, maximising the chance at detecting epistasis, should it exist.
+Because our focus was to ascertain the extent to which instances of epistasis arises from natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels.
+Transcription levels can be measured for thousands of genes and like most complex diseases, these expression traits are typically heritable \cite{Schadt2003}.
+But unlike complex diseases, genetic associations with gene expression commonly have very large effect sizes that explain large proportions of the genetic variance\cite{Powell2013}, making them good candidates to search for epistasis, should it exist.
 
@@
 First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods).
-Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $\alpha \approx 0.07$, Methods, Supplementary Figure \ref{fig:type1}).
+Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $0.05 \leq \alpha \leq 0.14$, Methods, Supplementary Figure \ref{fig:type1}).
 Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}).
@@
 
-It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice for several reasons.
+It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistasis is difficult in practice.
 For example, LD between causal variants and observed markers plays an important role.
 Not only is the dependence on LD much greater for epistatic effects than for additive effects (Supplementary Figure \ref{fig:ld_sampling_mean}), but when estimating epistatic variance it is more sensitive to changes in LD between observed SNPs and causal variants between independent samples when compared to additive effects (Supplementary Figure \ref{fig:ld_sampling_sd}), and this has a direct effect on statistical power for replication.
-Indeed, in addition to heterogeneity between populations, sampling variance of LD $r$ alone leads to over three fold decrease in observed variance for the most extreme case of $D \times D$ effects when compared to the average decrease for additive variance on average (Methods, Supplementary Figure \ref{fig:ld_reduction}).
-Therefore these results are encouraging with regards to the detection and replication of epistasis.
+The sampling variance of LD $r$ leads to the ascertainment of marker associations with higher sample $r$ in the discovery stage in comparison to the replication stage.
+However, the average decrease in $\hat{r}^{x}$ in replication samples becomes larger as $x$ increases (Methods, Supplementary Figure \ref{fig:ld_reduction}).
+For example, the decrease in $\hat{r}^8$ (which is proportional to the power of deteting $D \times D$ effects), is on average three fold greater than the decrease in $\hat{r}^2$ (which is proportional to the power of detecting additive effects).
 
@@
 Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question.
-In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance.
-Taking published results from a standard eQTL study \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the additive variance of a locus was $2.1\%$ or greater.
+In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance.
+Taking results from our previously published eQTL \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater.
 Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.
@@
 Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components.
-Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}.
-Taking the additive effects detected in Powell \emph{et al} (2012) where the minimum additive variance was $2.1\%$, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$.
-By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain a total of 0.22\% of phenotypic variance, approximately ten times lower than the additive variance.
+Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that estimated additive variance is overall a larger component than estimated epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}.
+Taking all additive effects detected in Powell \emph{et al} (2012) that have additive variance explaining $2.1\%$ or greater of phenotypic variance, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$.
+By contrast, the estimated epistatic variance from the interacting SNPs detected in this study on average explain a total of $0.22\%$ of phenotypic variance, approximately ten times lower than the estimated additive variance.
 There are several caveats to this comparison.
@@
 Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
-And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance.
+Thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance.
 This is due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.
+And forthly, the extent of winner's curse in estimation of effect sizes may differ between the the two studies.
 
@@
 Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
-The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
+The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
 Of these, 434 passed filtering in both replication datasets.
@@
 	\includegraphics[width=5in]{type1}
-	\caption{\textbf{Type 1 error rate of two stage design} In stage 1 SNPs are tested for full genetic effects (8 d.f.) and those that surpass a threshold for multiple testing are then tested for significant interaction terms in stage 2.
+	\caption{\textbf{Type 1 error rate of two stage design assuming a null model of one large additive effect and no epistasis} In stage 1 SNPs are tested for full genetic effects (8 d.f.) and those that surpass a threshold for multiple testing are then tested for significant interaction terms in stage 2.
 These interaction $p$-values are then adjusted (Bonferroni) for the total number of tests that passed stage 1.
@@
 	\includegraphics[width=5in]{ld_sampling_mean}
-	\caption{\textbf{Sampling mean for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$.
-For a given a population value of LD $r$ (columns of plots), plotted is the sample mean ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines).
+	\caption{\textbf{Sampling mean for different power terms of population $r$ values} Power of detection and replication of epistatic interactions depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$.
+For a given population value of LD $r$ (columns of plots), plotted is the sample mean ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines).
 As true $r$ reduces the statistical power to detect epistatic variants drops dramatically under the assumption that statistical power is proportional to higher moments of $r$.}
@@
 	\includegraphics[width=5in]{ld_sampling_sd}
-	\caption{\textbf{Sampling standard deviation for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$.
+	\caption{\textbf{Sampling standard deviation for different power terms of population $r$ values} Power of detection and replication of epistatic interactions depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$.
 For a given a population value of LD $r$ (columns of plots), plotted is the sampling standard deviation ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines).

### 2013-09-19 docs/manuscript/manuscript.tex
updated the manuscript with new cis-cis statistics

@@
 
-Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome.
+Epistasis is the phenomenon whereby a polymorphism's effect on a trait depends on other polymorphisms present in the genome.
 The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics.
@@
 Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}).
+We used strict quality control measures to avoid statistical associations being driven by technical artifacts (Methods).
+However it remains possible that unexplained technical artifacts may have led to the significant discovery interactions.
 Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
@@
 For example, LD between causal variants and observed markers plays an important role.
-Not only is the dependence on LD much greater for epistatic effects than for additive effects (Supplementary Figure \ref{fig:ld_sampling_mean}), but when estimating epistatic variance it is more sensitive to changes in LD between observed SNPs and causal variants between independent samples when compared to additive effects (Supplementary Figure \ref{fig:ld_sampling_sd}), and this has a direct effect on statistical power for replication.
+Not only is the dependence on LD much greater for epistatic effects than for additive effects (Supplementary Figure \ref{fig:ld_sampling_mean}), but when estimating epistatic variance it is more sensitive to changes in LD between observed SNPs and causal variants between independent samples when compared to additive effects (Supplementary Figure \ref{fig:ld_sampling_sd}).
+This has a direct effect on statistical power for replication.
 The sampling variance of LD $r$ leads to the ascertainment of marker associations with higher sample $r$ in the discovery stage in comparison to the replication stage.
@@
 
-Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene, median distance between interacting SNPs is 1.83Mb), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
+Of the discovery interactions, 26 were \emph{cis}-\emph{cis} acting (within 1Mb of the transcription start site, mean distance between SNPs was 0.53Mb), 462 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting.
 We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
@@
 The bioinformatic analysis of the significant epistatic loci suggests that there are a large number of possible mechanisms that can lead to non-additive genetic variation.
-Further research into such epistatic effects may provide a useful framework to understanding molecular mechanisms and complex trait variation in greater detail.
+Further research into such epistatic effects may provide a useful framework for understanding molecular mechanisms and complex trait variation in greater detail.
 With computational techniques and data now widely available the search for epistasis in larger datasets for traits of broader interest is warranted.
@@
 \begin{threeparttable}
-\caption{\textbf{Details on linkage disequilibrium and relative positions of all discovery \emph{cis-cis} interactions}}
+\caption{\textbf{Details on linkage disequilibrium and relative positions of all discovery interactions with SNPs on the same chromosome}}
 \label{tab:ld}

### 2013-09-23 docs/manuscript/manuscript.tex
manuscript is sent for appeal but now making changes in the meantime

@@
 Three hundred and forty-five interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data The remaining 89 interactions that did not replicate are depicted in grey.
-It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
+It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.
+An interactive version of this graph can be found here: }
 	\label{fig:fireworks}

### 2013-10-09 presentation/Brisbane_stat_genetics/BSGS.tex
presentation for BSGF - Friday 11th Oct 2013

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
% Epistasis: Brisbane Statistical Genetics Forum 
% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%   Author: Joseph Powell; joseph.powell@uq.edu.au
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{beamer}
\usepackage{beamerthemeshadow}
\usepackage[latin1]{inputenc}
\usepackage{comment}
\usetheme{Frankfurt}
%\usefonttheme{serif}

%\setbeamercolor{structure}{rgb}{}
\usecolortheme[RGB={0,160,176}]{structure}

\usepackage[absolute,overlay]{textpos} 
\usepackage{booktabs}
\usepackage{color}
\usepackage{threeparttable}
\usepackage{multirow}
\usepackage[normalem]{ulem}

\begin{document}

\title[Epistasis influencing transcription in humans]{Detection and replication of epistasis influencing transcription in humans}
\author{Joseph Powell and Gibran Hemani}
\institute{The University of Queensland Diamantina Institute \\
\vspace{.2cm}
and \\
\vspace{.2cm}
Queensland Brain Institute}
\date{11th Oct 2013}

\begin{frame}
\titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\subsection{}
\begin{frame}{What is epistasis?}
\begin{definition}
{The effect on the phenotype caused by locus A depends on the genotype at locus B....}
\end{definition}
\vspace{0.5cm}
Epistasis has been reported in model organisms through artificial gene knockouts, artificial line crosses and hybridisation. \\
\vspace{0.5cm}
Our aim was to systematically search for instances of epistasis amongst common variants for genetic variation that has arisen in natural populations.\end{frame}

\begin{frame}{Gene expression}
\begin{columns}[c]
\column{.5\textwidth} 
\begin{itemize}
\item Transcription measured for thousands of genes
\vspace{0.3cm}
\item Typically heritable 
\vspace{0.3cm}
\item Loci (eQTLs) commonly have very large effect sizes
\vspace{0.3cm}
\item Good candidates to search for epistasis
\end{itemize}
\column{.5\textwidth}
\begin{center}
\includegraphics[width=2.0cm]{images/ht12.png}
\end{center}
\end{columns}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Methods}
\subsection{}
\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods1.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods2.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{columns}[c]
\column{.5\textwidth}
\includegraphics[width=4.5cm]{images/BSGS.png} \\
\column{.5\textwidth} 
\begin{itemize}
\item 846 healthy individuals 
\vspace{0.3cm}
\item 528,509 autosomal SNPs
\vspace{0.3cm}
\item RNA measued for peripheral blood (Illumina HT-12v4.0)
\vspace{0.3cm}
\item Expression levels for 7,339 probes 
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods3.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{columns}[c]
\column{.5\textwidth}
\includegraphics[width=4.5cm]{images/exhaustive.png} \\
\column{.5\textwidth} 
\begin{itemize}
\item Exhaustive SNP x SNP testing for each probe
\vspace{0.3cm}
\item epiGPU software and GPU clusters
\vspace{0.3cm}
\item 8 d.f. F-test 
\vspace{0.3cm}
\item Over quadrillion tests  
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods4.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{columns}[c]
\column{.5\textwidth}
\includegraphics[width=4.5cm]{images/fam_wise.pdf} \\
\column{.5\textwidth} 
\begin{itemize}
\item Permutation and bonferoni used to give 5\% FWER
\vspace{0.2cm} 
\item Permutation: Single probe FWER;  $T*= 2.13x10^{-12}$ 
\vspace{0.2cm} 
\item Correct for 7,339 probes
\vspace{0.2cm} 
\item Experiment wide FWER; $T*/7,339=2.91x10^{-16}$  
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods5.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{columns}[c]
\column{.5\textwidth}
\includegraphics[width=4.5cm]{images/filter.png} \\
\column{.5\textwidth} 
\begin{itemize}
\item For SNP pairs that passed the $2.91x10^{-16}$ threshold
\vspace{0.1cm} 
\item All 9 genotypes classes present
\vspace{0.1cm} 
\item Minimum class size of 5
\vspace{0.1cm} 
\item No LD between SNP pairs ($r^2 < 0.1$ and $D'^2 < 0.1$)
\vspace{0.1cm} 
\item No single loci additive or dominance effects
\vspace{0.1cm} 
\item 11,155 pairs carried forward
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods6.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{columns}[c]
\column{.5\textwidth}
\includegraphics[width=4.5cm]{images/second.png} \\
\column{.5\textwidth} 
\begin{itemize}
\item Nested ANOVA for 11,155 pairs 
\vspace{0.2cm} 
\item Contrasts full genetic (8 d.f.) vs marginal effects (4 d.f)
\vspace{0.2cm} 
\item Thus, testing for the contribution of epistatic variance 
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods7.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{columns}[c]
\column{.5\textwidth}
\includegraphics[width=4.5cm]{images/bonf.png} \\
\column{.5\textwidth} 
\begin{itemize}
\item Epistatic effects significant at $p < 0.05/11,155 = 4.48x10^{-6}$
\vspace{0.2cm} 
\item 501 SNP pairs with significant interaction terms 
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods8.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{columns}[c]
\column{.5\textwidth}
\includegraphics[width=4.5cm]{images/rep1.png} \\
\column{.5\textwidth} 
Fehrmann
\begin{itemize}
\item 1,240 individuals (Netherlands) 
\end{itemize}
\vspace{0.3cm}
EGCUT
\begin{itemize}
\item 891 individuals (Estonian)
\vspace{0.3cm}
\item RNA measued for peripheral blood (Illumina HT-12v3.0)
\vspace{0.1cm} 
\item Genotyped with Illumina arrays
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods9.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{columns}[c]
\column{.5\textwidth}
\includegraphics[width=4.5cm]{images/rep2.png} \\
\column{.5\textwidth} 
\begin{itemize}
\item Replication of 501 epistatic hits using two independent cohorts 
\vspace{0.1cm}
\item Same filtering criteria
\vspace{0.1cm}
\item 434 pairs which could be tested in both cohorts 
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=7cm]{images/methods10.png} \\
\end{center}
\end{frame}

\begin{frame}
\begin{columns}[c]
\column{.5\textwidth}
\includegraphics[width=4.5cm]{images/func.png} \\
\column{.5\textwidth} 
Analyses to elucidate possible functional mechanisms
\begin{itemize}
\item Non-synonymous mutations
\vspace{0.1cm}
\item GWAS and known eQTL overlap
\vspace{0.1cm}
\item Genome segmentation
\vspace{0.1cm}
\item Chromosome interactions
\vspace{0.1cm}
\item Tissue specific transcription regions
\end{itemize}
\end{columns}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}
\subsection{}
\begin{frame}{Q-Q plots of interaction p-values from replication}
\begin{columns}[c]
\column{.5\textwidth}
The top panel shows all 434 discovery SNPs that were tested for interactions. \\
\vspace{0.3cm}
The bottom panel shows the same data as the top panel but excluding the 30 most significant interactions \\
\vspace{0.3cm}
Matched direction of epistatic effects $p=5.56x10^{-31}$ 
\column{.5\textwidth} 
\begin{center}
\includegraphics[height=6.5cm]{images/qqMeta.pdf} \\
\end{center}
\end{columns}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=8cm]{images/fireworks2.png} \\
\end{center}
\end{frame}

\begin{frame}{MBNL1}
\begin{columns}[c]
\column{.5\textwidth}
Involved in RNA modification and regulation of splicing \\
\vspace{0.3cm}
Has one cis-effect, controlled by 13-trans SNPs \\   
\vspace{0.3cm}
Each pair is independent \\
\vspace{0.3cm}
Masking pattern: when the trans-SNP is homozygous for the masking allele the decreasing allele of the cis-SNP no longer has an effect
\column{.5\textwidth} 
\begin{center}
\includegraphics[height=3.0cm]{images/MBNL1.pdf} \\
\end{center}
\end{columns}
\end{frame}

\begin{frame}{Chromosome interactions}
\begin{columns}[c]
\column{.5\textwidth}
Are epistatic SNPs located in positions known to interact? 
\begin{itemize}
\item Chromosome interactions identifyed in K562 cell lines using Hi-C and ChIP-seq
\item red lines = $N$ SNPs within 20kb, 500kb, 2Mb and 10Mb of known interacting regions
\item Histograms = null distributions
\end{itemize}
\column{.5\textwidth} 
\begin{center}
\includegraphics[height=4.5cm]{images/chr_int.pdf} \\
\end{center}
\end{columns}
\end{frame}

\begin{frame}{Conclusions}
This study presents the first evidence for multiple instances of segregating common polymorphisms interacting to influence human traits. \\
\vspace{0.4cm}
Computational and statistical frameworks can identify and replicate epistasis in complex traits. \\
\vspace{0.4cm}
Functional characterization of epistatic loci suggests a large number of possible mechanisms that can lead to non-additive genetic variation.
\end{frame}

\section*{}
\begin{frame}{Acknowledgements}
\begin{columns}
\column{.5\textwidth}
{\tiny
Complex Trait Genomics Group
\begin{itemize}
\item Gib Hemani
\item Kostya Shakhbazov
\item Jian Yang
\item Allan Mcrae
\item Peter Visscher
\end{itemize}
University of Groningen
\begin{itemize}
\item Harm-Jan Westra
\item Lude Franke
\end{itemize}
}
\column{.5\textwidth}
{\tiny
QIMR Berghofer
\begin{itemize}
\item Nick Martin
\item Anjali Henders
\item Grant Montgomery
\end{itemize}
Georgia Institute of Technology
\begin{itemize}
\item Greg Gibson
\end{itemize}
University of Tartu
\begin{itemize}
\item Andres Metspalu
\item Tonu Esko
\end{itemize}
Visit us at www.complextraitgenomics.com
}
\end{columns}
\end{frame}

\end{document}

### 2013-12-09 docs/manuscript/methods.tex
updated response to reviewers
manuscript accepted in principle! need to make it shorter.

@@
 \author{}
-\linenumber
 
@@
 \tableofcontents
+
+\section{Methods Summary}
+We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs.
+Each individual had gene expression levels measured in peripheral blood at 47,323 probes.
+Only the probes that passed quality control and had significant expression in $\geq 90\%$ of individuals were used in the analysis (7,339 probes representing 6,158 RefSeq genes).
+Recent hardware and software \cite{Hemani2011} advances that use graphics processing units (GPUs) made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
+We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
+SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
+Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}.
+Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
+All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects.
+All SNP pairs were required to have at least five data points in all nine genotype classes.
+If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained.
+Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
+The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
+Of these, 434 passed filtering in both replication datasets.
+A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.
 

### 2013-12-09 docs/manuscript/shorter.tex
manuscript accepted in principle! need to make it shorter.

--- 2013-09-23 docs/manuscript/manuscript.tex
@@
 \usepackage{lineno}
+\usepackage{hyperref}
 \linenumbers
@@
 
-Epistasis is the phenomenon whereby a polymorphism's effect on a trait depends on other polymorphisms present in the genome.
+Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome.
+The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics.
+Though often demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been reported in other species \cite{Carlborg2006}, few examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
+Its absence from empirical findings may simply be due to low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical and computational issues \cite{Cordell2009}.
+Here we show that, using advanced computation \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
+In a cohort of 846 individuals with 7339 gene expression levels measured in peripheral blood, we found 501 significant pairwise interactions between common SNPs influencing the expression of 238 genes ($p < 2.91 \times 10^{-16}$).
+Replication of these interactions in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$.
+Forty-four of the genetic interactions are located within 2Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
+Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs.
+For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
+This study presents the first evidence for multiple instances of segregating common polymorphisms interacting to influence human traits.
+
+% Epistasis is the phenomenon whereby a polymorphism's effect on a trait depends on other polymorphisms present in the genome.
 The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics.
@@
 In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions.
-First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods).
+First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood ($1.03 \times 10^{15}$ statistical tests, family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods).
 Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $0.05 \leq \alpha \leq 0.14$, Methods, Supplementary Figure \ref{fig:type1}).
@@
 A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}).
-
-It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistasis is difficult in practice.
-For example, LD between causal variants and observed markers plays an important role.
-Not only is the dependence on LD much greater for epistatic effects than for additive effects (Supplementary Figure \ref{fig:ld_sampling_mean}), but when estimating epistatic variance it is more sensitive to changes in LD between observed SNPs and causal variants between independent samples when compared to additive effects (Supplementary Figure \ref{fig:ld_sampling_sd}).
-This has a direct effect on statistical power for replication.
-The sampling variance of LD $r$ leads to the ascertainment of marker associations with higher sample $r$ in the discovery stage in comparison to the replication stage.
-However, the average decrease in $\hat{r}^{x}$ in replication samples becomes larger as $x$ increases (Methods, Supplementary Figure \ref{fig:ld_reduction}).
-For example, the decrease in $\hat{r}^8$ (which is proportional to the power of deteting $D \times D$ effects), is on average three fold greater than the decrease in $\hat{r}^2$ (which is proportional to the power of detecting additive effects).
+It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistasis is difficult in practice due to LD (Methods).
 
@@
 We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 5Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
-Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Osborne2004, Rieder2012}.
+Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Rieder2012}.
 
@@
 In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}.
-Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components.
-Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that estimated additive variance is overall a larger component than estimated epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}.
 Taking all additive effects detected in Powell \emph{et al} (2012) that have additive variance explaining $2.1\%$ or greater of phenotypic variance, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$.
@@
 Thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance.
-This is due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.
 And forthly, the extent of winner's curse in estimation of effect sizes may differ between the the two studies.
@@
 
-\subsection*{Methods Summary}
-We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs.
-Each individual had gene expression levels measured in peripheral blood at 47,323 probes.
-Only the probes that passed quality control and had significant expression in $\geq 90\%$ of individuals were used in the analysis (7,339 probes representing 6,158 RefSeq genes).
-Recent hardware and software \cite{Hemani2011} advances that use graphics processing units (GPUs) made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
-We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
-SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
-Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}.
-Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
-All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects.
-All SNP pairs were required to have at least five data points in all nine genotype classes.
-If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained.
-Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
-The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
-Of these, 434 passed filtering in both replication datasets.
-A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.
-
-\subsection*{Acknowledgements}
-
-We are grateful to the volunteers for their generous participation in these studies. We thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.
-
-This work could not have been completed without access to high performance GPGPU compute clusters.
+% \subsection*{Acknowledgements}
+
+% We are grateful to the volunteers for their generous participation in these studies. We thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.
+
+% This work could not have been completed without access to high performance GPGPU compute clusters.
 We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au).
@@
 
-The University of Queensland group is supported by the Australian National Health and Medical Research Council (NHMRC) grants 389892, 496667, 613601, 1010374 and 1046880, the Australian Research Council (ARC) grant (DE130100691), and by National Institutes of Health (NIH) grants GM057091 and GM099568.
-
-The QIMR researchers acknowledge funding from the Australian National Health and Medical Research Council (grants 241944, 389875, 389891, 389892, 389938, 442915, 442981, 496739, 496688 and 552485), the and the National Institutes of Health (grants AA07535, AA10248, AA014041, AA13320, AA13321, AA13326 and DA12854).
+% The University of Queensland group is supported by the Australian National Health and Medical Research Council (NHMRC) grants 389892, 496667, 613601, 1010374 and 1046880, the Australian Research Council (ARC) grant (DE130100691), and by National Institutes of Health (NIH) grants GM057091 and GM099568.
+
+% The QIMR researchers acknowledge funding from the Australian National Health and Medical Research Council (grants 241944, 389875, 389891, 389892, 389938, 442915, 442981, 496739, 496688 and 552485), the and the National Institutes of Health (grants AA07535, AA10248, AA014041, AA13320, AA13321, AA13326 and DA12854).
 We thank Anthony Caracella and Lisa Bowdler for technical assistance with the micro-array hybridisations.
 
-The CHDWB study funding support from the Georgia Institute of Technology Research Foundation.
+% The CHDWB study funding support from the Georgia Institute of Technology Research Foundation.
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript
 
-The Fehrmann study was supported by grants from the Celiac Disease Consortium (an innovative cluster approved by the Netherlands Genomics Initiative and partly funded by the Dutch Government (grant BSIK03009), the Netherlands Organization for Scientific Research (NWO-VICI grant 918.66.620, NWO-VENI grant 916.10.135 to L.F.), the Dutch Digestive Disease Foundation (MLDS WO11-30), and a Horizon Breakthrough grant from the Netherlands Genomics Initiative (grant 92519031 to L.F.).
+% The Fehrmann study was supported by grants from the Celiac Disease Consortium (an innovative cluster approved by the Netherlands Genomics Initiative and partly funded by the Dutch Government (grant BSIK03009), the Netherlands Organization for Scientific Research (NWO-VICI grant 918.66.620, NWO-VENI grant 916.10.135 to L.F.), the Dutch Digestive Disease Foundation (MLDS WO11-30), and a Horizon Breakthrough grant from the Netherlands Genomics Initiative (grant 92519031 to L.F.).
 This project was supported by the Prinses Beatrix Fonds, VSB fonds, H.
@@
 
-The EGCUT study received targeted financing from Estonian Government SF0180142s08, Center of Excellence in Genomics (EXCEGEN) and University of Tartu (SP1GVARENG).
+% The EGCUT study received targeted financing from Estonian Government SF0180142s08, Center of Excellence in Genomics (EXCEGEN) and University of Tartu (SP1GVARENG).
 We acknowledge EGCUT technical personnel, especially Mr V.
@@
 It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.
-An interactive version of this graph can be found here: }
+An interactive version of this graph can be found here: \url{http://kn3in.github.io/detecting_epi/}}
 	\label{fig:fireworks}

### 2013-12-10 docs/manuscript/methods.tex
still have a few words to cut

@@
 
+\subsection{Limitations of this type of comparison}
+
+Though it is useful to compare the relative variances of epistatic and additive effects, it must be stressed that our results here are approximations that are very limited by the study design.
+We estimate that additive effects explain approximately 10 times more variance than epistatic effects, but this could be an overestimate or an underestimate due to a number of different caveats.
+Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used.
+Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
+Thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance.
+And forthly, the extent of winner's curse in estimation of effect sizes may differ between the two studies.
+
 \subsection{Pedigree estimates}

### 2013-12-10 docs/manuscript/shorter.tex
still have a few words to cut

@@
 
-% Epistasis is the phenomenon whereby a polymorphism's effect on a trait depends on other polymorphisms present in the genome.
-The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics.
-Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and examples have been reported in other species \cite{Carlborg2006}, few convincing examples with independent replication exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
-Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}.
-Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
-In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
-Replication of these interactions in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$.
-There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
-Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs.
-For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction.
-This study presents the first evidence for multiple instances of segregating common polymorphisms interacting to influence human traits.
-
 \end{abstract}
@@
 For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects.
-Thus, when combined with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
+Thus, with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
 There are two simple ways to overcome this problem.
@@
 We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/501$, Table \ref{tab:bonferroni}).
-To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods).
+To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A1 \times A2$, $A1 \times D2$, $D1 \times A2$, $D1 \times D2$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods).
 Sign concordance between the discovery and both replication datasets was observed in 22 out of the 30 significantly replicated interactions (expected value $= 7.5$ under the null hypothesis of no interactions, $p = 3.76 \times 10^{-8}$).
@@
 % \subsection{Patterns of epistasis}
-Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows that this is unlikely to be the most effective strategy for its detection.
+Though seldom the focus of association studies, SNPs with known main effects are often tested for $A \times A$ genetic interactions \cite{Cordell2009}, but our analysis suggests this is unlikely to be the best strategy for its detection.
 The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \cite{Powell2013} (439 out of 501, Methods).
 Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects.
-Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst additive $\times$ additive, additive $\times$ dominance, dominance $\times$ additive and dominance $\times$ dominance at the discovery stage ($p = 0.22$ for departure from expectation).
-This is not surprising because the patterns of epistasis used for statistical decomposition (\emph{i.e.} $A \times A$, $A \times D$, $D \times A$, $D \times D$) are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.
+Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst $A \times A$, $A \times D$, $D \times A$ and $D \times D$ at the discovery stage ($p = 0.22$ for departure from expectation).
+This is not surprising because these patterns of epistasis used for statistical decomposition are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.
 
@@
 % \subsection{Contribution of epistatic variance}
-Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question.
-In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance.
+Quantifying the importance of epistasis in complex traits in humans remains an open question.
+Here we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance.
 Taking results from our previously published eQTL \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater.
@@
 By contrast, the estimated epistatic variance from the interacting SNPs detected in this study on average explain a total of $0.22\%$ of phenotypic variance, approximately ten times lower than the estimated additive variance.
-There are several caveats to this comparison.
-Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used.
-Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
-Thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance.
-And forthly, the extent of winner's curse in estimation of effect sizes may differ between the the two studies.
+There are several caveats to this comparison which we discuss in the Methods.
 

### 2013-12-10 docs/manuscript/shorter.tex
minor changes / typo

@@
 Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power.
-For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects.
+For example, increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects.
 Thus, with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.

### 2013-12-19 docs/manuscript/methods.tex
sent revised manuscript to nature

--- 2013-09-13 docs/manuscript/methods.tex
@@
 \usepackage{lineno}
+% \linenumbers
 
@@
 After this stringent QC 7339 probes remained for 2D-eQTL mapping.
+These data are accessible through GEO Series accession number GSE53195.
 
@@
 RNA levels were quantified using the Illumina HT-12 V3.0 platform.
+These data are accessible through GEO Series accession numbers GSE20332 and GSE20142.
 
@@
 RNA levels were quantified using the Illumina HT-12 V3.0 platform.
+These data are accessible through GEO Series accession number GSE48348.
 

### 2013-12-19 docs/manuscript/shorter.tex
sent revised manuscript to nature

@@
 \usepackage{lscape}
-\usepackage{lineno}
+% \usepackage{lineno}
 \usepackage{hyperref}
-\linenumbers
+% \linenumbers
 
@@
 \author[1,2]{Allan F McRae}
-\author[2]{Jian Yang}
+\author[1]{Jian Yang}
 \author[8]{Greg Gibson}
@@
 
-\affil[1]{University of Queensland Diamantina Institute, University of Queensland, Princess Alexandra Hospital, Brisbane, Queensland, Australia}
-\affil[2]{Queensland Brain Institute, University of Queensland, Brisbane, QLD, Australia}
+\affil[1]{Queensland Brain Institute, University of Queensland, Brisbane, QLD, Australia}
+\affil[2]{University of Queensland Diamantina Institute, University of Queensland, Princess Alexandra Hospital, Brisbane, Queensland, Australia}
 \affil[3]{Department of Genetics, University Medical Center Groningen, University of Groningen, Hanzeplein 1, Groningen, the Netherlands}
@@
 A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}).
-It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistasis is difficult in practice due to LD (Methods).
+It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistatic effects in independent samples is difficult in practice due to LD (Methods).
 
@@
 
-% \subsection*{Acknowledgements}
-
-% We are grateful to the volunteers for their generous participation in these studies. We thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.
-
-% This work could not have been completed without access to high performance GPGPU compute clusters.
+\section*{Methods Summary}
+We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs.
+Each individual had gene expression levels measured in peripheral blood at 7,339 probes representing 6,158 RefSeq genes (significant expression in $\geq 90\%$ of individuals).
+SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
+We used permutation analysis to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
+All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects.
+All SNP pairs were required to have at least five data points in all nine genotype classes.
+If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained.
+Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair.
+The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood Fehrmann \cite{Fehrmann2011}, $n=1240$; EGCUT \cite{Metspalu2004}, ($n=891$).
+A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.
+
+\subsection*{Acknowledgements}
+
+We are grateful to the volunteers for their generous participation in these studies. We thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.
+
+This work could not have been completed without access to high performance GPGPU compute clusters.
 We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au).
@@
 
-% The University of Queensland group is supported by the Australian National Health and Medical Research Council (NHMRC) grants 389892, 496667, 613601, 1010374 and 1046880, the Australian Research Council (ARC) grant (DE130100691), and by National Institutes of Health (NIH) grants GM057091 and GM099568.
-
-% The QIMR researchers acknowledge funding from the Australian National Health and Medical Research Council (grants 241944, 389875, 389891, 389892, 389938, 442915, 442981, 496739, 496688 and 552485), the and the National Institutes of Health (grants AA07535, AA10248, AA014041, AA13320, AA13321, AA13326 and DA12854).
+The University of Queensland group is supported by the Australian National Health and Medical Research Council (NHMRC) grants 389892, 496667, 613601, 1010374 and 1046880, the Australian Research Council (ARC) grant (DE130100691), and by National Institutes of Health (NIH) grants GM057091 and GM099568.
+
+The QIMR researchers acknowledge funding from the Australian National Health and Medical Research Council (grants 241944, 389875, 389891, 389892, 389938, 442915, 442981, 496739, 496688 and 552485), the and the National Institutes of Health (grants AA07535, AA10248, AA014041, AA13320, AA13321, AA13326 and DA12854).
 We thank Anthony Caracella and Lisa Bowdler for technical assistance with the micro-array hybridisations.
 
-% The CHDWB study funding support from the Georgia Institute of Technology Research Foundation.
+The CHDWB study funding support from the Georgia Institute of Technology Research Foundation.
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript
 
-% The Fehrmann study was supported by grants from the Celiac Disease Consortium (an innovative cluster approved by the Netherlands Genomics Initiative and partly funded by the Dutch Government (grant BSIK03009), the Netherlands Organization for Scientific Research (NWO-VICI grant 918.66.620, NWO-VENI grant 916.10.135 to L.F.), the Dutch Digestive Disease Foundation (MLDS WO11-30), and a Horizon Breakthrough grant from the Netherlands Genomics Initiative (grant 92519031 to L.F.).
+The Fehrmann study was supported by grants from the Celiac Disease Consortium (an innovative cluster approved by the Netherlands Genomics Initiative and partly funded by the Dutch Government (grant BSIK03009), the Netherlands Organization for Scientific Research (NWO-VICI grant 918.66.620, NWO-VENI grant 916.10.135 to L.F.), the Dutch Digestive Disease Foundation (MLDS WO11-30), and a Horizon Breakthrough grant from the Netherlands Genomics Initiative (grant 92519031 to L.F.).
 This project was supported by the Prinses Beatrix Fonds, VSB fonds, H.
@@
 
-% The EGCUT study received targeted financing from Estonian Government SF0180142s08, Center of Excellence in Genomics (EXCEGEN) and University of Tartu (SP1GVARENG).
+The EGCUT study received targeted financing from Estonian Government SF0180142s08, Center of Excellence in Genomics (EXCEGEN) and University of Tartu (SP1GVARENG).
 We acknowledge EGCUT technical personnel, especially Mr V.
@@
 Data analyzes were carried out in part in the High Performance Computing Center of University of Tartu.
+
+\subsection*{Author contributions}
+
+G.H., J.E.P., P.M.V., and G.W.M.
+conceived and designed the study.
+G.H., J.E.P., K.S., H-J.W., and J.Y.
+performed the analysis.
+T.E.
+and A.M.
+provided the EGCUT data.
+A.K.H., A.F.M., G.W.M., N.G.M., and J.E.P.
+provided the BSGS data.
+H-J.W.
+and L.F.
+provided the Fehrmann data.
+G.H.
+and J.E.P.
+wrote the manuscript with the participation of all authors.
+
+\subsection*{Author information}
+
+The authors declare no financial competing interests.
 
@@
 	\centering
-	\includegraphics[width=2.5in]{qqMeta.pdf}
+	% \includegraphics[width=2.5in]{qqMeta.pdf}
 	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets} The top panel shows all 434 discovery SNPs that were tested for interactions.
@@
 	\centering
-	\includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
+	% \includegraphics[width=5in]{pale_gray_graph_of_interactions_2_lists_GrayEdges}
 	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes).
@@
 
+\setcounter{page}{1}
 \section*{Supplementary Figures}
@@
 	\includegraphics[width=5in]{qqMetaNonsig}
-	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets, excluding the 30 points significant at the Bonferroni level} The right panel (Null) shows the interaction $p$-values from a meta analysis across two independent datasets on 434 randomly drawn SNP pairs.
+	\caption{\textbf{Q-Q plots of interaction $p$-values from replication datasets, excluding the 30 points significant at the Bonferroni level} The right panel (Null) shows the interaction $p$-values from a meta analysis across two independent datasets on 434 SNP pairs where one SNP has a marginal effect.
 The left panel (Empirical) shows the interaction $p$-values from the 404 putative interactions that were not significant at the Bonferroni correction threshold.

### 2013-12-19 docs/manuscript/manuscript.tex
updated manuscript
submitted revised manuscript
sent manuscript to dave evans

@@
 \usepackage{lscape}
-\usepackage{lineno}
-\linenumbers
+% \usepackage{lineno}
+\usepackage{hyperref}
+% \linenumbers
 
@@
 
-Epistasis is the phenomenon whereby a polymorphism's effect on a trait depends on other polymorphisms present in the genome.
+Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome.
 The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics.
-Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and examples have been reported in other species \cite{Carlborg2006}, few convincing examples with independent replication exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
-Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}.
-Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
-In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$).
+Though often demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been reported in other species \cite{Carlborg2006}, few examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
+Its absence from empirical findings may simply be due to low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical and computational issues \cite{Cordell2009}.
+Here we show that, using advanced computation \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
+In a cohort of 846 individuals with 7339 gene expression levels measured in peripheral blood, we found 501 significant pairwise interactions between common SNPs influencing the expression of 238 genes ($p < 2.91 \times 10^{-16}$).
 Replication of these interactions in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$.
-There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
+Forty-four of the genetic interactions are located within 2Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
 Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs.
@@
 Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power.
-For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects.
-Thus, when combined with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
+For example, increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects.
+Thus, with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
 There are two simple ways to overcome this problem.
@@
 A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}).
-
-It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistasis is difficult in practice.
-For example, LD between causal variants and observed markers plays an important role.
-Not only is the dependence on LD much greater for epistatic effects than for additive effects (Supplementary Figure \ref{fig:ld_sampling_mean}), but when estimating epistatic variance it is more sensitive to changes in LD between observed SNPs and causal variants between independent samples when compared to additive effects (Supplementary Figure \ref{fig:ld_sampling_sd}).
-This has a direct effect on statistical power for replication.
-The sampling variance of LD $r$ leads to the ascertainment of marker associations with higher sample $r$ in the discovery stage in comparison to the replication stage.
-However, the average decrease in $\hat{r}^{x}$ in replication samples becomes larger as $x$ increases (Methods, Supplementary Figure \ref{fig:ld_reduction}).
-For example, the decrease in $\hat{r}^8$ (which is proportional to the power of deteting $D \times D$ effects), is on average three fold greater than the decrease in $\hat{r}^2$ (which is proportional to the power of detecting additive effects).
+It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistatic effects in independent samples is difficult in practice due to LD (Methods).
 
 % \subsection{Patterns of epistasis}
-Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows that this is unlikely to be the most effective strategy for its detection.
+Though seldom the focus of association studies, SNPs with known main effects are often tested for $A \times A$ genetic interactions \cite{Cordell2009}, but our analysis suggests this is unlikely to be the best strategy for its detection.
 The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \cite{Powell2013} (439 out of 501, Methods).
 Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects.
-Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst additive $\times$ additive, additive $\times$ dominance, dominance $\times$ additive and dominance $\times$ dominance at the discovery stage ($p = 0.22$ for departure from expectation).
-This is not surprising because the patterns of epistasis used for statistical decomposition (\emph{i.e.} $A \times A$, $A \times D$, $D \times A$, $D \times D$) are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.
+Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst $A \times A$, $A \times D$, $D \times A$ and $D \times D$ at the discovery stage ($p = 0.22$ for departure from expectation).
+This is not surprising because these patterns of epistasis used for statistical decomposition are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.
 
@@
 We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 5Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
-Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Osborne2004, Rieder2012}.
+Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Rieder2012}.
 
 % \subsection{Contribution of epistatic variance}
-Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question.
-In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance.
+Quantifying the importance of epistasis in complex traits in humans remains an open question.
+Here we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance.
 Taking results from our previously published eQTL \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater.
@@
 In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}.
-Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components.
-Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that estimated additive variance is overall a larger component than estimated epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}.
 Taking all additive effects detected in Powell \emph{et al} (2012) that have additive variance explaining $2.1\%$ or greater of phenotypic variance, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$.
 By contrast, the estimated epistatic variance from the interacting SNPs detected in this study on average explain a total of $0.22\%$ of phenotypic variance, approximately ten times lower than the estimated additive variance.
-There are several caveats to this comparison.
-Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used.
-Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
-Thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance.
-This is due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.
-And forthly, the extent of winner's curse in estimation of effect sizes may differ between the the two studies.
+There are several caveats to this comparison which we discuss in the Methods.
 
@@
 
-\subsection*{Methods Summary}
+\section*{Methods Summary}
 We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs.
-Each individual had gene expression levels measured in peripheral blood at 47,323 probes.
-Only the probes that passed quality control and had significant expression in $\geq 90\%$ of individuals were used in the analysis (7,339 probes representing 6,158 RefSeq genes).
-Recent hardware and software \cite{Hemani2011} advances that use graphics processing units (GPUs) made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis.
-We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
+Each individual had gene expression levels measured in peripheral blood at 7,339 probes representing 6,158 RefSeq genes (significant expression in $\geq 90\%$ of individuals).
 SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms.
-Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}.
-Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone.
+We used permutation analysis to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER).
 All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects.
@@
 If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained.
-Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs.
-The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$).
-Of these, 434 passed filtering in both replication datasets.
+Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair.
+The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood Fehrmann \cite{Fehrmann2011}, $n=1240$; EGCUT \cite{Metspalu2004}, ($n=891$).
 A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

### 2014-01-07 docs/manuscript/manuscript.tex
references updated, manuscript has now been accepted. woo

@@
 \documentclass{article}
-\usepackage[superscript,biblabel]{cite}
+\usepackage{amsmath}
+% \usepackage[superscript,biblabel]{cite}
 \usepackage{graphicx}
@@
 % \usepackage{lineno}
+\usepackage{verbatim}
 \usepackage{hyperref}
 % \linenumbers
+% \usepackage{bibunits}
+% \usepackage{splitbib}
+\usepackage[style=nature, sorting=none]{biblatex}
+\addbibresource{refs.bib}
+
+\DeclareBibliographyCategory{Category1}
+\addtocategory{Category1}{Hill2008a}
+\addtocategory{Category1}{Costanzo2010}
+\addtocategory{Category1}{Carlborg2006}
+\addtocategory{Category1}{Strange2010}
+\addtocategory{Category1}{Cordell2009}
+\addtocategory{Category1}{Hemani2011}
+\addtocategory{Category1}{Lieberman-Aiden2009}
+\addtocategory{Category1}{Visscher2012}
+\addtocategory{Category1}{Carlborg2004}
+\addtocategory{Category1}{Weinreich2006}
+\addtocategory{Category1}{Breen2012}
+\addtocategory{Category1}{Weir2008}
+\addtocategory{Category1}{Marchini2005}
+\addtocategory{Category1}{LangoAllen2010}
+\addtocategory{Category1}{Schadt2003}
+\addtocategory{Category1}{Powell2013}
+\addtocategory{Category1}{Powell2012}
+\addtocategory{Category1}{Fehrmann2011}
+\addtocategory{Category1}{Metspalu2004}
+\addtocategory{Category1}{Preininger2013}
+\addtocategory{Category1}{Cockerham1954}
+\addtocategory{Category1}{Ho2004}
+\addtocategory{Category1}{Trynka2013}
+\addtocategory{Category1}{Hoffman2012}
+\addtocategory{Category1}{Lan2012}
+\addtocategory{Category1}{Rieder2012}
+\addtocategory{Category1}{Crow2010}
+\addtocategory{Category1}{Hemani2013}
+\addtocategory{Category1}{Evans2011}
+\addtocategory{Category1}{Bloom2013}
+
+\DeclareBibliographyCategory{Category2}
+\addtocategory{Category2}{Medland2009}
+\addtocategory{Category2}{Aulchenko2007}
+\addtocategory{Category2}{Yang2011}
+\addtocategory{Category2}{Westra2011}
+\addtocategory{Category2}{Williams1976}
+\addtocategory{Category2}{Alvarez-Castro2008}
+\addtocategory{Category2}{Koch2007}
+\addtocategory{Category2}{Rietveld2013}
+\addtocategory{Category2}{Stormo2000}
+\addtocategory{Category2}{HoSui2005}
 
@@
 \graphicspath{{../../analysis/images/}}
-\bibliographystyle{naturemag}
+% \bibliographystyle{naturemag}
 
@@
 Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome.
-The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics.
-Though often demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been reported in other species \cite{Carlborg2006}, few examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}.
-Its absence from empirical findings may simply be due to low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical and computational issues \cite{Cordell2009}.
-Here we show that, using advanced computation \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
+The extent to which epistasis influences complex traits \supercite{Carlborg2004} and contributes to their variation \supercite{Hill2008a,Crow2010} is a fundamental question in evolution and human genetics.
+Though often demonstrated in artificial gene manipulation studies in model organisms \supercite{Costanzo2010,Bloom2013}, and some examples have been reported in other species \supercite{Carlborg2006}, few examples exist for epistasis amongst natural polymorphisms in human traits \supercite{Strange2010,Evans2011}.
+Its absence from empirical findings may simply be due to low incidence in the genetic control of complex traits \supercite{Hill2008a,Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical and computational issues \supercite{Cordell2009}.
+Here we show that, using advanced computation \supercite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs).
 In a cohort of 846 individuals with 7339 gene expression levels measured in peripheral blood, we found 501 significant pairwise interactions between common SNPs influencing the expression of 238 genes ($p < 2.91 \times 10^{-16}$).
-Replication of these interactions in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$.
-Forty-four of the genetic interactions are located within 2Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
+Replication of these interactions in two independent data sets \supercite{Metspalu2004,Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$.
+Forty-four of the genetic interactions are located within 2Mb of regions of known physical chromosome interactions \supercite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$).
 Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs.
@@
 In the genetic analysis of complex traits it is usual for SNP effects to be estimated using an additive model where they are assumed to contribute independently and cumulatively to the mean of a trait.
-This framework has been successful in identifying thousands of associations \cite{Visscher2012}.
-But to date, though its contribution to phenotypic variance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}, there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}.
-Beyond the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}.
+This framework has been successful in identifying thousands of associations \supercite{Visscher2012}.
+But to date, though its contribution to phenotypic variance is frequently the subject of debate \supercite{Carlborg2004,Hill2008a,Crow2010}, there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \supercite{Strange2010, Evans2011}.
+Beyond the prism of human association studies there is evidence for epistasis, not only at the molecular scale from artificially induced mutations \supercite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \supercite{Weinreich2006} and speciation \supercite{Breen2012}.
 
 Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power.
-For example, increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects.
-Thus, with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly.
+For example, increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \supercite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \supercite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \supercite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects.
+Thus, with small genetic effect sizes, as is expected in most complex traits of interest \supercite{Visscher2012}, the power to detect epistasis diminishes rapidly.
 There are two simple ways to overcome this problem.
-One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants.
+One is by using extremely large sample sizes \supercite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants.
 Because our focus was to ascertain the extent to which instances of epistasis arises from natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels.
-Transcription levels can be measured for thousands of genes and like most complex diseases, these expression traits are typically heritable \cite{Schadt2003}.
-But unlike complex diseases, genetic associations with gene expression commonly have very large effect sizes that explain large proportions of the genetic variance\cite{Powell2013}, making them good candidates to search for epistasis, should it exist.
+Transcription levels can be measured for thousands of genes and like most complex diseases, these expression traits are typically heritable \supercite{Schadt2003}.
+But unlike complex diseases, genetic associations with gene expression commonly have very large effect sizes that explain large proportions of the genetic variance\supercite{Powell2013}, making them good candidates to search for epistasis, should it exist.
 
 % \subsection{Initial search in discovery set and replication}
-In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions.
+In our discovery dataset (Brisbane Systems Genetics Study, BSGS \supercite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions.
 First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood ($1.03 \times 10^{15}$ statistical tests, family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods).
@@
 However it remains possible that unexplained technical artifacts may have led to the significant discovery interactions.
-Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication.
+Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \supercite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \supercite{Metspalu2004}, in which we saw convincing evidence for replication.
 We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/501$, Table \ref{tab:bonferroni}).
@@
 The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent datasets.
-A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}).
+A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \supercite{Preininger2013}, but only 20 of the SNP pairs passed filtering because the sample size was small ($n=139$), and likely due to insufficient power we found no evidence for replication (Supplementary Figure \ref{fig:qqgg}).
 It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistatic effects in independent samples is difficult in practice due to LD (Methods).
@@
 % \subsection{Patterns of epistasis}
-Though seldom the focus of association studies, SNPs with known main effects are often tested for $A \times A$ genetic interactions \cite{Cordell2009}, but our analysis suggests this is unlikely to be the best strategy for its detection.
-The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \cite{Powell2013} (439 out of 501, Methods).
+Though seldom the focus of association studies, SNPs with known main effects are often tested for $A \times A$ genetic interactions \supercite{Cordell2009}, but our analysis suggests this is unlikely to be the best strategy for its detection.
+The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \supercite{Powell2013} (439 out of 501, Methods).
 Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects.
 Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst $A \times A$, $A \times D$, $D \times A$ and $D \times D$ at the discovery stage ($p = 0.22$ for departure from expectation).
-This is not surprising because these patterns of epistasis used for statistical decomposition are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.
+This is not surprising because these patterns of epistasis used for statistical decomposition are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \supercite{Cockerham1954}.
 
@@
 We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP.
-For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an effect (Supplementary Figure \ref{fig:MBNL1_3D}).
+For example, MBNL1 (involved in RNA modification and regulation of splicing \supercite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an effect (Supplementary Figure \ref{fig:MBNL1_3D}).
 Each of these interactions has evidence for replication in at least one dataset and six are significantly replicated at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).
 We see similar epistatic networks involving multiple (eight or more) \emph{trans}-acting SNPs for other gene expresson levels too, for example TMEM149 (Supplementary Figure \ref{fig:TMEM149_3D}), NAPRT1 (Supplementary Figure \ref{fig:NAPRT1_3D}), TRAPPC5 (Supplementary Figure \ref{fig:TRAPPC5_3D}), and CAST (Supplementary Figure \ref{fig:CAST_3D}).
-We observed that from pedigree analysis these five gene expression phenotypes had non-additive variance component estimates within the 95th percentile of the 17,994 gene expression phenotypes that were analysed previously \cite{Powell2013} (Supplementary Table \ref{tab:vc}, Methods).
+We observed that from pedigree analysis these five gene expression phenotypes had non-additive variance component estimates within the 95th percentile of the 17,994 gene expression phenotypes that were analysed previously \supercite{Powell2013} (Supplementary Table \ref{tab:vc}, Methods).
 
@@
 In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment (Methods).
-We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,tri-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}).
+We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,tri-methylation (H3K4me3) chromatin marks, in 34 cell types \supercite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}).
 There was significant enrichment for \emph{cis}-acting SNPs in haematopoietic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing).
@@
 This difference between \emph{cis} and \emph{trans} SNPs suggests different roles in epistatic interactions where tissue specificity is provided by the \emph{cis} SNPs.
-There is also enrichment for \emph{cis}-SNPs to be localised in regions with regulatory genomic features as measured by chromatin states \cite{Hoffman2012} (Supplementary Figure \ref{fig:snplocations}).
+There is also enrichment for \emph{cis}-SNPs to be localised in regions with regulatory genomic features as measured by chromatin states \supercite{Hoffman2012} (Supplementary Figure \ref{fig:snplocations}).
 
 We also demonstrate physical organisation of interacting loci within the cell, suggesting a mechanism by which biological function can lead to epistatic genetic variance.
-It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}.
-We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 5Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
-Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \cite{Rieder2012}.
+It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \supercite{Lieberman-Aiden2009}.
+We cross-referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \supercite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 5Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}).
+Interaction of distant loci may occur through physical proximity in transcriptional factories that organise across different chromosome regions and can regulate transcription of related genes \supercite{Rieder2012}.
 
@@
 Here we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance.
-Taking results from our previously published eQTL \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater.
+Taking results from our previously published eQTL \supercite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater.
 Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.
 
-In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}.
+In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \supercite{Hill2008a}.
 Taking all additive effects detected in Powell \emph{et al} (2012) that have additive variance explaining $2.1\%$ or greater of phenotypic variance, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$.
@@
 \section*{Methods Summary}
-We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs.
+We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \supercite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs.
 Each individual had gene expression levels measured in peripheral blood at 7,339 probes representing 6,158 RefSeq genes (significant expression in $\geq 90\%$ of individuals).
@@
 Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair.
-The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood Fehrmann \cite{Fehrmann2011}, $n=1240$; EGCUT \cite{Metspalu2004}, ($n=891$).
+The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood Fehrmann \supercite{Fehrmann2011}, $n=1240$; EGCUT \supercite{Metspalu2004}, ($n=891$).
 A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.
@@
 provided the BSGS data.
+G.G.
+provided the CHDWB data.
 H-J.W.
@@
 
-\bibliography{refs}
+% \bibliography{refs}{References 1}
+\printbibliography[category=Category1]
+\clearpage
+
+\section*{Online methods}
+
+\section{Discovery data}
+
+\subsection{Data description}
+
+The Brisbane Systems Genetics Study (BSGS) comprises 846 individuals of European descent from 274 independent families \supercite{Powell2012}.
+DNA samples from each individual were genotyped on the Illumina 610-Quad Beadchip by the Scientific Services Division at deCODE Genetics Iceland.
+Full details of genotyping procedures are given in Medland et al.
+\supercite{Medland2009} Standard quality control (QC) filters were applied and the remaining 528,509 autosomal SNPs were carried forward for further analysis.
+
+Gene expression profiles were generated from peripheral blood collected with PAXgene TM tubes (QIAGEN, Valencia, CA) using Illumina HT12-v4.0 bead arrays.
+The Illumina HT-12 v4.0 chip contains 47,323 probes, although some probes are not assigned to RefSeq genes.
+We removed any probes that did not match the following criteria: contained a SNP within the probe sequence with MAF $ > 0.05$ within 1000 genomes data; did not map to a listed RefSeq gene; were not significantly expressed (based on a detection $p$-value $< 0.05$) in at least 90\% of samples.
+After this stringent QC 7339 probes remained for 2D-eQTL mapping.
+These data are accessible through GEO Series accession number GSE53195.
+
+\subsection{Normalisation}
+
+Gene expression profiles were normalised and adjusted for batch and polygenic effects.
+Profiles were first adjusted for raw background expression in each sample.
+Expression levels were then adjusted using quantile and $\log_2$ transformation to standardise distributions between samples.
+Batch and polygenic effects were adjusted using the linear model
+
+\begin{equation}
+y = \mu + \beta_{1}c + \beta_{2}p + \beta_{3}s + \beta_{4}a + g + e
+\label{eq:lm}
+\end{equation}
+where $\mu$ is the population mean expression levels, $c$, $p$, $s$ and $a$ are vectors of chip, chip position, sex and generation respectively, fitted as fixed effects; and $g$ is a random additive polygenic effect with a variance covariance matrix
+\begin{equation}
+G_{jk} = \left \{ 
+\begin{array}{ll}
+\sigma _a ^2             &          j = k \\ 
+2\phi _{jk} \sigma _a ^2& 			j \neq k \\
+\end{array} \right.
+\end{equation}
+The parameter $\sigma_a^2$ is the variance component for additive background genetic.
+Here, we are using family based pedigree information rather than SNP based IBD to account for relationships between individuals and so $\phi _{jk}$ is the kinship coefficient between individuals $j$ and $k$.
+The residual, $e$, from equation \ref{eq:lm} is assumed to follow a multivariate normal distribution with a mean of zero.
+Residuals were normalised by rank transformation and used as the adjusted phenotype for the pairwise epistasis scan to remove any skewness and avoid results being driven by outliers.
+The GenABEL package for R was used to perform the normalisation \supercite{Aulchenko2007}.
+
+\section{Exhaustive 2D-eQTL analysis}
+
+\subsection{Two stage search}
+
+We used epiGPU \supercite{Hemani2011} software to perform an exhaustive scan for pairwise interactions, such that each SNP is tested against all other SNPs for statistical association with the expression values for each of the 7339 probes.
+This uses the massively parallel computational architecture of graphical processing units (GPUs) to speed up the exhaustive search.
+For each SNP pair there are 9 possible genotype classes.
+We treat each genotype class as a fixed effect and fit an 8 \emph{d.f.} $F$-test to test the following hypotheses:
+
+\begin{equation}
+H _0 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 = 0; 
+\end{equation}
+
+\begin{equation}
+H _1 : \sum _{i=1} ^3 \sum _{j=1} ^3 (\bar x _{ij} - \mu) ^2 > 0; 
+\label{eq:8df}
+\end{equation}
+where $\mu$ is the mean expression level and $x _{ij}$ is the pairwise genotype class mean for genotype $i$ at SNP 1 and genotype $j$ at SNP 2.
+This type of test does not parameterize for specific types of epistasis, rather it tests for the joint genetic effects at two loci.
+This has been demonstrated to be statistically more efficient when searching for a wide range of epistatic patterns, although will also include any marginal effects of SNPs which must be dealt with post-hoc \supercite{Hemani2013}.
+
+\subsubsection{Stage 1}
+
+The complete exhaustive scan for 7339 probes comprises $1.03 \times 10^{15}$ $F$-tests.
+We used permutation analysis to estimate an appropriate significance threshold for the study.
+To do this we performed a further 1600 exhaustive 2D scans on permuted phenotypes to generate a null distribution of the extreme $p$-values expected to be obtained from this number of multiple tests given the correlation structure between the SNPs.
+We took the most extreme $p$-value from each of the 1600 scans and set the 5\% FWER to be the 95\% most extreme of these $p$-values, $T_{*} = 2.13 \times 10^{-12}$.
+The effective number of tests in one 2D scan being performed is therefore $N_{*} = 0.05 / T_{*} \approx 2.33 \times 10^{10}$.
+To correct for the testing of multiple traits we established an experiment wide threshold of $T_{e} = 0.05 / (N_{*} \times 7339) = 2.91 \times 10^{-16}$.
+This is likely to be conservative as it assumes independence between probes.
+
+\paragraph{Filtering} We used two approaches to filter SNPs from stage 1 to be tested for significant interaction effects in stage 2.
+
+\paragraph{Filter 1} After keeping SNP pairs that surpassed the $2.91 \times 10^{-16}$ threshold in stage 1 only SNP pairs with at least 5 data points in all 9 genotype classes were kept.
+We then calculated the LD between interacting SNPs (amongst unrelated individuals within the discovery sample and also from 1000 genomes data) and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition.
+If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained.
+At this stage 6404 filtered SNP pairs remained.
+
+\paragraph{Filter 2} We also performed a second filtering screen applied to the list of SNP pairs from stage 1 that was identical to filter 1 but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) were removed \supercite{Powell2013}, creating a second set of 4751 unique filtered SNP pairs.
+
+\subsubsection{Stage 2}
+
+To ensure that interacting SNPs were driven by epistasis and not marginal effects we performed a nested ANOVA on each pair in the filtered set to test if the interaction terms were significant.
+We did this by contrasting the full genetic model (8 \emph{d.f.}) against the reduced marginal effects model which included the additive and dominance terms at both SNPs (4 \emph{d.f.}).
+Thus, a 4 \emph{d.f.} $F$-test was performed on the residual genetic variation, representing the contribution of epistatic variance.
+Significance of epistasis was determined using a Bonferroni threshold of $0.05 / (6404+4751) = 4.48 \times 10^{-6}$.
+This resulted in 406 and 95 SNP pairs with significant interaction terms from filters 1 and 2, respectively.
+
+\subsection{Type 1 error rate}
+
+Using a Bonferroni correction of $0.05$ in the second stage of the two stage discovery scan implies a type 1 error rate of $\alpha = 0.05$.
+However, this could be underestimated because the number tests performed in the second stage depends on the number of tests in the first stage, and this depends on statistical power and model choice.
+We performed simulations to estimate the type 1 error rate of this study design.
+
+We assumed a null model where there was one true additive effect and 7 other terms with no effect.
+To simulate a test statistic we simulated 8 z-scores, $z_1 \sim N(\sqrt{NCP}, 1)$ and $z_{2..8} \sim N(0,1)$.
+Thus $z_{full} = \sum^8_{i=1}{z_i} \sim \chi^{2}_{8}$ (representing the 8 d.f.
+test) and $z_{int} = \sum^8_{i=5}{z_i} \sim \chi^{2}_{4}$ (representing the 4 d.f.
+test where the null hypothesis of no epistasis is true).
+For a particular value of $NCP$ we simulated 100,000 $z$ values, and calculated the $p_{full}$-value for the $z_{full}$ test statistic.
+The $n_{int}$ test statistics with $p_{full} < 2.31 \times 10^{-16}$ were kept for the second stage, where the type 1 error rate of stage 2 was calculated as the proportion of $p_{int} < 0.05 / n_{int}$.
+The power at stage 1 was calculated as $n_{int} / 100,000$.
+This procedure was performed for a range of $NCP$ parameters that represented power ranging from $\sim 0$ to $\sim 1$.
+
+\subsection{Population stratification}
+
+We ruled out population stratification as a possible cause of inflated test statistics.
+To test for cryptic relatedness driving the interaction terms we tested for increased LD among the SNPs \supercite{Yang2011}.
+We calculated the mean of the off-diagonal elements of the correlation matrix of all unique SNPs from the 501 interactions (731 SNPs) using only unrelated individuals, $\bar{r^2} = 0.0039$.
+This is not significantly different from the null hypothesis of zero (sampling error $= 1/n_{\textrm{unrelated}} = 0.0039$).
+
+\subsection{Probe mapping}
+
+To avoid possibility that epistatic signals might arise due to expression probes hybridising in multiple locations we verified that probe sequences for genes with significant interactions mapped to only a single location.
+As an initial verification we performed a BLAST search of the full probe sequence against 1000 genomes phase 1 version 3 human genome reference and ensured that only one genomic location aligned significantly ($p < 0.05$).
+As a second step, to mitigate the possibility of weak hybridisation elsewhere in the genome we divided the probe sequence into three sections (1-25bp, 13-37bp, 26-50bp) and performed a BLAST search of these probe sequence fragments.
+No probe sequemces or probe sequence fragments mapped to positions other than the single expected genomic target ($p < 0.05$).
+
+\section{Replication}
+
+\subsection{Data description}
+
+We attempted replication of the 501 significant interactions from the discovery set using three independent cohorts; Fehrmann, EGCUT, and CHDWB.
+It was required that LD $r^2 < 0.1$ and $D'^{2} < 0.1$ between interacting SNPs (as measured in the replication sample directly), and all nine genotype classes had at least 5 individuals present in order to proceed with statistical testing for replication in both datasets.
+We also excluded any putative SNPs that had discordant allele frequencies in any of the datasets.
+Details of the cohorts are as follows.
+
+\paragraph{Fehrmann: $n=1240$}
+The Fehrmann dataset \supercite{Fehrmann2011} consists of peripheral blood samples of 1240 unrelated individuals from the United Kingdom and the Netherlands.
+Some of these individuals are patients, while others are healthy controls.
+Individuals were genotyped using the Illumina HumanHap300, Illumina HumanHap370CNV, and Illumina 610 Quad platforms.
+RNA levels were quantified using the Illumina HT-12 V3.0 platform.
+These data are accessible through GEO Series accession numbers GSE20332 and GSE20142.
+
+\paragraph{EGCUT: $n=891$}
+The Estonian Genome Center of the University of Tartu (EGCUT) study \supercite{Metspalu2004} consists of peripheral blood samples of 891 unrelated individuals from Estonia.
+They were genotyped using the Illumina HumanHap370CNV platform.
+RNA levels were quantified using the Illumina HT-12 V3.0 platform.
+These data are accessible through GEO Series accession number GSE48348.
+
+\paragraph{CDHWB: $n=139$}
+The Center for Health Discovery and Well Being (CDHWB) Study \supercite{Preininger2013} is a population based cohort consisting of 139 individuals of European descent collected in Atlanta USA.
+Gene expression profiles were generated with Illumina HT-12 V3.0 arrays from peripheral blood collected from Tempus tubes that preserve RNA.
+Whole genome genotypes were measured using Illumina OmniQuad arrays.
+Due to the small sample size, most SNP pairs did not pass filtering in this dataset (20 SNP pairs remained) and so we have excluded it from the rest of the analysis.
+
+\subsection{Meta Analysis}
+
+The 4 \emph{d.f.} interaction $p$-values for each independent replication dataset were calculated using the same statistical test as was performed in the discovery dataset.
+We then took the interaction $p$-values from EGCUT and Fehrmann and calculated a joint $p$-value using Fisher's method of combining $p$-values for a meta analysis as $-2 \ln p_{1} - 2 \ln p_{2} \sim \chi^{2}_{4 d.f.}$.
+As in the discovery analysis, all gene expression levels were normalised using rank transformation to avoid skew or outliers in the distribution \supercite{Westra2011}.
+
+\subsection{Concordance of direction of effects}
+
+We used four methods to calculate the concordance of the direction of effects between the discovery and replication datasets.
+
+\paragraph{Test 1} Is the most significant epistatic effect in the discovery set in the same direction as the same epistatic effect in the replication sets?
+We decomposed the genetic variance into 8 orthogonal effects, four of which are epistatic ($A \times A$, $A \times D$, $D \times A$, $D \times D$).
+The sign of the epistatic effect that had the largest variance in the discovery was recorded, and then was compared to the same epistatic effect in the two replication datasets (regardless of whether or not the same epistatic effect was the largest in the replication datasets).
+The probability of the sign being the same in one dataset is $1/2$.
+The probability of the sign being the same in two is $1/4$.
+
+\paragraph{Test 2} Is the most significant epistatic effect in the discovery the same as the largest epistatic effect in the replication set with the sign being concordant.
+As in Test 1, but this time we required that the largest effect was the same in the discovery and the replication, and that they had the same sign (\emph{e.g.} if the largest effect in the discovery is $A \times A$, with a positive effect, then concordance is achieved if the same is true in the replication).
+The probability of one replication dataset being concordant by chance is $1/8$, and concordance in both is $1/64$.
+
+\paragraph{Test 3} Do the epistatic effects that are significant at nominal $p < 0.05$ in the discovery have the same direction of effect as in the replication?
+Here we count all the epistatic variance components in the discovery that have $p < 0.05$ (1133 amongst the 434 discovery SNP pairs, $i.e.$ each SNP pair has at least 1 and at most 4 significant epistatic variance components).
+Then we compare the direction of the effect in the replication dataset.
+The probability of the sign being the same in one dataset for any one significant effect is $1/2$.
+The probability of the sign being the same in two is $1/4$.
+
+\paragraph{Test 4} If we count how many of the 4 epistatic effects are concordant between the discovery and replication data for each interaction then is this significant from what we expect by chance?
+There can be either 0, 1, 2, 3 or 4 concordant signs at each interaction, each with expectation of $p=1/16, 4/16, 6/16, 4/16, 1/16$ under the null, respectively.
+Observed counts are multinomially distributed, and we tested if the observed proportions were statistically different from the expected proportions using an approximation of the multinomial test \supercite{Williams1976}.
+
+The probability of observing the number of concordant signs in tests 1-3 is calculated using a binomial test.
+All variance decompositions were calculated using the NOIA method \supercite{Alvarez-Castro2008}.
+
+\section{Effects of LD on detection and replication}
+
+The power to detect genetic effects, when the observed markers are in LD with the causal variants, is proportional to $r^x$.
+For additive effects $x = 2$, but for non-additive effects $x$ is larger, \emph{i.e.} $x = 4$ for dominance or $A \times A$, $x = 6$ for $A \times D$ or $D \times A$, and $x = 8$ for $D \times D$.
+Many biologically realistic GP maps may be comprised of all 8 variance components \supercite{Hemani2013}.
+
+This is important for both detection and for replication of epistasis.
+For detection, if the epistatic effect includes the $D \times D$ term then if the two causal variants are tagged by observed markers that are each in LD $r = 0.9$, then if the true variance is $V_{t}$ then the observed variance $V_{o}$ at the markers will be $0.9^{8}V_{t} = 0.43V_{t}$.
+Therefore, it is important to consider the sampling variation of $\hat{r}^x$ in a sample given some true population value of $r$.
+
+\subsection{Simulation 1}
+
+For some values of fixed population parameters, $p_{1}$ (minor allele frequency at observed marker), $q_{1}$ (minor allele frequency at causal variant), and $r$ (LD between marker and causal variant), the expected haplotype frequencies are
+
+\begin{align}
+h_{11} &= r \sqrt{p_{1}q_{1}p_{2}q_{2}} + p_{1}q_{1} \\
+h_{12} &= p_{1}q_{2} - r \sqrt{p_{1}q_{1}p_{2}q_{2}} \\
+h_{21} &= p_{2}q_{1} - r \sqrt{p_{1}q_{1}p_{2}q_{2}} \\
+h_{22} &= r \sqrt{p_{1}q_{1}p_{2}q_{2}} + p_{2}q_{2}
+\end{align}
+where $p_{2} = 1 - p_{1}$ and $q_{2} = 1 - q_{1}$.
+For a range of population parameters we randomly sampled $2n$ haplotypes where the expected haplotype frequencies were $h_{11}, h_{12}, h_{21}, h_{22}$.
+From the sample haplotype frequencies we then calculated sample estimates of $\hat{r}$ where
+
+\begin{equation}
+\hat{r} = \frac{\hat{h}_{11} - \hat{p}_{1}\hat{q}_{1}}{\sqrt{\hat{p}_{1}\hat{q}_{1}\hat{p}_{2}\hat{q}_{2}}}
+\end{equation}
+
+For each value of combination of the parameters $p_{1}, q_{1}, r, n$ 1000 simulations were performed and the sampling mean and sampling standard deviation of $\hat{r}, \hat{r}^2, \hat{r}^4, \hat{r}^6, \hat{r}^8$ were recorded.
+It was observed that sampling variance increases for increasing $x$ in $\hat{r}^x$.
+
+\subsection{Simulation 2}
+
+We assume that the discovery SNP pairs are ascertained (from a very large number of tests) have high $\hat{r}$ between observed SNPs and causal variants because otherwise power of detection would be low.
+We can hypothesis that the distribution of $\hat{r}$ in this ascertained sample will be a mixture of $r$ that is high and $r$ that is lower but with ascertained higher values from sampling.
+Therefore, we would expect those with truly high $r$ to have a higher replication rate in independent datasets, and those with ascertained high $\hat{r}$ to have lower replication because resampling is unlikely to result in the same extreme ascertainment.
+To obtain empirical estimates of $\hat{r}$ in discovery and replication datasets we conducted the following simulation.
+
+\begin{enumerate}
+\item Using 1000 genomes data (phase 1, version 3, 379 European samples) we selected the 528,509 ``markers'' used in the original discovery analysis, plus 100,000 randomly chosen ``causal variants'' (CVs) with minor allele frequence $> 0.05$.
+\item The 379 individuals were split into discovery (190) and replication (189) sets.
+\item For each CV the marker with the maximum $\hat{r}^2_{D}$ from the marker panel was recorded in the discovery set. This marker was known as the ``discovery marker'' (DM).
+\item The $\hat{r}^2_{R}$ for each CV/DM pair was then calculated in the replication set where the discovery LD was ascertained to be high, such that $\hat{r}^2_{D} > 0.9$.
+\end{enumerate}
+
+We observed that there was an average decrease in $\hat{r}^x_{R}$ relative to $\hat{r}^x_{D}$, and that this decrease was larger with increasing $x$.
+We observed that $(\hat{r}^2_{R} - \hat{r}^2_{D}) / \hat{r}^2_{D} = 0.029$ whereas $(\hat{r}^8_{R} - \hat{r}^8_{D}) / \hat{r}^8_{D} = 0.092$.
+The average drop in in replication $\hat{r}^8$ was 3 times higher than the drop in $\hat{r}^2$.
+
+\subsection{Interpretation}
+
+Simulation 1 shows that sampling variance of $r^x$ increases as $x$ increases.
+Detection of epistatis is highly dependent upon high $\hat{r}$.
+Amongst the discovery SNPs there will be a mixture of interactions where observed SNPs are either in true high LD with causal variants, or will have highly inflated sample $\hat{r}^{x}$ compared to the population $r^x$.
+Simulation 2 shows that as $x$ gets larger, the average decrease in $\hat{r}^x$ between discovery and replication becomes larger, likely to be a result of ascertained high $\hat{r}$ in the discovery and increased sampling variance with increasing $x$ in the replication.
+These results demonstrate that if all else is equal, the impact of sampling variance of $r$ alone will reduce the replication rate of epistatic effects compared to additive effects.
+
+\section{Additive and non-additive variance estimation}
+
+\subsection{Fixed effects}
+
+To compare the relative contribution to the phenotypic variance of gene expression levels between additive and epistatic effects we are constrained by the problem that non-additive variance components for a phenotype cannot be calculated directly.
+Here, we only have SNP pairs that exceed a threshold of $p < 2.91 \times 10^{-16} = T_{e}$.
+A strong conclusion cannot be made about the genome-wide variance contribution, but we can compare the variance explained by SNP effects at this threshold for additive scans and epistatic scans.
+
+In Powell \emph{et al} 2012 \supercite{Powell2012} an expression quantitative trait locus (eQTL) study was performed searching for additive effects in the same BSGS dataset as was used for the discovery here.
+Using the threshold $T_{e}$ for the additive eQTL study, 453 of the 7339 probes analysed here had at least one significant additive effect.
+Assuming that the phenotypic variance for each of the probes is normalised to 1, the total phenotypic variance of all 7339 explained by the significant additive effects was 1.73\%.
+
+Following the same procedure, at the threshold $T_{e}$ there were 238 gene expression probes with at least one significant pairwise epistatic interaction out of the 7339 tested.
+In total the proportion of the phenotypic variance explained by the epistatic effects at these SNP pairs was 0.25\%.
+
+\subsection{Limitations of this type of comparison}
+
+Though it is useful to compare the relative variances of epistatic and additive effects, it must be stressed that our results here are approximations that are very limited by the study design.
+We estimate that additive effects explain approximately 10 times more variance than epistatic effects, but this could be an overestimate or an underestimate due to a number of different caveats.
+Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used.
+Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test.
+Thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance.
+And forthly, the extent of winner's curse in estimation of effect sizes may differ between the two studies.
+
+\subsection{Pedigree estimates}
+
+The gene expression levels for MBNL1, TMEM149, NAPRT1, TRAPPC5 and CAST are influenced by large \emph{cis}-\emph{trans} epistatic networks (eight interactions or more).
+Though it is not possible to orthogonally estimate the non-additive genetic variance for non-clonal populations, an approximation of a component of non-additive variance can be estimated using pedigree information.
+The BSGS data is comprised of some related individuals and standard quantitative genetic analysis was used to calculate the additive and dominance variance components for each gene expression phenotype in Powell \emph{et al} 2013 \supercite{Powell2013}.
+The dominance effect is likely to capture additive $\times$ additive genetic variance plus some fraction of other epistatic variance components.
+We found that the aforementioned genes had dominance variance component estimates within the top 5\% of all 17,994 gene expression probes that were analysed in Powell \emph{et al} 2013.
+
+\section{Functional enrichment analysis}
+
+\subsection{Tissue specific transcriptionally active regions}
+
+We employed a recently published method (\url{http://www.broadinstitute.org/mpg/epigwas/}) \supercite{Trynka2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \supercite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions.
+The exact method used to perform this analysis has been described previously \supercite{Rietveld2013}.
+Briefly, we tested the hypothesis that the 731 SNPs were more likely to be in transcriptionally active regions (as measured by chromatin marks) than a random set of SNPs selected from the same SNP chip.
+This hypothesis was tested for 34 cell types across four broad tissue types (haematopoietic, gastrointenstinal, musculoskeletal and endocrine, and brain).
+
+\subsection{Chromosome interactions}
+
+It has been shown \supercite{Lieberman-Aiden2009} that different regions on different chromosomes or within chromosomes spatially colocalise within the cell.
+We shall refer to the colocalisation of two chromosome regions as a chromosome interaction.
+A map of pairwise chromosome interactions for K562 blood cell lines was recently produced \supercite{Lan2012}, and we hypothesised that part of the underlying biological mechanism behind some of the 501 epistatic interactions may arise from chromosome interactions.
+We found that 44 of the putative epistatic interactions were amongst SNPs that were within 5Mb of known chromosome interactions.
+This means that SNP A was no more than 2.5Mb from the focal point of the chromosome interaction on chromosome A, and SNP B was no more than 2.5Mb from the focal point on chromosome B.
+
+We performed simulations to test how extreme the observation of 44 epistatic interactions overlapping with chromosome interactions is compared to chance.
+Chromosome interactions fall within functional genomic regions \supercite{Lieberman-Aiden2009, Lan2012}, and the SNPs in our epistatic interactions are enriched for functional genomic regions.
+Therefore, we designed the simulations to ensure that the null distribution was of chromosome interactions between SNPs enriched for functional genomic regions but with no known epistatic interactions.
+To do this we used the 731 SNPs that form the 501 putative epistatic interactions and randomly shuffled them to create new sets of 501 pairs, disallowing any SNP combinations that were in the original set.
+Therefore, each new random set was enriched for functional regions but had no genetic interactions.
+We scanned the map of chromosome interactions for overlaps with the new sets and then repeated the random shuffling process.
+We performed 1,000 such permutations to generate a null distribution of chromosome interaction overlaps.
+
+We repeated this process, searching for overlaps within 1Mb, 250kb, and 10kb.
+
+\subsection{SNP colocalisation with genomic features}
+\label{sub:snp_colocalisation_with_genomic_features}
+
+We tested for enrichment of genomic features for the 687 IndexSNPs that comprise the 434 epistatic interactions with data present in discovery and replication datasets.
+For each of the 687 IndexSNPs we calculated LD with all regional SNPs within a radius of 0.5Mb and kept all regional SNPs with LD $r^{2} > 0.8$.
+We then cross-referenced the remaining regional SNPs with the annotated chromatin structure reference \supercite{Hoffman2012}) querying whether the regional SNPs fell in Predicted promoter region including TSS (TSS), Predicted promoter flanking region (PF), Predicted enhancer (E), Predicted weak enhancer or open chromatin cis regulatory element (WE), CTCF enriched element (CTCF), Predicted transcribed region (T), or Predicted Repressed or Low Activity region (R) positions.
+Therefore a particular IndexSNP might cover multiple genomic features through LD.
+
+We then performed the whole querying process for each of the 528,509 SNPs present in the SNP chip used in the scan, and used the results from this second analysis to establish a null distribution for the expected proportion of SNPs for each genomic feature.
+We calculated $p$-values for enrichment of each of the seven genomic features independently, and for \emph{cis}- and \emph{trans}-SNPs separately, using a binomial test.
+For each genomic feature we used the expected proportion of SNPs as the expected probability of ``success'' ($p$).
+Here, a success is defined as an IndexSNP residing in a region that includes the genomic feature.
+The observed number of successes for each IndexSNP ($k$) out of the total count of IndexSNPs ($n$) was then modelled as $\Pr (X = k) = {n \choose k} p^{k}(1-p)^{n-k}$.
+
+\subsection{Transcription factor enrichment}
+
+To test for enrichment of transcription factor binding sites (TFBS) we followed a procedure similar to that described in Section \ref{sub:snp_colocalisation_with_genomic_features}.
+For each of the 687 IndexSNPs we extracted regional SNPs as previously described.
+We then used the PWMEnrich package in Bioconductor (\url{http://www.bioconductor.org/packages/2.12/bioc/html/PWMEnrich.html}) to identify which TFBSs each of the regional SNPs for one IndexSNP falls in (within a radius of 250bp).
+Thus, the number of occurrences of a particular TFBS was counted for each IndexSNP.
+We used the ``Threshold-free affinity'' method for identifying TFBSs \supercite{Stormo2000}.
+
+We constructed a null distribution of expected TFBS occurrences based on the same null hypothesis as described in Section \ref{sub:snp_colocalisation_with_genomic_features} - the probability of an IndexSNP covering a particular TFBS is identical to any of the 528,509 SNPs in the discovery SNP chip.
+To do this, we performed the same procedure for each SNP in the discovery SNP chip as was performed for each IndexSNP to obtain an expected probability of covering a particular TFBS.
+We then tested the IndexSNPs for enrichment of each TFBS independently, and for \emph{cis}- and \emph{trans}-SNPs separately.
+$p$-values were obtained using Z-scores, calculated by using a normal approximation to the sum of binomial random variables representing motif hits along the sequence \supercite{HoSui2005}.
+
+\subsection{Defining previously identified SNP associations}
+
+The discovery dataset (BSGS) had previously been analysed for additive and dominant marginal effects for all gene expression levels \supercite{Powell2012, Powell2013}.
+To define SNPs that had been previously detected to have effects for a particular gene expression level we used a significance threshold accounting for multiple testing across SNPs and expression probes, $T_{m} = 0.05 / (528509 \times 7339) = 1.29 \times 10^{-11}$.
+From this, we found that only nine of the 501 discovery interactions had known main effects, 64 were between SNPs that had no known marginal effects, and 439 were between a SNP with a known marginal effect and a SNP with no known marginal effect.
+
+\printbibliography[category=Category2]
+
 \clearpage
@@
 	\includegraphics[width=5in]{ChromatinStatesOverlapFixedOrder}
-	\caption{\textbf{Location of SNPs relative to genomic features} We used chromatin segmentation \cite{Hoffman2012} as a method for labelling genomic features.
+	\caption{\textbf{Location of SNPs relative to genomic features} We used chromatin segmentation \supercite{Hoffman2012} as a method for labelling genomic features.
 All SNPs within 1Mb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic features (\emph{x}-axis) were covered by the SNPs that compose the 501 significant interactions.
@@
 	\includegraphics[width=5in]{H3K4me3Enrichment501_list}
-	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3 \cite{Trynka2013}.
+	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3 \supercite{Trynka2013}.
 Enrichment $p$-values are calculated using permutation analysis for 34 different cell types ($y$-axis) in four tissue types (Rows of boxes).
@@
 \begin{threeparttable}
-\caption{\textbf{Estimation of additive and non-additive variance components from pedigree information} Taken from previous analysis in Powell et al 2013 \cite{Powell2013}}
+\caption{\textbf{Estimation of additive and non-additive variance components from pedigree information} Taken from previous analysis in Powell et al 2013 \supercite{Powell2013}}
 \label{tab:vc}

### 2014-03-06 Frayling_et_al/gib/isnp1.R
added scripts and data for frayling analysis

@@ -1,0 +1,110 @@
+library(lattice)
+library(latticeExtra)
+library(gridExtra)
+
+plot3dGp <- function(gp, title="", snp1="SNP1", snp2="SNP2", z=-45)
+{
+	p <- cloud(
+		gp, # This has to be a table!?
+		panel.3d.cloud=panel.3dbars, 
+		col="black", 
+		col.facet=c("#e5f5e0", "#A1D99B", "#31A354"), 
+		# col.facet=c("#31A354"), 
+		xbase=0.4, 
+		ybase=0.4,
+		xlab=list(
+			label=snp1, 
+			cex=0.8), 
+		ylab=list(
+			label=snp2, 
+			cex=0.8), 
+		scales=list(
+			arrows=F, 
+			z = list(cex = 0), 
+			y = list(cex = 0.5), 
+			x=list(cex=0.5)),
+		zlab="",
+		cex.title = 0.5,
+		screen = list(
+			z = z, 
+			x = -60, 
+			y = 3),
+		main = list(
+			label=title, 
+			cex=0.8)
+	)
+	return(p)
+}
+
+
+
+snp1 <- bsgs[,which(colnames(bsgs) == "rs8106959")]
+snp2 <- bsgs[,which(colnames(bsgs) == "rs6718480")]
+
+1 - p[which(colnames(bsgs) == "rs8106959")]
+1 - p[which(colnames(bsgs) == "rs6718480")]
+
+
+(1 - 0.82)^2 * 0.42^2 * 450
+
+
+subset(sig, snp2 == "rs6718480" & snp1 == "rs8106959")$gcs
+subset(sig, snp2 == "rs6718480" & snp1 == "rs8106959")$gcm
+
+
+all(rownames(bsgs) == a$V2) 
+
+
+phen <- phenlist[[1]][,which(colnames(phenlist[[1]]) == "ILMN_1786426")]
+
+res <- residuals(lm(phen ~ snp))
+res2 <- residuals(lm(phen ~ snp1))
+
+gp1 <- tapply(phen, list(snp1, snp2), mean)
+gp1 <- gp1 - min(gp1)
+gp2 <- tapply(res, list(snp1, snp2), mean)
+gp2 <- gp2 - min(gp2)
+gp3 <- tapply(res2, list(snp1, snp2), mean)
+gp3 <- gp3 - min(gp3)
+
+plot3dGp(gp1, z=30)
[…]

### 2015-01-30 investigation/docs/overview.tex
overview docs

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Investigation into the architecture of epistasis
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\title{Gene expression underlying Endo susceptability}
%
%%% Preamble

\documentclass[paper=a4, fontsize=11pt]{scrartcl}					% Article class of KOMA-script with 11pt font and a4 format
\usepackage[T1]{fontenc}
\usepackage{fourier}
\usepackage{float}
\usepackage{placeins}
\usepackage[english]{babel}											% English language/hyphenation
\usepackage[protrusion=true,expansion=true]{microtype}				% Better typography
\usepackage{amsmath,amsfonts,amsthm}								% Math packages
\usepackage[pdftex]{graphicx}										% Enable pdflatex
\usepackage{url}
\usepackage{multicol}
\usepackage{lscape}
\usepackage{parskip}

%%% Custom sectioning (sectsty package)
\usepackage{sectsty}												% Custom sectioning (see below)
\allsectionsfont{\centering \normalfont\scshape}					% Change font of al section commands

%%% Custom headers/footers (fancyhdr package)
\usepackage{fancyhdr}
\pagestyle{fancyplain}
\fancyhead{}														% No page header
\fancyfoot[L]{}														% Empty 
\fancyfoot[C]{}														% Empty
\fancyfoot[R]{\thepage}												% Pagenumbering
\renewcommand{\headrulewidth}{0pt}									% Remove header underlines
\renewcommand{\footrulewidth}{0pt}									% Remove footer underlines
\setlength{\headheight}{13.6pt}
\setlength{\parindent}{0pt}

\usepackage{tabularx,ragged2e,booktabs,caption}
\newcolumntype{C}[1]{>{\Centering}m{#1}}
\renewcommand\tabularxcolumn[1]{C{#1}}

%%% Equation and float numbering
\numberwithin{equation}{section}									% Equationnumbering: section.eq#
\numberwithin{figure}{section}										% Figurenumbering: section.fig#
\numberwithin{table}{section}										% Tablenumbering: section.tab#

%%% Referencing
\bibliographystyle{naturemag}

%%% Maketitle metadata
\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} 					% Horizontal rule

\title{
		%\vspace{-1in} 	
		\usefont{OT1}{bch}{b}{n}
		\normalfont \normalsize \textsc{University of Queensland} \\ [25pt]
		\horrule{0.5pt} \\[0.4cm]
		\huge Epistasis follow-up \\[0.3cm]
        \huge *** summary *** \\
		\horrule{2pt} \\[0.5cm]
		\date{28th Nov 2014}
}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Begin document
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section{Analysis summary}
\noindent

Original discovery analysis identified 501 interactions comprising of 781 unique SNPs and 238 genes (probes). These were 26 cis-cis, 462 cis-trans and 13 trans trans. The majority of our discovery interactions were composed of one SNP that was significantly associated with the gene expression level in the discovery data set, and one SNP that had no previous association (439 out of 501, Methods). Only nine interactions were between SNPs that both had known main effects, whereas 64 were between SNPs that had no known main effects.

The following analyses have been conducted; \\

1. Determining the empirical p-values for each of the 501 interactions \\

The initial analysis used F-tests and some simulation work to determine the expected Type 1 error rate in the 1st stage of the discovery process. Subsequent simulations and theoretical calculations have suggested that this is not correct when there is a large main effect and / or in the presence of LD. \\

a. Identify which of the two snps in the original epistasis pair has the largest additive effect. Then, treating that as a fixed SNP, perform a genome-wide analysis using the 8df and 4df epistasis model. Use the interaction p-values to determine the empirical distribution of null p-values. This has been completed after filtering all SNPs on the fixed SNPs chromosome as well as +/- 5MB around the position of the original second snp.  \\

b. Identify the largest additive eQTL for the probe (irrespective of effect size). Regress out the effect of the additive loci and use the adjusted phenotype for 8df and 4df model analysis. Results reported are the 8df and 4df of original pair and the empirical p-values from genome-wide analysis fitting each of the two snps as fixed. As before these are determined by filtering out the snps on the same chromosome as the original fixed snp and also within +/- 5MB of the second snp. \\

\vspace{1cm}

2. Prediction \\

Of the 501 snp pairs, 484 have both snps in the EGCUT data. Most of the Inchianti snps need to be imputed, but we expect most to pass filtering. For pairs without and Inchiamti snp I propose using the snp with the largest additive effect in the egcut data. \\

For each pair; \\

a. Predict the phenotype in egcut data using a predictor with effects estimated from \\

4df model (estimated in BSGS) \\
8df model (estimated in BSGS) \\
1df model (estimated in BSGS using the Inchianti snp) \\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Methods}

%%% End document
\end{document}

### 2015-02-11 investigation/docs/overview.tex
edits to overview doc

@@
 
+\section{Methods}
+
 The following analyses have been conducted; \\
@@
 The initial analysis used F-tests and some simulation work to determine the expected Type 1 error rate in the 1st stage of the discovery process.
-Subsequent simulations and theoretical calculations have suggested that this is not correct when there is a large main effect and / or in the presence of LD.
+The 1st stage was followed by a 2nd stage where the interaction model was fitted.
+Subsequent simulations and theoretical calculations have suggested that the Type 1 error rate of the 2nd stage is not correct when there is a large main effect and / or in the presence of LD.
 \\
 
-a.
-Identify which of the two snps in the original epistasis pair has the largest additive effect.
-Then, treating that as a fixed SNP, perform a genome-wide analysis using the 8df and 4df epistasis model.
-Use the interaction p-values to determine the empirical distribution of null p-values.
-This has been completed after filtering all SNPs on the fixed SNPs chromosome as well as +/- 5MB around the position of the original second snp.
+a. Identify which of the two snps in the original epistasis pair has the largest additive effect. \\
+
+b. Treat the largest additive SNP as a fixed SNP and perform a genome-wide analysis using the 8df and 4df epistasis model. \\
+
+c.
+This generates $\approx 500,000$ interaction p-values.
+Apply the same snp-pair filtering as used in the manuscript.
+Namely, LD ($r^2$ < 0.1), nclass $=$ 9, and minclass $>$ 5.
+Any SNP with +/- 5MB of original epistasis SNP pairs were also removed.
 \\
 
-b.
+d.  Use the (filtered) interaction p-values to determine the empirical distribution of null p-values. \\
+
+\newpage
+
+2.
 Identify the largest additive eQTL for the probe (irrespective of effect size).

### 2015-02-11 investigation/docs/overview.tex
edits to overview doc

@@
 		\horrule{2pt} \\[0.5cm]
-		\date{28th Nov 2014}
+		\date{}
 }
@@
 \newpage
-\section{Analysis summary}
+\section*{Original analysis summary}
 \noindent
@@
 Only nine interactions were between SNPs that both had known main effects, whereas 64 were between SNPs that had no known main effects.
+\\
 
-\section{Methods}
+In the original analysis the following thresholds were used; \\
+
+\subsubsection*{Stage 1}
+
+The complete exhaustive scan for 7339 probes comprises $1.03 \times 10^{15}$ $F$-tests.
+We used permutation analysis to estimate an appropriate significance threshold for the study.
+To do this we performed a further 1600 exhaustive 2D scans on permuted phenotypes to generate a null distribution of the extreme $p$-values expected to be obtained from this number of multiple tests given the correlation structure between the SNPs.
+We took the most extreme $p$-value from each of the 1600 scans and set the 5\% FWER to be the 95\% most extreme of these $p$-values, $T_{*} = 2.13 \times 10^{-12}$.
+The effective number of tests in one 2D scan being performed is therefore $N_{*} = 0.05 / T_{*} \approx 2.33 \times 10^{10}$.
+To correct for the testing of multiple probes we established an experiment wide threshold of $T_{e} = 0.05 / (N_{*} \times 7339) = 2.91 \times 10^{-16}$.
+
+\paragraph{Filtering} We used two approaches to filter SNPs from stage 1 to be tested for significant interaction effects in stage 2.
+
+\paragraph{Filter 1} After keeping SNP pairs that surpassed the $2.91 \times 10^{-16}$ threshold in stage 1 only SNP pairs with at least 5 data points in all 9 genotype classes were kept.
+We then calculated the LD between interacting SNPs (amongst unrelated individuals within the discovery sample and also from 1000 genomes data) and removed any pairs with $r^2 > 0.1$ or $D'^{2} > 0.1$ to avoid the inclusion of haplotype effects and to increase the accuracy of genetic variance decomposition.
+If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained, \emph{i.e.} if a probe had multiple SNP pairs that were on chromosomes one and two then only the SNP pair with the most significant $p$-value was retained.
+At this stage 6404 filtered SNP pairs remained.
+
+\paragraph{Filter 2} We also performed a second filtering screen applied to the list of SNP pairs from stage 1 that was identical to filter 1 but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) were removed \cite{Powell2013}, creating a second set of 4751 unique filtered SNP pairs.
+
+\subsubsection*{Stage 2}
+
+To ensure that interacting SNPs were driven by epistasis and not marginal effects we performed a nested ANOVA on each pair in the filtered set to test if the interaction terms were significant.
+We did this by contrasting the full genetic model (8 \emph{d.f.}) against the reduced marginal effects model which included the additive and dominance terms at both SNPs (4 \emph{d.f.}).
+Thus, a 4 \emph{d.f.} $F$-test was performed on the residual genetic variation, representing the contribution of epistatic variance.
+Significance of epistasis was determined using a Bonferroni threshold of $0.05 / (6404+4751) = 4.48 \times 10^{-6}$.
+This resulted in 406 and 95 SNP pairs with significant interaction terms from filters 1 and 2, respectively.
+
+\subsection*{Type 1 error rate}
+
+Using a Bonferroni correction of $0.05$ in the second stage of the two stage discovery scan implies a type 1 error rate of $\alpha = 0.05$.
+However, this could be underestimated because the number tests performed in the second stage depends on the number of tests in the first stage, and this depends on statistical power and model choice.
+We performed simulations to estimate the type 1 error rate of this study design.
+
+We assumed a null model where there was one true additive effect and 7 other terms with no effect.
+To simulate a test statistic we simulated 8 z-scores, $z_1 \sim N(\sqrt{NCP}, 1)$ and $z_{2..8} \sim N(0,1)$.
+Thus $z_{full} = \sum^8_{i=1}{z_i} \sim \chi^{2}_{8}$ (representing the 8 d.f.
+test) and $z_{int} = \sum^8_{i=5}{z_i} \sim \chi^{2}_{4}$ (representing the 4 d.f.
+test where the null hypothesis of no epistasis is true).
+For a particular value of $NCP$ we simulated 100,000 $z$ values, and calculated the $p_{full}$-value for the $z_{full}$ test statistic.
+The $n_{int}$ test statistics with $p_{full} < 2.31 \times 10^{-16}$ were kept for the second stage, where the type 1 error rate of stage 2 was calculated as the proportion of $p_{int} < 0.05 / n_{int}$.
+The power at stage 1 was calculated as $n_{int} / 100,000$.
+This procedure was performed for a range of $NCP$ parameters that represented power ranging from $\sim 0$ to $\sim 1$.
+
+\section*{Methods and results}
 
@@
 
-\newpage
-\section{Results}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\section{Methods}
-
 %%% End document

### 2015-02-11 investigation/scripts/epi_empirical.R
edits to overview doc

@@ -104,7 +104,7 @@ analysis.fun <- function(probe, snp1, snp2, bsgs, geno, bim) {
 
 
 output <- analysis.fun(probe, snp1, snp2, bsgs, geno, bim)
-save(output, file=paste(outdir, probe, "_output.RData", sep=""))
+save(output, file=paste(outdir, probe, "_", snp1, "_", snp2, "_output.RData", sep=""))
 
 
 

### 2015-02-11 investigation/scripts/epi_empirical_analysis.R

@@ -10,4 +10,31 @@ setwd("/Users/jpowell/repo/eQTL-2D/investigation/data/output/")
 
 
 lf <- list.files()
+
+summarize.fun <- function(lf) {
+
+	out <- matrix(0, nrow=length(lf), ncol=6)
+
+	for(i in 1:length(lf)) {
+		load(lf[i])
+		index <- which(output$nclass==9 & output$minclass > 5 & output$LD < 0.1)
+		foo <- output[index,]
+
+		# Calculate lambda (median)
+
+		out[i,1] <- substr(lf[i], 1, 12)
+		out[i,2] <- nrow(foo)
+		out[i,3] <- min(foo$P) 	
+		
+
+		rm(foo)
+		rm(output)
+	}
+
+
+Z <- qnorm(1-(foo$P/2))
+lambda <- (median(na.omit(Z)))^2/0.456
+
+}
+
 for

### 2015-02-11 investigation/docs/overview.tex
edits to overview doc

@@
 
-\paragraph{Filter 2} We also performed a second filtering screen applied to the list of SNP pairs from stage 1 that was identical to filter 1 but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) were removed \cite{Powell2013}, creating a second set of 4751 unique filtered SNP pairs.
+\paragraph{Filter 2} We also performed a second filtering screen applied to the list of SNP pairs from stage 1 that was identical to filter 1 but an additional step was included where any SNPs that had previously been shown to have a significant additive or dominant effect ($p < 1.29 \times 10^{-11}$) were removed, creating a second set of 4751 unique filtered SNP pairs.
 
@@
 This procedure was performed for a range of $NCP$ parameters that represented power ranging from $\sim 0$ to $\sim 1$.
+
+\newpage
 

### 2015-02-11 investigation/scripts/epi_empirical_analysis.R

@@ -44,6 +44,7 @@ out <- summarize.fun(lf)
 
 
 
+
 # Calculate the additive eQTL effects for each pair
 
 

### 2015-02-12 investigation/docs/overview.tex
figures added and scripts to make figures edited

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\title{Gene expression underlying Endo susceptability}
 %
@@
 \bibliographystyle{naturemag}
+
+%%% Figures
+\graphicspath{{/Users/jpowell/repo/eQTL-2D/investigation/docs/figures/}}
 
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics[width=3.5in]{lambda_107.png}
+\caption*{Figure | p-value distribution - ok}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=3.5in]{lambda_305.png}
+\caption*{Figure | p-value distribution - bad}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=4in]{lambda.png}
+\caption*{Figure | Distribution of median lambda from the 501 fixed SNP analysis.
+Red denotes original pairs identified from filter 2 - i.e.
+no previous study-wide additive effects.
+Black denotes the original filter 1 pairs}
+\end{figure}
+
 2.
@@
 Regress out the effect of the additive loci and use the adjusted phenotype for 8df and 4df model analysis.
-Results reported are the 8df and 4df of original pair and the empirical p-values from genome-wide analysis fitting each of the two snps as fixed.
-As before these are determined by filtering out the snps on the same chromosome as the original fixed snp and also within +/- 5MB of the second snp.
+Results reported are the 8df and 4df of original pair and the empirical p-values from genome-wide analysis fitting each of the two SNPs as fixed.
+As before these are determined by filtering out the SNPs on the same chromosome as the original fixed SNP and also within +/- 5MB of the second SNP.
 \\
+
+[Currently running]
 
@@
 
-Of the 501 snp pairs, 484 have both snps in the EGCUT data.
-Most of the Inchianti snps need to be imputed, but we expect most to pass filtering.
-For pairs without and Inchiamti snp I propose using the snp with the largest additive effect in the egcut data.
+Of the 501 SNP pairs, 484 have both SNP in the EGCUT data.
+Most of the Inchianti SNPs need to be imputed, but we expect most to pass filtering.
+For pairs without and Inchianti SNP I propose using the SNP with the largest additive effect in the egcut data.
 \\

### 2015-02-12 investigation/scripts/epi_empirical_analysis.R

@@ -39,7 +39,7 @@ gs <- filter_add.fun(sig, gs)
 # Determine the summary of the output
 
 lambda <- summarize.fun(lf)
-gs <- cbind(gs, lambda[,5:6])
+gs <- cbind(gs, lambda[,5:7])
 
 
 ##################################################################
@@ -71,10 +71,20 @@ hist(as.numeric(as.matrix(gs$lambda)), breaks=25,
 dev.off()
 
 
+load(lf[3])
+index <- which(output$nclass==9 & output$minclass > 5 & output$LD < 0.1)
+foo <- output[index,]
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/lambda_107.png")
+hist(foo$P, main="ILMN_1651886 - lambda = 1.07", xlab="p-value", col="lightgrey")
+dev.off()
 
 
-
-
+load(lf[17])
+index <- which(output$nclass==9 & output$minclass > 5 & output$LD < 0.1)
+foo <- output[index,]
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/lambda_305.png")
+hist(foo$P, main="ILMN_1660549 - lambda = 3.05", xlab="p-value", col="lightgrey")
+dev.off()
 
 
 

### 2015-02-12 investigation/scripts/epi_empirical_analysis_fun.R

@@ -8,7 +8,7 @@
 # provide summary formation for the output results
 summarize.fun <- function(lf) {
 
-	out <- matrix(0, nrow=length(lf), ncol=6)
+	out <- matrix(0, nrow=length(lf), ncol=7)
 
 	for(i in 1:length(lf)) {
 		load(lf[i])
@@ -25,6 +25,7 @@ summarize.fun <- function(lf) {
 		Z <- qnorm(1-(foo$P/2))
 		out[i,5] <- round((median(na.omit(Z)))^2/0.456, 2)
 		out[i,6] <- length(which(foo$P < 4.48e-6))
+		out[i,7] <- length(which(foo$P < 0.05/nrow(foo)))
 
 		# 
 
@@ -35,7 +36,7 @@ summarize.fun <- function(lf) {
 	}
 
 	out <- as.data.frame(out)
-	names(out) <- c("probename", "snp1", "snp2", "nsnps", "lambda", "nthreshold")	
+	names(out) <- c("probename", "snp1", "snp2", "nsnps", "lambda", "nthreshold", "nadjustedthreshold")	
 	return(out)
 }
 

### 2015-02-12 investigation/docs/overview.tex
figures added and scripts to make figures edited

@@
 
+\begin{landscape}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrr}
+  \hline
+ & F & P & nclass & minclass & LD \\ 
+  \hline
+1 & 0.36 & 0.78 & 8.00 & 1.00 & 0.00 \\ 
+  2 & 0.44 & 0.72 & 8.00 & 7.00 & 0.00 \\ 
+  3 & 0.33 & 0.85 & 9.00 & 2.00 & 0.00 \\ 
+  4 & 0.36 & 0.84 & 9.00 & 6.00 & 0.01 \\ 
+  5 & 0.29 & 0.75 & 6.00 & 1.00 & 0.00 \\ 
+  6 & 0.12 & 0.89 & 6.00 & 1.00 & 0.00 \\
+  ... & ... & ... & ... & ...& ... \\ 
+   \hline
+\end{tabular}
+\caption*{Header of output for a single pair}
+\end{table}
+
+\vspace{2cm}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rllllllllrlll}
+  \hline
+ & probename & snp1 & snp2 & nclass9 & minclass5 & LD01 & nsnpspass & nsnps & filter & lambda & nthres & nadjthres \\ 
+  \hline
+1 & ILMN\_1651385 & rs7989895 & rs4846085 & 419387 & 285750 & 506818 & 269121 & 506818 & 1.00 & 1.61 & 4 & 0 \\ 
+  2 & ILMN\_1651705 & rs872311 & rs11032695 & 419021 & 278721 & 511121 & 260851 & 511121 & 1.00 & 1.12 & 1 & 0 \\ 
+  3 & ILMN\_1651886 & rs7108734 & rs12784396 & 427787 & 319167 & 501291 & 305496 & 501291 & 1.00 & 1.07 & 0 & 0 \\ 
+  4 & ILMN\_1652333 & rs898095 & rs9892064 & 442554 & 336649 & 515007 & 322786 & 515007 & 2.00 & 1.34 & 7 & 0 \\ 
+  5 & ILMN\_1653205 & rs12429804 & rs2896452 & 383323 & 213815 & 507099 & 183628 & 507099 & 1.00 & 1.78 & 24 & 4 \\ 
+  6 & ILMN\_1653205 & rs12454561 & rs2896452 & 386460 & 215536 & 511390 & 185098 & 511390 & 1.00 & 1.78 & 25 & 4 \\ 
+  ... & ... & ... & ... & ... & ... & ... & ... & ... & ... & ... & ... & ... \\ 
+   \hline
+\end{tabular}
+\caption*{Header of summary output for 501 pairs}
+\end{table}
+\end{landscape}
+
+\newpage
+
 \begin{figure}[H]
@@
 \end{figure}
+
+\newpage
 

### 2015-02-12 investigation/scripts/epi_empirical_analysis.R

@@ -87,6 +87,10 @@ hist(foo$P, main="ILMN_1660549 - lambda = 3.05", xlab="p-value", col="lightgrey"
 dev.off()
 
 
+hist(-log10(as.numeric(as.matrix(gs$P_emp))), breaks=20,
+	main="", xlab="-log10 p-values", col="lightgrey")
+
+
 
 
 

### 2015-02-18 investigation/docs/overview.tex
updtaes to scripts and report

@@
 
+\begin{table}[ht]
+\centering
+\begin{tabular}{rlllll}
+  \hline
+probename & snp1 & snp2 & meanlambda & npairs \\ 
+  \hline
+ILMN\_1704730 & rs1884655 & rs10255470 & 2.88 & 10 \\ 
+ILMN\_1710752 & rs2123758 & rs2786014 & 2.15 & 8 \\ 
+ILMN\_1717234 & rs1157079 & rs7733671 & 4.31 & 17 \\ 
+ILMN\_1720059 & rs12435486 & rs7837237 & 2.29 & 7 \\ 
+ILMN\_1738784 & rs10930170 & rs12120009 & 2.24 & 6 \\ 
+ILMN\_1755589 & rs11080134 & rs11169322 & 1.16 & 6 \\ 
+ILMN\_1786426 & rs2839013 & rs8106959 & 5.65 & 20 \\ 
+ILMN\_1804396 & rs1293455 & rs2655991 & 1.38 & 7 \\ 
+ILMN\_2313158 & rs10869600 & rs13069559 & 3.15 & 15 \\ 
+ILMN\_2372639 & rs17159840 & rs10059004 & 4.17 & 17 \\ 
+ILMN\_3231952 & rs12947580 & rs8079215 & 2.16 & 6 \\ 
+   \hline
+\end{tabular}
+\caption*{Mean lambda per probe, when 5 or more interactions are identified amongst the 501 pairs}
+\end{table}
+
 \newpage

### 2015-02-18 investigation/scripts/epi_empirical_analysis.R

@@ -3,6 +3,8 @@
 # analysis and interpretation of the output from epi_empirical.R
 # joseph.powell@uq.edu.au
 
+library(xtable)
+
 # read in the functions
 source("/Users/jpowell/repo/eQTL-2D/investigation/scripts/epi_empirical_analysis_fun.R")
 
@@ -11,6 +13,9 @@ load("/Users/jpowell/repo/eQTL-2D/investigation/data/investigation_data.RData")
 
 # set the data dir
 setwd("/Users/jpowell/repo/eQTL-2D/investigation/data/output/")
+
+# Read in 30 sig list
+sig <- read.csv("/Users/jpowell/repo/eQTL-2D/investigation/data/sig_list.csv", header=T)
 
 # list files
 lf <- list.files()
@@ -105,6 +110,7 @@ hist(-log10(as.numeric(as.matrix(gs$P_emp))), breaks=20,
 # 4. Calculate a mean lambda per probe, for those probes that were in the 501 selected trios. I am guessing that the lambdas for probes MBLN1 and TMEM149 are large.
 
 multi_lambda <- multi_lambda.fun(gs,5)
+xtable(multi_lambda)
 
 
 

### 2015-02-18 investigation/scripts/epi_empirical_analysis_fun.R

@@ -10,7 +10,7 @@
 # provide summary formation for the output results
 summarize.fun <- function(lf) {
 
-	out <- matrix(0, nrow=length(lf), ncol=9)
+	out <- matrix(0, nrow=length(lf), ncol=10)
 
 	for(i in 1:length(lf)) {
 		load(lf[i])
@@ -27,7 +27,7 @@ summarize.fun <- function(lf) {
 		Z <- qnorm(1-(foo$P/2))
 		out[i,5] <- round((median(na.omit(Z)))^2/0.456, 2)
 		out[i,6] <- length(which(foo$P < 4.48e-6))
-		out[i,7] <- length(which(foo$P < 0.05/nrow(foo)))
+		# out[i,7] <- length(which(foo$P < 0.05/nrow(foo)))
 
 		# Calculate N pairs with F_i > H0-F from 4.48x10-6
 		F_thres <- qf(1-(4.48e-6), df1=4, df2=846)
@@ -35,16 +35,31 @@ summarize.fun <- function(lf) {
 
 		if(Q==0) {
 
-			out[i,8] <- NA
-			out[i,9] <- NA
+			F_sort <- sort(foo$F, decreasing=T)
+			F_emp <- F_sort[1]
+			P_emp <- round(1-pf(F_emp, df1=4, df2=846), 3)
+			out[i,9] <- F_emp
+			out[i,10] <- P_emp
+
+			# Emp Type 1 error rate
+			F_empN <- qf(1-(0.05/nrow(foo)), df1=4, df2=846)
+			out[i,7] <- round(F_empN,2)
+			out[i,8] <- length(which(F_sort > F_empN))
+
 		}
 
 		else{
-			F_emp <- sort(foo$F, decreasing=T)[Q]
-			P_emp <- 1-pf(F_emp, df1=4, df2=842)
+			F_sort <- sort(foo$F, decreasing=T)
+			F_emp <- F_sort[Q]
+			P_emp <- round(1-pf(F_emp, df1=4, df2=842), 3)
 
-			out[i,8] <- F_emp
-			out[i,9] <- P_emp
+			out[i,9] <- F_emp
+			out[i,10] <- P_emp
+
+			# Emp Type1 error
+			F_empN <- qf(1-(0.05/nrow(foo)), df1=4, df2=846)
+			out[i,7] <- round(F_empN,2)
+			out[i,8] <- length(which(F_sort > F_empN))
 		}
 		
 			
[…]

### 2015-02-18 investigation/docs/overview.tex
updtaes to scripts and report

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage
@@
 \includegraphics[width=3.5in]{lambda_107.png}
-\caption*{Figure | p-value distribution - ok}
+\caption*{Figure | Example of p-value distribution - ok}
 \end{figure}
@@
 \includegraphics[width=3.5in]{lambda_305.png}
-\caption*{Figure | p-value distribution - bad}
+\caption*{Figure | Example of p-value distribution - bad}
 \end{figure}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
 \begin{figure}[H]
@@
 \end{figure}
+
+\vspace{1cm}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+
+\begin{figure}[H]
+\centering
+\includegraphics[height=3in]{empF.png}
+\caption*{Figure | The F-statistic corresponding to $p=4.48^{-6}$ and df1=4, df2=842 from a $H_0$ table is 7.67.
+This figure shows the empirical (ranked) F-statistic corresponding to $n * 4.48^{-6}$.
+Where $n * 4.48^{-6} < 1$, the largest F-statistic taken.}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[height=3in]{empPval.png}
+\caption*{Figure | The -log10 p-values corresponding to above F-statistics (df1=4, df2=842). The -log10 $p=4.48^{-6} = 5.35$}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=4in]{Npairs_above_empthres.png}
+\caption*{Figure | For each pair (n=501) the number of tests where the test statistic is greater than the 95th percentile of an F-dist}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage

### 2015-02-18 investigation/scripts/epi_empirical_analysis.R

@@ -44,7 +44,7 @@ gs <- filter_add.fun(sig, gs)
 # Determine the summary of the output
 
 lambda <- summarize.fun(lf)
-gs <- cbind(gs, lambda[,5:9])
+gs <- cbind(gs, lambda[,5:10])
 
 
 ##################################################################
@@ -91,18 +91,34 @@ png(filename="~/repo/eQTL-2D/investigation/docs/figures/lambda_305.png")
 hist(foo$P, main="ILMN_1660549 - lambda = 3.05", xlab="p-value", col="lightgrey")
 dev.off()
 
-
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/empPval.png")
 hist(-log10(as.numeric(as.matrix(gs$P_emp))), breaks=20,
 	main="", xlab="-log10 p-values", col="lightgrey")
+dev.off()
 
-
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/empF.png")
+hist(as.numeric(as.matrix(gs$F_emp)), breaks=20,
+	main="", xlab="Empirical F-value", col="lightgrey")
+dev.off()
 
 
 # 1. For each probe-SNP pair (where SNP = fixed), calculate an empirical 5% type-I error rate, as you suggested yourself this morning. This will be useful to assess replication results. So count the number of observed test statistics greater than the 95th percentile of an F-distribution with [4,N-5] degrees of freedom. I'm not sure what the df in the denominator are, something like N-4 or N-5. 
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/Npairs_above_empthres.png")
+hist(as.numeric(as.matrix(gs$N_F_empNtests)), breaks=25, 
+	xlab="N pairs above emp 0.05 threshold", col="lightgrey", main="")
+dev.off()
 
-# 2. Given results from 1., display the empirical type-I error rate for the probe-SNP-SNP trios that are in the Nature article table (the top 30).
 
-# 3. For each probe-SNP pair, list the largest observed F-statistic from the empirical results, list the corresponding p-value (= 1/n I think, where n = number of SNPs that passed the filters), and list the F-statistic(s) from the probe-SNP-SNP trios that were among the 501.
+
+
+
+# 2. Given results from 1.
+# a) display the empirical type-I error rate for the probe-SNP-SNP trios that are in the Nature article table (the top 30).
+
+# 3. For each probe-SNP pair, list 
+# a) the largest observed F-statistic from the empirical results
[…]

### 2015-02-18 investigation/scripts/epi_empirical_analysis_fun.R

@@ -37,7 +37,7 @@ summarize.fun <- function(lf) {
 
 			F_sort <- sort(foo$F, decreasing=T)
 			F_emp <- F_sort[1]
-			P_emp <- round(1-pf(F_emp, df1=4, df2=846), 3)
+			P_emp <- 1-pf(F_emp, df1=4, df2=846)
 			out[i,9] <- F_emp
 			out[i,10] <- P_emp
 
@@ -51,7 +51,7 @@ summarize.fun <- function(lf) {
 		else{
 			F_sort <- sort(foo$F, decreasing=T)
 			F_emp <- F_sort[Q]
-			P_emp <- round(1-pf(F_emp, df1=4, df2=842), 3)
+			P_emp <- 1-pf(F_emp, df1=4, df2=842)
 
 			out[i,9] <- F_emp
 			out[i,10] <- P_emp

### 2015-02-19 investigation/docs/overview.tex
updtaes to scripts and report

@@
 \centering
-\begin{tabular}{rllllllllrlll}
+\tiny
+\begin{tabular}{llllllllrrrlllllll}
   \hline
- & probename & snp1 & snp2 & nclass9 & minclass5 & LD01 & nsnpspass & nsnps & filter & lambda & nthres & nadjthres \\ 
+ probename & snp1 & snp2 & nclass9 & minclass5 & LD01 & npass & nsnps & filter & p\_egcut & p\_fehr & $\lambda$ & nthres & Fe & N\_Fe & F\_emp & P\_emp & type1 \\ 
   \hline
-1 & ILMN\_1651385 & rs7989895 & rs4846085 & 419387 & 285750 & 506818 & 269121 & 506818 & 1.00 & 1.61 & 4 & 0 \\ 
-  2 & ILMN\_1651705 & rs872311 & rs11032695 & 419021 & 278721 & 511121 & 260851 & 511121 & 1.00 & 1.12 & 1 & 0 \\ 
-  3 & ILMN\_1651886 & rs7108734 & rs12784396 & 427787 & 319167 & 501291 & 305496 & 501291 & 1.00 & 1.07 & 0 & 0 \\ 
-  4 & ILMN\_1652333 & rs898095 & rs9892064 & 442554 & 336649 & 515007 & 322786 & 515007 & 2.00 & 1.34 & 7 & 0 \\ 
-  5 & ILMN\_1653205 & rs12429804 & rs2896452 & 383323 & 213815 & 507099 & 183628 & 507099 & 1.00 & 1.78 & 24 & 4 \\ 
-  6 & ILMN\_1653205 & rs12454561 & rs2896452 & 386460 & 215536 & 511390 & 185098 & 511390 & 1.00 & 1.78 & 25 & 4 \\ 
-  ... & ... & ... & ... & ... & ... & ... & ... & ... & ... & ... & ... & ... \\ 
+ILMN\_1651385 & rs7989895 & rs4846085 & 419387 & 285750 & 506818 & 269121 & 506818 & 1.00 & 0.25 & 0.61 & 1.61 & 4 & 9.43 & 0 & 8.95 & 6.35 & 0.10 \\ 
+ILMN\_1651705 & rs872311 & rs11032695 & 419021 & 278721 & 511121 & 260851 & 511121 & 1.00 & 0.30 & 0.26 & 1.12 & 1 & 9.41 & 0 & 7.75 & 5.41 & 0.05 \\ 
+ILMN\_1651886 & rs7108734 & rs12784396 & 427787 & 319167 & 501291 & 305496 & 501291 & 1.00 & 0.01 & 0.21 & 1.07 & 0 & 9.5 & 0 & 7.4 & 5.13 & 0.05 \\ 
+ILMN\_1652333 & rs898095 & rs9892064 & 442554 & 336649 & 515007 & 322786 & 515007 & 2.00 & 29.39 & 28.24 & 1.34 & 7 & 9.53 & 0 & 9.48 & 6.77 & 0.07 \\ 
+ILMN\_1653205 & rs12429804 & rs2896452 & 383323 & 213815 & 507099 & 183628 & 507099 & 1.00 & 0.02 & 0.29 & 1.78 & 24 & 9.22 & 4 & 10.56 & 7.63 & 0.11 \\ 
+ILMN\_1653205 & rs12454561 & rs2896452 & 386460 & 215536 & 511390 & 185098 & 511390 & 1.00 &  & 0.31 & 1.78 & 25 & 9.22 & 4 & 10.56 & 7.63 & 0.11 \\ 
+ILMN\_1653205 & rs2896452 & rs1004564 & 373839 & 208706 & 494954 & 179170 & 494954 & 1.00 & 0.18 & 0.38 & 1.78 & 23 & 9.2 & 2 & 10.08 & 7.25 & 0.11 \\ 
+ILMN\_1653205 & rs7152284 & rs2896452 & 385471 & 215221 & 509793 & 184859 & 509793 & 1.00 & 0.07 & 2.18 & 1.78 & 23 & 9.22 & 3 & 10.56 & 7.63 & 0.11 \\ 
+ILMN\_1653205 & rs8051751 & rs2896452 & 386341 & 215644 & 511501 & 185188 & 511501 & 1.00 & 0.18 & 1.39 & 1.78 & 24 & 9.22 & 4 & 10.56 & 7.63 & 0.11 \\ 
+ILMN\_1654545 & rs4333645 & rs1455268 & 421112 & 307876 & 495374 & 294544 & 495374 & 2.00 & 0.01 & 0.10 & 1.45 & 4 & 9.48 & 0 & 9.14 & 6.5 & 0.08 \\ 
+ILMN\_1656378 & rs10906857 & rs12490878 & 418379 & 314668 & 489831 & 301252 & 489831 & 1.00 & 0.34 & 0.42 & 1.1 & 1 & 9.49 & 0 & 7.69 & 5.36 & 0.05 \\ 
+ILMN\_1658247 & rs11613438 & rs1047944 & 402228 & 251785 & 490871 & 234981 & 490871 & 2.00 & 1.55 & 1.27 & 1.53 & 21 & 9.35 & 1 & 9.73 & 6.97 & 0.09 \\ 
    \hline
@@
 \end{table}
+
 \end{landscape}

### 2015-02-19 investigation/scripts/epi_empirical_analysis.R

@@ -15,7 +15,7 @@ load("/Users/jpowell/repo/eQTL-2D/investigation/data/investigation_data.RData")
 setwd("/Users/jpowell/repo/eQTL-2D/investigation/data/output/")
 
 # Read in 30 sig list
-sig <- read.csv("/Users/jpowell/repo/eQTL-2D/investigation/data/sig_list.csv", header=T)
+sig30 <- read.csv("/Users/jpowell/repo/eQTL-2D/investigation/data/sig_list.csv", header=T)
 
 # list files
 lf <- list.files()
@@ -44,7 +44,7 @@ gs <- filter_add.fun(sig, gs)
 # Determine the summary of the output
 
 lambda <- summarize.fun(lf)
-gs <- cbind(gs, lambda[,5:10])
+gs <- cbind(gs, lambda[,5:11])
 
 
 ##################################################################
@@ -92,7 +92,7 @@ hist(foo$P, main="ILMN_1660549 - lambda = 3.05", xlab="p-value", col="lightgrey"
 dev.off()
 
 png(filename="~/repo/eQTL-2D/investigation/docs/figures/empPval.png")
-hist(-log10(as.numeric(as.matrix(gs$P_emp))), breaks=20,
+hist(as.numeric(as.matrix(gs$P_emp)), breaks=20,
 	main="", xlab="-log10 p-values", col="lightgrey")
 dev.off()
 

### 2015-02-19 investigation/scripts/epi_empirical_analysis_fun.R

@@ -10,7 +10,7 @@
 # provide summary formation for the output results
 summarize.fun <- function(lf) {
 
-	out <- matrix(0, nrow=length(lf), ncol=10)
+	out <- matrix(0, nrow=length(lf), ncol=11)
 
 	for(i in 1:length(lf)) {
 		load(lf[i])
@@ -37,7 +37,7 @@ summarize.fun <- function(lf) {
 
 			F_sort <- sort(foo$F, decreasing=T)
 			F_emp <- F_sort[1]
-			P_emp <- 1-pf(F_emp, df1=4, df2=846)
+			P_emp <- round(-log10(1-pf(F_emp, df1=4, df2=846)),2)
 			out[i,9] <- F_emp
 			out[i,10] <- P_emp
 
@@ -51,7 +51,7 @@ summarize.fun <- function(lf) {
 		else{
 			F_sort <- sort(foo$F, decreasing=T)
 			F_emp <- F_sort[Q]
-			P_emp <- 1-pf(F_emp, df1=4, df2=842)
+			P_emp <- round(-log10(1-pf(F_emp, df1=4, df2=842)),2)
 
 			out[i,9] <- F_emp
 			out[i,10] <- P_emp
@@ -62,6 +62,10 @@ summarize.fun <- function(lf) {
 			out[i,8] <- length(which(F_sort > F_empN))
 		}
 		
+		# calculate the type 1 error rate
+		F <- qf(1-(0.05), df1=4, df2=842) 
+		out[i,11] <- round(length(which(foo$F > F))/nrow(foo),3)
+
 			
 		rm(foo)
 		rm(output)
@@ -70,7 +74,7 @@ summarize.fun <- function(lf) {
 	}
 
 	out <- as.data.frame(out)
-	names(out) <- c("probename", "snp1", "snp2", "nsnps", "lambda", "nthreshold", "F_empNtests", "N_F_empNtests", "F_emp", "P_emp")	
+	names(out) <- c("probename", "snp1", "snp2", "nsnps", "lambda", "nthreshold", "F_empNtests", "N_F_empNtests", "F_emp", "P_emp", "Type1")	
 	return(out)
 }
 

### 2015-02-19 investigation/docs/overview.tex
updtaes to scripts and report

@@
 \centering
-\includegraphics[width=4in]{Npairs_above_empthres.png}
+\includegraphics[height=3in]{Npairs_above_empthres.png}
 \caption*{Figure | For each pair (n=501) the number of tests where the test statistic is greater than the 95th percentile of an F-dist}
 \end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[height=3in]{type1.png}
+\caption*{Figure | Distribution of the empirical type 1 error rate. Determined as the proportion of tests with $F > 2.38$ (95th percentile of F-dist with df1=4, df2=842)}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{llllll}
+  \hline
+gene & probename & snp1 & snp2 & $\lambda$ & Type1 \\ 
+  \hline
+ADK & ILMN\_2358626 & rs2395095 & rs10824092 & 1.15 & 0.06 \\ 
+ATP13A1 & ILMN\_2134224 & rs4284750 & rs873870 & 1.82 & 0.119 \\ 
+C21ORF57 & ILMN\_1795836 & rs9978658 & rs11701361 & 1.43 & 0.085 \\ 
+CSTB & ILMN\_1761797 & rs9979356 & rs3761385 & 1.05 & 0.053 \\ 
+CTSC & ILMN\_2242463 & rs7930237 & rs556895 & 1.12 & 0.054 \\ 
+FN3KRP & ILMN\_1652333 & rs898095 & rs9892064 & 1.34 & 0.075 \\ 
+GAA & ILMN\_2410783 & rs11150847 & rs12602462 & 1.38 & 0.081 \\ 
+HNRPH1 & ILMN\_2101920 & rs6894268 & rs4700810 & 1.47 & 0.091 \\ 
+LAX1 & ILMN\_1769782 & rs1891432 & rs10900520 & 1.22 & 0.066 \\ 
+MBLN1 & ILMN\_2313158 & rs11981513 & rs13069559 & 3.36 & 0.242 \\ 
+MBLN1 & ILMN\_2313158 & rs16864367 & rs13079208 & 2.28 & 0.161 \\ 
+MBLN1 & ILMN\_2313158 & rs2030926 & rs13069559 & 3.39 & 0.241 \\ 
+MBLN1 & ILMN\_2313158 & rs218671 & rs13069559 & 3.36 & 0.242 \\ 
+MBLN1 & ILMN\_2313158 & rs2614467 & rs13069559 & 3.37 & 0.242 \\ 
+MBLN1 & ILMN\_2313158 & rs7710738 & rs13069559 & 3.38 & 0.243 \\ 
+MBP & ILMN\_2398939 & rs8092433 & rs4890876 & 1.19 & 0.066 \\ 
+NAPRT1 & ILMN\_1710752 & rs2123758 & rs3889129 & 2.14 & 0.155 \\ 
+NCL & ILMN\_2121437 & rs7563453 & rs4973397 & 1.2 & 0.065 \\ 
+PRMT2 & ILMN\_1675038 & rs2839372 & rs11701058 & 2.77 & 0.199 \\ 
+SNORD14A & ILMN\_1799381 & rs2634462 & rs6486334 & 2.45 & 0.168 \\ 
+TMEM149 & ILMN\_1786426 & rs807491 & rs7254601 & 2.69 & 0.194 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs1843357 & 5.79 & 0.396 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs2351458 & 5.77 & 0.395 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs6718480 & 5.82 & 0.397 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs6926382 & 5.75 & 0.396 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs914940 & 5.76 & 0.395 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs9509428 & 5.82 & 0.398 \\ 
+VASP & ILMN\_1743646 & rs1264226 & rs2276470 & 1.37 & 0.08 \\ 
+RPL13 & ILMN\_2413278 & rs352935 & rs2965817 & 1.11 & 0.058 \\ 
+TRA2A & ILMN\_1731043 & rs7776572 & rs11770192 & 1.36 & 0.079 \\ 
+   \hline
+\end{tabular}
+\caption*{The $\lambda$ and type 1 error rates for the 30 (bonferroni-correction) significant replicated pairs mentioned in Hemani \emph{et al}.}
+\end{table}
 

### 2015-02-19 investigation/scripts/epi_empirical_analysis.R

@@ -108,12 +108,17 @@ hist(as.numeric(as.matrix(gs$N_F_empNtests)), breaks=25,
 	xlab="N pairs above emp 0.05 threshold", col="lightgrey", main="")
 dev.off()
 
-
-
-
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/type1.png")
+hist(as.numeric(as.matrix(gs$Type1)), breaks=35,
+	xlab="Empirical type 1 error rate", col="lightgrey", main="")
+dev.off()
 
 # 2. Given results from 1.
 # a) display the empirical type-I error rate for the probe-SNP-SNP trios that are in the Nature article table (the top 30).
+gs30 <- type1_30.fun(gs, sig30)
+xtable(gs30[,c(1:4,13,19)])
+
+
 
 # 3. For each probe-SNP pair, list 
 # a) the largest observed F-statistic from the empirical results

### 2015-02-19 investigation/scripts/epi_empirical_analysis_fun.R

@@ -218,6 +218,26 @@ multi_lambda.fun <- function(gs, n) {
 
 
 
+##################################################################
+##################################################################
+##################################################################
+# subset gs table to match the 30 paper significant pairs
+
+type1_30.fun <- function(gs, sig30) {
+
+	index <- rep(0, 30)
+	gene <- 
+	for(i in 1:nrow(sig30)) {
+		index[i] <- which(gs$probename==as.character(sig30$Probe[i]) & gs$snp1==as.character(sig30$SNP1[i]) & gs$snp2==as.character(sig30$SNP2[i]))	
+	}	
+
+	gs30 <- gs[index,]
+	gs30 <- cbind(sig30$GENE, gs30)
+	names(gs30)[1] <- "gene"
+	return(gs30)
+
+}
+
 
 
 

### 2015-02-19 investigation/docs/overview.tex
uopdate

@@
 This figure shows the empirical (ranked) F-statistic corresponding to $n * 4.48^{-6}$.
-Where $n * 4.48^{-6} < 1$, the largest F-statistic taken.}
+Where $n * 4.48^{-6} < 1$, the largest F-statistic taken.
+There are 96 pairs where the corresponding F statistic is less than 7.67}
 \end{figure}
@@
 \end{tabular}
-\caption*{The $\lambda$ and type 1 error rates for the 30 (bonferroni-correction) significant replicated pairs mentioned in Hemani \emph{et al}.}
+\caption*{The $\lambda$ and type 1 error rates for the 30 (Bonferroni-correction) significant replicated pairs mentioned in Hemani \emph{et al}.}
 \end{table}

### 2015-04-15 investigation/docs/overview.tex
edits to the report doc

@@
 
-1. Determining the empirical p-values for each of the 501 interactions \\
+\textbf{1. GWAS based determination of the empirical p-values for each of the 501 interactions} \\
 
@@
 
-a. Identify which of the two snps in the original epistasis pair has the largest additive effect. \\
+a. The SNP with largest additive effect was identified for each pair of the 501 original epistasis SNPs. \\
 
-b. Treat the largest additive SNP as a fixed SNP and perform a genome-wide analysis using the 8df and 4df epistasis model. \\
+b. The largest additive SNP was treated as a fixed SNP and a genome-wide analysis using the 8df and 4df epistasis model was performed. \\
 
 c.
-This generates $\approx 500,000$ interaction p-values.
-Apply the same snp-pair filtering as used in the manuscript.
+This generated $\approx 500,000$ interaction p-values.
+The sample snp-pair filtering as used in the manuscript was applied.
 Namely, LD ($r^2$ < 0.1), nclass $=$ 9, and minclass $>$ 5.
@@
 
-d.  Use the (filtered) interaction p-values to determine the empirical distribution of null p-values. \\
+d.  The filtered interaction p-values were used to determine the empirical distribution of null p-values. \\
+
+e. Summary information such as median lambda were calculated from the filtered interaction p-values. \\
+
+\textbf{2. Permutation based determination of the empirical p-values for each of the 501 interactions} \\
+
+a. As before, the SNP with largest additive effect was identified for each pair of the 501 original epistasis SNPs. \\
+
+b. Genotypes at the corresponding epistatitic ''co-SNP'' were randomly shuffled (no replacement) amongst individuals. \\
+
+c. Interaction p-values and text statistics were calculated using the same methods. \\
+
+d. For each pair (\emph{n}=501) this analysis was performed 10,000,000 times. \\
+
+e. The same filtering was applied (although $r^2$ not required). \\
+
+f. The empirical \emph{p}-value determined based on rank of observed \emph{F}-Statistic
 
@@
 \end{landscape}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=3.5in]{lambda_107.png}
-\caption*{Figure | Example of p-value distribution - ok}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=3.5in]{lambda_305.png}
-\caption*{Figure | Example of p-value distribution - bad}
-\end{figure}
 

### 2015-04-15 investigation/docs/overview.tex
edits to the report doc and scripts for the analysis

@@
 \centering
-\includegraphics[width=4in]{lambda.png}
-\caption*{Figure | Distribution of median lambda from the 501 fixed SNP analysis.
-Red denotes original pairs identified from filter 2 - i.e.
-no previous study-wide additive effects.
-Black denotes the original filter 1 pairs}
+\includegraphics[width=3in]{lambdaF1.png}
+\caption*{Figure | Distribution of median lambda from the 406 filter 1 pairs from the GWAS based analysis. The mean lambda is 1.94}
+
+\end{figure}
+\begin{figure}[H]
+\centering
+\includegraphics[width=3in]{lambdaF2.png}
+\caption*{Figure | Distribution of median lambda from the 95 filter 2 pairs from the GWAS based analysis. The mean lambda is 1.31}
 \end{figure}
@@
   \hline
-probename & snp1 & snp2 & meanlambda & npairs \\ 
+ probename & snp1 & snp2 & gene & meanlambda & npairs \\ 
   \hline
-ILMN\_1704730 & rs1884655 & rs10255470 & 2.88 & 10 \\ 
-ILMN\_1710752 & rs2123758 & rs2786014 & 2.15 & 8 \\ 
-ILMN\_1717234 & rs1157079 & rs7733671 & 4.31 & 17 \\ 
-ILMN\_1720059 & rs12435486 & rs7837237 & 2.29 & 7 \\ 
-ILMN\_1738784 & rs10930170 & rs12120009 & 2.24 & 6 \\ 
-ILMN\_1755589 & rs11080134 & rs11169322 & 1.16 & 6 \\ 
-ILMN\_1786426 & rs2839013 & rs8106959 & 5.65 & 20 \\ 
-ILMN\_1804396 & rs1293455 & rs2655991 & 1.38 & 7 \\ 
-ILMN\_2313158 & rs10869600 & rs13069559 & 3.15 & 15 \\ 
-ILMN\_2372639 & rs17159840 & rs10059004 & 4.17 & 17 \\ 
-ILMN\_3231952 & rs12947580 & rs8079215 & 2.16 & 6 \\ 
+ILMN\_1704730 & rs1884655 & rs10255470 & CD93 & 2.88 & 10 \\ 
+ILMN\_1710752 & rs2123758 & rs2786014 & NAPRT1 & 2.15 & 8 \\ 
+ILMN\_1717234 & rs1157079 & rs7733671 & CAST & 4.31 & 17 \\ 
+ILMN\_1720059 & rs12435486 & rs7837237 & HMBOX1 & 2.29 & 7 \\ 
+ILMN\_1738784 & rs10930170 & rs12120009 & PPP2R5A & 2.24 & 6 \\ 
+ILMN\_1755589 & rs11080134 & rs11169322 & DIP2B & 1.16 & 6 \\ 
+ILMN\_1786426 & rs2839013 & rs8106959 & TMEM149 & 5.65 & 20 \\ 
+ILMN\_1804396 & rs1293455 & rs2655991 & C14ORF4 & 1.38 & 7 \\ 
+ILMN\_2313158 & rs10869600 & rs13069559 & MBNL1 & 3.15 & 15 \\ 
+ILMN\_2372639 & rs17159840 & rs10059004 & TRAPPC5 & 4.17 & 17 \\ 
+ILMN\_3231952 & rs12947580 & rs8079215 & ARL17B & 2.16 & 6 \\ 
    \hline
@@
 This figure shows the empirical (ranked) F-statistic corresponding to $n * 4.48^{-6}$.
-Where $n * 4.48^{-6} < 1$, the largest F-statistic taken.
-There are 96 pairs where the corresponding F statistic is less than 7.67}
+Where $n * 4.48^{-6} < 1$, the largest F-statistic is taken.
+There are 96 pairs where the corresponding F statistic is less than 7.67}.
+None of these are amongst the 30 bonferroni correction significant pairs.
 \end{figure}
@@
 \includegraphics[height=3in]{Npairs_above_empthres.png}
-\caption*{Figure | For each pair (n=501) the number of tests where the test statistic is greater than the 95th percentile of an F-dist}
+\caption*{Figure | For each pair (\emph{n}=501) the number of tests where the test statistic is greater than the 95th percentile of an F-dist}
 \end{figure}
@@
 \includegraphics[height=3in]{type1.png}
-\caption*{Figure | Distribution of the empirical type 1 error rate. Determined as the proportion of tests with $F > 2.38$ (95th percentile of F-dist with df1=4, df2=842)}
+\caption*{Figure | Distribution of the empirical type 1 error rate. Determined as the proportion of tests with $F > 2.38$ (95th percentile of \emph{F}-dist with df1=4, df2=842)}
 \end{figure}

### 2015-04-15 investigation/scripts/epi_empirical_analysis.R

@@ -70,12 +70,23 @@ dev.off()
 
 
 
-# Make figures for the different 
-png(filename="~/repo/eQTL-2D/investigation/docs/figures/lambda.png")
-hist(as.numeric(as.matrix(gs$lambda)), breaks=25, 
-	col=as.numeric(as.matrix(gs$filter)),
-	xlab="lambda", main="")
+# Make figures for the different filters
+f1 <- which(gs$filter==1)
+f2 <- which(gs$filter==2)
+
+# F1
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/lambdaF1.png")
+hist(as.numeric(as.matrix(gs$lambda[f1])), breaks=25, 
+	xlab="lambda", main="Filter 1", col="lightgrey")
 dev.off()
+
+
+# F2
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/lambdaF2.png")
+hist(as.numeric(as.matrix(gs$lambda[f2])), breaks=25, 
+	xlab="lambda", main="Filter 2", col="lightgrey")
+dev.off()
+
 
 
 load(lf[3])

### 2015-04-15 investigation/scripts/epi_empirical_analysis_fun.R

@@ -163,7 +163,7 @@ genome_sum.fun <- function(lf) {
 filter_add.fun <- function(sig, gs) {
 
 
-	out <- matrix(0, nrow=nrow(gs), ncol=3)
+	out <- matrix(0, nrow=nrow(gs), ncol=4)
 	for(i in 1:nrow(gs)) {
 
 		index <- which(sig$probename==gs$probename[i] & sig$snp1==gs$snp1[i] & sig$snp2==gs$snp2[i])
@@ -176,11 +176,12 @@ filter_add.fun <- function(sig, gs) {
 		out[i,1] <- sig$filter[index]
 		out[i,2] <- sig$pnest_egcut[index]
 		out[i,3] <- sig$pnest_fehr[index]
+		out[i,4] <- sig$probegene[index]
 
 	}
 
 	out <- as.data.frame(out)
-	names(out) <- c("filter", "pnest_egcut", "pnest_fehr")
+	names(out) <- c("filter", "pnest_egcut", "pnest_fehr", "gene")
 	out <- cbind(gs, out)
 	return(out)
 
@@ -198,18 +199,18 @@ multi_lambda.fun <- function(gs, n) {
 
 	mp <- which(table(gs$probename)>n)
 
-	out <- matrix(0, nrow=length(mp), ncol=5)
+	out <- matrix(0, nrow=length(mp), ncol=6)
 	for(i in 1:length(mp)) {
 
 		foo <- gs[which(gs$probename==names(mp[i])),]
-		out[i,4] <- round(mean(as.numeric(as.matrix(foo$lambda))),2)
-		out[i,1:3] <- as.matrix(foo[1,1:3])
-		out[i,5] <- nrow(foo)
+		out[i,5] <- round(mean(as.numeric(as.matrix(foo$lambda))),2)
+		out[i,1:4] <- as.matrix(foo[1,c(1:3, 12)])
+		out[i,6] <- nrow(foo)
 
 	}
 
 	out <- as.data.frame(out)
-	names(out) <- c("probename", "snp1", "snp2", "meanlambda", "npairs")
+	names(out) <- c("probename", "snp1", "snp2", "gene", "meanlambda", "npairs")
 	return(out)
 
 }

### 2015-04-21 investigation/docs/overview.tex
edits to varioys code and report docs

@@
 \end{tabular}
-\caption*{Mean lambda per probe, when 5 or more interactions are identified amongst the 501 pairs}
+\caption*{This table shows the mean lambda per probe, when 5 or more interactions are identified amongst the 501 pairs}
 \end{table}
@@
 
-2.
-Identify the largest additive eQTL for the probe (irrespective of effect size).
-Regress out the effect of the additive loci and use the adjusted phenotype for 8df and 4df model analysis.
-Results reported are the 8df and 4df of original pair and the empirical p-values from genome-wide analysis fitting each of the two SNPs as fixed.
-As before these are determined by filtering out the SNPs on the same chromosome as the original fixed SNP and also within +/- 5MB of the second SNP.
-\\
+\begin{landscape}
 
-[Currently running]
+\begin{table}[ht]
+\centering
+\begin{tabular}{llllllllll}
+  \hline
+ probename & snp1 & snp2 & gene & chr1 & chr2 & lambda & P\_emp & P\_emp\_ferh & P\_emp\_egcut \\ 
+  \hline
+ILMN\_1652333 & rs898095 & rs9892064 & FN3KRP & 17 & 17 & 1.34 & 6.77 & 7.21 & 7.11 \\ 
+ILMN\_1710752 & rs2123758 & rs3889129 & NAPRT1 & 8 & 8 & 2.14 & 7.36 & 7.42 & 6.99 \\ 
+  ILMN\_1731043 & rs7776572 & rs11770192 & TRA2A & 7 & 7 & 1.36 & 6.84 & 4.19 & 3.67 \\ 
+  ILMN\_1761797 & rs9979356 & rs3761385 & CSTB & 21 & 21 & 1.05 & 7.51 & 6.88 & 7.24 \\ 
+  ILMN\_1769782 & rs1891432 & rs10900520 & LAX1 & 1 & 1 & 1.22 & 6.64 & 5.88 & 6.47 \\ 
+  ILMN\_1786426 & rs807491 & rs7254601 & TMEM149 & 19 & 19 & 2.69 & 7.4 & 6.73 & 6.57 \\ 
+  ILMN\_1795836 & rs9978658 & rs11701361 & C21ORF57 & 21 & 21 & 1.43 & 5.65 & 6.84 & 5.12 \\ 
+  ILMN\_1799381 & rs2634462 & rs6486334 & SNORD14A & 11 & 11 & 2.45 & 7.22 & 6.63 & 6.82 \\ 
+  ILMN\_2101920 & rs6894268 & rs4700810 & HNRPH1 & 5 & 5 & 1.47 & 7.14 & 6.59 & 7.43 \\ 
+  ILMN\_2121437 & rs7563453 & rs4973397 & NCL & 2 & 2 & 1.2 & 5.88 & 5.87 & 5.96 \\ 
+  ILMN\_2242463 & rs7930237 & rs556895 & CTSC & 11 & 11 & 1.12 & 5.59 & 7.21 & 6.34 \\ 
+  ILMN\_2313158 & rs16864367 & rs13079208 & MBNL1 & 3 & 3 & 2.28 & 7.22 & 6.84 & 6.35 \\ 
+  ILMN\_2358626 & rs2395095 & rs10824092 & ADK & 10 & 10 & 1.15 & 7.74 & 5.68 & 5.77 \\ 
+  ILMN\_2398939 & rs8092433 & rs4890876 & MBP & 18 & 18 & 1.19 & 5.9 & 6.59 & 6.11 \\ 
+  ILMN\_2410783 & rs11150847 & rs12602462 & GAA & 17 & 17 & 1.38 & 6.91 & 6.67 & 6.86 \\ 
+   \hline
+\end{tabular}
+\caption*{Summary information for the pairs of SNPs that had permutation empirical $p<4.48^{-6}$.
+P\_emp ferh and egcut are the permutation empirical p-values corresponding to the f-statistics from the original replication}
 
-\vspace{1cm}
+\end{table}
+\end{landscape}
 
-2. Prediction \\
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
 
+Suggestions following discussions \\
+
+1. Prediction \\
 Of the 501 SNP pairs, 484 have both SNP in the EGCUT data.
@@
 1df model (estimated in BSGS using the Inchianti snp) \\
+
+2. New permutation \\
+
+Explore a permutation analysis where $a+d$ for SNP1 and SNP2 are held constant. The suggestion was to permute the ''residual'' 4 df terms:
+
+The model is: \\
+
+$y_i=a1*xa1_i*xd1_i+a2*xa2_j+\left [ aa*xa1_i*x2a_i+...+dd*xd1_i*xd2_i \right ]+\epsilon $
+
+with xa and xd indicator variables (dummies).
+I think that the suggestion was to permute the entire term in square brackets (say, $R_i$) across individuals, holding $y$ and the first 4 terms (dummies for $a$ and $d$) constant.
+An alternative is to permute each of the 4 residual terms across $y$ and $a+d$ dummies.
 

### 2015-04-21 investigation/scripts/epi_empirical_analysis.R

@@ -161,17 +161,33 @@ l1 <- paste(pemp$probename, pemp$snp1, pemp$snp2, sep="_")
 l2 <- paste(gs$probename, gs$snp1, gs$snp2, sep="_")
 index <- which(l2 %in% l1)
 gs500 <- gs[index,]
+n2 <- paste(gs500$probename, gs500$snp1, gs500$snp2, sep="_")
 
 # P1
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/perm_vs_gwas.png")
 plot(as.numeric(as.matrix(pemp$neglog10pemp)), as.numeric(as.matrix(gs500$P_emp)))
-
+dev.off()
 
 # P2
 length(which(as.numeric(as.matrix(pemp$neglog10pemp)) > 5.34))
+index <- which(as.numeric(as.matrix(pemp$neglog10pemp)) > 5.34)
+tmp <- gs500[which(n2 %in% l1[index]),]
+length(which(tmp$chr1!=tmp$chr2))
+
 
 
 # P3
 pemp30 <- type1_30.fun(pemp, sig30)
+n1 <- paste(pemp30$probegene, pemp30$snp1, pemp30$snp2, sep="_")
+n2 <- paste(gs500$gene, gs500$snp1, gs500$snp2, sep="_")
+
+index <- which(as.numeric(as.matrix(pemp30$neglog10pemp)) > 5.34)
+n3 <- n1[index]
+index <- which(n2 %in% n3)
+tmp <- gs500[index,]
+xtable(tmp[,c(1:3,12:14,17,22)])
+
+
 length(which(as.numeric(as.matrix(pemp30$neglog10pemp)) > 5.34))
 
 #p4

### 2015-04-21 investigation/scripts/epi_empirical_analysis_fun.R

@@ -163,7 +163,7 @@ genome_sum.fun <- function(lf) {
 filter_add.fun <- function(sig, gs) {
 
 
-	out <- matrix(0, nrow=nrow(gs), ncol=4)
+	out <- matrix(0, nrow=nrow(gs), ncol=8)
 	for(i in 1:nrow(gs)) {
 
 		index <- which(sig$probename==gs$probename[i] & sig$snp1==gs$snp1[i] & sig$snp2==gs$snp2[i])
@@ -177,11 +177,15 @@ filter_add.fun <- function(sig, gs) {
 		out[i,2] <- sig$pnest_egcut[index]
 		out[i,3] <- sig$pnest_fehr[index]
 		out[i,4] <- sig$probegene[index]
+		out[i,5] <- sig$chr1[index]
+		out[i,6] <- sig$chr2[index]
+		out[i,7] <- sig$pos1[index]
+		out[i,8] <- sig$pos2[index]
 
 	}
 
 	out <- as.data.frame(out)
-	names(out) <- c("filter", "pnest_egcut", "pnest_fehr", "gene")
+	names(out) <- c("filter", "pnest_egcut", "pnest_fehr", "gene", "chr1", "chr2", "pos1", "pos2")
 	out <- cbind(gs, out)
 	return(out)
 

### 2015-04-21 investigation/docs/overview.tex
edits to varioys code and report docs

@@
 
-b. Genotypes at the corresponding epistatitic ''co-SNP'' were randomly shuffled (no replacement) amongst individuals. \\
+b. Genotypes at the corresponding epistatic ''co-SNP'' were randomly shuffled (no replacement) amongst individuals. \\
 
@@
 There are 96 pairs where the corresponding F statistic is less than 7.67}.
-None of these are amongst the 30 bonferroni correction significant pairs.
+None of these are amongst the 30 Bonferroni correction significant pairs.
 \end{figure}
@@
 \centering
-\begin{tabular}{llllll}
+\begin{tabular}{llllllll}
   \hline
-gene & probename & snp1 & snp2 & $\lambda$ & Type1 \\ 
+ gene & probename & snp1 & snp2 & filter & gene.1 & nthreshold & P\_emp \\ 
   \hline
-ADK & ILMN\_2358626 & rs2395095 & rs10824092 & 1.15 & 0.06 \\ 
-ATP13A1 & ILMN\_2134224 & rs4284750 & rs873870 & 1.82 & 0.119 \\ 
-C21ORF57 & ILMN\_1795836 & rs9978658 & rs11701361 & 1.43 & 0.085 \\ 
-CSTB & ILMN\_1761797 & rs9979356 & rs3761385 & 1.05 & 0.053 \\ 
-CTSC & ILMN\_2242463 & rs7930237 & rs556895 & 1.12 & 0.054 \\ 
-FN3KRP & ILMN\_1652333 & rs898095 & rs9892064 & 1.34 & 0.075 \\ 
-GAA & ILMN\_2410783 & rs11150847 & rs12602462 & 1.38 & 0.081 \\ 
-HNRPH1 & ILMN\_2101920 & rs6894268 & rs4700810 & 1.47 & 0.091 \\ 
-LAX1 & ILMN\_1769782 & rs1891432 & rs10900520 & 1.22 & 0.066 \\ 
-MBLN1 & ILMN\_2313158 & rs11981513 & rs13069559 & 3.36 & 0.242 \\ 
-MBLN1 & ILMN\_2313158 & rs16864367 & rs13079208 & 2.28 & 0.161 \\ 
-MBLN1 & ILMN\_2313158 & rs2030926 & rs13069559 & 3.39 & 0.241 \\ 
-MBLN1 & ILMN\_2313158 & rs218671 & rs13069559 & 3.36 & 0.242 \\ 
-MBLN1 & ILMN\_2313158 & rs2614467 & rs13069559 & 3.37 & 0.242 \\ 
-MBLN1 & ILMN\_2313158 & rs7710738 & rs13069559 & 3.38 & 0.243 \\ 
-MBP & ILMN\_2398939 & rs8092433 & rs4890876 & 1.19 & 0.066 \\ 
-NAPRT1 & ILMN\_1710752 & rs2123758 & rs3889129 & 2.14 & 0.155 \\ 
-NCL & ILMN\_2121437 & rs7563453 & rs4973397 & 1.2 & 0.065 \\ 
-PRMT2 & ILMN\_1675038 & rs2839372 & rs11701058 & 2.77 & 0.199 \\ 
-SNORD14A & ILMN\_1799381 & rs2634462 & rs6486334 & 2.45 & 0.168 \\ 
-TMEM149 & ILMN\_1786426 & rs807491 & rs7254601 & 2.69 & 0.194 \\ 
-TMEM149 & ILMN\_1786426 & rs8106959 & rs1843357 & 5.79 & 0.396 \\ 
-TMEM149 & ILMN\_1786426 & rs8106959 & rs2351458 & 5.77 & 0.395 \\ 
-TMEM149 & ILMN\_1786426 & rs8106959 & rs6718480 & 5.82 & 0.397 \\ 
-TMEM149 & ILMN\_1786426 & rs8106959 & rs6926382 & 5.75 & 0.396 \\ 
-TMEM149 & ILMN\_1786426 & rs8106959 & rs914940 & 5.76 & 0.395 \\ 
-TMEM149 & ILMN\_1786426 & rs8106959 & rs9509428 & 5.82 & 0.398 \\ 
-VASP & ILMN\_1743646 & rs1264226 & rs2276470 & 1.37 & 0.08 \\ 
-RPL13 & ILMN\_2413278 & rs352935 & rs2965817 & 1.11 & 0.058 \\ 
-TRA2A & ILMN\_1731043 & rs7776572 & rs11770192 & 1.36 & 0.079 \\ 
+ADK & ILMN\_2358626 & rs2395095 & rs10824092 & 1 & ADK & 3 & 7.74 \\ 
+ATP13A1 & ILMN\_2134224 & rs4284750 & rs873870 & 2 & ATP13A1 & 23 & 7.08 \\ 
+C21ORF57 & ILMN\_1795836 & rs9978658 & rs11701361 & 2 & C21ORF57 & 6 & 5.65 \\ 
+CSTB & ILMN\_1761797 & rs9979356 & rs3761385 & 2 & CSTB & 4 & 7.51 \\ 
+CTSC & ILMN\_2242463 & rs7930237 & rs556895 & 1 & CTSC & 1 & 5.59 \\ 
+FN3KRP & ILMN\_1652333 & rs898095 & rs9892064 & 2 & FN3KRP & 7 & 6.77 \\ 
+GAA & ILMN\_2410783 & rs11150847 & rs12602462 & 2 & GAA & 9 & 6.91 \\ 
+HNRPH1 & ILMN\_2101920 & rs6894268 & rs4700810 & 2 & HNRPH1 & 13 & 7.14 \\ 
+LAX1 & ILMN\_1769782 & rs1891432 & rs10900520 & 2 & LAX1 & 5 & 6.64 \\ 
+MBLN1 & ILMN\_2313158 & rs11981513 & rs13069559 & 1 & MBNL1 & 86 & 9.06 \\ 
+MBLN1 & ILMN\_2313158 & rs16864367 & rs13079208 & 2 & MBNL1 & 79 & 7.22 \\ 
+MBLN1 & ILMN\_2313158 & rs2030926 & rs13069559 & 1 & MBNL1 & 91 & 9.06 \\ 
+MBLN1 & ILMN\_2313158 & rs218671 & rs13069559 & 1 & MBNL1 & 98 & 9.06 \\ 
+MBLN1 & ILMN\_2313158 & rs2614467 & rs13069559 & 1 & MBNL1 & 95 & 9.06 \\ 
+MBLN1 & ILMN\_2313158 & rs7710738 & rs13069559 & 1 & MBNL1 & 89 & 9.06 \\ 
+MBP & ILMN\_2398939 & rs8092433 & rs4890876 & 2 & MBP & 3 & 5.9 \\ 
+NAPRT1 & ILMN\_1710752 & rs2123758 & rs3889129 & 1 & NAPRT1 & 80 & 7.36 \\ 
+NCL & ILMN\_2121437 & rs7563453 & rs4973397 & 2 & NCL & 4 & 5.88 \\ 
+PRMT2 & ILMN\_1675038 & rs2839372 & rs11701058 & 2 & PRMT2 & 39 & 7.07 \\ 
+SNORD14A & ILMN\_1799381 & rs2634462 & rs6486334 & 2 & SNORD14A & 63 & 7.22 \\ 
+TMEM149 & ILMN\_1786426 & rs807491 & rs7254601 & 1 & TMEM149 & 74 & 7.4 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs1843357 & 1 & TMEM149 & 148 & 9.71 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs2351458 & 1 & TMEM149 & 154 & 9.71 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs6718480 & 1 & TMEM149 & 146 & 9.71 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs6926382 & 1 & TMEM149 & 157 & 9.71 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs914940 & 1 & TMEM149 & 156 & 9.71 \\ 
+TMEM149 & ILMN\_1786426 & rs8106959 & rs9509428 & 1 & TMEM149 & 160 & 9.71 \\ 
+VASP & ILMN\_1743646 & rs1264226 & rs2276470 & 2 & VASP & 10 & 6.81 \\ 
+RPL13 & ILMN\_2413278 & rs352935 & rs2965817 & 2 & RPL13 & 2 & 6.57 \\ 
+TRA2A & ILMN\_1731043 & rs7776572 & rs11770192 & 2 & TRA2A & 7 & 6.84 \\ 
    \hline
 \end{tabular}
-\caption*{The $\lambda$ and type 1 error rates for the 30 (Bonferroni-correction) significant replicated pairs mentioned in Hemani \emph{et al}.}
+\caption*{The $\lambda$ and type 1 error rates for the 30 (Bonferroni-correction) significant replicated pairs mentioned in Hemani \emph{et al}. P\_emp demotes the permutation empirical p-value}
 \end{table}
+
+\newpage
+\begin{figure}[H]
+\centering
+\includegraphics[height=3in]{pemp_hist.png}
+\caption*{Figure | Distribution of the permutation empirical p-values}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[height=3in]{perm_vs_gwas.png}
+\caption*{Figure | the relationship between the empirical p-values from the permutation and GWAS-based methods.
+It is unclear why there is this relationship.
+I've checked the results and code several times and believe this is correct.}
+\end{figure}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+
+Response to specific email questions \\
+
+Q1. How many pairs from the permutation analysis passed the Bonferoni threshold?
+
+A1. Bonferoni threshold is $= 4.48^{e10-6}$. There are 305 / 501 pairs with empirical p-values from the permutation that are smaller
+
+\vspace{1cm}
+
+Q2. How many pairs that ended up in Table 1 of Hemani passed the Bonferoni threshold?
+
+A2. 15
+
+\vspace{1cm}
+
+Q3. How many of the 95 pairs (filter 2) are above the B threshold and how many were in Table 1 of Hemani?
+
+A3. 1
+
+\vspace{1cm}
+
+Q4.
+My guess (from your previous results) is that all cis-trans pairs are among the 30 that didn't pass the threshold - is that correct?
+Are there any cis-trans pairs among the 305 ones that survived the Bonferoni threshold with the new empirical p-value?
+
+A4. Yes, that's correct. It's just cis-cis ones that pass the threshold. Although, of the 305 that do, 273 are cis-trans.
 

### 2015-04-21 investigation/scripts/epi_empirical_analysis.R

@@ -133,7 +133,7 @@ dev.off()
 # 2. Given results from 1.
 # a) display the empirical type-I error rate for the probe-SNP-SNP trios that are in the Nature article table (the top 30).
 gs30 <- type1_30.fun(gs, sig30)
-xtable(gs30[,c(1:4,13,19)])
+xtable(gs30[,c(1:4,10,13,19, 23)])
 
 
 # 3. For each probe-SNP pair, list 
@@ -151,6 +151,13 @@ xtable(multi_lambda)
 
 # 5. For those probe-SNP-SNP trios that pass the empirical threshold (100 out of 400?), what proportion are cis and trans? I'm guessing that most will be cis.
 
+##################################################################
+##################################################################
+##################################################################
+# make histogram of the empirical p-values from permuation
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/pemp_hist.png")
+hist(-log10(pemp$pemp), breaks=25, xlab="empirical (10,000,000) permutation p-value NegLog scale", main="", col="lightgray")
+dev.off()
 
 ##################################################################
 ##################################################################
@@ -165,7 +172,8 @@ n2 <- paste(gs500$probename, gs500$snp1, gs500$snp2, sep="_")
 
 # P1
 png(filename="~/repo/eQTL-2D/investigation/docs/figures/perm_vs_gwas.png")
-plot(as.numeric(as.matrix(pemp$neglog10pemp)), as.numeric(as.matrix(gs500$P_emp)))
+plot(as.numeric(as.matrix(pemp$neglog10pemp)), as.numeric(as.matrix(gs500$P_emp)),
+	xlab="permutation", ylab="GWAS", main="")
 dev.off()
 
 # P2

### 2015-04-28 investigation/docs/overview.tex
edits to the .tex file. needs re-comiling on desktop
data sig files added , report updated (again)

@@
 \usepackage{parskip}
+\usepackage{longtable}
 
@@
 \end{tabular}
-\caption*{The $\lambda$ and type 1 error rates for the 30 (Bonferroni-correction) significant replicated pairs mentioned in Hemani \emph{et al}. P\_emp demotes the permutation empirical p-value}
+\caption*{The $\lambda$ and type 1 error rates for the 30 (Bonferroni-correction) significant replicated pairs mentioned in Hemani \emph{et al}.
+P\_emp demotes the p-value from the observed F-statistic (df1=4, df2=846) which corresponds to $n*4.44e^{-6}$ ranked test statistic from the ‘gwas’ analysis.
+If $n*4.4e^{-6} < 1$ then the top ranked F-stat was used.
+}
 \end{table}
@@
 
+\newpage
+
 \begin{figure}[H]
 \centering
-\includegraphics[height=3in]{perm_vs_gwas.png}
-\caption*{Figure | the relationship between the empirical p-values from the permutation and GWAS-based methods.
-It is unclear why there is this relationship.
-I've checked the results and code several times and believe this is correct.}
+\includegraphics[height=3in]{perm_vs_lambdaF.png}
+\caption*{Figure | the relationship between the observed p values corrected for lambdaF, which was calculated from the GWAS-based analysis and the permutation empitical p-values}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[height=3in]{perm_vs_lambdaC.png}
+\caption*{Figure | the relationship between the observed p values corrected for lambdaChisq, which was calculated from the GWAS-based analysis and the permutation empitical p-values}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[height=3in]{lambda.png}
+\caption*{Figure | the relationship between the lambdaChisq and lambdaF from the GWAS-based analysis}
+\end{figure}
+
+\begin{landscape}
+
+\begin{figure}[H]
+\centering
+\includegraphics[height=4in]{TreeOfLife.png}
+\caption*{Figure | This figure denotes the numbers of Probe by SNP pairs that pass verious thresholds and split on original filtering and relative positional information.
+All pairs that pass both the permutation empirical p value threshold of $4.48e-6$ and the $\lambda_F$ corrected threshold of $4.48e-6$ are summerized in the table \emph{datasig.csv} and \emph{datasig\_lite.csv}.
+The total number of paris in this table is 155}
 \end{figure}
@@
 \newpage
-
-\begin{landscape}
 

### 2015-04-28 investigation/scripts/epi_empirical_analysis.R

@@ -292,7 +292,9 @@ length(which(as.numeric(as.matrix(dataf2ct$PlamC)) < 4.48e-6))
 
 datasig <- TableOfTruth.fun(data, sig30)
 write.csv(datasig, "~/repo/eQTL-2D/investigation/data/datasig.csv", quote=F, row.names=F)
+write.csv(datasig[,c(1:3,9,12,30,35:36, 39:41)], "~/repo/eQTL-2D/investigation/data/datasig_lite.csv", quote=F, row.names=F)
 
+xtable(datasig[,c(1:3,9,12,30,35:36, 39:41)])
 
 
 

### 2015-04-30 investigation/docs/overview.tex
col figures added and code edited to make them

@@
 \includegraphics[height=3in]{perm_vs_lambdaF.png}
-\caption*{Figure | the relationship between the observed p values corrected for lambdaF, which was calculated from the GWAS-based analysis and the permutation empitical p-values}
+\caption*{Figure | the relationship between the observed p values corrected for lambdaF, which was calculated from the GWAS-based analysis and the permutation empitical p-values.
+Red are \emph{cis-cis}}
 \end{figure}
@@
 \includegraphics[height=3in]{perm_vs_lambdaC.png}
-\caption*{Figure | the relationship between the observed p values corrected for lambdaChisq, which was calculated from the GWAS-based analysis and the permutation empitical p-values}
+\caption*{Figure | the relationship between the observed p values corrected for lambdaChisq, which was calculated from the GWAS-based analysis and the permutation empitical p-values.
+Red are \emph{cis-cis}}
 \end{figure}

### 2015-04-30 investigation/scripts/epi_empirical_analysis.R

@@ -294,8 +294,21 @@ datasig <- TableOfTruth.fun(data, sig30)
 write.csv(datasig, "~/repo/eQTL-2D/investigation/data/datasig.csv", quote=F, row.names=F)
 write.csv(datasig[,c(1:3,9,12,30,35:36, 39:41)], "~/repo/eQTL-2D/investigation/data/datasig_lite.csv", quote=F, row.names=F)
 
-xtable(datasig[,c(1:3,9,12,30,35:36, 39:41)])
 
+
+
+# Lambda Chi-sq
+col_index <- as.numeric(data$chr1==data$chr2)+1
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/perm_vs_lambdaC.png")
+plot(as.numeric(as.matrix(data$neglog10pemp)), -log10(as.numeric(as.matrix(data$PlamC))),
+	xlab="permutation", ylab="GWAS - lambda Chisq", main="", pch=16, col=col_index)
+dev.off()
+
+# Lambda F4
+png(filename="~/repo/eQTL-2D/investigation/docs/figures/perm_vs_lambdaF.png")
+plot(as.numeric(as.matrix(data$neglog10pemp)), -log10(as.numeric(as.matrix(data$PlamF))),
+	xlab="permutation", ylab="GWAS - lambda F (4df)", main="", pch=16, col=col_index)
+dev.off()
 
 
 

### 2019-09-22 null_replication/scripts/analyse_data.sh
added polygenic

@@ -43,10 +43,24 @@ do
 	varexp=`Rscript -e "cat(runif(1, $minvar, $maxvar))"`
 	echo $varexp
 
-	# Create phenotypes using sentinel SNP
+	# Create phenotypes using sentinel SNP and polygenic score
 
-	Rscript makephen.r ${sd}/${sensnp}_rep.raw ${sd}/${i}_${sim}_rep.fam ${varexp}
-	Rscript makephen.r ${sd}/${sensnp}_disc.raw ${sd}/${i}_${sim}_disc.fam ${varexp}
+	# Create polygenic score
+	awk '{print $1,$2,rand()}' ${sd}/polygenic.txt > ${sd}/${i}_${sim}_score.txt
+	plink --bfile ${sd}/polygenic --keep ${disc}list.txt --score ${sd}/${i}_${sim}_score.txt --out ${sd}/${i}_${sim}_disc_score.txt
+	plink --bfile ${sd}/polygenic --keep ${rep}list.txt --score ${sd}/${i}_${sim}_score.txt --out ${sd}/${i}_${sim}_rep_score.txt
+
+
+	if (( RANDOM % 2 ));
+	then
+		polyvar=0.3
+	else
+		polyvar=0
+	fi
+	echo "$polyvar"
+
+	Rscript makephen.r ${sd}/${sensnp}_rep.raw ${sd}/${i}_${sim}_rep.fam ${varexp} ${sd}/${i}_${sim}_rep_score.txt.profile ${polyvar}
+	Rscript makephen.r ${sd}/${sensnp}_disc.raw ${sd}/${i}_${sim}_disc.fam ${varexp} ${sd}/${i}_${sim}_disc_score.txt.profile ${polyvar}
 
 
 	# Perform scan
@@ -59,7 +73,9 @@ do
 	cat ${sd}/${i}_${sim}_rep_out[0-9]* > ${sd}/${i}_${sim}_rep_out
 	rm ${sd}/${i}_${sim}_rep_out[0-9]*
 
+	echo "analysing"
 	Rscript formatres.r ${sd}/${i}_${sim}_disc_out ${sd}/${i}_${sim}_rep_out ${varexp} ${i} ${sim} ${cissnp} ${sensnp} ${sd}/${i}_${sim}.rdata ${sd}/${cissnp}_exclude_disc.bim.orig
+	echo "done analysing"
 
 	rm ${sd}/${i}_${sim}_*
 

### 2019-09-22 null_replication/scripts/create_data.sh

@@ -59,10 +59,13 @@ plink --bfile ${sd}/${sensnp}_disc --recode A --out ${sd}/${sensnp}_disc
 plink --bfile ${sd}/${sensnp}_rep --recode A --out ${sd}/${sensnp}_rep
 
 
+# Polygenic snps
 
+shuf ${datafile}.bim | head -n 1000 | awk '{print $2,$5,rand()}' > ${sd}/polygenic.txt
+cat ${rep}list.txt ${disc}list.txt > temp.txt
 
-
-
+plink --bfile ${datafile} --keep temp.txt --extract ${sd}/polygenic.txt --make-bed --out ${sd}/polygenic
+rm temp.txt
 
 
 

### 2019-09-22 null_replication/scripts/makephen.r

@@ -7,6 +7,8 @@ args <- commandArgs(T)
 rawfile <- args[1]
 famfile <- args[2]
 varexp <- as.numeric(args[3])
+polyfile <- args[4]
+polyvar <- as.numeric(args[5])
 
 print(rawfile)
 print(famfile)
@@ -15,8 +17,16 @@ print(varexp)
 gen <- fread(rawfile) %>% as_tibble()
 names(gen)[ncol(gen)] <- "geno"
 
-gen$PHENOTYPE <- make_phen(sqrt(varexp), gen$geno)
+score <- fread(polyfile) %>% as_tibble()
+
+stopifnot(all(score$IID == gen$IID))
+
+gen$PHENOTYPE <- make_phen(c(sqrt(varexp), sqrt(polyvar)), cbind(gen$geno, score$SCORE))
 cor(gen$PHENOTYPE, gen$geno, use="pair")^2
+cor(gen$PHENOTYPE, score$SCORE, use="pair")^2
+
 gen <- subset(gen, select=-c(geno))
 write.table(gen, file=famfile, row=F, col=F, qu=F, sep=" ")
+
+
 

### 2019-09-22 null_replication/scripts/analyse_results.rmd
updated

---
title: Analysis of epistatic inflation under large single additive effects, with a focus on the replication rate
author: Gibran Hemani
---

## Background

A 4 d.f. test for epistatic interactions is intended to be orthogonal from additive effects. However we are now seeing that a variant which is in LD with a large causal variant will be susceptible to large 4df test statistics across the genome. This can lead to a high false discovery rate at stringent significance thresholds. If the causal variant is available with low measurement error, fitting it as a covariate is sufficient to eliminate these false positives.

In these simulations we are looking at two things:

1. The influence of the magnitude of the additive effect on the false discovery rate
2. The replication rate, in an independent sample, of false positives from the discovery analysis.

## Simulation scenario

Here we attempt to mimic the scenario in Hemani et al 2014. There were 846 samples in the discovery and a combined 2131 in the replication. We reported the MBNL1 gene being influenced by several cis-trans epistatic interactions, where the cis variant was rs13069559. These associations replicated at the Bonferroni level in the independent replication sample. Wood et al 2014 showed that fitting the fine-mapped additive cis-variant rs67903230 abrogated the cis-trans signals involving rs13069559.

So what we did here was:

1. Use ALSPAC imputed genotype data to create a discovery data set and replication dataset of the same sizes as used in the original analysis. 
2. Simulate a phenotype where the fine-mapped variant, rs67903230, had a large additive effect.
3. Performed a 4df interaction test of the original cis variant, rs13069559, against 502510 genotyped markers in the discovery
4. Tested any variants at Bonferroni significance for replication in the independent dataset

As in the original paper, only interactions between SNP pairs that had all 9 genotype classes were retained, and only one trans variant per chromosome was retained for replication. All SNPs from the cis chromosome except the cis-proxy SNP were excluded, so we're only looking for cis-trans effects here.

```{r}
library(tidyverse)
load("../data/rs67903230.rdata")
```

We performed `r nrow(res)` simulations, where the variance explained by the causal variant was sampled uniformly between 0% and 50%

```{r}
hist(res$varexp)
```

The lambdaGC value for the genome wide interaction effects were recorded for each simulation from the discovery and replication datasets

```{r}
plot(lambda_disc ~ lambda_repl, res)
```

And the lambdaGC value was strongly related to the variance explained by the causal variant

```{r}
plot(lambda_disc ~ varexp, res)
```

The number of significant hits at a Bonferroni threshold (p < 0.05 / 500k):

```{r}
hist(res$nsig_disc_bonf)
```

And the number significant at the threshold used in the paper pf p < 1e-16:

```{r}
hist(res$nsig_disc_bonf2, xlim=c(0,max(res$nsig_disc_bonf)))
```

Note that we found no significant hits at an experiment wide threshold, this is following `r nrow(res) * 500000` multiple tests. In the original paper we found 501 hits after performing over `r 1e15` tests, however in these simulations we have looked specifically at interactions between only cis-proxy variants which are in principle vulnerable to cis-trans false positives. The number of tests that we did using cis-proxy variants is likely similar in volume to the number performed empirically (as a small fraction of all SNPs are cis-proxy variants).

The number of independent discovery associations per simulation was higher with larger causal variant effects (p = `r summary(lm(nsig_disc_bonf ~ varexp, res))$coefficients[2,4]`)

```{r}
plot(nsig_disc_bonf ~ varexp, res)
```

The average correlation of F statistics from the discovery and replication analyses (without thresholding) was `r mean(res$fstat_cor, na.rm=TRUE)`

```{r}
hist(res$fstat_cor)
```

The larger the causal variant effect, the larger the F statistic correlation

```{r}
plot(fstat_cor ~ varexp, res)
```

## Is the replication rate high under the null simulations?

If the false positive rate is such that we obtain at least one false positive cis-trans association in the discovery, what is the chance of it replicating in an independent dataset?

We can attempt to replicate at a Bonferroni threshold (within one simulation, how many associations were attempted to replicate at corrected p < 0.05) or FDR (within one simulation, how many associations repliace at FDR < 0.05).

Note that these are much more relaxed thresholds than what we used in the original paper - which was 0.05 / 501.

```{r}

res$lambda_disc_bin <- cut(res$lambda_disc, 5)
temp <- res %>%
	group_by(lambda_disc_bin, nsig_disc_bonf) %>%
	summarise(count=n(), rate=n()/sum(res$lambda_disc_bin == .$lambda_disc_bin[1]))

ggplot(temp %>% filter(nsig_disc_bonf > 0), aes(x=lambda_disc_bin, y=count)) +
geom_bar(stat="identity", aes(fill=nsig_disc_bonf))

table(res$lambda_disc_bin)
hist(res$lambda_disc)

plot(cond_bonf_bonf2 ~ varexp, res)

plot(I(cond_bonf_bonf2/nsig_disc_bonf) ~ varexp, res)

mean(res$cond_bonf_bonf / res$nsig_disc_bonf, na.rm=T)
mean(res$cond_bonf_bonf2 / res$nsig_disc_bonf, na.rm=T)

res$bonf_bonf_rr <- res$cond_bonf_bonf / res$nsig_disc_bonf
res$bonf_fdr_rr <- res$cond_bonf_fdr / res$nsig_disc_bonf
res$bonf_bonf2_rr <- res$cond_bonf_bonf2 / res$nsig_disc_bonf

mean(res$bonf_bonf_rr, na.rm=TRUE)
mean(res$bonf_bonf2_rr, na.rm=TRUE)
tapply(res$bonf_bonf_rr, res$lambda_disc_bin, function(x) mean(x, na.rm=T))
tapply(res$bonf_fdr_rr, res$lambda_disc_bin, function(x) mean(x, na.rm=T))

tapply(res$bonf_bonf_rr, res$nsig_disc_bonf, function(x) mean(x, na.rm=T))
tapply(res$bonf_fdr_rr, res$nsig_disc_bonf, function(x) mean(x, na.rm=T))

```

### 2019-09-30 null_replication/scripts/analyse_results.rmd
updated analysis of null simulations

@@
 ---
+
+## Summary
+
+These simulations suggest
+
+- that discovering associations at experiment wide thresholds (e.g. 1e-16) is unlikely even when the lambda is very high
+- the replication rate could be quite high using FDR
+- the replication rate is very low at experiment wide thresholds
+- we might get higher experiment wide replication rates if we found a way that the discovery effects surpassed the experiment wide threshold
 
@@
 
+We also added an extra variable to test.
+Most expression traits have a large cis effect and a complex, polygenic trans component.
+For half of the simulations we also created a polygenic component that explained 30% of the variance, which was composed of 1000 SNPs which each had an effect sampled from a normal distribution.
+
 ```{r}
@@
 load("../data/rs67903230.rdata")
+res$polygenic <- res$i %% 2 == 0
 ```
@@
 ```{r}
-plot(lambda_disc ~ varexp, res)
+ggplot(res, aes(x=varexp, y=lambda_disc)) +
+geom_point(aes(colour=polygenic)) +
+geom_smooth(aes(colour=polygenic))
 ```
@@
 ```{r}
-hist(res$nsig_disc_bonf)
+ggplot(res, aes(x=nsig_disc_bonf)) +
+geom_histogram(aes(fill=polygenic), position="dodge")
 ```
@@
 
+There looks to be no affect of having a polygenic trans component, so from now on we will not stratify the results on that basis.
+
 The number of independent discovery associations per simulation was higher with larger causal variant effects (p = `r summary(lm(nsig_disc_bonf ~ varexp, res))$coefficients[2,4]`)
@@
 ```{r}
-plot(nsig_disc_bonf ~ varexp, res)
+ggplot(res, aes(x=varexp, y=nsig_disc_bonf)) +
+geom_point()
 ```
@@
 
+Here's a look at the average number of discovery hits per simulation, binned by levels of genome wide lambda
+
+```{r}
+res$lambda_disc_bin <- Hmisc::cut2(res$lambda_disc, g=30)
+group_by(res, lambda_disc_bin) %>% summarise(n=mean(nsig_disc_bonf, na.rm=T)) %>% 
+ggplot(aes(x=lambda_disc_bin, y=n)) +
+geom_point() +
+theme(axis.text.x=element_text(angle=90, hjust=0.5, vjust=0.5)) +
+labs(x="Simulation's lambda", y="Mean number of discovery hits")
+```
+
 ## Is the replication rate high under the null simulations?
@@
 
-We can attempt to replicate at a Bonferroni threshold (within one simulation, how many associations were attempted to replicate at corrected p < 0.05) or FDR (within one simulation, how many associations repliace at FDR < 0.05).
+We can attempt to replicate at
 
-Note that these are much more relaxed thresholds than what we used in the original paper - which was 0.05 / 501.
+* Bonferroni threshold (within one simulation, how many discovery associations were replicated at Bonferroni corrected p < 0.05)
+* FDR (within one simulation, how many discovery associations replicate at FDR < 0.05).
+* experiment threshold (within one simulation, how many discovery associations replicated at p < 0.05/501, the threshold used in the paper)
 
 ```{r}
-
-res$lambda_disc_bin <- cut(res$lambda_disc, 5)
-temp <- res %>%
-	group_by(lambda_disc_bin, nsig_disc_bonf) %>%
-	summarise(count=n(), rate=n()/sum(res$lambda_disc_bin == .$lambda_disc_bin[1]))
-
-ggplot(temp %>% filter(nsig_disc_bonf > 0), aes(x=lambda_disc_bin, y=count)) +
-geom_bar(stat="identity", aes(fill=nsig_disc_bonf))
-
-table(res$lambda_disc_bin)
-hist(res$lambda_disc)
-
-plot(cond_bonf_bonf2 ~ varexp, res)
-
-plot(I(cond_bonf_bonf2/nsig_disc_bonf) ~ varexp, res)
-
-mean(res$cond_bonf_bonf / res$nsig_disc_bonf, na.rm=T)
-mean(res$cond_bonf_bonf2 / res$nsig_disc_bonf, na.rm=T)
-
 res$bonf_bonf_rr <- res$cond_bonf_bonf / res$nsig_disc_bonf
@@
 
-mean(res$bonf_bonf_rr, na.rm=TRUE)
-mean(res$bonf_bonf2_rr, na.rm=TRUE)
-tapply(res$bonf_bonf_rr, res$lambda_disc_bin, function(x) mean(x, na.rm=T))
-tapply(res$bonf_fdr_rr, res$lambda_disc_bin, function(x) mean(x, na.rm=T))
+repl_rate <- bind_rows(
+	group_by(res, lambda_disc_bin) %>% summarise(m=mean(bonf_bonf_rr, na.rm=T)) %>% mutate(threshold="Bonferroni"),
+	group_by(res, lambda_disc_bin) %>% summarise(m=mean(bonf_fdr_rr, na.rm=T)) %>% mutate(threshold="FDR"),
+	group_by(res, lambda_disc_bin) %>% summarise(m=mean(bonf_bonf2_rr, na.rm=T)) %>% mutate(threshold="Experiment")
+)
+repl_rate %>% ggplot(aes(x=lambda_disc_bin, y=m, group=threshold)) +
+geom_line(aes(colour=threshold)) +
+scale_colour_brewer(type="qual") +
+theme(axis.text.x=element_text(angle=90, hjust=0.5, vjust=0.5)) +
+labs(x="Simulation's lambda", y="Replication rate")
+```
 
-tapply(res$bonf_bonf_rr, res$nsig_disc_bonf, function(x) mean(x, na.rm=T))
-tapply(res$bonf_fdr_rr, res$nsig_disc_bonf, function(x) mean(x, na.rm=T))
-
-```
+Overall, using FDR there is a reasonably high replication rate which doesn't correspond to discovery lambda.
+But the experiment wide threshold can only be met when the discovery lambda is very high, but even then the replication rate is extremely low.

### 2019-09-30 null_replication/docs/analyse_results.rmd
reorganised

--- 2019-09-30 null_replication/scripts/analyse_results.rmd
@@
 - that discovering associations at experiment wide thresholds (e.g. 1e-16) is unlikely even when the lambda is very high
-- the replication rate could be quite high using FDR
+- the replication rate could be quite high using FDR, even if the lambda is very low
 - the replication rate is very low at experiment wide thresholds

### 2019-09-30 null_replication/docs/quick.rmd
reorganised

---
title: All probe lambda values
---

Read in data and get FDR values for replication

```{r}
a <- read.csv("total_analysis_data.csv")
a$pnest_egcut <- 10^-a$pnest_egcut
a$pnest_fehr <- 10^-a$pnest_fehr
a$fdr_egcut <- p.adjust(a$pnest_egcut, "fdr")
a$fdr_fehr <- p.adjust(a$pnest_fehr, "fdr")
a$pnest_both <- NA
for(i in 1:nrow(a))
{
	a$pnest_both[i] <- pchisq(-2 * sum(log(c(a$pnest_fehr[i], a$pnest_egcut[i]))), df=4, low=FALSE)
}
a$fdr_both <- p.adjust(a$pnest_both, "fdr")
```

Remove cis-cis effects, how many remain?

```{r}
a <- subset(a, chr1 != chr2)
nrow(a)
```

How many replicate in both at FDR < 0.05

```{r}
b1 <- subset(a, (fdr_both < 0.05))
nrow(b1)
```

What are the lambda values of these probes

```{r}
plot(b1$lambda)
```

How many replicate in both at p < 0.05

```{r}
b2 <- subset(a, pnest_both < 0.05)
nrow(b2)
```

What are the lambda values of these probes

```{r}
plot(b2$lambda)
```

Do again but with no duplicate probes

```{r}
plot(subset(b2, !duplicated(gene))$lambda)
```

It looks like at replication significance threshold of FDR < 0.05 there is one cis-trans interaction that replicates that has lambda value close to 1, and at nominal p < 0.05 threshold there are 15.

Questions

1. What is the standard error around these lambda values?

### 2019-12-26 ftest_paper/docs/theory.rmd
updated

---
title: F test theory
---

```{r}
suppressWarnings(suppressPackageStartupMessages({
	library(dplyr)
	library(knitr)
	library(simulateGP)
}))
knitr::opts_chunk$set(warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, error=FALSE)
```

Simulate a simple 3 locus haplotype system with two loci in LD of 0.5, and each with MAF = 0.1

$$
\begin{aligned}
y_1 &\sim y_2 + y_3 + e_1 \\
y_1 &\sim y_2 + y_3 + y_{2}y_{3} + e_2
\end{aligned}
$$

Method to create y1, y2, y3:

```{r}
gendat <- function(n)
{
	y1 <- rep(0, n)
	y2 <- rep(0, n)
	y1[856:900] <- 1
	y2[901:945] <- 1
	y1[946:1000] <- 1
	y2[946:1000] <- 1
	y3 <- rbinom(n, 1, 0.5)
	return(tibble(y1,y2,y3))
}
```

Method to run the simulation:

```{r}
run_sim <- function(i, n=1000)
{
	set.seed(i)
	dat <- gendat(n)
	mod1 <- lm(y1 ~ y2 + y3, dat)
	mod2 <- lm(y1 ~ y2 + y3 + y2*y3, dat)
	amod <- anova(mod1, mod2)
	vare2 <- var(residuals(mod2))
	sse1 <- sum(residuals(mod1)^2)
	sse2 <- sum(residuals(mod2)^2)
	return(list(amod=amod, vare2=vare2, sse1=sse1, sse2=sse2))
}
```

Run

```{r}
l <- lapply(1:1000, function(x) run_sim(x))
```

Get the mean and variance of the F values

```{r}
sapply(l, function(x) x$amod$F[2]) %>% mean
sapply(l, function(x) x$amod$F[2]) %>% var
```

What about two trans effects? Expect the mean and variance of the F value to be 1 and 2, respectively.

```{r}
run_sim2 <- function(i, n=1000)
{
	set.seed(i)
	dat <- tibble(
		y1=rbinom(n, 1, 0.5),
		y2=rbinom(n, 1, 0.5),
		y3=rbinom(n, 1, 0.5)
	)
	mod1 <- lm(y1 ~ y2 + y3, dat)
	mod2 <- lm(y1 ~ y2 + y3 + y2*y3, dat)
	amod <- anova(mod1, mod2)
	vare2 <- var(residuals(mod2))
	sse1 <- sum(residuals(mod1)^2)
	sse2 <- sum(residuals(mod2)^2)
	return(list(amod=amod, vare2=vare2, sse1=sse1, sse2=sse2))
}
l2 <- lapply(1:1000, function(x) run_sim2(x))
sapply(l2, function(x) x$amod$F[2]) %>% mean
sapply(l2, function(x) x$amod$F[2]) %>% var
```

Peter wrote that

$$
\begin{aligned}
F &= \frac{SSE_1 - SSE_2}{SSE_2 / (n-4)} \\
  &= \frac{Q}{var(e_2)} 
\end{aligned}
$$

where in the case of no interaction we expect

$$
\begin{aligned}
SSE_2 / (n-4) \approx var(e1) = var(e2) &= var(y_1)(1 - r^2) \\
&= p_1(1-p_1)(1 - r^2)
\end{aligned}
$$

and $Q = SSE_1 - SSE_2$. Test this:

```{r}
dat <- tibble(
	sim = 1:1000,
	F = sapply(l, function(x) x$amod$F[2]),
	vare2 = sapply(l, function(x) x$vare2),
	sse1 = sapply(l, function(x) x$sse1),
	sse2 = sapply(l, function(x) x$sse2)
)
dat2 <- tibble(
	sim = 1:1000,
	F = sapply(l2, function(x) x$amod$F[2]),
	vare2 = sapply(l2, function(x) x$vare2),
	sse1 = sapply(l2, function(x) x$sse1),
	sse2 = sapply(l2, function(x) x$sse2)
)
```

For the cis-trans association:

```{r}
plot(sse2 ~ sse1, dat)
```

For the trans-trans association:

```{r}
plot(sse2 ~ sse1, dat2)
```

Checking the equation

```{r}
plot((dat$sse1-dat$sse2)/dat$vare2, dat$F)
```

This works great but note that in the original document it was $SSE_2 - SSE_1$ which should be switched.

## Interaction test

Let the genotypic value of $y_{1.ij} = (y_1 = 1 | y_2 = i, y_3 = j)$ where $i,j \in \{0,1\}$, and the counts for each value be

$$
n_{ij} = n ( 1- p_2 + i(2p_2) - 1))(1 - p_3 + i(2p_3-1))
$$

We can therefore calculate the numbers in the 2 x 2 combinations of $y_2$ and $y_3$:

$$
\begin{aligned}
y_2\; & y_3 & n_{ij}/n \\
0\; & 0 & (1 - p_2)(1 - p_3) \\
0\; & 1 & p_2(1 - p_3) \\
1\; & 0 & (1 - p_2)p_3 \\
1\; & 1 & p_2p_3 \\
\end{aligned}
$$

A test for interaction can be defined as

$$
\delta = mean(y_{1.11}) + mean(y_{1.00}) - mean(y_{1.10}) - mean(y_{1.01})
$$

with $mean(y_{1.ij}) = \sum y_{1.ij} / n_{ij}$.

Check this. First generate a dataset with a strong interaction

```{r}
set.seed(sapply(l, function(x) x$amod$F[2]) %>% which.max)
dat <- gendat(1000)
mod1 <- lm(y1 ~ y2 + y3, dat)
mod2 <- lm(y1 ~ y2 + y3 + y2*y3, dat)
amod <- anova(mod1, mod2)

SSE1 <- sum(residuals(mod1)^2)
SSE2 <- sum(residuals(mod2)^2)
pchisq((SSE1 - SSE2) / (SSE2 / 996), 1, low=F)
pf((SSE1 - SSE2) / (SSE2 / 996), 1, 996, low=F)
```

Now create one with a weak interaction:

```{r}
set.seed(sapply(l, function(x) x$amod$F[2]) %>% which.min)
datw <- gendat(1000)
modw1 <- lm(y1 ~ y2 + y3, datw)
modw2 <- lm(y1 ~ y2 + y3 + y2*y3, datw)
amodw <- anova(modw1, modw2)
```

Test for interaction:

```{r}
est_delta <- function(dat)
{
	n <- nrow(dat)
	obs = table(dat$y1, dat$y2) / nrow(dat) %>% c	
	tab <- with(dat, tapply(y1, list(y2, y3), mean)) %>% c
	return(tibble(
		y1.00 = tab[1],
		y1.10 = tab[2],
		y1.01 = tab[3],
		y1.11 = tab[4],
		delta = y1.00 + y1.11 - y1.10 - y1.01,
		p1 = mean(dat$y1),
		p2 = mean(dat$y2),
		p3 = mean(dat$y3),
		p00 = obs[1],
		p10 = obs[2],
		p01 = obs[3],
		p11 = obs[4],
		D = obs[4] - mean(dat$y1) * mean(dat$y2)
	))
}

# strong interaction
est_delta(dat)

# weak interaction
est_delta(datw)
```

In the linear model the error variance is assumed to be the same in each cell of the 2 x 2 table and a pooled estimate is used.

Variance of $\delta$ can be calculated from the allele and haplotype frequencies and compared to its variance under a linear model (LM). Summarise each of 1000 simulations for cis-trans and trans-trans examples:

```{r}
# cis-trans
run_sim3 <- function(i, n=1000)
{
	set.seed(i)
	dat <- gendat(n)
	return(est_delta(dat))
}
m <- lapply(1:1000, function(i) run_sim3(i)) %>% bind_rows()

# trans-trans
run_sim4 <- function(i, n=1000)
{
	set.seed(i)
	dat <- tibble(
		y1=rbinom(n, 1, 0.5),
		y2=rbinom(n, 1, 0.5),
		y3=rbinom(n, 1, 0.5)
	)
	return(est_delta(dat))
}
m2 <- lapply(1:1000, function(i) run_sim4(i)) %>% bind_rows()
```

From the given haplotype frequencies:

$$
\begin{aligned}
E(y_{1.11}) = E(y_{1.10}) &= p(y_1 = 1 | y_2 = 1) \\
                          &= p_{11} / p2 \\
                          &= p_1 + D / p_2
\end{aligned}
$$

Check:

```{r}
mean(m$y1.11)
mean(m$y1.10)
(m$p11/m$p2)[1]
(m$p1 + m$D/m$p2)[1]
```

and similarly

$$
\begin{aligned}
E(y_{1.00}) = E(y_{1.01}) &= p(y_1 = 0 | y_2 = 0) \\
                          &= p_{00} / p2 \\
                          &= p_1 - D/(1 - p_2)
\end{aligned}
$$

Check:

```{r}
mean(m$y1.00)
mean(m$y1.01)
(m$p1 - m$D/(1-m$p2))[1]
```

Each of the terms has a binomial variance

$$
var(mean(y_{1.ij})) = E(y_{1.ij})(1 - E(y1.ij)) / n_{ij}
$$

Putting all the terms together gives the exact variance of the test statistic as

$$
var(\delta) = 
\frac{p_1(1-p_1) + \frac{(1-2p_1)(1-2p_2)D}{p2(1-p2)} - \frac{D^2(1-3p_2(1-p_2))}{p_2^2(1-p_2])^2} }
{n p_2(1-p_2)p_3(1-p_3)}
$$

Check:

```{r}

# Variance of mean(y_1.11)
(sd(m$y1.11) / sqrt(nrow(m)))^2

# According to the formula
mean(m$y1.11) * (1 - mean(m$y1.11)) / mean((m$p2) * (m$p3) * 1000)
```

**Don't seem to be getting agreement here.**

The general idea that the variance of each term is differentseems to be correct though, as the variance of $var(mean(y_{1.ij}))$ is not consistent over $i$ and $j$

```{r}
(sd(m$y1.11) / sqrt(nrow(m)))^2
(sd(m$y1.10) / sqrt(nrow(m)))^2
(sd(m$y1.01) / sqrt(nrow(m)))^2
(sd(m$y1.00) / sqrt(nrow(m)))^2
```

## Other aspects

### Adjusting for additive effects with varying degrees of measurement error

```{r}
gendatp <- function(n, p1, p2, p3, r1, shufprop)
{
	dat <- simulateGP:::simulate_haplotypes(n, r1, p1, p2) %>% as_tibble %>% dplyr::select(y1=A1, y2=B1)
	dat$y3 <- rbinom(n, 1, p3)
	dat$y4 <- dat$y1
	index <- sample(1:nrow(dat), shufprop * nrow(dat), replace=FALSE)
	dat$y4[index] <- sample(dat$y4[index])
	return(dat)
}

run_simp <- function(param, i)
{
	set.seed(i*10)
	dat <- gendatp(param$n[i], param$p1[i], param$p2[i], param$p3[i], param$r1[i], param$shufprop[i])
	# dat$y1r <- residuals(lm(y1 ~ y4, dat))
	x <- dat$y1 + rnorm(nrow(dat), sd=sd(dat$y1)/2)
	mod1 <- lm(x ~ y4 + y2 + y3, dat)
	mod2 <- lm(x ~ y4 + y2 + y3 + y2*y3, dat)
	amod <- anova(mod1, mod2)
	param$r2[i] <- cor(dat$y1, dat$y4)
	param$F[i] <- amod$F[2]
	return(param[i,])
}

param <- expand.grid(
	p1=0.1,
	p2=0.1,
	p3=0.5,
	p4=0.1,
	shufprop=seq(0, 1, by=0.1),
	n=1000,
	r1=seq(0, 1, by=0.2),
	sim=1:500,
	r2=NA,
	F=NA
)

resp <- lapply(1:nrow(param), function(x) run_simp(param, x)) %>% bind_rows()
```

Plot

```{r}
resp %>%
group_by(r1, shufprop) %>%
summarise(n=n(), mF=mean(F, na.rm=T), vF=var(F, na.rm=T), r2=mean(r2, na.rm=T)) %>% as.data.frame
resp$R2 <- 1 - resp$shufprop
ggplot(resp, aes(x=as.factor(R2), y=F)) +
geom_boxplot(aes(fill=as.factor(r1))) +
scale_fill_brewer(type="seq") +
labs(y="Interaction F value", x="Measurement precision of causal additive variant", fill="LD between tagging\nvariant and causal variant")
```

## Point estimation

A model where the average additive effect is different from the per-individual additive effect. The error term in the model is now a mixture of variance not explained by the SNP and heterogeneity in the effect of the SNP across the population.

```{r}
simulate_phen <- function(x, meanb, varb, r2)
{
	n <- length(x)
	b <- rnorm(n, meanb, sqrt(varb))
	yp <- x * b
	myb <- x * meanb
	vhet <- var(yp) - var(x * meanb)
	r2het <- cor(myb, yp)^2
	ve <- (var(myb) - r2 * var(myb)) / r2
	y <- yp + rnorm(n, 0, sqrt(ve - vhet))
	m <- try(lm(y ~ x)$coef[2])
	m <- ifelse(class(m) == 'try-error', NA, m)
	return(
		list(
			y=y,
			info=tibble(
				varb=varb,
				r2=r2,
				r2het=r2het,
				rsqxy=cor(x,y)^2,
				ve=ve,
				vy=var(y),
				vyp=var(yp),
				vhet=vhet,
				bhat=m
			)
		)	
	)
}

x <- rbinom(10000, 1, 0.5)
out <- expand.grid(meanb=1, varb=seq(0.0, 0.5, by=0.02), r2=seq(0.1, 1, by=0.1)) %>%
	split(., 1:nrow(.)) %>%
	lapply(., function(i) simulate_phen(x, i$meanb, i$varb, i$r2)$info) %>% bind_rows()

plot(rsqxy ~ r2, out)
ggplot(out, aes(x=vyp, y=ve-vhet)) +
geom_point() +
facet_wrap(~ r2 )

```

Run simulations to see how point estimation heterogeneity impacts F statistic in interaction test.

```{r}
run_simp2 <- function(param, i)
{
	set.seed(i)
	dat <- gendatp(param$n[i], param$p1[i], param$p2[i], param$p3[i], param$r1[i], param$shufprop[i])
	o <- simulate_phen(dat$y1, param$meanb[i], param$varb[i], param$r2add[i])
	x <- o$y
	info <- o$info
	mod1 <- lm(x ~ y1 + y3, dat)
	mod2 <- lm(x ~ y1 + y3 + y1*y3, dat)
	amod <- anova(mod1, mod2)
	param$r2[i] <- cor(dat$y1, x)
	param$F[i] <- amod$F[2]
	param$ve <- info$ve
	param$vhet <- info$vhet
	param$rsqxy <- info$rsqxy
	param$vy <- info$vy

	return(param[i,])
}

param <- expand.grid(
	p1=0.1,
	p2=0.1,
	p3=0.5,
	p4=0.1,
	shufprop=0,
	r2add=0.5,
	meanb=1,
	varb=seq(0, 0.4, by=0.01),
	n=1000,
	r1=0.5,
	sim=1:1000,
	r2=NA,
	F=NA
)

res <- lapply(1:nrow(param), function(x) run_simp2(param, x)) %>% bind_rows()

res %>%
group_by(r1, varb) %>%
summarise(
	n=n(), 
	mF=mean(F, na.rm=T), 
	vF=var(F, na.rm=T), 
	r2=mean(r2, na.rm=T), 
	ve=mean(ve, na.rm=T), 
	vhet=mean(vhet, na.rm=T), 
	rsqxy=mean(rsqxy, na.rm=T),
	vy=mean(vy, na.rm=T)
) %>% as.data.frame
res$vhetvy <- res$vhet/res$vy
res$vhetvy[res$vhetvy < 0] <- 0
res$vhetvy <- cut(res$vhetvy, breaks=10)
ggplot(res, aes(x=varb, y=F)) +
geom_boxplot(aes(group=as.factor(varb), fill=varb==0)) +
scale_fill_brewer(type="seq") +
labs(y="Interaction F value", x="Proportion of error variance due to point estimation error", fill="No inflation") +
theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))

```

### 2019-12-26 null_replication/docs/analyse_results.rmd
updated

@@
 ---
+
+```{r}
+suppressWarnings(suppressPackageStartupMessages({
+	library(knitr)
+	library(tidyverse)
+}))
+opts_chunk$set(cache=TRUE)
+```
 
@@
 ```{r}
-library(tidyverse)
 load("../data/rs67903230.rdata")
@@
 ```{r}
-plot(lambda_disc ~ lambda_repl, res)
+ggplot(res, aes(y=lambda_repl, x=lambda_disc)) +
+geom_point() +
+labs(x = "Genomic inflation (discovery)", y = "Genomic inflation (replication)")
 ```
@@
 geom_smooth(aes(colour=polygenic))
+```
+
+```{r}
+ggplot(res, aes(x=varexp, y=lambda_disc)) +
+geom_point() +
+labs(x="Additive variance explained", y="Genomic inflation")
 ```
@@
 ```{r}
-ggplot(res, aes(x=varexp, y=nsig_disc_bonf)) +
-geom_point()
+res$varexp_bin <- Hmisc::cut2(res$varexp, g=30)
+ggplot(res, aes(x=varexp_bin, y=nsig_disc_bonf)) +
+geom_boxplot() +
+theme(axis.text.x=element_text(angle=90, hjust=0.5, vjust=0.5)) +
+labs(x="Additive variance explained", y="Mean number of discovery hits")
 ```
@@
 hist(res$fstat_cor)
+```
+
+```{r}
+ggplot(res, aes(x=fstat_cor)) +
+geom_histogram(fill="white", colour="black") +
+labs(x="Correlation of F statistics between discovery and replication")
+
 ```
@@
 res$lambda_disc_bin <- Hmisc::cut2(res$lambda_disc, g=30)
-group_by(res, lambda_disc_bin) %>% summarise(n=mean(nsig_disc_bonf, na.rm=T)) %>% 
-ggplot(aes(x=lambda_disc_bin, y=n)) +
-geom_point() +
+res %>% 
+ggplot(aes(x=lambda_disc_bin, y=nsig_disc_bonf)) +
+geom_boxplot() +
 theme(axis.text.x=element_text(angle=90, hjust=0.5, vjust=0.5)) +
-labs(x="Simulation's lambda", y="Mean number of discovery hits")
+labs(x="Genomic inflation of simulation", y="Mean number of discovery hits")
 ```
@@
 theme(axis.text.x=element_text(angle=90, hjust=0.5, vjust=0.5)) +
-labs(x="Simulation's lambda", y="Replication rate")
+labs(x="Genomic inflation factor", y="Replication rate")
 ```

### 2020-01-21 null_replication/scripts/analyse_data_scen1.sh
updated

@@ -81,6 +81,6 @@ echo "analysing"
 Rscript formatres.r ${sd}/${sim}_disc_out ${sd}/${sim}_rep_out ${varexp} ${code} ${sim} ${cissnp} ${sensnp} ${sd}/${sim}.scen1.rdata ${sd}/${cissnp}_exclude_disc.bim.orig
 echo "done analysing"
 
-rm ${sd}/${sim}_*
+# rm ${sd}/${sim}_*
 
 

### 2020-01-21 null_replication/scripts/formatres.r

@@ -14,11 +14,13 @@ outfile <- args[8]
 mapfile <- args[9]
 
 
-disc <- fread(discfile)  %>% as_tibble()
+disc <- fread(discfile)  %>% as_tibble() %>% filter(V3 == 8)
+print(nrow(disc))
 disc$p4 <- pf(disc$V6, 4, disc$V4, low=FALSE)
 disc$fdr <- p.adjust(disc$p4, "fdr")
 
-repl <- fread(replfile) %>% as_tibble()
+repl <- fread(replfile) %>% as_tibble() %>% filter(V3 == 8)
+print(nrow(repl))
 repl$p4 <- pf(repl$V6, 4, repl$V4, low=FALSE)
 repl$fdr <- p.adjust(repl$p4, "fdr")
 

### 2020-01-21 null_replication/docs/analyse_results.rmd

--- 2019-09-30 null_replication/docs/analyse_results.rmd
@@
 ```{r}
-library(tidyverse)
-load("../data/rs67903230.rdata")
-res$polygenic <- res$i %% 2 == 0
+l <- list()
+# load("../data/rs67903230.rdata")
+# res$polygenic <- res$sim %% 2 == 0
+# res$median_p_disc <- pchisq(res$lambda_disc * qchisq(0.5, 1), 1, low=FALSE)
+# res$median_p_repl <- pchisq(res$lambda_repl * qchisq(0.5, 1), 1, low=FALSE)
+# res$lambda_disc <- pnorm(1 - res$median_p_disc / 2)^2 / qchisq(0.5, 1)
+# res$lambda_repl <- pnorm(1 - res$median_p_repl / 2)^2 / qchisq(0.5, 1)
+# res$gene <- "MBNL1"
+# res$i <- paste(res$gene, res$sentinel, res$cis, sep="_")
+# res$scenario <- 1
+
+load("../data/aggregate1.rdata")
+res$polygenic <- res$sim %% 2 == 0
+res$gene <- sapply(res$i, function(x) strsplit(x, "_")[[1]][1])
+res$scenario <- 1
+l[[1]] <- res
+
+load("../data/aggregate.rdata")
+res$polygenic <- res$sim %% 2 == 0
+res$gene <- sapply(res$i, function(x) strsplit(x, "_")[[1]][1])
+res$scenario <- 2
+l[[2]] <- res
+
+load("../data/aggregate2.rdata")
+res$polygenic <- res$sim %% 2 == 0
+res$gene <- sapply(res$i, function(x) strsplit(x, "_")[[1]][1])
+res$scenario <- 3
+l[[3]] <- res
+
+res <- bind_rows(l)
+
 ```
@@
 ```{r}
-hist(res$varexp)
+ggplot(res, aes(x=varexp)) +
+geom_histogram() +
+facet_grid(scenario ~ .)
 ```
@@
 ```{r}
-plot(lambda_disc ~ lambda_repl, res)
+ggplot(res, aes(y=lambda_repl, x=lambda_disc)) +
+geom_point() +
+labs(x = "Genomic inflation (discovery)", y = "Genomic inflation (replication)") +
+facet_grid(scenario ~ .)
 ```
@@
 ggplot(res, aes(x=varexp, y=lambda_disc)) +
-geom_point(aes(colour=polygenic)) +
-geom_smooth(aes(colour=polygenic))
+geom_point(size=0.03, alpha=0.1) +
+labs(x="Additive variance explained", y="Genomic inflation") +
+facet_grid(scenario ~ .) +
+theme_bw()
 ```
@@
 ggplot(res, aes(x=nsig_disc_bonf)) +
-geom_histogram(aes(fill=polygenic), position="dodge")
+geom_histogram(aes(fill=polygenic), position="dodge") +
+facet_grid(scenario ~ .)
 ```
@@
 ```{r}
-ggplot(res, aes(x=varexp, y=nsig_disc_bonf)) +
-geom_point()
+res$varexp_bin <- Hmisc::cut2(res$varexp, g=30)
+ggplot(res, aes(x=varexp_bin, y=nsig_disc_bonf)) +
+geom_boxplot() +
+theme(axis.text.x=element_text(angle=90, hjust=0.5, vjust=0.5)) +
+labs(x="Additive variance explained", y="Mean number of discovery hits") +
+facet_grid(scenario ~ .)
 ```
@@
 hist(res$fstat_cor)
+```
+
+```{r}
+ggplot(res, aes(x=fstat_cor)) +
+geom_histogram(fill="white", colour="black") +
+labs(x="Correlation of F statistics between discovery and replication") +
+facet_grid(scenario ~ ., scale="free_y")
 ```
@@
 res$lambda_disc_bin <- Hmisc::cut2(res$lambda_disc, g=30)
-group_by(res, lambda_disc_bin) %>% summarise(n=mean(nsig_disc_bonf, na.rm=T)) %>% 
-ggplot(aes(x=lambda_disc_bin, y=n)) +
-geom_point() +
+res$varexp_bin <- Hmisc::cut2(res$varexp, g=30)
+res %>% 
+ggplot(aes(x=lambda_disc_bin, y=nsig_disc_bonf)) +
+geom_boxplot() +
 theme(axis.text.x=element_text(angle=90, hjust=0.5, vjust=0.5)) +
-labs(x="Simulation's lambda", y="Mean number of discovery hits")
+labs(x="Genomic inflation of simulation", y="Mean number of discovery hits") +
+facet_grid(scenario ~ .)
+
+res %>% 
+ggplot(aes(x=varexp_bin, y=nsig_disc_bonf)) +
+geom_boxplot() +
+theme(axis.text.x=element_text(angle=90, hjust=0.5, vjust=0.5)) +
+labs(x="Variance explained by cis causal variant", y="Mean number of discovery hits") +
+facet_grid(scenario ~ .)
 ```
@@
 
-```{r}
+```{r, fig.width=12}
 res$bonf_bonf_rr <- res$cond_bonf_bonf / res$nsig_disc_bonf
@@
 repl_rate <- bind_rows(
-	group_by(res, lambda_disc_bin) %>% summarise(m=mean(bonf_bonf_rr, na.rm=T)) %>% mutate(threshold="Bonferroni"),
-	group_by(res, lambda_disc_bin) %>% summarise(m=mean(bonf_fdr_rr, na.rm=T)) %>% mutate(threshold="FDR"),
-	group_by(res, lambda_disc_bin) %>% summarise(m=mean(bonf_bonf2_rr, na.rm=T)) %>% mutate(threshold="Experiment")
+	group_by(res, varexp_bin, scenario) %>% summarise(m=mean(bonf_bonf_rr, na.rm=T)) %>% mutate(threshold="Bonferroni"),
+	group_by(res, varexp_bin, scenario) %>% summarise(m=mean(bonf_fdr_rr, na.rm=T)) %>% mutate(threshold="FDR"),
+	group_by(res, varexp_bin, scenario) %>% summarise(m=mean(bonf_bonf2_rr, na.rm=T)) %>% mutate(threshold="Experiment")
 )
-repl_rate %>% ggplot(aes(x=lambda_disc_bin, y=m, group=threshold)) +
+repl_rate %>% ggplot(aes(x=varexp_bin, y=m, group=threshold)) +
 geom_line(aes(colour=threshold)) +
@@
 theme(axis.text.x=element_text(angle=90, hjust=0.5, vjust=0.5)) +
-labs(x="Simulation's lambda", y="Replication rate")
+labs(x="Variance explained by cis additive variant", y="Replication rate") +
+facet_grid(. ~ scenario) +
+ylim(c(0,0.3))
 ```

### 2020-01-23 null_replication/docs/analyse_results.rmd
updated

@@
 
-load("../data/aggregate.rdata")
+load("../data/aggregate2.rdata")
 res$polygenic <- res$sim %% 2 == 0
@@
 
-load("../data/aggregate2.rdata")
+load("../data/aggregate3.rdata")
 res$polygenic <- res$sim %% 2 == 0
@@
 labs(x="Variance explained by cis additive variant", y="Replication rate") +
-facet_grid(. ~ scenario) +
-ylim(c(0,0.3))
+facet_grid(. ~ scenario)
 ```
@@
 But the experiment wide threshold can only be met when the discovery lambda is very high, but even then the replication rate is extremely low.
+
+## Replication rate of MBNL1
+
+Analysing the MBNL1 locus - what is the discovery and replication rate?
+
+There are 40000 simulations, where the cis-additive variant explains some amount of the variance in the phenotype (ranging from 0% to 50%).
+If a signal is found to be significant then it is tested for replication.
+In this plot, the x axis shows how many independent associations found to be significant for a simulations, and the y axis is the proportion of those that replicate at p < 0.05/501.
+The simulations are binned based on the simulated additive variance (columns of plots).
+The blue dots are from the simulations, and the red dot is the empirical result for MBNL1.
+Note that the cis-additive variant explains 10.5% of the variance of MBNL1, so the relevent boxes for comparison against the simulations are the second and third from the left.
+There are very few discovered loci here, and the replication rate of those is always zero.
+When the additive locus explains more variance, then the number of discovered hits increases and the replication rate increases.
+
+```{r, fig.width=12}
+temp <- subset(res, scenario == 1)
+
+temp$vbin <- cut(temp$varexp, breaks=10)
+o <- group_by(temp, vbin, cond_bonf_bonf2, nsig_disc_bonf) %>%
+summarise(n=n())
+o$what <- "simulation"
+
+e <- tibble(cond_bonf_bonf2=5, nsig_disc_bonf=11, n=1, what="empirical", vbin = unique(temp$vbin))
+oe <- bind_rows(o,e)
+oe$rr <- oe$cond_bonf_bonf2 / oe$nsig_disc_bonf
+oe$rr[is.na(oe$rr)] <- 0
+
+ggplot(oe, aes(x=nsig_disc_bonf, y=rr)) +
+geom_point(aes(color=what, size=n)) +
+labs(x="Number discovered", y="Proportion replicated") +
+facet_grid(. ~ vbin)
+```

### 2020-01-26 null_replication/docs/analyse_results.rmd
updated

@@
 
+First check the distribution under the null.
+
+```{r}
+library(dplyr)
+load("../data/aggregate4.rdata")
+
+hist(res$lambda_disc, breaks=40)
+hist(res$lambda_disc2, breaks=40)
+
+# chi(1) version of lambda
+mean(res$lambda_disc)
+sd(res$lambda_disc) / sqrt(nrow(res))
+
+# chi(4) version of lambda
+mean(res$lambda_disc2)
+sd(res$lambda_disc2) / sqrt(nrow(res))
+```
+
+Can I convert the chi(1) lambda to the chi(4) lambda?
+
+```{r}
+res$lambda_disc
+
+res$median_p_disc <- pchisq(res$lambda_disc * qchisq(0.5, 1), 1, low=FALSE)
+res$median_f4_disc <- qf(res$median_p_disc, 4, 840, low=FALSE)
+res$lambda_disc2_2 <- res$median_f4_disc / qchisq(0.5, 4, low=FALSE) * 4
+
+plot(lambda_disc ~ lambda_disc2, res)
+plot(lambda_disc2_2 ~ lambda_disc2, res)
+```
+
+Read in main results
+
 ```{r}
 l <- list()
-# load("../data/rs67903230.rdata")
-# res$polygenic <- res$sim %% 2 == 0
-# res$median_p_disc <- pchisq(res$lambda_disc * qchisq(0.5, 1), 1, low=FALSE)
-# res$median_p_repl <- pchisq(res$lambda_repl * qchisq(0.5, 1), 1, low=FALSE)
-# res$lambda_disc <- pnorm(1 - res$median_p_disc / 2)^2 / qchisq(0.5, 1)
-# res$lambda_repl <- pnorm(1 - res$median_p_repl / 2)^2 / qchisq(0.5, 1)
-# res$gene <- "MBNL1"
-# res$i <- paste(res$gene, res$sentinel, res$cis, sep="_")
-# res$scenario <- 1
 
@@
 res <- bind_rows(l)
+res$median_p_disc <- pchisq(res$lambda_disc * qchisq(0.5, 1), 1, low=FALSE)
+res$median_f4_disc <- qf(res$median_p_disc, 4, 840, low=FALSE)
+res$lambda_disc <- res$median_f4_disc / qchisq(0.5, 4, low=FALSE) * 4
+res$median_p_repl <- pchisq(res$lambda_repl * qchisq(0.5, 1), 1, low=FALSE)
+res$median_f4_repl <- qf(res$median_p_repl, 4, 2112, low=FALSE)
+res$lambda_repl <- res$median_f4_repl / qchisq(0.5, 4, low=FALSE) * 4
 

### 2020-01-27 null_replication/scripts/mbnl1_analysis.rmd
upd

---
title: Example output from simulations for MBNL1 where Va = 0
---

Perform 4df test of rs13069559 against 500k SNPs

```{r}
suppressPackageStartupMessages(suppressWarnings({
	library(dplyr)
	library(data.table)
	library(parallel)
	library(knitr)
}))
opts_chunk$set(cache=TRUE)
```

```{r}
a <- fread("../data/scratch/MBNL1_rs67903230_rs13069559/18778_disc_out")
names(a) <- c("snp1", "snp2", "df1", "df2", "fast_f8", "fast_f4")

b <- subset(a, df1 == 8)
b$fast_p4 <- pf(b$fast_f4, 4, b$df2, low=FALSE)
hist(b$fast_p4)
```

Replicate using native R statistical tests

```{r}
map <- fread("temp.map")
snplist <- c(map$V2[1], map$V2[-1][b$snp2])
write.table(snplist, file="snplist8df.txt", row=F, col=F, qu=F)
```

```{r engine="bash"}
episcan -De temp.ped temp.map ../data/scratch/MBNL1_rs67903230_rs13069559/rs13069559_merge_disc.egu
plink --file temp --extract snplist8df.txt --recode A --out temp
```

Read in the data, calculate the 4df F values

```{r}
ped <- fread("temp.raw", he=T)
fam <- fread("../data/scratch/MBNL1_rs67903230_rs13069559/18778_disc.fam")
mat <- ped[,-c(1:6)] %>% as.data.frame()

out <- mclapply(2:ncol(mat), function(x)
{
	mod <- anova(
		lm(fam$V6 ~ as.factor(mat[,1]) + as.factor(mat[,x])),
		lm(fam$V6 ~ as.factor(mat[,1]) : as.factor(mat[,x]))
	)
	return(mod$F[2])

}, mc.cores=16)

b$r_f4 <- unlist(out)
b$r_p4 <- pf(b$r_f4, 4, b$df2, low=FALSE)
```

Distribution of p-values from R version of F test

```{r}
hist(b$r_p4)
```

Correlation between fast F test and R version

```{r}
cor(b$fast_f4, b$r_f4)
```

Correlation between fast and R versions of the p-values

```{r}
cor(b$fast_f4, b$r_f4)
```

Range of p-values

```{r}
range(b$fast_p4)
range(b$r_p4)
```

Lambda values

```{r}
median(b$fast_f4) / qchisq(0.5, 4) * 4
median(b$r_f4) / qchisq(0.5, 4) * 4
```

The fast version looks a bit more conservative. This is probably because when LD isn't accounted for between the two variants in the fast model, it slightly overestimates the additive variance and the residual non-additive variance is smaller.

### 2020-01-27 null_replication/scripts/analyse_data_scen1.sh

@@ -24,7 +24,7 @@ disc="../data/disc"
 rep="../data/rep"
 
 code="MBNL1_rs67903230_rs13069559"
-IFS='_' read -r -a array <<< ${code}
+IFS='_' read -r -a array <<< "$code"
 gene="${array[0]}"
 sensnp="${array[1]}"
 cissnp="${array[2]}"

### 2020-01-27 null_replication/scripts/analyse_data_scen2.sh

@@ -24,7 +24,7 @@ disc="../data/disc"
 rep="../data/rep"
 
 code="$1"
-IFS='_' read -r -a array <<< ${code}
+IFS='_' read -r -a array <<< "$code"
 gene="${array[0]}"
 sensnp="${array[1]}"
 cissnp="${array[2]}"

### 2020-01-27 null_replication/scripts/analyse_data_scen4.sh

@@ -24,7 +24,7 @@ disc="../data/disc"
 rep="../data/rep"
 
 code="$1"
-IFS='_' read -r -a array <<< ${code}
+IFS='_' read -r -a array <<< "$code"
 gene="${array[0]}"
 sensnp="${array[1]}"
 cissnp="${array[2]}"
@@ -77,6 +77,6 @@ echo "analysing"
 Rscript formatres.r ${sd}/${sim}_disc_out ${sd}/${sim}_rep_out ${varexp} ${code} ${sim} ${cissnp} ${sensnp} ${sd}/${sim}.scen4.rdata ${sd}/${cissnp}_exclude_disc.bim.orig
 echo "done analysing"
 
-rm ${sd}/${sim}_*
+# rm ${sd}/${sim}_*
 
 

### 2020-01-27 null_replication/scripts/create_data.sh

@@ -6,7 +6,7 @@
 
 datafile="../data/combined"
 code="$1"
-IFS='_' read -r -a array <<< ${code}
+IFS='_' read -r -a array <<< "$code"
 
 gene="${array[0]}"
 sensnp="${array[1]}"

### 2020-01-27 null_replication/scripts/mbnl1_analysis.rmd
upd

@@
 This is probably because when LD isn't accounted for between the two variants in the fast model, it slightly overestimates the additive variance and the residual non-additive variance is smaller.
+
+## Other plots
+
+Histograms of F values for fast and R versions
+
+```{r}
+hist(b$r_f4)
+hist(b$fast_f4)
+```
+
+Is the rs13069559 more correlated with the rest of the genome than expected by chance?
+
+Summary of rs13069559
+
+```{r}
+r_rs13069559 <- cor(mat[,1], mat[,-1])
+median(r_rs13069559)
+hist(r_rs13069559)
+```
+
+Choose some random snps, get median LD r of each SNP against all other SNPs
+
+```{r}
+set.seed(1)
+snps <- sample(1:ncol(mat), 32, replace=FALSE)
+
+rval <- mclapply(1:length(snps), function(x)
+{
+	r_rand <- cor(mat[,snps[x]], mat[,-snps[x]])
+	return(median(r_rand))
+}, mc.cores=16)
+```
+
+Plot median r values
+
+```{r}
+library(ggplot2)
+o <- data.frame(what = c("rs13069559", rep("random", length(snps))), median_r = c(median(r_rs13069559), unlist(rval))) %>%
+	arrange(median_r) %>%
+	mutate(ord = 1:n())
+
+ggplot(o, aes(x=ord, y=median_r)) +
+geom_point(aes(colour=what))
+```
+
+Looks like the median LD across the genome of rs13069559 is pretty high.

### 2020-01-28 null_replication/scripts/mbnl1_analysis.rmd
upd

@@
 mat <- ped[,-c(1:6)] %>% as.data.frame()
+```
+
+Allele frequency of rs13069559:
+
+```{r}
+sum(mat[,1], na.rm=T) / (2 * sum(!is.na(mat[,1])))
+```
+
+```{r}
 
@@
 Looks like the median LD across the genome of rs13069559 is pretty high.
+
+## Restricting tests based on minimum cell count
+
+Find pairs where minimum cell number is 5 or more, as was used in the original paper
+
+```{r}
+getcells <- function(mincount)
+{
+	cells <- mclapply(2:ncol(mat), function(x)
+	{
+		tab <- table(mat[,1], mat[,x])
+		return(length(tab) == 9 & min(tab) >= mincount)
+	}, mc.cores=16)
+
+	cells <- unlist(cells)
+	return(cells)
+}
+
+cells5 <- getcells(5)
+cells3 <- getcells(3)
+table(cells5)
+table(cells3)
+```
+
+So, this drastically reduces the number of variants remaining in the analysis
+
+```{r}
+estlambda <- function(fval)
+{
+	median(fval) / qchisq(0.5, 4) * 4
+}
+
+estlambda(b$fast_f4)
+estlambda(b$r_f4)
+estlambda(b$fast_f4[cells3])
+estlambda(b$r_f4[cells3])
+estlambda(b$fast_f4[cells5])
+estlambda(b$r_f4[cells5])
+```

### 2020-01-31 null_replication/docs/analyse_results.rmd
upd

@@
 sd(res$lambda_disc2) / sqrt(nrow(res))
+
+group_by(res, cis, sentinel, i) %>%
+summarise(m=mean(lambda_disc2), v=sd(lambda_disc2)) %>%
+plot(m ~ v, .)
+
+library(ggplot2)
+ggplot(res, aes(y=lambda_disc2, x=paste(i, sentinel, cis))) +
+geom_boxplot()
+
 ```
